---
title: "Mathematical Statistics"
subtitle: "The Gamma, Exponential and Chi-Square Distributions"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - Gamma, Exponential and Chi-Square"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
library(ggplot2)
theme_set(theme_minimal(base_size = 18))
```

## 🎯 Learning Objectives

::: {style="font-size: 32px"}
By the end of this lecture, you will be able to:

- **State** the gamma density (Definition 4.9) and use $\Gamma(n) = (n-1)!$

- **Derive** $E(Y) = \alpha\beta$ and $V(Y) = \alpha\beta^2$ by recognising a gamma integral (Theorem 4.8)

- **Compute** gamma probabilities and quantiles in R, minding that R wants the rate $1/\beta$

- **Identify** the chi-square and exponential distributions as gamma special cases

- **Use** the memoryless property of the exponential, and say when it is a poor model
:::

---

## 🗺️ Where We Are

::: {style="font-size: 32px"}
**Wackerly §4.6**

Wednesday: the normal distribution, symmetric about $\mu$ and spread over the whole real line.

::: {.fragment}
Many quantities in finance are neither. A waiting time, a lifetime, a repair time cannot be negative, and most of their mass sits near zero with a long tail to the right.
:::

::: {.fragment}
Today's family is built for exactly that shape, and it contains two special cases we will use for the rest of the course: the **exponential** and the **chi-square**.
:::
:::

---

## ❓ Motivating Question

::: {style="font-size: 30px"}
On an interbank AZN/USD desk, the gap between two **large trades** (above 1 million USD) averages 25 minutes, with a standard deviation of about 25 minutes.

::: {.fragment}
Model the gap as normal with $\mu = \sigma = 25$ and it gives
$$P(Y < 0) = P(Z < -1) = 0.1587$$
**Sixteen percent of gaps would be negative.** The model is not slightly wrong; it is wrong in kind.
:::

::: {.fragment}
We need a density that lives on $[0, \infty)$, is skewed right, and still has a mean and a variance we can compute.
:::
:::

---

## 📐 Definition 4.9: The Gamma

::: {style="font-size: 30px"}
::: {.callout-important}
## Definition 4.9

$Y$ has a **gamma distribution** with parameters $\alpha > 0$ and $\beta > 0$ if and only if
$$f(y) = \begin{cases} \dfrac{y^{\alpha-1} e^{-y/\beta}}{\beta^{\alpha}\,\Gamma(\alpha)}, & 0 \le y < \infty, \\[4pt] 0, & \text{elsewhere,} \end{cases} \qquad \Gamma(\alpha) = \int_0^\infty y^{\alpha-1} e^{-y}\,dy.$$
:::

::: {.fragment}
$\alpha$ is the **shape** parameter; $\beta$ is the **scale** parameter. Multiplying $Y$ by a positive constant changes $\beta$ and leaves $\alpha$ alone.
:::
:::

---

## 🔢 The Gamma Function

::: {style="font-size: 30px"}
$\Gamma(\alpha)$ is only there to make the density integrate to 1. Three facts do all the work:

- $\Gamma(1) = 1$, by direct integration
- $\Gamma(\alpha) = (\alpha - 1)\,\Gamma(\alpha - 1)$ for $\alpha > 1$, by parts
- $\Gamma(n) = (n-1)!$ for a positive integer $n$

```{r}
#| label: gamma-fn
#| code-fold: false
gamma(c(1, 2, 4, 7))     # 0!, 1!, 3!, 6!
```

So the gamma function is the factorial, shifted by one and extended to every $\alpha > 0$.
:::

---

## 📈 What the Shape Parameter Does

```{r}
#| label: shape-figure
#| echo: false
#| fig-width: 10
#| fig-height: 4.6
yy <- seq(0, 10, by = 0.02)
shape_df <- rbind(
  data.frame(y = yy, f = dgamma(yy, shape = 1, rate = 1), a = "alpha == 1"),
  data.frame(y = yy, f = dgamma(yy, shape = 2, rate = 1), a = "alpha == 2"),
  data.frame(y = yy, f = dgamma(yy, shape = 4, rate = 1), a = "alpha == 4"))
ggplot(shape_df, aes(y, f, colour = a)) +
  geom_line(linewidth = 1.4) +
  scale_colour_manual(values = c("#14130f", "#3d6e8f", "#b3833a"),
                      labels = scales::label_parse()) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  scale_x_continuous(breaks = seq(0, 10, 2)) +
  labs(x = "y", y = "f(y)", colour = NULL,
       title = expression("Gamma densities with" ~ beta == 1 ~ "(Figure 4.16)")) +
  theme(legend.position = "top", legend.text = element_text(size = 20))
```

::: {style="font-size: 28px"}
At $\alpha = 1$ the mode is at zero; as $\alpha$ grows the peak moves right and the tail thins.
:::

---

## 📐 Theorem 4.8: Mean and Variance

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorem 4.8
If $Y$ has a gamma distribution with parameters $\alpha$ and $\beta$, then $\mu = E(Y) = \alpha\beta$ and $\sigma^2 = V(Y) = \alpha\beta^2$.
:::

::: {.fragment}
**The trick.** The density integrates to 1, so $\int_0^\infty y^{\alpha-1}e^{-y/\beta}\,dy = \beta^{\alpha}\Gamma(\alpha)$ for **every** $\alpha > 0$. Then
$$E(Y) = \frac{1}{\beta^{\alpha}\Gamma(\alpha)}\int_0^\infty y^{\alpha}e^{-y/\beta}\,dy = \frac{\beta^{\alpha+1}\Gamma(\alpha+1)}{\beta^{\alpha}\Gamma(\alpha)} = \alpha\beta.$$
:::

::: {.fragment}
The same step with $\alpha + 2$ gives $E(Y^2) = \alpha(\alpha+1)\beta^2$, so $V(Y) = \alpha\beta^2$.
:::
:::

---

## 🏢 Worked Example: Waiting for Claims

::: {style="font-size: 30px"}
A Baku property insurer tracks the time $Y$, in days, until its **third** large claim of the quarter. $Y$ is gamma with $\alpha = 3$, $\beta = 4$.

::: {.fragment}
**Moments.** $E(Y) = 3 \times 4 = 12$ days; $V(Y) = 3 \times 16 = 48$, so $\sigma = 6.93$ days.
:::

::: {.fragment}
**A tail probability.** With integer $\alpha$, Exercise 4.99 turns the gamma tail into a Poisson sum with mean $20/\beta = 5$:
$$P(Y > 20) = \sum_{x=0}^{2} \frac{5^x e^{-5}}{x!} = 18.5\,e^{-5} = 0.1247$$
:::

::: {.fragment}
One quarter in eight, the claims desk waits more than 20 days for its third large claim.
:::
:::

---

## 💻 The Same Numbers in R

```{r}
#| label: claims-r
#| code-fold: false
alpha <- 3; beta <- 4
c(mean = alpha * beta, var = alpha * beta^2)
1 - pgamma(20, shape = alpha, rate = 1 / beta)   # P(Y > 20)
sum(dpois(0:2, lambda = 20 / beta))              # Exercise 4.99 route
qgamma(0.95, shape = alpha, rate = 1 / beta)     # 95th percentile, days
```

::: {style="font-size: 28px"}
**The trap:** R's third argument is the **rate** $1/\beta$, as in the book's `pgamma(y0, α, 1/β)`. Pass $\beta$ there and the answer is silently wrong.
:::

---

## 📐 Definition 4.10: Chi-Square

::: {style="font-size: 30px"}
::: {.callout-important}
## Definition 4.10
Let $\nu$ be a positive integer. $Y$ has a **chi-square distribution with $\nu$ degrees of freedom** if and only if $Y$ is gamma with $\alpha = \nu/2$ and $\beta = 2$.
:::

::: {.callout-important}
## Theorem 4.9
If $Y$ is $\chi^2$ with $\nu$ degrees of freedom, then $E(Y) = \nu$ and $V(Y) = 2\nu$.
:::

::: {.fragment}
Proof: Theorem 4.8 with $\alpha = \nu/2$, $\beta = 2$. The name "degrees of freedom" is explained in Theorem 6.4; for now it is just the parameter.
:::
:::

---

## 🖥️ Gamma to Chi-Square: A Repair Time

::: {style="font-size: 30px"}
The time $Y$, in hours, to restore a failed rack in a bank's data centre is gamma with $\alpha = 1.5$, $\beta = 4$. What is $P(Y < 3.5)$?

::: {.fragment}
Exercise 6.46 will show: if $\alpha = n/2$, then $2Y/\beta$ is $\chi^2$ with $n$ degrees of freedom. Here $2Y/4 = Y/2$ is $\chi^2_3$, so
$$P(Y < 3.5) = P(Y/2 < 1.75) = P(\chi^2_3 < 1.75)$$
:::

```{r}
#| label: chisq-link
#| code-fold: false
c(gamma = pgamma(3.5, shape = 1.5, rate = 1/4), chisq = pchisq(1.75, df = 3))
```

About 37% of rack failures are fixed within 3.5 hours.
:::

---

## 📐 Definition 4.11: The Exponential

::: {style="font-size: 30px"}
::: {.callout-important}
## Definition 4.11
$Y$ has an **exponential distribution** with parameter $\beta > 0$ if and only if
$$f(y) = \begin{cases} \dfrac{1}{\beta}\, e^{-y/\beta}, & 0 \le y < \infty, \\[4pt] 0, & \text{elsewhere.} \end{cases}$$
:::

::: {.callout-important}
## Theorem 4.10
If $Y$ is exponential with parameter $\beta$, then $E(Y) = \beta$ and $V(Y) = \beta^2$.
:::

::: {.fragment}
It is the gamma with $\alpha = 1$, and the one member whose distribution function has a closed form: $P(Y > y) = e^{-y/\beta}$.
:::
:::

---

## 💾 Worked Example: Drive Lifetimes

::: {style="font-size: 30px"}
Solid-state drives in a data centre fail after an exponential lifetime with mean $\beta = 5$ years.

::: {.fragment}
**Survival past the warranty.** $P(Y > 3) = e^{-3/5} = 0.5488$. Just over half the drives outlast a 3-year warranty.
:::

::: {.fragment}
**An early-replacement threshold.** Which age $t$ do only 10% of drives fail before?
$$1 - e^{-t/5} = 0.10 \;\Rightarrow\; t = -5\ln(0.90) = 0.527 \text{ years} \approx 6.3 \text{ months}$$
:::

::: {.fragment}
Because $\sigma = \beta = 5$ years, lifetimes are as variable as they are long: a fleet mean says little about one drive.
:::
:::

---

## 🧠 Example 4.10: Memoryless

::: {style="font-size: 30px"}
If $Y$ is exponential and $a, b > 0$, then
$$P(Y > a + b \mid Y > a) = \frac{P(Y > a + b)}{P(Y > a)} = \frac{e^{-(a+b)/\beta}}{e^{-a/\beta}} = e^{-b/\beta} = P(Y > b)$$

::: {.fragment}
The intersection of $(Y > a + b)$ and $(Y > a)$ is just $(Y > a + b)$; that is the whole proof.
:::

::: {.fragment}
**Having survived $a$ units tells you nothing about the next $b$.** The geometric distribution in Chapter 3 had the same property (Exercise 4.95 links the two).
:::
:::

---

## ☎️ Memoryless at a Bank Call Centre

::: {style="font-size: 28px"}
Hold time at a bank's call centre is exponential with mean 4 minutes, so $P(Y > 2) = e^{-0.5} = 0.6065$. A caller has already held 5 minutes. Simulating 100,000 calls:
:::

```{r}
#| label: memoryless-sim
#| code-fold: false
hold <- rexp(100000, rate = 1/4)
c(fresh_caller  = mean(hold > 2),
  waited_5_mins = mean(hold[hold > 5] > 7),
  exact         = exp(-2/4))
```

::: {style="font-size: 28px"}
The caller who has waited 5 minutes is no closer to being answered. Whether that is **realistic** is a modelling question: a queue that clears in order does remember.
:::

---

## 🔬 Same Mean, Different Shape

```{ojs}
//| echo: false

viewof shape = {
  const input = Inputs.range([1, 20], {value: 1, step: 0.5,
    label: "Shape α (mean fixed at 10):"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}
```

```{ojs}
//| echo: false

lgam = x => {
  const g = 7, c = [0.99999999999980993, 676.5203681218851, -1259.1392167224028,
    771.32342877765313, -176.61502916214059, 12.507343278686905,
    -0.13857109526572012, 9.9843695780195716e-6, 1.5056327351493116e-7];
  x -= 1; let a = c[0]; const t = x + g + 0.5;
  for (let i = 1; i < 9; i++) a += c[i] / (x + i);
  return 0.5 * Math.log(2 * Math.PI) + (x + 0.5) * Math.log(t) - t + Math.log(a);
}
scaleB = 10 / shape
dens = y => y <= 0 ? (shape === 1 ? 1 / scaleB : 0) :
  Math.exp((shape - 1) * Math.log(y) - y / scaleB - shape * Math.log(scaleB) - lgam(shape))
grid = Array.from({length: 601}, (_, i) => ({y: i * 0.05, f: dens(i * 0.05)}))
tail = {
  let s = 0; const h = 0.01;
  for (let y = 15; y < 200; y += h) s += h * (dens(y) + dens(y + h)) / 2;
  return s;
}

md`α = **${shape}**, β = **${scaleB.toFixed(2)}**: σ = **${(10 / Math.sqrt(shape)).toFixed(2)}**, and P(Y > 15) = **${tail.toFixed(4)}**`

Plot.plot({
  width: 1150,
  height: 310,
  marginTop: 40,
  marginLeft: 78,
  marginBottom: 58,
  style: {fontSize: "18px"},
  x: {label: "Waiting time y (mean 10)", domain: [0, 30]},
  y: {label: "f(y)", grid: true},
  marks: [
    Plot.areaY(grid.filter(d => d.y >= 15), {x: "y", y: "f", fill: "#cbb8a9"}),
    Plot.line(grid, {x: "y", y: "f", stroke: "#14130f", strokeWidth: 2.5}),
    Plot.ruleX([10], {stroke: "#8b2635", strokeDasharray: "4 4"}),
    Plot.ruleY([0])
  ]
})
```

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 30px"}
Back to the AZN/USD desk. Gaps between large trades are now modelled as **exponential with mean 25 minutes**.

**Four minutes, in pairs:**

1. What is the probability that a gap exceeds 30 minutes?

2. No large trade for 40 minutes. What is the probability that one arrives in the next 10?

3. The time $W$ until the third large trade is gamma with $\alpha = 3$, $\beta = 25$. Find $E(W)$ and the standard deviation of $W$.
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 30px"}
1. $P(Y > 30) = e^{-30/25} = e^{-1.2} = 0.3012$.

2. Memoryless: the 40 minutes are irrelevant.
$$P(Y \le 50 \mid Y > 40) = P(Y \le 10) = 1 - e^{-0.4} = 0.3297$$

3. Theorem 4.8: $E(W) = 3 \times 25 = 75$ minutes, $V(W) = 3 \times 625 = 1875$, so $\sigma_W = 43.30$ minutes.

::: {.fragment}
Compare with the normal model in the motivating question: this one never puts mass below zero, and its $\sigma$ equals its mean for a single gap, as the desk observed.
:::
:::

---

## 📝 Quiz #1: Chi-Square Moments {.quiz-question}

$Y$ has a chi-square distribution with 8 degrees of freedom. What are $E(Y)$ and $V(Y)$?

- [$E(Y) = 8$, $V(Y) = 16$]{.correct data-explanation="✅ Theorem 4.9: E(Y) = ν and V(Y) = 2ν. Equivalently, gamma with α = 4 and β = 2 gives αβ = 8 and αβ² = 16."}
- $E(Y) = 4$, $V(Y) = 8$
- $E(Y) = 8$, $V(Y) = 64$
- $E(Y) = 16$, $V(Y) = 32$

---

## 📝 Quiz #2: A Server Fan {.quiz-question}

A cooling fan's lifetime is exponential with mean 4 years. It has already run 2 years. What is the probability it runs at least 1 more year?

- [$e^{-1/4} = 0.7788$]{.correct data-explanation="✅ Memoryless: P(Y > 3 | Y > 2) = P(Y > 1) = e^(−1/4). The two years already run do not enter."}
- $e^{-3/4} = 0.4724$
- $1 - e^{-1/4} = 0.2212$
- $e^{-2/4} = 0.6065$

---

## 📝 Quiz #3: Back Out the Parameters {.quiz-question}

A payment-processing delay is gamma distributed with mean 6 seconds and variance 12 seconds². What are $\alpha$ and $\beta$?

- [$\alpha = 3$, $\beta = 2$]{.correct data-explanation="✅ Divide the variance by the mean: αβ²/(αβ) = β = 12/6 = 2, then α = 6/2 = 3."}
- $\alpha = 2$, $\beta = 3$
- $\alpha = 6$, $\beta = 1$
- $\alpha = 0.5$, $\beta = 12$

---

## 📋 Key Formulas

::: {style="font-size: 28px"}
| | Density on $y \ge 0$ | $E(Y)$ | $V(Y)$ |
|---|---|---|---|
| Gamma, Def. 4.9 | $\dfrac{y^{\alpha-1}e^{-y/\beta}}{\beta^{\alpha}\Gamma(\alpha)}$ | $\alpha\beta$ | $\alpha\beta^2$ |
| $\chi^2_\nu$, Def. 4.10 | gamma, $\alpha = \nu/2$, $\beta = 2$ | $\nu$ | $2\nu$ |
| Exponential, Def. 4.11 | $\dfrac{1}{\beta}e^{-y/\beta}$ | $\beta$ | $\beta^2$ |

- $\Gamma(n) = (n-1)!$ and $\Gamma(\alpha) = (\alpha-1)\Gamma(\alpha-1)$
- Exponential: $P(Y > y) = e^{-y/\beta}$ and $P(Y > a+b \mid Y > a) = P(Y > b)$
- In R: `pgamma(y, alpha, 1/beta)`, `qgamma(p, alpha, 1/beta)`
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- The gamma family models quantities that are **nonnegative and skewed right**: waiting times, lifetimes, repair times

- $\alpha$ sets the shape, $\beta$ the scale; with the mean fixed, a larger $\alpha$ means a thinner tail

- Theorem 4.8 follows from one idea: every gamma integral is $\beta^{\alpha}\Gamma(\alpha)$

- Chi-square ($\alpha = \nu/2$, $\beta = 2$) and exponential ($\alpha = 1$) are gamma special cases

- The exponential is **memoryless**, which is both its convenience and its limitation
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition**

- Exercises at the end of §4.6: start with 4.81, 4.82, 4.88, 4.89 and 4.91, then 4.96, 4.104 and 4.106 – 4.109

- Re-set Exercise 4.91 as a bank's cash-demand problem: what vault holding keeps the chance of a shortfall at 1%?

**Week 10, Problem Set 2** is open now and closes **Sunday 22 November at 23:59** on WeBWorK, covering §4.6.

**Next class:** 18 November, the beta distribution and how to choose a continuous model (Wackerly §4.7 – 4.8).
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**

📧 sorujov@ada.edu.az\
🏢 Building D, Room D325\
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- A drive that has run 4 years is as good as new under the exponential model. What would you look for in real failure data to reject that?

- Why does the gamma tail have a Poisson-sum form only when $\alpha$ is an integer?

- With the mean held fixed, what happens to the gamma density as $\alpha \to \infty$, and which distribution does it start to resemble?
:::
