---
title: "Mathematical Statistics"
subtitle: "The Event-Composition Method"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - The Event-Composition Method"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
```

## 🎯 Learning Objectives

::: {style="font-size: 32px"}
By the end of this lecture, you will be able to:

- **Follow** the four steps of the event-composition method (Wackerly §2.9)

- **Write** a compound event as unions, intersections and complements of events whose probabilities are known

- **Test** a composition against the sample points before trusting it

- **Compute** the reliability of series and parallel systems, and of a chain that mixes them

- **Handle** "at least one" and sequential draws without replacement by composition
:::

---

## 🗺️ Where We Are

::: {style="font-size: 32px"}
**Wackerly §2.9**

Last class ended on a promise: *the event-composition method — how to decide, for a compound event with no obvious decomposition, which events to name in the first place.*

::: {.fragment}
We now own the tools: Theorem 2.5 for $\cap$, Theorem 2.6 for $\cup$, Theorem 2.7 for complements.
:::

::: {.fragment}
Today is about the step **before** the tools — choosing the events so that every probability the laws ask for is one you actually know.
:::
:::

---

## 💳 Motivating Question

::: {style="font-size: 30px"}
::: {.callout-important}
## A card payment in a Baku supermarket

The payment passes a **terminal**, the **acquirer's switch**, the **card network** (run from two redundant data centres) and the **issuing bank**. Each piece has a published availability.

What is the probability that the payment is authorised?
:::

::: {.fragment}
Nobody publishes "the probability the payment goes through". It has to be **built** from the pieces — and the building is the method.
:::
:::

---

## 📐 The Event-Composition Method

::: {style="font-size: 30px"}
::: {.callout-note}
## Wackerly §2.9: the four steps

1. Define the experiment.
2. Visualize the nature of the sample points. Identify a few to clarify your thinking.
3. Write an equation expressing the event of interest, $A$, as a composition of two or more events, using unions, intersections and/or complements. Make certain that $A$ and the composition represent the **same set of sample points**.
4. Apply the additive and multiplicative laws to the composition to find $P(A)$.
:::

Unlike the sample-point method, no list of $S$ is needed — only a clear picture of a typical sample point.
:::

---

## 🔑 Step 3 Is the Hard One

::: {style="font-size: 30px"}
Many compositions equal $A$. **Pick the one in which every probability in step 4 is known.** If one is unknown, the method fails.

::: {.fragment}
Two kinds of pieces make step 4 easy:

- **mutually exclusive** pieces, so the additive law has no subtraction
- **independent** pieces, so the multiplicative law has no conditionals
:::

::: {.fragment}
The most common error is a composition that is simply not equal to $A$. **Test the equality**: take a sample point in $A$ and check it lands in the composition, and the other way round.
:::
:::

---

## 🚢 Example 2.17: Export Shipments

::: {style="font-size: 28px"}
An exporter sends 40% of its shipments to Europe through **Alat port** ($R$) and 60% overland through the **Red Bridge** crossing ($D$). Customs clears 70% of Alat shipments and 80% of Red Bridge shipments within 48 hours ($F$). A shipment is picked at random.

::: {.fragment}
**Step 3.** Every shipment takes exactly one route, so
$$F = (F \cap R) \cup (F \cap D), \qquad \text{two mutually exclusive pieces.}$$
:::

::: {.fragment}
**Step 4.** $P(F \cap R) = P(F \mid R)P(R) = 0.7 \times 0.4 = 0.28$ and $P(F \cap D) = 0.8 \times 0.6 = 0.48$, so
$$P(F) = 0.28 + 0.48 = 0.76$$
:::
:::

---

## 🏦 Example 2.19: A Clearing Mandate

::: {style="font-size: 28px"}
Five correspondent banks bid to clear a company's euro payments; two quote the **lowest fees**. The treasurer shortlists **two at random**. What is $P(A)$, $A$ = *exactly one of the two cheapest is shortlisted*?

::: {.fragment}
Let $B_1$: cheapest bank on the first pick, $B_2$: one of the three dearer banks on the second, and $B_3, B_4$ the same the other way round. Define $C_1, \ldots, C_4$ likewise for the **second-cheapest** bank. Then
$$A = (B_1 \cap B_2) \cup (B_3 \cap B_4) \cup (C_1 \cap C_2) \cup (C_3 \cap C_4)$$
:::

::: {.fragment}
**Test it:** every shortlist with exactly one cheap bank lies in exactly one of the four pieces, and no other shortlist lies in any.
:::
:::

---

## 🧮 Example 2.19: Step 4, and a Check

::: {style="font-size: 28px"}
The four pieces are mutually exclusive, and each is a two-stage draw:
$$P(B_1 \cap B_2) = P(B_1)P(B_2 \mid B_1) = \tfrac{1}{5} \times \tfrac{3}{4} = \tfrac{3}{20}, \qquad P(A) = 4 \times \tfrac{3}{20} = \tfrac{3}{5}$$
:::

```{r}
#| label: clearing-check
# Sample-point check: list all C(5,2) = 10 shortlists; banks 1 and 2 are cheapest
shortlists <- combn(5, 2)
n_cheap    <- colSums(shortlists <= 2)
c(shortlists = ncol(shortlists), exactly_one = sum(n_cheap == 1),
  P_A = mean(n_cheap == 1))
```

::: {style="font-size: 28px"}
Same answer as the sample-point method, reached without writing $S$ down.
:::

---

## ⚡ Example 2.20: Power Feeds

::: {style="font-size: 28px"}
A Baku data centre hosting a bank's core system has **three independent power feeds**, each live during a grid event with probability 0.9. What is $P(A)$, $A$ = *at least one feed is live*?

::: {.fragment}
Let $B_i$: feed $i$ fails, so $P(B_i) = 0.1$. Then $\bar{A} = B_1 \cap B_2 \cap B_3$, and by Theorem 2.7 and independence
$$P(A) = 1 - P(B_1)P(B_2)P(B_3) = 1 - (0.1)^3 = 0.999$$
:::

::: {.fragment}
As a union, $A$ needs seven terms of inclusion–exclusion. As a complement, it is **one intersection**. Frequently $P(\bar{A})$ is the easy one.
:::
:::

---

## 🔗 Series and Parallel Systems

::: {style="font-size: 28px"}
$n$ independent components, each working with probability $p$.

:::: {.columns}
::: {.column width="50%"}
::: {.callout-note}
## Series: all must work
A payment settlement chain.
$$W = W_1 \cap W_2 \cap \cdots \cap W_n$$
$$P(W) = p^n$$
:::
:::
::: {.column width="50%"}
::: {.callout-note}
## Parallel: one is enough
Redundant power feeds.
$$\bar{W} = \bar{W}_1 \cap \bar{W}_2 \cap \cdots \cap \bar{W}_n$$
$$P(W) = 1 - (1-p)^n$$
:::
:::
::::

::: {.fragment}
Same components, opposite compositions: every link added to a series **costs** reliability, every feed added in parallel **buys** it.
:::
:::

---

## 💻 Series vs Parallel in R

```{r}
#| label: series-parallel
p <- 0.95                          # each component works 95% of the time
n <- c(1, 2, 4, 8, 12, 20)
data.frame(components = n,
           series     = round(p^n, 4),
           parallel   = round(1 - (1 - p)^n, 4),
           parallel_fails = signif((1 - p)^n, 3))
```

::: {style="font-size: 28px"}
Twelve 95% links in series work only 54% of the time; two 95% feeds in parallel already give 99.75%.
:::

---

## 🔬 Interactive: Series Versus Parallel {.smaller}

```{ojs}
//| echo: false

viewof p_comp = {
  const input = Inputs.range([0.50, 0.99], {value: 0.95, step: 0.01,
    label: "Reliability of each component p:"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}
```

```{ojs}
//| echo: false

rel = Array.from({length: 20}, (_, i) => i + 1).flatMap(n => [
  {n, R: Math.pow(p_comp, n), system: "series"},
  {n, R: 1 - Math.pow(1 - p_comp, n), system: "parallel"}
])
halfN = (rel.find(d => d.system === "series" && d.R < 0.5) || {n: "more than 20"}).n

md`At **p = ${p_comp.toFixed(2)}**, a series chain drops below 50% at **${halfN}** components.`

Plot.plot({
  width: 1150,
  height: 300,
  marginLeft: 78,
  marginBottom: 58,
  marginRight: 110,
  marginTop: 40,
  style: {fontSize: "18px"},
  x: {label: "Number of components n", domain: [1, 20], ticks: [1, 5, 10, 15, 20]},
  y: {label: "P(system works)", domain: [0, 1], tickFormat: ".1f"},
  color: {domain: ["series", "parallel"], range: ["#8b2635", "#2f5d8a"]},
  marks: [
    Plot.ruleY([0.5], {stroke: "#cbb8a9", strokeDasharray: "4 4"}),
    Plot.line(rel, {x: "n", y: "R", stroke: "system", strokeWidth: 3}),
    Plot.text(rel.filter(d => d.n === 20), {x: "n", y: "R", text: "system",
      fill: "system", dx: 12, textAnchor: "start", fontSize: 20}),
    Plot.ruleY([0])
  ]
})
```

---

## 💳 Back to the Card Payment

::: {style="font-size: 28px"}
Terminal $T$ (0.995), acquirer switch $Q$ (0.99), network data centres $N_1, N_2$ (0.97 each), issuer $I$ (0.99), all independent. Authorisation needs every stage, and the network needs **either** centre:
$$\text{Auth} = T \cap Q \cap (N_1 \cup N_2) \cap I$$

::: {.fragment}
$$P(N_1 \cup N_2) = 1 - (0.03)^2 = 0.9991$$
$$P(\text{Auth}) = 0.995 \times 0.99 \times 0.9991 \times 0.99 = 0.9743$$
:::

::: {.fragment}
With a single data centre it would be $0.995 \times 0.99 \times 0.97 \times 0.99 = 0.9459$. The second centre halves the failure rate, from 5.4% to 2.6%.
:::
:::

---

## 🏠 Example 2.21: Mortgage Pipeline

::: {style="font-size: 28px"}
Each application arriving at a bank's mortgage desk needs **manual underwriting** with probability $p = 1/6$, independently. $A_i$: application $i$ does **not**. The $r$th is the first manual case:
$$E_r = A_1 \cap \cdots \cap A_{r-1} \cap \bar{A}_r, \qquad P(E_r) = (5/6)^{r-1}(1/6)$$
:::

```{r}
#| label: first-manual
r   <- 1:6
P_r <- (5/6)^(r - 1) * (1/6)
round(c(setNames(P_r, paste0("r=", r)), within_6 = sum(P_r)), 4)
```

::: {style="font-size: 28px"}
Over all $r$ the geometric series sums to $\frac{1/6}{1 - 5/6} = 1$.
:::

---

## 📦 Draws Without Replacement

::: {style="font-size: 28px"}
A customs post holds **8 containers**, of which **2** carry undeclared goods. Inspectors open them one at a time in random order. $N_i$: the $i$th container opened is clean.

::: {.fragment}
**(a) At least one flagged container among the first three.** The complement is "the first three are all clean", a chain in which each factor conditions on what came before:
$$P = 1 - P(N_1)P(N_2 \mid N_1)P(N_3 \mid N_1 \cap N_2) = 1 - \tfrac{6}{8}\cdot\tfrac{5}{7}\cdot\tfrac{4}{6} = 1 - \tfrac{5}{14} = 0.6429$$
:::

::: {.fragment}
The factors are **not** $\tfrac{6}{8}$ three times: a clean container that has been opened is not put back.
:::
:::

---

## 📦 Found on the Fourth Opening

::: {style="font-size: 28px"}
**(b)** Exactly one flagged container in the first three, then a flagged one: three **mutually exclusive** orders, FNNF, NFNF, NNFF, each with the probability of FNNF:
$$\tfrac{2}{8}\cdot\tfrac{6}{7}\cdot\tfrac{5}{6}\cdot\tfrac{1}{5} = \tfrac{1}{28}, \qquad P = 3 \times \tfrac{1}{28} = \tfrac{3}{28} = 0.1071$$
:::

```{r}
#| label: containers
# The two flagged containers occupy 2 of 8 opening positions, all C(8,2) = 28 equally likely
pos <- combn(8, 2)
c(at_least_one_in_first_3 = mean(pos[1, ] <= 3),
  second_found_on_4th     = mean(pos[2, ] == 4))
```

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 30px"}
A cotton shipment from Baku to Poti leaves on time only if **customs** clears it (0.92), a **rail slot** is booked through either of two forwarders (0.75 each), and the **port** accepts it (0.96). All independent.

**Four minutes, in pairs:**

1. Write the event "leaves on time" as a composition. Then find its probability.

2. Which is worth more: a **third forwarder**, or raising customs to **0.97**?
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 28px"}
1. $\text{OnTime} = C \cap (F_1 \cup F_2) \cap P$, and $P(F_1 \cup F_2) = 1 - 0.25^2 = 0.9375$:
$$P(\text{OnTime}) = 0.92 \times 0.9375 \times 0.96 = 0.828$$

::: {.fragment}
2. Third forwarder: $0.92 \times (1 - 0.25^3) \times 0.96 = 0.8694$.
   Better customs: $0.97 \times 0.9375 \times 0.96 = 0.8730$.
:::

::: {.fragment}
Customs wins, narrowly. The parallel block is already at 94%; the weakest **series** link is where the reliability leaks. The composition tells you where to spend money.
:::
:::

---

## 📝 Quiz #1: A Settlement Chain {.quiz-question}

A cross-border payment passes three independent links, each working with probability 0.98. What is the probability the payment settles?

- [$0.98^3 = 0.9412$]{.correct data-explanation="✅ Settlement needs every link, an intersection of independent events: 0.98 × 0.98 × 0.98 = 0.9412."}
- $1 - 0.02^3 = 0.999992$
- $0.98$
- $3 \times 0.02 = 0.06$

---

## 📝 Quiz #2: Test the Composition {.quiz-question}

$A$: the export licence is approved; $B$: the import licence is approved. Which composition is the event **exactly one** licence is approved?

- $A \cup B$
- [$(A \cap \bar{B}) \cup (\bar{A} \cap B)$]{.correct data-explanation="✅ A sample point with exactly one approval lies in exactly one of these two disjoint pieces. A ∪ B also contains the point where both are approved, so it fails the test."}
- $\overline{A \cap B}$
- $\bar{A} \cap \bar{B}$

---

## 📝 Quiz #3: Without Replacement {.quiz-question}

Of 5 letters of credit, 2 contain discrepancies. They are checked one at a time in random order. What is the probability the first two checked are both clean?

- [$\tfrac{3}{5} \times \tfrac{2}{4} = 0.30$]{.correct data-explanation="✅ Theorem 2.5 in sequence: 3/5 clean on the first check, then 2 clean left among 4. The second factor conditions on the first."}
- $\left(\tfrac{3}{5}\right)^2 = 0.36$
- $\tfrac{3}{5} = 0.60$
- $\tfrac{2}{5} \times \tfrac{1}{4} = 0.10$

---

## 📋 Key Formulas

::: {style="font-size: 30px"}
| | Statement |
|---|---|
| partition by route | $P(F) = P(F \mid R)P(R) + P(F \mid D)P(D)$ |
| series, $n$ components | $P(W) = p^n$ |
| parallel, $n$ components | $P(W) = 1 - (1-p)^n$ |
| at least one | $P(A) = 1 - P(\bar{A}_1 \cap \cdots \cap \bar{A}_n)$ |
| first success on trial $r$ | $P(E_r) = (1-p)^{r-1}p$ |
| sequential, no replacement | $P(A_1 \cap A_2 \cap A_3) = P(A_1)P(A_2 \mid A_1)P(A_3 \mid A_1 \cap A_2)$ |
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- The method has four steps; step 3, **writing the composition**, is where the work is

- Choose pieces whose probabilities you know, preferably mutually exclusive or independent

- Test the composition against sample points before applying any law

- Series is an intersection, parallel is the complement of an intersection

- "At least one" goes through the complement; draws without replacement go through the chain of conditionals
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition**

- Exercises at the end of §2.9 — start with 2.110, 2.112, 2.116, then 2.120 and 2.121

- For 2.116, draw the three lines as a series or parallel diagram **before** writing any probability

**Week 4, Problem Set 1** opens today and closes **10 October at 23:59** on WeBWorK, covering §2.9.

**Next class:** what to do when the event depends on which state the world is in — the law of total probability and Bayes' rule.
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**

📧 sorujov@ada.edu.az
🏢 Building D, Room D325
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- Example 2.17 split $F$ by route. What goes wrong if some shipments could use both routes?

- A system of four 0.9 components can be wired as two parallel pairs in series, or two series pairs in parallel. Which is more reliable?

- Why does the method fail, rather than approximate, when one probability in step 4 is unknown?
:::
