---
title: "Mathematical Statistics"
subtitle: "Chapter 5: Multivariate Distributions – Conditional Distributions & Independence"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ../ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    margin: 0.04
    min-scale: 0.2
    max-scale: 2.0
    footer: "Mathematical Statistics - Ch.5: Multivariate Distributions (Part 2)"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    mouse-wheel: true
    overview: true
    zoom: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::::: learning-objectives
:::: {style="font-size:38px"}
📚 By the end of this lecture, you will be able to:

::: incremental
-   **Derive** conditional probability functions for discrete random variables

-   **Calculate** conditional density functions for continuous random variables

-   **Apply** the definition of independence to verify whether random variables are independent

-   **Use** the factorization theorem to quickly check independence

-   **Interpret** conditional distributions in portfolio risk and asset allocation scenarios
:::
::::
:::::

## 📋 Overview

::::: callout-note
## 📚 Topics Covered Today

:::: {style="font-size:32px"}
::: incremental
-   **Conditional Distributions (Discrete)** – Probability given information about another variable

-   **Conditional Distributions (Continuous)** – Conditional density functions

-   **Independent Random Variables** – Definition and properties

-   **Factorization Theorem** – Quick test for independence

-   **Case Study** – Conditional analysis of banking sector returns
:::
::::
:::::

## 📖 Why Conditional Distributions?

::::::::: callout-note
## 🎯 Motivation

:::::::: {style="font-size:33px"}
Understanding how one variable behaves **given information about another** is crucial for prediction, risk management, and decision-making.

::::::: columns
:::: {.column width="50%"}
::: fragment
**Finance Applications:**

-   Credit risk given macroeconomic conditions
-   Stock returns given market movements
-   Default probability given credit rating
-   Portfolio VaR given stress scenarios
:::
::::

:::: {.column width="50%"}
::: fragment
**Azerbaijan Context:**

-   Oil revenue given OPEC decisions
-   Manat volatility given oil prices
-   PASHA Bank profitability given interest rates
-   Insurance claims given weather events
:::
::::
:::::::
::::::::
:::::::::

## 📖 Definition: Conditional Probability Function (Discrete) {.larger}

:::::: {.callout-note style="font-size:40px;"}
## 📝 Definition 5.5: Conditional Probability Function

If $Y_1$ and $Y_2$ are jointly discrete random variables with joint probability function $p(y_1, y_2)$ and marginal probability functions $p_1(y_1)$ and $p_2(y_2)$, then the **conditional discrete probability function** of $Y_1$ given $Y_2$ is:

$$p(y_1|y_2) = P(Y_1 = y_1 | Y_2 = y_2) = \frac{p(y_1, y_2)}{p_2(y_2)}$$

provided that $p_2(y_2) > 0$.

::: {.fragment style="font-size:30px;"}
**Key insight:** This is Bayes' rule $P(A|B) = P(A \cap B)/P(B)$ applied to random variables!
:::
::::::

## 📌 Example 1: Portfolio Asset Selection {.smaller}

**Problem:** A fund manager selects 2 assets from a pool of 3 tech stocks, 2 bank stocks, and 1 commodity ETF. Let $Y_1$ = number of tech stocks, $Y_2$ = number of bank stocks selected. Find $P(Y_1 = 1 | Y_2 = 1)$.

::: {.fragment}
**Solution:** From the joint distribution:

| $y_1 \backslash y_2$ | 0 | 1 | 2 | **Total** |
|:---:|:---:|:---:|:---:|:---:|
| 0 | 0 | 2/15 | 1/15 | 3/15 |
| 1 | 3/15 | 6/15 | 0 | 9/15 |
| 2 | 3/15 | 0 | 0 | 3/15 |
| **Total** | 6/15 | **8/15** | 1/15 | 1 |
:::

::: {.fragment}
$$P(Y_1 = 1 | Y_2 = 1) = \frac{p(1,1)}{p_2(1)} = \frac{6/15}{8/15} = \frac{6}{8} = \frac{3}{4}$$
:::

## 📌 Example 1: Complete Conditional Distribution

::::: {style="font-size: 34px;"}
**Problem:** Find the complete conditional distribution of $Y_1$ (tech stocks) given $Y_2 = 1$ (one bank stock selected).

::: {.columns}
::: {.column width="50%"}
$$P(Y_1 = 0 | Y_2 = 1) = \frac{2/15}{8/15} = \frac{1}{4}$$

$$P(Y_1 = 1 | Y_2 = 1) = \frac{6/15}{8/15} = \frac{3}{4}$$

$$P(Y_1 = 2 | Y_2 = 1) = \frac{0}{8/15} = 0$$
:::

::: {.column width="50%"}
::: {.fragment}
**Portfolio Interpretation:**

Given one bank stock in portfolio:

- 25% chance of zero tech stocks
- 75% chance of one tech stock  
- 0% chance of two tech stocks

**Verification:** $\frac{1}{4} + \frac{3}{4} + 0 = 1$ ✓
:::
:::
:::
:::::

## 📖 Definition: Conditional Density Function (Continuous) {.larger}

:::::: {.callout-note style="font-size:36px;"}
## 📝 Definition 5.7: Conditional Density Function

Let $Y_1$ and $Y_2$ be jointly continuous random variables with joint density $f(y_1, y_2)$ and marginal densities $f_1(y_1)$ and $f_2(y_2)$. 

For any $y_2$ such that $f_2(y_2) > 0$, the **conditional density** of $Y_1$ given $Y_2 = y_2$ is:

$$f(y_1|y_2) = \frac{f(y_1, y_2)}{f_2(y_2)}$$

Similarly: $f(y_2|y_1) = \frac{f(y_1, y_2)}{f_1(y_1)}$ when $f_1(y_1) > 0$

::: {.fragment style="font-size:28px;"}
**Note:** The conditional density is **undefined** when the marginal density equals zero!
:::
::::::

## 📌 Example 2: Trading Volume Given Liquidity {.smaller}

**Problem:** Let $Y_2$ = available liquidity (millions AZN), $Y_1$ = actual trade volume ($Y_1 \leq Y_2$). Joint density:
$$f(y_1, y_2) = \begin{cases} 1/2, & 0 \leq y_1 \leq y_2 \leq 2 \\ 0, & \text{elsewhere} \end{cases}$$

Find $f(y_1|y_2)$.

::: {.fragment}
**Solution:** First find the marginal density of liquidity $f_2(y_2)$:
$$f_2(y_2) = \int_0^{y_2} \frac{1}{2} \, dy_1 = \frac{y_2}{2}, \quad 0 \leq y_2 \leq 2$$
:::

::: {.fragment}
Then:
$$f(y_1|y_2) = \frac{1/2}{y_2/2} = \frac{1}{y_2}, \quad 0 \leq y_1 \leq y_2$$

**Interpretation:** Given liquidity $y_2$, trade volume is **uniform** on $[0, y_2]$!
:::

## 📌 Example 2: Conditional Probability Calculation {.smaller}

**Problem:** Find $P(Y_1 \leq 0.5 | Y_2 = 1.5)$ — probability trade volume ≤ 0.5M given liquidity of 1.5M AZN.

::: {.fragment}
**Solution:** Using $f(y_1|y_2 = 1.5) = \frac{1}{1.5}$ for $0 \leq y_1 \leq 1.5$:

$$P(Y_1 \leq 0.5 | Y_2 = 1.5) = \int_0^{0.5} \frac{1}{1.5} \, dy_1 = \frac{0.5}{1.5} = \frac{1}{3}$$
:::

::: {.fragment}
**Compare with higher liquidity $Y_2 = 2$M:**
$$P(Y_1 \leq 0.5 | Y_2 = 2) = \int_0^{0.5} \frac{1}{2} \, dy_1 = \frac{1}{4}$$

**Trading Insight:** More liquidity → lower probability of small trades!
:::

## 📖 Definition: Independent Random Variables {.larger}

:::::: {.callout-note style="font-size:36px;"}
## 📝 Definition 5.8: Independence

Let $Y_1$ have distribution function $F_1(y_1)$, $Y_2$ have distribution function $F_2(y_2)$, and let $(Y_1, Y_2)$ have joint distribution function $F(y_1, y_2)$. 

Then $Y_1$ and $Y_2$ are **independent** if and only if:

$$F(y_1, y_2) = F_1(y_1) \cdot F_2(y_2)$$

for every pair of real numbers $(y_1, y_2)$.

If $Y_1$ and $Y_2$ are not independent, they are said to be **dependent**.

::: {.fragment style="font-size:28px;"}
**Portfolio Intuition:** Joint CDF factors into a product of marginal CDFs → no information about one asset tells you anything about the other!
:::
::::::

## 🧮 Theorem: Characterization of Independence {.larger}

::::::: {.callout-important style="font-size:36px;"}
## Theorem 5.4: Independence via Probability/Density Functions

::::: columns
::: {.column .fragment width="50%"}
**Discrete case:** $Y_1$ and $Y_2$ are independent if and only if:
$$p(y_1, y_2) = p_1(y_1) \cdot p_2(y_2) \quad \text{for all } (y_1, y_2)$$
:::

::: {.column .fragment width="50%"}
**Continuous case:** $Y_1$ and $Y_2$ are independent if and only if:
$$f(y_1, y_2) = f_1(y_1) \cdot f_2(y_2) \quad \text{for all } (y_1, y_2)$$
:::
:::::

::: {.fragment style="font-size:28px;"}
**Key insight:** Independence means the joint distribution **factors** into the product of marginals for **ALL** values — essential for portfolio diversification theory!
:::
:::::::

## 📌 Example 3: Testing Independence (Loan Defaults) {.smaller}

**Problem:** A bank has two independent loan portfolios. Let $Y_1$ = default category for portfolio 1, $Y_2$ = default category for portfolio 2 (each rated 1-6 risk levels). Each has uniform distribution. Are $Y_1$ and $Y_2$ independent?

::: {.fragment}
Since portfolios are managed independently: $p(y_1, y_2) = 1/36$ for all $y_1, y_2 \in \{1,2,3,4,5,6\}$.
:::

::: {.fragment}
Marginals: $p_1(y_1) = 1/6$ and $p_2(y_2) = 1/6$
:::

::: {.fragment}
Check: $p_1(y_1) \cdot p_2(y_2) = \frac{1}{6} \cdot \frac{1}{6} = \frac{1}{36} = p(y_1, y_2)$ ✓
:::

::: {.fragment}
**Conclusion:** $Y_1$ and $Y_2$ are **INDEPENDENT**.

**Banking interpretation:** Separately managed portfolios → diversification benefit!
:::

## 📌 Example 4: Dependence in Portfolio Selection {.smaller}

**Problem:** In Example 1 (asset selection), are $Y_1$ (tech stocks) and $Y_2$ (bank stocks) independent?

::: {.fragment}
From the joint distribution: $p(0, 0) = 0$
:::

::: {.fragment}
But marginals give: $p_1(0) = 3/15$ and $p_2(0) = 6/15$
:::

::: {.fragment}
$$p_1(0) \cdot p_2(0) = \frac{3}{15} \cdot \frac{6}{15} = \frac{2}{25} \neq 0 = p(0, 0)$$
:::

::: {.fragment}
**Conclusion:** $Y_1$ and $Y_2$ are **DEPENDENT**.

**Portfolio intuition:** If we select zero tech stocks, we must have selected bank stocks or the commodity ETF!
:::

## 📌 Example 5: Testing Independence (Stock Returns) {.smaller}

**Problem:** Let $Y_1$ = normalized return of AAPL, $Y_2$ = normalized return of a commodity ETF. Joint density: $f(y_1, y_2) = 6y_1y_2^2$ for $0 \leq y_1 \leq 1, 0 \leq y_2 \leq 1$. Are they independent?

::: {.fragment}
Find marginals:
$$f_1(y_1) = \int_0^1 6y_1y_2^2 \, dy_2 = 6y_1 \cdot \frac{1}{3} = 2y_1, \quad 0 \leq y_1 \leq 1$$

$$f_2(y_2) = \int_0^1 6y_1y_2^2 \, dy_1 = 6y_2^2 \cdot \frac{1}{2} = 3y_2^2, \quad 0 \leq y_2 \leq 1$$
:::

::: {.fragment}
Check: $f_1(y_1) \cdot f_2(y_2) = 2y_1 \cdot 3y_2^2 = 6y_1y_2^2 = f(y_1, y_2)$ ✓

**Conclusion:** $Y_1$ and $Y_2$ are **INDEPENDENT** — great for diversification!
:::

## 📌 Example 6: Dependent Bond Returns {.smaller}

**Problem:** Let $Y_1$ = short-term bond return, $Y_2$ = long-term bond return, where short-term never exceeds long-term. Joint density: $f(y_1, y_2) = 2$ for $0 \leq y_2 \leq y_1 \leq 1$. Are they independent?

::: {.fragment}
Find marginals:
$$f_1(y_1) = \int_0^{y_1} 2 \, dy_2 = 2y_1, \quad f_2(y_2) = \int_{y_2}^1 2 \, dy_1 = 2(1-y_2)$$
:::

::: {.fragment}
Check: $f_1(y_1) \cdot f_2(y_2) = 4y_1(1-y_2) \neq 2 = f(y_1, y_2)$
:::

::: {.fragment}
**Conclusion:** $Y_1$ and $Y_2$ are **DEPENDENT**.

**Key observation:** The triangular region $0 \leq y_2 \leq y_1 \leq 1$ signals dependence — term structure creates relationship!
:::

## 🧮 Theorem: Factorization Criterion {.larger}

::::::: {.callout-important style="font-size:32px;"}
## Theorem 5.5: Quick Test for Independence

Let $Y_1$ and $Y_2$ have a joint density $f(y_1, y_2)$ that is positive if and only if $a \leq y_1 \leq b$ and $c \leq y_2 \leq d$, for constants $a, b, c, d$; and $f(y_1, y_2) = 0$ otherwise.

Then $Y_1$ and $Y_2$ are **independent** if and only if:
$$f(y_1, y_2) = g(y_1) \cdot h(y_2)$$

where $g(y_1)$ is a nonnegative function of $y_1$ alone and $h(y_2)$ is a nonnegative function of $y_2$ alone.

::: {.fragment style="font-size:26px;"}
**Risk Management Insight:** Support must be **rectangular**. Non-rectangular support (like budget constraints) → typically dependent!
:::
:::::::

## 📌 Example 7: Interest Rate Factorization {.smaller}

**Problem:** Let $Y_1$ = normalized US interest rate, $Y_2$ = normalized EU interest rate. Joint density: $f(y_1, y_2) = 2y_1$ for $0 \leq y_1 \leq 1, 0 \leq y_2 \leq 1$. Are they independent?

::: {.fragment}
**Step 1:** Check the region — $[0,1] \times [0,1]$ is rectangular ✓
:::

::: {.fragment}
**Step 2:** Can we factor?

$$f(y_1, y_2) = 2y_1 = \underbrace{y_1}_{g(y_1)} \cdot \underbrace{2}_{h(y_2)}$$

Yes! The joint density factors.
:::

::: {.fragment}
**Conclusion:** $Y_1$ and $Y_2$ are **INDEPENDENT**.

**Economic note:** If central banks set rates independently, this model applies!
:::

## 🎮 Interactive: Why Support Shape Determines Independence {.smaller}

:::::: {style="font-size: 0.60em; margin-top: -12px;"}
::::: columns
::: {.column width="30%"}

```{ojs}
//| echo: false

viewof power_a = {
  const input = Inputs.range([0, 3], {
    value: 1,
    step: 0.5,
    label: "Power a (for y₁):"
  });

  input.addEventListener('pointerdown', e => e.stopPropagation());
  input.addEventListener('touchstart', e => e.stopPropagation());
  input.addEventListener('mousedown', e => e.stopPropagation());
  input.addEventListener('click', e => e.stopPropagation());
  input.addEventListener('wheel', e => e.stopPropagation());
  input.addEventListener('pointermove', e => e.stopPropagation());
  input.addEventListener('touchmove', e => e.stopPropagation());

  return input;
}

viewof power_b = {
  const input = Inputs.range([0, 3], {
    value: 1,
    step: 0.5,
    label: "Power b (for y₂):"
  });

  input.addEventListener('pointerdown', e => e.stopPropagation());
  input.addEventListener('touchstart', e => e.stopPropagation());
  input.addEventListener('mousedown', e => e.stopPropagation());
  input.addEventListener('click', e => e.stopPropagation());
  input.addEventListener('wheel', e => e.stopPropagation());
  input.addEventListener('pointermove', e => e.stopPropagation());
  input.addEventListener('touchmove', e => e.stopPropagation());

  return input;
}

viewof region_type = {
  const input = Inputs.radio(["Rectangular", "Triangular"], {
    value: "Rectangular",
    label: "Support Region:"
  });

  input.addEventListener('pointerdown', e => e.stopPropagation());
  input.addEventListener('touchstart', e => e.stopPropagation());
  input.addEventListener('mousedown', e => e.stopPropagation());
  input.addEventListener('click', e => e.stopPropagation());

  return input;
}

viewof sliceY1 = {
  const input = Inputs.range([0.05, 0.95], {
    value: 0.6,
    step: 0.05,
    label: "Slice at Y₁ ="
  });

  input.addEventListener('pointerdown', e => e.stopPropagation());
  input.addEventListener('touchstart', e => e.stopPropagation());
  input.addEventListener('mousedown', e => e.stopPropagation());
  input.addEventListener('click', e => e.stopPropagation());
  input.addEventListener('wheel', e => e.stopPropagation());
  input.addEventListener('pointermove', e => e.stopPropagation());
  input.addEventListener('touchmove', e => e.stopPropagation());

  return input;
}

formula_html = html`<span style="font-size: 1.05em;">f(y₁, y₂) = c · y₁<sup>${power_a}</sup> · y₂<sup>${power_b}</sup></span>`;

verdict_html = region_type === "Rectangular"
  ? html`<span style="color: #1a7f37; font-weight: 700;">INDEPENDENT</span>`
  : html`<span style="color: #b42318; font-weight: 700;">DEPENDENT</span>`;

lesson_html = region_type === "Rectangular"
  ? html`<div style="background: #d4edda; padding: 6px; border-radius: 5px; border-left: 3px solid #1a7f37; font-size: 0.95em;"><strong>Support:</strong> 0 &lt; Y₁ &lt; 1, 0 &lt; Y₂ &lt; 1<br>Y₂'s range <strong>never</strong> changes with Y₁.</div>`
  : html`<div style="background: #f8d7da; padding: 6px; border-radius: 5px; border-left: 3px solid #b42318; font-size: 0.95em;"><strong>Support:</strong> 0 &lt; Y₂ ≤ Y₁ &lt; 1<br>Knowing Y₁ <strong>restricts</strong> Y₂ (lower triangle!).</div>`;

md`**Key insight:** Independence needs both (1) density factors AND (2) rectangular support.

${viewof power_a}
${viewof power_b}
${viewof region_type}
${viewof sliceY1}

**Formula:** ${formula_html}

**Verdict:** ${verdict_html}

${lesson_html}`
```

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

jointDataInteractive = {
  const n = 80;
  const data = [];
  for (let i = 0; i < n; i++) {
    for (let j = 0; j < n; j++) {
      const y1 = i / (n - 1);
      const y2 = j / (n - 1);

      const inRegion = region_type === "Triangular" ? (y2 <= y1) : true;

      if (inRegion) {
        const density = Math.pow(y1 + 0.01, power_a) * Math.pow(y2 + 0.01, power_b);
        data.push({y1, y2, density});
      }
    }
  }
  
  // For triangular region, add zero-density boundary points to prevent contour bleeding
  if (region_type === "Triangular") {
    for (let i = 0; i < n; i++) {
      const y1 = i / (n - 1);
      for (let j = Math.ceil(y1 * (n-1)) + 1; j < n; j++) {
        const y2 = j / (n - 1);
        data.push({y1, y2, density: 0});
      }
    }
  }
  
  return data;
}

constraintLine = region_type === "Triangular"
  ? Array.from({length: 100}, (_, i) => ({y1: i / 99, y2: i / 99}))
  : [];

// Create a vertical band to highlight the conditioning slice
sliceBand = (() => {
  const bandwidth = 0.015;
  const ymax = region_type === "Triangular" ? sliceY1 : 1;
  
  return [{
    x1: sliceY1 - bandwidth,
    x2: sliceY1 + bandwidth,
    y1: 0,
    y2: ymax
  }];
})();

// Points for the conditional range marker
conditionalDots = (() => {
  const points = [];
  const ymax = region_type === "Triangular" ? sliceY1 : 1;
  for (let i = 0; i <= 35; i++) {
    points.push({y1: sliceY1, y2: (i / 35) * ymax});
  }
  return points;
})();

plot = Plot.plot({
  width: 600,
  height: 380,
  marginLeft: 55,
  marginBottom: 45,
  color: {scheme: "YlOrRd", legend: true, label: "Density"},
  x: {domain: [0, 1], label: "Y₁"},
  y: {domain: [0, 1], label: "Y₂"},
  title: region_type === "Rectangular"
    ? "Rectangle: Support is 0<Y₁<1, 0<Y₂<1 (full square)"
    : "Triangle: Support is 0<Y₂≤Y₁<1 (LOWER triangle)",
  marks: [
    Plot.contour(jointDataInteractive, {x: "y1", y: "y2", fill: "density", blur: 1, thresholds: 14}),
    // Vertical band highlighting the slice
    Plot.rect(sliceBand, {x1: "x1", x2: "x2", y1: "y1", y2: "y2", fill: "#3b82f6", fillOpacity: 0.12}),
    // Constraint boundary
    constraintLine.length ? Plot.line(constraintLine, {x: "y1", y: "y2", stroke: "#1f2937", strokeWidth: 3, strokeDasharray: "8,4"}) : null,
    // Conditional range dots
    Plot.dot(conditionalDots, {x: "y1", y: "y2", r: 3, fill: "#1d4ed8", opacity: 0.9}),
    // Slice marker line
    Plot.ruleX([sliceY1], {stroke: "#1d4ed8", strokeWidth: 3}),
    // Constraint label
    constraintLine.length ? Plot.text([{y1: 0.78, y2: 0.84, text: "y₂ = y₁"}], {x: "y1", y: "y2", text: "text", fontSize: 13, fontWeight: "bold", fill: "#1f2937"}) : null,
    Plot.frame()
  ].filter(Boolean)
});

html`<div>
  ${plot}
  <div style="margin-top: 5px; font-size: 0.9em; line-height: 1.3;">
    <strong style="color: #1d4ed8;">Blue band at Y₁=${sliceY1.toFixed(2)}:</strong> Shows Y₂'s feasible range. ${region_type === "Triangular" ? `Only [0, ${sliceY1.toFixed(2)}] allowed.` : "Always [0, 1] allowed."}
  </div>
</div>`
```

:::
:::::
::::::

## 🤝 Think-Pair-Share: Independence Check {.larger}

::::: {style="font-size: 36px;"}
:::: callout-note
## 💭 Activity (4 minutes)

**Problem:** A portfolio model has joint density $f(y_1, y_2) = 24y_1y_2$ for normalized returns where $0 < y_1 < 1$, $0 < y_2 < 1$, and $y_1 + y_2 < 1$ (budget constraint).

::: columns
::: {.column width="33%"}
**🧠 Think (1 min):**

- What shape is the support region?
- Can the density factor?
:::

::: {.column width="33%"}
**👫 Pair (2 min):**

- Sketch the region together
- Discuss: rectangular or not?
:::

::: {.column width="33%"}
**🗣️ Share (1 min):**

- Conclusion: Independent or dependent?
- Why?
:::
:::
::::
:::::

```{r}
#| echo: false
library(countdown)
countdown(minutes = 4, seconds = 0,
          bottom = 0, right = 0,
          font_size = "1.5em",
          color_running_background = "#31b09e",
          color_running_text = "white",
          color_finished_background = "#cc3311",
          color_finished_text = "white",
          warn_when = 60,
          play_sound = TRUE)
```

## 💰 Case Study: Conditional Analysis of Bank Returns {.smaller}

:::::: {style="font-size: 0.75em; margin-top: -5px;"}

::::: columns
::: {.column width="50%"}

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "📊 Show Data Loading Code"
#| message: false
#| warning: false
#| eval: true

library(tidyverse)
library(tidyquant)

# Get Azerbaijan-relevant bank stocks: JPM and BAC
# (similar to PASHA Bank and Kapital Bank dynamics)
symbols <- c("JPM", "BAC")
prices <- tq_get(symbols, from = "2022-01-01")

returns <- prices %>%
  group_by(symbol) %>%
  tq_transmute(select = adjusted,
               mutate_fun = periodReturn,
               period = "daily") %>%
  pivot_wider(names_from = symbol, 
              values_from = daily.returns) %>%
  na.omit()

# Conditional analysis: JPM given BAC quartiles
returns <- returns %>%
  mutate(
    BAC_state = case_when(
      BAC < quantile(BAC, 0.25) ~ "BAC Down (Q1)",
      BAC > quantile(BAC, 0.75) ~ "BAC Up (Q4)",
      TRUE ~ "BAC Normal"
    )
  )

# Conditional statistics
cond_stats <- returns %>%
  group_by(BAC_state) %>%
  summarise(
    `Mean JPM (%)` = mean(JPM) * 100,
    `SD JPM (%)` = sd(JPM) * 100,
    n = n()
  )

knitr::kable(cond_stats, digits = 3,
             caption = "JPM Returns Conditional on BAC State")
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "📊 Show Visualization Code"
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 4.5

# Visualize conditional distributions
ggplot(returns, aes(x = JPM, fill = BAC_state)) +
  geom_density(alpha = 0.5) +
  labs(title = "Conditional Density: f(JPM | BAC state)",
       subtitle = "Bank sector dependence",
       x = "JPM Daily Return",
       y = "Density",
       fill = "Condition") +
  theme_minimal(base_size = 14) +
  scale_fill_manual(values = c("red", "gray50", "green4"))
```

:::
:::::
::::::

## 💰 Case Study: Testing Independence {.smaller}

:::::: {style="font-size: 0.75em; margin-top: -5px;"}

::::: columns
::: {.column width="50%"}

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "📊 Show Chi-Square Test Code"
#| message: false
#| warning: false
#| eval: true

# Statistical test for independence
returns_discrete <- returns %>%
  mutate(
    JPM_cat = cut(JPM, breaks = 5),
    BAC_cat = cut(BAC, breaks = 5)
  )

# Contingency table and chi-square test
cont_table <- table(returns_discrete$JPM_cat, 
                    returns_discrete$BAC_cat)
chi_test <- chisq.test(cont_table)

cat("Chi-Square Test for Independence:\n")
cat("X² =", round(chi_test$statistic, 2), "\n")
cat("df =", chi_test$parameter, "\n")
cat("p-value <", format.pval(chi_test$p.value), "\n\n")
cat("Conclusion:", 
    ifelse(chi_test$p.value < 0.05, 
           "REJECT H₀ → Returns are DEPENDENT", 
           "Cannot reject H₀"))
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "📊 Show Scatter Plot Code"
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 4.5

# Scatter plot showing dependence
ggplot(returns, aes(x = BAC, y = JPM)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "JPM vs BAC Daily Returns",
       subtitle = paste("Correlation:", 
                        round(cor(returns$JPM, returns$BAC), 3)),
       x = "BAC Daily Return",
       y = "JPM Daily Return") +
  theme_minimal(base_size = 14) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5)
```

:::
:::::
::::::

## 💰 Case Study: Key Findings

::::::: {.callout-important style="font-size:32px;"}
## 📊 Analysis Results

::::: columns
:::: {.column width="33%"}
::: fragment
**Conditional Behavior:**

-   When BAC falls, JPM tends to fall
-   Conditional mean shifts significantly
-   Conditional volatility also changes
:::
::::

:::: {.column width="33%"}
::: fragment
**Independence Test:**

-   Chi-square strongly rejects independence
-   Clear positive dependence
-   Returns move together
:::
::::

:::: {.column width="33%"}
::: fragment
**Portfolio Implications:**

1.  **Diversification limited**: Same-sector stocks dependent
2.  **Risk modeling**: Must use joint distributions
3.  **Stress testing**: Conditional analysis essential
:::
::::
:::::
:::::::

## 📝 Quiz #1: Conditional Probability {.quiz-question}

::::: {style="font-size: 36px;"}
If $p(1,2) = 0.15$ and $p_2(2) = 0.30$, what is $P(Y_1 = 1 | Y_2 = 2)$?

- [0.50]{.correct data-explanation="✅ Correct! By Definition 5.5: $P(Y_1 = 1 | Y_2 = 2) = p(1,2)/p_2(2) = 0.15/0.30 = 0.50$."}
- 0.15
- 0.30
- 0.45
:::::

## 📝 Quiz #2: Independence Condition {.quiz-question}

::::: {style="font-size: 32px;"}
For continuous random variables, which condition is equivalent to independence?

- [$f(y_1, y_2) = f_1(y_1) \cdot f_2(y_2)$ for all $(y_1, y_2)$]{.correct data-explanation="✅ Correct! By Theorem 5.4, independence requires the joint density to factor into the product of marginals for ALL values."}
- $f(y_1|y_2) = f_2(y_2)$ for all $y_2$
- $f(y_1, y_2) = f_1(y_1) + f_2(y_2)$ for all $(y_1, y_2)$
- $f(y_1, y_2) > 0$ for all $(y_1, y_2)$
:::::

## 📝 Quiz #3: Factorization Theorem {.quiz-question}

::::: {style="font-size: 32px;"}
$f(y_1, y_2) = 12y_1^2y_2$ for $0 \leq y_1 \leq 1, 0 \leq y_2 \leq 1$. Are $Y_1$ and $Y_2$ independent?

- [Yes, because $f(y_1,y_2) = g(y_1) \cdot h(y_2)$ with rectangular support]{.correct data-explanation="✅ Correct! The region is rectangular and $f(y_1,y_2) = 12y_1^2y_2 = (y_1^2)(12y_2)$ factors into a function of $y_1$ times a function of $y_2$."}
- No, because the powers are different
- No, because the constant 12 cannot be split
- Cannot determine without computing marginals
:::::

## 📝 Quiz #4: Dependent Variables {.quiz-question}

::::: {style="font-size: 34px;"}
$f(y_1, y_2) = 8y_1y_2$ for $0 < y_1 < y_2 < 1$. Why are $Y_1$ and $Y_2$ dependent?

- [The region of support is not rectangular]{.correct data-explanation="✅ Correct! The constraint $y_1 < y_2$ creates a triangular (non-rectangular) region. When the support is not rectangular, Theorem 5.5 cannot be applied and variables are typically dependent."}
- The density cannot be factored
- The coefficient 8 is too large
- The marginal densities don't exist
:::::

## 📝 Summary {.larger}

:::::: {.callout-note style="font-size:36px;"}
## ✅ Key Takeaways

::: incremental
-   **Conditional distributions** describe behavior of one variable given information about another

-   **Discrete case:** $p(y_1|y_2) = p(y_1, y_2)/p_2(y_2)$ when $p_2(y_2) > 0$

-   **Continuous case:** $f(y_1|y_2) = f(y_1, y_2)/f_2(y_2)$ when $f_2(y_2) > 0$

-   **Independence:** Joint = Product of marginals for ALL values

-   **Factorization theorem:** Rectangular support + density factors → independence
:::
::::::

## 📚 Practice Problems

::::: {style="font-size: 32px;"}
::: {.callout-tip}
## 📝 Homework Problems

**Problem 1 (Conditional Discrete):** Using the portfolio asset selection example, find the complete conditional distribution of $Y_2$ (bank stocks) given $Y_1 = 1$ (one tech stock selected).

**Problem 2 (Conditional Continuous):** If stock returns have $f(y_1, y_2) = 4y_1y_2$ for $0 < y_1 < 1, 0 < y_2 < 1$, find $f(y_1|y_2)$ and verify it's a valid density.

**Problem 3 (Independence):** A risk model has $f(y_1, y_2) = 2$ for $0 < y_1 + y_2 < 1$. Determine if the risk factors are independent.

**Problem 4 (Azerbaijan Application):** Download SOCAR and Brent crude oil price data. Test whether their daily returns are independent using a chi-square test.
:::
:::::

## 👋 Thank You! {.center}

::::: {style="font-size: 32px;"}
::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business, ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Expected Values, Covariance, and Correlation

**Reading:** Chapter 5, Sections 5.5-5.8

**Preparation:** Review variance and expected value formulas

**⏰ Reminders:**

- ✅ Complete Practice Problems 1-4
- ✅ Review integration by parts
- ✅ Think about correlation vs dependence
:::
:::
:::::

## ❓ Questions? {.center}

::::: {style="font-size: 36px;"}
::: {.callout-note}
## 💬 Open Discussion

**Key Topics for Discussion:**

- How does conditional expectation relate to regression?

- Can two variables be dependent but have zero correlation?

- How do we use independence in portfolio theory?

- What's the relationship between statistical independence and economic independence?
:::
:::::
