---
title: "Mathematical Statistics"
subtitle: "Conditional Distributions and Independence"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics - Conditional Distributions and Independence"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::: {.learning-objectives}
By the end of this lecture, you will be able to:

- Compute **conditional probability distributions** for both discrete and continuous bivariate random variables

- Apply the **multiplicative law** to decompose joint distributions into marginal and conditional components

- Define and verify **statistical independence** of random variables using distribution function and density criteria

- Use **Theorem 5.5** to efficiently check independence without computing marginal distributions

- Interpret independence and dependence in **financial contexts** such as portfolio diversification and correlated asset returns
:::

## 📋 Overview {.smaller}

::: {style="font-size:50px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Conditional Distributions (Discrete)** – Computing probabilities given partial information

- **Conditional Distributions (Continuous)** – Conditional densities and the "slicing" interpretation

- **Independence of Random Variables** – Definition, properties, and verification methods

- **Theorem 5.5: Easy Independence Check** – Factorization criterion for rectangular support

- **Case Study** – Are stock and bond returns independent? Testing with real market data
:::
:::
:::

## 📖 Motivation: Why Conditional Distributions? {.smaller}

::: {.callout-note}
## 🎯 The Power of Partial Information

In many situations, we have **partial information** about a random outcome and want to update our probabilities accordingly:

::: {.columns}
::: {.column width="50%"}
::: {.fragment}
**Finance Examples:**

- Given that the market is up, what's the probability a specific stock is up?
- Given yesterday's return, what's today's expected return?
- Given a credit rating, what's the default probability?
:::
:::

::: {.column width="50%"}
::: {.fragment}
**Other Examples:**

- Given a patient's age, what's the probability of a disease?
- Given high temperature, what's the probability of rain?
- Given a test score in math, what's the expected score in physics?
:::
:::
:::
:::

**Key Insight:** Conditional distributions capture how knowledge of one variable **changes our beliefs** about another.

## 📖 Recall: Multiplicative Law of Probability {.smaller}

::: {.callout-tip}
## 🔗 Connection to Chapter 2

From basic probability, we know the **multiplicative law**:

$$P(A \cap B) = P(A) \cdot P(B|A) = P(B) \cdot P(A|B)$$

For random variables, this becomes:

$$p(y_1, y_2) = p_1(y_1) \cdot p(y_2|y_1) = p_2(y_2) \cdot p(y_1|y_2)$$

**Interpretation:** The joint probability can be decomposed into a marginal probability times a conditional probability.
:::

This relationship leads us to the formal definition of conditional distributions.

## 📖 Definition: Conditional Probability Function (Discrete) {.smaller}

::: {.callout-note}
## 📝 Definition 5.5: Conditional Discrete Probability Function

If $Y_1$ and $Y_2$ are discrete random variables with joint probability function $p(y_1, y_2)$ and marginal $p_2(y_2) > 0$, then the **conditional probability function** of $Y_1$ given $Y_2 = y_2$ is:

$$p(y_1|y_2) = \frac{p(y_1, y_2)}{p_2(y_2)}$$

Similarly, the conditional probability function of $Y_2$ given $Y_1 = y_1$ (when $p_1(y_1) > 0$) is:

$$p(y_2|y_1) = \frac{p(y_1, y_2)}{p_1(y_1)}$$
:::

**Verification:** For any fixed $y_2$, the conditional pmf sums to 1: $\sum_{y_1} p(y_1|y_2) = 1$

## 📌 Example 1: Committee Selection - Conditional Probability {.smaller}

**Problem:** In the committee selection example (4 Republicans, 3 Democrats, committee of 3), find $P(Y_1 = 2 | Y_2 = 1)$—the probability of exactly 2 Republicans given exactly 1 Democrat is selected.

**Solution:**

. . .

Recall: $Y_1$ = number of Republicans, $Y_2$ = number of Democrats, with $Y_1 + Y_2 = 3$.

. . .

**Step 1: Find $p(2, 1)$ from the joint distribution**

$$p(2, 1) = \frac{\binom{4}{2}\binom{3}{1}}{\binom{7}{3}} = \frac{6 \cdot 3}{35} = \frac{18}{35}$$

. . .

**Step 2: Find $p_2(1)$ (marginal probability of 1 Democrat)**

$$p_2(1) = p(2, 1) = \frac{18}{35}$$ (only one way to get exactly 1 Democrat with the constraint)

. . .

**Step 3: Compute conditional probability**

$$p(Y_1 = 2 | Y_2 = 1) = \frac{p(2, 1)}{p_2(1)} = \frac{18/35}{18/35} = 1$$

**Interpretation:** If exactly 1 Democrat is selected, we **must** have exactly 2 Republicans (since the committee has 3 members).

## 📖 Definition: Conditional Density Function (Continuous) {.smaller}

::: {.callout-note}
## 📝 Definition 5.7: Conditional Density Function

If $Y_1$ and $Y_2$ are jointly continuous random variables with joint pdf $f(y_1, y_2)$ and marginal pdf $f_2(y_2) > 0$, the **conditional density function** of $Y_1$ given $Y_2 = y_2$ is:

$$f(y_1|y_2) = \frac{f(y_1, y_2)}{f_2(y_2)}$$

The **conditional distribution function** is:

$$F(y_1|y_2) = P(Y_1 \leq y_1 | Y_2 = y_2) = \int_{-\infty}^{y_1} f(t|y_2) \, dt$$
:::

**Geometric Interpretation:** The conditional density is like taking a "slice" through the joint density surface at a fixed value of $y_2$, then renormalizing to integrate to 1.

## 📌 Example 2: Soft-Drink Machine (Conditional Density) {.smaller}

**Problem:** A soft-drink machine has $Y_1$ = amount sold (in gallons) and $Y_2$ = amount in supply, with joint pdf:

$$f(y_1, y_2) = \frac{1}{2} \quad \text{for } 0 \leq y_1 \leq y_2 \leq 2$$

Find $P(Y_1 < 0.5 | Y_2 = 1.5)$ and $P(Y_1 < 0.5 | Y_2 = 2)$.

**Solution:**

. . .

**Step 1: Find marginal $f_2(y_2)$**

$$f_2(y_2) = \int_0^{y_2} \frac{1}{2} \, dy_1 = \frac{y_2}{2} \quad \text{for } 0 \leq y_2 \leq 2$$

. . .

**Step 2: Find conditional density $f(y_1|y_2)$**

$$f(y_1|y_2) = \frac{f(y_1, y_2)}{f_2(y_2)} = \frac{1/2}{y_2/2} = \frac{1}{y_2} \quad \text{for } 0 \leq y_1 \leq y_2$$

. . .

**Step 3: Compute conditional probabilities**

$$P(Y_1 < 0.5 | Y_2 = 1.5) = \int_0^{0.5} \frac{1}{1.5} \, dy_1 = \frac{0.5}{1.5} = \boxed{\frac{1}{3}}$$

$$P(Y_1 < 0.5 | Y_2 = 2) = \int_0^{0.5} \frac{1}{2} \, dy_1 = \frac{0.5}{2} = \boxed{\frac{1}{4}}$$

## 📌 Example 2: Interpretation {.smaller}

::: {.callout-important}
## 📊 Key Insight: Conditional Probabilities Change!

Notice that:

- $P(Y_1 < 0.5 | Y_2 = 1.5) = 1/3 \approx 33.3\%$
- $P(Y_1 < 0.5 | Y_2 = 2) = 1/4 = 25\%$

**The probability of selling less than 0.5 gallons depends on how much is in supply!**
:::

**Business Interpretation:**

- With a smaller supply (1.5 gallons), there's a higher chance of low sales because customers may not find what they want
- With a larger supply (2 gallons), sales tend to be higher, so low sales are less likely
- This demonstrates **dependence** between supply and sales—they are NOT independent!

## 🎮 Interactive: Conditional Distribution Visualizer {.smaller}

::: {style="font-size: 0.8em;"}

**Explore Conditional Densities:** See how the conditional distribution changes as you vary the conditioning value.

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof y2_cond = Inputs.range([0.5, 2], {
  value: 1.5, 
  step: 0.1, 
  label: "Conditioning value y₂:"
})

// For f(y1,y2) = 1/2 on 0 <= y1 <= y2 <= 2
// Marginal f2(y2) = y2/2
// Conditional f(y1|y2) = 1/y2 on [0, y2]

md`**Joint density:**  
$f(y_1, y_2) = \\frac{1}{2}$  
on $0 \\leq y_1 \\leq y_2 \\leq 2$

**Marginal:**  
$f_2(y_2) = \\frac{y_2}{2} = ${(y2_cond/2).toFixed(3)}$

**Conditional:**  
$f(y_1|y_2=${y2_cond.toFixed(1)}) = \\frac{1}{${y2_cond.toFixed(1)}} = ${(1/y2_cond).toFixed(3)}$

on $[0, ${y2_cond.toFixed(1)}]$`
```

**Observations:**  

- As $y_2$ increases, the conditional density becomes flatter (more spread out)

- The conditional density always integrates to 1 over its support

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// Generate conditional density data
cond_y1_vals = d3.range(0, y2_cond + 0.01, 0.01)

cond_density_data = cond_y1_vals.map(y1 => ({
  y1: y1,
  density: 1/y2_cond
}))

// Also show what happens outside support
full_y1_range = d3.range(0, 2.1, 0.01)

full_density_data = full_y1_range.map(y1 => ({
  y1: y1,
  density: y1 <= y2_cond ? 1/y2_cond : 0
}))

Plot.plot({
  width: 750,
  height: 450,
  marginLeft: 60,
  marginBottom: 50,
  x: {
    label: "y₁ (amount sold)",
    domain: [0, 2.2]
  },
  y: {
    label: "Conditional Density f(y₁|y₂)",
    domain: [0, 2.5]
  },
  marks: [
    Plot.areaY(cond_density_data, {
      x: "y1", 
      y: "density", 
      fill: "steelblue", 
      opacity: 0.5
    }),
    Plot.line(full_density_data, {
      x: "y1", 
      y: "density", 
      stroke: "darkblue", 
      strokeWidth: 3
    }),
    Plot.ruleX([y2_cond], {
      stroke: "red", 
      strokeWidth: 2, 
      strokeDasharray: "5,5"
    }),
    Plot.ruleY([0]),
    Plot.text([[y2_cond + 0.1, 0.5]], {
      text: [`y₂ = ${y2_cond.toFixed(1)}`],
      fill: "red",
      fontSize: 14
    })
  ],
  caption: html`<b>Conditional density:</b> f(y₁|y₂=${y2_cond.toFixed(1)}) = ${(1/y2_cond).toFixed(3)} on [0, ${y2_cond.toFixed(1)}]. <span style="color:red">┊</span> Upper bound`
})
```

:::

:::

:::

## 📖 Definition: Independent Random Variables {.smaller}

::: {.callout-note}
## 📝 Definition 5.8: Independence of Random Variables

Two random variables $Y_1$ and $Y_2$ are said to be **independent** if and only if:

$$F(y_1, y_2) = F_1(y_1) \cdot F_2(y_2) \quad \text{for all } (y_1, y_2)$$

where $F(y_1, y_2)$ is the joint CDF and $F_1, F_2$ are the marginal CDFs.

**Equivalent Conditions (Theorem 5.4):**

- **Discrete case:** $p(y_1, y_2) = p_1(y_1) \cdot p_2(y_2)$ for all $(y_1, y_2)$

- **Continuous case:** $f(y_1, y_2) = f_1(y_1) \cdot f_2(y_2)$ for all $(y_1, y_2)$
:::

**Intuition:** Independence means knowing the value of $Y_1$ provides **no information** about $Y_2$, and vice versa. The variables do not "influence" each other.

## 📌 Example 3: Dice Tossing - Showing Independence {.smaller}

**Problem:** For two fair dice tosses, verify that $Y_1$ (die 1) and $Y_2$ (die 2) are independent.

**Solution:**

. . .

**Joint probability function:**
$$p(y_1, y_2) = \frac{1}{36} \quad \text{for all } y_1, y_2 \in \{1, 2, 3, 4, 5, 6\}$$

. . .

**Marginal distributions:**
$$p_1(y_1) = \sum_{y_2=1}^{6} \frac{1}{36} = \frac{6}{36} = \frac{1}{6} \quad \text{for } y_1 \in \{1, \ldots, 6\}$$

$$p_2(y_2) = \frac{1}{6} \quad \text{(by symmetry)}$$

. . .

**Check factorization:**
$$p_1(y_1) \cdot p_2(y_2) = \frac{1}{6} \cdot \frac{1}{6} = \frac{1}{36} = p(y_1, y_2) \checkmark$$

**Conclusion:** Since $p(y_1, y_2) = p_1(y_1) \cdot p_2(y_2)$ for all $(y_1, y_2)$, the dice outcomes are **independent**.

## 📌 Example 4: Committee Selection - Showing Dependence {.smaller}

**Problem:** For the committee selection (4 Republicans, 3 Democrats, select 3), are $Y_1$ and $Y_2$ independent?

**Solution:**

. . .

Recall: $p(2, 1) = \frac{18}{35}$

**Marginal distributions:**

$p_1(2) = \frac{18}{35}$ (computed by summing over $y_2$)

$p_2(1) = \frac{18}{35}$ (computed by summing over $y_1$)

. . .

**Check if $p(2, 1) = p_1(2) \cdot p_2(1)$:**

$$p_1(2) \cdot p_2(1) = \frac{18}{35} \cdot \frac{18}{35} = \frac{324}{1225} \neq \frac{18}{35} = p(2, 1)$$

. . .

**Conclusion:** Since $p(y_1, y_2) \neq p_1(y_1) \cdot p_2(y_2)$, the variables are **NOT independent**.

**Intuition:** Knowing the number of Republicans tells you exactly how many Democrats (since total is fixed at 3).

## 📌 Example 5: Continuous Case - Showing Independence {.smaller}

**Problem:** Let $f(y_1, y_2) = 6y_1 y_2^2$ for $0 < y_1 < 1$ and $0 < y_2 < 1$. Are $Y_1$ and $Y_2$ independent?

**Solution:**

. . .

**Find marginal $f_1(y_1)$:**
$$f_1(y_1) = \int_0^1 6y_1 y_2^2 \, dy_2 = 6y_1 \cdot \frac{1}{3} = 2y_1 \quad \text{for } 0 < y_1 < 1$$

. . .

**Find marginal $f_2(y_2)$:**
$$f_2(y_2) = \int_0^1 6y_1 y_2^2 \, dy_1 = 6y_2^2 \cdot \frac{1}{2} = 3y_2^2 \quad \text{for } 0 < y_2 < 1$$

. . .

**Check factorization:**
$$f_1(y_1) \cdot f_2(y_2) = 2y_1 \cdot 3y_2^2 = 6y_1 y_2^2 = f(y_1, y_2) \checkmark$$

**Conclusion:** $Y_1$ and $Y_2$ are **independent**.

## 📌 Example 6: Non-Rectangular Support - Showing Dependence {.smaller}

**Problem:** For the gasoline problem with $f(y_1, y_2) = 3y_1$ on $0 \leq y_2 \leq y_1 \leq 1$, are $Y_1$ and $Y_2$ independent?

**Solution:**

. . .

From Lecture 1, we found:

- $f_1(y_1) = 3y_1^2$ for $0 \leq y_1 \leq 1$
- $f_2(y_2) = \frac{3}{2}(1 - y_2^2)$ for $0 \leq y_2 \leq 1$

. . .

**Check at a specific point, say $(y_1, y_2) = (0.5, 0.3)$:**

$$f_1(0.5) \cdot f_2(0.3) = 3(0.5)^2 \cdot \frac{3}{2}(1 - 0.09) = 0.75 \cdot 1.365 = 1.024$$

$$f(0.5, 0.3) = 3(0.5) = 1.5$$

. . .

Since $1.024 \neq 1.5$, we have $f(y_1, y_2) \neq f_1(y_1) \cdot f_2(y_2)$.

**Conclusion:** $Y_1$ and $Y_2$ are **NOT independent**.

**Key Insight:** The constraint $y_2 \leq y_1$ (you can't sell more than you have) creates dependence!

## 🧮 Theorem 5.5: Easy Independence Check {.smaller}

::: {.callout-important}
## Theorem 5.5: Factorization Criterion for Independence

Let $Y_1$ and $Y_2$ have joint pdf $f(y_1, y_2)$. If:

1. The region where $f(y_1, y_2) > 0$ is a **rectangle** (possibly infinite): $\{(y_1, y_2): a < y_1 < b, c < y_2 < d\}$

2. The joint density **factors** as: $f(y_1, y_2) = g(y_1) \cdot h(y_2)$ for some functions $g$ and $h$

Then $Y_1$ and $Y_2$ are **independent**, with marginals proportional to $g$ and $h$.
:::

**Power of Theorem 5.5:** You can verify independence **without computing marginals explicitly!**

**Example:** $f(y_1, y_2) = 6y_1 y_2^2$ on $[0,1] \times [0,1]$ (rectangle) factors as $g(y_1) = y_1$ and $h(y_2) = y_2^2$, so $Y_1 \perp Y_2$.

## 📌 Example 7: Using Theorem 5.5 {.smaller}

**Problem:** Let $f(y_1, y_2) = 2y_1$ for $0 < y_1 < 1$ and $0 < y_2 < 1$. Use Theorem 5.5 to determine if $Y_1$ and $Y_2$ are independent.

**Solution:**

. . .

**Step 1: Check the support**

The region $\{0 < y_1 < 1, 0 < y_2 < 1\}$ is the unit square—a **rectangle**. ✓

. . .

**Step 2: Check factorization**

$$f(y_1, y_2) = 2y_1 = (2y_1) \cdot (1) = g(y_1) \cdot h(y_2)$$

where $g(y_1) = 2y_1$ and $h(y_2) = 1$. ✓

. . .

**Conclusion:** By Theorem 5.5, $Y_1$ and $Y_2$ are **independent**.

**Observation:** Even though $f(y_1, y_2)$ only depends on $y_1$, this doesn't automatically mean dependence! The rectangular support allows factorization, implying $Y_2 \sim \text{Uniform}(0,1)$.

## 🎮 Interactive: Independence vs Dependence {.smaller}

::: {style="font-size: 0.8em;"}

**Compare Distributions:** Toggle between independent and dependent bivariate densities.

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof dist_type = Inputs.radio(
  ["Independent", "Dependent"], 
  {value: "Independent", label: "Distribution Type:"}
)

md`**${dist_type} Variables**

${dist_type === "Independent" ? 
  "**Joint pdf:** $f(y_1, y_2) = 4y_1 y_2$\n\non $[0,1] \\times [0,1]$\n\n**Marginals:**\n- $f_1(y_1) = 2y_1$\n- $f_2(y_2) = 2y_2$\n\n**Check:** $f = f_1 \\cdot f_2$ ✓" : 
  "**Joint pdf:** $f(y_1, y_2) = 3y_1$\n\non $0 \\leq y_2 \\leq y_1 \\leq 1$\n\n**Marginals:**\n- $f_1(y_1) = 3y_1^2$\n- $f_2(y_2) = \\frac{3}{2}(1-y_2^2)$\n\n**Check:** $f \\neq f_1 \\cdot f_2$ ✗"}`
```

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// Generate contour data based on distribution type
grid_size2 = 40
y1_vals2 = d3.range(0.02, 1, 1/grid_size2)
y2_vals2 = d3.range(0.02, 1, 1/grid_size2)

indep_dep_data = {
  let data = [];
  for (let y1 of y1_vals2) {
    for (let y2 of y2_vals2) {
      let density;
      if (dist_type === "Independent") {
        // f(y1, y2) = 4*y1*y2 on [0,1]x[0,1]
        density = 4 * y1 * y2;
      } else {
        // f(y1, y2) = 3*y1 on 0 <= y2 <= y1 <= 1
        density = (y2 <= y1) ? 3 * y1 : 0;
      }
      data.push({y1: y1, y2: y2, density: density});
    }
  }
  return data;
}

Plot.plot({
  width: 750,
  height: 500,
  marginLeft: 60,
  marginBottom: 50,
  color: {
    type: "linear",
    scheme: dist_type === "Independent" ? "blues" : "oranges",
    legend: true,
    label: "Density"
  },
  x: {
    label: "y₁",
    domain: [0, 1]
  },
  y: {
    label: "y₂",
    domain: [0, 1]
  },
  marks: [
    Plot.contour(indep_dep_data, {
      x: "y1",
      y: "y2",
      fill: "density",
      blur: 3,
      thresholds: 15
    }),
    dist_type === "Dependent" ? Plot.line([[0,0], [1,1]], {stroke: "white", strokeWidth: 2, strokeDasharray: "5,5"}) : null,
    Plot.frame()
  ],
  caption: dist_type === "Independent" ? 
    html`<b>Independent:</b> Contours are "aligned" with axes (rectangular pattern)` :
    html`<b>Dependent:</b> Contours follow constraint y₂ ≤ y₁ (triangular support)`
})
```

:::

:::

:::

## 🤝 Think-Pair-Share: Risk Assessment {.r-fit-text}

::: {style="font-size: 28px"}
::: {.callout-note}
## 💭 Student Engagement Activity (5 minutes)

**Scenario:** A company is evaluating two investment projects with the following joint probability distribution for success:

| Project A \ Project B | Fail (0) | Succeed (1) |
|:---:|:---:|:---:|
| **Fail (0)** | 0.20 | 0.15 |
| **Succeed (1)** | 0.10 | 0.55 |

**Think (1.5 minutes):** Work individually

- If Project A succeeds, what is $P(\text{Project B succeeds} | \text{Project A succeeds})$?
- If Project A fails, what is $P(\text{Project B succeeds} | \text{Project A fails})$?
- Are the project outcomes independent?

**Pair (2 minutes):** Discuss with a partner

- Compare your calculations
- What does this dependence/independence mean for the company's risk?

**Share (1.5 minutes):** Class discussion

- How should this affect the company's diversification strategy?
:::
:::

## 💰 Case Study: Are Stock Returns Independent? (Real Data)

::: {style="font-size:35px"}
::: {.columns}
::: {.column width="50%"}
::: {.callout-note}
## 📈 Testing Independence with Real Data

**Context**: A fundamental question in portfolio theory is whether asset returns are independent. If returns are independent, diversification is maximally effective.

**Key Questions**:

- Are stock (SPY) and bond (TLT) returns independent?

- How do we test independence with real data?

- What does dependence imply for portfolio construction?

:::
:::

::: {.column width="50%"}
::: {.callout-tip}
## 📊 Data Source

We analyze **daily returns** for:

- **SPY**: S&P 500 ETF (stocks)
- **TLT**: 20+ Year Treasury Bond ETF (bonds)

**Source**: Yahoo Finance API via `quantmod` package

**Period**: Last 3 years of daily data

**Why these assets?**: Traditional portfolio theory suggests stocks and bonds should have low or negative correlation
:::
:::
:::
:::

## 💰 Case Study: Data Collection and Summary {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Load required libraries
library(quantmod)
library(tidyverse)
library(knitr)

# Set date range (last 3 years)
end_date <- Sys.Date()
start_date <- end_date - (3 * 365)

# Download data
getSymbols("SPY", from = start_date, 
           to = end_date, auto.assign = TRUE)
getSymbols("TLT", from = start_date, 
           to = end_date, auto.assign = TRUE)

# Calculate daily log returns
spy_returns <- dailyReturn(SPY, type = "log")
tlt_returns <- dailyReturn(TLT, type = "log")

# Combine into data frame
returns_df <- data.frame(
  Date = index(spy_returns),
  SPY = as.numeric(spy_returns),
  TLT = as.numeric(tlt_returns)
) %>% na.omit()

cat(sprintf("Observations: %d\n", nrow(returns_df)))
cat(sprintf("Period: %s to %s\n", 
            min(returns_df$Date), max(returns_df$Date)))
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Summary statistics
cat("=== SPY (Stocks) ===\n")
cat(sprintf("Mean: %.4f%%\n", mean(returns_df$SPY) * 100))
cat(sprintf("Std Dev: %.4f%%\n", sd(returns_df$SPY) * 100))

cat("\n=== TLT (Bonds) ===\n")
cat(sprintf("Mean: %.4f%%\n", mean(returns_df$TLT) * 100))
cat(sprintf("Std Dev: %.4f%%\n", sd(returns_df$TLT) * 100))

# Correlation (preview of Section 5.7)
corr_val <- cor(returns_df$SPY, returns_df$TLT)
cat(sprintf("\n=== Correlation ===\n"))
cat(sprintf("Corr(SPY, TLT): %.4f\n", corr_val))

# If independent, correlation should be near 0
cat(sprintf("\nNote: If independent, correlation ≈ 0\n"))
cat(sprintf("Observed correlation suggests %s\n",
            ifelse(abs(corr_val) < 0.1, "possible independence",
                   "dependence")))
```

:::
:::
:::

## 💰 Case Study: Visual Tests of Independence {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

# Scatter plot - if independent, no pattern
ggplot(returns_df, aes(x = SPY, y = TLT)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(title = "SPY vs TLT Daily Returns",
       subtitle = "Looking for patterns (independence = no pattern)",
       x = "SPY Return (Stocks)",
       y = "TLT Return (Bonds)") +
  theme_minimal(base_size = 11)
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

# Conditional density comparison
# If independent, density of TLT should not change based on SPY

# Divide SPY into terciles
returns_df <- returns_df %>%
  mutate(SPY_tercile = cut(SPY, 
                           breaks = quantile(SPY, c(0, 1/3, 2/3, 1)),
                           labels = c("SPY Low", "SPY Mid", "SPY High"),
                           include.lowest = TRUE))

# Plot conditional distributions
ggplot(returns_df, aes(x = TLT, fill = SPY_tercile)) +
  geom_density(alpha = 0.5) +
  labs(title = "TLT Returns by SPY Performance",
       subtitle = "If independent, distributions should overlap",
       x = "TLT Return (Bonds)",
       y = "Density",
       fill = "SPY Tercile") +
  theme_minimal(base_size = 11) +
  scale_fill_manual(values = c("red", "gray", "green"))
```

:::
:::
:::

## 💰 Case Study: Formal Independence Test {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Chi-square test of independence
# Discretize returns into categories
returns_df <- returns_df %>%
  mutate(
    SPY_cat = cut(SPY, breaks = c(-Inf, -0.01, 0.01, Inf),
                  labels = c("Down", "Flat", "Up")),
    TLT_cat = cut(TLT, breaks = c(-Inf, -0.01, 0.01, Inf),
                  labels = c("Down", "Flat", "Up"))
  )

# Contingency table
cont_table <- table(returns_df$SPY_cat, returns_df$TLT_cat)
cat("Contingency Table:\n")
print(cont_table)

# Chi-square test
chi_test <- chisq.test(cont_table)
cat(sprintf("\nChi-square statistic: %.2f\n", chi_test$statistic))
cat(sprintf("Degrees of freedom: %d\n", chi_test$parameter))
cat(sprintf("P-value: %.6f\n", chi_test$p.value))
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Interpretation
cat("=== Test Interpretation ===\n\n")

if(chi_test$p.value < 0.05) {
  cat("Result: REJECT independence (p < 0.05)\n\n")
  cat("SPY and TLT returns are NOT independent.\n")
  cat("The relationship is statistically significant.\n\n")
} else {
  cat("Result: Cannot reject independence (p >= 0.05)\n\n")
  cat("Insufficient evidence against independence.\n\n")
}

# Expected vs observed
cat("Expected counts (if independent):\n")
print(round(chi_test$expected, 1))

cat("\nResiduals (Obs - Expected):\n")
print(round(chi_test$observed - chi_test$expected, 1))
```

:::
:::
:::

## 💰 Case Study: Key Findings {.smaller}

::: {style="font-size: 32.5px;"}
::: {.callout-important}
## 📊 Analysis Results

::: {.columns}

::: {.column width="33%"}
::: {.fragment}
**Evidence Against Independence:**

- Chi-square test typically rejects independence

- Scatter plot shows negative relationship

- Conditional distributions shift based on SPY performance
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Economic Interpretation:**

- Stocks and bonds often move in **opposite directions**

- "Flight to safety": When stocks fall, investors buy bonds

- This negative correlation is the basis for stock-bond portfolio diversification
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Portfolio Implications:**

1. **Diversification works**: Negative correlation reduces overall portfolio risk

2. **Not perfect hedge**: Correlation varies over time

3. **Crisis periods**: Correlations can break down (both assets fall together)
:::
:::

:::
:::
:::

## 📝 Quiz #1: Conditional Probability Computation {.smaller .quiz-question}

If $p(y_1, y_2) = 0.15$ and $p_2(y_2) = 0.30$, what is $p(y_1 | y_2)$?

- [0.50]{.correct data-explanation="✅ Correct! p(y₁|y₂) = p(y₁,y₂)/p₂(y₂) = 0.15/0.30 = 0.50"}
- 0.15
- 0.45
- 2.00

## 📝 Quiz #2: Identifying Independence {.smaller .quiz-question}

Let $f(y_1, y_2) = 2e^{-y_1}e^{-2y_2}$ for $y_1 > 0$ and $y_2 > 0$. Are $Y_1$ and $Y_2$ independent?

- [Yes, because the joint density factors and the support is rectangular]{.correct data-explanation="✅ Correct! The density factors as g(y₁)·h(y₂) = (e^{-y₁})·(2e^{-2y₂}) and the support (0,∞)×(0,∞) is rectangular, so by Theorem 5.5, they are independent."}
- Yes, because the density is positive everywhere
- No, because the density depends on both $y_1$ and $y_2$
- Cannot determine without computing marginals

## 📝 Quiz #3: Independence and Conditional Distributions {.smaller .quiz-question}

If $Y_1$ and $Y_2$ are independent, what is true about the conditional distribution $f(y_1|y_2)$?

- [It equals the marginal distribution $f_1(y_1)$]{.correct data-explanation="✅ Correct! If independent, f(y₁|y₂) = f(y₁,y₂)/f₂(y₂) = f₁(y₁)·f₂(y₂)/f₂(y₂) = f₁(y₁). Knowing y₂ doesn't change the distribution of Y₁."}
- It equals zero
- It equals the joint distribution $f(y_1, y_2)$
- It depends on the specific value of $y_2$

## 📝 Summary {.smaller}

::: {style="font-size: 32.5px"}
::: {.summary-box}
**✅ Key Takeaways**

- **Conditional distributions** describe the probability distribution of one variable given the value of another: $p(y_1|y_2) = p(y_1, y_2)/p_2(y_2)$

- **Independence** means the joint distribution factors into marginals: $f(y_1, y_2) = f_1(y_1) \cdot f_2(y_2)$, equivalently, knowing $Y_2$ provides no information about $Y_1$

- **Theorem 5.5** provides a quick check: if the support is rectangular and the density factors, the variables are independent

- **Non-rectangular support** (like $y_2 \leq y_1$) automatically implies dependence

- **Financial insight**: Stock and bond returns are typically NOT independent—their negative correlation enables diversification benefits
:::
:::

## 📚 Practice Problems

::: {.callout-tip}
## 📝 Homework Problems

**Problem 1 (Conditional PMF):** Given a joint probability table for two assets' returns, compute all conditional probabilities and verify they sum to 1.

**Problem 2 (Conditional PDF):** For $f(y_1, y_2) = 8y_1y_2$ on $0 < y_1 < y_2 < 1$, find $f(y_1|y_2)$ and compute $P(Y_1 < 0.3 | Y_2 = 0.5)$.

**Problem 3 (Testing Independence):** Determine whether the following pairs are independent:
(a) $f(y_1, y_2) = e^{-y_1-y_2}$ on $y_1 > 0, y_2 > 0$
(b) $f(y_1, y_2) = 24y_1y_2$ on $0 < y_1 < 1, 0 < y_2 < 1-y_1$

**Problem 4 (Financial Application):** A portfolio has two assets. If returns are independent with given marginal distributions, find the probability both assets have positive returns.
:::

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business

ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Expected Values, Covariance, and Correlation

**Reading:** Chapter 5, Sections 5.5-5.7

**Preparation:** Review expected value and variance from Chapter 4

**⏰ Reminders:**

✅ Complete Practice Problems 1-4

✅ Think about real-world examples of dependent variables

✅ Consider: How does dependence affect portfolio risk?

✅ Work hard!
:::
:::

## ❓ Questions? {.center}

::: {.callout-note}
## 💬 Open Discussion

**Key Topics for Discussion:**

- Why is independence such an important concept in statistics and finance?

- When is it reasonable to assume independence between random variables?

- How does dependence between asset returns affect the risk-return tradeoff in portfolios?

- What are some real-world examples where independence is clearly violated?
:::
