---
title: "Mathematical Statistics"
subtitle: "Bivariate and Multivariate Probability Distributions"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ../ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics - Bivariate and Multivariate Distributions"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::: {.learning-objectives}
By the end of this lecture, you will be able to:

- Define and interpret **joint probability functions** for discrete bivariate random variables and verify their properties

- Compute **joint probability density functions** for continuous bivariate random variables and calculate probabilities over specified regions

- Derive **marginal probability distributions** from joint distributions using summation (discrete) or integration (continuous)

- Visualize bivariate distributions using **3D surfaces**, contour plots, and understand geometric interpretations

- Apply multivariate distribution concepts to **portfolio analysis** and financial risk assessment using real market data
:::

## 📋 Overview

::: {style="font-size:38px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Introduction to Multivariate Distributions** – Why study joint distributions? Applications in finance and economics

- **Joint Probability Functions (Discrete)** – Definition, properties, and examples with probability tables

- **Joint Probability Density Functions (Continuous)** – Definition, properties, and integration over regions

- **Marginal Distributions** – Extracting individual variable distributions from joint distributions

- **Case Study** – Portfolio returns analysis using real stock market data
:::
:::
:::

## 📖 Why Study Multivariate Distributions?

::: {.callout-note}
## 🎯 Motivation

In the real world, random variables rarely occur in isolation. We often need to study **relationships between multiple random variables**:

::: {.columns}
::: {.column width="50%"}
::: {.fragment}
**Finance Applications:**

- Portfolio returns (multiple stocks)
- Risk factors affecting asset prices
- Interest rates and inflation
- Credit risk and market risk
:::
:::

::: {.column width="50%"}
::: {.fragment}
**Other Applications:**

- Height and weight of individuals
- Temperature and humidity
- Supply and demand quantities
- Test scores across subjects
:::
:::
:::
:::

**Key Question:** How do we mathematically describe the **joint behavior** of two or more random variables?

## 📖 Definition: Joint Probability Function (Discrete) {.larger}

::: {.callout-note}
## 📝 Definition 5.1: Bivariate Probability Function

Let $Y_1$ and $Y_2$ be discrete random variables. The **joint (bivariate) probability function** for $Y_1$ and $Y_2$ is:

$$p(y_1, y_2) = P(Y_1 = y_1, Y_2 = y_2)$$

for all pairs $(y_1, y_2)$ in the joint sample space.

**Interpretation:** $p(y_1, y_2)$ gives the probability that $Y_1$ takes value $y_1$ **and** $Y_2$ takes value $y_2$ simultaneously.
:::

**Notation:** We write $(Y_1, Y_2)$ to denote a **bivariate random variable** (or random vector) consisting of two components.

## 🧮 Theorem: Properties of Joint Probability Functions {.larger}

::: {.callout-important}
## Theorem 5.1: Properties of Joint Probability Functions

For any joint probability function $p(y_1, y_2)$:

::: {.columns}

::: {.column width='50%' .fragment}

**Property 1 (Non-negativity):**
$$p(y_1, y_2) \geq 0 \quad \text{for all } (y_1, y_2)$$

:::

::: {.column width='50%' .fragment}

**Property 2 (Normalization):**
$$\sum_{y_1} \sum_{y_2} p(y_1, y_2) = 1$$

The double sum is taken over all possible pairs $(y_1, y_2)$ in the sample space.
:::
:::
:::

**Intuition:** Probabilities must be non-negative and sum to 1 across all possible outcomes—the same principles as univariate distributions, extended to pairs!

## 📌 Example 1: Dice Tossing Experiment {.smaller}

**Problem:** Consider tossing two fair dice. Let $Y_1$ = value on die 1 and $Y_2$ = value on die 2. Find the joint probability function.

**Solution:**

. . .

Since each die has 6 faces and the tosses are independent:

$$p(y_1, y_2) = P(Y_1 = y_1) \cdot P(Y_2 = y_2) = \frac{1}{6} \cdot \frac{1}{6} = \frac{1}{36}$$

. . .

for all $y_1, y_2 \in \{1, 2, 3, 4, 5, 6\}$.

::: {.columns}
::: {.column weight='50%' .fragment}
**Sample Space:** 36 equally likely outcomes

$(1,1), (1,2), \ldots, (6,6)$
:::

::: {.column weight='50%' .fragment}
**Verification:**
$$\sum_{y_1=1}^{6} \sum_{y_2=1}^{6} \frac{1}{36} = 36 \cdot \frac{1}{36} = 1 \checkmark$$
:::
:::

## 📌 Example 2: Committee Selection (Discrete) {.smaller}

**Problem:** A committee of 3 is selected at random from 4 Republicans and 3 Democrats. Let $Y_1$ = number of Republicans and $Y_2$ = number of Democrats selected.

**Solution:**

::: {.columns}
::: {.column weight='50%' .fragment}
The joint probability function is:

$$p(y_1, y_2) = \frac{\binom{4}{y_1} \binom{3}{y_2}}{\binom{7}{3}}$$

where $y_1 + y_2 = 3$, $y_1 \in \{0, 1, 2, 3\}$, $y_2 \in \{0, 1, 2, 3\}$.
:::

::: {.column weight='50%' .fragment}
**Joint Probability Table:**

| $y_1 \backslash y_2$ | 0 | 1 | 2 | 3 |
|:---:|:---:|:---:|:---:|:---:|
| **0** | 0 | 0 | 0 | $\frac{1}{35}$ |
| **1** | 0 | 0 | $\frac{12}{35}$ | 0 |
| **2** | 0 | $\frac{18}{35}$ | 0 | 0 |
| **3** | $\frac{4}{35}$ | 0 | 0 | 0 |

**Note:** Only cells where $y_1 + y_2 = 3$ have non-zero probabilities.
:::
:::

## 📖 Definition: Joint Distribution Function {.smaller}

::: {.callout-note}
## 📝 Definition 5.2: Joint Distribution Function (CDF)

The **joint distribution function** (or joint CDF) for random variables $Y_1$ and $Y_2$ is:

$$F(y_1, y_2) = P(Y_1 \leq y_1, Y_2 \leq y_2)$$

for $-\infty < y_1 < \infty$ and $-\infty < y_2 < \infty$.

**Interpretation:** $F(y_1, y_2)$ gives the probability that $Y_1$ is at most $y_1$ **and** $Y_2$ is at most $y_2$.
:::

::: {.callout-tip}
## 📐 Geometric Interpretation

**The "Southwest Rectangle":** $F(y_1, y_2)$ represents the probability accumulated in the rectangular region from $(-\infty, -\infty)$ to $(y_1, y_2)$.


::: {.columns}
::: {.column width="50%"}
**For Discrete Variables:**

- Sum all probabilities $p(y_1', y_2')$ where $y_1' \leq y_1$ and $y_2' \leq y_2$
- Think: "All points to the left and below $(y_1, y_2)$"

$$F(y_1, y_2) = \sum_{y_1' \leq y_1} \sum_{y_2' \leq y_2} p(y_1', y_2')$$
:::

::: {.column width="50%"}
**For Continuous Variables:**

- Integrate the joint pdf over the region
- The volume under $f(y_1, y_2)$ from $-\infty$ to the point $(y_1, y_2)$

$$F(y_1, y_2) = \int_{-\infty}^{y_1} \int_{-\infty}^{y_2} f(t_1, t_2) \, dt_2 \, dt_1$$
:::
:::

**Visual:** Imagine looking at the $(y_1, y_2)$ plane. The CDF at point $(a, b)$ captures all probability mass in the quadrant below and to the left of that point.
:::

## 🧮 Theorem: Properties of Joint CDFs {.larger}

::: {.callout-important}
## Theorem 5.2: Properties of Bivariate CDFs

For any joint distribution function $F(y_1, y_2)$:

1. $\lim_{y_1 \to -\infty} F(y_1, y_2) = \lim_{y_2 \to -\infty} F(y_1, y_2) = 0$

2. $\lim_{y_1 \to \infty, y_2 \to \infty} F(y_1, y_2) = 1$

3. If $y_1^* \geq y_1$ and $y_2^* \geq y_2$, then $F(y_1^*, y_2^*) \geq F(y_1, y_2)$ (non-decreasing)

4. $F(y_1, y_2)$ is right-continuous in each argument
:::

**Intuition:** The CDF starts at 0, increases monotonically, and approaches 1 as both variables go to infinity.

## 📖 Definition: Joint Probability Density Function (Continuous) {.larger}

::: {.callout-note style="margin-top: -10px;"}
## 📝 Definition 5.3: Joint Probability Density Function

Random variables $Y_1$ and $Y_2$ are said to be **jointly continuous** if their joint distribution function $F(y_1, y_2)$ can be written as:

$$F(y_1, y_2) = \int_{-\infty}^{y_1} \int_{-\infty}^{y_2} f(t_1, t_2) \, dt_2 \, dt_1$$

for some non-negative function $f(y_1, y_2)$ called the **joint probability density function** (pdf).

**Key Relationship:**
$$f(y_1, y_2) = \frac{\partial^2 F(y_1, y_2)}{\partial y_1 \, \partial y_2}$$

where this derivative exists.
:::

## 🧮 Theorem: Properties of Joint PDFs {.larger}

::: {.callout-important}
## Theorem 5.3: Properties of Joint Probability Density Functions

For any joint pdf $f(y_1, y_2)$:

::: {.columns}
::: {.column weight='50%'}
**Property 1 (Non-negativity):**
$$f(y_1, y_2) \geq 0 \quad \text{for all } (y_1, y_2)$$

:::

::: {.column weight='50%'}

**Property 2 (Normalization):**
$$\int_{-\infty}^{\infty} \int_{-\infty}^{\infty} f(y_1, y_2) \, dy_1 \, dy_2 = 1$$

:::
:::

**Probability Calculation:**
$$P[(Y_1, Y_2) \in A] = \iint_A f(y_1, y_2) \, dy_1 \, dy_2$$
:::

<div style="margin-top: -20px !important;">

**Geometric Interpretation:** The **volume** under the surface $z = f(y_1, y_2)$ over region $A$ gives the probability.

</div>

## 📌 Example 3: Uniform Distribution on Unit Square {.smaller}

**Problem:** A radioactive particle is equally likely to land anywhere in a unit square. Let $(Y_1, Y_2)$ denote the landing coordinates. Find the joint pdf and $P(Y_1 + Y_2 \leq 1)$.

**Solution:**

::: {.columns}
::: {.column weight='50%' .fragment}

Since the particle is equally likely to land anywhere:

$$f(y_1, y_2) = \begin{cases} 1 & \text{if } 0 \leq y_1 \leq 1, \, 0 \leq y_2 \leq 1 \\ 0 & \text{otherwise} \end{cases}$$


**Verification:** $\int_0^1 \int_0^1 1 \, dy_1 \, dy_2 = 1 \checkmark$

:::

::: {.column weight='50%' .fragment}

**Finding $P(Y_1 + Y_2 \leq 1)$:**

$$P(Y_1 + Y_2 \leq 1) = \int_0^1 \int_0^{1-y_1} 1 \, dy_2 \, dy_1 = \int_0^1 (1 - y_1) \, dy_1$$
$$= \left[ y_1 - \frac{y_1^2}{2} \right]_0^1 = \frac{1}{2}$$

:::
:::

. . .

**Geometric:** This is the area of the triangle below the line $y_1 + y_2 = 1$ within the unit square.

## 📌 Example 4: Gasoline Tank Inventory {.smaller}

**Problem:** A firm stocks gasoline for sale. Let $Y_1$ = amount stocked (in 1000s of gallons) at the start of the week, and $Y_2$ = amount sold during the week. The joint pdf is:

$$f(y_1, y_2) = \begin{cases} 3y_1 & \text{if } 0 \leq y_2 \leq y_1 \leq 1 \\ 0 & \text{otherwise} \end{cases}$$

Find the probability that less than half of the inventory is sold.

**Solution:**

::: {.columns}
::: {.column weight='50%' .fragment}
We need $P(Y_2 < Y_1/2)$.

$$P\left(Y_2 < \frac{Y_1}{2}\right) = \int_0^1 \int_0^{y_1/2} 3y_1 \, dy_2 \, dy_1$$
:::

::: {.column weight='50%' .fragment}
**Calculation:**

$$= \int_0^1 3y_1 \cdot \frac{y_1}{2} \, dy_1 = \frac{3}{2} \int_0^1 y_1^2 \, dy_1$$
$$= \frac{3}{2} \cdot \frac{1}{3} = \boxed{\frac{1}{2}}$$
:::
:::

. . .

::: {style="margin-top: -10px;"}

**Business Insight:** There's a 50% chance of selling less than half the inventory—important for inventory planning!

:::


## 🎮 Interactive: Bivariate Probability Surface {.smaller}

::: {style="font-size: 0.65em;"}

**Explore Bivariate Densities:** Adjust parameters to see how the joint density surface changes.

::: {.columns}

::: {.column width="28%"}

```{ojs}
//| echo: false

viewof alpha = Inputs.range([0.5, 3], {
  value: 1, 
  step: 0.25, 
  label: "α (power of y₁):"
})

viewof beta = Inputs.range([0.5, 3], {
  value: 1, 
  step: 0.25, 
  label: "β (power of y₂):"
})

// Calculate normalizing constant for f(y1,y2) = c * y1^alpha * y2^beta on [0,1]x[0,1]
c_constant = (alpha + 1) * (beta + 1)

html`<div style="font-size: 0.9em;">
<p><strong>Density Function:</strong></p>
<p><em>f</em>(y<sub>1</sub>, y<sub>2</sub>) = ${c_constant.toFixed(2)} · y<sub>1</sub><sup>${alpha.toFixed(2)}</sup> · y<sub>2</sub><sup>${beta.toFixed(2)}</sup></p>
<p><strong>on</strong> [0,1] × [0,1]</p>
<p><strong>Normalizing constant:</strong><br/>
<em>c</em> = (α+1)(β+1) = ${c_constant.toFixed(2)}</p>
</div>`
```

**Observations:**  

- Higher α → density near y₁ = 1

- Higher β → density near y₂ = 1

- Surface integrates to 1

:::

::: {.column width="72%"}

```{ojs}
//| echo: false

// Generate grid for 3D-like visualization using contours
grid_size = 30
y1_vals = d3.range(0.02, 1, 1/grid_size)
y2_vals = d3.range(0.02, 1, 1/grid_size)

// Generate density values
density_data = {
  let data = [];
  for (let y1 of y1_vals) {
    for (let y2 of y2_vals) {
      let density = c_constant * Math.pow(y1, alpha) * Math.pow(y2, beta);
      data.push({y1: y1, y2: y2, density: density});
    }
  }
  return data;
}

Plot.plot({
  width: 800,
  height: 480,
  marginLeft: 50,
  marginBottom: 45,
  color: {
    type: "linear",
    scheme: "viridis",
    legend: true,
    label: "Density"
  },
  x: {
    label: "y₁",
    domain: [0, 1]
  },
  y: {
    label: "y₂",
    domain: [0, 1]
  },
  marks: [
    Plot.contour(density_data, {
      x: "y1",
      y: "y2",
      fill: "density",
      blur: 5,
      thresholds: 20
    }),
    Plot.frame()
  ]
})
```

:::

:::

:::

## 📖 Definition: Marginal Probability Distributions {.smaller}

::: {.callout-note}
## 📝 Definition 5.4: Marginal Distributions

Given a joint distribution of $(Y_1, Y_2)$, the **marginal distributions** describe the probability distribution of each variable individually.

**Discrete Case:**
$$p_1(y_1) = \sum_{y_2} p(y_1, y_2) \quad \text{and} \quad p_2(y_2) = \sum_{y_1} p(y_1, y_2)$$

**Continuous Case:**
$$f_1(y_1) = \int_{-\infty}^{\infty} f(y_1, y_2) \, dy_2 \quad \text{and} \quad f_2(y_2) = \int_{-\infty}^{\infty} f(y_1, y_2) \, dy_1$$
:::

**Intuition:** To find the marginal distribution of $Y_1$, we "accumulate" (sum or integrate) the joint probabilities over all possible values of $Y_2$.

## 📌 Example 5: Finding Marginal Distributions (Discrete) {.smaller}

**Problem:** Using the committee selection example, find the marginal distributions of $Y_1$ (number of Republicans) and $Y_2$ (number of Democrats).

**Solution:**

. . .

Recall the joint probability table with constraint $y_1 + y_2 = 3$:

::: {.columns}
::: {.column weight='50%' .fragment}
**Marginal of $Y_1$ (sum across rows):**

$p_1(0) = \frac{1}{35}$ (all Democrats)

$p_1(1) = \frac{12}{35}$

$p_1(2) = \frac{18}{35}$

$p_1(3) = \frac{4}{35}$ (all Republicans)
:::

::: {.column weight='50%' .fragment}
**Marginal of $Y_2$ (sum down columns):**

$p_2(0) = \frac{4}{35}$ (no Democrats)

$p_2(1) = \frac{18}{35}$

$p_2(2) = \frac{12}{35}$

$p_2(3) = \frac{1}{35}$ (all Democrats)
:::
:::

. . .

**Verification:** Both marginals sum to 1: $\frac{1+12+18+4}{35} = \frac{35}{35} = 1 \checkmark$

## 📌 Example 6: Finding Marginal Distributions (Continuous) {.smaller}

**Problem:** For the gasoline inventory problem with $f(y_1, y_2) = 3y_1$ for $0 \leq y_2 \leq y_1 \leq 1$, find the marginal distributions.

**Solution:**

::: {.columns}
::: {.column weight='50%' .fragment}
**Marginal of $Y_1$ (amount stocked):**

$$f_1(y_1) = \int_0^{y_1} 3y_1 \, dy_2 = 3y_1 \cdot y_1 = 3y_1^2$$

for $0 \leq y_1 \leq 1$

**Verification:** $\int_0^1 3y_1^2 \, dy_1 = [y_1^3]_0^1 = 1 \checkmark$
:::

::: {.column weight='50%' .fragment}
**Marginal of $Y_2$ (amount sold):**

$$f_2(y_2) = \int_{y_2}^{1} 3y_1 \, dy_1 = \frac{3}{2}[y_1^2]_{y_2}^{1}$$
$$= \frac{3}{2}(1 - y_2^2)$$

for $0 \leq y_2 \leq 1$
:::
:::

. . .

**Business Insight:** The marginal distribution of amount sold helps the firm understand sales patterns independently of stocking decisions.

## 🤝 Think-Pair-Share: Investment Decision {.smaller .r-fit-text}

::: {style="font-size: 26px"}
::: {.callout-note}
## 💭 Student Engagement Activity (4 minutes)

**Scenario:** An investor is considering two assets with the following joint probability distribution for annual returns (in %):

| $Y_1 \backslash Y_2$ | -5% | 0% | 10% | 20% |
|:---:|:---:|:---:|:---:|:---:|
| **-10%** | 0.05 | 0.05 | 0.02 | 0.00 |
| **5%** | 0.10 | 0.15 | 0.15 | 0.05 |
| **15%** | 0.03 | 0.10 | 0.20 | 0.10 |

**Think (1 minute):** Work individually

- Calculate the marginal distribution of Asset 1's returns ($Y_1$)
- Calculate the marginal distribution of Asset 2's returns ($Y_2$)

**Pair (2 minutes):** Discuss with a partner

- Which asset has higher expected return based on the marginals?
- Which asset appears riskier (more spread in the marginal distribution)?

**Share (1 minute):** Class discussion

- How does knowing the joint distribution help us beyond just the marginals?
:::
:::

## 💰 Case Study: Portfolio Returns Analysis (Real Data) {.larger}

::: {style="font-size:40px"}
::: {.columns}
::: {.column width="50%"}
::: {.callout-note}
## 📈 Financial Market Application

**Context**: In portfolio theory, we study the joint distribution of multiple asset returns to understand diversification benefits and overall portfolio risk.

**Key Questions**:

- What does the joint distribution of two stock returns look like?

- How do we extract marginal distributions from observed data?

- What patterns emerge in the scatter of returns?

:::
:::

::: {.column width="50%"}
::: {.callout-tip}
## 📊 Data Source

We analyze **daily returns** for **Apple (AAPL)** and **Microsoft (MSFT)** stocks.

**Source**: Yahoo Finance API via `quantmod` package

**Period**: Last 252 trading days (1 year)

**Data Type**: Adjusted closing prices, converted to daily log returns

**Verification**: Cross-checked with financial data providers
:::
:::
:::
:::

## 💰 Case Study: Data Collection {.smaller}

::: {style="font-size:20px"}
::: {.columns}
::: {.column width="33%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Load required libraries
library(quantmod)
library(tidyverse)
library(knitr)

# Set date range (last 252 trading days)
end_date <- Sys.Date()
start_date <- end_date - 365

# Download stock data
getSymbols("AAPL", from = start_date, 
           to = end_date, auto.assign = TRUE)
getSymbols("MSFT", from = start_date, 
           to = end_date, auto.assign = TRUE)
```

:::

::: {.column width="33%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Calculate daily log returns
aapl_returns <- dailyReturn(AAPL, type = "log")
msft_returns <- dailyReturn(MSFT, type = "log")

# Combine into data frame
returns_df <- data.frame(
  Date = index(aapl_returns),
  AAPL = as.numeric(aapl_returns),
  MSFT = as.numeric(msft_returns)
) %>% na.omit()

cat(sprintf("Observations: %d\n", nrow(returns_df)))
cat(sprintf("Range: %s to %s\n", 
            min(returns_df$Date), max(returns_df$Date)))
```

:::

::: {.column width="33%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Summary statistics
cat("=== AAPL Returns ===\n")
cat(sprintf("Mean: %.4f%%\n", mean(returns_df$AAPL) * 100))
cat(sprintf("Std Dev: %.4f%%\n", sd(returns_df$AAPL) * 100))

cat("\n=== MSFT Returns ===\n")
cat(sprintf("Mean: %.4f%%\n", mean(returns_df$MSFT) * 100))
cat(sprintf("Std Dev: %.4f%%\n", sd(returns_df$MSFT) * 100))

cat("\n=== Correlation ===\n")
cat(sprintf("Corr: %.4f\n", 
            cor(returns_df$AAPL, returns_df$MSFT)))
```

:::
:::
:::

## 💰 Case Study: Joint Distribution Visualization {.smaller}

::: {style="font-size:20px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

# Scatter plot of joint returns
ggplot(returns_df, aes(x = AAPL, y = MSFT)) +
  geom_point(alpha = 0.6, color = "steelblue", size = 2) +
  geom_smooth(method = "lm", color = "red", 
              se = TRUE, alpha = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", 
             color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", 
             color = "gray50") +
  labs(title = "Joint Distribution: AAPL vs MSFT Returns",
       subtitle = "Daily log returns with linear fit",
       x = "AAPL Daily Return",
       y = "MSFT Daily Return") +
  theme_minimal(base_size = 11) +
  coord_fixed(ratio = 1)
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

# 2D density plot (empirical joint distribution)
ggplot(returns_df, aes(x = AAPL, y = MSFT)) +
  stat_density_2d(aes(fill = after_stat(density)), 
                  geom = "raster", contour = FALSE) +
  scale_fill_viridis_c(option = "plasma") +
  geom_point(alpha = 0.3, color = "white", size = 0.5) +
  labs(title = "Empirical Joint Density",
       subtitle = "2D kernel density estimate",
       x = "AAPL Daily Return",
       y = "MSFT Daily Return",
       fill = "Density") +
  theme_minimal(base_size = 11) +
  coord_fixed(ratio = 1)
```

:::
:::
:::

## 💰 Case Study: Marginal Distributions {.smaller}

::: {style="font-size:40px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

# Marginal distribution of AAPL
ggplot(returns_df, aes(x = AAPL)) +
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 40, fill = "steelblue", 
                 alpha = 0.7, color = "black") +
  stat_function(fun = dnorm, 
                args = list(mean = mean(returns_df$AAPL),
                           sd = sd(returns_df$AAPL)),
                color = "red", linewidth = 1.2) +
  geom_vline(xintercept = mean(returns_df$AAPL),
             color = "darkgreen", linetype = "dashed",
             linewidth = 1) +
  labs(title = "Marginal Distribution: AAPL Returns",
       subtitle = "Histogram with normal overlay",
       x = "Daily Return",
       y = "Density") +
  theme_minimal(base_size = 11)
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

# Marginal distribution of MSFT
ggplot(returns_df, aes(x = MSFT)) +
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 40, fill = "coral", 
                 alpha = 0.7, color = "black") +
  stat_function(fun = dnorm, 
                args = list(mean = mean(returns_df$MSFT),
                           sd = sd(returns_df$MSFT)),
                color = "red", linewidth = 1.2) +
  geom_vline(xintercept = mean(returns_df$MSFT),
             color = "darkgreen", linetype = "dashed",
             linewidth = 1) +
  labs(title = "Marginal Distribution: MSFT Returns",
       subtitle = "Histogram with normal overlay",
       x = "Daily Return",
       y = "Density") +
  theme_minimal(base_size = 11)
```

:::
:::
:::

## 💰 Case Study: Key Findings {.larger}

::: {style="font-size: 40px;"}
::: {.callout-important}
## 📊 Analysis Results

::: {.columns}

::: {.column width="33%"}
::: {.fragment}
**Joint Distribution Characteristics:**

- Strong positive correlation between AAPL and MSFT returns

- Both tend to move together (same sector, similar market factors)

- Elliptical scatter pattern suggests approximate bivariate normality
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Marginal Distributions:**

- Both stocks show approximately symmetric return distributions

- Similar volatility levels (tech sector characteristics)

- Marginals alone don't capture the dependence structure
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Portfolio Implications:**

1. **Limited Diversification**: High correlation means holding both doesn't reduce risk much

2. **Sector Risk**: Both exposed to similar tech sector factors

3. **Better Diversification**: Would need assets from different sectors with lower correlation
:::
:::

:::
:::
:::

## 📝 Quiz #1: Joint Probability Functions {.smaller .quiz-question}

For a valid joint probability function $p(y_1, y_2)$, which of the following must be true?

- [$p(y_1, y_2) \geq 0$ for all $(y_1, y_2)$ and $\sum_{y_1}\sum_{y_2} p(y_1, y_2) = 1$]{.correct data-explanation="✅ Correct! Joint probability functions must be non-negative everywhere and sum to 1 over all possible pairs."}
- $p(y_1, y_2) \geq 0$ for all $(y_1, y_2)$ and $\sum_{y_1}\sum_{y_2} p(y_1, y_2) \leq 1$
- $0 \leq p(y_1, y_2) \leq 1$ for all $(y_1, y_2)$ and no constraint on sum
- $p(y_1, y_2) > 0$ for all $(y_1, y_2)$ and $\sum_{y_1}\sum_{y_2} p(y_1, y_2) = 1$

## 📝 Quiz #2: Computing Marginal Distributions {.smaller .quiz-question}

To find the marginal distribution $f_1(y_1)$ from a continuous joint pdf $f(y_1, y_2)$, you should:

- [Integrate $f(y_1, y_2)$ with respect to $y_2$ over all possible values of $y_2$]{.correct data-explanation="✅ Correct! The marginal density f₁(y₁) = ∫f(y₁, y₂)dy₂ is found by integrating out the other variable."}
- Differentiate $f(y_1, y_2)$ with respect to $y_2$
- Set $y_2 = 0$ in $f(y_1, y_2)$
- Multiply $f(y_1, y_2)$ by $y_2$ and integrate

## 📝 Quiz #3: Bivariate Density Interpretation {.smaller .quiz-question}

For a continuous bivariate random variable $(Y_1, Y_2)$ with joint pdf $f(y_1, y_2)$, the probability $P(Y_1 \leq a, Y_2 \leq b)$ equals:

- [The volume under the surface $f(y_1, y_2)$ over the region $(-\infty, a] \times (-\infty, b]$]{.correct data-explanation="✅ Correct! For continuous bivariate distributions, probabilities are calculated as volumes under the density surface over the specified region."}
- The value $f(a, b)$
- The area under the curve $f(y_1, y_2) = a + b$
- The double derivative of $f(y_1, y_2)$ at $(a, b)$

## 📝 Summary {.smaller}

::: {style="font-size: 28px"}
::: {.summary-box}
**✅ Key Takeaways**

- **Joint probability functions** (discrete) and **joint pdfs** (continuous) describe the simultaneous behavior of two or more random variables, satisfying non-negativity and normalization properties

- **Joint CDFs** $F(y_1, y_2) = P(Y_1 \leq y_1, Y_2 \leq y_2)$ accumulate probability in the "southwest rectangle"

- **Marginal distributions** are obtained by summing (discrete) or integrating (continuous) the joint distribution over all values of the other variable

- The joint distribution contains **more information** than the marginals alone—it captures the dependence structure between variables

- **Financial applications** include modeling portfolio returns, where understanding joint behavior is crucial for risk management and diversification
:::
:::

## 📚 Practice Problems

::: {.callout-tip}
## 📝 Homework Problems

**Problem 1 (Discrete Joint Distribution):** A portfolio contains stocks A and B. Let $Y_1$ = return on stock A and $Y_2$ = return on stock B, with joint pmf given by a $3 \times 3$ table. Verify it's a valid pmf and find both marginal distributions.

**Problem 2 (Continuous Joint Distribution):** Let $f(y_1, y_2) = cy_1y_2$ for $0 < y_1 < 2$ and $0 < y_2 < 1$. Find the constant $c$ and compute $P(Y_1 > 1, Y_2 < 0.5)$.

**Problem 3 (Marginal from Joint):** Given $f(y_1, y_2) = 2$ for $0 < y_1 < 1$, $0 < y_2 < y_1$, find $f_1(y_1)$ and $f_2(y_2)$.

**Problem 4 (Financial Application):** For two assets with given joint distribution, compute the probability that both assets have positive returns simultaneously.
:::

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business

ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Conditional Distributions and Independence

**Reading:** Chapter 5, Sections 5.3-5.4

**Preparation:** Review conditional probability from Chapter 2

**⏰ Reminders:**

✅ Complete Practice Problems 1-4

✅ Review integration techniques (double integrals)

✅ Think about when two random variables might be independent

✅ Work hard!
:::
:::

## ❓ Questions? {.center}

::: {.callout-note}
## 💬 Open Discussion

**Key Topics for Discussion:**

- How do joint distributions help us understand portfolio risk beyond individual asset risks?

- Why is the marginal distribution not sufficient for understanding the relationship between two variables?

- What are some real-world examples where knowing the joint distribution is crucial for decision-making?

- How would you estimate a joint distribution from observed data in practice?
:::

