---
title: "Mathematical Statistics"
subtitle: "Chapter 5: Multivariate Distributions – Conditional Distributions & Independence"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics - Ch.5: Multivariate Distributions (Part 2)"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::: {.learning-objectives}
By the end of this lecture, you will be able to:

- **Derive** conditional probability functions for discrete random variables

- **Calculate** conditional density functions for continuous random variables

- **Apply** the definition of independence to verify whether random variables are independent

- **Use** the factorization theorem to quickly check independence

- **Interpret** conditional distributions in financial risk scenarios
:::

## 📋 Overview {.smaller}

::: {style="font-size:38px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Conditional Distributions (Discrete)** – Probability given information about another variable

- **Conditional Distributions (Continuous)** – Conditional density functions

- **Independent Random Variables** – Definition and properties

- **Factorization Theorem** – Quick test for independence

- **Case Study** – Conditional VaR (Value-at-Risk) in portfolio management
:::
:::
:::

## 📖 Why Conditional Distributions? {.smaller .r-fit-text}

::: {.callout-note}
## 🎯 Motivation

Understanding how one variable behaves **given information about another** is crucial for prediction, risk management, and decision-making.

::: {.columns}
::: {.column width="50%"}
::: {.fragment}
**Finance Applications:**

- Credit risk given macroeconomic conditions
- Stock returns given market movements
- Default probability given credit rating
- Portfolio VaR given stress scenarios
:::
:::

::: {.column width="50%"}
::: {.fragment}
**Azerbaijan Context:**

- Oil revenue given OPEC decisions
- Manat volatility given oil prices
- Bank profitability given interest rates
- Insurance claims given weather events
:::
:::
:::
:::

**Key Question:** How do we describe the distribution of $Y_1$ when we **know** the value of $Y_2$?

## 📖 Definition: Conditional Probability Function (Discrete) {.smaller}

::: {.callout-note}
## 📝 Definition 5.5: Conditional Probability Function

If $Y_1$ and $Y_2$ are jointly discrete random variables with joint probability function $p(y_1, y_2)$ and marginal probability functions $p_1(y_1)$ and $p_2(y_2)$, then the **conditional discrete probability function** of $Y_1$ given $Y_2$ is:

$$p(y_1|y_2) = P(Y_1 = y_1 | Y_2 = y_2) = \frac{P(Y_1 = y_1, Y_2 = y_2)}{P(Y_2 = y_2)} = \frac{p(y_1, y_2)}{p_2(y_2)}$$

provided that $p_2(y_2) > 0$.
:::

**Key insight:** This is just the conditional probability formula $P(A|B) = P(A \cap B)/P(B)$ applied to random variables!

## 📌 Example 1: Committee Selection {.smaller}

**Problem:** From 3 Republicans, 2 Democrats, 1 Independent, select a committee of 2. Let $Y_1$ = number of Republicans, $Y_2$ = number of Democrats. Find $P(Y_1 = 1 | Y_2 = 1)$.

**Solution:**

. . .

From Lecture 1, we have the joint distribution:

| $y_1 \backslash y_2$ | 0 | 1 | 2 | **Total** |
|:---:|:---:|:---:|:---:|:---:|
| 0 | 0 | 2/15 | 1/15 | 3/15 |
| 1 | 3/15 | 6/15 | 0 | 9/15 |
| 2 | 3/15 | 0 | 0 | 3/15 |
| **Total** | 6/15 | **8/15** | 1/15 | 1 |

. . .

$$P(Y_1 = 1 | Y_2 = 1) = \frac{p(1,1)}{p_2(1)} = \frac{6/15}{8/15} = \frac{6}{8} = \frac{3}{4}$$

## 📌 Example 1: Complete Conditional Distribution {.smaller}

**Problem:** Find the complete conditional distribution of $Y_1$ given $Y_2 = 1$.

**Solution:**

::: {.columns}
::: {.column width="50%"}
$$P(Y_1 = 0 | Y_2 = 1) = \frac{p(0,1)}{p_2(1)} = \frac{2/15}{8/15} = \frac{1}{4}$$

$$P(Y_1 = 1 | Y_2 = 1) = \frac{p(1,1)}{p_2(1)} = \frac{6/15}{8/15} = \frac{3}{4}$$

$$P(Y_1 = 2 | Y_2 = 1) = \frac{p(2,1)}{p_2(1)} = \frac{0}{8/15} = 0$$
:::

::: {.column width="50%"}
::: {.fragment}
**Interpretation:**

Given that exactly one Democrat is on the committee:

- 25% chance of zero Republicans
- 75% chance of one Republican  
- 0% chance of two Republicans

**Verification:** $\frac{1}{4} + \frac{3}{4} + 0 = 1$ ✓
:::
:::
:::

## 📖 Definition: Conditional Density Function (Continuous) {.smaller}

::: {.callout-note}
## 📝 Definition 5.7: Conditional Density Function

Let $Y_1$ and $Y_2$ be jointly continuous random variables with joint density $f(y_1, y_2)$ and marginal densities $f_1(y_1)$ and $f_2(y_2)$. 

For any $y_2$ such that $f_2(y_2) > 0$, the **conditional density** of $Y_1$ given $Y_2 = y_2$ is:

$$f(y_1|y_2) = \frac{f(y_1, y_2)}{f_2(y_2)}$$

Similarly, for any $y_1$ such that $f_1(y_1) > 0$:

$$f(y_2|y_1) = \frac{f(y_1, y_2)}{f_1(y_1)}$$
:::

**Note:** The conditional density is **undefined** when the marginal density equals zero!

## 📌 Example 2: Soft-Drink Machine {.smaller}

**Problem:** Let $Y_2$ = amount in supply (gallons), $Y_1$ = amount sold ($Y_1 \leq Y_2$). Joint density:
$$f(y_1, y_2) = \begin{cases} 1/2, & 0 \leq y_1 \leq y_2 \leq 2 \\ 0, & \text{elsewhere} \end{cases}$$

Find the conditional density of $Y_1$ given $Y_2 = y_2$.

. . .

**Solution:** First find $f_2(y_2)$:
$$f_2(y_2) = \int_0^{y_2} \frac{1}{2} \, dy_1 = \frac{y_2}{2}, \quad 0 \leq y_2 \leq 2$$

. . .

Then:
$$f(y_1|y_2) = \frac{f(y_1, y_2)}{f_2(y_2)} = \frac{1/2}{y_2/2} = \frac{1}{y_2}, \quad 0 \leq y_1 \leq y_2$$

**Interpretation:** Given supply $y_2$, sales are **uniform** on $[0, y_2]$!

## 📌 Example 2: Conditional Probability {.smaller}

**Problem:** Find $P(Y_1 \leq 0.5 | Y_2 = 1.5)$ — the probability that sales are at most 0.5 gallons given supply of 1.5 gallons.

**Solution:**

. . .

Using the conditional density $f(y_1|y_2 = 1.5) = \frac{1}{1.5}$ for $0 \leq y_1 \leq 1.5$:

$$P(Y_1 \leq 0.5 | Y_2 = 1.5) = \int_0^{0.5} \frac{1}{1.5} \, dy_1 = \frac{0.5}{1.5} = \frac{1}{3}$$

. . .

**Compare with $Y_2 = 2$:**
$$P(Y_1 \leq 0.5 | Y_2 = 2) = \int_0^{0.5} \frac{1}{2} \, dy_1 = \frac{0.5}{2} = \frac{1}{4}$$

**Insight:** The conditional probability depends on the conditioning value!

## 📖 Definition: Independent Random Variables {.smaller}

::: {.callout-note}
## 📝 Definition 5.8: Independence

Let $Y_1$ have distribution function $F_1(y_1)$, $Y_2$ have distribution function $F_2(y_2)$, and let $Y_1$ and $Y_2$ have joint distribution function $F(y_1, y_2)$. 

Then $Y_1$ and $Y_2$ are **independent** if and only if:

$$F(y_1, y_2) = F_1(y_1) \cdot F_2(y_2)$$

for every pair of real numbers $(y_1, y_2)$.

If $Y_1$ and $Y_2$ are not independent, they are said to be **dependent**.
:::

**Intuition:** Joint CDF factors into a product of marginal CDFs.

## 🧮 Theorem: Characterization of Independence {.smaller}

::: {.callout-important}
## Theorem 5.4: Independence via Probability/Density Functions

**Discrete case:** $Y_1$ and $Y_2$ are independent if and only if:
$$p(y_1, y_2) = p_1(y_1) \cdot p_2(y_2) \quad \text{for all } (y_1, y_2)$$

**Continuous case:** $Y_1$ and $Y_2$ are independent if and only if:
$$f(y_1, y_2) = f_1(y_1) \cdot f_2(y_2) \quad \text{for all } (y_1, y_2)$$
:::

**Key insight:** Independence means the joint distribution **factors** into the product of marginals for **ALL** values.

## 📌 Example 3: Testing Independence (Discrete) {.smaller}

**Problem:** For the die-tossing experiment ($Y_1$ = die 1, $Y_2$ = die 2), are $Y_1$ and $Y_2$ independent?

**Solution:**

. . .

We have $p(y_1, y_2) = 1/36$ for all $y_1, y_2 \in \{1,2,3,4,5,6\}$.

. . .

Marginal distributions: $p_1(y_1) = 1/6$ and $p_2(y_2) = 1/6$

. . .

Check: $p_1(y_1) \cdot p_2(y_2) = \frac{1}{6} \cdot \frac{1}{6} = \frac{1}{36} = p(y_1, y_2)$ ✓

. . .

**Conclusion:** $Y_1$ and $Y_2$ are **INDEPENDENT**.

This makes intuitive sense: the outcome on one die doesn't affect the other!

## 📌 Example 4: Testing Independence (Discrete) {.smaller}

**Problem:** In Example 1 (committee selection), are $Y_1$ (Republicans) and $Y_2$ (Democrats) independent?

**Solution:**

. . .

From the joint distribution table, we have $p(0, 0) = 0$.

. . .

But the marginals give: $p_1(0) = 3/15$ and $p_2(0) = 6/15$

. . .

$$p_1(0) \cdot p_2(0) = \frac{3}{15} \cdot \frac{6}{15} = \frac{18}{225} = \frac{2}{25} \neq 0 = p(0, 0)$$

. . .

**Conclusion:** $Y_1$ and $Y_2$ are **DEPENDENT**.

**Intuition:** If we select zero Republicans, we must have selected either Democrats or Independents, affecting $Y_2$.

## 📌 Example 5: Testing Independence (Continuous) {.smaller}

**Problem:** Let $f(y_1, y_2) = 6y_1y_2^2$ for $0 \leq y_1 \leq 1, 0 \leq y_2 \leq 1$. Are $Y_1$ and $Y_2$ independent?

**Solution:**

. . .

Find marginals:
$$f_1(y_1) = \int_0^1 6y_1y_2^2 \, dy_2 = 6y_1 \cdot \frac{1}{3} = 2y_1, \quad 0 \leq y_1 \leq 1$$

$$f_2(y_2) = \int_0^1 6y_1y_2^2 \, dy_1 = 6y_2^2 \cdot \frac{1}{2} = 3y_2^2, \quad 0 \leq y_2 \leq 1$$

. . .

Check: $f_1(y_1) \cdot f_2(y_2) = 2y_1 \cdot 3y_2^2 = 6y_1y_2^2 = f(y_1, y_2)$ ✓

**Conclusion:** $Y_1$ and $Y_2$ are **INDEPENDENT**.

## 📌 Example 6: Dependent Variables {.smaller}

**Problem:** Let $f(y_1, y_2) = 2$ for $0 \leq y_2 \leq y_1 \leq 1$. Are $Y_1$ and $Y_2$ independent?

**Solution:**

. . .

Find marginals:
$$f_1(y_1) = \int_0^{y_1} 2 \, dy_2 = 2y_1, \quad 0 \leq y_1 \leq 1$$

$$f_2(y_2) = \int_{y_2}^1 2 \, dy_1 = 2(1-y_2), \quad 0 \leq y_2 \leq 1$$

. . .

Check: $f_1(y_1) \cdot f_2(y_2) = 2y_1 \cdot 2(1-y_2) = 4y_1(1-y_2) \neq 2 = f(y_1, y_2)$

. . .

**Conclusion:** $Y_1$ and $Y_2$ are **DEPENDENT**.

**Key observation:** The region $0 \leq y_2 \leq y_1 \leq 1$ (triangular) immediately suggests dependence!

## 🧮 Theorem: Factorization Criterion {.smaller}

::: {.callout-important}
## Theorem 5.5: Quick Test for Independence

Let $Y_1$ and $Y_2$ have a joint density $f(y_1, y_2)$ that is positive if and only if $a \leq y_1 \leq b$ and $c \leq y_2 \leq d$, for constants $a, b, c, d$; and $f(y_1, y_2) = 0$ otherwise.

Then $Y_1$ and $Y_2$ are **independent** if and only if:
$$f(y_1, y_2) = g(y_1) \cdot h(y_2)$$

where $g(y_1)$ is a nonnegative function of $y_1$ alone and $h(y_2)$ is a nonnegative function of $y_2$ alone.
:::

**Important:** The region of support must be **rectangular** (constants $a, b, c, d$). If the region is not rectangular, variables are typically dependent!

## 📌 Example 7: Using Factorization Theorem {.smaller}

**Problem:** Does $f(y_1, y_2) = 2y_1$ for $0 \leq y_1 \leq 1, 0 \leq y_2 \leq 1$ give independent variables?

**Solution:**

. . .

**Step 1:** Check the region — $[0,1] \times [0,1]$ is rectangular ✓

. . .

**Step 2:** Can we factor?

$$f(y_1, y_2) = 2y_1 = \underbrace{y_1}_{g(y_1)} \cdot \underbrace{2}_{h(y_2)}$$

. . .

Yes! The joint density factors into a function of $y_1$ only times a function of $y_2$ only.

**Conclusion:** $Y_1$ and $Y_2$ are **INDEPENDENT**.

Note: $g(y_1) = y_1$ and $h(y_2) = 2$ are not themselves density functions, but that's okay!

## 🎮 Interactive: Independence Test {.smaller}

::: {style="font-size: 0.8em;"}

**Explore: When does $f(y_1, y_2) = c \cdot y_1^a \cdot y_2^b$ give independent variables?**

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof power_a = Inputs.range([0, 2], {
  value: 1, 
  step: 0.5, 
  label: "Power a:"
})

viewof power_b = Inputs.range([0, 2], {
  value: 1, 
  step: 0.5, 
  label: "Power b:"
})

independence_check = {
  // f(y1,y2) = c * y1^a * y2^b factors automatically!
  return "✅ INDEPENDENT (factors into g(y₁)×h(y₂))";
}
```

**Status:** ${independence_check}

**Formula:** $f(y_1, y_2) = c \cdot y_1^{${power_a}} \cdot y_2^{${power_b}}$

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// Generate contour data for joint density
jointData = {
  const n = 40;
  const data = [];
  for (let i = 0; i < n; i++) {
    for (let j = 0; j < n; j++) {
      const y1 = i / (n - 1);
      const y2 = j / (n - 1);
      const density = Math.pow(y1, power_a) * Math.pow(y2, power_b);
      data.push({y1, y2, density});
    }
  }
  return data;
}

Plot.plot({
  width: 450,
  height: 400,
  color: {scheme: "YlOrRd", legend: true, label: "Density"},
  x: { domain: [0, 1], label: "Y₁" },
  y: { domain: [0, 1], label: "Y₂" },
  marks: [
    Plot.contour(jointData, {
      x: "y1", y: "y2", fill: "density",
      blur: 3, thresholds: 20
    }),
    Plot.text([{x: 0.5, y: 1.1, text: "Joint Density Contours (rectangular region → can factor)"}], {
      x: "x", y: "y", text: "text", fontSize: 12
    })
  ]
})
```

:::
:::
:::

## 💰 Case Study: Conditional Analysis of Bank Returns {.smaller}

::: {style="font-size:22px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

library(tidyverse)
library(tidyquant)

# Get bank stock data
symbols <- c("JPM", "BAC")
prices <- tq_get(symbols, from = "2022-01-01")

returns <- prices %>%
  group_by(symbol) %>%
  tq_transmute(select = adjusted,
               mutate_fun = periodReturn,
               period = "daily") %>%
  pivot_wider(names_from = symbol, 
              values_from = daily.returns) %>%
  na.omit()

# Conditional analysis: JPM returns given BAC performance
returns <- returns %>%
  mutate(
    BAC_state = case_when(
      BAC < quantile(BAC, 0.25) ~ "BAC Down (Q1)",
      BAC > quantile(BAC, 0.75) ~ "BAC Up (Q4)",
      TRUE ~ "BAC Normal"
    )
  )

# Conditional summary statistics
cond_stats <- returns %>%
  group_by(BAC_state) %>%
  summarise(
    `Mean JPM` = mean(JPM) * 100,
    `SD JPM` = sd(JPM) * 100,
    `n` = n()
  )

knitr::kable(cond_stats, digits = 3,
             caption = "JPM Returns (%) Conditional on BAC")
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

# Visualize conditional distributions
ggplot(returns, aes(x = JPM, fill = BAC_state)) +
  geom_density(alpha = 0.5) +
  labs(title = "Conditional Density of JPM Returns",
       subtitle = "f(JPM | BAC state)",
       x = "JPM Daily Return",
       y = "Density",
       fill = "Condition") +
  theme_minimal(base_size = 12) +
  scale_fill_manual(values = c("red", "gray", "green"))
```

:::
:::
:::

## 💰 Case Study: Testing Independence {.smaller}

::: {style="font-size:22px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Statistical test for independence
# Chi-square test on discretized returns
returns_discrete <- returns %>%
  mutate(
    JPM_cat = cut(JPM, breaks = 5),
    BAC_cat = cut(BAC, breaks = 5)
  )

# Contingency table
cont_table <- table(returns_discrete$JPM_cat, 
                    returns_discrete$BAC_cat)

# Chi-square test
chi_test <- chisq.test(cont_table)

cat("Chi-Square Test for Independence:\n")
cat("X² =", round(chi_test$statistic, 2), "\n")
cat("df =", chi_test$parameter, "\n")
cat("p-value <", format.pval(chi_test$p.value), "\n")
cat("\nConclusion:", 
    ifelse(chi_test$p.value < 0.05, 
           "REJECT independence (dependent)", 
           "Cannot reject independence"))
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

# Scatter plot with regression
ggplot(returns, aes(x = BAC, y = JPM)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "JPM vs BAC Returns",
       subtitle = "Strong dependence visible",
       x = "BAC Daily Return",
       y = "JPM Daily Return") +
  theme_minimal(base_size = 12) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed")
```

:::
:::
:::

## 💰 Case Study: Key Findings {.smaller}

::: {style="font-size: 26px;"}
::: {.callout-important}
## 📊 Analysis Results

::: {.columns}

::: {.column width="33%"}
::: {.fragment}
**Conditional Behavior:**

- When BAC falls, JPM tends to fall

- Conditional mean shifts significantly

- Conditional volatility also changes
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Independence Test:**

- Chi-square strongly rejects independence

- Clear positive dependence

- Returns move together
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Portfolio Implications:**

1. **Diversification limited**: Same-sector stocks dependent

2. **Risk modeling**: Must use joint distributions

3. **Stress testing**: Conditional analysis essential
:::
:::

:::
:::
:::

## 📝 Quiz #1: Conditional Probability {.smaller .quiz-question}

If $p(1,2) = 0.15$ and $p_2(2) = 0.30$, what is $P(Y_1 = 1 | Y_2 = 2)$?

- [0.50]{.correct data-explanation="✅ Correct! By Definition 5.5: $P(Y_1 = 1 | Y_2 = 2) = p(1,2)/p_2(2) = 0.15/0.30 = 0.50$."}
- 0.15
- 0.30
- 0.45

## 📝 Quiz #2: Independence Condition {.smaller .quiz-question}

For continuous random variables, which condition is equivalent to independence?

- [$f(y_1, y_2) = f_1(y_1) \cdot f_2(y_2)$ for all $(y_1, y_2)$]{.correct data-explanation="✅ Correct! By Theorem 5.4, independence requires the joint density to factor into the product of marginals for ALL values."}
- $f(y_1|y_2) = f_2(y_2)$ for all $y_2$
- $f(y_1, y_2) = f_1(y_1) + f_2(y_2)$ for all $(y_1, y_2)$
- $f(y_1, y_2) > 0$ for all $(y_1, y_2)$

## 📝 Quiz #3: Factorization Theorem {.smaller .quiz-question}

$f(y_1, y_2) = 12y_1^2y_2$ for $0 \leq y_1 \leq 1, 0 \leq y_2 \leq 1$. Are $Y_1$ and $Y_2$ independent?

- [Yes, because $f(y_1,y_2) = g(y_1) \cdot h(y_2)$ with rectangular support]{.correct data-explanation="✅ Correct! The region is rectangular and $f(y_1,y_2) = 12y_1^2y_2 = (y_1^2)(12y_2)$ factors into a function of $y_1$ times a function of $y_2$."}
- No, because the powers are different
- No, because the constant 12 cannot be split
- Cannot determine without computing marginals

## 📝 Quiz #4: Dependent Variables {.smaller .quiz-question}

$f(y_1, y_2) = 8y_1y_2$ for $0 < y_1 < y_2 < 1$. Why are $Y_1$ and $Y_2$ dependent?

- [The region of support is not rectangular]{.correct data-explanation="✅ Correct! The constraint $y_1 < y_2$ creates a triangular (non-rectangular) region. When the support is not rectangular, Theorem 5.5 cannot be applied and variables are typically dependent."}
- The density cannot be factored
- The coefficient 8 is too large
- The marginal densities don't exist

## 📝 Summary {.smaller}

::: {style="font-size: 28px"}
::: {.summary-box}
**✅ Key Takeaways**

- **Conditional distributions** describe behavior of one variable given information about another

- **Discrete case:** $p(y_1|y_2) = p(y_1, y_2)/p_2(y_2)$ when $p_2(y_2) > 0$

- **Continuous case:** $f(y_1|y_2) = f(y_1, y_2)/f_2(y_2)$ when $f_2(y_2) > 0$

- **Independence:** Joint = Product of marginals for ALL values; independent $\Rightarrow$ $f(y_1|y_2) = f_1(y_1)$

- **Factorization theorem:** Rectangular support + density factors $\Rightarrow$ independence
:::
:::

## 📚 Practice Problems

::: {.callout-tip}
## 📝 Homework Problems

**Problem 1 (Conditional Discrete):** Using the committee selection example, find the complete conditional distribution of $Y_2$ given $Y_1 = 1$.

**Problem 2 (Conditional Continuous):** If $f(y_1, y_2) = 4y_1y_2$ for $0 < y_1 < 1, 0 < y_2 < 1$, find $f(y_1|y_2)$ and verify it's a valid density.

**Problem 3 (Independence):** Determine if $f(y_1, y_2) = 2$ for $0 < y_1 + y_2 < 1$ gives independent variables.

**Problem 4 (Financial Application):** If stock $A$ returns have conditional distribution $f(r_A|r_B) = \frac{1}{\sigma\sqrt{2\pi}}\exp\left(-\frac{(r_A - \rho r_B)^2}{2\sigma^2}\right)$ given market return $r_B$, are $R_A$ and $R_B$ independent?
:::

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business

ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Expected Values, Covariance, and Correlation

**Reading:** Chapter 5, Sections 5.5-5.8

**Preparation:** Review variance and expected value formulas

**⏰ Reminders:**

✅ Complete Practice Problems 1-4

✅ Review integration by parts

✅ Think about how correlation relates to dependence

✅ Work hard!
:::
:::

## ❓ Questions? {.center}

::: {.callout-note}
## 💬 Open Discussion

**Key Topics for Discussion:**

- How does conditional expectation relate to regression?

- Can two variables be dependent but have zero correlation?

- How do we use independence in portfolio theory?

- What's the relationship between statistical independence and economic independence?
:::
