---
title: "Mathematical Statistics"
subtitle: "Bivariate and Multivariate Probability Distributions"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - Bivariate and Multivariate Distributions"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
library(ggplot2)
theme_set(theme_minimal(base_size = 22))
```

## 🎯 Learning Objectives

::: {style="font-size: 32px"}
By the end of this lecture, you will be able to:

- **Explain** why two marginal probabilities do not determine the probability of a joint event

- **Construct** a joint probability function from a sample space, and check it against Theorem 5.1

- **Evaluate** a joint distribution function $F(y_1, y_2)$ from a table, including the rectangle rule of Theorem 5.2

- **Compute** a joint probability as a double integral of a joint density over a sketched region

- **Extend** the definitions from two random variables to $n$
:::

---

## 🗺️ Where We Are

::: {style="font-size: 32px"}
**Wackerly §5.1–5.2 — the start of Chapter 5**

Saturday closed Chapter 4: one continuous random variable at a time — its density, its expected values, its MGF, and Tchebysheff's bound on how far it strays.

::: {.fragment}
Every model so far has described **one** quantity. A loan book, a portfolio, a payment system is never one quantity.
:::

::: {.fragment}
Today the object becomes a **pair** $(Y_1, Y_2)$, and the question becomes how two random variables behave *together*. This chapter carries the rest of the course.
:::
:::

---

## ❓ The Question This Lecture Answers

::: {style="font-size: 30px"}
Two SME borrowers of a Baku bank, both in construction, each default within the year with probability $0.07$. What is the probability that **both** default?

::: {.fragment}
- If they are unrelated: $0.07 \times 0.07 = 0.0049$
- If they always move together: $0.07$
:::

::: {.fragment}
A fourteen-fold range, and the two marginals cannot choose between them. The answer lives in the **joint** distribution, which §5.2 defines.
:::

::: {.fragment}
§5.1: every sample $(y_1, \ldots, y_n)$ is also such an intersection, $(Y_1 = y_1, \ldots, Y_n = y_n)$. We build the tools with $n = 2$, then extend.
:::
:::

---

## 📐 Definition 5.1 and Theorem 5.1

::: {style="font-size: 29px"}
::: {.callout-note}
## Definition 5.1
Let $Y_1$ and $Y_2$ be discrete random variables. The **joint (or bivariate) probability function** for $Y_1$ and $Y_2$ is
$$p(y_1, y_2) = P(Y_1 = y_1, Y_2 = y_2), \quad -\infty < y_1 < \infty,\ -\infty < y_2 < \infty.$$
:::

::: {.callout-important}
## Theorem 5.1
1. $p(y_1, y_2) \ge 0$ for all $y_1, y_2$.
2. $\sum_{y_1, y_2} p(y_1, y_2) = 1$, summing over all pairs given nonzero probability.
:::
:::

---

## 🏦 Routing Two Loan Applications

::: {style="font-size: 28px"}
Two applications are each sent **at random, independently**, to committee 1 (Corporate), 2 (SME) or 3 (Retail). $Y_1$ = number sent to Corporate, $Y_2$ = number to SME. All $3 \times 3 = 9$ routings $\{i, j\}$ are equally likely.

| | $y_1 = 0$ | $y_1 = 1$ | $y_1 = 2$ |
|---|:---:|:---:|:---:|
| $y_2 = 0$ | $1/9$ | $2/9$ | $1/9$ |
| $y_2 = 1$ | $2/9$ | $2/9$ | $0$ |
| $y_2 = 2$ | $1/9$ | $0$ | $0$ |

::: {.fragment}
$p(1,1) = 2/9$: both $\{1,2\}$ and $\{2,1\}$ give it. The zeros say $Y_1 + Y_2 \le 2$: **the support is a region, not a rectangle**.
:::
:::

---

## 📈 Joint Returns of Two Assets

::: {style="font-size: 28px"}
Monthly returns (%) on a Baku-listed bank share, $Y_1$, and an oil-sector bond fund, $Y_2$:

| | $y_2 = -1$ | $y_2 = 1$ | $y_2 = 3$ |
|---|:---:|:---:|:---:|
| $y_1 = -2$ | $0.10$ | $0.08$ | $0.02$ |
| $y_1 = 0$ | $0.06$ | $0.30$ | $0.09$ |
| $y_1 = 4$ | $0.02$ | $0.12$ | $0.21$ |

::: {.fragment}
- Both gain: $p(4,1) + p(4,3) = 0.33$. Both lose: $p(-2,-1) = 0.10$.
- An equal-weight portfolio gains when $y_1 + y_2 > 0$: six cells, total $0.76$.
:::
:::

---

## 💻 The Table in R

```{r}
#| label: joint-returns
p <- matrix(c(0.10, 0.08, 0.02,
              0.06, 0.30, 0.09,
              0.02, 0.12, 0.21), nrow = 3, byrow = TRUE,
            dimnames = list(y1 = c(-2, 0, 4), y2 = c(-1, 1, 3)))
y1 <- as.numeric(rownames(p)); y2 <- as.numeric(colnames(p))
Fjoint <- function(a, b) sum(p[y1 <= a, y2 <= b])     # Definition 5.2
c(min = min(p), total = sum(p),                        # Theorem 5.1
  both_gain = sum(p[y1 > 0, y2 > 0]),
  port_gain = sum(p[outer(y1, y2, "+") > 0]),
  F01 = Fjoint(0, 1), F4m1 = Fjoint(4, -1))
```

---

## 🖼️ Mass Sits on the Diagonal

```{r}
#| label: joint-returns-fig
#| echo: false
#| fig-width: 10
#| fig-height: 4.6
d <- expand.grid(y1 = y1, y2 = y2)
d$p <- as.vector(p)
ggplot(d, aes(y1, y2)) +
  geom_point(aes(size = p), colour = "#8b2635", alpha = 0.8) +
  geom_text(aes(label = sprintf("%.2f", p)), nudge_x = 0.75, size = 7) +
  scale_size_area(max_size = 26, guide = "none") +
  scale_x_continuous(breaks = y1, limits = c(-2.7, 5.2)) +
  scale_y_continuous(breaks = y2, limits = c(-1.8, 3.8)) +
  labs(x = "Bank share return (%)", y = "Bond fund return (%)")
```

::: {style="font-size: 28px"}
The large masses run from bottom left to top right: the two assets tend to move together.
:::

---

## 📐 Definition 5.2: The Joint CDF

::: {style="font-size: 29px"}
::: {.callout-note}
## Definition 5.2
For any random variables $Y_1$ and $Y_2$, the **joint distribution function** is
$F(y_1, y_2) = P(Y_1 \le y_1, Y_2 \le y_2)$. For discrete variables, $F(y_1, y_2) = \sum_{t_1 \le y_1} \sum_{t_2 \le y_2} p(t_1, t_2)$.
:::

On the returns table:

- $F(0, 1) = 0.10 + 0.08 + 0.06 + 0.30 = 0.54$
- $F(0.5, 1.5) = 0.54$ as well — no new mass between the grid points
- $F(-3, 5) = 0$ and $F(5, 5) = 1$
:::

---

## 📐 Theorem 5.2: Properties of $F$

::: {style="font-size: 28px"}
::: {.callout-important}
## Theorem 5.2
1. $F(-\infty, -\infty) = F(-\infty, y_2) = F(y_1, -\infty) = 0$.
2. $F(\infty, \infty) = 1$.
3. If $y_1^* \ge y_1$ and $y_2^* \ge y_2$, then $F(y_1^*, y_2^*) - F(y_1^*, y_2) - F(y_1, y_2^*) + F(y_1, y_2) \ge 0$.
:::

::: {.fragment}
Part 3 is $P(y_1 < Y_1 \le y_1^*,\ y_2 < Y_2 \le y_2^*)$: the rectangle rule. On the returns table,
$$P(-2 < Y_1 \le 4,\ -1 < Y_2 \le 3) = 1 - 0.18 - 0.20 + 0.10 = 0.72,$$
which is the four cells $0.30 + 0.09 + 0.12 + 0.21$.
:::
:::

---

## 📝 Quiz #1: Reading $F$ From a Table {.quiz-question}

In the joint returns table, what is $F(4, -1) = P(Y_1 \le 4, Y_2 \le -1)$?

- [$0.18$]{.correct data-explanation="✅ Every value of Y1 is at most 4, so only the condition Y2 ≤ −1 bites: the whole y2 = −1 column, 0.10 + 0.06 + 0.02 = 0.18."}
- $0.02$
- $0.10$
- $1$

---

## 📐 Definition 5.3 and Theorem 5.3

::: {style="font-size: 28px"}
::: {.callout-note}
## Definition 5.3
If there is a nonnegative $f(y_1, y_2)$ with
$F(y_1, y_2) = \int_{-\infty}^{y_1} \int_{-\infty}^{y_2} f(t_1, t_2)\, dt_2\, dt_1$ for all $y_1, y_2$, then $Y_1, Y_2$ are **jointly continuous**, and $f$ is the **joint probability density function**.
:::

::: {.callout-important}
## Theorem 5.3
1. $f(y_1, y_2) \ge 0$ for all $y_1, y_2$.
2. $\int_{-\infty}^{\infty} \int_{-\infty}^{\infty} f(y_1, y_2)\, dy_1\, dy_2 = 1$.
:::

Probabilities are now **volumes**: $P(a_1 \le Y_1 \le a_2,\ b_1 \le Y_2 \le b_2) = \int_{b_1}^{b_2} \int_{a_1}^{a_2} f(y_1, y_2)\, dy_1\, dy_2$.
:::

---

## ⏱️ Two Settlement Times

::: {style="font-size: 29px"}
An AZN/USD foreign-exchange trade settles in two legs. $Y_1$ = hours until the **first** leg settles, $Y_2$ = hours until the **second**. By construction $Y_1 \le Y_2$. Suppose
$$f(y_1, y_2) = \begin{cases} 2e^{-(y_1 + y_2)}, & 0 \le y_1 \le y_2 < \infty,\\ 0, & \text{elsewhere.} \end{cases}$$

::: {.fragment}
**Check Theorem 5.3.** $f \ge 0$, and, integrating $y_1$ first over $0 \le y_1 \le y_2$,
$$\int_0^\infty 2e^{-y_2}\left(1 - e^{-y_2}\right) dy_2 = 2\left(1 - \tfrac{1}{2}\right) = 1.$$
:::
:::

---

## ✏️ Worked Example: Two Probabilities

::: {style="font-size: 28px"}
**Both legs settled within one hour.** Only $Y_2 \le 1$ bites, because $Y_1 \le Y_2$:
$$F(1, 1) = \int_0^1 \int_0^{y_2} 2e^{-(y_1 + y_2)}\, dy_1\, dy_2 = 2(1 - e^{-1}) - (1 - e^{-2}) = (1 - e^{-1})^2 = 0.3996$$

::: {.fragment}
**First leg within 30 minutes, second after an hour.** The rectangle $0 \le y_1 \le 0.5,\ y_2 > 1$ lies wholly inside the support:
$$\int_0^{0.5} \int_1^{\infty} 2e^{-(y_1 + y_2)}\, dy_2\, dy_1 = 2e^{-1}\left(1 - e^{-0.5}\right) = 0.2895$$
:::
:::

---

## ⚠️ Settlement Risk: The Gap

::: {style="font-size: 28px"}
While one leg has settled and the other has not, the bank is exposed to its counterparty. What is $P(Y_2 - Y_1 > 1)$?

::: {.fragment}
**Sketch the region first.** It is the part of the support above the line $y_2 = y_1 + 1$. For each $y_1$, $y_2$ runs from $y_1 + 1$ to $\infty$:
$$\int_0^{\infty} \int_{y_1 + 1}^{\infty} 2e^{-(y_1 + y_2)}\, dy_2\, dy_1 = \int_0^{\infty} 2e^{-1}e^{-2y_1}\, dy_1 = e^{-1} = 0.3679$$
:::

::: {.fragment}
More than a third of trades leave the bank exposed for over an hour.
:::
:::

---

## 💻 Checking the Integrals by Simulation

```{r}
#| label: settlement-sim
# Two independent exponential(1) legs; the earlier is Y1, the later Y2.
# This pair has exactly the joint density 2exp(-(y1 + y2)) on 0 <= y1 <= y2.
n  <- 100000
a  <- rexp(n); b <- rexp(n)
Y1 <- pmin(a, b); Y2 <- pmax(a, b)
rbind(simulated = c(F_1_1 = mean(Y1 <= 1 & Y2 <= 1),
                    rect  = mean(Y1 <= 0.5 & Y2 > 1),
                    gap   = mean(Y2 - Y1 > 1)),
      exact     = c((1 - exp(-1))^2, 2 * exp(-1) * (1 - exp(-0.5)), exp(-1))) |>
  round(4)
```

::: {style="font-size: 28px"}
A hundred thousand simulated trades land within $0.003$ of all three integrals — the region and the limits were right.
:::

---

## 🖼️ Where the Gap Lives

```{r}
#| label: settlement-fig
#| echo: false
#| fig-width: 10
#| fig-height: 4.8
s <- data.frame(y1 = Y1[1:3000], y2 = Y2[1:3000])
s$gap <- ifelse(s$y2 - s$y1 > 1, "Gap over 1 hour", "Gap under 1 hour")
ggplot(s, aes(y1, y2, colour = gap)) +
  geom_point(alpha = 0.45, size = 1.6) +
  geom_abline(intercept = 0, slope = 1, linewidth = 0.9) +
  geom_abline(intercept = 1, slope = 1, linetype = "dashed", linewidth = 0.9) +
  scale_colour_manual(values = c("Gap over 1 hour" = "#8b2635",
                                 "Gap under 1 hour" = "#8ba3c7"), name = NULL) +
  coord_cartesian(xlim = c(0, 3), ylim = c(0, 5)) +
  labs(x = expression("First leg " * Y[1] * " (hours)"),
       y = expression("Second leg " * Y[2] * " (hours)")) +
  theme(legend.position = "top", legend.text = element_text(size = 18))
```

::: {style="font-size: 28px"}
No point falls below the solid line $y_2 = y_1$; the red points above the dashed line are the event.
:::

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 29px"}
Back to the two construction borrowers. $Y_i = 1$ if borrower $i$ defaults, $0$ if not:

| | $y_2 = 0$ | $y_2 = 1$ |
|---|:---:|:---:|
| $y_1 = 0$ | $0.88$ | $0.05$ |
| $y_1 = 1$ | $0.05$ | $c$ |

**Four minutes, in pairs:** (1) find $c$; (2) find the probability that at least one defaults; (3) compare $P(\text{both})$ with $0.07 \times 0.07$.
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 29px"}
1. Theorem 5.1: the entries sum to 1, so $c = 1 - 0.88 - 0.05 - 0.05 = 0.02$. Each borrower still defaults with probability $0.05 + 0.02 = 0.07$.

2. "At least one" fails only at $(0, 0)$: $1 - p(0,0) = 1 - 0.88 = 0.12$.

::: {.fragment}
3. $P(Y_1 = 1, Y_2 = 1) = 0.02$, about **four times** the $0.0049$ that unrelated borrowers would give.
:::

::: {.fragment}
Same marginals, very different risk of a double loss. That is the opening question answered: **only the joint table could say.**
:::
:::

---

## 📝 Quiz #2: Making a Density Valid {.quiz-question}

A credit line's limit $Y_1$ and the amount drawn $Y_2$ (million AZN) have constant density $f(y_1, y_2) = k$ on the triangle $0 \le y_2 \le y_1 \le 2$. What is $k$?

- [$1/2$]{.correct data-explanation="✅ Theorem 5.3 needs the volume to be 1. The triangle has area (1/2)(2)(2) = 2, so k × 2 = 1 and k = 1/2."}
- $1$
- $1/4$
- $2$

---

## 📋 Key Formulas

::: {style="font-size: 28px"}
| | Statement |
|---|---|
| Definition 5.1 | $p(y_1, y_2) = P(Y_1 = y_1, Y_2 = y_2)$ |
| Theorem 5.1 | $p \ge 0$, $\ \sum_{y_1, y_2} p(y_1, y_2) = 1$ |
| Definition 5.2 | $F(y_1, y_2) = P(Y_1 \le y_1, Y_2 \le y_2)$ |
| Theorem 5.2, part 3 | $P(y_1 < Y_1 \le y_1^*, y_2 < Y_2 \le y_2^*) = F(y_1^*, y_2^*) - F(y_1^*, y_2) - F(y_1, y_2^*) + F(y_1, y_2)$ |
| Theorem 5.3 | $f \ge 0$, $\ \iint f(y_1, y_2)\, dy_1\, dy_2 = 1$ |
| $n$ variables | $F(y_1, \ldots, y_n) = \int_{-\infty}^{y_1} \cdots \int_{-\infty}^{y_n} f(t_1, \ldots, t_n)\, dt_n \cdots dt_1$ |
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- Marginal probabilities alone cannot price a joint event; the joint distribution can

- A joint probability function is a table of nonnegative entries summing to 1

- $F(y_1, y_2)$ accumulates mass below and to the left; differences of $F$ give rectangles

- For jointly continuous variables, probabilities are volumes under $f(y_1, y_2)$

- **Sketch the support and the event before writing the limits**

- Everything extends to $(Y_1, \ldots, Y_n)$ — the setting of every sample
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition**

- Exercises at the end of §5.2 — start with 5.1, 5.3 and 5.4 (tables), then 5.5, 5.7, 5.9 and 5.11 (densities: sketch every region first)

- On the settlement model, find $P(Y_1 + Y_2 < 2)$ and check it with the simulation code above

**Week 12, Problem Set 1** is open now and closes **Sunday 6 December at 23:59** on WeBWorK, covering §5.1–5.2.

**Next class:** 28 November — marginal and conditional probability distributions (Wackerly §5.3): how to recover each borrower's 7% from the joint table, and what one default says about the other.
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**\
📧 sorujov@ada.edu.az\
🏢 Building D, Room D325\
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- Given two marginal default probabilities of $0.07$, what are the largest and smallest values $P(\text{both default})$ can take?

- Why must a joint density's support be sketched before the limits are written, when a single density's never had to be?

- In the settlement model, is $P(Y_2 - Y_1 > h)$ a shape worth knowing for every $h$? Guess its form from $h = 1$.
:::
