---
title: "Mathematical Statistics I"
subtitle: "SESSION TITLE — Wackerly §§X.Y–X.Z"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I — SHORT TITLE"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# knitr is named explicitly so engine selection cannot drift to jupyter and
# quietly stop running the R chunks.
engine: knitr
execute:
  echo: true        # students see the code that produced the number
  warning: false
  message: false
  freeze: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
# One seed for the whole deck: every render produces the same numbers, so the
# figures on the slides match what is said out loud about them.
set.seed(2026)
library(ggplot2)
theme_set(theme_minimal(base_size = 18))
```

## 🎯 Learning Objectives

::: {style="font-size: 34px"}
By the end of this lecture, you will be able to:

- **Verb** the first objective, in a financial or economic setting

- **Verb** the second objective

- **Verb** the third objective

- **Verb** the fourth objective
:::

---

## Motivating Question

::: {.callout-important}
## The question this lecture answers
State the question in one sentence, in economic terms, before any notation.
:::

---

## Definition {.smaller}

::: {.callout-note}
## Definition (Wackerly §X.Y)
State the definition exactly as the book gives it, then restate it in one
plain sentence.
:::

$$
\text{display the notation here}
$$

---

## Worked Example — Finance Context

::: {style="font-size: 30px"}
Set the scene in two lines: a portfolio, a default count, a waiting time.

**Step 1.** What we know.

**Step 2.** The calculation, done in full.

**Step 3.** The answer, read back in economic terms.
:::

---

## Computation in R

```{r}
#| label: worked-example
# Executable: this runs at render time and its output appears below.
returns <- rnorm(500, mean = 0.05, sd = 1.2)
round(c(mean = mean(returns), sd = sd(returns)), 3)
```

---

## The Picture

```{r}
#| label: worked-figure
#| fig-height: 4.2
#| fig-width: 9
#| echo: false
ggplot(data.frame(r = returns), aes(r)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "grey85", colour = "white") +
  labs(x = "Daily return (%)", y = "Relative frequency", title = NULL)
```

---

## 🧠 Think–Pair–Share

```{r}
#| label: tps-timer
#| echo: false
# R-native timer. Renders to self-contained HTML/JS, so it runs offline in
# the hall with no clicking away from the slide.
countdown::countdown(
  minutes = 3, seconds = 0,
  top = 0, right = 0,
  font_size = "2em",
  warn_when = 30
)
```

::: {style="font-size: 32px"}
**Think** (1 min) — on your own, write down your answer to the question below.

**Pair** (1 min) — compare with your neighbour; find where you disagree.

**Share** (1 min) — one pair reports back.

> Put the single discussion question here, phrased so that a wrong answer is
> as interesting as a right one.
:::

---

## Explore It

```{ojs}
//| echo: false
viewof n = Inputs.range([5, 200], {step: 1, value: 30, label: "Sample size n"})
```

```{ojs}
//| echo: false
Plot.plot({
  width: 1050,
  height: 320,
  marginLeft: 60,
  x: {label: "Value"},
  y: {label: "Density"},
  marks: [
    Plot.rectY(sample, Plot.binX({y: "proportion"}, {x: d => d, fill: "#8ba3c7"})),
    Plot.ruleY([0])
  ]
})
```

```{ojs}
//| echo: false
sample = Array.from({length: n}, d3.randomNormal(0, 1))
```

---

## Q1: Question text goes here {.quiz-question}

- A distractor that a student could plausibly pick
- [The correct option]{.correct data-explanation="Say why it is correct in one sentence, and why the tempting distractor is not."}
- Another distractor
- A fourth option

---

## Key Formulas

::: {.callout-important}
## Take these away
$$\text{formula one}$$

$$\text{formula two}$$
:::

---

## Summary

::: {style="font-size: 32px"}
- The first thing we established

- The second

- What it sets up for next session
:::

---

## Next Session

::: {style="font-size: 32px"}
**Reading:** Wackerly §§X.Y–X.Z

**Problem Set N** closes DATE at 23:59.
:::
