---
title: "Mathematical Statistics"
subtitle: "Continuous Random Variables: Distribution and Density Functions"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - Continuous Random Variables"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
library(ggplot2)
theme_set(theme_minimal(base_size = 18))
```

## 🎯 Learning Objectives

::: {style="font-size: 32px"}
By the end of this lecture, you will be able to:

- **Write** the distribution function $F(y) = P(Y \le y)$ of any random variable, discrete or continuous

- **Explain** why $P(Y = y) = 0$ for a continuous $Y$, and what replaces a probability function

- **Move** between a density $f(y)$ and its distribution function $F(y)$ by integrating or differentiating

- **Compute** $P(a \le Y \le b)$ as an area under $f(y)$, and find the constant that makes $f(y)$ a density

- **Find** a quantile $\phi_p$, such as the median settlement time of a payment
:::

---

## 🗺️ Where We Are

::: {style="font-size: 30px"}
**Wackerly §§4.1–4.2**

Friday closed Chapter 3 with a map of the discrete models: name the counting mechanism, pick the model, read off $E(Y)$ and $V(Y)$. Every one of them assigned a probability $p(y)$ to each value.

::: {.fragment}
Today we leave counts behind. The time a payment takes to settle, the share of a credit line a firm draws, a day's move in the exchange rate: these take **any value in an interval**.
:::

::: {.fragment}
For those, $p(y)$ cannot work. Chapter 4 replaces it with two functions, $F(y)$ and $f(y)$, and today is about both.
:::
:::

---

## ❓ Motivating Question

::: {.callout-important}
## The payments desk's problem
A Baku bank sends cross-border transfers that settle anywhere between 0 and 4 hours after release. The desk promises corporate clients settlement **within 3 hours**.

*What is the probability a transfer settles in exactly 2 hours? And in at most 3?*
:::

::: {.fragment style="font-size: 30px"}
The first question has the surprising answer **zero**, and yet the second has a perfectly good answer. By the end of today we will compute it: $27/32 \approx 0.844$.
:::

---

## 📐 Definition 4.1 and Theorem 4.1

::: {style="font-size: 30px"}
::: {.callout-note}
## Definition 4.1
Let $Y$ denote any random variable. The **distribution function** of $Y$, denoted by $F(y)$, is such that $F(y) = P(Y \le y)$ for $-\infty < y < \infty$.
:::

::: {.callout-important}
## Theorem 4.1: Properties of a Distribution Function
1. $F(-\infty) \equiv \lim_{y \to -\infty} F(y) = 0$
2. $F(\infty) \equiv \lim_{y \to \infty} F(y) = 1$
3. $F(y)$ is nondecreasing: if $y_1 < y_2$ then $F(y_1) \le F(y_2)$
:::

Note the word **any**: $F(y)$ exists for discrete variables too.
:::

---

## 🪜 A Discrete F Is a Staircase

::: {style="font-size: 28px"}
Example 4.1, in a loan office: two SME applications, each approved independently with probability $1/2$. $Y$ = number approved, so $p(0) = 1/4$, $p(1) = 1/2$, $p(2) = 1/4$ and, for instance, $F(1.5) = 1/4 + 1/2 = 3/4$.
:::

```{r}
#| label: staircase
#| echo: false
#| fig-width: 10
#| fig-height: 3.9
steps <- data.frame(x0 = c(-1, 0, 1, 2), x1 = c(0, 1, 2, 3),
                    F  = c(0, 0.25, 0.75, 1))
ggplot(steps) +
  geom_segment(aes(x = x0, xend = x1, y = F, yend = F), linewidth = 1.4) +
  geom_point(aes(x = x0, y = F), data = steps[-1, ], size = 4) +
  geom_point(aes(x = x1, y = F), data = steps[-4, ], size = 4,
             shape = 21, fill = "white", stroke = 1.3) +
  scale_x_continuous(breaks = -1:3) +
  scale_y_continuous(breaks = c(0, 0.25, 0.75, 1),
                     labels = c("0", "1/4", "3/4", "1")) +
  labs(x = "y (applications approved)", y = "F(y)") +
  theme(panel.grid.minor = element_blank())
```

---

## 📐 Definition 4.2: When F Has No Jumps

::: {style="font-size: 30px"}
The staircase jumps at 0, 1, 2 by exactly $p(0)$, $p(1)$, $p(2)$. It is flat in between because those values carry no probability.

::: {.callout-note}
## Definition 4.2
A random variable $Y$ with distribution function $F(y)$ is said to be **continuous** if $F(y)$ is continuous, for $-\infty < y < \infty$.
:::

::: {.fragment}
**Consequence.** If $Y$ is continuous, $P(Y = y) = 0$ for every real $y$. A point with $P(Y = y_0) = p_0 > 0$ would put a jump of size $p_0$ into $F$. So a transfer settling in *exactly* 2.000... hours has probability zero, while one settling between 2 and 3 hours does not.
:::
:::

---

## 📐 Definition 4.3: The Density

::: {style="font-size: 29px"}
::: {.callout-note}
## Definition 4.3
Let $F(y)$ be the distribution function for a continuous random variable $Y$. Then $f(y) = \dfrac{dF(y)}{dy}$, wherever the derivative exists, is called the **probability density function** for $Y$. Equivalently, $F(y) = \int_{-\infty}^{y} f(t)\, dt$.
:::

::: {.callout-important}
## Theorem 4.2: Properties of a Density Function
1. $f(y) \ge 0$ for all $y$, $-\infty < y < \infty$
2. $\int_{-\infty}^{\infty} f(y)\, dy = 1$
:::
:::

---

## 💳 Worked Example: A Credit Line

::: {style="font-size: 29px"}
Example 4.3, re-set. $Y$ is the **fraction** of its approved credit line a corporate client has drawn at month-end:
$$f(y) = \begin{cases} 3y^2, & 0 \le y \le 1, \\ 0, & \text{elsewhere.} \end{cases}$$

::: {.fragment}
Integrate from $-\infty$ up to $y$. Below 0 nothing accumulates; above 1 everything already has:
$$F(y) = \begin{cases} 0, & y < 0, \\ \int_0^y 3t^2\, dt = y^3, & 0 \le y \le 1, \\ 1, & y > 1. \end{cases}$$
:::

::: {.fragment}
So $P(Y \le 0.5) = 0.5^3 = 0.125$: only one client in eight uses half the line or less. Clients draw heavily.
:::
:::

---

## 📈 f and F Side by Side

```{r}
#| label: credit-line-figure
#| echo: false
#| fig-width: 11
#| fig-height: 4.6
yy <- seq(0, 1, by = 0.005)
cl <- rbind(
  data.frame(y = yy, v = 3 * yy^2, panel = "Density~f(y)==3*y^2"),
  data.frame(y = yy, v = yy^3,     panel = "Distribution~'function'~F(y)==y^3"))
med <- 0.5^(1/3)
marks <- data.frame(panel = c("Density~f(y)==3*y^2", "Distribution~'function'~F(y)==y^3"),
                    v = c(3 * med^2, 0.5))
ggplot(cl, aes(y, v)) +
  geom_area(data = subset(cl, panel == "Density~f(y)==3*y^2" & y <= med),
            fill = "#8ba3c7", alpha = 0.6) +
  geom_line(linewidth = 1.3) +
  geom_segment(data = marks, aes(x = med, xend = med, y = 0, yend = v),
               linetype = "dashed", colour = "#8b2635", linewidth = 1) +
  facet_wrap(~ panel, scales = "free_y", labeller = label_parsed) +
  labs(x = "y (fraction of credit line drawn)", y = NULL) +
  theme(strip.text = element_text(size = 19, face = "bold"))
```

::: {style="font-size: 28px"}
The shaded area on the left, 0.5, is the **height** of $F$ on the right at the same $y$. And $f(1) = 3$: a density may exceed 1; $F$ never does.
:::

---

## 📐 Definition 4.4: Quantiles

::: {style="font-size: 30px"}
::: {.callout-note}
## Definition 4.4
Let $Y$ denote any random variable. If $0 < p < 1$, the $p$th **quantile** of $Y$, denoted by $\phi_p$, is the smallest value such that $F(\phi_p) = P(Y \le \phi_p) \ge p$. If $Y$ is continuous, $\phi_p$ is the smallest value such that $F(\phi_p) = p$.
:::

::: {.fragment}
Credit line: solve $\phi_p^3 = p$.

- Median: $\phi_{.5} = 0.5^{1/3} = 0.7937$. On a 500,000 AZN line, half of clients have drawn more than 396,850 AZN.
- $\phi_{.90} = 0.9^{1/3} = 0.9655$: the bank's liquidity desk plans for 96.5% utilisation in the top decile.
:::
:::

---

## 📐 Theorem 4.3: Probability Is Area

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorem 4.3
If the random variable $Y$ has density function $f(y)$ and $a < b$, then the probability that $Y$ falls in the interval $[a, b]$ is
$$P(a \le Y \le b) = \int_a^b f(y)\, dy = F(b) - F(a)$$
:::

::: {.fragment}
Because $P(Y = a) = P(Y = b) = 0$, the endpoints do not matter:
$$P(a < Y < b) = P(a \le Y < b) = P(a < Y \le b) = P(a \le Y \le b)$$
**This is false for discrete variables.** In the loan office, $P(0 < Y < 2) = 1/2$ but $P(0 \le Y \le 2) = 1$.
:::
:::

---

## 🏦 Worked Example: Settlement Time

::: {style="font-size: 29px"}
Examples 4.4–4.5, re-set. Settlement time $Y$ (hours) has $f(y) = c\,y(4 - y)$ for $0 \le y \le 4$, zero elsewhere. **Step 1: find $c$.**
$$\int_0^4 c(4y - y^2)\, dy = c\left[2y^2 - \frac{y^3}{3}\right]_0^4 = c\left(32 - \frac{64}{3}\right) = \frac{32}{3}c = 1 \;\Rightarrow\; c = \frac{3}{32}$$

::: {.fragment}
**Step 2: the distribution function.** For $0 \le y \le 4$, $F(y) = \dfrac{6y^2 - y^3}{32}$.
:::

::: {.fragment}
**Step 3: read off the desk's answers.**
$$P(Y \le 3) = \frac{54 - 27}{32} = \frac{27}{32} = 0.8438, \qquad P(1 \le Y \le 3) = \frac{27}{32} - \frac{5}{32} = 0.6875$$
The 3-hour promise is broken for about **one transfer in six**, $5/32 = 0.156$.
:::
:::

---

## 💻 Checking It in R

```{r}
#| label: settlement-check
#| code-fold: false
f <- function(y) ifelse(y >= 0 & y <= 4, y * (4 - y), 0)
c_const <- 1 / integrate(f, 0, 4)$value            # should be 3/32
dens <- function(y) c_const * f(y)
Fy   <- function(y) integrate(dens, 0, y)$value

c(c = c_const, P_le_3 = Fy(3), P_1_to_3 = Fy(3) - Fy(1))

# quantiles: solve F(phi) = p numerically
q <- function(p) uniroot(function(y) Fy(y) - p, c(0, 4))$root
round(c(median = q(0.5), phi_90 = q(0.9)), 4)
```

::: {style="font-size: 28px"}
The median is 2 hours, by symmetry. Nine transfers in ten settle within 3.22 hours.
:::

---

## 🔬 Interactive: Area Under f Is the Height of F {.smaller}

```{ojs}
//| echo: false

viewof b = {
  const input = Inputs.range([0, 4], {value: 3, step: 0.05,
    label: "Settlement deadline b (hours):"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}
```

```{ojs}
//| echo: false

dens = y => (3 / 32) * y * (4 - y)
cdf = y => (6 * y * y - y * y * y) / 32
grid = Array.from({length: 161}, (_, i) => ({y: i * 0.025, f: dens(i * 0.025)}))
shaded = grid.filter(d => d.y <= b + 1e-9)

md`Shaded area = F(${b.toFixed(2)}) = P(Y ≤ ${b.toFixed(2)}) = **${cdf(b).toFixed(4)}**`

Plot.plot({
  width: 1150,
  height: 320,
  marginLeft: 78,
  marginBottom: 58,
  style: {fontSize: "18px"},
  x: {label: "Settlement time y (hours)", domain: [0, 4], ticks: [0, 0.5, 1, 1.5, 2, 2.5, 3, 3.5, 4]},
  y: {label: "Density f(y)", domain: [0, 0.4], ticks: [0, 0.1, 0.2, 0.3, 0.4], tickFormat: ".1f"},
  marks: [
    Plot.areaY(shaded, {x: "y", y: "f", fill: "#8ba3c7", fillOpacity: 0.7}),
    Plot.line(grid, {x: "y", y: "f", stroke: "#14130f", strokeWidth: 2.5}),
    Plot.ruleX([b], {stroke: "#8b2635", strokeWidth: 2, strokeDasharray: "5 4"}),
    Plot.ruleY([0])
  ]
})
```

---

## 🏧 From F to f: Service Time at a Branch

::: {style="font-size: 29px"}
Example 4.2 runs the other way: start from $F$ and differentiate. A branch's teller service time $Y$ (minutes) has
$$F(y) = 1 - e^{-y^2/50}, \quad y \ge 0 \qquad \Rightarrow \qquad f(y) = \frac{dF(y)}{dy} = \frac{y}{25}\,e^{-y^2/50}, \quad y \ge 0$$

::: {.fragment}
- A customer takes longer than 10 minutes: $1 - F(10) = e^{-2} = 0.1353$
- Between 5 and 10 minutes: $F(10) - F(5) = e^{-0.5} - e^{-2} = 0.4712$
- Median: $1 - e^{-y^2/50} = 0.5$ gives $\phi_{.5} = \sqrt{50 \ln 2} = 5.89$ minutes
:::

::: {.fragment}
No integral was needed: when $F$ is given, **every probability is a subtraction**.
:::
:::

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 30px"}
The daily move $Y$ (in %) of the EUR/AZN rate is modelled by
$$f(y) = 1 - |y|, \quad -1 \le y \le 1, \qquad 0 \text{ elsewhere.}$$

**Four minutes, in pairs:**

1. Check that $f$ is a density. (Sketch it first.)

2. Find $P(|Y| > 0.5)$, the chance of a move of more than half a percent.

3. Find $\phi_{.95}$. And what is $P(Y = 0)$, where $f$ is highest?
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 29px"}
1. $f \ge 0$, and the graph is a triangle with base 2 and height 1: area $= \tfrac{1}{2} \times 2 \times 1 = 1$.

2. Each tail beyond $\pm 0.5$ is a triangle with base and height $0.5$, area $0.125$. So $P(|Y| > 0.5) = 0.25$.

::: {.fragment}
3. For $0 \le y \le 1$, $F(y) = 1 - \tfrac{(1-y)^2}{2}$. Setting $F(\phi) = 0.95$ gives $(1 - \phi)^2 = 0.1$, so
$$\phi_{.95} = 1 - \sqrt{0.1} = 0.6838\%$$
A move above 0.68% happens on about one trading day in twenty.
:::

::: {.fragment}
And $P(Y = 0) = 0$. **$f(0) = 1$ is a height, not a probability**; probability lives only in areas.
:::
:::

---

## 📝 Quiz #1: Height or Probability? {.quiz-question}

The settlement density has $f(2) = 0.375$. What is $P(Y = 2)$?

- [$0$]{.correct data-explanation="✅ Y is continuous, so F has no jumps and P(Y = 2) = 0. The value f(2) = 0.375 is a density height; probability is area, and a single point has none."}
- $0.375$
- $0.5$, because 2 is the median
- It cannot be found without $F(y)$

---

## 📝 Quiz #2: Making a Density {.quiz-question}

A density is $f(y) = cy$ for $0 \le y \le 4$ and $0$ elsewhere. What is $c$?

- $1/4$
- [$1/8$]{.correct data-explanation="✅ Theorem 4.2 needs the total area to be 1: the integral of cy from 0 to 4 is 8c, so c = 1/8. (It is a triangle of base 4 and height 4c.)"}
- $1/16$
- $1/2$

---

## 📝 Quiz #3: Reading F {.quiz-question}

The credit-line fraction has $F(y) = y^3$ on $[0, 1]$. What is the probability a client has drawn **more than half** the line?

- $0.125$
- $0.5$
- [$0.875$]{.correct data-explanation="✅ P(Y > 0.5) = 1 − F(0.5) = 1 − 0.125 = 0.875. The 0.125 is the complementary event, Y ≤ 0.5."}
- $0.75$

---

## 📋 Key Formulas

::: {style="font-size: 30px"}
| | Statement |
|---|---|
| Definition 4.1 | $F(y) = P(Y \le y)$, for any random variable |
| Theorem 4.1 | $F(-\infty) = 0$, $F(\infty) = 1$, $F$ nondecreasing |
| continuous $Y$ | $P(Y = y) = 0$ for every $y$ |
| Definition 4.3 | $f(y) = dF(y)/dy$, $\;F(y) = \int_{-\infty}^{y} f(t)\,dt$ |
| Theorem 4.2 | $f(y) \ge 0$, $\;\int_{-\infty}^{\infty} f(y)\,dy = 1$ |
| Theorem 4.3 | $P(a \le Y \le b) = \int_a^b f(y)\,dy = F(b) - F(a)$ |
| Definition 4.4 | $F(\phi_p) = p$, $Y$ continuous |
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- $F(y) = P(Y \le y)$ describes **every** random variable; a discrete one has a staircase, a continuous one has none

- For continuous $Y$, single points carry probability zero, so endpoints never matter

- The density $f$ is the derivative of $F$; $F$ is the accumulated area under $f$

- A density is a height, not a probability, and may exceed 1

- Probabilities are areas, $F(b) - F(a)$; quantiles invert $F$
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition**

- §4.2: Exercises 4.1 – 4.19; start with 4.8, 4.11, 4.12 and 4.14

- Exercise 4.7 is the discrete-endpoints warning from Theorem 4.3, worked from the binomial table

- Redo the settlement example with the deadline moved to 2.5 hours: how often is the promise broken?

**Week 9, Problem Set 1** is open now and closes **Sunday 15 November at 23:59** on WeBWorK, covering §§4.1–4.2.

**Next class:** 7 November, expected values for continuous random variables, and our first named continuous model, the uniform distribution (Wackerly §§4.3–4.4).
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**

📧 sorujov@ada.edu.az\
🏢 Building D, Room D325\
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- A distribution function can have jumps **and** rise smoothly in between. What kind of variable would that describe in insurance?

- If $f(y) = 3y^2$ on $[0,1]$, why is $f(1) = 3$ not a contradiction of $P \le 1$?

- The bank reports the 90th percentile of settlement time rather than the mean. Why might a client prefer that number?
:::
