---
title: "Mathematical Statistics"
subtitle: "Continuous Random Variables and Probability Distributions"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics - Continuous Random Variables"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives

::: {style="font-size: 30px"}
::: {.learning-objectives}
By the end of this lecture, you will be able to:

- Define continuous random variables and distinguish them from discrete random variables in financial contexts (such as stock price movements)

- Derive and interpret distribution functions $F(y)$ and probability density functions $f(y)$ for continuous random variables

- Apply properties of distribution and density functions to calculate probabilities over intervals

- Compute quantiles and percentiles (including the median) for continuous distributions to support risk analysis

- Use real financial data to model continuous random variables and evaluate probabilistic outcomes
:::
:::

## 📋 Overview

::: {style="font-size:38px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Introduction** – Continuous vs discrete random variables with economic examples

- **Distribution Functions** – Definitions, properties, and cumulative distribution interpretation

- **Probability Density Functions** – Derivatives of CDFs and their properties

- **Quantiles & Percentiles** – Median and other measures for financial risk assessment

- **Applications** – Real-world financial data analysis and probability calculations
:::
:::
:::

## 📖 Introduction to Continuous Random Variables

::: {.callout-note}
## 🔍 Motivation

In probability theory and statistics, random variables model real-world phenomena:

::: {.incremental}
- **Discrete random variables** – Take countable values (e.g., number of defaults in a loan portfolio)

- **Continuous random variables** – Take any value within an interval (uncountably infinite outcomes)

- **Economic example** – Daily stock price movements: prices vary continuously and can fluctuate between any two real values

- **Key insight** – For continuous variables defined over positive real line segments, no single value can be entirely ruled out as a potential outcome
:::
:::

## 📖 Definition: Distribution Function {.larger}

::: {.callout-note}
## 📝 Definition 1: Distribution Function

Let $Y$ denote any random variable. The **distribution function** of $Y$, denoted by $F(y)$, is defined as:
$$F(y) = P(Y \leq y) \quad \text{for } -\infty < y < \infty$$

::: {.incremental}
- **Property 1** – $F(y)$ represents the probability that $Y$ takes on a value less than or equal to $y$ for all possible values of $y$ (cumulative probability interpretation)

- **Property 2** – $F(y)$ is also called the **cumulative distribution function (CDF)**

- **Property 3** – The CDF applies to both discrete and continuous random variables
:::
:::

## 📌 Example 1: Binomial Distribution Function {.smaller}

**Problem:** Suppose $Y$ has a binomial distribution with $n = 2$ and $p = \frac{1}{2}$. Find the distribution function $F(y)$.

. . .

**Solution:**

. . .

**Step 1:** For $y < 0$, no binomial value exists:
$$F(y) = P(Y \leq y) = 0$$

. . .

**Step 2:** For $0 \leq y < 1$, only $Y = 0$ is included:
$$F(y) = P(Y = 0) = \binom{2}{0} \left(\frac{1}{2}\right)^0 \left(\frac{1}{2}\right)^{2} = \frac{1}{4}$$

::: {.columns}
::: {.column widht="50%"}

**Step 3:** For $1 \leq y < 2$, include $Y = 0$ and $Y = 1$:
$$F(y) = P(Y = 0) + P(Y = 1) = \frac{1}{4} + \frac{1}{2} = \frac{3}{4}$$
:::

::: {.column widht="50%" .fragment}
**Step 4:** For $y \geq 2$, all values included:
$$F(y) = P(Y = 0) + P(Y = 1) + P(Y = 2) = \frac{1}{4} + \frac{1}{2} + \frac{1}{4} = 1$$

:::
:::

## 📌 Example 1: Piecewise CDF {.smaller}

::: {style="font-size:30px"}
The distribution function for the binomial random variable $Y$ with $n=2$ and $p=0.5$ is:

$$F(y) = \begin{cases}
0, & \text{if } y < 0 \\
\frac{1}{4}, & \text{if } 0 \leq y < 1 \\
\frac{3}{4}, & \text{if } 1 \leq y < 2 \\
1, & \text{if } y \geq 2
\end{cases}$$

**Key observations:**

- This is a **step function** (discrete distribution)

- The CDF jumps at integer values (0, 1, 2)

- Between jumps, $F(y)$ remains constant
:::

## 🧮 Theorem: Properties of Distribution Functions {.smaller}

::: {.callout-important}
## Theorem 1: Properties of $F(y)$

If $F(y)$ is a distribution function, then it possesses the following properties:
:::

. . .

**Property 1:** $F(-\infty) \equiv \lim_{y \to -\infty} F(y) = 0$

- As $y$ approaches negative infinity, the probability approaches zero (no outcomes below $-\infty$)

. . .

**Property 2:** $F(\infty) \equiv \lim_{y \to \infty} F(y) = 1$

- As $y$ approaches positive infinity, the cumulative probability approaches one (all outcomes eventually included)

. . .

**Property 3:** $F(y)$ is a non-decreasing function of $y$

- For any $y_1 < y_2$, we have $F(y_1) \leq F(y_2)$ (probabilities cannot decrease as we move right on the number line)

## 📖 Definition: Continuous Random Variable {.larger}

::: {.callout-note}
## 📝 Definition 2: Continuous Random Variable

A random variable $Y$ with distribution function $F(y)$ is said to be **continuous** if the distribution function $F(y)$ is continuous for $-\infty < y < \infty$.

::: {.incremental}
- **Interpretation** – $Y$ is continuous if its distribution function forms a smooth, unbroken curve (no jumps or discontinuities)

- **Distinction from discrete** – Discrete variables have step-function CDFs; continuous variables have smooth CDFs

- **Examples** – Stock returns, interest rates, asset prices, portfolio values
:::
:::

## 📖 Property: Zero Point Probability

::: {.callout-important}
## Property: Point Probability for Continuous Variables

If $Y$ is a continuous random variable, then for any real number $y$:
$$P(Y = y) = 0$$

**Key implications:**

- The probability of $Y$ taking on any **exact** value is zero

- Continuous random variables can assume an uncountably infinite number of values within an interval

- Probabilities are meaningful only over **intervals**, not at individual points

- This is why $P(a \leq Y \leq b) = P(a < Y < b)$ for continuous $Y$
:::

## 📖 Definition: Probability Density Function {.larger}

::: {.callout-note}
## 📝 Definition 3: Probability Density Function (PDF)

Let $F(y)$ be the distribution function for a continuous random variable $Y$. The function $f(y)$, given by:
$$f(y) = \frac{dF(y)}{dy} = F'(y)$$
wherever the derivative exists, is called the **probability density function (PDF)** for $Y$.

::: {.incremental}
- **Property 1** – $f(y)$ is the derivative of the CDF (rate of change of cumulative probability)

- **Property 2** – Conversely, $F(y) = \int_{-\infty}^{y} f(t) \, dt$ (CDF is the integral of PDF)

- **Property 3** – $f(y)$ itself is not a probability; it is a density (probability per unit length)
:::
:::

## 🧮 Theorem: Properties of Density Functions {.smaller}

::: {.callout-important}
## Theorem 2: Properties of PDF

If $f(y)$ is a probability density function for a continuous random variable, then:
:::

. . .

**Property 1:** $f(y) \geq 0$ for all $y$, $-\infty < y < \infty$

- The density function is always non-negative (densities cannot be negative)

. . .

**Property 2:** $\int_{-\infty}^{\infty} f(y) \, dy = 1$

- The total area under the density curve equals one (total probability is 1)

. . .

**Verification:** To check if $f(y)$ is a valid PDF, verify both properties above.


## 📌 Example 2: Finding PDF from CDF {.smaller}

**Problem:** Given the distribution function:
$$F(y) = \begin{cases}
0, & \text{for } y < 0 \\
y, & \text{for } 0 \leq y \leq 1 \\
1, & \text{for } y > 1
\end{cases}$$
Find the probability density function $f(y)$.

. . .

**Solution:** Take the derivative $f(y) = \frac{dF(y)}{dy}$:

. . .

**Region 1:** For $y < 0$, $F(y) = 0$ (constant), so $f(y) = 0$

. . .

**Region 2:** For $0 \leq y \leq 1$, $F(y) = y$ (linear), so $f(y) = 1$

. . .

**Region 3:** For $y > 1$, $F(y) = 1$ (constant), so $f(y) = 0$

. . .

**Result:** This is the **uniform distribution** on $[0,1]$:
$$f(y) = \begin{cases}
1, & 0 \leq y \leq 1 \\
0, & \text{elsewhere}
\end{cases}$$

## 📌 Example 3: Finding CDF from PDF {.smaller}

**Problem:** Let $Y$ be a continuous random variable with PDF:
$$f(y) = \begin{cases}
3y^2, & 0 \leq y \leq 1 \\
0, & \text{elsewhere}
\end{cases}$$
Find the distribution function $F(y)$.

. . .

**Solution:** Integrate $F(y) = \int_{-\infty}^{y} f(t) \, dt$:

. . .

**Region 1:** For $y < 0$:
$$F(y) = \int_{-\infty}^{y} 0 \, dt = 0$$

::: {.columns}

::: {.column width="50%" .fragment}

**Region 2:** For $0 \leq y \leq 1$:
$$F(y) = \int_{-\infty}^{0} 0 \, dt + \int_{0}^{y} 3t^2 \, dt = \left[t^3\right]_0^y = y^3$$

:::

::: {.column width="50%" .fragment}

**Region 3:** For $y > 1$:
$$F(y) = \int_{0}^{1} 3t^2 \, dt = \left[t^3\right]_0^1 = 1$$

:::
:::

## 📌 Example 3: Result {.smaller}

::: {style="font-size:30px"}
The distribution function is:
$$F(y) = \begin{cases}
0, & y < 0 \\
y^3, & 0 \leq y \leq 1 \\
1, & y > 1
\end{cases}$$

**Key observations:**

- The CDF is continuous (smooth curve, no jumps)

- $F(y)$ is strictly increasing on $[0,1]$

- The PDF $f(y) = 3y^2$ is continuous on $(0,1)$, though it jumps to zero at the boundaries

**Remark:** The distribution function for a continuous random variable must be continuous, but the density function need not be everywhere continuous.
:::

## 🎮 Interactive: PDF and CDF Visualization {.smaller}

::: {style="font-size: 0.8em;"}

**Explore the Relationship:** Adjust the parameter $a$ to see how the PDF $f(y) = ay^{a-1}$ on $[0,1]$ changes the CDF.

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof a_param = Inputs.range([0.5, 5], {
  value: 3, 
  step: 0.5, 
  label: "Parameter a:"
})

md`**Current Parameter:**  
a = ${a_param.toFixed(1)}  

**PDF:** f(y) = ${a_param.toFixed(1)} y<sup>${(a_param-1).toFixed(1)}</sup>  

**CDF:** F(y) = y<sup>${a_param.toFixed(1)}</sup>`
```

**Observations:**

- When a > 1, density increases toward y=1

- When a < 1, density decreases toward y=1

- When a = 1, uniform distribution

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// Generate data for PDF and CDF
y_values = d3.range(0, 1.01, 0.01)

pdf_data = y_values.map(y => ({
  y: y,
  value: y === 0 && a_param < 1 ? null : a_param * Math.pow(y, a_param - 1)
})).filter(d => d.value !== null)

cdf_data = y_values.map(y => ({
  y: y,
  value: Math.pow(y, a_param)
}))

Plot.plot({
  width: 800,
  height: 450,
  marginLeft: 60,
  marginBottom: 40,
  x: {
    label: "y",
    domain: [0, 1],
    grid: true
  },
  y: {
    label: "Function Value",
    domain: [0, Math.max(5, a_param * 1.2)],
    grid: true
  },
  marks: [
    Plot.line(pdf_data, {x: "y", y: "value", stroke: "steelblue", strokeWidth: 3, tip: true}),
    Plot.line(cdf_data, {x: "y", y: "value", stroke: "orange", strokeWidth: 2.5, strokeDasharray: "5,5", tip: true}),
    Plot.ruleX([0]),
    Plot.ruleY([0]),
    Plot.ruleY([1], {stroke: "red", strokeDasharray: "3,3", strokeWidth: 1})
  ],
  caption: html`<span style="color: steelblue; font-weight: bold;">━━</span> PDF f(y) | <span style="color: orange; font-weight: bold;">━ ━</span> CDF F(y) | <span style="color: red;">- -</span> Probability = 1`
})
```

:::
:::
:::

## 📖 Definition: Quantiles and Percentiles {.larger}

::: {.callout-note}
## 📝 Definition 4: Quantiles

Let $Y$ denote any random variable. If $0 < p < 1$, the **p-th quantile** of $Y$, denoted by $\phi_p$, is the smallest value such that:
$$P(Y \leq \phi_p) = F(\phi_p) \geq p$$

For continuous $Y$, $\phi_p$ satisfies $F(\phi_p) = p$.

::: {.incremental}
- **Percentile notation** – Some prefer to call $\phi_p$ the $100p$-th percentile of $Y$

- **Median** – If $p = 0.5$, then $\phi_{0.5}$ is the **median** of $Y$

- **Financial interpretation** – Quantiles are used in Value-at-Risk (VaR) calculations for portfolio risk assessment
:::
:::

## 📌 Example 4: Computing the Median {.smaller}

**Problem:** For the random variable $Y$ from Example 3 with $f(y) = 3y^2$ on $[0,1]$, find the median.

. . .

**Solution:** The median $\phi_{0.5}$ satisfies $F(\phi_{0.5}) = 0.5$.

. . .

**Step 1:** Recall that $F(y) = y^3$ for $0 \leq y \leq 1$.

. . .

**Step 2:** Set up the equation:
$$(\phi_{0.5})^3 = 0.5$$

. . .

**Step 3:** Solve for $\phi_{0.5}$:
$$\phi_{0.5} = (0.5)^{1/3} = 0.7937$$

. . .

**Interpretation:** The median of $Y$ is approximately 0.794. This means that $P(Y \leq 0.794) = 0.5$, so half the probability mass lies below this value (useful for summarizing skewed distributions in finance).

## 🧮 Theorem: Probability Over Intervals {.smaller}

::: {.callout-important}
## Theorem 3: Interval Probabilities

If the random variable $Y$ has density function $f(y)$ and $a < b$, then:
$$P(a \leq Y \leq b) = \int_{a}^{b} f(y) \, dy$$
:::

. . .

**Key consequences:**

. . .

**Consequence 1:** $P(a \leq Y \leq b) = F(b) - F(a)$

- The probability over $[a,b]$ is the difference in CDF values

. . .

**Consequence 2:** For continuous $Y$:
$$P(a \leq Y \leq b) = P(a < Y < b) = P(a \leq Y < b) = P(a < Y \leq b)$$

- Whether endpoints are included or excluded does not matter (since $P(Y = a) = P(Y = b) = 0$)


## 📌 Example 5: Finding Valid PDF Constants {.smaller}

**Problem:** Given $f(y) = cy^2$ for $0 \leq y \leq 2$, and $f(y) = 0$ elsewhere, find the value of $c$ for which $f(y)$ is a valid density function.

. . .

**Solution:** For $f(y)$ to be a valid PDF, it must satisfy:
$$\int_{-\infty}^{\infty} f(y) \, dy = 1$$

. . .

**Step 1:** Compute the integral over the support:
$$\int_{0}^{2} cy^2 \, dy = 1$$

::: {.columns}

::: {.column width="50%" .fragment}

**Step 2:** Evaluate:
$$c \int_{0}^{2} y^2 \, dy = c \left[\frac{y^3}{3}\right]_0^2 = c \cdot \frac{8}{3} = 1$$
:::

::: {.column width="50%" .fragment}
**Step 3:** Solve for $c$:
$$c = \frac{3}{8}$$

**Result:** The valid PDF is $f(y) = \frac{3}{8}y^2$ for $0 \leq y \leq 2$.

:::
:::

## 📌 Example 6: Computing Interval Probabilities {.smaller}

**Problem:** Given $Y$ with density $f(y) = \frac{3}{8}y^2$ for $0 \leq y \leq 2$, find $P(1 \leq Y \leq 2)$ and $P(1 < Y < 2)$.

. . .

**Solution:**

::: {.columns}

::: {.column width="33%" .fragment}

**Step 1:** Compute the integral:
$$P(1 \leq Y \leq 2) = \int_{1}^{2} \frac{3}{8}y^2 \, dy $$
$$= \frac{3}{8} \int_{1}^{2} y^2 \, dy$$

:::

::: {.column width = "33%" .fragment}

**Step 2:** Evaluate:
$$\frac{3}{8} \left[\frac{y^3}{3}\right]_1^2 = \frac{3}{8} \left(\frac{8}{3} - \frac{1}{3}\right)$$
$$ = \frac{3}{8} \cdot \frac{7}{3} = \frac{7}{8}$$

:::

::: {.column width="33%" .fragment}

**Step 3:** For continuous $Y$, $P(Y = 1) = P(Y = 2) = 0$, so:
$$P(1 < Y < 2) = P(1 \leq Y \leq 2) = \frac{7}{8}$$

**Interpretation:** There is an 87.5% probability that $Y$ falls between 1 and 2.

:::
:::

## 💰 Case Study: Stock Returns (Real Data) {.smaller}

::: {style="font-size:28px"}
::: {.columns}
::: {.column width="50%"}
::: {.callout-note}
## 📈 Stock Return Analysis

**Context:** Modeling daily log-returns of S&P 500 index as a continuous random variable.

**Key Questions:**

- Can we model stock returns as continuous distributions?

- What is the empirical CDF and how does it compare to theoretical distributions?

- What are key quantiles (VaR measures) for risk management?
:::
:::

::: {.column width="50%" .fragment}
::: {.callout-tip}
## 📊 Data Source

We analyze daily S&P 500 log-returns from **January 2020 to October 2024**.

**Source:** Yahoo Finance API (via quantmod)

**Period:** 2020-01-01 to 2024-10-31

**Data Quality:** Adjusted closing prices

**Verification:** Cross-checked with multiple financial data providers
:::
:::
:::
:::

## 💰 Case Study: Data Loading and Processing {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="30%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

library(quantmod)
library(tidyverse)
library(lubridate)

# Load S&P 500 data
getSymbols("^GSPC", from = "2020-01-01", 
           to = "2024-10-31", auto.assign = TRUE)

# Compute log-returns
sp500 <- data.frame(
  Date = index(GSPC),
  Close = as.numeric(GSPC$GSPC.Adjusted)
) %>%
  mutate(LogReturn = c(NA, diff(log(Close))))

# Remove NA
sp500_clean <- sp500 %>% filter(!is.na(LogReturn))

cat("Number of observations:", nrow(sp500_clean), "
")
cat("Date range:", min(sp500_clean$Date), 
    "to", max(sp500_clean$Date), "
")
```

:::

::: {.column width="70%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Summary statistics
summary_stats <- sp500_clean %>%
  summarise(
    Mean = mean(LogReturn),
    StdDev = sd(LogReturn),
    Median = median(LogReturn),
    Min = min(LogReturn),
    Max = max(LogReturn),
    Q5 = quantile(LogReturn, 0.05),
    Q95 = quantile(LogReturn, 0.95)
  )

knitr::kable(summary_stats, digits = 5, 
             caption = "Summary Statistics of S&P 500 Log-Returns")
```

**Key insights:**

- Mean return is slightly positive

- Standard deviation shows volatility

- 5th percentile (VaR) indicates risk
:::
:::
:::

## 💰 Case Study: Empirical CDF and Density {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 4.5

# Empirical CDF
ggplot(sp500_clean, aes(x = LogReturn)) +
  stat_ecdf(geom = "step", color = "steelblue", 
            size = 1.2) +
  labs(title = "Empirical CDF of S&P 500 Log-Returns",
       subtitle = "January 2020 - October 2024",
       x = "Log-Return",
       y = "Cumulative Probability F(y)") +
  theme_minimal(base_size = 13) +
  geom_hline(yintercept = c(0.05, 0.5, 0.95), 
             linetype = "dashed", color = "red", alpha = 0.5)
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 4.5

# Empirical PDF (histogram + density)
ggplot(sp500_clean, aes(x = LogReturn)) +
  geom_histogram(aes(y = ..density..), 
                 bins = 50, fill = "lightblue", 
                 color = "black", alpha = 0.6) +
  geom_density(color = "darkblue", size = 1.2) +
  labs(title = "Empirical PDF of S&P 500 Log-Returns",
       subtitle = "Histogram and Kernel Density Estimate",
       x = "Log-Return",
       y = "Density f(y)") +
  theme_minimal(base_size = 13)
```

:::
:::
:::

## 💰 Case Study: Quantile Analysis {.smaller}

::: {style="font-size:28px"}

::: {.columns}

::: {.column width="50%"}
```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Compute key quantiles
quantiles <- sp500_clean %>%
  summarise(
    `1% (VaR 99%)` = quantile(LogReturn, 0.01),
    `5% (VaR 95%)` = quantile(LogReturn, 0.05),
    `25% (Q1)` = quantile(LogReturn, 0.25),
    `50% (Median)` = quantile(LogReturn, 0.50),
    `75% (Q3)` = quantile(LogReturn, 0.75),
    `95%` = quantile(LogReturn, 0.95),
    `99%` = quantile(LogReturn, 0.99)
  )

knitr::kable(t(quantiles), digits = 5, 
             col.names = c("Log-Return Value"),
             caption = "Key Quantiles of S&P 500 Log-Returns")
```

:::

::: {.column width="50%"}
**Financial interpretation:**

- **VaR 95%** (5th percentile): There is a 5% chance daily losses exceed this value (risk measure for portfolio management)

- **Median**: The middle value of the distribution (50% probability above/below)

- **VaR 99%** (1st percentile): Extreme loss threshold for stress testing
:::
:::
:::

## 📝 Quiz #1: Distribution Functions {.quiz-question}

What is the value of the distribution function $F(y)$ as $y$ approaches positive infinity?

- [1]{.correct data-explanation="✅ Correct! As y approaches infinity, the cumulative probability includes all outcomes, so F(infinity) = 1."}
- 0
- 0.5
- Undefined

## 📝 Quiz #2: Continuous Random Variables {.quiz-question}

For a continuous random variable Y, what is the probability that Y equals exactly 5?

- [0]{.correct data-explanation="✅ Correct! For continuous variables, the probability of any single exact value is zero."}
- 0.5
- Depends on the distribution
- Cannot be determined

## 📝 Quiz #3: PDF Properties {.quiz-question}

Which of the following must be true for a valid probability density function f(y)?

- [The integral of f(y) over all y equals 1]{.correct data-explanation="✅ Correct! The total area under the PDF curve must equal 1 (total probability)."}
- f(y) can be negative for some values of y
- f(y) must be continuous everywhere
- The maximum value of f(y) cannot exceed 1

## 📝 Summary {.smaller}

::: {.summary-box}
**✅ Key Takeaways**

- **Continuous random variables** can take any value in an interval; individual point probabilities are zero

- **Distribution function** $F(y) = P(Y \leq y)$ is cumulative and non-decreasing, with limits 0 and 1 at negative and positive infinity

- **Probability density function** $f(y) = F'(y)$ satisfies $f(y) \geq 0$ and $\int f(y) dy = 1$; probabilities are found by integrating $f(y)$ over intervals

- **Quantiles and percentiles** (including the median) provide important summary measures for risk assessment in financial applications

- **Real financial data** (such as S&P 500 returns) can be modeled using continuous distributions to compute risk measures like Value-at-Risk
:::

## 📚 Practice Problems

::: {.callout-tip}
## 📝 Homework Problems

**Problem 1:** Given $f(y) = 2y$ for $0 \leq y \leq 1$ and zero elsewhere, verify this is a valid PDF and find $F(y)$. Then compute $P(0.25 \leq Y \leq 0.75)$.

**Problem 2:** For the PDF $f(y) = ke^{-y}$ for $y > 0$ and zero elsewhere, find the constant $k$ that makes this a valid density, and determine the median.

**Problem 3:** A stock's daily log-return is modeled by $f(y) = \frac{1}{2}$ for $-0.02 \leq y \leq 0.02$. Find the probability that the return is between -0.01 and 0.01, and find the 5th percentile (VaR 95%).

**Problem 4:** Using the real S&P 500 data from the case study, compute the interquartile range (IQR) and interpret it in the context of financial risk.
:::

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov

Assistant Professor

School of Business

ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

Topic: Expected Value and Variance of Continuous Random Variables

Reading: Review properties of integrals and moments

Preparation: Refresh calculus (integration by parts)

**⏰ Reminders:**

✅ Complete practice problems

✅ Review quantile computations

✅ Work hard and stay curious!
:::
:::

## ❓ Questions? {.center}

::: {.callout-note}
## 💬 Open Discussion (5 minutes)

**Discussion Point 1:** How do continuous distributions differ from discrete distributions in practical financial modeling?

**Discussion Point 2:** Why is the median sometimes preferred over the mean for skewed distributions (e.g., stock returns during crises)?

**Discussion Point 3:** How can quantiles be used to set risk limits in portfolio management?

**Discussion Point 4:** What are the limitations of using historical data to estimate distribution functions?
:::
