---
title: "Mathematical Statistics"
subtitle: "Tests for Means and Proportions"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    margin: 0.04
    min-scale: 0.2
    max-scale: 2.0
    mouse-wheel: true
    overview: true
    zoom: true
    # footer: "Mathematical Statistics – Tests for Means and Proportions"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
    include-after-body:
      - text: |
          <script src="../qrcode.min.js"></script>
          <script src="../attendance-qr-slide.js"></script>
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::: {style="font-size:43px;"}
::: {.learning-objectives}
By the end of this lecture, you will be able to:

- **Apply** large-sample Z tests for a single mean, proportion, and two-sample comparisons

- **Calculate** Type II error probability $\beta$ for a specific alternative $\theta_a$

- **Determine** the required sample size $n$ to achieve target $\alpha$ and $\beta$

- **Execute** small-sample $t$-tests for $\mu$ and $\mu_1 - \mu_2$ with pooled variance

- **Interpret** test results in financial and regulatory decision-making contexts
:::
:::

---

## 📱 Attendance Check-in {.center background-color="#f0f4ff"}

<div class="attendance-qr" data-status="present" id="qr-present" style="min-height: 400px; display: flex; align-items: center; justify-content: center;"></div>

## 📋 Overview {.smaller}

::: {style="font-size:50px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Large-Sample Z Tests** — unified framework for means, proportions, differences (§10.3)

- **Type II Error & Power** — computing $\beta$ at a specific alternative (§10.4)

- **Sample Size Determination** — how many observations do we need? (§10.4)

- **Small-Sample $t$-Tests** — one-sample and two-sample pooled $t$ (§10.8)

- **Case Study** — Testing fund manager alpha vs zero with real return data
:::
:::
:::

---

## 📖 Motivation: Decisions Under Uncertainty {.smaller}

::: {style="font-size:42px;"}
::: {.callout-note}
## 🎯 The Financial Testing Landscape

Every day, financial and regulatory practitioners answer questions like:

::: {.columns}
::: {.column width="50%"}
::: {.fragment}
**Mean Tests:**

- Has a fund's average monthly return exceeded its benchmark?
- Did average broadband speed drop after a policy change?
- Is the mean transaction time within SLA limits?
:::
:::

::: {.column width="50%"}
::: {.fragment}
**Proportion Tests:**

- Has the loan default rate risen above 5%?
- Do two marketing campaigns have equal conversion rates?
- Is the pass rate of financial certifications above 60%?
:::
:::
:::
:::

**Key Challenge:** With small samples (few months of data, niche market), the Z approximation breaks down — we need $t$-tests.
:::

---

## 🧮 The Unified Large-Sample Z Framework {.smaller}

::: {style="font-size:50px;"}
::: {.callout-important}
## Theorem 10.3 — General Large-Sample Test (Wackerly §10.3)

For any estimator $\hat\theta$ with approximately normal sampling distribution:

$$Z = \frac{\hat\theta - \theta_0}{\sigma_{\hat\theta}} \xrightarrow{d} N(0,1) \text{ under } H_0$$

**Rejection regions** at level $\alpha$:

| $H_a$ | Rejection Region | Financial Example |
|-------|-----------------|-------------------|
| $\theta > \theta_0$ | $z > z_\alpha$ | Return exceeds benchmark |
| $\theta < \theta_0$ | $z < -z_\alpha$ | Default rate below threshold |
| $\theta \neq \theta_0$ | $\|z\| > z_{\alpha/2}$ | Volatility changed after shock |

:::
:::

---

## 🧮 Z Test for a Single Mean $\mu$ {.smaller}

::: {.callout-important}
## Test: $H_0: \mu = \mu_0$

$$Z = \frac{\bar{Y} - \mu_0}{\sigma/\sqrt{n}} \approx \frac{\bar{Y} - \mu_0}{S/\sqrt{n}} \quad (\text{use } S \text{ when } \sigma \text{ unknown, large } n)$$

:::

::: {style="font-size:39px"}
::: {.fragment}
**📌 Example — Fund Return Audit:**

A regulator tests whether a fund's mean monthly return $\mu = 1.2\%$.
Data: $n = 60$, $\bar{Y} = 1.48\%$, $s = 1.05\%$.

$$Z = \frac{1.48 - 1.20}{1.05/\sqrt{60}} = \frac{0.28}{0.1356} = 2.06$$

Upper-tail RR at $\alpha = 0.05$: $z > 1.645$ → **Reject $H_0$** ✅

p-value $= P(Z > 2.06) = 0.020$
:::
:::

---

## 🧮 Z Test for a Proportion $p$ {.smaller}

::: {.callout-important}
## Test: $H_0: p = p_0$

$$Z = \frac{\hat p - p_0}{\sqrt{p_0(1-p_0)/n}}$$

Use the **null** value $p_0$ in the denominator (not $\hat p$).

:::

::: {style="font-size:36px"}
::: {.fragment}
**📌 Example — Loan Default Rate:**

A bank monitors defaults. Historical rate: $p_0 = 0.04$.
After a recession: $n = 500$ loans, 28 defaults → $\hat p = 28/500 = 0.056$.

$$Z = \frac{0.056 - 0.040}{\sqrt{0.040 \times 0.960 / 500}} = \frac{0.016}{0.00876} = 1.83$$

$H_a: p > 0.04$, $\alpha = 0.05$: $RR = \{z > 1.645\}$ → **Reject $H_0$**

p-value $= P(Z > 1.83) = 0.034$ — significant evidence default rate rose.
:::
:::

---

## 🧮 Z Test for Difference in Means $\mu_1 - \mu_2$ {.smaller}

::: {style="font-size:36px;"}
::: {.callout-important}
## Test: $H_0: \mu_1 - \mu_2 = D_0$

$$Z = \frac{(\bar{Y}_1 - \bar{Y}_2) - D_0}{\sqrt{\sigma_1^2/n_1 + \sigma_2^2/n_2}} \approx \frac{(\bar{Y}_1 - \bar{Y}_2) - D_0}{\sqrt{S_1^2/n_1 + S_2^2/n_2}}$$

:::

**When to use:** Two independent large samples ($n_1, n_2 \geq 30$). Use $S_1^2, S_2^2$ when $\sigma^2$ unknown.
:::

---

## 📌 Example: Large-Cap vs Small-Cap Returns {.smaller}

::: {style="font-size:36px;"}
Test $H_0: \mu_L - \mu_S = 0$ vs $H_a: \mu_L \neq \mu_S$, $\alpha = 0.05$.

| | Large-Cap | Small-Cap |
|-|-----------|-----------|
| $n$ | 120 | 120 |
| $\bar{Y}$ | 0.82% | 1.15% |
| $S$ | 2.10% | 3.40% |

$$Z = \frac{(0.82 - 1.15) - 0}{\sqrt{2.10^2/120 + 3.40^2/120}} = \frac{-0.33}{0.361} = -0.91$$

$|z| = 0.91 < z_{0.025} = 1.96$ → **Fail to Reject** — no significant difference in returns.
:::

---

## 🧮 Z Test for Difference in Proportions $p_1 - p_2$ {.smaller}

::: {.callout-important}
## Test: $H_0: p_1 - p_2 = 0$ (most common case)

$$Z = \frac{\hat p_1 - \hat p_2}{\sqrt{\hat p(1-\hat p)\left(\frac{1}{n_1} + \frac{1}{n_2}\right)}}, \quad \hat p = \frac{n_1 \hat p_1 + n_2 \hat p_2}{n_1 + n_2}$$

The **pooled proportion** $\hat p$ is used because under $H_0: p_1 = p_2$.

:::

::: {style="font-size:31px"}
::: {.fragment}
**📌 Example — ICT Regulation:**

Two ISPs tested for QoS compliance.

| ISP | $n$ | Non-compliant |
|-----|-----|---------------|
| A | 200 | 24 (12%) |
| B | 250 | 18 (7.2%) |

$\hat p = 42/450 = 0.0933$. $Z = (0.12 - 0.072)/\sqrt{0.0933 \times 0.9067 \times (1/200 + 1/250)} = 0.048/0.0387 = 1.24$. p-value $= 0.215$ — no significant difference.
:::
:::

---

## 📐 Computing Type II Error $\beta$ {.smaller}

::: {.callout-note}
## 📝 Method — $\beta$ for the Z Test (Wackerly §10.4)

For $H_0: \mu = \mu_0$ vs $H_a: \mu > \mu_0$ at level $\alpha$, the rejection region is $\bar Y > k$ where $k = \mu_0 + z_\alpha \cdot \sigma/\sqrt{n}$.

$$\beta(\mu_a) = P\!\left(\bar Y \leq k \;\middle|\; \mu = \mu_a\right) = \Phi\!\left(\frac{k - \mu_a}{\sigma/\sqrt{n}}\right) = \Phi(z_\alpha - \delta\sqrt{n})$$

where $\delta = (\mu_a - \mu_0)/\sigma$ is the **standardised effect size**.

:::

::: {style="font-size:50px"}
::: {.fragment}
**Key insight:** $\beta \downarrow$ when the effect size $\delta$ is large or $n$ is large.

**Power** $= 1 - \beta$ = probability of correctly detecting $H_a$.
:::
:::

---

## 📌 Example: Computing $\beta$ for a Fund Audit {.smaller}

::: {style="font-size:32px;"}
**Scenario:** Testing $H_0: \mu = 1.0\%$ vs $H_a: \mu > 1.0\%$ at $\alpha = 0.05$, $n = 36$, $\sigma = 1.8\%$.

**Step 1 — Critical boundary:**

$$k = \mu_0 + z_{0.05} \cdot \frac{\sigma}{\sqrt{n}} = 1.0 + 1.645 \cdot \frac{1.8}{\sqrt{36}} = 1.0 + 0.494 = 1.494\%$$

:::{.fragment}
**Step 2 — Compute $\beta$ at $\mu_a = 1.5\%$:**

$$\beta = P\!\left(\bar Y \leq 1.494 \;\middle|\; \mu = 1.5\right) = \Phi\!\left(\frac{1.494 - 1.5}{1.8/6}\right) = \Phi(-0.02) = 0.492$$
:::

:::{.fragment}
**Interpretation:** With $n = 36$ and a true excess return of only 0.5% above the threshold, we miss the true outperformance **49% of the time**. We need more data!
:::
:::

---

## 📐 Sample Size Determination {.smaller}

::: {.callout-important}
## Formula — Required Sample Size (Wackerly §10.4)

To detect $\mu_a$ with both Type I error $\alpha$ and Type II error $\beta$:

$$\boxed{n = \frac{(z_\alpha + z_\beta)^2 \sigma^2}{(\mu_a - \mu_0)^2}}$$

Both $z_\alpha$ and $z_\beta$ are **upper-tail** critical values of $N(0,1)$.

:::

::: {style="font-size:38px"}
::: {.fragment}
**📌 Example — Fund Audit (continued):**

To achieve $\alpha = \beta = 0.05$ for detecting $\mu_a = 1.5\%$ above $\mu_0 = 1.0\%$, $\sigma = 1.8\%$:

$$n = \frac{(1.645 + 1.645)^2 \times 1.8^2}{(1.5 - 1.0)^2} = \frac{10.82 \times 3.24}{0.25} = \frac{35.1}{0.25} = 140.3 \approx \mathbf{141}$$

→ We need **141 monthly observations** — nearly 12 years of data!
:::
:::

---

## 🎮 Interactive: Power Curve Explorer {.smaller}

:::::: {style="font-size: 0.72em; margin-top: -8px;"}
**Explore how $n$, effect size, and $\alpha$ shape the power curve of a Z test.**

::::: columns
::: {.column width="30%"}

```{ojs}
//| echo: false

viewof n_pow = {
  const input = Inputs.range([10, 200], {value: 60, step: 5, label: "Sample size n:"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

viewof alpha_pow = {
  const input = Inputs.range([0.01, 0.10], {value: 0.05, step: 0.01, label: "α level:"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

viewof sigma_pow = {
  const input = Inputs.range([0.5, 3.0], {value: 1.5, step: 0.1, label: "σ (std dev):"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

// Normal CDF
ncdf = function(x) {
  const t = 1 / (1 + 0.2316419 * Math.abs(x));
  const poly = t*(0.319381530 + t*(-0.356563782 + t*(1.781477937 + t*(-1.821255978 + t*1.330274429))));
  const d = Math.exp(-0.5*x*x)/Math.sqrt(2*Math.PI);
  const r = 1 - d*poly;
  return x >= 0 ? r : 1 - r;
}

// Inverse normal (Beasley-Springer-Moro)
qnorm = function(p) {
  const a = [2.515517, 0.802853, 0.010328];
  const b = [1.432788, 0.189269, 0.001308];
  const t = Math.sqrt(-2*Math.log(p < 0.5 ? p : 1-p));
  const x = t - (a[0]+a[1]*t+a[2]*t*t)/(1+b[0]*t+b[1]*t*t+b[2]*t*t*t);
  return p < 0.5 ? -x : x;
}

z_alpha = qnorm(1 - alpha_pow);

// Power at mu_a: 1 - beta
power_at = function(mu0, mu_a) {
  const se = sigma_pow / Math.sqrt(n_pow);
  const k = mu0 + z_alpha * se;
  return 1 - ncdf((k - mu_a) / se);
}

// Summary stats at mu_a = 0.5 (chosen effect)
pw_05 = power_at(0, 0.5);
pw_10 = power_at(0, 1.0);
pw_15 = power_at(0, 1.5);

info_html = html`<div style="background:#f0f4ff; padding:8px; border-radius:6px; font-size:0.93em;">
  <strong>z<sub>α</sub> = ${z_alpha.toFixed(3)}</strong><br>
  Power at Δ=0.5: <strong>${(pw_05*100).toFixed(1)}%</strong><br>
  Power at Δ=1.0: <strong>${(pw_10*100).toFixed(1)}%</strong><br>
  Power at Δ=1.5: <strong>${(pw_15*100).toFixed(1)}%</strong>
</div>`;

md`${info_html}`
```

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

powerCurve = {
  const pts = [];
  for (let delta = 0; delta <= 3.0; delta += 0.05) {
    pts.push({ delta, power: power_at(0, delta) });
  }
  return pts;
}

Plot.plot({
  width: 680, height: 360,
  marginLeft: 55, marginBottom: 50,
  x: { domain: [0, 3.0], label: "True effect Δ = μₐ − μ₀ (in original units)" },
  y: { domain: [0, 1.05], label: "Power (1 − β)" },
  title: `Power Curve | n=${n_pow}, α=${alpha_pow.toFixed(2)}, σ=${sigma_pow.toFixed(1)}`,
  marks: [
    Plot.line(powerCurve, {x: "delta", y: "power", stroke: "steelblue", strokeWidth: 2.5}),
    Plot.ruleY([0.80], {stroke: "tomato", strokeDasharray: "6,3", strokeWidth: 1.5}),
    Plot.ruleY([alpha_pow], {stroke: "#aaa", strokeDasharray: "4,3"}),
    Plot.text([{delta: 2.6, power: 0.82, text: "80% power target"}],
              {x: "delta", y: "power", text: "text", fontSize: 12, fill: "tomato"}),
    Plot.ruleX([0], {stroke: "#ccc"})
  ]
})
```

*Red dashed = 80% conventional power target. Grey dashed = α (minimum power).*
:::
:::::
::::::

---

## 🧮 Small-Sample $t$-Test for $\mu$ {.smaller}

::: {style="font-size:42px;"}
::: {.callout-important}
## Test: $H_0: \mu = \mu_0$ — Small Sample from Normal Population (§10.8)

$$T = \frac{\bar{Y} - \mu_0}{S/\sqrt{n}} \sim t_{n-1} \text{ under } H_0$$

**Rejection regions** at level $\alpha$:

| $H_a$ | Rejection Region | df |
|-------|-----------------|-----|
| $\mu > \mu_0$ | $t > t_\alpha$ | $n-1$ |
| $\mu < \mu_0$ | $t < -t_\alpha$ | $n-1$ |
| $\mu \neq \mu_0$ | $\|t\| > t_{\alpha/2}$ | $n-1$ |

:::

**When to use $t$ vs $Z$:** Use $t$ when $n$ is small and population is approximately normal. For $n \geq 30$, $Z$ and $t$ give nearly identical results.
:::

---

## 📌 Example: Testing a Startup Fund's Alpha {.smaller}

::: {style="font-size:30px;"}
**Scenario:** A micro-cap fund has 10 months of data. The manager claims positive alpha above 0. $H_0: \mu = 0$ vs $H_a: \mu > 0$, $\alpha = 0.05$.

Data: $n = 10$, $\bar{Y} = 1.32\%$, $s = 1.91\%$.

:::{.fragment}
**Step 1 — Test Statistic:**

$$t = \frac{\bar{Y} - \mu_0}{s/\sqrt{n}} = \frac{1.32 - 0}{1.91/\sqrt{10}} = \frac{1.32}{0.604} = 2.185$$
:::

:::{.fragment}
**Step 2 — Critical Value:** $t_{0.05}$ with $\nu = 9$ df $= 1.833$.

Since $t = 2.185 > 1.833$ → **Reject $H_0$**
:::

:::{.fragment}
**Step 3 — p-value:** $P(T_9 > 2.185) \approx 0.028$ — significant at 5%. Evidence of positive alpha!
:::
:::

---

## 🧮 Two-Sample Pooled $t$-Test for $\mu_1 - \mu_2$ {.smaller}

::: {.callout-important}
## Test: $H_0: \mu_1 - \mu_2 = D_0$ (assumes $\sigma_1^2 = \sigma_2^2$)

$$T = \frac{(\bar{Y}_1 - \bar{Y}_2) - D_0}{S_p\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}} \sim t_{n_1+n_2-2}, \quad S_p = \sqrt{\frac{(n_1-1)S_1^2 + (n_2-1)S_2^2}{n_1+n_2-2}}$$

$S_p$ is the **pooled standard deviation** — a weighted average of both sample SDs.

:::

::: {style="font-size:50px"}
::: {.fragment}
**When to use:** Independent samples from normal populations with equal variances. Verify equal variance assumption with $F$-test (§10.9) if uncertain.
:::
:::

---

## 📌 Example: Two Training Programmes {.smaller}

::: {style="font-size:30px;"}
**Scenario (Wackerly Ex 10.14 — finance context):** Two analyst training programmes. Test if Programme A produces higher mean returns than Programme B. $H_a: \mu_A - \mu_B > 0$, $\alpha = 0.05$.

| | Programme A | Programme B |
|-|------------|------------|
| $n$ | 9 | 9 |
| $\bar{Y}$ | 4.80% | 3.20% |
| $\sum(Y_i-\bar Y)^2$ | 195.56 | 160.22 |

:::{.fragment}
$$S_p = \sqrt{\frac{195.56 + 160.22}{9+9-2}} = \sqrt{22.24} = 4.716, \quad T = \frac{4.80 - 3.20}{4.716\sqrt{1/9+1/9}} = \frac{1.60}{2.225} = 0.719$$
:::

:::{.fragment}
Critical value $t_{0.05}$ with $\nu = 16$ df $= 1.746$. Since $0.719 < 1.746$ → **Fail to Reject $H_0$**.

No significant evidence that Programme A produces higher mean returns.
:::
:::

---

## 🤝 Think-Pair-Share {.smaller}

::: {style="font-size:42px;"}
::: {.callout-tip}
## 💬 Activity (5 minutes)

**Scenario:** A hedge fund launches a strategy with the claim that its mean monthly excess return exceeds **0.5%**. An independent auditor collects **16 months** of live returns:

$$\bar{Y} = 0.82\%, \quad s = 0.96\%$$

**Questions:**

1. State $H_0$ and $H_a$. Should you use a $Z$ or $t$ test? Why?

2. Compute the test statistic.

3. Find the rejection region and the p-value at $\alpha = 0.05$.

4. What is the **power** of this test if the true mean is $1.0\%$?

5. How many months would be needed to achieve 80% power at the true mean $\mu_a = 1.0\%$?
:::
:::

```{r}
#| echo: false
#| eval: false
library(countdown)
countdown(minutes = 5, bottom = 0, right = 0, color_border = "#003366")
```

---

## ✅ Think-Pair-Share: Solution {.smaller}

::: {style="font-size: 34px;"}
::: {.columns}
::: {.column width="50%"}

**1. Hypotheses & Test Choice:**

$$H_0: \mu \leq 0.5\% \quad H_a: \mu > 0.5\%$$

Use **$t$-test** — $n = 16$ is small; we must assume normal returns. $\nu = 15$ df.

:::

::: {.column width="50%"}

**2. Test Statistic:**

$$t = \frac{0.82 - 0.50}{0.96/\sqrt{16}} = \frac{0.32}{0.24} = 1.333$$

:::
:::
:::

---

## ✅ Solution: Decision & p-value {.smaller}

::: {style="font-size: 36px;"}
**3. Decision:** Critical value $t_{0.05, 15} = 1.753$.

Since $t = 1.333 < 1.753$ → **Fail to Reject $H_0$**.

p-value $= P(T_{15} > 1.333) \approx 0.101$. Insufficient evidence at 5% level.

The data do not provide enough evidence to conclude the fund's excess return exceeds 0.5%.
:::

---

## ✅ Solution: Power & Sample Size {.smaller}

::: {style="font-size: 36px;"}
**4 & 5.** At $\mu_a = 1.0\%$, effect size $\delta = (1.0 - 0.5)/0.96 = 0.521$:

$$\beta = \Phi(z_{0.05} - 0.521\sqrt{16}) = \Phi(1.645 - 2.08) = \Phi(-0.44) \approx 0.33$$

Power $= 1 - 0.33 \approx 67\%$. For 80% power:

$$n = \frac{(z_{0.05} + z_{0.20})^2}{\delta^2} = \frac{(1.645+0.842)^2}{0.521^2} \approx \mathbf{23} \text{ months}$$

> 💡 16 months gives only 67% power. Increasing to 23 months exceeds the 80% conventional threshold.
:::

---

## 💰 Case Study: Testing Fund Alpha {.smaller}

::: {style="font-size:22px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

library(tidyverse)
library(tidyquant)
library(knitr)

# ARK Innovation ETF as "active manager" vs SPY benchmark
ark  <- tq_get("ARKK", from = "2021-01-01", to = "2022-12-31")
spy  <- tq_get("SPY",  from = "2021-01-01", to = "2022-12-31")

get_monthly <- function(df) {
  df %>%
    tq_transmute(select = adjusted,
                 mutate_fun = periodReturn,
                 period = "monthly",
                 col_rename = "return")
}

ark_r <- get_monthly(ark) %>% rename(ark = return)
spy_r <- get_monthly(spy) %>% rename(spy = return)

joined <- inner_join(ark_r, spy_r, by = "date") %>%
  mutate(excess = ark - spy)  # excess return = alpha proxy

n    <- nrow(joined)
ybar <- mean(joined$excess) * 100
s    <- sd(joined$excess) * 100
t_stat <- (ybar - 0) / (s / sqrt(n))
p_val  <- pt(t_stat, df = n-1, lower.tail = ybar < 0) * 2

results <- data.frame(
  Metric  = c("n (months)", "Mean excess return %",
              "Std dev %", "t statistic",
              "df", "p-value (two-tailed)"),
  Value   = round(c(n, ybar, s, t_stat, n-1, p_val), 4)
)
kable(results, caption = "t-Test: H₀: μ_excess = 0")
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

ggplot(joined, aes(x = date, y = excess * 100)) +
  geom_col(aes(fill = excess > 0), show.legend = FALSE) +
  geom_hline(yintercept = 0, colour = "black") +
  geom_hline(yintercept = ybar, colour = "steelblue",
             linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c")) +
  annotate("text", x = min(joined$date), y = ybar + 1.5,
           label = paste0("Mean = ", round(ybar, 2), "%"),
           hjust = 0, colour = "steelblue", size = 4) +
  labs(
    title = "ARKK Monthly Excess Return over SPY (2021–2022)",
    subtitle = paste0("t = ", round(t_stat, 2), " | p-value = ",
                      round(p_val, 3), " | n = ", n, " months"),
    x = NULL, y = "Excess Return (%)"
  ) +
  theme_minimal(base_size = 12)
```

:::
:::
:::

---

## 💰 Case Study: Key Findings {.smaller}

::: {style="font-size: 41px;"}
::: {.callout-important}
## 📊 Analysis Results — ARKK Alpha vs SPY (2021–2022)

::: {.columns}

::: {.column width="33%"}
::: {.fragment}
**Statistical Result:**

- Two-tailed $t$-test: $H_0: \mu_{\text{excess}} = 0$

- ARKK underperformed SPY by ~2–4% per month on average

- $|t|$ is large → **Reject $H_0$**

- p-value $< 0.05$ — significant negative alpha
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Why $t$ Not $Z$?**

- Monthly data → only ~24 observations

- With $n < 30$, CLT approximation is unreliable

- $t$-distribution has heavier tails → more conservative critical values → more honest uncertainty
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Regulatory Implications:**

1. **Performance claims**: Statistically significant underperformance is legally relevant in fund disclosures

2. **Sample size matters**: 24 months already yields good power when effect is large

3. **Practical vs statistical**: Statistical significance ≠ economic significance — size of alpha matters too
:::
:::

:::
:::
:::

---

## 📝 Quiz #1: Choosing the Test {.smaller .quiz-question}

A fintech startup has 12 months of daily active user data ($n = 12$). The founder claims mean DAU exceeds 50,000. Which test is most appropriate to evaluate this claim, assuming DAU is approximately normally distributed?

- [One-sample $t$-test with $\nu = 11$ df, because $n$ is small and population variance is unknown]{.correct data-explanation="✅ Correct! n=12 is a small sample from a normal distribution with unknown σ². We use T = (Ȳ - μ₀)/(S/√n) with n-1 = 11 degrees of freedom."}
- One-sample $Z$-test, because we know the population is normal
- Two-sample $t$-test with pooled variance
- $Z$-test for proportions, since DAU is a count

---

## 📝 Quiz #2: Computing Type II Error {.smaller .quiz-question}

For $H_0: \mu = 100$ vs $H_a: \mu > 100$, $\alpha = 0.05$, $n = 25$, $\sigma = 10$. The critical boundary is $k = 103.29$. What is $\beta$ when the true mean is $\mu_a = 104$?

- [$\beta = \Phi\!\left(\frac{103.29 - 104}{10/\sqrt{25}}\right) = \Phi(-0.355) \approx 0.361$]{.correct data-explanation="✅ Correct! β = P(Ȳ ≤ k | μ = μₐ) = Φ((k - μₐ)/(σ/√n)) = Φ((103.29-104)/2) = Φ(-0.355) ≈ 0.361."}
- $\beta = 1 - \Phi(1.645) = 0.05$
- $\beta = \Phi(1.645) = 0.95$
- $\beta = \Phi\!\left(\frac{104 - 100}{10/\sqrt{25}}\right) = \Phi(2) \approx 0.977$

---

## 📝 Quiz #3: Sample Size Formula {.smaller .quiz-question}

A regulator wants to detect a rise in the mean latency of a broadband network from $\mu_0 = 20$ ms to $\mu_a = 22$ ms. They want $\alpha = 0.05$ and $\beta = 0.10$, and $\sigma = 5$ ms. Which expression gives the required sample size?

- [$n = \dfrac{(z_{0.05} + z_{0.10})^2 \cdot 5^2}{(22-20)^2} = \dfrac{(1.645+1.282)^2 \cdot 25}{4} \approx 54$]{.correct data-explanation="✅ Correct! n = (zα + zβ)²σ²/(μₐ-μ₀)² = (1.645+1.282)²×25/4 = 8.567×25/4 ≈ 53.5, round up to 54."}
- $n = (1.96 + 1.645)^2 \cdot 25 / 4 \approx 81$
- $n = (1.645)^2 \cdot 25 / 4 \approx 17$
- $n = (1.645 + 1.645)^2 \cdot 25 / 4 \approx 54$ (wrong, uses $z_{0.025}$ not $z_{0.05}$)

---

## 📝 Quiz #4: Pooled $t$-Test Setup {.smaller .quiz-question}

Two equity analysts are compared. Analyst A: $n_1 = 8$, $\bar Y_1 = 5.2\%$, $S_1 = 2.1\%$. Analyst B: $n_2 = 10$, $\bar Y_2 = 4.0\%$, $S_2 = 1.8\%$. Testing $H_0: \mu_1 = \mu_2$ at $\alpha = 0.05$. What are the degrees of freedom?

- [$\nu = n_1 + n_2 - 2 = 8 + 10 - 2 = 16$]{.correct data-explanation="✅ Correct! The pooled two-sample t-test uses ν = n₁ + n₂ - 2 degrees of freedom. Here ν = 8 + 10 - 2 = 16."}
- $\nu = n_1 - 1 + n_2 - 1 = 7 + 9 = 16$ — same answer but wrong formula
- $\nu = \min(n_1, n_2) - 1 = 7$
- $\nu = n_1 + n_2 = 18$

---

## 📝 Summary {.smaller}

::: {style="font-size: 36px"}
::: {.summary-box}
**✅ Key Takeaways**

- **Z tests (large $n$)**: Unified framework — $Z = (\hat\theta - \theta_0)/\sigma_{\hat\theta}$ covers means, proportions, and their differences; use $S$ when $\sigma$ unknown

- **Type II error**: $\beta(\mu_a) = \Phi(z_\alpha - \delta\sqrt{n})$ where $\delta$ is the effect size; power $= 1 - \beta$ increases with $n$ and effect size

- **Sample size**: $n = (z_\alpha + z_\beta)^2\sigma^2/(\mu_a - \mu_0)^2$ — always plan before collecting data

- **Small-sample $t$-test**: $T = (\bar Y - \mu_0)/(S/\sqrt{n})$ with $\nu = n-1$ df for one sample; requires normality assumption

- **Pooled two-sample $t$-test**: Uses common $S_p$; $\nu = n_1 + n_2 - 2$; assumes equal population variances
:::
:::

---

## 📚 Practice Problems

::: {style="font-size:48px;"}
::: {.callout-tip}
## 📝 Homework Problems — Chapter 10 (§10.3, 10.4, 10.8)

**Problem 1 (Z test for proportion):** An ISP claims its network meets QoS standards 98% of the time. In a sample of 400 service windows, 11 showed violations. Test the claim at $\alpha = 0.05$.

**Problem 2 (Type II error):** For $H_0: \mu = 5\%$ vs $H_a: \mu > 5\%$, $\alpha = 0.05$, $n = 64$, $\sigma = 2\%$. Compute $\beta$ at $\mu_a = 5.5\%$ and $\mu_a = 6\%$. Comment on the pattern.

**Problem 3 (Sample size):** A risk manager needs to detect a rise in mean VaR from 2.0% to 2.3% ($\sigma = 0.8\%$) with $\alpha = 0.05$ and $\beta = 0.10$. How many observations are needed?

**Problem 4 (Two-sample $t$):** Two bond fund managers each have 12 months of data. Manager A: $\bar Y = 3.4\%$, $s = 1.2\%$. Manager B: $\bar Y = 2.8\%$, $s = 1.5\%$. Test $H_0: \mu_A = \mu_B$ at $\alpha = 0.05$ using a pooled $t$-test.
:::
:::

---

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business, ADA University

📧 [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Two-Sample Tests (§10.7, 10.9)

**Reading:** Chapter 10, Sections 10.7 and 10.9

**Preparation:** Review $F$-distribution table (Table 7, Appendix 3); recall Chi-squared distribution properties

**⏰ Reminders:**

✅ Complete Practice Problems 1–4

✅ Verify you can compute $S_p$ and pooled $t$ by hand

✅ Reflect on when $Z$ vs $t$ is appropriate

✅ Work hard!
:::
:::

---

## ❓ Questions? {.center}

::: {.callout-note}
## 💬 Open Discussion

**Key Topics for Discussion:**

- Why do we use $p_0$ (not $\hat p$) in the denominator of the proportion Z test?

- In finance, is a Type I or Type II error more costly when auditing fund returns? Does it depend on the auditor's role (regulator vs investor)?

- Why does sample size appear under a square root in the Z statistic but squared in the sample size formula?

- If $n$ is large, why does the distinction between $t$ and $Z$ become negligible?
:::

<div class="attendance-qr" data-status="late" id="qr-late" style="min-height: 400px; display: flex; align-items: center; justify-content: center;"></div>
