---
title: "Mathematical Statistics"
subtitle: "Two-Sample Tests & Variance Testing"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    margin: 0.04
    min-scale: 0.2
    max-scale: 2.0
    mouse-wheel: true
    overview: true
    zoom: true
    # footer: "Mathematical Statistics – Two-Sample Tests & Variance Testing"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
    include-after-body:
      - text: |
          <script src="../qrcode.min.js"></script>
          <script src="../attendance-qr-slide.js"></script>
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::: {style="font-size:42px;"}
::: {.learning-objectives}
By the end of this lecture, you will be able to:

- **Justify** the choice between one-tailed and two-tailed tests based on financial decision context

- **Apply** the $\chi^2$ test to assess whether a population variance equals a specified value

- **Construct** the $F$-test to compare two population variances

- **Interpret** $F$-ratio critical values and p-values from tables and software

- **Connect** variance testing to risk management decisions in finance and regulation
:::
:::

---

## 📱 Attendance Check-in {.center background-color="#f0f4ff"}

<div class="attendance-qr" data-status="present" id="qr-present" style="min-height: 400px; display: flex; align-items: center; justify-content: center;"></div>

## 📋 Overview {.smaller}

::: {style="font-size:50px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **One-Tail vs Two-Tail Logic** — when each is appropriate (§10.7)

- **$\chi^2$ Test for $\sigma^2$** — testing a single variance against a benchmark (§10.9)

- **$F$-Test for $\sigma_1^2 = \sigma_2^2$** — comparing two population variances (§10.9)

- **Two-Tailed $F$-Test** — using the larger-over-smaller convention

- **Case Study** — Comparing volatility of two asset classes with real market data
:::
:::
:::

---

## 📖 Motivation: Why Variance Matters {.smaller}

::: {style="font-size:42px;"}
::: {.callout-note}
## 🎯 Variance = Risk in Finance

In finance, variance and standard deviation are the **primary measures of risk**. Testing variances is not abstract — it drives billion-dollar decisions.

::: {.columns}
::: {.column width="50%"}
::: {.fragment}
**Single Variance Tests:**

- Has a portfolio's volatility breached its risk mandate?
- Has a broadband network's latency variance exceeded the SLA?
- Did a new process reduce the variability of loan processing times?
- Is a trading strategy's return distribution within expected bounds?
:::
:::

::: {.column width="50%"}
::: {.fragment}
**Two-Variance Comparisons:**

- Is Fund A riskier than Fund B (before choosing a pooled $t$-test)?
- Are two markets equally volatile after a regulatory change?
- Does automated underwriting produce more consistent decisions?
- Do two ISPs have equal QoS variance across customers?
:::
:::
:::
:::
:::

---

## 📖 One-Tail vs Two-Tail: Choosing Correctly (§10.7) {.smaller}

::: {style="font-size:33px;"}
The choice of $H_a$ should reflect the **practical question**, not data snooping.

**Use a one-tailed test when:**

- The loss from one direction is much greater (e.g., volatility *exceeding* a limit)
- Prior knowledge restricts the plausible direction of departure
- A regulator cares only if a metric *exceeds* a threshold

**Use a two-tailed test when:**

- Any departure from $\theta_0$ is problematic (too high *or* too low)
- The research question is "has something changed?" not "has it increased?"
- You have no prior directional hypothesis
:::

---

## ⚠️ The Pre-commitment Rule {.smaller}

::: {style="font-size:36px;"}
::: {.callout-important}
## ⚠️ Pre-commit to $H_a$ Before Seeing the Data

**$H_a$ must be specified before seeing the data.** Choosing the tail after observing results inflates the effective Type I error rate.

| Action | Stated $\alpha$ | True $\alpha$ |
|--------|----------------|---------------|
| Pre-specified $H_a$ | 0.05 | 0.05 |
| Post-hoc tail choice | 0.05 | **0.10** |

This is sometimes called **p-hacking** or **data dredging** — it invalidates the test.
:::
:::

---

## 📖 The $\chi^2$ Distribution — Quick Review {.smaller}

::: {style="font-size:33px"}

::: {.columns}
::: {.column width="55%"}

::: {.fragment}
If $Y_1, \ldots, Y_n \sim N(\mu, \sigma^2)$ independently, then:

$$\frac{(n-1)S^2}{\sigma^2} \sim \chi^2_{n-1}$$

This is **Theorem 7.3** from Chapter 7.
:::

::: {.fragment}
**Properties of $\chi^2_\nu$:**

- Always $\geq 0$ — right-skewed
- Mean $= \nu$, Variance $= 2\nu$
- As $\nu \to \infty$, approaches Normal
- $\chi^2_\alpha$: upper-tail critical value; $P(\chi^2 > \chi^2_\alpha) = \alpha$
:::
:::

::: {.column width="45%"}
::: {.fragment}
**Critical value notation:**

| Tail | Symbol | Meaning |
|------|--------|---------|
| Upper | $\chi^2_\alpha$ | $P(\chi^2 > \chi^2_\alpha) = \alpha$ |
| Lower | $\chi^2_{1-\alpha}$ | $P(\chi^2 < \chi^2_{1-\alpha}) = \alpha$ |

From Table 6, Appendix 3 — always indexed by **df = $n-1$**.
:::
:::
:::
:::

---

## 🧮 $\chi^2$ Test for a Single Variance $\sigma^2$ {.smaller}

::: {style="font-size:43px;"}
::: {.callout-important}
## Test: $H_0: \sigma^2 = \sigma^2_0$ — One Sample from Normal Population (§10.9)

$$\chi^2 = \frac{(n-1)S^2}{\sigma^2_0} \sim \chi^2_{n-1} \text{ under } H_0$$

**Rejection regions** at level $\alpha$:

| $H_a$ | Rejection Region |
|-------|-----------------|
| $\sigma^2 > \sigma^2_0$ | $\chi^2 > \chi^2_\alpha$ |
| $\sigma^2 < \sigma^2_0$ | $\chi^2 < \chi^2_{1-\alpha}$ |
| $\sigma^2 \neq \sigma^2_0$ | $\chi^2 > \chi^2_{\alpha/2}$ or $\chi^2 < \chi^2_{1-\alpha/2}$ |

:::

**Intuition:** If $S^2 \gg \sigma^2_0$, the ratio is large → evidence $\sigma^2 > \sigma^2_0$. The $\chi^2$ distribution tells us how large is "surprisingly large."
:::

---

## 📌 Example 1: Fund Volatility Mandate {.smaller}

::: {style="font-size:37px;"}
**Scenario:** A risk mandate requires that a fund's monthly return standard deviation be **no greater than $\sigma_0 = 3\%$**. Over $n = 25$ months, the observed $s = 3.8\%$.

Test $H_0: \sigma^2 = 0.0009$ vs $H_a: \sigma^2 > 0.0009$ at $\alpha = 0.05$.

:::{.fragment}
**Test statistic:**

$$\chi^2 = \frac{(n-1)s^2}{\sigma^2_0} = \frac{24 \times (0.038)^2}{(0.030)^2} = \frac{24 \times 0.001444}{0.0009} = \frac{0.03466}{0.0009} = 38.5$$
:::

:::{.fragment}
**Critical value:** $\chi^2_{0.05}$ with $\nu = 24$ df $= 36.415$.

Since $38.5 > 36.415$ → **Reject $H_0$** ✅
:::

:::{.fragment}
**Conclusion:** At $\alpha = 0.05$, there is significant evidence that the fund's volatility **breaches its 3% mandate**. The risk committee must be notified.
:::
:::

---

## 📌 Example 2: p-Value for the $\chi^2$ Test {.smaller}

**Continuing Example 1:** Bound the p-value using Table 6.

::: {style="font-size:33px"}

For $\nu = 24$ df:

| $\alpha$ | $\chi^2_\alpha$ |
|----------|----------------|
| 0.050 | 36.415 |
| 0.025 | 39.364 |

Our observed $\chi^2 = 38.5$ falls between $36.415$ and $39.364$:

$$0.025 < p\text{-value} < 0.050$$

::: {.fragment}
**Exact p-value** (from R): $P(\chi^2_{24} > 38.5) = 0.029$

**Practical interpretation:** The 2.9% probability confirms the volatility breach is statistically real — not merely sampling noise. The fund has a roughly 1-in-35 chance of seeing this variance under the null.
:::
:::

---

## 📖 The $F$-Distribution — Quick Review {.smaller}

::: {style="font-size:31px"}

::: {.callout-note}
## 📝 Key Property

If $U \sim \chi^2_{\nu_1}$ and $V \sim \chi^2_{\nu_2}$ are independent, then:

$$F = \frac{U/\nu_1}{V/\nu_2} \sim F_{\nu_1, \nu_2}$$

:::

::: {.columns}
::: {.column width="55%"}
::: {.fragment}
**Properties of $F_{\nu_1, \nu_2}$:**

- Always $\geq 0$, right-skewed
- $\nu_1$ = numerator df, $\nu_2$ = denominator df
- $F_\alpha$: upper-tail critical value; $P(F > F_\alpha) = \alpha$
- Reciprocal property: $F_{1-\alpha, \nu_1, \nu_2} = 1/F_{\alpha, \nu_2, \nu_1}$
:::
:::

::: {.column width="45%"}
::: {.fragment}
**Key financial link:**

$$F = \frac{S_1^2/\sigma_1^2}{S_2^2/\sigma_2^2}$$

Under $H_0: \sigma_1^2 = \sigma_2^2$, this simplifies to:

$$F = \frac{S_1^2}{S_2^2} \sim F_{n_1-1,\; n_2-1}$$
:::
:::
:::
:::

---

## 🧮 $F$-Test for Equality of Two Variances {.smaller}

::: {style="font-size:41px;"}
::: {.callout-important}
## Test: $H_0: \sigma_1^2 = \sigma_2^2$ — Two Independent Normal Samples (§10.9)

$$F = \frac{S_1^2}{S_2^2} \sim F_{n_1-1,\; n_2-1} \text{ under } H_0$$

**Rejection regions** at level $\alpha$:

| $H_a$ | Rejection Region | Convention |
|-------|-----------------|-----------|
| $\sigma_1^2 > \sigma_2^2$ | $F > F_{\alpha, \nu_1, \nu_2}$ | Label suspected larger variance as $S_1^2$ |
| $\sigma_1^2 \neq \sigma_2^2$ | $F > F_{\alpha/2, \nu_1, \nu_2}$ | Put larger $S^2$ in numerator |

:::

::: {.fragment}
**Two-tailed shortcut (Wackerly §10.9):** Always place the **larger** $S^2$ in the numerator. Compare to $F_{\alpha/2}$ (upper tail only). This avoids needing lower-tail $F$ values.
:::
:::

---

## 📌 Example 3: Comparing Volatilities {.smaller}

::: {style="font-size:36px;"}
**Scenario:** Two equity funds. Fund A ($n_1=10$): $s_1^2 = 0.0003$. Fund B ($n_2=20$): $s_2^2 = 0.0001$.

Test $H_0: \sigma_A^2 = \sigma_B^2$ vs $H_a: \sigma_A^2 > \sigma_B^2$ at $\alpha = 0.05$.

**Test statistic** (Fund A in numerator — suspected larger):

$$F = \frac{s_1^2}{s_2^2} = \frac{0.0003}{0.0001} = 3.0, \quad \nu_1 = 9,\; \nu_2 = 19$$
:::

---

## 📌 Example 3: Decision & p-Value {.smaller}

::: {style="font-size:36px;"}
**Critical value:** $F_{0.05, 9, 19} = 2.42$ (Table 7, App. 3).

Since $F = 3.0 > 2.42$ → **Reject $H_0$** ✅

Fund A is significantly more volatile than Fund B.

**p-value bounds from table ($\nu_1=9$, $\nu_2=19$):**

| $\alpha$ | $F_\alpha$ |
|----------|-----------|
| 0.025 | 2.88 |
| 0.010 | 3.52 |

Since $2.88 < F = 3.0 < 3.52$: $\quad 0.01 < p\text{-value} < 0.025$.

Exact p-value $= 0.021$.
:::

---

## 📌 Example 4: Two-Tailed $F$-Test (Wackerly Ex 10.21) {.smaller}

::: {style="font-size:30px;"}
**Scenario:** Compare return volatility of a Growth fund ($n_1=14$, $s_1^2 = 12.7$) and Value fund ($n_2=10$, $s_2^2 = 26.4$).

$H_0: \sigma_G^2 = \sigma_V^2$ vs $H_a: \sigma_G^2 \neq \sigma_V^2$, $\alpha = 0.10$.

:::{.fragment}
**Two-tailed shortcut:** Place larger variance in numerator:

$$F = \frac{s_{\text{larger}}^2}{s_{\text{smaller}}^2} = \frac{26.4}{12.7} = 2.079$$

df: $\nu_1 = n_V - 1 = 9$, $\nu_2 = n_G - 1 = 13$
:::

:::{.fragment}
**Critical value:** Compare to $F_{0.05, 9, 13} = 2.71$ (using $\alpha/2 = 0.05$).

Since $2.079 < 2.71$ → **Fail to Reject $H_0$**
:::

:::{.fragment}
Exact p-value $= 2 \times P(F_{9,13} > 2.079) = 2 \times 0.1005 = 0.201$ — no significant difference in volatility.
:::
:::

---

## 🎮 Interactive: $\chi^2$ and $F$ Critical Values {.smaller}

:::::: {style="font-size: 0.72em; margin-top: -8px;"}
**Explore how degrees of freedom shape the $\chi^2$ and $F$ distributions and their critical values.**

::::: columns
::: {.column width="30%"}

```{ojs}
//| echo: false

viewof dist_type = {
  const input = Inputs.select(["Chi-squared", "F distribution"], {label: "Distribution:"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

viewof df1 = {
  const input = Inputs.range([1, 40], {value: 10, step: 1, label: "df₁ (or ν):"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

viewof df2 = {
  const input = Inputs.range([1, 40], {value: 15, step: 1, label: "df₂ (F only):"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

viewof alpha_var = {
  const input = Inputs.range([0.01, 0.10], {value: 0.05, step: 0.01, label: "α level:"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

// Gamma function approximation
lgamma = function(n) {
  const c = [76.18009172947146, -86.50532032941677, 24.01409824083091,
             -1.231739572450155, 0.1208650973866179e-2, -0.5395239384953e-5];
  let x = n, y = n, tmp = x + 5.5;
  tmp -= (x + 0.5) * Math.log(tmp);
  let ser = 1.000000000190015;
  for (let j = 0; j < 6; j++) ser += c[j] / ++y;
  return -tmp + Math.log(2.5066282746310005 * ser / x);
}

// Chi-squared PDF
chi2pdf = function(x, k) {
  if (x <= 0) return 0;
  return Math.exp((k/2-1)*Math.log(x) - x/2 - (k/2)*Math.log(2) - lgamma(k/2));
}

// F PDF approximation
fpdf = function(x, d1, d2) {
  if (x <= 0) return 0;
  const lnum = (d1/2)*Math.log(d1) + (d2/2)*Math.log(d2) + (d1/2-1)*Math.log(x);
  const lden = lgamma(d1/2) + lgamma(d2/2) - lgamma((d1+d2)/2) + ((d1+d2)/2)*Math.log(d2+d1*x);
  return Math.exp(lnum - lden);
}

// Approximate upper-tail critical value via bisection on CDF (numerical integration)
trapCDF = function(pdf, lo, hi, n) {
  const h = (hi - lo) / n;
  let s = 0;
  for (let i = 0; i <= n; i++) {
    const x = lo + i * h;
    const w = (i === 0 || i === n) ? 0.5 : 1;
    s += w * pdf(x);
  }
  return s * h;
}

critVal = {
  const xmax = dist_type === "Chi-squared" ? Math.max(df1 * 4, 50) : 15;
  const total = trapCDF(
    dist_type === "Chi-squared" ? (x => chi2pdf(x, df1)) : (x => fpdf(x, df1, df2)),
    0.001, xmax, 2000
  );
  // Bisect to find upper alpha critical value
  let lo = 0.001, hi = xmax;
  for (let i = 0; i < 60; i++) {
    const mid = (lo + hi) / 2;
    const tail = trapCDF(
      dist_type === "Chi-squared" ? (x => chi2pdf(x, df1)) : (x => fpdf(x, df1, df2)),
      mid, xmax, 1000
    ) / total;
    tail > alpha_var ? (lo = mid) : (hi = mid);
  }
  return (lo + hi) / 2;
}

info_html2 = html`<div style="background:#f0f4ff; padding:8px; border-radius:6px; font-size:0.93em; line-height:1.6;">
  <strong>Distribution:</strong> ${dist_type === "Chi-squared" ? `χ²(${df1})` : `F(${df1},${df2})`}<br>
  <strong>α = ${alpha_var.toFixed(2)}</strong><br>
  <strong>Critical value:</strong> ${critVal.toFixed(3)}<br>
  <em style="color:#555;">Reject H₀ if statistic > ${critVal.toFixed(3)}</em>
</div>`;

md`${info_html2}`
```

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

plotDist = {
  const xmax = dist_type === "Chi-squared" ? Math.max(df1 * 3.5, 30) : 10;
  const pts = [];
  for (let x = 0.05; x <= xmax; x += xmax / 300) {
    const y = dist_type === "Chi-squared" ? chi2pdf(x, df1) : fpdf(x, df1, df2);
    pts.push({x, y, region: x >= critVal ? "reject" : "accept"});
  }
  return pts;
}

Plot.plot({
  width: 680, height: 360,
  marginLeft: 55, marginBottom: 50,
  x: { label: dist_type === "Chi-squared" ? "χ² statistic" : "F statistic" },
  y: { label: "Density" },
  title: `${dist_type === "Chi-squared" ? `χ²(${df1})` : `F(${df1}, ${df2})`} | α=${alpha_var.toFixed(2)} | Critical value = ${critVal.toFixed(3)}`,
  marks: [
    Plot.areaY(plotDist.filter(d => d.region === "reject"),
               {x: "x", y: "y", fill: "tomato", fillOpacity: 0.45}),
    Plot.areaY(plotDist.filter(d => d.region === "accept"),
               {x: "x", y: "y", fill: "steelblue", fillOpacity: 0.25}),
    Plot.line(plotDist, {x: "x", y: "y", stroke: "#333", strokeWidth: 2}),
    Plot.ruleX([critVal], {stroke: "tomato", strokeWidth: 2.5, strokeDasharray: "6,3"}),
    Plot.text([{x: critVal, y: plotDist[0]?.y * 0.6 ?? 0.05,
                text: ` ← c.v. = ${critVal.toFixed(2)}`}],
              {x: "x", y: "y", text: "text", fontSize: 13, fill: "tomato", textAnchor: "start"})
  ]
})
```

*Red shaded area = rejection region ($\alpha$). Toggle distribution, df, and $\alpha$ with sliders.*
:::
:::::
::::::

---

## 🤝 Think-Pair-Share {.smaller}

::: {style="font-size:37px;"}
::: {.callout-tip}
## 💬 Activity (5 minutes)

**Scenario:** A fintech regulator monitors two payment processors, FinPay A and FinPay B, for consistency of transaction processing times (in seconds). Random samples yield:

| | FinPay A | FinPay B |
|-|----------|----------|
| $n$ | 16 | 21 |
| $\bar{Y}$ (sec) | 1.42 | 1.38 |
| $s^2$ | 0.18 | 0.07 |

**Questions:**

1. The regulator suspects FinPay A is more variable. State $H_0$ and $H_a$.

2. Compute the $F$-statistic and identify the degrees of freedom.

3. Find the rejection region at $\alpha = 0.05$. What is your decision?

4. Before running a pooled $t$-test for the means, why does this $F$-test matter?
:::
:::

```{r}
#| echo: false
#| eval: false
library(countdown)
countdown(minutes = 5, bottom = 0, right = 0, color_border = "#003366")
```

---

## ✅ Think-Pair-Share: Solution {.smaller}

::: {style="font-size: 36px;"}
::: {.columns}
::: {.column width="50%"}

**1. Hypotheses:**

$$H_0: \sigma_A^2 = \sigma_B^2 \quad H_a: \sigma_A^2 > \sigma_B^2$$

One-tailed — regulator suspects A is more variable.

:::

::: {.column width="50%"}

**2. $F$-Statistic:**

Place $s_A^2 = 0.18$ (suspected larger) in numerator:

$$F = \frac{0.18}{0.07} = 2.571, \quad \nu_1 = 15,\; \nu_2 = 20$$

:::
:::
:::

---

## ✅ Solution (Continued): Decision & Implications {.smaller}

::: {style="font-size: 36px;"}
::: {.columns}
::: {.column width="50%"}

**3. Decision:**

$F_{0.05, 15, 20} \approx 2.20$ (Table 7).

Since $F = 2.571 > 2.20$ → **Reject $H_0$** ✅

p-value $\approx 0.024$ — FinPay A significantly more variable.

:::

::: {.column width="50%"}

**4. Connection to $t$-Test:**

Pooled $t$-test requires $\sigma_A^2 = \sigma_B^2$. Since rejected, use **Welch's $t$-test** instead.

> 💡 Always run $F$-test before pooled $t$. An invalid pooled $t$ gives wrong p-values.

:::
:::
:::

---

## 💰 Case Study: Volatility Comparison {.smaller}

::: {style="font-size:22px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

library(tidyverse)
library(tidyquant)
library(knitr)

# Growth ETF (QQQ) vs Value ETF (IVE) — comparing volatility
qqq <- tq_get("QQQ", from = "2020-01-01", to = "2023-12-31")
ive <- tq_get("IVE", from = "2020-01-01", to = "2023-12-31")

get_monthly_ret <- function(df, nm) {
  df %>%
    tq_transmute(select = adjusted,
                 mutate_fun = periodReturn,
                 period = "monthly",
                 col_rename = "return") %>%
    mutate(asset = nm)
}

returns <- bind_rows(
  get_monthly_ret(qqq, "Growth (QQQ)"),
  get_monthly_ret(ive, "Value (IVE)")
)

# F-test for equal variances
qqq_r <- returns %>% filter(asset == "Growth (QQQ)") %>% pull(return)
ive_r <- returns %>% filter(asset == "Value (IVE)")  %>% pull(return)

ftest <- var.test(qqq_r, ive_r)

summary_tbl <- returns %>%
  group_by(asset) %>%
  summarise(n       = n(),
            mean    = mean(return) * 100,
            std_dev = sd(return) * 100,
            variance = var(return) * 10000)

kable(summary_tbl, digits = 3,
      caption = "Monthly Return Statistics (2020-2023)")
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 5.2

# Visualise
ggplot(returns, aes(x = return * 100, fill = asset)) +
  geom_density(alpha = 0.45, colour = NA) +
  geom_vline(data = returns %>%
               group_by(asset) %>%
               summarise(m = mean(return) * 100),
             aes(xintercept = m, colour = asset),
             linetype = "dashed", linewidth = 1) +
  annotate("text", x = 7, y = 0.115,
           label = paste0("F = ",   round(ftest$statistic, 3), "\n",
                          "p = ",   round(ftest$p.value, 3)),
           size = 4, colour = "black",
           hjust = 0) +
  scale_fill_manual(values  = c("Growth (QQQ)" = "steelblue",
                                "Value (IVE)"  = "tomato")) +
  scale_colour_manual(values = c("Growth (QQQ)" = "steelblue",
                                 "Value (IVE)"  = "tomato")) +
  labs(title    = "Monthly Return Distributions: Growth vs Value",
       subtitle = "F-test for equal variances",
       x = "Monthly Return (%)", y = "Density",
       fill = NULL, colour = NULL) +
  theme_minimal(base_size = 12)
```

:::
:::
:::

---

## 💰 Case Study: Key Findings {.smaller}

::: {style="font-size: 38px;"}
::: {.callout-important}
## 📊 Growth vs Value Volatility — F-Test Results (2020–2023)

::: {.columns}

::: {.column width="33%"}
::: {.fragment}
**Test Setup:**

- $H_0: \sigma^2_{\text{QQQ}} = \sigma^2_{\text{IVE}}$
- $H_a: \sigma^2_{\text{QQQ}} \neq \sigma^2_{\text{IVE}}$
- Two-tailed $F$-test, $\alpha = 0.05$
- QQQ ($n=48$): $s \approx 6.1\%$
- IVE ($n=48$): $s \approx 4.8\%$
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Results:**

- $F = s^2_{\text{QQQ}}/s^2_{\text{IVE}} \approx 1.62$
- df: $47$ and $47$
- $F_{0.025, 47, 47} \approx 1.70$
- $F < F_{crit}$ → **Fail to Reject $H_0$**
- p-value $\approx 0.09$ — marginal

Growth and Value variances are not significantly different at 5%.
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Practical Implications:**

1. **Portfolio construction**: The two ETFs can be treated as having similar risk levels for mean-variance optimisation

2. **Pooled $t$-test validity**: Equal variance assumption is not rejected → pooled $t$ is defensible for comparing mean returns

3. **Period sensitivity**: The 2020–2023 window includes COVID crash — always check robustness to time period choice
:::
:::

:::
:::
:::

---

## 📝 Quiz #1: Choosing the Right Test {.smaller .quiz-question}

A risk manager has 30 months of monthly return data for a bond fund and wants to verify whether the volatility has **increased above** the benchmark $\sigma_0 = 1.5\%$. Which test statistic is appropriate?

- [$\chi^2 = (n-1)S^2/\sigma_0^2$ with $\nu = 29$ df, upper-tail rejection region]{.correct data-explanation="✅ Correct! Single variance test against a specified benchmark uses χ² = (n−1)S²/σ₀² with df = n−1 = 29. Upper tail because Hₐ: σ² > σ₀²."}
- $F = S_1^2/S_2^2$ with two independent samples
- $Z = (S - \sigma_0)/(\sigma_0/\sqrt{2n})$, standard normal
- $t = (\bar Y - \mu_0)/(S/\sqrt{n})$ with $\nu = 29$ df

---

## 📝 Quiz #2: $F$-Test Critical Region {.smaller .quiz-question}

Testing $H_0: \sigma_A^2 = \sigma_B^2$ vs $H_a: \sigma_A^2 \neq \sigma_B^2$ at $\alpha = 0.10$ with $n_A = 11$, $n_B = 16$, and $s_A^2 > s_B^2$. Using the larger-over-smaller convention, which is the correct rejection criterion?

- [Reject if $F = s_A^2/s_B^2 > F_{0.05,\, 10,\, 15}$]{.correct data-explanation="✅ Correct! Two-tailed test at α=0.10 → compare to F_{α/2} = F_{0.05}. Larger variance in numerator gives df₁ = nₐ−1 = 10, df₂ = n_B−1 = 15. Reject if F > F_{0.05, 10, 15}."}
- Reject if $F > F_{0.10,\, 10,\, 15}$
- Reject if $F > F_{0.05,\, 15,\, 10}$
- Reject if $F > F_{0.025,\, 10,\, 15}$

---

## 📝 Quiz #3: Computing the $\chi^2$ Statistic {.smaller .quiz-question}

A telecom regulator requires that the variance of download speed across customers be at most $\sigma_0^2 = 25$ Mbps². A sample of $n = 20$ customers gives $s^2 = 34$ Mbps². What is the $\chi^2$ statistic and what are the degrees of freedom?

- [$\chi^2 = 19 \times 34/25 = 25.84$, with $\nu = 19$ df]{.correct data-explanation="✅ Correct! χ² = (n−1)S²/σ₀² = 19×34/25 = 25.84. Degrees of freedom = n−1 = 19."}
- $\chi^2 = 20 \times 34/25 = 27.2$, with $\nu = 20$ df
- $\chi^2 = 19 \times 25/34 = 13.97$, with $\nu = 19$ df
- $\chi^2 = \sqrt{34/25} = 1.166$, with $\nu = 19$ df

---

## 📝 Quiz #4: F-Test Interpretation {.smaller .quiz-question}

An analyst computes $F = 2.15$ for a two-tailed variance test with $\nu_1 = 12$, $\nu_2 = 18$ df. From tables: $F_{0.05, 12, 18} = 2.34$ and $F_{0.025, 12, 18} = 2.77$. What is the correct conclusion at $\alpha = 0.10$?

- [Reject $H_0$; since $F = 2.15 > F_{0.05, 12, 18} = 2.34$ is not satisfied — actually fail to reject. Wait, $2.15 < 2.34$ → Fail to reject at $\alpha = 0.10$]{.correct data-explanation="✅ Correct! Two-tailed test at α=0.10 compares F to F_{0.05, 12, 18} = 2.34. Since F = 2.15 < 2.34, we fail to reject H₀. There is insufficient evidence of unequal variances."}
- Reject $H_0$; since $F = 2.15 > 1$ we have evidence of unequal variances
- Reject $H_0$; since $F = 2.15 > F_{0.025, 12, 18} = 2.77$ — incorrect, 2.15 < 2.77
- Cannot determine without the exact p-value

---

## 📝 Summary {.smaller}

::: {style="font-size: 34px"}
::: {.summary-box}
**✅ Key Takeaways**

- **Tail choice (§10.7)**: Pre-specify $H_a$ based on what decision you care about; one-tailed when only one direction matters, two-tailed when any change is relevant

- **$\chi^2$ test**: $\chi^2 = (n-1)S^2/\sigma_0^2$ with $\nu = n-1$ df tests a single variance against a benchmark; right-skewed distribution means rejection regions differ for upper vs lower tail

- **$F$-test**: $F = S_1^2/S_2^2$ compares two variances; use larger $S^2$ in numerator and compare to $F_{\alpha/2}$ for two-tailed tests

- **Gateway test**: The $F$-test for equal variances should precede any pooled two-sample $t$-test — it validates the equal-variance assumption

- **Risk management link**: Variance tests directly answer whether volatility mandates are breached or whether two assets have comparable risk profiles
:::
:::

---

## 📚 Practice Problems

::: {style="font-size:42px;"}
::: {.callout-tip}
## 📝 Homework Problems — Chapter 10 (§10.7, §10.9)

**Problem 1 ($\chi^2$ test):** A derivatives desk requires that daily P&L variance be at most $\sigma_0^2 = 4$ (USD millions)². Over 25 trading days, $s^2 = 5.8$. Test $H_0: \sigma^2 = 4$ vs $H_a: \sigma^2 > 4$ at $\alpha = 0.05$. Find the exact p-value in R.

**Problem 2 (Tail choice):** For each scenario below, state whether you would use a one-tailed or two-tailed test and explain: (a) Testing whether a fund's Sharpe ratio has *changed* after a manager switch; (b) Testing whether default rates have *increased* after a recession; (c) Testing whether two ISPs have *equal* complaint rates.

**Problem 3 ($F$-test):** Fund A ($n_1=25$, $s_1=2.4\%$) and Fund B ($n_2=31$, $s_2=1.7\%$). Test $H_0: \sigma_A^2 = \sigma_B^2$ vs $H_a: \sigma_A^2 > \sigma_B^2$ at $\alpha = 0.05$. Is the pooled $t$-test assumption valid?

**Problem 4 (Combined):** Using the datasets from Problem 3, perform a pooled $t$-test (if justified) or Welch's $t$-test for $H_0: \mu_A = \mu_B$ given $\bar Y_A = 1.9\%$ and $\bar Y_B = 1.6\%$. What is the full inference chain?
:::
:::

---

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business, ADA University

📧 [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Linear Models & Least Squares (Chapter 11)

**Reading:** Chapter 11, Sections 11.1–11.5

**Preparation:** Review covariance and correlation from Chapter 5; recall the formula for $\hat\beta_1 = S_{xy}/S_{xx}$

**⏰ Reminders:**

✅ Complete Practice Problems 1–4

✅ Make sure you can read $F$-tables (Table 7, Appendix 3)

✅ Think about where you would apply each variance test in your field

✅ Work hard!
:::
:::

---

## ❓ Questions? {.center}

::: {.callout-note}
## 💬 Open Discussion

**Key Topics for Discussion:**

- Why is the $F$-distribution always right-skewed, and what does this imply about using only the upper tail?

- In risk management, is it more dangerous to make a Type I or Type II error when testing a volatility mandate? Does it depend on who you are (regulator vs manager)?

- Why does placing the larger $S^2$ in the numerator of the $F$-test help avoid needing lower-tail $F$ critical values?

- When would you prefer a Levene test or Bartlett test over the $F$-test for equal variances?
:::

<div class="attendance-qr" data-status="late" id="qr-late" style="min-height: 400px; display: flex; align-items: center; justify-content: center;"></div>
