---
title: "Mathematical Statistics"
subtitle: "Random Variables, Random Sampling, and Chapter 2 in Review"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - Random Variables and Random Sampling"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
library(ggplot2)
theme_set(theme_minimal(base_size = 18))
```

## 🎯 Learning Objectives

::: {style="font-size: 30px"}
By the end of this lecture, you will be able to:

- **Define** a random variable as a real-valued function on $S$ (Definition 2.12) — *load-bearing*

- **Compute** $P(Y = y)$ by summing the probabilities of the sample points in $\{Y = y\}$ — *load-bearing*

- **Count** the $\binom{N}{n}$ possible simple random samples and show each unit is included with probability $n/N$

- **Choose** the right Chapter 2 tool for a problem, and combine several in one calculation

The first two carry all of Chapter 3. Sampling and the summary we meet more briefly.
:::

---

## 🗺️ Where We Are

::: {style="font-size: 30px"}
**Wackerly §§2.11–2.13**

Last class ended: *turning outcomes into numbers — random variables, random sampling, and a synthesis of Chapter 2.*

::: {.fragment}
A bank's risk committee holding three SME loans does not ask *which* loans default. It asks **how many**, and **how much money** is lost.
:::

::: {.fragment}
Those are numbers, not lists of outcomes. Today we build the bridge from a sample space to a number — and from there, in Chapter 3, to a distribution.
:::
:::

---

## 📐 Definition 2.12: Random Variable

::: {style="font-size: 30px"}
::: {.callout-note}
## Definition 2.12 (Wackerly §2.11)
A **random variable** is a real-valued function for which the domain is a sample space.
:::

::: {.fragment}
So $Y$ takes each sample point $E_i \in S$ to a number $Y(E_i)$. Several points may share a value; no point gets two.
:::

::: {.fragment}
The **numerical event** $\{Y = y\}$ is the set of all sample points assigned the value $y$. These sets **partition** $S$: they are mutually exclusive, and together they cover it.
:::
:::

---

## 🏦 Three SME Loans

::: {style="font-size: 28px"}
A Baku bank holds three loans that default independently: construction, $P = 0.10$; retail, $P = 0.05$; agriculture, $P = 0.20$. Let $Y$ = number of defaults.

| Sample point | Probability | $Y$ |
|---|---|---|
| NNN | $0.9 \times 0.95 \times 0.8 = 0.684$ | 0 |
| DNN, NDN, NND | $0.076,\ 0.036,\ 0.171$ | 1 |
| DDN, DND, NDD | $0.004,\ 0.019,\ 0.009$ | 2 |
| DDD | $0.1 \times 0.05 \times 0.2 = 0.001$ | 3 |

The points are **not** equally likely, so we cannot count — we must sum.
:::

---

## ➕ From Sample Points to $P(Y = y)$

::: {style="font-size: 30px"}
$P(Y = y)$ is the sum of the probabilities of the sample points assigned the value $y$.

::: {.fragment}
$$P(Y = 1) = P(\text{DNN}) + P(\text{NDN}) + P(\text{NND}) = 0.076 + 0.036 + 0.171 = 0.283$$
:::

::: {.fragment}
$$P(Y = 0) = 0.684, \quad P(Y = 2) = 0.032, \quad P(Y = 3) = 0.001$$
:::

::: {.fragment}
The four values sum to 1, because the events $\{Y = y\}$ partition $S$. **That check is free — always make it.**
:::
:::

---

## 💻 The Same Calculation in R

```{r}
#| label: loans-enumerate
#| code-fold: false
p  <- c(construction = 0.10, retail = 0.05, agriculture = 0.20)
S  <- expand.grid(c1 = 0:1, c2 = 0:1, c3 = 0:1)          # 8 sample points
S$prob <- apply(S, 1, function(d) prod(ifelse(d == 1, p, 1 - p)))
S$Y    <- rowSums(S[, 1:3])                             # the function Y(E_i)
pY <- tapply(S$prob, S$Y, sum)                           # sum within {Y = y}
round(pY, 3); sum(pY)
```

---

## 📊 The Picture: $P(Y = y)$

```{r}
#| label: loans-figure
#| echo: false
#| fig-width: 9
#| fig-height: 4.0
df <- data.frame(y = as.integer(names(pY)), p = as.numeric(pY))
ggplot(df, aes(factor(y), p)) +
  geom_col(fill = "#8ba3c7", width = 0.6) +
  geom_text(aes(label = sprintf("%.3f", p)), vjust = -0.4, size = 7) +
  scale_y_continuous(limits = c(0, 0.78), breaks = seq(0, 0.7, 0.1)) +
  labs(x = "Number of defaults, y", y = "P(Y = y)")
```

::: {style="font-size: 28px"}
Eight sample points collapsed into four numbers. In Chapter 3 this list of values gets a name: the **probability distribution** of $Y$.
:::

---

## 💰 A Second Function on the Same $S$

::: {style="font-size: 28px"}
Exposures are 200, 150 and 100 thousand AZN. Let $L$ = loss in default (thousand AZN).

| NNN | NND | NDN | DNN | NDD | DND | DDN | DDD |
|---|---|---|---|---|---|---|---|
| 0 | 100 | 150 | 200 | 250 | 300 | 350 | 450 |

::: {.fragment}
Same experiment, same $S$, a **different** function — and a different partition: every point gets its own value.
:::

::: {.fragment}
$P(L \ge 250) = 0.009 + 0.019 + 0.004 + 0.001 = 0.033$. Here $\{L \ge 250\} = \{Y \ge 2\}$ exactly: it is one set of sample points, described by two random variables.
:::
:::

---

## 🎲 §2.12: The Design Matters

::: {style="font-size: 28px"}
A statistical experiment observes a **sample** from a **population**, and uses the probability of the observed sample to make an inference. So the sampling method — the **design** — must be stated.

An auditor picks $n = 2$ of a bank's $N = 5$ branches. Probability of drawing one specific pair $\{a, b\}$:

::: {.fragment}
- **Without replacement:** $\binom{5}{2} = 10$ equally likely pairs, so $1/10$.
:::

::: {.fragment}
- **With replacement:** 25 equally likely ordered draws; $\{a, b\}$ arises as $(a, b)$ or $(b, a)$, so $2/25$.
:::
:::

---

## 📐 Definition 2.13: Random Sample

::: {style="font-size: 30px"}
::: {.callout-note}
## Definition 2.13 (Wackerly §2.12)
Let $N$ and $n$ be the numbers of elements in the population and sample. If the sampling is conducted so that each of the $\binom{N}{n}$ samples has an equal probability of being selected, the sampling is **random**, and the result is a **random sample**.
:::

::: {.fragment}
Each sample therefore has probability $1 / \binom{N}{n}$. In practice the draw is made by software — or, in the book, from a random number table (Table 12, Appendix 3).
:::
:::

---

## 🏛️ Worked Example: A Branch Audit

::: {style="font-size: 28px"}
The Central Bank audits a simple random sample of $n = 6$ of a commercial bank's $N = 30$ branches.

::: {.fragment}
**Number of possible samples:** $\binom{30}{6} = 593{,}775$, each with probability $1/593{,}775$.
:::

::: {.fragment}
**Is the Ganjlik branch audited?** Count the samples containing it: fix it, choose the other 5 from 29.
$$P(\text{included}) = \frac{\binom{29}{5}}{\binom{30}{6}} = \frac{6}{30} = 0.2 = \frac{n}{N}$$
:::

::: {.fragment}
Every branch faces the same $n/N$. That is the sense in which an SRS is "fair".
:::
:::

---

## 💻 Checking $n/N$ by Simulation

```{r}
#| label: srs-sim
#| code-fold: false
N <- 30; n <- 6
choose(N, n)                                   # number of possible samples
audits <- replicate(20000, sample(1:N, n))     # 20,000 simple random samples
incl <- sapply(1:N, function(b) mean(colSums(audits == b)))  # share containing b
round(range(incl), 3)                          # every branch near n/N = 0.2
```

::: {style="font-size: 28px"}
Each of the 30 branches was audited in about 20% of the simulated samples.
:::

---

## 🏘️ Not Every Survey Is One SRS

::: {style="font-size: 30px"}
A household budget survey needs 800 households across Azerbaijan. One SRS could, by pure chance, draw them mostly from Baku.

::: {.fragment}
Instead: allot a number of households to each region, and draw an SRS **within** each region. The pieces are then combined.
:::

::: {.fragment}
Such designs are only partially random, and are a course in themselves. For us, Definition 2.13 is the building block — and the random variables of Chapter 3 will be defined on its sample space.
:::
:::

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 30px"}
A survey firm interviews an SRS of $n = 3$ of the $N = 10$ households on one street. Four of the ten hold a bank loan.

**Four minutes, in pairs:**

1. How many different samples are possible?

2. What is the probability that a given household is interviewed?

3. Let $Y$ = number of indebted households in the sample. Find $P(Y = 0)$ and $P(Y = 2)$.
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 30px"}
1. $\binom{10}{3} = 120$ samples, each with probability $1/120$.

2. $\binom{9}{2}/\binom{10}{3} = 36/120 = 3/10 = n/N$.

::: {.fragment}
3. $Y$ is a random variable on the 120-point sample space. Count the points in each event:
$$P(Y = 0) = \frac{\binom{6}{3}}{120} = \frac{20}{120} = 0.167, \qquad P(Y = 2) = \frac{\binom{4}{2}\binom{6}{1}}{120} = \frac{36}{120} = 0.3$$
:::

::: {.fragment}
Equally likely points, so here summing **is** counting.
:::
:::

---

## 🧭 §2.13: Which Tool, When?

::: {style="font-size: 28px"}
| The problem gives you | Reach for |
|---|---|
| A small $S$ you can list | Sample-point method (§2.5) |
| Equally likely points, too many to list | Counting rules (§2.6) |
| "Given that …", or a two-way table | Conditional probability (§2.7) |
| "Unrelated", "separately" | Independence: multiply (§2.7) |
| "Both" / "either" / "at least one" | Multiplicative, additive laws, complement (§2.8) |
| A compound event to decompose | Event-composition method (§2.9) |
| Stages or states, then a reversal | Total probability, Bayes (§2.10) |
:::

---

## 🔗 Synthesis: Up-Days on an Index

::: {style="font-size: 28px"}
A stock index is in a **calm** regime with probability 0.7, **volatile** with 0.3. Given the regime, days are independent: $P(\text{up}) = 0.6$ if calm, $0.4$ if volatile. Let $Y$ = number of up-days in the next 3 trading days.

::: {.fragment}
**Given calm** (independence, then counting the $\binom{3}{y}$ orderings):
$$P(Y = 3 \mid \text{calm}) = 0.6^3 = 0.216, \qquad P(Y = 3 \mid \text{volatile}) = 0.4^3 = 0.064$$
:::

::: {.fragment}
**Total probability** over the two regimes:
$$P(Y = 3) = 0.7(0.216) + 0.3(0.064) = 0.1512 + 0.0192 = 0.1704$$
:::
:::

---

## 🔗 Synthesis: Reading the Regime Back

::: {style="font-size: 28px"}
**Bayes' rule:** three up-days in a row — how sure are we the market is calm?
$$P(\text{calm} \mid Y = 3) = \frac{0.1512}{0.1704} = 0.887$$

```{r}
#| label: synthesis
#| code-fold: false
y <- 0:3
calm <- choose(3, y) * 0.6^y * 0.4^(3 - y)     # P(Y = y | calm): orderings x product
vol  <- choose(3, y) * 0.4^y * 0.6^(3 - y)     # P(Y = y | volatile)
pY   <- 0.7 * calm + 0.3 * vol                 # total probability
out  <- rbind(P_Y = round(pY, 4), P_calm_given_Y = round(0.7 * calm / pY, 3))
colnames(out) <- paste0("y=", y); out
```

Prior 0.7, posterior 0.887 after three up-days — and 0.409 after none.
:::

---

## 📝 Quiz #1: Summing Sample Points {.quiz-question}

Two bond coupons are due; each is paid with probability 0.9, independently. $Y$ = number of missed coupons. What is $P(Y = 1)$?

- [$0.18$]{.correct data-explanation="✅ {Y = 1} = {paid-missed, missed-paid}, two sample points of 0.9 × 0.1 = 0.09 each, so P(Y = 1) = 0.18."}
- $0.09$
- $0.10$
- $0.81$

---

## 📝 Quiz #2: Inclusion Probability {.quiz-question}

A compliance officer draws an SRS of 5 client files from 40. What is the probability that a particular client's file is drawn?

- $1/40$
- $1/\binom{40}{5}$
- [$5/40 = 0.125$]{.correct data-explanation="✅ Samples containing the file: choose(39, 4); all samples: choose(40, 5). The ratio is 5/40 = n/N."}
- $5/39$

---

## 📝 Quiz #3: What Is a Random Variable? {.quiz-question}

A loan is observed for one year. Which of these is a **random variable**?

- The event that the loan defaults
- The sample space $\{\text{default}, \text{no default}\}$
- [The number of defaults: 1 if it defaults, 0 if not]{.correct data-explanation="✅ Definition 2.12: a real-valued function on S. The event 'defaults' is a subset of S, not a number; the 0/1 count assigns a number to each sample point."}
- The probability of default, 0.08

---

## 📋 Key Formulas

::: {style="font-size: 30px"}
| | Statement |
|---|---|
| Definition 2.12 | $Y: S \to$ the real line |
| numerical event | $\{Y = y\} = \{E_i \in S : Y(E_i) = y\}$ |
| probability of a value | $P(Y = y) = \sum_{E_i:\, Y(E_i) = y} P(E_i)$ |
| the check | $\sum_y P(Y = y) = 1$ |
| Definition 2.13 | each of $\binom{N}{n}$ samples has probability $1/\binom{N}{n}$ |
| inclusion probability | $\binom{N-1}{n-1} / \binom{N}{n} = n/N$ |
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- A random variable is a **function** on $S$; the events $\{Y = y\}$ partition $S$

- $P(Y = y)$ is a sum over sample points — a count only when they are equally likely

- One $S$ carries many random variables: defaults, losses, up-days

- An SRS makes all $\binom{N}{n}$ samples equally likely, so each unit is included with probability $n/N$

- Chapter 2 is a toolbox; hard problems use several tools in sequence
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition**

- §2.11: exercises 2.139 – 2.142; supplementary 2.150, 2.167, 2.170

- Chapter 2 supplementary exercises for review: 2.146, 2.149, 2.152, 2.157, 2.173, 2.174

**Week 5, Problem Set 1** is open now and closes **Sunday 18 October at 23:59** on WeBWorK, covering §§2.11–2.13.

**Quiz I (Chapters 1–2)** is on **17 October**, first 30 minutes of class, on WeBWorK. This is our last session on Chapter 2 — the supplementary exercises are your revision list.

**Next class:** Chapter 3 — discrete random variables, their probability distributions, and expected value [(§§3.1–3.3).]{style="white-space: nowrap"}
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**

📧 sorujov@ada.edu.az\
🏢 Building D, Room D325\
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- If $Y$ and $L$ are two random variables on the same $S$, is $Y + L$ one too? Is $Y^2$?

- Sampling with replacement gave $2/25$ for a specific pair, not $1/10$. Which design carries more information about the population?

- In the synthesis example, why are the up-days **not** independent once the regime is unknown?
:::
