---
title: "Mathematical Statistics"
subtitle: "Method of Transformations and Moment-Generating Functions"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ../ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    margin: 0.04
    min-scale: 0.2
    max-scale: 2.0
    footer: "Mathematical Statistics - Transformations and MGF Method"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    mouse-wheel: true
    overview: true
    zoom: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives

::: {.learning-objectives}
By the end of this lecture, you will be able to:

- Apply the **method of transformations** (change of variables) to find pdfs of transformed random variables

- Use the **Jacobian** for the univariate transformation formula

- Apply the **moment-generating function method** to identify distributions of sums

- Recognize when the MGF method is **more efficient** than direct integration

- Apply these techniques to derive distributions of **portfolio returns** and **risk measures**
:::

## 📋 Overview 

::: {style="font-size:50px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Method of Transformations** – Direct formula for monotonic functions

- **The Jacobian in 1D** – Why $|dy/du|$ appears in the formula

- **MGF Method** – Using moment-generating functions to identify distributions

- **Sums of Random Variables** – Convolutions and the MGF approach

- **Case Study** – Distribution of portfolio returns using MGF
:::
:::
:::

## 📖 Why Faster Transformation Methods?

::::::::: {.callout-note}
## 🎯 Motivation

:::::::: {style="font-size:30px; margin-top: -20px;"}
The distribution function method from last lecture **works** but can be tedious. In finance, we need **efficient tools**:

::::::: columns
:::: {.column width="50%" style="margin-top: -20px;"}
::: fragment
**Direct Formula Needed For:**

-   Modeling stock prices as exponentials of normal returns (log-normal)
-   Simulating asset paths from uniform random numbers
-   Pricing derivatives requiring PDF transformations
-   Converting between return types (simple vs log)
:::
::::

:::: {.column width="50%" style="margin-top: -20px;"}
::: fragment
**MGF Method Needed For:**

-   Finding portfolio return distributions (sums of asset returns)
-   Proving normality of diversified portfolios
-   Insurance claim aggregation (sum of exponentials)
-   Testing portfolio variance with chi-square distributions
:::
::::
:::::::
:::::::::
:::::::::

## 📖 Recall: Distribution Function Method

::: {.callout-tip}
## 🔗 From Last Lecture

The distribution function method works for **any** transformation:

1. Find $F_U(u) = P(U \leq u) = P(g(Y) \leq u)$
2. Express in terms of $Y$
3. Differentiate to get $f_U(u)$

**Limitation:** Can be tedious when the function is monotonic and differentiable.
:::

**Today's Question:** Can we get a **direct formula** for $f_U(u)$ without going through the CDF?

**Answer:** Yes! The method of transformations provides exactly this shortcut.

## 📖 Definition: Method of Transformations

::: {.callout-note}
## 📝 Theorem 6.2: Univariate Transformation

Let $Y$ be a continuous random variable with pdf $f_Y(y)$ on support $(a, b)$.

Let $U = g(Y)$ where $g$ is **strictly monotonic** and differentiable.

Let $h = g^{-1}$ be the inverse function, so $Y = h(U)$.

Then the pdf of $U$ is:

$$f_U(u) = f_Y(h(u)) \cdot \left| \frac{dh}{du} \right| = f_Y(h(u)) \cdot |h'(u)|$$

for $u$ in the range of $g$.

The term $|h'(u)| = |dy/du|$ is called the **Jacobian** of the transformation.
:::

## 🧮 Why Does the Jacobian Appear? 

::: {.callout-important}
## Intuition: Stretching and Compressing

Consider a small interval $[y, y + dy]$ that maps to $[u, u + du]$.

::: {.columns}
::: {.column width="50%"}
**Probability is conserved:**
$$f_Y(y) \, dy = f_U(u) \, du$$

:::

::: {.column width="50%"}
Solving for $f_U(u)$:
$$f_U(u) = f_Y(y) \cdot \frac{dy}{du} = f_Y(h(u)) \cdot |h'(u)|$$
:::
:::

The absolute value ensures the density is positive!
:::

::: {.columns}
::: {.column width="50%"}
**If $g$ is increasing:** $h'(u) > 0$

- Orientation preserved
- $|h'(u)| = h'(u)$
:::

::: {.column width="50%"}
**If $g$ is decreasing:** $h'(u) < 0$

- Orientation reversed
- $|h'(u)| = -h'(u)$
:::
:::

## 📌 Example 1: Simulating Default Times {.smaller}

**Problem:** A bank models time-to-default of a loan as $Y \sim \text{Exponential}(\beta)$. To simulate defaults using uniform random numbers, define $U = 1 - e^{-Y/\beta}$. Find the distribution of $U$.

**Solution:**

. . .

**Step 1:** Identify $g(y) = 1 - e^{-y/\beta}$. This is strictly increasing on $y > 0$.

. . .

::: {.columns}
::: {.column width="50%"}
::: {.fragment}
**Step 2:** Find the inverse. From $u = 1 - e^{-y/\beta}$:

$$e^{-y/\beta} = 1 - u \implies y = -\beta \ln(1-u) = h(u)$$
:::
:::

::: {.column width="50%"}
::: {.fragment}
**Step 3:** Compute the Jacobian:
$$h'(u) = \frac{d}{du}[-\beta \ln(1-u)] = \frac{\beta}{1-u}$$
:::
:::
:::

. . .

::: {style="margin-top: -20px;"}
**Step 4:** Apply the formula. Since $f_Y(y) = \frac{1}{\beta}e^{-y/\beta}$:
$$f_U(u) = \frac{1}{\beta}e^{-h(u)/\beta} \cdot \frac{\beta}{1-u} = \frac{1}{\beta}(1-u) \cdot \frac{\beta}{1-u} = 1 \; \text{for} \; 0 < u < 1 \rightarrow \text{Uniform}(0,1)!$$

:::

**Finance Application:** Any exponential default time can be simulated from Uniform(0,1) -- the basis of Monte Carlo credit risk models.

## 🧮 Theorem: Probability Integral Transform

::: {.callout-important}
## Theorem 6.3: Probability Integral Transform

Let $Y$ be a continuous random variable with CDF $F_Y(y)$.

::: {.columns}
::: {.column width="50%"}
**Part 1:** If $U = F_Y(Y)$, then $U \sim \text{Uniform}(0, 1)$.

:::

::: {.column width="50%"}
**Part 2:** Conversely, if $U \sim \text{Uniform}(0, 1)$, then $Y = F_Y^{-1}(U)$ has CDF $F_Y$.
:::
:::
:::


**Why This Matters:**

::: {.columns style="font-size: 0.75em; margin-top: -20px;"}

::: {.column width="33%"}
- **Simulation:** Generate any distribution from uniform random numbers!
:::

::: {.column width="33%"}
- **Goodness-of-fit:** Transform data to uniform for testing
:::

::: {.column width="34%"}
- **Copulas:** Model dependence using uniform marginals
:::
:::

. . .

**Finance Application:** Generate correlated asset returns by transforming correlated uniforms.

## 📌 Example 2: Log-Normal Distribution {.smaller}

**Problem:** If $Y \sim N(\mu, \sigma^2)$, find the distribution of $U = e^Y$.

**Solution:**

. . .

**Step 1:** $g(y) = e^y$ is strictly increasing. Range: $U > 0$.

. . .

**Step 2:** Inverse: $y = \ln(u) = h(u)$

. . .

**Step 3:** Jacobian: $h'(u) = \frac{1}{u}$

. . .

**Step 4:** Apply formula:
$$f_U(u) = \frac{1}{\sigma\sqrt{2\pi}} \exp\left[-\frac{(\ln u - \mu)^2}{2\sigma^2}\right] \cdot \frac{1}{u}$$

. . .

$$= \frac{1}{u\sigma\sqrt{2\pi}} \exp\left[-\frac{(\ln u - \mu)^2}{2\sigma^2}\right] \quad \text{for } u > 0$$

This is the **Log-Normal distribution**: $U \sim \text{LogNormal}(\mu, \sigma^2)$.

## 📖 The Log-Normal in Finance 

::: {.callout-note}
## 📝 Why Log-Normal for Stock Prices?

If log-returns are normal: $r_t = \ln(P_t/P_{t-1}) \sim N(\mu, \sigma^2)$

Then: $\ln(P_t) = \ln(P_0) + \sum_{i=1}^{t} r_i$

By CLT, $\ln(P_t)$ is approximately normal, so $P_t = e^{\ln(P_t)}$ is **log-normal**.
:::

::: {.columns style="font-size: 0.75em; margin-top: -20px;"}
::: {.column width="50%"}
**Properties of Log-Normal:**

- Always positive (prices can't be negative!)
- Right-skewed
- Mean: $E[U] = e^{\mu + \sigma^2/2}$
- Variance: $V(U) = e^{2\mu + \sigma^2}(e^{\sigma^2} - 1)$
:::

::: {.column width="50%"}
**Black-Scholes Model:**

Stock price follows geometric Brownian motion:
$$dS_t = \mu S_t \, dt + \sigma S_t \, dW_t$$

Solution: $S_t$ is log-normally distributed!
:::
:::

## 🎮 Interactive: Log-Normal Parameters {.smaller}

::: {style="font-size: 0.75em; margin-top: -8px;"}

**Explore:** How do $\mu$ and $\sigma$ affect the log-normal stock price distribution?

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof mu_ln = {
  const input = Inputs.range([-1, 2], {
    value: 0,
    step: 0.1,
    label: "μ (log-scale mean):"
  });

  ['pointerdown', 'touchstart', 'mousedown', 'click', 'wheel', 'pointermove', 'touchmove', 'mousemove'].forEach(e =>
    input.addEventListener(e, ev => ev.stopPropagation())
  );
  return input;
}

viewof sigma_ln = {
  const input = Inputs.range([0.1, 1.5], {
    value: 0.5,
    step: 0.1,
    label: "σ (log-scale std):"
  });

  ['pointerdown', 'touchstart', 'mousedown', 'click', 'wheel', 'pointermove', 'touchmove', 'mousemove'].forEach(e =>
    input.addEventListener(e, ev => ev.stopPropagation())
  );
  return input;
}

// Log-normal statistics
ln_mean = Math.exp(mu_ln + sigma_ln * sigma_ln / 2)
ln_median = Math.exp(mu_ln)
ln_mode = Math.exp(mu_ln - sigma_ln * sigma_ln)
```

**Mean:** ${ln_mean.toFixed(3)}

**Median:** ${ln_median.toFixed(3)}

**Mode:** ${ln_mode.toFixed(3)}

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// Generate log-normal PDF
lnPdf = {
  const points = [];
  for (let x = 0.01; x <= 8; x += 0.05) {
    const logx = Math.log(x);
    const pdf = Math.exp(-(logx - mu_ln) * (logx - mu_ln) / (2 * sigma_ln * sigma_ln)) / (x * sigma_ln * Math.sqrt(2 * Math.PI));
    points.push({x: x, pdf: pdf});
  }
  return points;
}

Plot.plot({
  width: 620,
  height: 400,
  x: { domain: [0, 8], label: "U = eʸ" },
  y: { domain: [0, 2], label: "Density" },
  marks: [
    Plot.line(lnPdf, {x: "x", y: "pdf", stroke: "steelblue", strokeWidth: 2}),
    Plot.ruleX([ln_mean], {stroke: "red", strokeDasharray: "5,5"}),
    Plot.ruleX([ln_median], {stroke: "green", strokeDasharray: "3,3"}),
    Plot.ruleY([0])
  ]
})
```

**Red dashed:** Mean | **Green dashed:** Median

:::
:::
:::

## 📖 Method of Moment-Generating Functions 

::: {.callout-note}
## 📝 Theorem 6.4: MGF Uniqueness

If two random variables have the **same moment-generating function** in a neighborhood of $t = 0$, they have the **same distribution**.

**Method:**
1. Find $M_U(t) = E[e^{tU}]$ for the transformed variable $U$
2. Recognize $M_U(t)$ as the MGF of a known distribution
3. Conclude $U$ has that distribution
:::

**When to Use MGF Method:**

- Finding distribution of **sums** of independent RVs
- When the MGF has a recognizable form
- Avoiding complex integration

## 🧮 Key MGF Property: Sums of Independent RVs 

::: {.callout-important}
## Theorem 6.5: MGF of Sums

If $Y_1, Y_2, \ldots, Y_n$ are **independent** random variables, then:

$$M_{Y_1 + Y_2 + \cdots + Y_n}(t) = M_{Y_1}(t) \cdot M_{Y_2}(t) \cdots M_{Y_n}(t)$$

**Proof:** 
$$M_{\sum Y_i}(t) = E[e^{t\sum Y_i}] = E[\prod e^{tY_i}] = \prod E[e^{tY_i}] = \prod M_{Y_i}(t)$$

The last equality uses **independence**.
:::

**Power of This Result:** Instead of complex convolution integrals, just multiply MGFs!

## 📌 Example 3: Two-Asset Portfolio Return {.smaller}

**Problem:** A portfolio holds two uncorrelated assets with returns $Y_1 \sim N(\mu_1, \sigma_1^2)$ and $Y_2 \sim N(\mu_2, \sigma_2^2)$, independent. Find the distribution of the total return $U = Y_1 + Y_2$.

**Solution:**

. . .

**Step 1:** Recall the normal MGF: $M_Y(t) = e^{\mu t + \sigma^2 t^2/2}$

. . .

**Step 2:** Compute product of MGFs:
$$M_U(t) = M_{Y_1}(t) \cdot M_{Y_2}(t) = e^{\mu_1 t + \sigma_1^2 t^2/2} \cdot e^{\mu_2 t + \sigma_2^2 t^2/2}$$

. . .

$$= e^{(\mu_1 + \mu_2)t + (\sigma_1^2 + \sigma_2^2)t^2/2}$$

. . .

**Step 3:** Recognize this as the MGF of $N(\mu_1 + \mu_2, \sigma_1^2 + \sigma_2^2)$

$$\boxed{Y_1 + Y_2 \sim N(\mu_1 + \mu_2, \sigma_1^2 + \sigma_2^2)}$$

**Portfolio Insight:** Variances **add** for uncorrelated assets -- this is why diversification across truly independent markets reduces risk most effectively!

## 📌 Example 4: Total Insurance Claim Processing Time {.smaller}

**Problem:** An insurance company processes $n$ independent claims, each taking $Y_i \sim \text{Exponential}(\beta)$ days. Find the distribution of total processing time $U = \sum_{i=1}^n Y_i$.

**Solution:**

. . .

**Step 1:** Exponential MGF: $M_Y(t) = (1 - \beta t)^{-1}$ for $t < 1/\beta$

. . .

**Step 2:** Product of MGFs:
$$M_U(t) = \prod_{i=1}^n M_{Y_i}(t) = [(1 - \beta t)^{-1}]^n = (1 - \beta t)^{-n}$$

. . .

**Step 3:** Recognize as Gamma MGF with $\alpha = n$:

$$\boxed{U = \sum_{i=1}^n Y_i \sim \text{Gamma}(n, \beta)}$$

**Insurance Insight:** Total time to process $n$ claims follows a Gamma distribution -- actuaries use Gamma quantiles to plan staffing and reserves.

## 📌 Example 5: Testing Portfolio Variance {.smaller}

**Problem:** A risk analyst standardizes $n$ independent asset returns to get $Z_i \sim N(0,1)$. To test portfolio variance, they compute $U = \sum_{i=1}^n Z_i^2$. What is the distribution of $U$?

**Solution:**

. . .

**Step 1:** We showed $Z^2 \sim \chi^2(1)$, which has MGF: $M_{Z^2}(t) = (1 - 2t)^{-1/2}$

. . .

**Step 2:** Product of MGFs:
$$M_U(t) = \prod_{i=1}^n (1 - 2t)^{-1/2} = (1 - 2t)^{-n/2}$$

. . .

**Step 3:** This is the MGF of $\chi^2(n)$:

$$\boxed{\sum_{i=1}^n Z_i^2 \sim \chi^2(n)}$$

::: {style="font-size: 0.75em; margin-top: -20px;"}
**Risk Testing:** The $\chi^2$ distribution underlies variance tests in portfolio risk management -- risk managers use it to test if realized volatility exceeds model predictions.
:::


## 💰 Case Study: Portfolio Return Data 

::: {style="font-size:35px"}
**Scenario:** 40% AAPL, 35% MSFT, 25% GOOG (daily log returns, 2020-2024)

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "📊 Show Code"
#| message: false
#| warning: false
#| eval: true

library(tidyverse)
library(tidyquant)
library(knitr)

# Download data for 3 assets
symbols <- c("AAPL", "MSFT", "GOOG")
prices <- tq_get(symbols, from = "2020-01-01", to = "2024-12-31")

# Calculate log returns
returns <- prices %>%
  group_by(symbol) %>%
  mutate(log_ret = log(adjusted / lag(adjusted))) %>%
  na.omit() %>%
  select(date, symbol, log_ret) %>%
  pivot_wider(names_from = symbol, values_from = log_ret)

# Portfolio weights
w <- c(AAPL = 0.4, MSFT = 0.35, GOOG = 0.25)

# Portfolio return (weighted sum)
returns <- returns %>%
  mutate(portfolio = w["AAPL"]*AAPL + w["MSFT"]*MSFT + w["GOOG"]*GOOG)

# Summary statistics table
summary_stats <- data.frame(
  Asset = c("AAPL", "MSFT", "GOOG", "Portfolio"),
  Weight = c("40%", "35%", "25%", "—"),
  Mean = c(mean(returns$AAPL), mean(returns$MSFT),
           mean(returns$GOOG), mean(returns$portfolio)) * 252 * 100,
  StdDev = c(sd(returns$AAPL), sd(returns$MSFT),
             sd(returns$GOOG), sd(returns$portfolio)) * sqrt(252) * 100
)

kable(summary_stats, digits = 2,
      col.names = c("Asset", "Weight", "Ann. Mean (%)", "Ann. SD (%)"),
      caption = "Daily Log Return Statistics (Annualized)")
```
:::

## 💰 Case Study: Portfolio Return Distribution {.smaller}

::: {style="font-size:30px"}
**Does the portfolio return look normal?** MGF theory predicts yes, if returns are approximately normal and independent.

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "📊 Show Code"
#| message: false
#| warning: false
#| eval: true
#| fig-width: 8
#| fig-height: 3.8
#| fig-align: center

# Plot portfolio return distribution
ggplot(returns, aes(x = portfolio)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "steelblue", alpha = 0.7) +
  stat_function(fun = dnorm,
                args = list(mean = mean(returns$portfolio),
                           sd = sd(returns$portfolio)),
                color = "red", linewidth = 1.2) +
  labs(title = "Portfolio Return Distribution",
       subtitle = "40% AAPL, 35% MSFT, 25% GOOG | Red = Normal fit",
       x = "Daily Log Return", y = "Density") +
  theme_minimal(base_size = 14)
```
:::

## 💰 Case Study: Correlation & Portfolio Math 

::: {style="font-size:38px;"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "📊 Show Analysis Code"
#| message: false
#| warning: false
#| eval: true

# Correlation matrix
cor_matrix <- cor(returns %>% select(AAPL, MSFT, GOOG))
kable(cor_matrix, digits = 3, caption = "Correlation Matrix")
```

:::

::: {.column width="50%"}
::: {.fragment}
**The Math (from MGF perspective):**

If returns were independent:
$$V(R_p) = w_1^2\sigma_1^2 + w_2^2\sigma_2^2 + w_3^2\sigma_3^2$$

With correlation:
$$V(R_p) = \sum_i w_i^2\sigma_i^2 + 2\sum_{i<j}w_i w_j \rho_{ij}\sigma_i\sigma_j$$


:::
:::
:::
:::

. . .

Since $\rho_{ij} < 1$, portfolio variance is **less than** the case with perfect correlation!

## 💰 Case Study: Diversification Benefit {.smaller}

::: {style="font-size:50px;"}

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "📊 Show Analysis Code"
#| message: false
#| warning: false
#| eval: true

# Weighted average of individual SDs
weighted_avg_sd <- sum(w * c(sd(returns$AAPL),
                              sd(returns$MSFT),
                              sd(returns$GOOG)))

# Actual portfolio SD
portfolio_sd <- sd(returns$portfolio)

# Risk reduction table
risk_comparison <- data.frame(
  Metric = c("Weighted Avg SD", "Actual Portfolio SD", "Risk Reduction"),
  Value = c(sprintf("%.4f", weighted_avg_sd),
            sprintf("%.4f", portfolio_sd),
            sprintf("%.1f%%", 100 * (1 - portfolio_sd/weighted_avg_sd)))
)

kable(risk_comparison, caption = "Diversification Benefit")
```

**This is the mathematical foundation of diversification!**
:::

## 💰 Case Study: Key Findings

::: {style="font-size: 47px"}
::: {.callout-important}
## 📊 Analysis Results

::: {.columns}

::: {.column width="33%"}
::: {.fragment}
**MGF Insight:**

- Portfolio return = weighted sum of individual returns

- If returns were normal and independent, portfolio would be normal

- MGF method: multiply individual MGFs
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Reality Check:**

- Returns are **not independent** (correlated)

- Correlations < 1 enable diversification

- Portfolio SD significantly less than weighted average
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Practical Implications:**

1. **Diversification works** because correlations < 1

2. **Risk reduction** is quantifiable using variance formulas

3. **Portfolio optimization** balances return vs. risk
:::
:::

:::
:::
:::

## 🤝 Think-Pair-Share: Choosing the Right

::::: {style="font-size: 40px;"}
:::: callout-note
## 💭 Activity (4 minutes)

A hedge fund analyst needs to find the distribution of a weighted portfolio return $R_p = 0.5R_1 + 0.3R_2 + 0.2R_3$ where each $R_i \sim N(\mu_i, \sigma_i^2)$ independently.

::: columns
::: {.column width="33%"}
**🧠 Think (1 min):**

- Which method is most efficient: distribution function, transformation, or MGF?
- What makes this problem easy for MGFs?
:::

::: {.column width="33%"}
**👫 Pair (2 min):**

- Compare your method choices
- Work out $M_{R_p}(t)$ together
- What distribution do you recognize?
:::

::: {.column width="33%"}
**🗣️ Share (1 min):**

- What is the final distribution of $R_p$?
- How would correlation change your approach?
:::
:::
::::
:::::

```{r}
#| echo: false
library(countdown)
countdown(minutes = 4, seconds = 0,
          top = 1, right = 0,
          font_size = "1.5em",
          color_running_background = "#31b09e",
          color_running_text = "white",
          color_finished_background = "#cc3311",
          color_finished_text = "white",
          color_warning_background = "#f7dc6f",
          warn_when = 60,
          play_sound = TRUE)
```

## 📝 Quiz #1: Transformation Method {.smaller .quiz-question}

A risk analyst models a stock return $Y$ with pdf $f_Y(y)$. To convert to a price ratio $U = e^Y$ (strictly increasing, inverse $h(u) = \ln(u)$), the pdf of $U$ is:

- [$f_U(u) = f_Y(\ln(u)) \cdot \frac{1}{u}$]{.correct data-explanation="✅ Correct! The transformation formula gives f_U(u) = f_Y(h(u)) * |h'(u)|. Here h(u) = ln(u) and h'(u) = 1/u, so f_U(u) = f_Y(ln(u)) / u. This is exactly how the log-normal distribution arises!"}
- $f_U(u) = f_Y(e^u) \cdot e^u$
- $f_U(u) = f_Y(\ln(u))$
- $f_U(u) = f_Y(\ln(u)) \cdot u$

## 📝 Quiz #2: Stock Price Distribution {.smaller .quiz-question}

If daily log-returns of a stock follow $Y \sim N(\mu, \sigma^2)$, and the price ratio is $S/S_0 = e^Y$, what distribution does $S/S_0$ follow?

- [Log-Normal$(\mu, \sigma^2)$]{.correct data-explanation="✅ Correct! Since S/S₀ = eʸ and Y is normal, the ratio follows a log-normal distribution. This is the theoretical foundation of the Black-Scholes model for stock prices."}
- Normal$(\mu, \sigma^2)$
- Exponential$(e^\mu)$
- Gamma$(\mu, \sigma^2)$

## 📝 Quiz #3: Portfolio Return MGF {.smaller .quiz-question}

Two uncorrelated assets have return MGFs $M_1(t)$ and $M_2(t)$. Total portfolio return is $R = R_1 + R_2$. What is $M_R(t)$?

- [$M_1(t) \cdot M_2(t)$]{.correct data-explanation="✅ Correct! For independent random variables, the MGF of the sum equals the product of individual MGFs. This makes analyzing portfolio return distributions remarkably simple when assets are uncorrelated."}
- $M_1(t) + M_2(t)$
- $M_1(t) / M_2(t)$
- $M_1(M_2(t))$

## 📝 Quiz #4: Insurance Claim Aggregation {.smaller .quiz-question}

An insurer processes 5 independent claims, each with processing time Exponential$(\beta)$. Using the MGF method, the distribution of total processing time is:

- [Gamma$(5, \beta)$]{.correct data-explanation="✅ Correct! Each exponential MGF is (1-βt)⁻¹. Multiplying 5 gives (1-βt)⁻⁵, the MGF of Gamma(5, β). Actuaries use this to plan staffing for claim processing."}
- Exponential$(5\beta)$
- Normal$(5\beta, 5\beta^2)$
- Chi-square$(5)$

## 📝 Summary {.smaller}

::: {style="font-size: 35px"}
::: {.summary-box}
**✅ Key Takeaways**

- **Transformation method**: $f_U(u) = f_Y(h(u)) \cdot |h'(u)|$ for monotonic $g$ with inverse $h$ — the Jacobian accounts for stretching/compressing

- **Probability integral transform**: $F_Y(Y) \sim \text{Uniform}(0,1)$ — foundation for Monte Carlo simulation

- **Log-normal distribution**: $e^Y$ where $Y \sim N(\mu, \sigma^2)$ — models stock prices and other positive quantities

- **MGF method**: For sums of independent RVs, $M_{\sum Y_i}(t) = \prod M_{Y_i}(t)$

- **Key results**: Normals sum to normal, exponentials to gamma, $\chi^2(1)$'s to $\chi^2(n)$ — proven elegantly via MGFs
:::
:::

## 📚 Practice Problems

::: {.callout-tip}
## 📝 Homework Problems

**Problem 1 (Transformation -- Risk):** An insurance loss is modeled as $Y \sim \text{Exponential}(2)$. Use the transformation method to find the pdf of the cube-root loss severity $U = Y^{1/3}$.

**Problem 2 (Log-Normal -- Stock Pricing):** A stock price follows $S \sim \text{LogNormal}(4, 0.09)$. Find $E[S]$ and $P(S > 60)$.

**Problem 3 (MGF -- Claim Counts):** Two branch offices have independent claim counts $Y_1 \sim \text{Poisson}(\lambda_1)$ and $Y_2 \sim \text{Poisson}(\lambda_2)$. Use MGFs to show total claims $Y_1 + Y_2 \sim \text{Poisson}(\lambda_1 + \lambda_2)$.

**Problem 4 (Chi-Square -- Variance Test):** A risk analyst standardizes 10 independent daily returns to get $Z_i \sim N(0,1)$. Find $P(\sum Z_i^2 > 18.31)$ to test if realized variance is unexpectedly high.

**Problem 5 (Portfolio):** A pension fund holds two uncorrelated asset classes with returns $N(0.08, 0.04)$ and $N(0.05, 0.01)$. Find the distribution of the 60-40 portfolio return using the MGF method.
:::

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business

ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Multivariate Transformations and Order Statistics

**Reading:** Chapter 6, Sections 6.6-6.7

**Preparation:** Review partial derivatives and Jacobian determinants

**⏰ Reminders:**

✅ Complete Practice Problems 1-5

✅ Review matrix determinants

✅ Think about min/max of random samples

✅ Work hard!
:::
:::

## ❓ Questions? {.center}

::: {.callout-note}
## 💬 Open Discussion

**Key Topics for Discussion:**

- When is the MGF method preferable to the distribution function method?

- Why do stock prices follow log-normal rather than normal distributions?

- How does the probability integral transform enable Monte Carlo simulation?

- What happens when random variables are not independent?
:::
