---
title: "Mathematical Statistics"
subtitle: "Method of Transformations and Moment-Generating Functions"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics - Transformations and MGF Method"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::: {.learning-objectives}
By the end of this lecture, you will be able to:

- Apply the **method of transformations** (change of variables) to find pdfs of transformed random variables

- Use the **Jacobian** for the univariate transformation formula

- Apply the **moment-generating function method** to identify distributions of sums

- Recognize when the MGF method is **more efficient** than direct integration

- Apply these techniques to derive distributions of **portfolio returns** and **risk measures**
:::

## 📋 Overview {.smaller}

::: {style="font-size:38px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Method of Transformations** – Direct formula for monotonic functions

- **The Jacobian in 1D** – Why $|dy/du|$ appears in the formula

- **MGF Method** – Using moment-generating functions to identify distributions

- **Sums of Random Variables** – Convolutions and the MGF approach

- **Case Study** – Distribution of portfolio returns using MGF
:::
:::
:::

## 📖 Recall: Distribution Function Method {.smaller}

::: {.callout-tip}
## 🔗 From Last Lecture

The distribution function method works for **any** transformation:

1. Find $F_U(u) = P(U \leq u) = P(g(Y) \leq u)$
2. Express in terms of $Y$
3. Differentiate to get $f_U(u)$

**Limitation:** Can be tedious when the function is monotonic and differentiable.
:::

**Today's Question:** Can we get a **direct formula** for $f_U(u)$ without going through the CDF?

**Answer:** Yes! The method of transformations provides exactly this shortcut.

## 📖 Definition: Method of Transformations {.smaller}

::: {.callout-note}
## 📝 Theorem 6.2: Univariate Transformation

Let $Y$ be a continuous random variable with pdf $f_Y(y)$ on support $(a, b)$.

Let $U = g(Y)$ where $g$ is **strictly monotonic** and differentiable.

Let $h = g^{-1}$ be the inverse function, so $Y = h(U)$.

Then the pdf of $U$ is:

$$f_U(u) = f_Y(h(u)) \cdot \left| \frac{dh}{du} \right| = f_Y(h(u)) \cdot |h'(u)|$$

for $u$ in the range of $g$.

The term $|h'(u)| = |dy/du|$ is called the **Jacobian** of the transformation.
:::

## 🧮 Why Does the Jacobian Appear? {.smaller}

::: {.callout-important}
## Intuition: Stretching and Compressing

Consider a small interval $[y, y + dy]$ that maps to $[u, u + du]$.

**Probability is conserved:**
$$f_Y(y) \, dy = f_U(u) \, du$$

Solving for $f_U(u)$:
$$f_U(u) = f_Y(y) \cdot \frac{dy}{du} = f_Y(h(u)) \cdot |h'(u)|$$

The absolute value ensures the density is positive!
:::

::: {.columns}
::: {.column width="50%"}
**If $g$ is increasing:** $h'(u) > 0$

- Orientation preserved
- $|h'(u)| = h'(u)$
:::

::: {.column width="50%"}
**If $g$ is decreasing:** $h'(u) < 0$

- Orientation reversed
- $|h'(u)| = -h'(u)$
:::
:::

## 📌 Example 1: Exponential to Uniform {.smaller}

**Problem:** Let $Y \sim \text{Exponential}(\beta)$. Find the distribution of $U = 1 - e^{-Y/\beta}$.

**Solution:**

. . .

**Step 1:** Identify $g(y) = 1 - e^{-y/\beta}$. This is strictly increasing on $y > 0$.

. . .

**Step 2:** Find the inverse. From $u = 1 - e^{-y/\beta}$:
$$e^{-y/\beta} = 1 - u \implies y = -\beta \ln(1-u) = h(u)$$

. . .

**Step 3:** Compute the Jacobian:
$$h'(u) = \frac{d}{du}[-\beta \ln(1-u)] = \frac{\beta}{1-u}$$

. . .

**Step 4:** Apply the formula. Since $f_Y(y) = \frac{1}{\beta}e^{-y/\beta}$:
$$f_U(u) = \frac{1}{\beta}e^{-h(u)/\beta} \cdot \frac{\beta}{1-u} = \frac{1}{\beta}(1-u) \cdot \frac{\beta}{1-u} = 1$$

for $0 < u < 1$. **This is Uniform(0,1)!**

## 🧮 Theorem: Probability Integral Transform {.smaller}

::: {.callout-important}
## Theorem 6.3: Probability Integral Transform

Let $Y$ be a continuous random variable with CDF $F_Y(y)$.

**Part 1:** If $U = F_Y(Y)$, then $U \sim \text{Uniform}(0, 1)$.

**Part 2:** Conversely, if $U \sim \text{Uniform}(0, 1)$, then $Y = F_Y^{-1}(U)$ has CDF $F_Y$.
:::

**Why This Matters:**

- **Simulation:** Generate any distribution from uniform random numbers!
- **Goodness-of-fit:** Transform data to uniform for testing
- **Copulas:** Model dependence using uniform marginals

**Finance Application:** Generate correlated asset returns by transforming correlated uniforms.

## 📌 Example 2: Log-Normal Distribution {.smaller}

**Problem:** If $Y \sim N(\mu, \sigma^2)$, find the distribution of $U = e^Y$.

**Solution:**

. . .

**Step 1:** $g(y) = e^y$ is strictly increasing. Range: $U > 0$.

. . .

**Step 2:** Inverse: $y = \ln(u) = h(u)$

. . .

**Step 3:** Jacobian: $h'(u) = \frac{1}{u}$

. . .

**Step 4:** Apply formula:
$$f_U(u) = \frac{1}{\sigma\sqrt{2\pi}} \exp\left[-\frac{(\ln u - \mu)^2}{2\sigma^2}\right] \cdot \frac{1}{u}$$

. . .

$$= \frac{1}{u\sigma\sqrt{2\pi}} \exp\left[-\frac{(\ln u - \mu)^2}{2\sigma^2}\right] \quad \text{for } u > 0$$

This is the **Log-Normal distribution**: $U \sim \text{LogNormal}(\mu, \sigma^2)$.

## 📖 The Log-Normal in Finance {.smaller}

::: {.callout-note}
## 📝 Why Log-Normal for Stock Prices?

If log-returns are normal: $r_t = \ln(P_t/P_{t-1}) \sim N(\mu, \sigma^2)$

Then: $\ln(P_t) = \ln(P_0) + \sum_{i=1}^{t} r_i$

By CLT, $\ln(P_t)$ is approximately normal, so $P_t = e^{\ln(P_t)}$ is **log-normal**.
:::

::: {.columns}
::: {.column width="50%"}
**Properties of Log-Normal:**

- Always positive (prices can't be negative!)
- Right-skewed
- Mean: $E[U] = e^{\mu + \sigma^2/2}$
- Variance: $V(U) = e^{2\mu + \sigma^2}(e^{\sigma^2} - 1)$
:::

::: {.column width="50%"}
**Black-Scholes Model:**

Stock price follows geometric Brownian motion:
$$dS_t = \mu S_t \, dt + \sigma S_t \, dW_t$$

Solution: $S_t$ is log-normally distributed!
:::
:::

## 🎮 Interactive: Log-Normal Parameters {.smaller}

::: {style="font-size: 0.8em;"}

**Explore:** How do $\mu$ and $\sigma$ affect the log-normal distribution?

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof mu_ln = Inputs.range([-1, 2], {
  value: 0, 
  step: 0.1, 
  label: "μ (log-scale mean):"
})

viewof sigma_ln = Inputs.range([0.1, 1.5], {
  value: 0.5, 
  step: 0.1, 
  label: "σ (log-scale std):"
})

// Log-normal statistics
ln_mean = Math.exp(mu_ln + sigma_ln * sigma_ln / 2)
ln_median = Math.exp(mu_ln)
ln_mode = Math.exp(mu_ln - sigma_ln * sigma_ln)
```

**Mean:** ${ln_mean.toFixed(3)}

**Median:** ${ln_median.toFixed(3)}

**Mode:** ${ln_mode.toFixed(3)}

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// Generate log-normal PDF
lnPdf = {
  const points = [];
  for (let x = 0.01; x <= 8; x += 0.05) {
    const logx = Math.log(x);
    const pdf = Math.exp(-(logx - mu_ln) * (logx - mu_ln) / (2 * sigma_ln * sigma_ln)) / (x * sigma_ln * Math.sqrt(2 * Math.PI));
    points.push({x: x, pdf: pdf});
  }
  return points;
}

Plot.plot({
  width: 480,
  height: 320,
  x: { domain: [0, 8], label: "U = eʸ" },
  y: { domain: [0, 2], label: "Density" },
  marks: [
    Plot.line(lnPdf, {x: "x", y: "pdf", stroke: "steelblue", strokeWidth: 2}),
    Plot.ruleX([ln_mean], {stroke: "red", strokeDasharray: "5,5"}),
    Plot.ruleX([ln_median], {stroke: "green", strokeDasharray: "3,3"}),
    Plot.ruleY([0])
  ]
})
```

**Red dashed:** Mean | **Green dashed:** Median

:::
:::
:::

## 📖 Method of Moment-Generating Functions {.smaller}

::: {.callout-note}
## 📝 Theorem 6.4: MGF Uniqueness

If two random variables have the **same moment-generating function** in a neighborhood of $t = 0$, they have the **same distribution**.

**Method:**
1. Find $M_U(t) = E[e^{tU}]$ for the transformed variable $U$
2. Recognize $M_U(t)$ as the MGF of a known distribution
3. Conclude $U$ has that distribution
:::

**When to Use MGF Method:**

- Finding distribution of **sums** of independent RVs
- When the MGF has a recognizable form
- Avoiding complex integration

## 🧮 Key MGF Property: Sums of Independent RVs {.smaller}

::: {.callout-important}
## Theorem 6.5: MGF of Sums

If $Y_1, Y_2, \ldots, Y_n$ are **independent** random variables, then:

$$M_{Y_1 + Y_2 + \cdots + Y_n}(t) = M_{Y_1}(t) \cdot M_{Y_2}(t) \cdots M_{Y_n}(t)$$

**Proof:** 
$$M_{\sum Y_i}(t) = E[e^{t\sum Y_i}] = E[\prod e^{tY_i}] = \prod E[e^{tY_i}] = \prod M_{Y_i}(t)$$

The last equality uses **independence**.
:::

**Power of This Result:** Instead of complex convolution integrals, just multiply MGFs!

## 📌 Example 3: Sum of Independent Normals {.smaller}

**Problem:** If $Y_1 \sim N(\mu_1, \sigma_1^2)$ and $Y_2 \sim N(\mu_2, \sigma_2^2)$ are independent, find the distribution of $U = Y_1 + Y_2$.

**Solution:**

. . .

**Step 1:** Recall the normal MGF: $M_Y(t) = e^{\mu t + \sigma^2 t^2/2}$

. . .

**Step 2:** Compute product of MGFs:
$$M_U(t) = M_{Y_1}(t) \cdot M_{Y_2}(t) = e^{\mu_1 t + \sigma_1^2 t^2/2} \cdot e^{\mu_2 t + \sigma_2^2 t^2/2}$$

. . .

$$= e^{(\mu_1 + \mu_2)t + (\sigma_1^2 + \sigma_2^2)t^2/2}$$

. . .

**Step 3:** Recognize this as the MGF of $N(\mu_1 + \mu_2, \sigma_1^2 + \sigma_2^2)$

$$\boxed{Y_1 + Y_2 \sim N(\mu_1 + \mu_2, \sigma_1^2 + \sigma_2^2)}$$

**Note:** Variances **add** for independent variables!

## 📌 Example 4: Sum of Independent Exponentials {.smaller}

**Problem:** If $Y_1, Y_2, \ldots, Y_n$ are independent $\text{Exponential}(\beta)$, find the distribution of $U = \sum_{i=1}^n Y_i$.

**Solution:**

. . .

**Step 1:** Exponential MGF: $M_Y(t) = (1 - \beta t)^{-1}$ for $t < 1/\beta$

. . .

**Step 2:** Product of MGFs:
$$M_U(t) = \prod_{i=1}^n M_{Y_i}(t) = [(1 - \beta t)^{-1}]^n = (1 - \beta t)^{-n}$$

. . .

**Step 3:** Recognize as Gamma MGF with $\alpha = n$:

$$\boxed{U = \sum_{i=1}^n Y_i \sim \text{Gamma}(n, \beta)}$$

**Interpretation:** Sum of $n$ independent exponentials is Gamma!

## 📌 Example 5: Chi-Square as Sum of Squared Normals {.smaller}

**Problem:** If $Z_1, Z_2, \ldots, Z_n$ are independent $N(0, 1)$, what is the distribution of $U = \sum_{i=1}^n Z_i^2$?

**Solution:**

. . .

**Step 1:** We showed $Z^2 \sim \chi^2(1)$, which has MGF: $M_{Z^2}(t) = (1 - 2t)^{-1/2}$

. . .

**Step 2:** Product of MGFs:
$$M_U(t) = \prod_{i=1}^n (1 - 2t)^{-1/2} = (1 - 2t)^{-n/2}$$

. . .

**Step 3:** This is the MGF of $\chi^2(n)$:

$$\boxed{\sum_{i=1}^n Z_i^2 \sim \chi^2(n)}$$

**Connection:** $\chi^2(n) = \text{Gamma}(n/2, 2)$, matching the Gamma MGF form!

## 💰 Case Study: Portfolio Return Distribution {.smaller}

::: {style="font-size:22px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

library(tidyverse)
library(tidyquant)

# Download data for 3 assets
symbols <- c("AAPL", "MSFT", "GOOG")
prices <- tq_get(symbols, from = "2020-01-01", to = "2024-12-31")

# Calculate log returns
returns <- prices %>%
  group_by(symbol) %>%
  mutate(log_ret = log(adjusted / lag(adjusted))) %>%
  na.omit() %>%
  select(date, symbol, log_ret) %>%
  pivot_wider(names_from = symbol, values_from = log_ret)

# Portfolio weights
w <- c(AAPL = 0.4, MSFT = 0.35, GOOG = 0.25)

# Portfolio return (weighted sum)
returns <- returns %>%
  mutate(portfolio = w["AAPL"]*AAPL + w["MSFT"]*MSFT + w["GOOG"]*GOOG)

# Compare individual vs portfolio stats
cat("Individual Asset Statistics:\n")
cat(sprintf("AAPL - Mean: %.4f, SD: %.4f\n", 
            mean(returns$AAPL), sd(returns$AAPL)))
cat(sprintf("MSFT - Mean: %.4f, SD: %.4f\n", 
            mean(returns$MSFT), sd(returns$MSFT)))
cat(sprintf("GOOG - Mean: %.4f, SD: %.4f\n", 
            mean(returns$GOOG), sd(returns$GOOG)))
cat(sprintf("\nPortfolio - Mean: %.4f, SD: %.4f\n", 
            mean(returns$portfolio), sd(returns$portfolio)))
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

# Plot portfolio return distribution
ggplot(returns, aes(x = portfolio)) +
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 50, fill = "steelblue", alpha = 0.7) +
  stat_function(fun = dnorm,
                args = list(mean = mean(returns$portfolio),
                           sd = sd(returns$portfolio)),
                color = "red", linewidth = 1.2) +
  labs(title = "Portfolio Return Distribution",
       subtitle = "40% AAPL, 35% MSFT, 25% GOOG",
       x = "Daily Log Return", y = "Density") +
  theme_minimal()
```

:::
:::
:::

## 💰 Case Study: Why Portfolio SD < Weighted Average SD {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Correlation matrix
cor_matrix <- cor(returns %>% select(AAPL, MSFT, GOOG))
cat("Correlation Matrix:\n")
print(round(cor_matrix, 3))

# Weighted average of individual SDs
weighted_avg_sd <- sum(w * c(sd(returns$AAPL), 
                              sd(returns$MSFT), 
                              sd(returns$GOOG)))

# Actual portfolio SD
portfolio_sd <- sd(returns$portfolio)

cat(sprintf("\nWeighted Avg SD: %.4f\n", weighted_avg_sd))
cat(sprintf("Actual Portfolio SD: %.4f\n", portfolio_sd))
cat(sprintf("Risk Reduction: %.1f%%\n", 
            100 * (1 - portfolio_sd/weighted_avg_sd)))
```

:::

::: {.column width="50%"}
::: {.fragment}
**The Math (from MGF perspective):**

If returns were independent:
$$V(R_p) = w_1^2\sigma_1^2 + w_2^2\sigma_2^2 + w_3^2\sigma_3^2$$

With correlation:
$$V(R_p) = \sum_i w_i^2\sigma_i^2 + 2\sum_{i<j}w_i w_j \rho_{ij}\sigma_i\sigma_j$$

Since $\rho_{ij} < 1$, portfolio variance is **less than** the case with perfect correlation!

**This is the mathematical foundation of diversification!**
:::
:::
:::
:::

## 💰 Case Study: Key Findings {.smaller}

::: {style="font-size: 26px;"}
::: {.callout-important}
## 📊 Analysis Results

::: {.columns}

::: {.column width="33%"}
::: {.fragment}
**MGF Insight:**

- Portfolio return = weighted sum of individual returns

- If returns were normal and independent, portfolio would be normal

- MGF method: multiply individual MGFs
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Reality Check:**

- Returns are **not independent** (correlated)

- Correlations < 1 enable diversification

- Portfolio SD significantly less than weighted average
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Practical Implications:**

1. **Diversification works** because correlations < 1

2. **Risk reduction** is quantifiable using variance formulas

3. **Portfolio optimization** balances return vs. risk
:::
:::

:::
:::
:::

## 📝 Quiz #1: Transformation Method {.smaller .quiz-question}

If $Y$ has pdf $f_Y(y)$ and $U = g(Y)$ where $g$ is strictly increasing with inverse $h = g^{-1}$, the pdf of $U$ is:

- [$f_U(u) = f_Y(h(u)) \cdot |h'(u)|$]{.correct data-explanation="✅ Correct! The transformation formula requires multiplying the original pdf evaluated at the inverse by the absolute value of the derivative of the inverse (the Jacobian)."}
- $f_U(u) = f_Y(g(u)) \cdot |g'(u)|$
- $f_U(u) = f_Y(h(u))$
- $f_U(u) = f_Y(h(u)) / |h'(u)|$

## 📝 Quiz #2: Log-Normal Distribution {.smaller .quiz-question}

If $Y \sim N(\mu, \sigma^2)$, what is the distribution of $U = e^Y$?

- [Log-Normal$(\mu, \sigma^2)$]{.correct data-explanation="✅ Correct! The exponential of a normal random variable follows a log-normal distribution. This is the theoretical foundation for stock price models in finance."}
- Normal$(\mu, \sigma^2)$
- Exponential$(e^\mu)$
- Gamma$(\mu, \sigma^2)$

## 📝 Quiz #3: MGF of Sums {.smaller .quiz-question}

If $Y_1$ and $Y_2$ are independent with MGFs $M_1(t)$ and $M_2(t)$, what is $M_{Y_1+Y_2}(t)$?

- [$M_1(t) \cdot M_2(t)$]{.correct data-explanation="✅ Correct! For independent random variables, the MGF of the sum is the product of the individual MGFs. This property makes the MGF method powerful for analyzing sums."}
- $M_1(t) + M_2(t)$
- $M_1(t) / M_2(t)$
- $M_1(M_2(t))$

## 📝 Quiz #4: Sum of Exponentials {.smaller .quiz-question}

If $Y_1, Y_2, \ldots, Y_5$ are independent Exponential$(\beta)$ random variables, what is the distribution of their sum?

- [Gamma$(5, \beta)$]{.correct data-explanation="✅ Correct! The sum of n independent exponential(β) random variables follows a Gamma(n, β) distribution. This can be proven using the MGF method by multiplying the exponential MGFs."}
- Exponential$(5\beta)$
- Normal$(5\beta, 5\beta^2)$
- Chi-square$(5)$

## 📝 Summary {.smaller}

::: {style="font-size: 28px"}
::: {.summary-box}
**✅ Key Takeaways**

- **Transformation method**: $f_U(u) = f_Y(h(u)) \cdot |h'(u)|$ for monotonic $g$ with inverse $h$

- **Jacobian** $|h'(u)|$ accounts for stretching/compressing of probability density

- **Probability integral transform**: $F_Y(Y) \sim \text{Uniform}(0,1)$ — foundation for simulation

- **Log-normal distribution**: $e^Y$ where $Y \sim N(\mu, \sigma^2)$ — models positive quantities like stock prices

- **MGF method**: For sums of independent RVs, $M_{\sum Y_i}(t) = \prod M_{Y_i}(t)$

- **Key results**: Sum of normals is normal; sum of exponentials is gamma; sum of $\chi^2(1)$ is $\chi^2(n)$
:::
:::

## 📚 Practice Problems

::: {.callout-tip}
## 📝 Homework Problems

**Problem 1 (Transformation):** If $Y \sim \text{Exponential}(2)$, use the transformation method to find the pdf of $U = Y^{1/3}$.

**Problem 2 (Log-Normal):** If stock price $S \sim \text{LogNormal}(4, 0.09)$, find $E[S]$ and $P(S > 60)$.

**Problem 3 (MGF - Sum of Poissons):** If $Y_1 \sim \text{Poisson}(\lambda_1)$ and $Y_2 \sim \text{Poisson}(\lambda_2)$ are independent, use MGFs to show $Y_1 + Y_2 \sim \text{Poisson}(\lambda_1 + \lambda_2)$.

**Problem 4 (Chi-Square):** If $Z_1, \ldots, Z_{10}$ are independent $N(0,1)$, find $P(\sum Z_i^2 > 18.31)$.

**Problem 5 (Portfolio):** Two assets have independent returns with $N(0.08, 0.04)$ and $N(0.05, 0.01)$. Find the distribution of a 60-40 portfolio return.
:::

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business

ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Multivariate Transformations and Order Statistics

**Reading:** Chapter 6, Sections 6.6-6.7

**Preparation:** Review partial derivatives and Jacobian determinants

**⏰ Reminders:**

✅ Complete Practice Problems 1-5

✅ Review matrix determinants

✅ Think about min/max of random samples

✅ Work hard!
:::
:::

## ❓ Questions? {.center}

::: {.callout-note}
## 💬 Open Discussion

**Key Topics for Discussion:**

- When is the MGF method preferable to the distribution function method?

- Why do stock prices follow log-normal rather than normal distributions?

- How does the probability integral transform enable Monte Carlo simulation?

- What happens when random variables are not independent?
:::
