---
title: "Mathematical Statistics"
subtitle: "Functions of Random Variables: Distribution Function Method"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ../ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics - Functions of Random Variables"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::: {.learning-objectives}
By the end of this lecture, you will be able to:

- Understand **why** finding distributions of functions of random variables is essential in statistics and finance

- Apply the **distribution function method** to find the pdf of $U = g(Y)$ for univariate transformations

- Handle **non-monotonic transformations** by partitioning the domain appropriately

- Find the distribution of **functions of multiple random variables** using the CDF approach

- Apply these techniques to **financial transformations** such as returns, log-returns, and portfolio values
:::

## 📋 Overview {.smaller}

::: {style="font-size:38px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Introduction to Transformations** – Why study functions of random variables?

- **The Distribution Function Method** – Finding $F_U(u) = P(U \leq u)$ then differentiating

- **Monotonic Transformations** – Increasing and decreasing functions

- **Non-Monotonic Transformations** – Handling functions like $U = Y^2$

- **Case Study** – Stock returns as transformations of prices
:::
:::
:::

## 📖 Why Study Functions of Random Variables? {.smaller .r-fit-text}

::: {.callout-note}
## 🎯 Motivation

In practice, we often observe one random variable but need to know the distribution of a **function** of that variable:

::: {.columns}
::: {.column width="50%"}
::: {.fragment}
**Finance Applications:**

- Stock prices → Returns: $R = \frac{P_t - P_{t-1}}{P_{t-1}}$
- Prices → Log-returns: $r = \ln(P_t/P_{t-1})$
- Individual returns → Portfolio value
- Asset values → Option payoffs
:::
:::

::: {.column width="50%"}
::: {.fragment}
**Statistical Applications:**

- Sample mean $\bar{Y}$ from individual $Y_i$'s
- Sample variance $S^2$ from data
- Test statistics (t, F, χ²)
- Maximum/minimum values
:::
:::
:::
:::

**Key Question:** Given the distribution of $Y$, how do we find the distribution of $U = g(Y)$?

## 📖 Three Main Methods {.smaller}

::: {.callout-note}
## 📝 Overview of Techniques

We will study three approaches to find the distribution of $U = g(Y)$:

**1. Distribution Function Method (Today)**
$$F_U(u) = P(U \leq u) = P(g(Y) \leq u)$$
Then differentiate: $f_U(u) = \frac{d}{du}F_U(u)$

**2. Method of Transformations (Next Lecture)**
Direct formula using the inverse function and its derivative

**3. Moment-Generating Function Method (Next Lecture)**
Find $M_U(t)$ and identify the distribution
:::

**Key Insight:** The distribution function method is the most **general** approach—it always works!

## 📖 Definition: The Distribution Function Method {.smaller}

::: {.callout-note}
## 📝 Method of Distribution Functions

To find the probability distribution of $U = g(Y)$:

**Step 1:** Find the CDF of $U$:
$$F_U(u) = P(U \leq u) = P(g(Y) \leq u)$$

**Step 2:** Express $P(g(Y) \leq u)$ in terms of events involving $Y$

**Step 3:** Use the known distribution of $Y$ to compute the probability

**Step 4:** Differentiate to get the pdf:
$$f_U(u) = \frac{d}{du}F_U(u)$$
:::

**Principle:** Convert the problem about $U$ into a problem about $Y$, which we already understand!

## 📌 Example 1: Linear Transformation {.smaller}

**Problem:** Let $Y$ have pdf $f_Y(y) = 2y$ for $0 < y < 1$. Find the distribution of $U = 3Y + 1$.

**Solution:**

. . .

**Step 1:** Find the range of $U$. Since $0 < Y < 1$, we have $1 < U < 4$.

. . .

**Step 2:** Find $F_U(u) = P(U \leq u) = P(3Y + 1 \leq u) = P(Y \leq \frac{u-1}{3})$

. . .

**Step 3:** Compute using $F_Y$:
$$F_U(u) = \int_0^{(u-1)/3} 2y \, dy = y^2 \Big|_0^{(u-1)/3} = \frac{(u-1)^2}{9}$$

. . .

**Step 4:** Differentiate:
$$f_U(u) = \frac{d}{du}\left[\frac{(u-1)^2}{9}\right] = \frac{2(u-1)}{9} \quad \text{for } 1 < u < 4$$

## 📌 Example 2: Square Transformation {.smaller}

**Problem:** Let $Y \sim N(0, 1)$. Find the distribution of $U = Y^2$.

**Solution:**

. . .

**Step 1:** Range of $U$: Since $Y \in (-\infty, \infty)$, we have $U \geq 0$.

. . .

**Step 2:** For $u > 0$:
$$F_U(u) = P(Y^2 \leq u) = P(-\sqrt{u} \leq Y \leq \sqrt{u})$$

. . .

**Step 3:** Using standard normal CDF $\Phi$:
$$F_U(u) = \Phi(\sqrt{u}) - \Phi(-\sqrt{u}) = 2\Phi(\sqrt{u}) - 1$$

. . .

**Step 4:** Differentiate using chain rule:
$$f_U(u) = 2\phi(\sqrt{u}) \cdot \frac{1}{2\sqrt{u}} = \frac{1}{\sqrt{2\pi u}}e^{-u/2}$$

This is the **Chi-square distribution with 1 degree of freedom**: $U \sim \chi^2(1)$!

## 🧮 Theorem: Chi-Square from Normal {.smaller}

::: {.callout-important}
## Theorem 6.1: Square of Standard Normal

If $Y \sim N(0, 1)$, then $U = Y^2 \sim \chi^2(1)$.

More generally, if $Y_1, Y_2, \ldots, Y_n$ are independent $N(0,1)$ random variables, then:

$$U = Y_1^2 + Y_2^2 + \cdots + Y_n^2 \sim \chi^2(n)$$

The chi-square distribution with $n$ degrees of freedom has pdf:
$$f(u) = \frac{1}{2^{n/2}\Gamma(n/2)}u^{n/2-1}e^{-u/2} \quad \text{for } u > 0$$
:::

**Financial Application:** This result is fundamental for testing portfolio variance!

## 🎮 Interactive: Transformation Visualizer {.smaller}

::: {style="font-size: 0.8em;"}

**Explore:** See how the distribution changes under $U = Y^2$

::: {.columns}

::: {.column width="35%"}

```{ojs}
//| echo: false

viewof n_samples = Inputs.range([100, 5000], {
  value: 1000, 
  step: 100, 
  label: "Sample size:"
})

// Generate standard normal samples
normalSamples = {
  const samples = [];
  for (let i = 0; i < n_samples; i++) {
    // Box-Muller transform
    const u1 = Math.random();
    const u2 = Math.random();
    const z = Math.sqrt(-2 * Math.log(u1)) * Math.cos(2 * Math.PI * u2);
    samples.push({y: z, u: z * z});
  }
  return samples;
}

// Chi-square(1) theoretical PDF
chiSqPdf = {
  const points = [];
  for (let u = 0.01; u <= 8; u += 0.05) {
    const pdf = Math.exp(-u/2) / Math.sqrt(2 * Math.PI * u);
    points.push({u: u, pdf: pdf});
  }
  return points;
}
```

**Transformation:** $U = Y^2$

**Samples:** ${n_samples}

:::

::: {.column width="65%"}

```{ojs}
//| echo: false

Plot.plot({
  width: 450,
  height: 300,
  x: { domain: [0, 8], label: "U = Y²" },
  y: { domain: [0, 1.5], label: "Density" },
  marks: [
    Plot.rectY(normalSamples, Plot.binX({y: "density"}, {x: "u", fill: "steelblue", fillOpacity: 0.6})),
    Plot.line(chiSqPdf, {x: "u", y: "pdf", stroke: "red", strokeWidth: 2}),
    Plot.ruleY([0])
  ]
})
```

**Red line:** Theoretical $\chi^2(1)$ pdf

:::
:::
:::

## 📌 Example 3: Functions of Two Variables {.smaller}

**Problem:** Let $Y_1, Y_2$ be independent Uniform(0,1). Find the distribution of $U = Y_1 + Y_2$.

**Solution:**

. . .

**Step 1:** Range of $U$: Since $0 < Y_1, Y_2 < 1$, we have $0 < U < 2$.

. . .

**Step 2:** For $0 < u < 1$:
$$F_U(u) = P(Y_1 + Y_2 \leq u) = \int_0^u \int_0^{u-y_1} 1 \, dy_2 \, dy_1 = \int_0^u (u - y_1) \, dy_1 = \frac{u^2}{2}$$

. . .

For $1 \leq u < 2$: (integration region changes!)
$$F_U(u) = 1 - \frac{(2-u)^2}{2}$$

. . .

**Step 3:** Differentiate:
$$f_U(u) = \begin{cases} u & 0 < u < 1 \\ 2 - u & 1 \leq u < 2 \end{cases}$$

This is the **triangular distribution**!

## 📖 Visualizing the Sum of Uniforms {.smaller}

::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 5
#| fig-height: 4

library(tidyverse)

# Generate sum of two uniforms
set.seed(42)
n <- 10000
y1 <- runif(n)
y2 <- runif(n)
u <- y1 + y2

# Theoretical triangular PDF
u_theory <- seq(0, 2, 0.01)
f_theory <- ifelse(u_theory < 1, u_theory, 2 - u_theory)

# Plot
ggplot() +
  geom_histogram(aes(x = u, y = after_stat(density)), 
                 bins = 50, fill = "steelblue", alpha = 0.6) +
  geom_line(aes(x = u_theory, y = f_theory), 
            color = "red", linewidth = 1.2) +
  labs(title = "Distribution of U = Y₁ + Y₂",
       subtitle = "Sum of two Uniform(0,1) variables",
       x = "U", y = "Density") +
  theme_minimal()
```

:::

::: {.column width="50%"}
::: {.fragment}
**Key Observations:**

- Sum of uniforms is **NOT** uniform!

- The triangular shape emerges from the **convolution** of densities

- Central Limit Theorem preview: Sum of more uniforms → approaches normal

**Financial Insight:** This is why portfolio returns don't follow the same distribution as individual asset returns!
:::
:::
:::

## 💰 Case Study: Stock Returns Distribution {.smaller}

::: {style="font-size:22px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

library(tidyverse)
library(tidyquant)

# Download Apple stock data
aapl <- tq_get("AAPL", from = "2020-01-01", to = "2024-12-31")

# Calculate different return transformations
returns <- aapl %>%
  mutate(
    # Simple return: R = (P_t - P_{t-1}) / P_{t-1}
    simple_return = (adjusted - lag(adjusted)) / lag(adjusted),
    # Log return: r = ln(P_t / P_{t-1})
    log_return = log(adjusted / lag(adjusted)),
    # Squared return (for volatility)
    sq_return = log_return^2
  ) %>%
  na.omit()

# Summary statistics
cat("Simple Returns:\n")
cat(sprintf("  Mean: %.4f\n", mean(returns$simple_return)))
cat(sprintf("  Std:  %.4f\n", sd(returns$simple_return)))

cat("\nLog Returns:\n")
cat(sprintf("  Mean: %.4f\n", mean(returns$log_return)))
cat(sprintf("  Std:  %.4f\n", sd(returns$log_return)))
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

# Compare simple vs log returns
ggplot(returns) +
  geom_histogram(aes(x = simple_return, y = after_stat(density), fill = "Simple"), 
                 alpha = 0.5, bins = 50) +
  geom_histogram(aes(x = log_return, y = after_stat(density), fill = "Log"), 
                 alpha = 0.5, bins = 50) +
  stat_function(fun = dnorm, 
                args = list(mean = mean(returns$log_return), 
                           sd = sd(returns$log_return)),
                color = "red", linewidth = 1) +
  labs(title = "AAPL Returns: Simple vs Log",
       subtitle = "Red: Normal fit to log returns",
       x = "Return", y = "Density", fill = "Type") +
  theme_minimal() +
  xlim(-0.15, 0.15)
```

:::
:::
:::

## 💰 Case Study: Why Log Returns? {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}

**The Transformation:**
$$r_t = \ln\left(\frac{P_t}{P_{t-1}}\right) = \ln(P_t) - \ln(P_{t-1})$$

If $\ln(P_t) \sim N(\mu, \sigma^2)$, then:
$$r_t \sim N(\mu_r, \sigma_r^2)$$

**Properties of Log Returns:**

1. **Additive over time**: $r_{t:t+n} = r_t + r_{t+1} + \cdots + r_{t+n-1}$

2. **Approximately normal** (by CLT for sums)

3. **Symmetric** around zero

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 5
#| fig-height: 4

# Q-Q plot for normality check
ggplot(returns, aes(sample = log_return)) +
  stat_qq(color = "steelblue") +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Log Returns vs Normal",
       subtitle = "Deviations in tails indicate fat tails",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()
```

**Observation:** Heavy tails! Real returns have more extreme events than normal predicts.

:::
:::
:::

## 💰 Case Study: Key Findings {.smaller}

::: {style="font-size: 26px;"}
::: {.callout-important}
## 📊 Analysis Results

::: {.columns}

::: {.column width="33%"}
::: {.fragment}
**Transformation Effects:**

- Simple and log returns nearly identical for small returns

- Log returns are symmetric; simple returns slightly skewed

- Transformation choice affects statistical properties
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Normality Assessment:**

- Log returns approximately normal in center

- **Fat tails**: More extreme events than normal

- Kurtosis > 3 indicates leptokurtic distribution
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Practical Implications:**

1. **Use log returns** for multi-period analysis

2. **Fat tails matter** for risk management

3. **VaR estimates** using normal assumption understate risk
:::
:::

:::
:::
:::

## 📝 Quiz #1: Distribution Function Method {.smaller .quiz-question}

To find the distribution of $U = g(Y)$ using the distribution function method, the correct first step is:

- [Find $F_U(u) = P(U \leq u) = P(g(Y) \leq u)$]{.correct data-explanation="✅ Correct! The distribution function method starts by expressing the CDF of U in terms of events involving Y, then differentiates to get the pdf."}
- Find the inverse function $g^{-1}(u)$
- Compute the moment-generating function of $U$
- Take the derivative of $g(Y)$

## 📝 Quiz #2: Square Transformation {.smaller .quiz-question}

If $Y \sim N(0, 1)$, what is the distribution of $U = Y^2$?

- [Chi-square with 1 degree of freedom]{.correct data-explanation="✅ Correct! When you square a standard normal random variable, you get a chi-square distribution with 1 degree of freedom. This is a fundamental result used in hypothesis testing."}
- Normal with mean 0 and variance 1
- Exponential with mean 1
- Uniform on (0, 1)

## 📝 Quiz #3: Non-Monotonic Functions {.smaller .quiz-question}

When applying the distribution function method to $U = Y^2$ where $Y$ can be positive or negative:

- [We must account for both $Y = \sqrt{u}$ and $Y = -\sqrt{u}$ since $Y^2 \leq u$ means $-\sqrt{u} \leq Y \leq \sqrt{u}$]{.correct data-explanation="✅ Correct! Since squaring is not monotonic, we need to consider all values of Y that satisfy the inequality, which includes both positive and negative square roots."}
- We only need the positive root $Y = \sqrt{u}$
- The method cannot be applied to non-monotonic functions
- We average the results from both roots

## 📝 Quiz #4: Sum of Uniforms {.smaller .quiz-question}

If $Y_1, Y_2$ are independent Uniform(0,1) random variables, what type of distribution does $U = Y_1 + Y_2$ have?

- [Triangular distribution on (0, 2)]{.correct data-explanation="✅ Correct! The sum of two independent Uniform(0,1) variables follows a triangular distribution, not a uniform distribution. The pdf is f(u) = u for 0 < u < 1 and f(u) = 2 - u for 1 ≤ u < 2."}
- Uniform distribution on (0, 2)
- Normal distribution
- Exponential distribution

## 📝 Summary {.smaller}

::: {style="font-size: 28px"}
::: {.summary-box}
**✅ Key Takeaways**

- **Functions of random variables** arise constantly in statistics and finance (returns, test statistics, portfolio values)

- The **distribution function method** is universal: Find $F_U(u) = P(g(Y) \leq u)$, then differentiate

- For **monotonic functions**, the inequality $g(Y) \leq u$ simplifies directly

- For **non-monotonic functions** like $Y^2$, partition the domain and sum probabilities

- **Chi-square distribution**: $Y^2 \sim \chi^2(1)$ when $Y \sim N(0,1)$ — fundamental for inference

- **Financial insight**: Log returns are preferred because they're additive and approximately normal
:::
:::

## 📚 Practice Problems

::: {.callout-tip}
## 📝 Homework Problems

**Problem 1 (Linear Transformation):** If $Y$ has pdf $f(y) = 3y^2$ for $0 < y < 1$, find the pdf of $U = 2Y - 1$.

**Problem 2 (Square Root):** If $Y \sim \text{Uniform}(0, 4)$, find the distribution of $U = \sqrt{Y}$.

**Problem 3 (Absolute Value):** If $Y \sim N(0, 1)$, find the pdf of $U = |Y|$. (This is the half-normal distribution.)

**Problem 4 (Financial Application):** If log-price $\ln(P_t) \sim N(4, 0.04)$, find the distribution of the price $P_t$.

**Problem 5 (Sum):** If $Y_1, Y_2, Y_3$ are independent Exponential(1), find $F_U(u)$ for $U = \min(Y_1, Y_2, Y_3)$.
:::

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business

ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Method of Transformations & MGF Method

**Reading:** Chapter 6, Sections 6.4-6.5

**Preparation:** Review derivatives and inverse functions

**⏰ Reminders:**

✅ Complete Practice Problems 1-5

✅ Review the chain rule from calculus

✅ Think about when transformations are invertible

✅ Work hard!
:::
:::

## ❓ Questions? {.center}

::: {.callout-note}
## 💬 Open Discussion

**Key Topics for Discussion:**

- Why is the distribution function method considered the most general approach?

- How do fat tails in return distributions affect risk management decisions?

- What are some other financial quantities that are functions of random variables?

- When might we prefer simple returns over log returns?
:::
