---
title: "Mathematical Statistics"
subtitle: "Multivariate Transformations and Order Statistics"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics - Jacobians and Order Statistics"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::: {.learning-objectives}
By the end of this lecture, you will be able to:

- Apply **multivariate transformations** using the Jacobian determinant to find joint distributions

- Derive the distribution of **sums and differences** of random variables using Jacobians

- Define and compute the distribution of **order statistics** $Y_{(1)}, Y_{(2)}, \ldots, Y_{(n)}$

- Find the distribution of the **sample range** and other functions of order statistics

- Apply order statistics to **Value-at-Risk (VaR)** and extreme value analysis in finance
:::

## 📋 Overview {.smaller}

::: {style="font-size:38px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Multivariate Jacobian Transformations** – Extending the change-of-variables technique to 2+ dimensions

- **The Jacobian Determinant** – Computing $|J|$ for bivariate transformations

- **Order Statistics** – Distributions of sorted sample values

- **Extreme Order Statistics** – $Y_{(1)} = \min$ and $Y_{(n)} = \max$

- **Case Study** – Value-at-Risk using order statistics
:::
:::
:::

## 📖 Motivation: Multivariate Transformations {.smaller}

::: {.callout-note}
## 🎯 Why Study Multivariate Transformations?

Many important quantities involve **functions of multiple random variables**:

::: {.columns}
::: {.column width="50%"}
::: {.fragment}
**Statistical Applications:**

- Sum $U = Y_1 + Y_2$ (sample total)
- Ratio $U = Y_1/Y_2$ (F-statistic)
- Sample mean and variance jointly
- Regression coefficients
:::
:::

::: {.column width="50%"}
::: {.fragment}
**Finance Applications:**

- Portfolio return = weighted sum
- Sharpe ratio = return/volatility
- Hedge ratio = covariance/variance
- Option payoffs involving multiple assets
:::
:::
:::
:::

**Key Question:** Given joint distribution of $(Y_1, Y_2)$, find joint distribution of $(U_1, U_2) = (g_1(Y_1, Y_2), g_2(Y_1, Y_2))$.

## 📖 Definition: Bivariate Jacobian Transformation {.smaller}

::: {.callout-note}
## 📝 Theorem 6.6: Bivariate Transformation

Let $(Y_1, Y_2)$ have joint pdf $f_{Y_1, Y_2}(y_1, y_2)$.

Define transformations: $U_1 = g_1(Y_1, Y_2)$ and $U_2 = g_2(Y_1, Y_2)$

Let the inverse be: $Y_1 = h_1(U_1, U_2)$ and $Y_2 = h_2(U_1, U_2)$

The **Jacobian** is:
$$J = \begin{vmatrix} \frac{\partial y_1}{\partial u_1} & \frac{\partial y_1}{\partial u_2} \\ \frac{\partial y_2}{\partial u_1} & \frac{\partial y_2}{\partial u_2} \end{vmatrix} = \frac{\partial y_1}{\partial u_1}\frac{\partial y_2}{\partial u_2} - \frac{\partial y_1}{\partial u_2}\frac{\partial y_2}{\partial u_1}$$

Then:
$$f_{U_1, U_2}(u_1, u_2) = f_{Y_1, Y_2}(h_1(u_1, u_2), h_2(u_1, u_2)) \cdot |J|$$
:::

## 📌 Example 1: Sum and Difference {.smaller}

**Problem:** Let $Y_1, Y_2$ be independent $N(0, 1)$. Find the joint distribution of $U_1 = Y_1 + Y_2$ and $U_2 = Y_1 - Y_2$.

**Solution:**

. . .

**Step 1:** Find inverse transformation:
$$y_1 = \frac{u_1 + u_2}{2}, \quad y_2 = \frac{u_1 - u_2}{2}$$

. . .

**Step 2:** Compute Jacobian:
$$J = \begin{vmatrix} \frac{1}{2} & \frac{1}{2} \\ \frac{1}{2} & -\frac{1}{2} \end{vmatrix} = \frac{1}{2} \cdot \left(-\frac{1}{2}\right) - \frac{1}{2} \cdot \frac{1}{2} = -\frac{1}{2}$$

So $|J| = \frac{1}{2}$

. . .

**Step 3:** Apply formula. Since $f_{Y_1,Y_2}(y_1, y_2) = \frac{1}{2\pi}e^{-(y_1^2 + y_2^2)/2}$:

$$f_{U_1,U_2}(u_1, u_2) = \frac{1}{2\pi}\exp\left[-\frac{(u_1+u_2)^2/4 + (u_1-u_2)^2/4}{2}\right] \cdot \frac{1}{2}$$

## 📌 Example 1: Sum and Difference (cont.) {.smaller}

**Simplifying the exponent:**

. . .

$$(u_1+u_2)^2 + (u_1-u_2)^2 = u_1^2 + 2u_1u_2 + u_2^2 + u_1^2 - 2u_1u_2 + u_2^2 = 2u_1^2 + 2u_2^2$$

. . .

So:
$$f_{U_1,U_2}(u_1, u_2) = \frac{1}{4\pi}\exp\left[-\frac{u_1^2 + u_2^2}{4}\right]$$

. . .

$$= \frac{1}{2\sqrt{\pi}}e^{-u_1^2/4} \cdot \frac{1}{2\sqrt{\pi}}e^{-u_2^2/4}$$

. . .

::: {.callout-important}
## Key Result

$U_1 = Y_1 + Y_2 \sim N(0, 2)$ and $U_2 = Y_1 - Y_2 \sim N(0, 2)$

Moreover, **$U_1$ and $U_2$ are independent!**
:::

## 📌 Example 2: Ratio of Normals {.smaller}

**Problem:** If $Z_1, Z_2$ are independent $N(0,1)$, find the distribution of $T = Z_1/\sqrt{Z_2^2/1}$.

**Solution outline:**

. . .

This is related to the **t-distribution**. Set $U_1 = Z_1$ and $U_2 = Z_2^2$.

. . .

We know $Z_2^2 \sim \chi^2(1)$, so:
$$T = \frac{Z_1}{\sqrt{Z_2^2}} = \frac{N(0,1)}{\sqrt{\chi^2(1)/1}} \sim t(1)$$

. . .

::: {.callout-important}
## Theorem 6.7: Student's t-Distribution

If $Z \sim N(0,1)$ and $W \sim \chi^2(\nu)$ are independent, then:
$$T = \frac{Z}{\sqrt{W/\nu}} \sim t(\nu)$$

The t-distribution with $\nu$ degrees of freedom has pdf:
$$f(t) = \frac{\Gamma((\nu+1)/2)}{\sqrt{\nu\pi}\Gamma(\nu/2)}\left(1 + \frac{t^2}{\nu}\right)^{-(\nu+1)/2}$$
:::

## 📖 Definition: Order Statistics {.smaller}

::: {.callout-note}
## 📝 Definition 6.2: Order Statistics

Let $Y_1, Y_2, \ldots, Y_n$ be a random sample from a distribution with pdf $f(y)$ and CDF $F(y)$.

The **order statistics** are the sample values arranged in ascending order:
$$Y_{(1)} \leq Y_{(2)} \leq \cdots \leq Y_{(n)}$$

where:
- $Y_{(1)} = \min(Y_1, \ldots, Y_n)$ is the **minimum**
- $Y_{(n)} = \max(Y_1, \ldots, Y_n)$ is the **maximum**
- $Y_{(k)}$ is the **$k$-th smallest** value
:::

**Notation:** Parentheses in subscript indicate ordered values!

## 🧮 Theorem: Distribution of Order Statistics {.smaller}

::: {.callout-important}
## Theorem 6.8: PDF of the k-th Order Statistic

The pdf of $Y_{(k)}$ is:

$$f_{Y_{(k)}}(y) = \frac{n!}{(k-1)!(n-k)!} [F(y)]^{k-1} [1-F(y)]^{n-k} f(y)$$

**Intuition:** 
- $[F(y)]^{k-1}$: probability that $k-1$ observations are less than $y$
- $[1-F(y)]^{n-k}$: probability that $n-k$ observations are greater than $y$
- $f(y)$: one observation equals $y$
- Multinomial coefficient: ways to arrange
:::

**Special Cases:**
- Minimum: $f_{Y_{(1)}}(y) = n[1-F(y)]^{n-1}f(y)$
- Maximum: $f_{Y_{(n)}}(y) = n[F(y)]^{n-1}f(y)$

## 📌 Example 3: Maximum of Uniform Sample {.smaller}

**Problem:** Let $Y_1, \ldots, Y_n$ be iid Uniform(0, 1). Find the distribution of $Y_{(n)} = \max$.

**Solution:**

. . .

For Uniform(0,1): $f(y) = 1$ and $F(y) = y$ for $0 < y < 1$.

. . .

Using the maximum formula:
$$f_{Y_{(n)}}(y) = n[F(y)]^{n-1}f(y) = n \cdot y^{n-1} \cdot 1 = ny^{n-1}$$

for $0 < y < 1$.

. . .

::: {.columns}
::: {.column width="50%"}
**Properties:**

- $E[Y_{(n)}] = \frac{n}{n+1}$
- As $n \to \infty$, $Y_{(n)} \to 1$
- This is Beta$(n, 1)$ distribution!
:::

::: {.column width="50%"}
**Financial Application:**

Best return in a sample of $n$ trading days — useful for performance attribution!
:::
:::

## 🎮 Interactive: Order Statistics Visualizer {.smaller}

::: {style="font-size: 0.8em;"}

**Explore:** Distribution of min and max from Uniform(0,1) samples

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof n_order = Inputs.range([2, 20], {
  value: 5, 
  step: 1, 
  label: "Sample size n:"
})

// Expected values
e_min = 1 / (n_order + 1)
e_max = n_order / (n_order + 1)
```

**E[Y₍₁₎]:** ${e_min.toFixed(4)}

**E[Y₍ₙ₎]:** ${e_max.toFixed(4)}

**Range:** ${(e_max - e_min).toFixed(4)}

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// Generate order statistics PDFs
orderPdfs = {
  const points = [];
  for (let y = 0.01; y <= 0.99; y += 0.01) {
    const f_min = n_order * Math.pow(1 - y, n_order - 1);
    const f_max = n_order * Math.pow(y, n_order - 1);
    points.push({y: y, f_min: f_min, f_max: f_max});
  }
  return points;
}

Plot.plot({
  width: 480,
  height: 320,
  x: { domain: [0, 1], label: "y" },
  y: { domain: [0, Math.max(n_order, 5)], label: "Density" },
  marks: [
    Plot.line(orderPdfs, {x: "y", y: "f_min", stroke: "red", strokeWidth: 2}),
    Plot.line(orderPdfs, {x: "y", y: "f_max", stroke: "blue", strokeWidth: 2}),
    Plot.ruleY([0])
  ]
})
```

**Red:** Minimum | **Blue:** Maximum

:::
:::
:::

## 📖 Definition: Sample Range {.smaller}

::: {.callout-note}
## 📝 Definition 6.3: Sample Range

The **sample range** is defined as:
$$R = Y_{(n)} - Y_{(1)} = \max - \min$$

**Interpretation:** Measures the spread of the sample data.

For a random sample from Uniform(0, $\theta$):
- The range $R$ is a **sufficient statistic** for $\theta$
- $E[R] = \frac{n-1}{n+1}\theta$
:::

**Finance Application:** The range of daily returns over a period measures **realized volatility** — the difference between the highest and lowest prices is the "trading range."

## 🧮 Joint Distribution of Extreme Order Statistics {.smaller}

::: {.callout-important}
## Theorem 6.9: Joint PDF of Min and Max

The joint pdf of $(Y_{(1)}, Y_{(n)})$ is:

$$f_{Y_{(1)}, Y_{(n)}}(y_1, y_n) = n(n-1)[F(y_n) - F(y_1)]^{n-2}f(y_1)f(y_n)$$

for $y_1 < y_n$.

**For Uniform(0,1):**
$$f_{Y_{(1)}, Y_{(n)}}(y_1, y_n) = n(n-1)(y_n - y_1)^{n-2}$$
:::

This allows us to find the distribution of the **range** $R = Y_{(n)} - Y_{(1)}$.

## 💰 Case Study: Value-at-Risk with Order Statistics {.smaller}

::: {style="font-size:22px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

library(tidyverse)
library(tidyquant)

# Download S&P 500 data
spy <- tq_get("SPY", from = "2015-01-01", to = "2024-12-31")

# Calculate daily returns
returns <- spy %>%
  mutate(ret = log(adjusted / lag(adjusted))) %>%
  na.omit()

n <- nrow(returns)
cat(sprintf("Sample size: %d trading days\n\n", n))

# Historical VaR using order statistics
# VaR at α = 5% means the α*n-th smallest return
alpha <- 0.05
k <- ceiling(alpha * n)  # k-th order statistic

sorted_returns <- sort(returns$ret)
var_5pct <- sorted_returns[k]

cat(sprintf("5%% Historical VaR: %.4f (%.2f%%)\n", 
            var_5pct, var_5pct * 100))
cat(sprintf("This is Y_(%d) from %d observations\n", k, n))

# Also compute 1% VaR
k_1pct <- ceiling(0.01 * n)
var_1pct <- sorted_returns[k_1pct]
cat(sprintf("\n1%% Historical VaR: %.4f (%.2f%%)\n", 
            var_1pct, var_1pct * 100))
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 5

# Visualize VaR
ggplot(returns, aes(x = ret)) +
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 100, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = var_5pct, color = "red", 
             linewidth = 1.2, linetype = "dashed") +
  geom_vline(xintercept = var_1pct, color = "darkred", 
             linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = var_5pct - 0.005, y = 30, 
           label = "5% VaR", color = "red", angle = 90) +
  annotate("text", x = var_1pct - 0.005, y = 30, 
           label = "1% VaR", color = "darkred", angle = 90) +
  labs(title = "SPY Return Distribution with VaR",
       subtitle = "Historical VaR using order statistics",
       x = "Daily Log Return", y = "Density") +
  theme_minimal()
```

:::
:::
:::

## 💰 Case Study: Extreme Returns Analysis {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Analyze extreme returns (order statistics)
cat("=== Extreme Return Analysis ===\n\n")

# Worst 10 days (smallest order statistics)
cat("10 Worst Days (Y_(1) to Y_(10)):\n")
worst_10 <- returns %>%
  arrange(ret) %>%
  head(10) %>%
  select(date, ret)
print(worst_10, n = 10)

# Best 10 days (largest order statistics)  
cat("\n10 Best Days (Y_(n-9) to Y_(n)):\n")
best_10 <- returns %>%
  arrange(desc(ret)) %>%
  head(10) %>%
  select(date, ret)
print(best_10, n = 10)
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Range statistics
range_ret <- max(returns$ret) - min(returns$ret)
cat(sprintf("\n=== Range Statistics ===\n"))
cat(sprintf("Maximum return: %.4f (%.2f%%)\n", 
            max(returns$ret), max(returns$ret)*100))
cat(sprintf("Minimum return: %.4f (%.2f%%)\n", 
            min(returns$ret), min(returns$ret)*100))
cat(sprintf("Sample range: %.4f (%.2f%%)\n", 
            range_ret, range_ret*100))

# Compare to theoretical for normal
sigma <- sd(returns$ret)
n <- nrow(returns)
# Expected range for normal is approximately 2*sigma*sqrt(2*log(n))
expected_range_normal <- 2 * sigma * sqrt(2 * log(n))
cat(sprintf("\nExpected range (if Normal): %.4f\n", expected_range_normal))
cat(sprintf("Actual range: %.4f\n", range_ret))
cat(sprintf("Ratio: %.2f (>1 suggests fat tails)\n", 
            range_ret / expected_range_normal))
```

:::
:::
:::

## 💰 Case Study: Key Findings {.smaller}

::: {style="font-size: 26px;"}
::: {.callout-important}
## 📊 Analysis Results

::: {.columns}

::: {.column width="33%"}
::: {.fragment}
**Order Statistics for VaR:**

- 5% VaR = $Y_{(k)}$ where $k = \lceil 0.05n \rceil$

- Non-parametric: no distribution assumption needed

- Directly interpretable as "worst $\alpha$% of days"
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Extreme Value Insights:**

- Worst days often cluster (market crises)

- Best days also cluster (recovery periods)

- Missing few best days dramatically hurts returns
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Practical Implications:**

1. **Fat tails**: Actual range exceeds normal prediction

2. **Risk management**: Order statistics provide robust VaR

3. **Timing matters**: Extreme days dominate long-term returns
:::
:::

:::
:::
:::

## 📝 Quiz #1: Jacobian Transformation {.smaller .quiz-question}

For the transformation $U_1 = Y_1 + Y_2$, $U_2 = Y_1 - Y_2$, the absolute value of the Jacobian is:

- [1/2]{.correct data-explanation="✅ Correct! The inverse transformation is y₁ = (u₁+u₂)/2, y₂ = (u₁-u₂)/2. The Jacobian determinant is (1/2)(-1/2) - (1/2)(1/2) = -1/2, so |J| = 1/2."}
- 1
- 2
- 1/4

## 📝 Quiz #2: Order Statistics Notation {.smaller .quiz-question}

In a sample of size $n = 10$, $Y_{(3)}$ represents:

- [The 3rd smallest value in the sample]{.correct data-explanation="✅ Correct! Order statistics Y₍ₖ₎ represent the k-th smallest value when the sample is arranged in ascending order. So Y₍₃₎ is the third smallest observation."}
- The 3rd observation in the original sample
- The 3rd largest value in the sample
- The median of the sample

## 📝 Quiz #3: Maximum of Uniform Sample {.smaller .quiz-question}

If $Y_1, \ldots, Y_5$ are iid Uniform(0,1), the pdf of $Y_{(5)} = \max$ is:

- [$5y^4$ for $0 < y < 1$]{.correct data-explanation="✅ Correct! Using the formula f_{Y_(n)}(y) = n[F(y)]^{n-1}f(y) with F(y) = y and f(y) = 1 for Uniform(0,1), we get f_{Y_(5)}(y) = 5·y⁴·1 = 5y⁴."}
- $y^5$ for $0 < y < 1$
- $5(1-y)^4$ for $0 < y < 1$
- $1$ for $0 < y < 1$

## 📝 Quiz #4: Historical VaR {.smaller .quiz-question}

To compute the 5% historical VaR from 1000 daily returns, you would use:

- [The 50th smallest return (approximately $Y_{(50)}$)]{.correct data-explanation="✅ Correct! The α% VaR corresponds to the ⌈αn⌉-th order statistic. For α = 0.05 and n = 1000, this is approximately the 50th smallest return."}
- The 5th smallest return
- The 950th smallest return
- The average of all returns

## 📝 Summary {.smaller}

::: {style="font-size: 28px"}
::: {.summary-box}
**✅ Key Takeaways**

- **Bivariate Jacobian**: $f_{U_1,U_2}(u_1,u_2) = f_{Y_1,Y_2}(h_1,h_2) \cdot |J|$ where $J$ is the determinant of partial derivatives

- **Sum and difference** of independent normals are independent normals — powerful result!

- **Order statistics** $Y_{(k)}$: k-th smallest value, with pdf involving $[F(y)]^{k-1}[1-F(y)]^{n-k}$

- **Extreme order statistics**: Min has pdf $n[1-F(y)]^{n-1}f(y)$; Max has pdf $n[F(y)]^{n-1}f(y)$

- **Sample range** $R = Y_{(n)} - Y_{(1)}$ measures spread; useful for volatility estimation

- **Historical VaR**: The $\alpha$-quantile is estimated by order statistic $Y_{(\lceil \alpha n \rceil)}$
:::
:::

## 📚 Practice Problems

::: {.callout-tip}
## 📝 Homework Problems

**Problem 1 (Jacobian):** Let $Y_1, Y_2$ be independent Exponential(1). Use Jacobian method to find the joint pdf of $U = Y_1 + Y_2$ and $V = Y_1/(Y_1 + Y_2)$. Show $U$ and $V$ are independent.

**Problem 2 (Order Statistics):** For a sample of size 5 from Exponential(β), find the pdf of the median $Y_{(3)}$.

**Problem 3 (Maximum):** If $Y_1, \ldots, Y_{10}$ are iid Exponential(1), find $P(Y_{(10)} > 3)$.

**Problem 4 (Range):** For $Y_1, \ldots, Y_n$ iid Uniform(0,1), find $E[R]$ where $R = Y_{(n)} - Y_{(1)}$.

**Problem 5 (VaR):** From 500 daily returns, you want to estimate the 1% VaR. Which order statistic would you use? What is the interpretation?
:::

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business

ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Sampling Distributions and the Central Limit Theorem (Chapter 7)

**Reading:** Chapter 7, Sections 7.1-7.3

**Preparation:** Review normal distribution properties

**⏰ Reminders:**

✅ Complete Practice Problems 1-5

✅ Review Chapter 6 concepts thoroughly

✅ Think about how sample statistics are distributed

✅ Work hard!
:::
:::

## ❓ Questions? {.center}

::: {.callout-note}
## 💬 Open Discussion

**Key Topics for Discussion:**

- How does the Jacobian generalize the univariate transformation formula?

- Why are order statistics useful for robust estimation?

- What are the advantages of historical VaR over parametric VaR?

- How do extreme value distributions extend order statistics theory?
:::
