---
title: "Mathematical Statistics"
subtitle: "Functions of Random Variables: Introduction and Distribution Functions"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ../ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    margin: 0.04
    min-scale: 0.2
    max-scale: 2.0
    footer: "Mathematical Statistics - Functions of Random Variables"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    mouse-wheel: true
    overview: true
    zoom: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::::: learning-objectives
:::: {style="font-size:35px"}
📚 By the end of this lecture, you will be able to:

::: incremental
-   Identify and formulate **functions of random variables** in financial contexts such as portfolio returns, option payoffs, and risk metrics

-   Compute probability distributions of **discrete transformations** using the probability function method for financial applications

-   Apply the **method of distribution functions** to derive probability distributions of transformed continuous random variables

-   Calculate probabilities for **derived distributions** such as sums, differences, and ratios of stock returns using CDFs

-   Implement statistical transformations in R to analyze **real financial data** and validate theoretical distributions
:::
::::
:::::

## 📋 Overview

::::: {.callout-note}
## 📚 Topics Covered Today

:::: {style="font-size:32px"}
::: incremental
-   **Why Functions of Random Variables?** – Motivation from portfolio theory, derivatives pricing, and risk management

-   **Finding Probability Distributions** – Techniques for deriving distributions of transformations

-   **Method of Distribution Functions** – Using CDFs to find distributions of $U = g(Y)$ for continuous random variables

-   **Financial Applications** – Portfolio returns, option payoffs, profit/loss distributions

-   **Case Study** – Analyzing transformed stock returns using real market data
:::
::::
:::::

## 📖 Why Study Functions of Random Variables?

::::::::: {.callout-note}
## 🎯 Motivation

:::::::: {style="font-size:30px"}
In finance and economics, we constantly work with **transformations of random variables**:

::::::: columns
:::: {.column width="50%"}
::: fragment
**Portfolio & Investment:**

-   Portfolio return = weighted sum of individual stock returns
-   Profit/Loss = Stock price - Purchase price
-   Return on investment = (Final - Initial) / Initial
-   Option payoffs = max(S - K, 0)
:::
::::

:::: {.column width="50%"}
::: fragment
**Risk Management:**

-   Value at Risk (VaR) transformations
-   Portfolio variance from covariances
-   Credit ratings from default probabilities
-   Exchange rate conversions
:::
::::
:::::::
::::::::
:::::::::

## 💡 Real-World Example {.larger}

:::::: {style="font-size:32px"}
::::: columns
:::: {.column width="60%"}
::: fragment
**Scenario:** You invest in a stock with uncertain return $Y$ (in %). Your initial investment is \$10,000.

Your **profit** is a function of the return:
$$U = 10000 \times \frac{Y}{100}$$

::: {.callout-important}
**Question:** If we know the probability distribution of $Y$ (the return), how do we find the probability distribution of $U$ (the profit)?
:::
:::
::::

:::: {.column .fragment width="40%"}
**This is the central problem:**

Given:

-   Random variable $Y$
-   Function $U = g(Y)$
-   Distribution of $Y$

Find:

-   Distribution of $U$
::::
:::::
::::::

## 📖 Key Problem Statement {.larger}

::::::: {.callout-important style="font-size:40px;"}
## 🎯 Central Question of Chapter 6

::::: columns
:::: {.column width="50%"}
Given:

1.  Random variable(s): $Y$ or $(Y_1, Y_2, \ldots, Y_n)$
2.  A function: $U = g(Y)$ or $U = g(Y_1, Y_2, \ldots, Y_n)$
3.  The probability distribution of $Y$ (or joint distribution)

**Find:** The probability distribution of $U = g(Y)$
::::

:::: {.column .fragment width="50%"}
**Why does this matter?**

-   Statistical inference often requires knowing distributions of **sample statistics** (e.g., $\bar{Y}$, $S^2$)
-   Financial models require distributions of **derived quantities** (portfolio returns, option values)
-   Risk management needs distributions of **transformed variables** (VaR, profit/loss)
::::
:::::
:::::::

## 🔍 Example 6.1: Profit from Investment (Discrete) 

:::::: {style="font-size:30px !important; margin-top: -20px;"}
::::: columns
:::: {.column width="55%"}
::: fragment
**Problem:** A stock's return $Y$ (in \$) has the following probability distribution:

| $y$    | -2   | -1   | 0    | 1    | 2    |
|--------|------|------|------|------|------|
| $p(y)$ | 0.10 | 0.20 | 0.30 | 0.25 | 0.15 |

You bought 1 share for \$50. Your **profit** is:
$$U = Y - 5 \text{ (subtracting transaction cost)}$$

**Find:** The probability distribution of $U$.
:::
::::

:::: {.column .fragment width="45%"}
**Solution:**

For each value of $y$, compute $u = y - 5$:

| $y$ | $u = y-5$ | $p(u)$ |
|-----|-----------|--------|
| -2  | -7        | 0.10   |
| -1  | -6        | 0.20   |
| 0   | -5        | 0.30   |
| 1   | -4        | 0.25   |
| 2   | -3        | 0.15   |

::: {.callout-tip}
The probabilities remain the same! Only the values shift.
:::
::::
:::::
::::::

## 📊 Visualizing the Transformation

```{ojs}
//| echo: false

viewof shift_param = {
  const wrapper = html`<div style="display: flex; align-items: center; gap: 10px;">
    <span style="white-space: nowrap; font-size: 14px; font-weight: 500; min-width: 180px;">Transaction Cost (shift):</span>
  </div>`;
  const slider = Inputs.range([-10, 10], {value: -5, step: 0.5, label: ""});
  wrapper.appendChild(slider);
  
  // Forward the value property from the slider to the wrapper
  Object.defineProperty(wrapper, 'value', {
    get: () => slider.value,
    set: (v) => { slider.value = v; }
  });
  
  ['pointerdown', 'touchstart', 'mousedown', 'click', 'wheel', 'pointermove', 'touchmove', 'mousemove'].forEach(e =>
    wrapper.addEventListener(e, ev => ev.stopPropagation())
  );
  return wrapper;
}

y_values = [-2, -1, 0, 1, 2]
probs = [0.10, 0.20, 0.30, 0.25, 0.15]
u_values = y_values.map(y => y + shift_param)

{
  const width = 1150;
  const height = 550;
  const margin = {top: 50, right: 50, bottom: 70, left: 80};
  
  const container = html`<div style="display: flex; justify-content: center; width: 100%;"></div>`;
  
  const svg = d3.create("svg")
    .attr("width", width)
    .attr("height", height)
    .attr("style", "background: white; font-size: 16px;");
  
  const x = d3.scaleLinear()
    .domain([
      Math.min(d3.min(y_values), d3.min(u_values)) - 2, 
      Math.max(d3.max(y_values), d3.max(u_values)) + 2
    ])
    .range([margin.left, width - margin.right]);
  
  const y = d3.scaleLinear()
    .domain([0, 0.35])
    .range([height - margin.bottom, margin.top]);
  
  // Original Y distribution
  svg.selectAll(".bar-y")
    .data(y_values.map((val, i) => ({x: val, y: probs[i]})))
    .join("rect")
    .attr("class", "bar-y")
    .attr("x", d => x(d.x) - 18)
    .attr("y", d => y(d.y))
    .attr("width", 36)
    .attr("height", d => y(0) - y(d.y))
    .attr("fill", "steelblue")
    .attr("opacity", 0.5);
  
  // Transformed U distribution
  svg.selectAll(".bar-u")
    .data(u_values.map((val, i) => ({x: val, y: probs[i]})))
    .join("rect")
    .attr("class", "bar-u")
    .attr("x", d => x(d.x) - 18)
    .attr("y", d => y(d.y))
    .attr("width", 36)
    .attr("height", d => y(0) - y(d.y))
    .attr("fill", "darkred")
    .attr("opacity", 0.7);
  
  // Axes
  svg.append("g")
    .attr("transform", `translate(0,${height - margin.bottom})`)
    .call(d3.axisBottom(x))
    .style("font-size", "16px");
  
  svg.append("g")
    .attr("transform", `translate(${margin.left},0)`)
    .call(d3.axisLeft(y))
    .style("font-size", "16px");
  
  // Labels
  svg.append("text")
    .attr("x", width / 2)
    .attr("y", height - 15)
    .attr("text-anchor", "middle")
    .style("font-size", "20px")
    .text("Value");
  
  svg.append("text")
    .attr("transform", "rotate(-90)")
    .attr("x", -height / 2)
    .attr("y", 25)
    .attr("text-anchor", "middle")
    .style("font-size", "20px")
    .text("Probability");
  
  svg.append("text")
    .attr("x", width / 2)
    .attr("y", 30)
    .attr("text-anchor", "middle")
    .style("font-size", "22px")
    .style("font-weight", "bold")
    .text(`Original Y (blue) vs Transformed U = Y + ${shift_param.toFixed(1)} (red)`);
  
  container.appendChild(svg.node());
  return container;
}
```

## 📖 General Approach for Discrete RVs {.larger}

:::::: {.callout-note style="font-size:40px;"}
## 📝 Method for Discrete Transformations

Given: Discrete random variable $Y$ with probability function $p_Y(y)$

Let: $U = g(Y)$ be a transformation

**Steps to find** $p_U(u)$:

::: incremental
1.  List all possible values of $Y$: $y_1, y_2, \ldots$

2.  Compute corresponding values of $U$: $u_i = g(y_i)$

3.  For each unique value $u$, find all $y$ values that map to it

4.  Sum probabilities: $p_U(u) = \sum_{y: g(y)=u} p_Y(y)$
:::
::::::

## 🔍 Example 6.2: Non-One-to-One Function {.larger}

:::::: {style="font-size:32px"}
::::: columns
:::: {.column .fragment width="46%"}
**Problem:** Stock return $Y$ has distribution:

| $y$    | -2   | -1   | 0    | 1    | 2    |
|--------|------|------|------|------|------|
| $p(y)$ | 0.10 | 0.20 | 0.30 | 0.25 | 0.15 |

Define **absolute return**:
$$U = |Y|^2$$

Find $p_U(u)$.
::::

:::: {.column .fragment width="30%"}
**Mapping:** Multiple $y$ values map to same $u$:

| $y$ | $u = y^2$ | $p_Y(y)$ |
|-----|-----------|----------|
| -2  | 4         | 0.10     |
| -1  | 1         | 0.20     |
| 0   | 0         | 0.30     |
| 1   | 1         | 0.25     |
| 2   | 4         | 0.15     |
::::

:::: {.column .fragment width="24%"}
**Solution:** Combine probabilities for same $u$:

| $u$ | $p_U(u)$ |
|-----|----------|
| 0   | 0.30     |
| 1   | 0.45     |
| 4   | 0.25     |

::: {.callout-tip}
$p_U(1) = p_Y(-1) + p_Y(1)$

$= 0.20 + 0.25 = 0.45$
:::
::::
:::::
::::::

## 📖 Method of Distribution Functions (Continuous) {.smaller}

::::::: {.callout-important style="font-size:32px;"}
## 🔧 Method of Distribution Functions

**For continuous random variables**, we use CDFs to find distributions:

::::: columns
:::: {.column width="50%"}
**Given:**

-   Continuous RV $Y$ with pdf $f_Y(y)$ and CDF $F_Y(y)$
-   Transformation $U = g(Y)$
::::

:::: {.column .fragment width="50%"}
**Steps:**

1.  Find CDF of $U$: $F_U(u) = P(U \leq u) = P(g(Y) \leq u)$

2.  Convert to inequality in $Y$: $P(Y \in A_u)$ where $A_u = \{y: g(y) \leq u\}$

3.  Differentiate to get pdf: $f_U(u) = \frac{d}{du}F_U(u)$
::::
:::::
:::::::

## 🔍 Example 6.3: Linear Transformation

:::::: {style="font-size:30px; margin-top:-20px;"}
::::: columns
:::: {.column width="50%"}
::: fragment
**Problem:** Daily stock return $Y \sim \text{Uniform}(0, 1)$, so $f_Y(y) = 1$ for $0 \leq y \leq 1$.

An investor scales and shifts the return: $U = 2Y + 1$

**Find:** The pdf of $U$.
:::

::: fragment
**Solution:**

1.  **Find support of U:** Since $0 \leq Y \leq 1$, then $1 \leq U \leq 3$

2.  **Find CDF:** For $1 \leq u \leq 3$:
    $$F_U(u) = P(U \leq u) = P(2Y + 1 \leq u) $$
    $$ = P\left(Y \leq \frac{u-1}{2}\right) = F_Y\left(\frac{u-1}{2}\right)$$
:::
::::

:::: {.column .fragment width="50%"}
**Solution (continued):**

3.  **Since** $Y \sim \text{Uniform}(0,1)$: $F_Y(y) = y$, so $F_U(u) = \frac{u-1}{2}$

4.  **Differentiate:**
    $$f_U(u) = \frac{d}{du}\left(\frac{u-1}{2}\right) = \frac{1}{2}, \quad 1 \leq u \leq 3$$

::: {.callout-tip}
$U \sim \text{Uniform}(1, 3)$ — Linear transformation of uniform is uniform!
:::
::::
:::::
::::::

## 📊 Interactive: Linear Transformation of Uniform

```{ojs}
//| echo: false

viewof scale_a = {
  const wrapper = html`<div style="display: flex; align-items: center; gap: 10px;">
    <span style="white-space: nowrap; font-size: 14px; font-weight: 500; min-width: 160px;">Scale parameter (a):</span>
  </div>`;
  const slider = Inputs.range([0.5, 5], {value: 2, step: 0.1, label: ""});
  wrapper.appendChild(slider);
  
  // Forward the value property from the slider to the wrapper
  Object.defineProperty(wrapper, 'value', {
    get: () => slider.value,
    set: (v) => { slider.value = v; }
  });
  
  ['pointerdown', 'touchstart', 'mousedown', 'click', 'wheel', 'pointermove', 'touchmove', 'mousemove'].forEach(e =>
    wrapper.addEventListener(e, ev => ev.stopPropagation())
  );
  return wrapper;
}

viewof shift_b = {
  const wrapper = html`<div style="display: flex; align-items: center; gap: 10px;">
    <span style="white-space: nowrap; font-size: 14px; font-weight: 500; min-width: 160px;">Shift parameter (b):</span>
  </div>`;
  const slider = Inputs.range([-2, 5], {value: 1, step: 0.1, label: ""});
  wrapper.appendChild(slider);
  
  // Forward the value property from the slider to the wrapper
  Object.defineProperty(wrapper, 'value', {
    get: () => slider.value,
    set: (v) => { slider.value = v; }
  });
  
  ['pointerdown', 'touchstart', 'mousedown', 'click', 'wheel', 'pointermove', 'touchmove', 'mousemove'].forEach(e =>
    wrapper.addEventListener(e, ev => ev.stopPropagation())
  );
  return wrapper;
}

u_min = shift_b
u_max = scale_a + shift_b
f_u = 1 / scale_a

{
  const width = 1000;
  const height = 380;
  const margin = {top: 40, right: 40, bottom: 60, left: 70};

  const svg = d3.create("svg")
    .attr("width", width)
    .attr("height", height)
    .attr("style", "background: white; font-size: 16px;");

  const maxX = Math.max(6, u_max + 1);
  const minX = -1;
  const maxY = Math.max(1.5, f_u + 0.3);

  const plotWidth = width - margin.left - margin.right;
  const plotHeight = height - margin.top - margin.bottom;

  // Equal scale: same pixels per unit on both axes
  const pixelsPerUnitX = plotWidth / (maxX - minX);
  const pixelsPerUnitY = plotHeight / maxY;
  const pixelsPerUnit = Math.min(pixelsPerUnitX, pixelsPerUnitY);

  const actualPlotWidth = pixelsPerUnit * (maxX - minX);
  const actualPlotHeight = pixelsPerUnit * maxY;

  const x = d3.scaleLinear()
    .domain([minX, maxX])
    .range([margin.left, margin.left + actualPlotWidth]);

  const y_scale = d3.scaleLinear()
    .domain([0, maxY])
    .range([height - margin.bottom, height - margin.bottom - actualPlotHeight]);

  // Original Y ~ Uniform(0,1)
  svg.append("rect")
    .attr("x", x(0))
    .attr("y", y_scale(1))
    .attr("width", x(1) - x(0))
    .attr("height", y_scale(0) - y_scale(1))
    .attr("fill", "steelblue")
    .attr("opacity", 0.4);

  // Transformed U ~ Uniform(b, a+b)
  svg.append("rect")
    .attr("x", x(u_min))
    .attr("y", y_scale(f_u))
    .attr("width", x(u_max) - x(u_min))
    .attr("height", y_scale(0) - y_scale(f_u))
    .attr("fill", "darkred")
    .attr("opacity", 0.6);

  // Axes
  svg.append("g")
    .attr("transform", `translate(0,${height - margin.bottom})`)
    .call(d3.axisBottom(x))
    .style("font-size", "14px");

  svg.append("g")
    .attr("transform", `translate(${margin.left},0)`)
    .call(d3.axisLeft(y_scale))
    .style("font-size", "14px");

  // Labels
  svg.append("text")
    .attr("x", width / 2)
    .attr("y", height - 10)
    .attr("text-anchor", "middle")
    .style("font-size", "18px")
    .text("Value");

  svg.append("text")
    .attr("transform", "rotate(-90)")
    .attr("x", -height / 2)
    .attr("y", 20)
    .attr("text-anchor", "middle")
    .style("font-size", "18px")
    .text("Density");

  svg.append("text")
    .attr("x", width / 2)
    .attr("y", 25)
    .attr("text-anchor", "middle")
    .style("font-size", "20px")
    .style("font-weight", "bold")
    .text(`Y ~ Uniform(0,1) → U = ${scale_a.toFixed(1)}Y + ${shift_b.toFixed(1)}`);

  return svg.node();
}
```

## 🔍 Example 6.4: Nonlinear Transformation {.larger}

:::::: {style="font-size:36px; margin-top:-20px;"}
::::: columns
:::: {.column width="50%"}
::: fragment
**Problem:** Stock return $Y \sim \text{Uniform}(0, 1)$.

Define **squared return**: $U = Y^2$

**Find:** The pdf of $U$.
:::

::: fragment
**Solution:**

1.  **Support:** If $0 \leq Y \leq 1$, then $0 \leq U \leq 1$

2.  **CDF:** For $0 \leq u \leq 1$:
    $$F_U(u) = P(U \leq u) = P(Y^2 \leq u) $$
    $$= P(Y \leq \sqrt{u}) = \sqrt{u}$$
:::
::::

:::: {.column .fragment width="50%"}
**Solution (continued):**

3.  **PDF:** Differentiate:
    $$f_U(u) = \frac{d}{du}(\sqrt{u}) = \frac{1}{2\sqrt{u}}, \quad 0 < u < 1$$

::: {.callout-warning}
**Note:** The density goes to infinity as $u \to 0^+$! This is valid because $\int_0^1 \frac{1}{2\sqrt{u}} du = 1$.
:::
::::
:::::
::::::

## 📊 Comparing Linear vs Nonlinear Transformations

::::: columns
:::: {.column width="30%"}
```{ojs}
//| echo: false

viewof trans_type = {
  const wrapper = html`<div style="margin-top: 20px;">
    <div style="font-size: 18px; font-weight: 600; margin-bottom: 15px; color: #2c3e50;">
      Select Transformation:
    </div>
  </div>`;
  
  const radio = Inputs.radio(
    ["Linear: U = 2Y + 1", "Quadratic: U = Y²", "Cubic: U = Y³", "Square Root: U = √Y"],
    {value: "Quadratic: U = Y²", label: ""}
  );
  wrapper.appendChild(radio);
  
  // Forward the value property
  Object.defineProperty(wrapper, 'value', {
    get: () => radio.value,
    set: (v) => { radio.value = v; }
  });
  
  return wrapper;
}
```

::: {.callout-tip style="font-size: 20px; margin-top: 28px;"}
**Key Insight:**

- **Linear**: Preserves uniform shape
- **Quadratic**: Density → ∞ as u → 0
- **Cubic**: Higher concentration near 0
- **Square Root**: Linear increase in density
:::
::::

:::: {.column width="70%"}
```{ojs}
//| echo: false

y_grid = Array.from({length: 300}, (_, i) => i / 299)

u_transformed = y_grid.map(y => {
  if (trans_type === "Linear: U = 2Y + 1") return 2 * y + 1;
  if (trans_type === "Quadratic: U = Y²") return y * y;
  if (trans_type === "Cubic: U = Y³") return y * y * y;
  if (trans_type === "Square Root: U = √Y") return Math.sqrt(y);
})

f_u_values = y_grid.map((y, i) => {
  const u = u_transformed[i];
  if (trans_type === "Linear: U = 2Y + 1") return {u: u, f: 0.5};
  if (trans_type === "Quadratic: U = Y²") return {u: u, f: 0.5 / Math.sqrt(Math.max(u, 0.001))};
  if (trans_type === "Cubic: U = Y³") return {u: u, f: (1/3) / Math.pow(Math.max(u, 0.001), 2/3)};
  if (trans_type === "Square Root: U = √Y") return {u: u, f: 2 * u};
}).filter(d => d.u >= 0 && d.u <= 5 && d.f < 100)

{
  const width = 850;
  const height = 550;
  const margin = {top: 60, right: 40, bottom: 70, left: 80};
  
  const container = html`<div style="display: flex; justify-content: center; width: 100%;"></div>`;
  
  const svg = d3.create("svg")
    .attr("width", width)
    .attr("height", height)
    .attr("style", "background: white; font-size: 16px;");
  
  // Fixed domains to zoom the plot while maintaining equal scale
  const maxF = 2;     // y-axis limit
  const maxU = 3.5;   // x-axis limit
  
  // Calculate scale to maintain aspect ratio (same pixels per unit on both axes)
  const plotHeight = height - margin.bottom - margin.top;
  const plotWidth = width - margin.right - margin.left;
  
  // Use the smaller scale to ensure both axes fit
  const pixelsPerUnitY = plotHeight / maxF;
  const pixelsPerUnitX = plotWidth / maxU;
  const pixelsPerUnit = Math.min(pixelsPerUnitY, pixelsPerUnitX);
  
  // Recalculate actual plot dimensions with equal scale
  const actualPlotWidth = pixelsPerUnit * maxU;
  const actualPlotHeight = pixelsPerUnit * maxF;
  
  const x = d3.scaleLinear()
    .domain([0, maxU])
    .range([margin.left, margin.left + actualPlotWidth]);
  
  const y_scale = d3.scaleLinear()
    .domain([0, maxF])
    .range([height - margin.bottom, height - margin.bottom - actualPlotHeight]);
  
  // Original Y ~ Uniform(0,1) distribution - now truly square with width=1, height=1
  svg.append("rect")
    .attr("x", x(0))
    .attr("y", y_scale(1))
    .attr("width", x(1) - x(0))
    .attr("height", y_scale(0) - y_scale(1))
    .attr("fill", "steelblue")
    .attr("opacity", 0.5)
    .attr("stroke", "steelblue")
    .attr("stroke-width", 4);
  
  // Add label for original distribution
  svg.append("text")
    .attr("x", x(0.5))
    .attr("y", y_scale(1) - 10)
    .attr("text-anchor", "middle")
    .style("font-size", "15px")
    .style("fill", "steelblue")
    .style("font-weight", "bold")
    .text("Y ~ Uniform(0,1)");
  
  // Transformed distribution - clip at maxF for stability
  const area = d3.area()
    .x(d => x(d.u))
    .y0(y_scale(0))
    .y1(d => y_scale(Math.min(d.f, maxF)));
  
  svg.append("path")
    .datum(f_u_values)
    .attr("fill", "darkred")
    .attr("fill-opacity", 0.5)
    .attr("stroke", "darkred")
    .attr("stroke-width", 3)
    .attr("d", area);
  
  // Axes
  svg.append("g")
    .attr("transform", `translate(0,${height - margin.bottom})`)
    .call(d3.axisBottom(x).ticks(8))
    .style("font-size", "15px");
  
  svg.append("g")
    .attr("transform", `translate(${margin.left},0)`)
    .call(d3.axisLeft(y_scale).ticks(6))
    .style("font-size", "15px");
  
  // Axis labels
  svg.append("text")
    .attr("x", width / 2)
    .attr("y", height - 20)
    .attr("text-anchor", "middle")
    .style("font-size", "19px")
    .style("font-weight", "600")
    .text("Value");
  
  svg.append("text")
    .attr("transform", "rotate(-90)")
    .attr("x", -height / 2)
    .attr("y", 25)
    .attr("text-anchor", "middle")
    .style("font-size", "19px")
    .style("font-weight", "600")
    .text("Density");
  
  // Title
  svg.append("text")
    .attr("x", width / 2)
    .attr("y", 30)
    .attr("text-anchor", "middle")
    .style("font-size", "21px")
    .style("font-weight", "bold")
    .text(`Transformation: ${trans_type}`);
  
  container.appendChild(svg.node());
  return container;
}
```
::::
:::::

<!-- ## 🧮 Theorem 6.1: Distribution Function Technique {.larger}

::::::: {.callout-important style="font-size:38px;"}
## Theorem 6.1: Distribution Function Technique

Let $Y$ have CDF $F_Y(y)$ and pdf $f_Y(y)$. Let $U = g(Y)$ where $g$ is a **monotone** function with inverse $g^{-1}$.

Then the pdf of $U$ is:

$$f_U(u) = f_Y(g^{-1}(u)) \left| \frac{d}{du} g^{-1}(u) \right|$$

:::: {.fragment}
::: {style="font-size:34px"}
**Derivation (for increasing $g$):** Starting from the CDF:
$$F_U(u) = P(g(Y) \leq u) = P(Y \leq g^{-1}(u)) = F_Y(g^{-1}(u))$$
Differentiating via the **chain rule**: $f_U(u) = f_Y(g^{-1}(u)) \cdot \frac{d}{du} g^{-1}(u)$

For **decreasing** $g$, the inequality flips, producing a sign change — hence the absolute value.
:::
::::
::::::: -->

## 💼 Think-Pair-Share: Portfolio Returns {.smaller}

::::: columns
:::: {.column width="45%"}
::: {.callout-note style="font-size:30px;"}
## 📋 Scenario

You invest in two stocks:

-   **Stock A:** $Y_A \sim \text{Normal}(\mu_A = 0.08, \sigma_A = 0.15)$
-   **Stock B:** $Y_B \sim \text{Normal}(\mu_B = 0.12, \sigma_B = 0.20)$

You allocate **60% to A** and **40% to B**.

Portfolio return:
$$U = 0.6 Y_A + 0.4 Y_B$$
:::
::::

:::: {.column width="55%"}
::: {.callout-tip style="font-size:28px;"}
## ❓ Discussion Questions

**1.** What method would you use to find the distribution of $U$?

**2.** If $Y_A$ and $Y_B$ are independent, what is the distribution of $U$?

**3.** How would correlation between $Y_A$ and $Y_B$ affect your answer?
:::

::: {style="margin-top:20px; font-size:26px; text-align:center; color:#2c3e50; font-weight:600;"}
⏱️ Discuss with your neighbor for 5 minutes
:::
::::
:::::

```{r}
#| echo: false
library(countdown)
countdown(minutes = 5, seconds = 0,
          top = 0, right = 0,
          font_size = "1.5em",
          color_running_background = "#31b09e",
          color_running_text = "white",
          color_finished_background = "#cc3311",
          color_finished_text = "white",
          color_warning_background = "#f7dc6f",
          warn_when = 60,
          play_sound = TRUE)
```

## 📊 Case Study: Demonstrating Y² Transformation

::::: columns
:::: {.column width="32%"}
::: {.callout-note style="font-size:32px"}
## 💼 Visualizing the Quadratic Transformation

**Key Learning:** When $Y \sim \text{Uniform}(0,1)$ and $U = Y^2$, we expect:

$$f_U(u) = \frac{1}{2\sqrt{u}}$$

Let's verify this with simulation!

**Method:**
- Generate 10,000 samples from Uniform(0,1)
- Transform to $U = Y^2$
- Compare empirical histogram to theoretical density
:::
::::

:::: {.column width="68%"}
```{r}
#| echo: true
#| code-fold: true
#| code-summary: "📊 Show Code"
#| fig-width: 9
#| fig-height: 4.5

library(tidyverse)

# Generate samples from Uniform(0,1)
set.seed(123)
n <- 10000
y_samples <- runif(n, 0, 1)
u_samples <- y_samples^2

# Theoretical density for U = Y²
u_seq <- seq(0.01, 1, length.out = 200)
f_u_theoretical <- 1 / (2 * sqrt(u_seq))

# Create comparison plot
ggplot() +
  # Empirical histogram
  geom_histogram(aes(x = u_samples, y = after_stat(density)), 
                 bins = 50, fill = "darkred", alpha = 0.5, color = "white") +
  # Theoretical density
  geom_line(aes(x = u_seq, y = f_u_theoretical), 
            color = "blue", linewidth = 2) +
  # Annotations
  annotate("text", x = 0.7, y = 3, 
           label = "Theoretical: f[U](u)==frac(1,2*sqrt(u))",
           parse = TRUE, size = 6, color = "blue") +
  annotate("text", x = 0.7, y = 2.5, 
           label = "Empirical: 10,000 samples",
           size = 5, color = "darkred") +
  labs(title = "Transformation U = Y² where Y ~ Uniform(0,1)",
       subtitle = "Notice: Density → ∞ as u → 0, matching our theoretical result!",
       x = "u = y²", y = "Density f_U(u)") +
  theme_minimal(base_size = 16) +
  coord_cartesian(ylim = c(0, 4))
```
::::
:::::

## 📊 Case Study: Real Stock Portfolio Transformation

::::: columns
:::: {.column width="35%"}
::: {.callout-note style="font-size:30px"}
## 💼 Linear Transformation of Stock Returns

**Scenario:** Portfolio with:
- 60% Apple (AAPL)
- 40% Microsoft (MSFT)
- Monthly data: 2020-2024

**Transformation:**
$$R_p = 0.6 \cdot R_{AAPL} + 0.4 \cdot R_{MSFT}$$

**Key Question:** How does the portfolio return distribution compare to individual stocks?
:::
::::

:::: {.column width="65%"}
```{r}
#| echo: true
#| code-fold: true
#| code-summary: "📊 Show Code"
#| message: false
#| warning: false
#| fig-width: 7.5
#| fig-height: 4

library(tidyverse)
library(tidyquant)

# Download monthly returns for AAPL and MSFT (2020-2024)
stocks <- c("AAPL", "MSFT")
prices <- tq_get(stocks, 
                 from = "2020-01-01", 
                 to = "2024-12-31",
                 get = "stock.prices")

# Calculate monthly returns
returns <- prices %>%
  group_by(symbol) %>%
  tq_transmute(select = adjusted,
               mutate_fun = periodReturn,
               period = "monthly",
               col_rename = "ret")

# Pivot to wide format
returns_wide <- returns %>%
  pivot_wider(names_from = symbol, values_from = ret) %>%
  drop_na()

# Create portfolio returns (60% AAPL, 40% MSFT)
returns_wide <- returns_wide %>%
  mutate(Portfolio = 0.6 * AAPL + 0.4 * MSFT)

# Plot distributions
returns_wide %>%
  select(date, AAPL, MSFT, Portfolio) %>%
  pivot_longer(-date, names_to = "Asset", values_to = "Return") %>%
  ggplot(aes(x = Return, fill = Asset, color = Asset)) +
  geom_density(alpha = 0.4, linewidth = 1.2) +
  scale_fill_manual(values = c("AAPL" = "#3498db", 
                                "MSFT" = "#e74c3c", 
                                "Portfolio" = "#2ecc71")) +
  scale_color_manual(values = c("AAPL" = "#2980b9", 
                                 "MSFT" = "#c0392b", 
                                 "Portfolio" = "#27ae60")) +
  labs(title = "Distribution of Monthly Returns (2020-2024)",
       subtitle = "Portfolio = 60% AAPL + 40% MSFT",
       x = "Monthly Return", y = "Density") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```
::::
:::::

## 📊 Case Study: Portfolio Statistics

::: {style="font-size:30px"}
```{r}
#| echo: true
#| code-fold: true
#| code-summary: "📊 Show Code"
#| message: false
#| warning: false

library(knitr)

# Calculate summary statistics
stats <- returns_wide %>%
  summarise(
    across(c(AAPL, MSFT, Portfolio),
           list(
             Mean = ~mean(., na.rm = TRUE) * 12 * 100,  # Annualized %
             SD = ~sd(., na.rm = TRUE) * sqrt(12) * 100,  # Annualized %
             Sharpe = ~(mean(., na.rm = TRUE) / sd(., na.rm = TRUE)) * sqrt(12)
           ),
           .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(everything(), 
               names_to = c("Asset", ".value"),
               names_pattern = "(.*)_(.*)") %>%
  mutate(across(where(is.numeric), ~round(., 2)))

kable(stats, caption = "Annualized Statistics: AAPL, MSFT, and 60/40 Portfolio")
```
:::

::: {.callout-important style="font-size:32px; margin-top:-10px;"}
### 🔑 Key Insights

**Linear Transformation Property:**

::: {.columns}

::: {.column width="50%" .fragment}
- Portfolio return is a **weighted sum** of individual returns
- $E[R_p] = 0.6 \cdot E[R_{AAPL}] + 0.4 \cdot E[R_{MSFT}]$ (linearity!)
:::

::: {.column width="50%" .fragment}
- Portfolio volatility depends on **correlation** between stocks
- **Diversification benefit:** Portfolio SD typically lower than weighted average of individual SDs
:::

:::
:::

## 📝 Quiz #1: Linear Transformations {.quiz-question}

::: {style="font-size:34px"}
A stock return Y follows a Uniform(0, 2) distribution. You transform it to U = 3Y + 1. What is the distribution of U?

- Uniform(0, 6)
- [Uniform(1, 7)]{.correct data-explanation="✅ Correct! Linear transformation of uniform is uniform. Since Y ∈ [0, 2], then U = 3Y + 1 ∈ [1, 7]. The pdf is f_U(u) = 1/6 for 1 ≤ u ≤ 7."}
- Uniform(3, 7)
- Normal(4, 1)
:::

## 📝 Quiz #2: Quadratic Transformation {.quiz-question}

::: {style="font-size:34px"}
If $Y \sim \text{Uniform}(0, 1)$ and $U = Y^2$, what is $f_U(0.25)$?

- 0.25
- 0.5
- [1.0]{.correct data-explanation="✅ Correct! Using f_U(u) = 1/(2√u) for 0 < u < 1: f_U(0.25) = 1/(2√0.25) = 1/(2×0.5) = 1.0"}
- 2.0
:::

## 📝 Quiz #3: Distribution Function Method {.quiz-question}

::: {style="font-size:34px"}
Why do we use the method of distribution functions for continuous RVs instead of directly transforming the pdf?

- It's computationally faster
- [The CDF method properly accounts for the change in probability measure]{.correct data-explanation="✅ Correct! The CDF method correctly handles probability transformations. Direct pdf transformation would miss the Jacobian factor needed to preserve total probability of 1."}
- PDFs cannot be transformed
- It gives the same result either way
:::

## 📝 Quiz #4: Non-One-to-One Functions {.quiz-question}

::: {style="font-size:34px"}
For discrete RV Y with transformation U = g(Y), if g is not one-to-one, what do we do?

- The transformation is invalid
- We must use a different method
- [Sum probabilities of all y values that map to each u]{.correct data-explanation="✅ Correct! We use p_U(u) = Σ_{y: g(y)=u} p_Y(y) to combine probabilities from all y values that produce the same u value."}
- The pdf becomes undefined
:::

## 📚 Summary {.larger}

:::::: {style="font-size:38px"}
::: incremental
**Key Takeaways:**

-   **Functions of random variables** are ubiquitous in finance: portfolio returns, option payoffs, risk metrics

-   **Discrete case:** Transform values and preserve/sum probabilities

-   **Continuous case:** Use the **method of distribution functions** (CDF approach)

-   **Linear transformations** preserve distributional families (e.g., uniform → uniform, normal → normal)

-   **Nonlinear transformations** can dramatically change the shape of distributions

-   **Real data** often requires empirical validation of theoretical distributions
:::
::::::

## 📝 Practice Problems {.smaller}

:::::: {style="font-size:34px"}
::::: columns
:::: {.column width="50%"}
**Problem 1:**

If $Y \sim \text{Exponential}(\lambda = 0.5)$ and $U = 2Y$, find $f_U(u)$.

**Problem 2:**

Stock return $Y \sim \text{Normal}(0.10, 0.04)$. An option payoff is $U = \max(Y - 0.08, 0)$. Describe how you would find $F_U(u)$.
::::

:::: {.column width="50%"}
**Problem 3:**

Two stocks have returns $Y_1$ and $Y_2$. Portfolio return is $U = 0.5Y_1 + 0.5Y_2$. If both are independent Uniform(0, 1), find $f_U(u)$.

**Problem 4:**

Prove that if $Y \sim \text{Uniform}(a, b)$ and $U = cY + d$ (with $c > 0$), then $U \sim \text{Uniform}(ca + d, cb + d)$.
::::
:::::
::::::

## 👋 Thank You!

::: {style="font-size: 32px"}
::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business

ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Method of Transformations (Theorem 6.2)

**Reading:** Wackerly Sections 6.4-6.5

**Preparation:** Review derivatives and inverse functions

**⏰ Reminders:**

✅ Complete Practice Problems 1-4

✅ Review the chain rule from calculus

✅ Think about when transformations are invertible

✅ Work hard!
:::
:::
:::

## ❓ Questions?

::: {style="font-size: 45px"}
::: {.callout-note}
## 💬 Open Discussion

**Key Topics for Discussion:**

- Why is the distribution function method considered the most general approach?

- How do fat tails in return distributions affect risk management decisions?

- What are some other financial quantities that are functions of random variables?

- When might we prefer simple returns over log returns?

- In practice, when would you use empirical methods vs theoretical transformation methods?
:::
:::
