---
title: "Mathematical Statistics"
subtitle: "The Hypergeometric and Poisson Distributions; Midterm I Review"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - Hypergeometric and Poisson; Midterm I Review"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
library(ggplot2)
theme_set(theme_minimal(base_size = 18))
```

## 🎯 Learning Objectives

::: {style="font-size: 32px"}
By the end of this lecture, you will be able to:

- **Recognise** sampling without replacement from a finite pool, and model it with the hypergeometric distribution

- **Compute** hypergeometric probabilities, mean and variance, and explain the factor $\frac{N-n}{N-1}$

- **Apply** the Poisson distribution to counts of events in time, including a Poisson process over $a$ units

- **Use** the Poisson as the limit of the binomial for rare events

- **Choose** among the five discrete models of Chapter 3 for Midterm I
:::

---

## 🗺️ Where We Are

::: {style="font-size: 32px"}
**Wackerly §3.7–3.8**

Saturday ended on one sentence: *the binomial fixes the trials; the geometric and negative binomial fix the successes and count the trials.* All three rest on the same two assumptions — **independent** trials with a **constant** $p$.

::: {.fragment}
Today we drop the assumptions themselves, one at a time. Draw **without replacement** from a small pool and independence fails: that is the hypergeometric. Let $n \to \infty$ while $p \to 0$ and there is no $n$ left to count: that is the Poisson.
:::
:::

---

## 🧾 Why the Binomial Fails in an Audit

::: {style="font-size: 30px"}
An auditor receives a batch of **25 invoices** from a Baku construction supplier. Unknown to her, **5** are misstated. She pulls **6** at random for testing.

::: {.fragment}
The first draw is misstated with probability $5/25 = 0.20$. If it was, the second is misstated with probability $4/24 = 0.167$; if it was not, $5/24 = 0.208$.
:::

::: {.fragment}
**The trials are not independent and $p$ is not constant**, so $Y$, the number of misstated invoices in the sample, is not binomial. It is a counting problem over subsets — Section 2.6 — and that is where the book derives the answer.
:::
:::

---

## 📐 Definition 3.10: Hypergeometric

::: {style="font-size: 30px"}
::: {.callout-important}
## Definition 3.10

A random variable $Y$ has a **hypergeometric probability distribution** if and only if
$$p(y) = \frac{\binom{r}{y}\binom{N-r}{n-y}}{\binom{N}{n}},$$
where $y$ is an integer $0, 1, 2, \ldots, n$, subject to $y \le r$ and $n - y \le N - r$.
:::

::: {.fragment}
Numerator: choose $y$ of the $r$ "red" elements **and** $n-y$ of the $N-r$ others (the $mn$ rule). Denominator: all $\binom{N}{n}$ equally likely samples. In R: `dhyper(y, r, N - r, n)`.
:::
:::

---

## 📊 Worked Example: The Audit Rule

::: {style="font-size: 30px"}
The firm's rule: **escalate to a full audit if two or more** of the 6 sampled invoices are misstated. With $N = 25$, $r = 5$, $n = 6$:

::: {.fragment}
$$p(0) = \frac{\binom{5}{0}\binom{20}{6}}{\binom{25}{6}} = \frac{38760}{177100} = 0.2189, \qquad p(1) = \frac{\binom{5}{1}\binom{20}{5}}{\binom{25}{6}} = \frac{77520}{177100} = 0.4377$$
:::

::: {.fragment}
$$P(Y \ge 2) = 1 - p(0) - p(1) = 1 - 0.2189 - 0.4377 = 0.3434$$
:::

::: {.fragment}
A batch with **one invoice in five** misstated still passes the test about two times in three. The complement did the work, exactly as in Chapter 2.
:::
:::

---

## 📐 Theorem 3.10: Mean and Variance

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorem 3.10

If $Y$ has a hypergeometric distribution,
$$\mu = E(Y) = \frac{nr}{N} \qquad \sigma^2 = V(Y) = n\left(\frac{r}{N}\right)\left(\frac{N-r}{N}\right)\left(\frac{N-n}{N-1}\right)$$
:::

::: {.fragment}
With $p = r/N$ this is $\mu = np$ and $\sigma^2 = npq\,\frac{N-n}{N-1}$: the binomial, times a **finite-population correction**.
:::

::: {.fragment}
Audit: $\mu = 6(5)/25 = 1.2$ and $\sigma^2 = 6(0.2)(0.8)(19/24) = 0.76$. The binomial would say $0.96$ — sampling without replacement removes about a fifth of the variance.
:::
:::

---

## 💻 When Does the Pool Size Matter?

```{r}
#| label: hyper-vs-binom
#| fig-height: 3.9
#| fig-width: 10
#| echo: false
y <- 0:6
pmf <- rbind(
  data.frame(y, p = dhyper(y, 5, 20, 6),    model = "Hypergeometric, N = 25"),
  data.frame(y, p = dhyper(y, 50, 200, 6),  model = "Hypergeometric, N = 250"),
  data.frame(y, p = dbinom(y, 6, 0.2),      model = "Binomial, p = 0.2"))
pmf$model <- factor(pmf$model, levels = unique(pmf$model))
ggplot(pmf, aes(factor(y), p, fill = model)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.8) +
  scale_fill_manual(values = c("#8b2635", "#cbb8a9", "#14130f")) +
  labs(x = "Misstated invoices in a sample of 6", y = "p(y)", fill = NULL) +
  theme(legend.position = "top", legend.text = element_text(size = 18))
```

::: {style="font-size: 28px"}
Same $r/N = 0.2$ throughout. At $N = 250$ the sample is 2.4% of the pool and the hypergeometric is indistinguishable from the binomial; at $N = 25$ it is 24% and the distribution is visibly tighter.
:::

---

## 🌧️ From Binomial to Poisson

::: {style="font-size: 30px"}
Split a day into $n$ tiny subintervals, each holding at most one event with probability $p$, independently. The count is binomial. Now let $n \to \infty$ with $\lambda = np$ fixed:
$$\lim_{n\to\infty}\binom{n}{y}p^y(1-p)^{n-y} = \frac{\lambda^y}{y!}e^{-\lambda}$$

::: {.callout-important}
## Definition 3.11

$Y$ has a **Poisson probability distribution** if and only if
$$p(y) = \frac{\lambda^y}{y!}e^{-\lambda}, \qquad y = 0, 1, 2, \ldots, \quad \lambda > 0.$$
:::
:::

---

## 📦 Worked Example: A Claims Desk

::: {style="font-size: 30px"}
The cargo-insurance desk of a Baku insurer receives on average $\lambda = 3$ claims per day, and can fully process **4** in a day. Let $Y$ be tomorrow's claim count.

::: {.fragment}
$$P(Y = 0) = e^{-3} = 0.0498 \quad P(Y = 1) = 3e^{-3} = 0.1494 \quad P(Y = 2) = 4.5e^{-3} = 0.2240$$
:::

::: {.fragment}
The desk falls behind when $Y \ge 5$:
$$P(Y \ge 5) = 1 - P(Y \le 4) = 1 - 0.8153 = 0.1847$$
:::

::: {.fragment}
Roughly **one working day in five** leaves a backlog. In R: `1 - ppois(4, 3)`, or Table 3 of Appendix 3.
:::
:::

---

## 📐 Theorem 3.11: Mean = Variance

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorem 3.11

If $Y$ has a Poisson distribution with parameter $\lambda$, then $\mu = E(Y) = \lambda$ and $\sigma^2 = V(Y) = \lambda$.
:::

::: {.fragment}
**Proof idea.** In $\sum_{y} y\,\lambda^y e^{-\lambda}/y!$ the $y = 0$ term vanishes; cancel $y$ against $y!$, factor out $\lambda$, and put $z = y - 1$. What remains is a Poisson sum, which equals 1. The variance is Exercise 3.138.
:::

::: {.fragment}
A practical test: if claim counts show a sample variance far above their mean, the Poisson model — not just its $\lambda$ — is in doubt.
:::
:::

---

## 🏧 A Poisson Process: One ATM

::: {style="font-size: 28px"}
If events occur as a Poisson process at $\lambda$ per unit, the count in $a$ units is Poisson with mean $a\lambda$. An ATM in Nizami Street averages **12 withdrawals per hour**.
:::

```{r}
#| label: atm
lambda_hr <- 12
hours <- c(1/6, 1/2, 1, 2)
data.frame(
  window     = c("10 min", "30 min", "1 hour", "2 hours"),
  mean       = hours * lambda_hr,
  P_none     = round(exp(-hours * lambda_hr), 4),
  P_over_30  = round(1 - ppois(30, hours * lambda_hr), 4))
```

::: {style="font-size: 28px"}
No withdrawal in 10 minutes: $e^{-2} = 0.1353$. A cassette good for 30 withdrawals runs out within 2 hours with probability $0.0958$.
:::

---

## ⚠️ The Poisson Limit: Rare Losses

::: {style="font-size: 28px"}
A bank's trading desk has a loss event (a failed settlement, a fat-finger trade) on any of **250** trading days with probability **0.008**, independently. Binomial, or Poisson with $\lambda = np = 2$?
:::

```{r}
#| label: oprisk
y <- 0:5
data.frame(y,
  binomial = round(dbinom(y, 250, 0.008), 4),
  poisson  = round(dpois(y, 2), 4))
```

::: {style="font-size: 28px"}
Agreement to the third decimal: $P(Y \ge 4)$ is $0.1421$ exact and $0.1429$ Poisson. The book's rule of thumb: large $n$, small $p$, $\lambda = np$ below about 7.
:::

---

## 🔬 Interactive: Binomial to Poisson {.smaller}

```{ojs}
//| echo: false

viewof n_sub = {
  const input = Inputs.range([4, 200], {value: 6, step: 1,
    label: "Number of trials n (λ = np = 2):"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}
```

```{ojs}
//| echo: false

lam = 2
binomPmf = (n, p) => {
  const out = [Math.pow(1 - p, n)];
  for (let k = 0; k < 10; k++) out.push(k + 1 > n ? 0 : out[k] * (n - k) / (k + 1) * p / (1 - p));
  return out;
}
poisPmf = Array.from({length: 11}, (_, k) => {
  let f = 1; for (let i = 2; i <= k; i++) f *= i;
  return Math.exp(-lam) * Math.pow(lam, k) / f;
})
bin = binomPmf(n_sub, lam / n_sub).map((p, k) => ({y: k, p}))
pois = poisPmf.map((p, k) => ({y: k, p}))
gap = d3.max(bin, (d, k) => Math.abs(d.p - pois[k].p))

md`With **n = ${n_sub}** and **p = ${(lam / n_sub).toFixed(4)}**, the largest gap between binomial bars and Poisson dots is **${gap.toFixed(4)}**.`

Plot.plot({
  width: 1150,
  height: 300,
  marginLeft: 78,
  marginBottom: 58,
  style: {fontSize: "18px"},
  x: {label: "y", domain: d3.range(0, 11), padding: 0.25},
  y: {label: "p(y)", domain: [0, 0.45], tickFormat: ".2f"},
  marks: [
    Plot.barY(bin, {x: "y", y: "p", fill: "#cbb8a9"}),
    Plot.dot(pois, {x: "y", y: "p", r: 8, fill: "#8b2635"}),
    Plot.ruleY([0])
  ]
})
```

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 30px"}
A bank's risk desk has recorded operational-loss events as a Poisson process at **2 per month** for years. Last quarter (3 months) there were **11**.

**Four minutes, in pairs:**

1. What is the distribution of $Y$, the number of events in a quarter?

2. Where does 11 fall relative to $\mu \pm 2\sigma$?

3. Is 11 alarming enough to report to the risk committee?
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 30px"}
1. A Poisson process over $a = 3$ months: $Y$ is Poisson with $\lambda^\star = 3 \times 2 = 6$.

2. Theorem 3.11: $\mu = 6$, $\sigma = \sqrt{6} = 2.45$, so $\mu + 2\sigma = 6 + 2(2.45) = 10.90$. The observed 11 lies **just outside**.

::: {.fragment}
3. Exactly: $P(Y \ge 11) = 1 - P(Y \le 10) = 1 - 0.9574 = 0.0426$, from `1 - ppois(10, 6)`.

Not impossible, but a one-in-25 quarter under the old rate — worth an investigation into whether $\lambda$ has risen, which is the reasoning of Example 3.22.
:::
:::

---

## 📝 Quiz #1: A Licence Lottery {.quiz-question}

The Energy Ministry awards 6 solar licences by lottery among 16 bidders, of whom 4 are joint ventures. What is the probability that **no** joint venture wins?

- [$\binom{12}{6}\big/\binom{16}{6} = 0.1154$]{.correct data-explanation="✅ Licences are drawn without replacement from a finite pool of 16, so Y is hypergeometric with N = 16, r = 4, n = 6: p(0) = 924/8008 = 0.1154."}
- $(0.75)^6 = 0.1780$
- $\binom{12}{6}\big/\binom{16}{4} = 0.5077$
- $12/16 = 0.75$

---

## 📝 Quiz #2: Scaling the Interval {.quiz-question}

Withdrawals at an ATM follow a Poisson process at 12 per hour. What is the probability of **no** withdrawal in a 5-minute window?

- [$e^{-1} = 0.368$]{.correct data-explanation="✅ Five minutes is a = 1/12 of an hour, so the count is Poisson with mean aλ = 1, and P(Y = 0) = e⁻¹ = 0.368."}
- $e^{-12} = 0.000006$
- $1 - e^{-1} = 0.632$
- $1/12 = 0.083$

---

## 🧭 Review: Which Discrete Model?

::: {style="font-size: 28px"}
| Model | $Y$ is… | $E(Y)$ | $V(Y)$ |
|--------|--------------|----|------|
| Binomial §3.4 | successes in $n$ independent trials | $np$ | $npq$ |
| Geometric §3.5 | the trial of the **first** success | $1/p$ | $q/p^2$ |
| Negative binomial §3.6 | the trial of the $r$-th success | $r/p$ | $rq/p^2$ |
| Hypergeometric §3.7 | "red" items in $n$ drawn **without replacement** | $nr/N$ | $npq\frac{N-n}{N-1}$ |
| Poisson §3.8 | events in a fixed interval | $\lambda$ | $\lambda$ |

::: {.fragment}
Ask in order: **is $n$ fixed?** Independent trials, binomial; a finite pool, hypergeometric. **Is the trial count the variable?** Geometric or negative binomial. **Only a rate?** Poisson.
:::
:::

---

## ✍️ Exam-Style Question 1

::: {style="font-size: 30px"}
Of a bank's 12 branches, 3 failed an internal compliance check. The Central Bank inspects 4 branches chosen at random. Find the probability that it finds **at least one** failing branch, and the mean and variance of the number it finds.

::: {.fragment}
Without replacement from $N = 12$ with $r = 3$, $n = 4$:
$$P(Y \ge 1) = 1 - \frac{\binom{3}{0}\binom{9}{4}}{\binom{12}{4}} = 1 - \frac{126}{495} = 0.7455$$
:::

::: {.fragment}
$\mu = 4(3)/12 = 1$ and $\sigma^2 = 4(0.25)(0.75)(8/11) = 0.5455$. Treating it as binomial would give $1 - 0.75^4 = 0.6836$ — wrong, and by 6 points.
:::
:::

---

## ✍️ Exam-Style Question 2

::: {style="font-size: 30px"}
Large claims (above 50,000 AZN) reach a reinsurer as a Poisson process at 0.5 per week. Call a 4-week month **quiet** if it has at most one large claim. Months are independent. Find the probability that at least 2 of the next 6 months are quiet.

::: {.fragment}
**Step 1 (§3.8).** A month has $\lambda = 4(0.5) = 2$: $w = P(Y \le 1) = e^{-2} + 2e^{-2} = 3e^{-2} = 0.4060$.
:::

::: {.fragment}
**Step 2 (§3.4).** Quiet months $X$ are binomial with $n = 6$, $p = w = 0.4060$:
$$P(X \ge 2) = 1 - (0.5940)^6 - 6(0.4060)(0.5940)^5 = 1 - 0.0439 - 0.1801 = 0.7759$$
:::
:::

---

## 📝 Quiz #3: Exam-Style — A Cost Built on a Count {.quiz-question}

Claims at a desk are Poisson with $\lambda = 4$ per day. Daily handling cost is $C = 500 + 800Y$ AZN. What is the **standard deviation** of $C$?

- [$1600$ AZN]{.correct data-explanation="✅ V(C) = 800² V(Y) = 640000 × 4 = 2,560,000, so σ = 800 × √4 = 1600 AZN. The constant 500 shifts the cost but adds no spread; E(C) = 500 + 800(4) = 3700 AZN."}
- $3200$ AZN
- $3700$ AZN
- $2100$ AZN

---

## 📋 Key Formulas

::: {style="font-size: 30px"}
| | Statement |
|---|---|
| Definition 3.10 | $p(y) = \binom{r}{y}\binom{N-r}{n-y}\big/\binom{N}{n}$ |
| Theorem 3.10 | $\mu = \frac{nr}{N}$, $\ \sigma^2 = n\frac{r}{N}\frac{N-r}{N}\frac{N-n}{N-1}$ |
| Definition 3.11 | $p(y) = \frac{\lambda^y}{y!}e^{-\lambda}$, $\ y = 0, 1, 2, \ldots$ |
| Theorem 3.11 | $\mu = \sigma^2 = \lambda$ |
| Poisson process | count in $a$ units is Poisson with mean $a\lambda$ |
| Poisson limit | $\binom{n}{y}p^y q^{n-y} \approx \frac{(np)^y}{y!}e^{-np}$, large $n$, small $p$ |
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- Drawing **without replacement** from a small pool breaks independence: the count is hypergeometric

- Its mean is the binomial's $np$; its variance is $npq$ shrunk by $\frac{N-n}{N-1}$, which matters when $n/N$ is not small

- Counts of rare events in an interval are Poisson, and the single parameter $\lambda$ is both the mean and the variance

- Over $a$ units of a Poisson process the mean is $a\lambda$; for large $n$ and small $p$ the binomial is close to Poisson($np$)

- For Midterm I, name the model **before** computing: fixed $n$? finite pool? trials to a success? only a rate?
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition**

- §3.7: Exercises 3.103, 3.105, 3.108, 3.112, 3.117
- §3.8: Exercises 3.121, 3.122, 3.126, 3.128, 3.134, 3.139

**Week 7, Problem Set 1** is open now and closes **Sunday 1 November at 23:59** on WeBWorK, covering §§3.7–3.8. It is the only set this week.

**Midterm Examination I** is on **Saturday 24 October**: Chapters 1–3, §§3.1–3.8.

**Next class:** 28 October, after the midterm — moments and moment-generating functions (Wackerly §3.9).
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**

📧 sorujov@ada.edu.az\
🏢 Building D, Room D325\
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- In the audit, how large must the batch $N$ be before the binomial answer is within 0.005 of the hypergeometric one?

- Why can a Poisson count exceed any bound, when a binomial count never exceeds $n$?

- Two independent ATMs average 12 and 8 withdrawals an hour. What would you guess the distribution of their total is — and how would you check it?
:::
