---
title: "Mathematical Statistics"
subtitle: "Expected Values for Continuous Random Variables"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics - Expected Values & Uniform Distribution"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::: {style="font-size: 30px"}
::: {.learning-objectives}
By the end of this lecture, you will be able to:

- Define and compute expected values for continuous random variables using integration techniques

- Apply theorems on expectations to calculate means and variances for continuous distributions

- Recognize and work with the continuous uniform distribution, including its density function and parameters

- Calculate probabilities for uniformly distributed random variables in business contexts (e.g., service arrival times, waiting times)

- Derive and apply formulas for mean and variance of the uniform distribution in financial and operational scenarios
:::
:::

## 📋 Overview {.smaller}

::: {style="font-size:38px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Expected Values** – Definition and computation for continuous random variables
- **Expectation Theorems** – Linearity properties and variance calculations
- **Uniform Distribution** – Definition, parameters, and empirical interpretation
- **Uniform Applications** – Poisson-arrival conditioning and service time modeling
- **Statistical Properties** – Mean and variance derivations for uniform distributions
:::
:::
:::

## 📖 Definition: Expected Value {.smaller}

::: {.callout-note}
## 📝 Definition 1: Expected Value of a Continuous Random Variable

The **expected value** (or mean) of a continuous random variable $Y$ with probability density function $f(y)$ is:

$$\boxed{E(Y) = \int_{-\infty}^{\infty} yf(y) \, dy}$$

provided that the integral exists (i.e., converges absolutely).

::: {.incremental}
- **Interpretation** – The expected value represents the "center of mass" or long-run average of the distribution

- **Financial Context** – For a continuous return distribution, $E(Y)$ represents the expected return on an investment

- **Existence Condition** – The integral must converge; distributions with "heavy tails" may not have finite expected values
:::
:::

## 🧮 Theorem: Expected Value of Functions {.smaller}

::: {.callout-important}
## Theorem 1: Expectation of Transformed Variables

Let $g(Y)$ be a function of a continuous random variable $Y$ with density $f(y)$. Then:

$$\boxed{E[g(Y)] = \int_{-\infty}^{\infty} g(y) f(y) \, dy}$$

provided the integral exists.
:::

. . .

**Key Insight:** You do NOT need to find the distribution of $g(Y)$ first – you can compute expectations directly using the original density $f(y)$.

. . .

**Example Applications:**

- $g(Y) = Y^2$ gives $E(Y^2)$, used to compute variance: $V(Y) = E(Y^2) - [E(Y)]^2$

- $g(Y) = e^{tY}$ gives the moment generating function $M(t) = E(e^{tY})$

- $g(Y) = (Y - \mu)^2$ gives variance directly: $V(Y) = E[(Y - \mu)^2]$

## 🧮 Theorem: Properties of Expectations {.smaller}

::: {.callout-important}
## Theorem 2: Linearity of Expectations

Let $c$ be a constant, and let $g(Y), g_1(Y), g_2(Y), \ldots, g_k(Y)$ be functions of continuous random variable $Y$. Then:
:::

. . .

**Property 1 (Constant):**
$$E(c) = c$$

. . .

**Property 2 (Scalar Multiplication):**
$$E[cg(Y)] = cE[g(Y)]$$

. . .

**Property 3 (Additivity):**
$$E[g_1(Y) + g_2(Y) + \ldots + g_k(Y)] = E[g_1(Y)] + E[g_2(Y)] + \ldots + E[g_k(Y)]$$

. . .

**Portfolio Application:** If a portfolio consists of $k$ assets with returns $R_1, R_2, \ldots, R_k$ and weights $w_1, w_2, \ldots, w_k$, then the expected portfolio return is $E(R_p) = w_1E(R_1) + w_2E(R_2) + \ldots + w_kE(R_k)$.

## 📌 Example 1: Computing Mean and Variance {.smaller}

::: {style="font-size:28px"}
**Problem:** A random variable $Y$ has density function:
$$f(y) = \frac{3}{8}y^2, \quad 0 \leq y \leq 2, \quad f(y) = 0 \text{ elsewhere}$$

Find the mean $\mu = E(Y)$ and variance $\sigma^2 = V(Y)$.

. . .

**Solution (Part 1 - Mean):**

By definition:
$$E(Y) = \int_{-\infty}^{\infty} y f(y) \, dy = \int_{0}^{2} y \left(\frac{3}{8}y^2\right) \, dy = \frac{3}{8} \int_{0}^{2} y^3 \, dy$$

. . .

$$= \frac{3}{8} \left[\frac{1}{4}y^4\right]_{0}^{2} = \frac{3}{8} \cdot \frac{16}{4} = \frac{3}{8} \cdot 4 = \boxed{1.5}$$

:::

## 📌 Example 1: Computing Mean and Variance (Part 2) {.smaller}

::: {style="font-size:28px"}
**Solution (Part 2 - Variance):**

First, compute $E(Y^2)$:
$$E(Y^2) = \int_{-\infty}^{\infty} y^2 f(y) \, dy = \int_{0}^{2} y^2 \left(\frac{3}{8}y^2\right) \, dy = \frac{3}{8} \int_{0}^{2} y^4 \, dy$$

. . .

$$= \frac{3}{8} \left[\frac{1}{5}y^5\right]_{0}^{2} = \frac{3}{8} \cdot \frac{32}{5} = \frac{96}{40} = 2.4$$

. . .

Now apply the variance formula:
$$\sigma^2 = V(Y) = E(Y^2) - [E(Y)]^2 = 2.4 - (1.5)^2 = 2.4 - 2.25 = \boxed{0.15}$$

. . .

**Interpretation:** The standard deviation is $\sigma = \sqrt{0.15} \approx 0.387$, indicating moderate variability around the mean of 1.5.
:::

## 📖 Definition: Uniform Distribution {.smaller}

::: {.callout-note}
## 📝 Definition 2: Continuous Uniform Distribution

If $\theta_1 < \theta_2$, a random variable $Y$ is said to have a **continuous uniform probability distribution** on the interval $(\theta_1, \theta_2)$ if and only if the density function is:

$$\boxed{f(y) = \begin{cases}
\frac{1}{\theta_2 - \theta_1}, & \text{if } \theta_1 \leq y \leq \theta_2, \\
0, & \text{elsewhere}
\end{cases}}$$

**Notation:** $Y \sim \text{Uniform}(\theta_1, \theta_2)$ or $Y \sim U(\theta_1, \theta_2)$
:::

. . .

::: {.callout-note}
## 📝 Definition 3: Parameters

The constants that determine the specific form of a density function are called **parameters** of the density function. For the uniform distribution, $\theta_1$ and $\theta_2$ are the parameters.
:::

## 🎮 Interactive: Visualizing the Uniform Distribution {.smaller}

::: {style="font-size: 0.8em;"}

**Explore the Uniform PDF:** Adjust the interval boundaries to see how the density changes.

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof theta1 = Inputs.range([0, 20], {
  value: 2, 
  step: 0.5, 
  label: "θ₁ (lower bound):"
})

viewof theta2 = Inputs.range([0, 20], {
  value: 8, 
  step: 0.5, 
  label: "θ₂ (upper bound):"
})

height_uniform = 1 / (theta2 - theta1)

md`**Current Parameters:**  
θ₁ = ${theta1}  
θ₂ = ${theta2}  
PDF height = ${height_uniform.toFixed(4)}`
```

**Observations:**  
As the interval width increases, the PDF height decreases to maintain total probability = 1.

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// Generate uniform PDF data
uniform_data = [
  {x: theta1 - 1, y: 0},
  {x: theta1, y: 0},
  {x: theta1, y: height_uniform},
  {x: theta2, y: height_uniform},
  {x: theta2, y: 0},
  {x: theta2 + 1, y: 0}
]

// Create shaded area data
shaded_area = d3.range(theta1, theta2 + 0.01, (theta2 - theta1) / 100).map(x => ({
  x: x,
  y: height_uniform
}))

Plot.plot({
  width: 800,
  height: 450,
  marginLeft: 60,
  marginBottom: 50,
  x: {
    label: "Y",
    grid: true,
    domain: [theta1 - 2, theta2 + 2]
  },
  y: {
    label: "f(y)",
    domain: [0, Math.max(0.5, height_uniform * 1.2)]
  },
  marks: [
    Plot.areaY(shaded_area, {x: "x", y: "y", fill: "steelblue", opacity: 0.3}),
    Plot.line(uniform_data, {x: "x", y: "y", stroke: "steelblue", strokeWidth: 3}),
    Plot.ruleY([0]),
    Plot.text([{x: (theta1 + theta2) / 2, y: height_uniform / 2, label: `Area = 1`}], {
      x: "x", 
      y: "y", 
      text: "label", 
      fill: "darkblue",
      fontSize: 16,
      fontWeight: "bold"
    })
  ],
  caption: html`Uniform PDF with constant density = 1 / (θ₂ - θ₁)`
})
```

:::
:::
:::

## 📖 Properties of the Uniform Distribution {.smaller}

::: {style="font-size:32px"}

**Key Characteristics:**

- **Constant Density** – The PDF has the same value throughout the interval $[\theta_1, \theta_2]$, representing "complete uncertainty" within the range

- **Equal Probability** – Any subinterval of fixed length has the same probability, regardless of location within $[\theta_1, \theta_2]$

- **Probability Calculation** – For any interval $[a, b] \subseteq [\theta_1, \theta_2]$:
$$P(a \leq Y \leq b) = \int_{a}^{b} \frac{1}{\theta_2 - \theta_1} dy = \frac{b - a}{\theta_2 - \theta_1}$$

- **Financial Interpretation** – Models scenarios with "equal ignorance" – when all outcomes in a range are equally likely (e.g., random stock selection from a portfolio, uniform price fluctuations within a band)

:::

## 📌 Example 2: Probability with Uniform Distribution {.smaller}

**Problem:** Consider a continuous uniform distribution with parameters $a = 2$ and $b = 8$. What is the probability that the random variable falls within the interval $[4, 6]$?

. . .

**Solution:**

The probability density function is:
$$f(x; 2, 8) = \frac{1}{8 - 2} = \frac{1}{6} \quad \text{for } 2 \leq x \leq 8$$

. . .

To find $P(4 \leq X \leq 6)$, integrate the PDF over the interval:
$$P(4 \leq X \leq 6) = \int_{4}^{6} f(x; 2, 8) \, dx = \int_{4}^{6} \frac{1}{6} \, dx$$

. . .

$$= \left[\frac{1}{6}x\right]_{4}^{6} = \frac{6}{6} - \frac{4}{6} = \frac{2}{6} = \boxed{\frac{1}{3}}$$

. . .

**Alternative Method:** Since probability is proportional to length: $P(4 \leq X \leq 6) = \frac{6-4}{8-2} = \frac{2}{6} = \frac{1}{3}$

## 📖 Application: Poisson Process & Uniform Conditioning {.smaller}

::: {.callout-tip}
## 🎯 Real-World Application

**Context:** In operations research and queueing theory, the arrival of events (customers, phone calls, server requests) often follows a Poisson process.

**Key Result:** If exactly one event occurred in the time interval $(0, t)$ under a Poisson process, then the **actual time of occurrence** is uniformly distributed over $(0, t)$.

**Examples:**

- **Call Centers** – If one customer called during a 60-minute hour, the call time is $\text{Uniform}(0, 60)$

- **Network Traffic** – If one packet arrived in a 100ms window, arrival time is uniformly distributed

- **Service Operations** – Conditional distribution of service completion time given exactly one completion
:::

## 📌 Example 3: Poisson Arrivals & Uniform Conditioning {.smaller}

**Problem:** Arrivals of customers at a checkout counter follow a Poisson distribution. During a given 30-minute period, exactly one customer arrived. Find the probability that the customer arrived during the last 5 minutes.

. . .

**Solution:**

Given that exactly one arrival occurred in $(0, 30)$, the arrival time $Y$ is uniformly distributed:
$$Y \sim \text{Uniform}(0, 30)$$

. . .

The probability of arrival in the last 5 minutes $[25, 30]$ is:
$$P(25 \leq Y \leq 30) = \int_{25}^{30} \frac{1}{30} \, dy = \frac{30 - 25}{30} = \frac{5}{30} = \boxed{\frac{1}{6}}$$

. . .

**Interpretation:** Any 5-minute interval has probability $\frac{1}{6}$ of containing the arrival. The probability is proportional to the interval length: $\frac{5}{30} = \frac{1}{6}$.

## 🧮 Theorem: Mean of Uniform Distribution {.smaller}

::: {.callout-important}
## Theorem 3: Expected Value of Uniform Distribution

If $\theta_1 < \theta_2$ and $Y \sim \text{Uniform}(\theta_1, \theta_2)$, then:

$$\boxed{\mu = E(Y) = \frac{\theta_1 + \theta_2}{2}}$$
:::

. . .

**Proof:**

By definition of expected value:
$$E(Y) = \int_{-\infty}^{\infty} y f(y) \, dy = \int_{\theta_1}^{\theta_2} y \left(\frac{1}{\theta_2 - \theta_1}\right) \, dy$$

. . .

$$= \frac{1}{\theta_2 - \theta_1} \int_{\theta_1}^{\theta_2} y \, dy = \frac{1}{\theta_2 - \theta_1} \left[\frac{1}{2}y^2\right]_{\theta_1}^{\theta_2}$$

. . .

$$= \frac{1}{\theta_2 - \theta_1} \cdot \frac{1}{2}(\theta_2^2 - \theta_1^2) = \frac{\theta_2^2 - \theta_1^2}{2(\theta_2 - \theta_1)}$$

. . .

$$= \frac{(\theta_2 - \theta_1)(\theta_2 + \theta_1)}{2(\theta_2 - \theta_1)} = \boxed{\frac{\theta_1 + \theta_2}{2}}$$

## 🧮 Theorem: Variance of Uniform Distribution {.smaller}

::: {.callout-important}
## Theorem 4: Variance of Uniform Distribution

If $\theta_1 < \theta_2$ and $Y \sim \text{Uniform}(\theta_1, \theta_2)$, then:

$$\boxed{\sigma^2 = V(Y) = \frac{(\theta_2 - \theta_1)^2}{12}}$$
:::

. . .

**Proof Strategy:**

1. Compute $E(Y^2) = \int_{\theta_1}^{\theta_2} y^2 \cdot \frac{1}{\theta_2 - \theta_1} \, dy$

. . .

2. Use the formula $V(Y) = E(Y^2) - [E(Y)]^2$

. . .

**Step 1: Calculate $E(Y^2)$**
$$E(Y^2) = \frac{1}{\theta_2 - \theta_1} \int_{\theta_1}^{\theta_2} y^2 \, dy = \frac{1}{\theta_2 - \theta_1} \left[\frac{1}{3}y^3\right]_{\theta_1}^{\theta_2}$$

. . .

$$= \frac{\theta_2^3 - \theta_1^3}{3(\theta_2 - \theta_1)} = \frac{(\theta_2 - \theta_1)(\theta_2^2 + \theta_2\theta_1 + \theta_1^2)}{3(\theta_2 - \theta_1)} = \frac{\theta_2^2 + \theta_2\theta_1 + \theta_1^2}{3}$$

## 🧮 Variance Proof (Continued) {.smaller}

**Step 2: Apply Variance Formula**

We have:
$$E(Y^2) = \frac{\theta_2^2 + \theta_2\theta_1 + \theta_1^2}{3}$$
$$[E(Y)]^2 = \left(\frac{\theta_1 + \theta_2}{2}\right)^2 = \frac{(\theta_1 + \theta_2)^2}{4} = \frac{\theta_1^2 + 2\theta_1\theta_2 + \theta_2^2}{4}$$

. . .

Therefore:
$$V(Y) = E(Y^2) - [E(Y)]^2 = \frac{\theta_2^2 + \theta_2\theta_1 + \theta_1^2}{3} - \frac{\theta_1^2 + 2\theta_1\theta_2 + \theta_2^2}{4}$$

. . .

Finding common denominator (12):
$$= \frac{4(\theta_2^2 + \theta_2\theta_1 + \theta_1^2) - 3(\theta_1^2 + 2\theta_1\theta_2 + \theta_2^2)}{12}$$

. . .

$$= \frac{4\theta_2^2 + 4\theta_2\theta_1 + 4\theta_1^2 - 3\theta_1^2 - 6\theta_1\theta_2 - 3\theta_2^2}{12}$$

. . .

$$= \frac{\theta_2^2 - 2\theta_2\theta_1 + \theta_1^2}{12} = \frac{(\theta_2 - \theta_1)^2}{12}$$

## 📌 Example 4: Uniform Distribution Mean & Variance {.smaller}

**Problem:** A stock price is equally likely to take any value between $\$50$ and $\$70$ during a trading session. Model this as a uniform distribution and find:

(a) The expected price

(b) The variance and standard deviation

. . .

**Solution:**

Let $Y \sim \text{Uniform}(50, 70)$ represent the stock price.

. . .

**(a) Expected price:**
$$\mu = E(Y) = \frac{\theta_1 + \theta_2}{2} = \frac{50 + 70}{2} = \boxed{\$60}$$

. . .

**(b) Variance and standard deviation:**
$$\sigma^2 = V(Y) = \frac{(\theta_2 - \theta_1)^2}{12} = \frac{(70 - 50)^2}{12} = \frac{400}{12} = \boxed{33.33}$$

$$\sigma = \sqrt{33.33} \approx \boxed{\$5.77}$$

## 🎮 Interactive: Uniform Mean and Variance {.smaller}

::: {style="font-size: 0.8em;"}

**Explore Mean and Variance:** Adjust the uniform distribution parameters to see how they affect the mean and variance.

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof a_unif = Inputs.range([0, 30], {
  value: 2, 
  step: 1, 
  label: "θ₁ (lower):"
})

viewof b_unif = Inputs.range([0, 30], {
  value: 18, 
  step: 1, 
  label: "θ₂ (upper):"
})

mean_unif = (a_unif + b_unif) / 2
var_unif = Math.pow(b_unif - a_unif, 2) / 12
sd_unif = Math.sqrt(var_unif)

html`<div style="font-size: 14px; line-height: 1.8;">
  <p><strong>Distribution:</strong><br>
  Y ~ Uniform(${a_unif}, ${b_unif})</p>
  
  <p><strong>Parameters:</strong><br>
  Mean μ = ${mean_unif.toFixed(2)}<br>
  Variance σ² = ${var_unif.toFixed(2)}<br>
  Std Dev σ = ${sd_unif.toFixed(2)}</p>
  
  <p><strong>Observation:</strong><br>
  Variance increases with the square of interval width!</p>
</div>`
```

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// Generate PDF data with explicit boundaries (like first interactive)
pdf_height2 = 1 / (b_unif - a_unif)

uniform_pdf_data2 = [
  {x: a_unif - 5, y: 0},
  {x: a_unif, y: 0},
  {x: a_unif, y: pdf_height2},
  {x: b_unif, y: pdf_height2},
  {x: b_unif, y: 0},
  {x: b_unif + 5, y: 0}
]

// Create shaded area data
shaded_area2 = d3.range(a_unif, b_unif + 0.01, (b_unif - a_unif) / 100).map(x => ({
  x: x,
  y: pdf_height2
}))

// Markers for mean and standard deviations
markers2 = [
  {x: mean_unif, y: pdf_height2 / 2, label: "μ", color: "red"},
  {x: mean_unif - sd_unif, y: pdf_height2 * 0.7, label: "μ - σ", color: "orange"},
  {x: mean_unif + sd_unif, y: pdf_height2 * 0.7, label: "μ + σ", color: "orange"}
]

Plot.plot({
  width: 800,
  height: 450,
  marginLeft: 60,
  x: {
    label: "Y",
    grid: true
  },
  y: {
    label: "Density f(y)",
    domain: [0, pdf_height2 * 1.3]
  },
  marks: [
    Plot.areaY(shaded_area2, {x: "x", y: "y", fill: "steelblue", opacity: 0.3}),
    Plot.line(uniform_pdf_data2, {x: "x", y: "y", stroke: "steelblue", strokeWidth: 3}),
    Plot.ruleX([mean_unif], {stroke: "red", strokeWidth: 2, strokeDasharray: "5,3"}),
    Plot.ruleX([mean_unif - sd_unif, mean_unif + sd_unif], {stroke: "orange", strokeWidth: 1.5, strokeDasharray: "3,3"}),
    Plot.dot(markers2, {x: "x", y: "y", fill: "color", r: 6, stroke: "white", strokeWidth: 2}),
    Plot.text(markers2, {x: "x", y: "y", text: "label", dy: -12, fontSize: 14, fontWeight: "bold"}),
    Plot.ruleY([0])
  ],
  caption: html`<span style="color: red;">━━</span> Mean | <span style="color: orange;">━ ━</span> ± 1 Standard Deviation`
})
```

:::
:::
:::

## 💰 Case Study: Investment Holding Period Returns (Real Data) {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}
::: {.callout-note}
## 📊 Investment Scenario

**Context**: An investor holds a tech stock (AAPL - Apple Inc.) for a short trading period. We analyze the intraday price range to model uncertainty using the uniform distribution.

**Key Questions**:

- What is the expected (average) price during the trading window?

- What is the price variability (standard deviation)?

- What is the probability the price falls within a target range?
:::
:::

::: {.column width="50%"}
::: {.callout-tip}
## 📊 Data Source

We analyze **Apple Inc. (AAPL)** daily trading data from the last 30 trading days.

**Source**: Yahoo Finance API (via quantmod R package)

**Period**: Last 30 trading days from today

**Data Quality**: High-frequency market data (OHLC - Open, High, Low, Close)

**Verification**: Cross-checked with NASDAQ official records
:::
:::
:::
:::

## 💰 Case Study: Data Analysis & Uniform Modeling {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Load required libraries
library(quantmod)
library(dplyr)
library(lubridate)

# Fetch AAPL data (last 30 days)
getSymbols("AAPL", src = "yahoo", 
           from = Sys.Date() - 40, 
           to = Sys.Date(), auto.assign = TRUE)

# Extract data
aapl_data <- as.data.frame(AAPL)
aapl_data$Date <- as.Date(rownames(aapl_data))

# Focus on most recent day
recent_day <- tail(aapl_data, 1)

# Extract High and Low (range)
low_price <- as.numeric(recent_day$AAPL.Low)
high_price <- as.numeric(recent_day$AAPL.High)

cat("Recent Trading Day Analysis\n")
cat("============================\n")
cat("Date:", format(recent_day$Date, "%Y-%m-%d"), "\n")
cat("Low Price: $", round(low_price, 2), "\n")
cat("High Price: $", round(high_price, 2), "\n")
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Model as Uniform(low, high)
# Calculate mean and variance
mean_price <- (low_price + high_price) / 2
var_price <- (high_price - low_price)^2 / 12
sd_price <- sqrt(var_price)

cat("\nUniform Distribution Model\n")
cat("===========================\n")
cat("Y ~ Uniform(", round(low_price, 2), 
    ", ", round(high_price, 2), ")\n\n", sep = "")

cat("Expected Price (μ): $", 
    round(mean_price, 2), "\n")
cat("Variance (σ²): ", 
    round(var_price, 2), "\n")
cat("Std Deviation (σ): $", 
    round(sd_price, 2), "\n\n")

# Probability within 1 std dev of mean
lower_bound <- mean_price - sd_price
upper_bound <- mean_price + sd_price
prob_within_1sd <- (upper_bound - lower_bound) / 
                   (high_price - low_price)

cat("P(μ - σ ≤ Y ≤ μ + σ) = ", 
    round(prob_within_1sd, 4), "\n")
```

:::
:::
:::

## 💰 Case Study: Visualization {.smaller}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 11
#| fig-height: 5

library(ggplot2)

# Create data for uniform PDF
x_vals <- seq(low_price - 2, high_price + 2, length.out = 1000)
density_vals <- ifelse(x_vals >= low_price & x_vals <= high_price, 
                       1/(high_price - low_price), 0)
plot_data <- data.frame(x = x_vals, density = density_vals)

# Create plot
ggplot(plot_data, aes(x = x, y = density)) +
  geom_area(fill = "steelblue", alpha = 0.3) +
  geom_line(color = "steelblue", size = 1.5) +
  geom_vline(xintercept = mean_price, color = "red", linetype = "dashed", size = 1.2) +
  geom_vline(xintercept = c(lower_bound, upper_bound), 
             color = "orange", linetype = "dotted", size = 1) +
  annotate("text", x = mean_price, y = max(density_vals) * 0.5, 
           label = paste0("Mean\n$", round(mean_price, 2)), 
           color = "red", fontface = "bold", size = 3.5) +
  annotate("text", x = lower_bound, y = max(density_vals) * 0.8, 
           label = paste0("μ - σ"), color = "orange", fontface = "bold", size = 3) +
  annotate("text", x = upper_bound, y = max(density_vals) * 0.8, 
           label = paste0("μ + σ"), color = "orange", fontface = "bold", size = 3) +
  labs(title = paste0("Uniform Distribution Model: AAPL Intraday Price (", 
                      format(recent_day$Date, "%Y-%m-%d"), ")"),
       subtitle = paste0("Assuming equal likelihood across trading range [$", 
                        round(low_price, 2), ", $", round(high_price, 2), "]"),
       x = "Stock Price ($)", y = "Probability Density") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11),
        axis.title = element_text(size = 12))
```

## 💰 Case Study: Financial Interpretation {.smaller}

::: {style="font-size:28px"}

**Key Insights from AAPL Data:**

- **Uniform Assumption**: The model assumes equal likelihood for all prices within the day's range – a simplification suitable for short-term uncertainty analysis

- **Expected Price**: The mean $\mu = \frac{\theta_1 + \theta_2}{2}$ represents the midpoint of the trading range, serving as a central estimate

- **Price Volatility**: The standard deviation $\sigma = \frac{\theta_2 - \theta_1}{2\sqrt{3}} \approx 0.289 \times (\text{Range})$ quantifies intraday price variability

- **Probability Calculations**: For any target price range $[a, b]$ within the day's range, probability is simply $P(a \leq Y \leq b) = \frac{b - a}{\theta_2 - \theta_1}$

**Limitations:**

- Real intraday prices are NOT uniformly distributed (they follow complex patterns)

- This model is most appropriate when we have minimal information beyond the range

- More sophisticated models (e.g., log-normal, jump-diffusion) are used for rigorous financial modeling

:::

## 📝 Quiz #1: Expected Value Definition {.quiz-question}

What is the correct formula for the expected value of a continuous random variable $Y$ with density $f(y)$?

- [$E(Y) = \int_{-\infty}^{\infty} yf(y) \, dy$]{.correct data-explanation="✅ Correct! This is the definition of expected value for continuous random variables."}
- $E(Y) = \int_{-\infty}^{\infty} f(y) \, dy$
- $E(Y) = \sum y f(y)$
- $E(Y) = \int_{-\infty}^{\infty} y^2 f(y) \, dy$

## 📝 Quiz #2: Uniform Distribution Parameters {.quiz-question}

For a uniform distribution on the interval $[3, 9]$, what is the value of the probability density function within the interval?

- [$f(y) = \frac{1}{6}$]{.correct data-explanation="✅ Correct! The density is 1 divided by the interval width: 1/(9-3) = 1/6."}
- $f(y) = \frac{1}{3}$
- $f(y) = \frac{1}{9}$
- $f(y) = 6$

## 📝 Quiz #3: Mean of Uniform Distribution {.quiz-question}

A random variable $Y$ follows a uniform distribution on $[10, 30]$. What is $E(Y)$?

- [$E(Y) = 20$]{.correct data-explanation="✅ Correct! The mean is the midpoint: (10 + 30)/2 = 20."}
- $E(Y) = 10$
- $E(Y) = 15$
- $E(Y) = 25$

## 📝 Quiz #4: Variance Formula {.quiz-question}

Which formula correctly expresses the variance of a random variable $Y$?

- [$V(Y) = E(Y^2) - [E(Y)]^2$]{.correct data-explanation="✅ Correct! This is the computational formula for variance."}
- $V(Y) = E(Y)^2 - E(Y^2)$
- $V(Y) = E(Y^2) + [E(Y)]^2$
- $V(Y) = [E(Y)]^2$

## 📝 Quiz #5: Uniform Distribution Probability {.quiz-question}

For $Y \sim \text{Uniform}(0, 10)$, what is $P(3 \leq Y \leq 7)$?

- [$P(3 \leq Y \leq 7) = 0.4$]{.correct data-explanation="✅ Correct! Probability equals length ratio: (7-3)/(10-0) = 4/10 = 0.4."}
- $P(3 \leq Y \leq 7) = 0.3$
- $P(3 \leq Y \leq 7) = 0.5$
- $P(3 \leq Y \leq 7) = 0.7$

## 📝 Summary {.smaller}

::: {.summary-box}
**✅ Key Takeaways**

- **Expected values** for continuous random variables are computed using integration: $E(Y) = \int yf(y)dy$, with linearity properties enabling efficient calculation of means and variances

- **Expectation of functions** $E[g(Y)]$ can be computed directly from the original density without deriving the distribution of $g(Y)$

- The **uniform distribution** $Y \sim \text{Uniform}(\theta_1, \theta_2)$ has constant density $\frac{1}{\theta_2 - \theta_1}$ over its interval, representing complete uncertainty

- **Mean and variance** of uniform distribution are $\mu = \frac{\theta_1 + \theta_2}{2}$ (midpoint) and $\sigma^2 = \frac{(\theta_2 - \theta_1)^2}{12}$

- **Poisson conditioning**: When exactly one event occurs in a Poisson process over $(0, t)$, the occurrence time is uniformly distributed on $(0, t)$

- **Financial applications** include modeling price ranges, arrival times in service operations, and scenarios with equal likelihood across an interval
:::

## 📚 Practice Problems {.smaller}

::: {.callout-tip}
## 📝 Homework Problems

**Problem 1 (Variance Derivation):** Complete the derivation of $V(Y) = \frac{(\theta_2 - \theta_1)^2}{12}$ for the uniform distribution. Show all algebraic steps clearly.

**Problem 2 (Investment Analysis):** A commodity price is uniformly distributed between $\$80$ and $\$120$. Find: (a) the expected price, (b) the probability the price exceeds $\$100$, and (c) the standard deviation of the price.

**Problem 3 (Expectation Computation):** Let $Y$ have density $f(y) = 2y$ for $0 \leq y \leq 1$. Find $E(Y)$, $E(Y^2)$, and $V(Y)$.

**Problem 4 (Poisson Application):** During a 2-hour period, exactly 3 customers arrived at a bank following a Poisson process. What is the probability that all 3 arrivals occurred in the first hour? (Hint: Use properties of uniform distribution conditioning.)

**Problem 5 (Function of Random Variable):** For $Y \sim \text{Uniform}(0, 1)$, find $E(Y^3)$ and $E(e^Y)$.
:::

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov  
Assistant Professor  
School of Business  
ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)  
🏢 Office: D312  
⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Normal Distribution and Properties

**Reading:** Chapter on Normal Distribution in textbook

**Preparation:** Review integration techniques and standard normal tables

**⏰ Reminders:**

✅ Complete practice problems (due next class)

✅ Review uniform distribution formulas

✅ Work hard and stay curious!
:::
:::

## ❓ Questions? {.smaller .center}

::: {.callout-note}
## 💬 Open Discussion

**Discussion Topics:**

- How do we know when the expected value integral converges?

- What other real-world scenarios can be modeled with the uniform distribution?

- How does the uniform distribution relate to other continuous distributions we'll study?

- What are the limitations of assuming uniform distribution in financial modeling?

- How can we extend these concepts to multivariate continuous random variables?
:::
