---
title: "Mathematical Statistics"
subtitle: "Probability-Generating Functions, Tchebysheff's Theorem, and Chapter 3 in Review"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - PGFs, Tchebysheff and Chapter 3 Review"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
library(ggplot2)
theme_set(theme_minimal(base_size = 18))
```

## 🎯 Learning Objectives

::: {style="font-size: 32px"}
By the end of this lecture, you will be able to:

- **Write** the probability-generating function $P(t) = E(t^Y)$ of a count, and read probabilities off its coefficients

- **Differentiate** $P(t)$ at $t = 1$ to obtain factorial moments, and from them $E(Y)$ and $V(Y)$

- **Bound** $P(|Y - \mu| < k\sigma)$ with Tchebysheff's theorem when only $\mu$ and $\sigma$ are known

- **Invert** the bound to size a buffer: find $C$ with $P(|Y-\mu| \ge C)$ below a target

- **Choose** among the Chapter 3 models for a count on a bank's risk desk
:::

---

## 🗺️ Where We Are

::: {style="font-size: 30px"}
**Wackerly §§3.10–3.11**

Wednesday ended on **uniqueness**: recognise the MGF $m(t) = E(e^{tY})$, and you know the distribution. It stored the moments.

::: {.fragment}
Today, two more tools, and then the chapter closes:

1. A **cousin** of $m(t)$ built for counts, $P(t) = E(t^Y)$, whose coefficients are the probabilities themselves. §3.10 is optional in the book, so we meet it briefly.
2. **Tchebysheff's theorem**, which needs no model at all: only $\mu$ and $\sigma$. This one is load-bearing.
:::
:::

---

## ❓ Motivating Question

::: {.callout-important}
## The risk desk's problem
A Baku bank's SME book has averaged **12 defaults a month**, with a standard deviation of **3**, over five years. Nobody on the desk is willing to say the count is binomial, Poisson, or anything else.

The regulator asks: *how confident are you that next month's defaults stay between 6 and 18?*
:::

::: {.fragment style="font-size: 30px"}
Without a distribution there is no $p(y)$ to add up. By the end of today there will still be an answer, and it will be a guarantee.
:::

---

## 📐 Definition 3.15: The PGF

::: {style="font-size: 30px"}
::: {.callout-note}
## Definition 3.15
Let $Y$ be an integer-valued random variable with $P(Y = i) = p_i$, $i = 0, 1, 2, \ldots$ The **probability-generating function** of $Y$ is
$$P(t) = E(t^Y) = p_0 + p_1 t + p_2 t^2 + \cdots = \sum_{i=0}^{\infty} p_i t^i$$
for all $t$ such that $P(t)$ is finite.
:::

::: {.fragment}
The name is literal: **the coefficient of $t^i$ is $P(Y = i)$**. Expand $P(t)$ as a series and the probability function falls out. Compare $m(t)$, whose coefficients are moments.
:::
:::

---

## 📐 Theorem 3.13: Factorial Moments

::: {style="font-size: 30px"}
::: {.callout-note}
## Definition 3.16
The $k$th **factorial moment** of $Y$ is $\mu_{[k]} = E[Y(Y-1)(Y-2)\cdots(Y-k+1)]$.
:::

::: {.callout-important}
## Theorem 3.13
$$\left.\frac{d^k P(t)}{dt^k}\right|_{t=1} = P^{(k)}(1) = \mu_{[k]}$$
:::

::: {.fragment}
Setting $t = 1$ removes the powers. Then $\mu = \mu_{[1]}$ and $V(Y) = \mu_{[2]} + \mu - \mu^2$.
:::
:::

---

## 🚗 Worked Example: Motor Claims

::: {style="font-size: 30px"}
An insurer's claims per motor policy per year: $p_0 = 0.70$, $p_1 = 0.20$, $p_2 = 0.08$, $p_3 = 0.02$.

$$P(t) = 0.70 + 0.20t + 0.08t^2 + 0.02t^3$$

::: {.fragment}
$P^{(1)}(t) = 0.20 + 0.16t + 0.06t^2$, so $\mu = P^{(1)}(1) = 0.42$ claims.
:::

::: {.fragment}
$P^{(2)}(t) = 0.16 + 0.12t$, so $\mu_{[2]} = P^{(2)}(1) = 0.28$.
:::

::: {.fragment}
$$V(Y) = 0.28 + 0.42 - 0.42^2 = 0.5236, \qquad \sigma = 0.724$$
:::
:::

---

## 💻 The Same Numbers, Two Ways

```{r}
#| label: pgf-check
#| code-fold: false
y <- 0:3
p <- c(0.70, 0.20, 0.08, 0.02)

mu    <- sum(y * p)              # P^(1)(1)
mu_2f <- sum(y * (y - 1) * p)    # P^(2)(1), the 2nd factorial moment
c(mean = mu, fact2 = mu_2f, var = mu_2f + mu - mu^2)

sims <- sample(y, 1e5, replace = TRUE, prob = p)   # 100,000 simulated policies
round(c(sim_mean = mean(sims), sim_var = var(sims)), 4)
```

::: {style="font-size: 28px"}
The PGF route and the simulation agree to about two decimal places.
:::

---

## 📐 The Geometric PGF

::: {style="font-size: 29px"}
Examples 3.26–3.27. A watch-listed borrower first misses a payment in month $Y$, geometric with $p = 0.25$. Since $p_0 = 0$,
$$P(t) = \sum_{y=1}^{\infty} t^y q^{y-1} p = \frac{pt}{1 - qt}, \qquad t < 1/q$$

::: {.fragment}
Then $P^{(1)}(t) = \dfrac{p}{(1-qt)^2}$ gives $\mu = P^{(1)}(1) = 1/p = 4$ months, and $P^{(2)}(1) = 2q/p^2$ gives
$$V(Y) = \frac{2q}{p^2} + \frac{1}{p} - \frac{1}{p^2} = \frac{q}{p^2} = 12$$
:::

::: {.fragment}
Why keep a second tool? Because $P(t)$ is sometimes the easier one to find. Note $m(t) = P(e^t)$.
:::
:::

---

## 📐 Theorem 3.14: Tchebysheff

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorem 3.14
Let $Y$ be a random variable with mean $\mu$ and finite variance $\sigma^2$. Then, for any constant $k > 0$,
$$P(|Y - \mu| < k\sigma) \ge 1 - \frac{1}{k^2} \qquad \text{or} \qquad P(|Y - \mu| \ge k\sigma) \le \frac{1}{k^2}$$
:::

::: {.fragment}
- It holds for **any** distribution, bell-shaped or not
- It is **conservative**: the true probability usually beats the bound comfortably
- The proof is deferred to §4.10
:::
:::

---

## 🏦 Answering the Risk Desk

::: {style="font-size: 30px"}
$\mu = 12$ and $\sigma = 3$. The interval $(6, 18)$ is $\mu \pm k\sigma$ with $k = 2$:
$$P(6 < Y < 18) = P(|Y - 12| < 2 \times 3) \ge 1 - \frac{1}{2^2} = \frac{3}{4}$$

::: {.fragment}
**At least 75%**, whatever the distribution. That is the sentence the desk can sign.
:::

::: {.fragment}
If the book were steadier, $\sigma = 2$, the same interval is $k = 3$ and the bound rises to $1 - 1/9 = 8/9 \approx 0.889$. As in Example 3.28, **$\sigma$ drives the guarantee**.
:::
:::

---

## 🧮 Running It Backwards: a Provision

::: {style="font-size: 30px"}
The desk wants a level $C$ with $P(|Y - 12| \ge C) \le 0.04$.

::: {.fragment}
Set $1/k^2 = 0.04$, so $k = 5$ and $C = k\sigma = 5 \times 3 = 15$.
:::

::: {.fragment}
Provision for $12 + 15 = 27$ defaults: the chance of **27 or more** is **at most 4%**, with no model assumed.
:::

::: {.fragment}
The price of assuming nothing is a wide band. A model buys a tighter one, but only if the model is right. Exercise 3.167(b) asks exactly this question.
:::
:::

---

## 💻 How Conservative Is the Bound?

```{r}
#| label: cheb-table
#| code-fold: false
# Three models, each measured against its own mean and sd
cover <- function(y, py, k) {
  mu <- sum(y * py); s <- sqrt(sum((y - mu)^2 * py))
  sum(py[abs(y - mu) < k * s])
}
k <- c(1.5, 2, 3)
data.frame(
  k           = k,
  binom_48_25 = sapply(k, \(k) cover(0:48,  dbinom(0:48, 48, 0.25), k)),
  poisson_12  = sapply(k, \(k) cover(0:200, dpois(0:200, 12), k)),
  geom_25     = sapply(k, \(k) cover(1:500, dgeom(0:499, 0.25), k)),
  tchebysheff = 1 - 1 / k^2
) |> round(4)
```

::: {style="font-size: 28px"}
All three beat the bound; at $k = 2$ by 18 to 19 percentage points.
:::

---

## 📈 The Bound Sits Under Every Model

```{r}
#| label: cheb-figure
#| echo: false
#| fig-width: 10
#| fig-height: 4.4
kk <- seq(1, 4, by = 0.01)
curve_df <- rbind(
  data.frame(k = kk, model = "Binomial(48, 0.25)",
             p = sapply(kk, \(k) cover(0:48, dbinom(0:48, 48, 0.25), k))),
  data.frame(k = kk, model = "Poisson(12)",
             p = sapply(kk, \(k) cover(0:200, dpois(0:200, 12), k))),
  data.frame(k = kk, model = "Geometric(0.25)",
             p = sapply(kk, \(k) cover(1:500, dgeom(0:499, 0.25), k))))
bound_df <- data.frame(k = kk, p = 1 - 1 / kk^2)
ggplot(curve_df, aes(k, p, colour = model)) +
  geom_step(linewidth = 1.1) +
  geom_line(data = bound_df, aes(k, p), inherit.aes = FALSE,
            colour = "#8b2635", linewidth = 1.4, linetype = "dashed") +
  annotate("text", x = 3.1, y = 0.72, label = "Tchebysheff:~1 - 1/k^2",
           parse = TRUE, colour = "#8b2635", size = 6.5) +
  scale_colour_manual(values = c("#14130f", "#3d6e8f", "#b3833a")) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.25)) +
  labs(x = "k (half-width in standard deviations)",
       y = expression(P(abs(Y - mu) < k * sigma)), colour = NULL) +
  theme(legend.position = "top", legend.text = element_text(size = 18))
```

---

## 🎯 But the Bound Cannot Be Improved

::: {style="font-size: 30px"}
Exercise 3.169, as a policy-rate decision. The central bank cuts by 1 point, holds, or raises by 1 point:
$$p(-1) = \tfrac{1}{18}, \qquad p(0) = \tfrac{16}{18}, \qquad p(1) = \tfrac{1}{18}$$

::: {.fragment}
$E(Y) = 0$ and $V(Y) = 2/18 = 1/9$, so $\sigma = 1/3$. Take $k = 3$, so $k\sigma = 1$:
$$P(|Y| \ge 1) = \tfrac{2}{18} = \tfrac{1}{9} = \tfrac{1}{k^2}$$
:::

::: {.fragment}
Equality. For any $k > 1$ some distribution attains the bound, so **no better model-free statement exists**.
:::
:::

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 30px"}
A microfinance lender opens on average **40 new arrears cases a month**, with variance **25**. No model is assumed.

**Four minutes, in pairs:**

1. Give a lower bound for $P(30 < Y < 50)$.

2. Find $C$ with $P(|Y - 40| \ge C) \le 1/16$.

3. A colleague says "it's 95%, by the empirical rule". When is she entitled to say that?
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 30px"}
1. $\sigma = \sqrt{25} = 5$ and $(30, 50) = 40 \pm 2 \times 5$, so $k = 2$:
$$P(30 < Y < 50) \ge 1 - \tfrac{1}{4} = 0.75$$

2. $1/k^2 = 1/16$ gives $k = 4$, so $C = 4 \times 5 = 20$: outside $(20, 60)$ with probability at most $1/16$.

::: {.fragment}
3. Only if the distribution is **roughly mound-shaped**. The empirical rule is an approximation for bell shapes; Tchebysheff is a guarantee for every shape. They do not contradict: 95% is above 75%.
:::
:::

---

## 🧭 Chapter 3 in Review: Which Model?

::: {style="font-size: 28px"}
| Model | $Y$ counts | $E(Y)$ | $V(Y)$ |
|---|---|---|---|
| Binomial | successes in $n$ independent trials | $np$ | $npq$ |
| Geometric | trial of the first success | $1/p$ | $q/p^2$ |
| Neg. binomial | trial of the $r$th success | $r/p$ | $rq/p^2$ |
| Hypergeometric | successes in $n$ drawn without replacement | $nr/N$ | $n\frac{r}{N}\frac{N-r}{N}\frac{N-n}{N-1}$ |
| Poisson | rare events per unit of time or space | $\lambda$ | $\lambda$ |
| No model | any count; only $\mu, \sigma$ known | $\mu$ | $\sigma^2$: Tchebysheff |
:::

---

## 🏦 The Risk Desk's Week

::: {style="font-size: 29px"}
- 25 mortgage borrowers, each defaulting independently with probability 0.03: how many default? **Binomial(25, 0.03)**

- A recovery agent phones debtors in turn: on which call does the first promise to pay come? **Geometric**

- An examiner pulls 8 of 60 loan files, 5 of which are misfiled: how many misfiled in the sample? **Hypergeometric**

- Card-fraud alerts in an hour, rare and independent: **Poisson**

- Only five years of monthly means and standard deviations: **Tchebysheff**

- The PGF and MGF sit behind every row: they deliver each $E(Y)$ and $V(Y)$
:::

---

## 📝 Quiz #1: Reading a PGF {.quiz-question}

A count has $P(t) = 0.5 + 0.3t + 0.2t^2$. What is $E(Y)$?

- [$0.7$]{.correct data-explanation="✅ Theorem 3.13: E(Y) = P⁽¹⁾(1) = 0.3 + 2(0.2)(1) = 0.7. The same as summing y p(y) = 0(0.5) + 1(0.3) + 2(0.2)."}
- $1.0$
- $0.3$
- $0.5$

---

## 📝 Quiz #2: The Guarantee {.quiz-question}

Daily FX transfers at a branch have mean 50 and standard deviation 5, shape unknown. What does Tchebysheff guarantee for $P(40 < Y < 60)$?

- [At least $0.75$]{.correct data-explanation="✅ (40, 60) is μ ± 2σ, so the bound is 1 − 1/2² = 0.75. The 95% figure needs a mound shape, which we were not given."}
- At least $0.95$
- Exactly $0.75$
- At most $0.25$

---

## 📝 Quiz #3: Which Model? {.quiz-question}

An auditor draws 10 of a bank's 200 guarantee contracts **without replacement** and counts how many are undocumented. Which model fits?

- [Hypergeometric]{.correct data-explanation="✅ A fixed sample drawn without replacement from a finite pool: the draws are dependent, which rules out the binomial unless the pool is very large."}
- Binomial
- Poisson
- Geometric

---

## 📋 Key Formulas

::: {style="font-size: 30px"}
| | Statement |
|---|---|
| Definition 3.15 | $P(t) = E(t^Y) = \sum_i p_i t^i$ |
| Theorem 3.13 | $P^{(k)}(1) = \mu_{[k]} = E[Y(Y-1)\cdots(Y-k+1)]$ |
| mean and variance | $\mu = P^{(1)}(1)$, $\;V(Y) = P^{(2)}(1) + \mu - \mu^2$ |
| geometric, Poisson | $\dfrac{pt}{1-qt}$, $\;e^{\lambda(t-1)}$ |
| Theorem 3.14 | $P(\lvert Y-\mu \rvert < k\sigma) \ge 1 - 1/k^2$ |
| equivalently | $P(\lvert Y-\mu \rvert \ge k\sigma) \le 1/k^2$ |
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- A PGF stores $P(Y = i)$ as the coefficient of $t^i$; its derivatives at $t = 1$ are factorial moments

- $V(Y) = \mu_{[2]} + \mu - \mu^2$ turns them into a variance

- Tchebysheff needs only $\mu$ and $\sigma$, and gives at least $1 - 1/k^2$ inside $\mu \pm k\sigma$

- The bound is conservative for familiar models but cannot be improved in general

- Chapter 3 is a map: name the counting mechanism, pick the model, read off $E(Y)$ and $V(Y)$
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition**

- §3.10 (optional): Exercises 3.164 – 3.166, the binomial and Poisson PGFs

- §3.11: Exercises 3.167 – 3.174; start with 3.167, 3.169 and 3.173

- Redo the risk-desk provision with $\sigma = 4$: what does $C$ become?

**Week 8, Problem Set 2** is open now and closes **Sunday 8 November at 23:59** on WeBWorK, covering §§3.10–3.11.

**Next class:** 4 November, we leave counts behind for continuous random variables: distribution and density functions (Wackerly §§4.1–4.2).
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**

📧 sorujov@ada.edu.az\
🏢 Building D, Room D325\
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- Tchebysheff says nothing for $k \le 1$. Why is $1 - 1/k^2$ useless there?

- The Poisson PGF is $e^{\lambda(t-1)}$. What is its second factorial moment, and hence its variance?

- A model gives a tighter band than Tchebysheff. When would a risk manager still report the Tchebysheff band?
:::
