---
title: "Mathematical Statistics"
subtitle: "Expected Values of Functions of Random Variables and Covariance"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - Expected Values and Covariance"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
library(ggplot2)
theme_set(theme_minimal(base_size = 18))
```

## 🎯 Learning Objectives

::: {style="font-size: 32px"}
By the end of this lecture, you will be able to:

- **Compute** $E[g(Y_1, Y_2)]$ from a joint probability table or a joint density (Definition 5.9)

- **Use** Theorems 5.6–5.8 to take expectations of sums term by term, with or without independence

- **Apply** Theorem 5.9 to factor $E[g(Y_1)h(Y_2)]$ when $Y_1$ and $Y_2$ are independent

- **Compute** $\text{Cov}(Y_1, Y_2)$ and $\rho$ by Theorem 5.10, and read their sign

- **Explain** why independence implies zero covariance, but zero covariance does not imply independence
:::

---

## 🗺️ Where We Are

::: {style="font-size: 30px"}
**Wackerly §§5.5–5.7**

Wednesday (§5.4): independence means the joint distribution is the product of the marginals, everywhere. It ended on co-movement: at correlation 0.6, $P(\text{both fall}) = 0.352$, not $0.5 \times 0.5 = 0.25$.

::: {.fragment}
We used that word, **correlation**, before defining it. Today we define it, as one expected value of a function of two variables: $E[(Y_1 - \mu_1)(Y_2 - \mu_2)]$.
:::

::: {.fragment}
First, the general tool: $E[g(Y_1, Y_2)]$ for total revenue, a product of shares, a spread.
:::
:::

---

## ❓ Motivating Question

::: {.callout-important}
## The appliance retailer's peak hour
A Baku home-appliance store sells air conditioners at **1,200 AZN** and refrigerators at **900 AZN**. In the Saturday peak hour it sells $Y_1$ air conditioners and $Y_2$ refrigerators.

What revenue should it **expect** in that hour, and do the two product lines sell **together**?
:::

::: {.fragment style="font-size: 30px"}
Revenue $R = 1200Y_1 + 900Y_2$ is a function of **two** random variables. Chapter 4 told us how to find $E[g(Y)]$ for one.
:::

---

## 📐 Definition 5.9

::: {style="font-size: 30px"}
::: {.callout-note}
## Definition 5.9
Let $g(Y_1, \ldots, Y_k)$ be a function of discrete random variables with probability function $p(y_1, \ldots, y_k)$. Then
$$E[g(Y_1, \ldots, Y_k)] = \sum_{\text{all } y_k} \cdots \sum_{\text{all } y_1} g(y_1, \ldots, y_k)\, p(y_1, \ldots, y_k).$$
For continuous variables with joint density $f$, the sums become integrals over $(-\infty, \infty)$.
:::

::: {.fragment}
In words: **weight each value of $g$ by the probability of the cell that produces it**, and add. Taking $g(Y_1, Y_2) = Y_1$ gives back Definition 4.5 through the marginal density.
:::
:::

---

## 🏬 Worked Example: The Peak Hour

::: {style="font-size: 28px"}
:::: {.columns}
::: {.column width="50%"}
| $y_1 \backslash y_2$ | 0 | 1 | 2 | $p_1$ |
|---|---|---|---|---|
| **0** | .30 | .15 | .05 | .50 |
| **1** | .10 | .15 | .10 | .35 |
| **2** | .02 | .05 | .08 | .15 |
| $p_2$ | .42 | .35 | .23 | 1 |
:::

::: {.column width="50%"}
Each cell carries a revenue $g(y_1, y_2) = 1200y_1 + 900y_2$: from 0 AZN (nothing sold) to 4200 AZN (two of each).
:::
::::

::: {.fragment}
Definition 5.9, cell by cell:
$$E(R) = 0(.30) + 900(.15) + 1800(.05) + \cdots + 4200(.08) = \mathbf{1509} \text{ AZN}$$
:::
:::

---

## 📐 Theorems 5.6–5.8: Special Theorems

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorems 5.6, 5.7, 5.8
For a constant $c$ and functions $g, g_1, \ldots, g_k$ of $Y_1$ and $Y_2$:
$$E(c) = c, \qquad E[c\,g(Y_1, Y_2)] = c\,E[g(Y_1, Y_2)],$$
$$E[g_1(Y_1, Y_2) + \cdots + g_k(Y_1, Y_2)] = E[g_1(Y_1, Y_2)] + \cdots + E[g_k(Y_1, Y_2)].$$
:::

::: {.fragment}
The proofs are the univariate ones: a sum or an integral is linear. **Nothing here assumes independence.** The expectation of a sum is the sum of the expectations, however the terms are related.
:::
:::

---

## ⚡ Expected Revenue the Short Way

::: {style="font-size: 28px"}
With $g_1 = 1200Y_1$ and $g_2 = 900Y_2$, Theorems 5.7 and 5.8 give
$$E(R) = 1200E(Y_1) + 900E(Y_2) = 1200(0.65) + 900(0.81) = 780 + 729 = 1509 \text{ AZN}.$$
Only the **marginals** were needed. The same 1509 as nine cells of arithmetic:
:::

```{r}
#| label: peak-hour
p <- matrix(c(.30, .15, .05,
              .10, .15, .10,
              .02, .05, .08), nrow = 3, byrow = TRUE)
y <- 0:2
R <- outer(1200 * y, 900 * y, "+")          # revenue in each cell
c(by_definition_5.9 = sum(R * p),
  by_linearity      = 1200 * sum(y * rowSums(p)) + 900 * sum(y * colSums(p)))
```

---

## ⚡ Worked Example: A Power Plant

::: {style="font-size: 28px"}
A gas-fired plant sells a share $Y_1$ of its monthly output on the spot market (the rest under contract), and a share $Y_2$ of those spot sales clears at peak-hour prices. Model:
$$f(y_1, y_2) = 2(1 - y_1), \qquad 0 \le y_1 \le 1,\; 0 \le y_2 \le 1.$$

::: {.fragment}
$Y_1 Y_2$ is the share of **total** output sold at the peak spot price. By Definition 5.9,
$$E(Y_1Y_2) = \int_0^1\!\!\int_0^1 2y_1y_2(1 - y_1)\,dy_2\,dy_1 = \int_0^1 (y_1 - y_1^2)\,dy_1 = \frac{1}{2} - \frac{1}{3} = \frac{1}{6}.$$
:::

::: {.fragment}
On 180 GWh of monthly output, the plant should expect **30 GWh** sold at the peak spot price.
:::
:::

---

## 📐 Theorem 5.9: Products Split

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorem 5.9
If $Y_1$ and $Y_2$ are independent and $g(Y_1)$, $h(Y_2)$ are functions of $Y_1$ only and $Y_2$ only, then
$$E[g(Y_1)h(Y_2)] = E[g(Y_1)]\,E[h(Y_2)],$$
provided the expectations exist.
:::

::: {.fragment}
The plant's density is $2(1 - y_1) \times 1$ on a rectangle, so by Wednesday's Theorem 5.5 $Y_1$ and $Y_2$ are independent, with $E(Y_1) = 1/3$ and $E(Y_2) = 1/2$:
$$E(Y_1Y_2) = \tfrac{1}{3} \cdot \tfrac{1}{2} = \tfrac{1}{6}, \text{ as before, with no double integral.}$$
:::
:::

---

## 📈 The Sign of the Deviations

```{r}
#| label: deviation-figure
#| echo: false
#| fig-width: 10
#| fig-height: 4.3
n  <- 120
z1 <- rnorm(n); z2 <- rnorm(n)
bankA <- 1.0 + 2.12 * z1
bankB <- 1.2 + 2.96 * (0.516 * z1 + sqrt(1 - 0.516^2) * z2)
dev  <- data.frame(a = bankA, b = bankB,
                   sign = ifelse((bankA - 1.0) * (bankB - 1.2) > 0,
                                 "product of deviations > 0", "product of deviations < 0"))
share_pos <- mean(dev$sign == "product of deviations > 0")
ggplot(dev, aes(a, b, colour = sign)) +
  geom_vline(xintercept = 1.0, linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = 1.2, linetype = "dashed", colour = "grey40") +
  geom_point(size = 2.6, alpha = 0.85) +
  annotate("text", x = 1.0, y = max(bankB) + 0.8, label = "mu[1]", parse = TRUE, size = 7) +
  annotate("text", x = min(bankA) - 0.3, y = 1.2 + 0.9, label = "mu[2]", parse = TRUE, size = 7) +
  scale_colour_manual(values = c("#8b2635", "#3d6e8f")) +
  labs(x = "Bank A monthly return (%)", y = "Bank B monthly return (%)", colour = NULL) +
  theme(legend.position = "top", legend.text = element_text(size = 18))
```

::: {style="font-size: 28px"}
`r round(100 * share_pos)`% of 120 simulated months land where $(y_1 - \mu_1)(y_2 - \mu_2) > 0$, so the **average** product is positive.
:::

---

## 📐 Definition 5.10 and Theorem 5.10

::: {style="font-size: 28px"}
::: {.callout-note}
## Definition 5.10
If $Y_1$ and $Y_2$ have means $\mu_1$ and $\mu_2$, the **covariance** of $Y_1$ and $Y_2$ is
$$\text{Cov}(Y_1, Y_2) = E[(Y_1 - \mu_1)(Y_2 - \mu_2)].$$
Its scale-free version is the **correlation coefficient** $\rho = \dfrac{\text{Cov}(Y_1, Y_2)}{\sigma_1 \sigma_2}$, with $-1 \le \rho \le 1$.
:::

::: {.callout-important}
## Theorem 5.10
$$\text{Cov}(Y_1, Y_2) = E(Y_1Y_2) - E(Y_1)E(Y_2).$$
:::

Proof: expand the product, then apply Theorems 5.7 and 5.8 term by term.
:::

---

## 🏦 Worked Example: Two Bank Shares

::: {style="font-size: 28px"}
Monthly returns (%) of two bank shares on the Baku Stock Exchange, $Y_1$ for bank A and $Y_2$ for bank B:

| $y_1 \backslash y_2$ | $-3$ | $1$ | $5$ | $p_1$ |
|---|---|---|---|---|
| $-2$ | .15 | .08 | .02 | .25 |
| $1$ | .08 | .30 | .12 | .50 |
| $4$ | .02 | .07 | .16 | .25 |
| $p_2$ | .25 | .45 | .30 | 1 |

::: {.fragment}
$E(Y_1) = 1.0$, $E(Y_2) = 1.2$, and summing $y_1y_2\,p(y_1, y_2)$ over the nine cells gives $E(Y_1Y_2) = 4.44$.

$$\text{Cov}(Y_1, Y_2) = 4.44 - (1.0)(1.2) = 3.24, \qquad \rho = \frac{3.24}{\sqrt{4.5}\,\sqrt{8.76}} = 0.516.$$
:::
:::

---

## 💻 Covariance and Correlation in R

```{r}
#| label: bank-cov
#| code-fold: false
y1 <- c(-2, 1, 4); y2 <- c(-3, 1, 5)
p  <- matrix(c(.15, .08, .02,
               .08, .30, .12,
               .02, .07, .16), nrow = 3, byrow = TRUE)
mu1 <- sum(y1 * rowSums(p)); mu2 <- sum(y2 * colSums(p))
cov_def  <- sum(outer(y1 - mu1, y2 - mu2) * p)       # Definition 5.10
cov_5.10 <- sum(outer(y1, y2) * p) - mu1 * mu2        # Theorem 5.10
s1 <- sqrt(sum(y1^2 * rowSums(p)) - mu1^2); s2 <- sqrt(sum(y2^2 * colSums(p)) - mu2^2)
round(c(cov_def = cov_def, cov_5.10 = cov_5.10, rho = cov_5.10 / (s1 * s2)), 3)
```

::: {style="font-size: 28px"}
The covariance of 3.24 is in "percent squared", hard to judge. $\rho = 0.52$ says: **moderate** positive linear dependence.
:::

---

## 📐 Theorem 5.11

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorem 5.11
If $Y_1$ and $Y_2$ are independent, then $\text{Cov}(Y_1, Y_2) = 0$.
:::

Proof: Theorem 5.9 gives $E(Y_1Y_2) = \mu_1\mu_2$, and Theorem 5.10 finishes it. For the power plant, $\text{Cov} = \tfrac{1}{6} - \tfrac{1}{3}\cdot\tfrac{1}{2} = 0$.

::: {.fragment}
Read it backwards as a test: the bank shares have $\text{Cov} = 3.24 \ne 0$, so they **cannot** be independent.

**The converse is false.** Zero covariance does not make two variables independent.
:::
:::

---

## 🛢️ Zero Covariance, Still Dependent

::: {style="font-size: 28px"}
An energy desk holds $Y_1$ Brent and $Y_2$ natural-gas contracts each day ($-1$ short, $0$ flat, $+1$ long). The desk mandate forbids being **flat in both** (Wackerly Table 5.3):

| $y_1 \backslash y_2$ | $-1$ | $0$ | $+1$ |
|---|---|---|---|
| $-1$ | 1/16 | 3/16 | 1/16 |
| $0$ | 3/16 | **0** | 3/16 |
| $+1$ | 1/16 | 3/16 | 1/16 |

::: {.fragment}
Dependent: $p(0, 0) = 0 \ne p_1(0)p_2(0) = (6/16)^2$. Yet $E(Y_1) = E(Y_2) = 0$, and the four corners give $E(Y_1Y_2) = \tfrac{1}{16} - \tfrac{1}{16} - \tfrac{1}{16} + \tfrac{1}{16} = 0$, so **Cov = 0**.
:::
:::

---

## 📉 Dependent but Uncorrelated

```{r}
#| label: straddle-figure
#| echo: false
#| fig-width: 10
#| fig-height: 4.3
brent  <- rnorm(250, mean = 0, sd = 2)          # daily Brent move, %
payoff <- abs(brent)                            # at-the-money straddle payoff
r_str  <- cor(brent, payoff)
ggplot(data.frame(x = brent, y = payoff), aes(x, y)) +
  geom_point(colour = "#3d6e8f", size = 2.4, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, colour = "#8b2635", linewidth = 1.3) +
  labs(x = "Daily Brent move (%)", y = "Straddle payoff (% of spot)")
```

::: {style="font-size: 28px"}
The payoff is **completely determined** by the move, yet over 250 simulated days the sample correlation is $r = `r sprintf("%.2f", r_str)`$ and the fitted line (red) is almost flat. Covariance sees only **linear** dependence; a V shape cancels itself out.
:::

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 30px"}
:::: {.columns}
::: {.column width="42%"}
| $y_1 \backslash y_2$ | 0 | 1 | 2 |
|---|---|---|---|
| **0** | .30 | .15 | .05 |
| **1** | .10 | .15 | .10 |
| **2** | .02 | .05 | .08 |
:::
::: {.column width="58%"}
The peak hour again: $E(Y_1) = 0.65$ air conditioners, $E(Y_2) = 0.81$ refrigerators.
:::
::::

**Four minutes, in pairs:**

1. Find $E(Y_1Y_2)$.

2. Find $\text{Cov}(Y_1, Y_2)$. What sign did you expect, and why?

3. Are $Y_1$ and $Y_2$ independent? Give **two** ways to tell.
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 28px"}
1. Only cells with $y_1, y_2 \ge 1$ contribute:
$$E(Y_1Y_2) = 1(.15) + 2(.10) + 2(.05) + 4(.08) = 0.77$$

2. Theorem 5.10: $\text{Cov}(Y_1, Y_2) = 0.77 - (0.65)(0.81) = 0.2435 > 0$. A customer refitting a kitchen or a flat buys both.

::: {.fragment}
3. Not independent. By Theorem 5.11, a nonzero covariance rules independence out. Directly: $p(0, 0) = 0.30 \ne p_1(0)p_2(0) = 0.50 \times 0.42 = 0.21$.

Note that $E(R) = 1509$ AZN did **not** need any of this: linearity ignores dependence.
:::
:::

---

## 📝 Quiz #1: Linearity {.quiz-question}

A telecom operator's daily new mobile contracts $Y_1$ and fibre contracts $Y_2$ are **dependent**, with $E(Y_1) = 2$ and $E(Y_2) = 3$ (hundreds). What is $E(4Y_1 - Y_2 + 10)$?

- [$15$]{.correct data-explanation="✅ Theorems 5.6–5.8: 4(2) − 3 + 10 = 15. Linearity holds whatever the dependence between Y1 and Y2."}
- $21$
- $9$
- Cannot be found without the joint distribution

---

## 📝 Quiz #2: From Moments to ρ {.quiz-question}

$E(Y_1Y_2) = 5$, $E(Y_1) = 1$, $E(Y_2) = 2$, $V(Y_1) = 4$, $V(Y_2) = 9$. What is $\rho$?

- [$0.5$]{.correct data-explanation="✅ Theorem 5.10: Cov = 5 − (1)(2) = 3. Then ρ = 3 / (2 × 3) = 0.5, dividing by the standard deviations, not the variances."}
- $3$
- $0.083$
- $0.833$

---

## 📝 Quiz #3: Reading a Zero {.quiz-question}

Two exchange-rate changes are found to have $\text{Cov}(Y_1, Y_2) = 0$. Which conclusion is justified?

- They are independent
- [There is no linear dependence; they may still be dependent]{.correct data-explanation="✅ Theorem 5.11 runs one way only. Example 5.24 (the energy desk) and the straddle both have zero covariance and strong dependence."}
- $E(Y_1Y_2) = 0$
- Their correlation could still be positive

---

## 📋 Key Formulas

::: {style="font-size: 28px"}
| | Statement |
|---|---|
| Definition 5.9 | $E[g(Y_1, Y_2)] = \sum\sum g(y_1, y_2)\,p(y_1, y_2)$ or $\iint g\,f\,dy_1\,dy_2$ |
| Theorems 5.6–5.8 | $E(c) = c$, $E(cg) = cE(g)$, $E(g_1 + g_2) = E(g_1) + E(g_2)$ |
| Theorem 5.9 | independent: $E[g(Y_1)h(Y_2)] = E[g(Y_1)]E[h(Y_2)]$ |
| Definition 5.10 | $\text{Cov}(Y_1, Y_2) = E[(Y_1 - \mu_1)(Y_2 - \mu_2)]$ |
| Theorem 5.10 | $\text{Cov}(Y_1, Y_2) = E(Y_1Y_2) - \mu_1\mu_2$ |
| correlation | $\rho = \text{Cov}(Y_1, Y_2)/(\sigma_1\sigma_2)$, $\;-1 \le \rho \le 1$ |
| Theorem 5.11 | independent $\Rightarrow$ Cov $= 0$ (not conversely) |
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- Definition 5.9: weight each value of $g$ by its joint probability or density, and add

- The expectation of a sum is the sum of the expectations, with **no** independence needed

- Under independence, expectations of products split into products of expectations (Theorem 5.9)

- Covariance is the average product of deviations; its sign says whether two variables move together

- $\rho$ removes the units, so it can be read as weak, moderate or strong

- Independence forces zero covariance; zero covariance allows dependence that is not linear
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition**

- §§5.5–5.6: Exercises 5.72 – 5.88; start with 5.72, 5.74, 5.77, 5.80 and 5.84

- §5.7: Exercises 5.89 – 5.97; start with 5.89, 5.91, 5.93, 5.95 and 5.97

- Change the bank-share table so that $\rho < 0$ while keeping both marginals. Which cells must move?

**Week 13, Problem Set 2** is open now and closes **Sunday 13 December at 23:59** on WeBWorK, covering §§5.5–5.7.

**Next class:** 9 December, the expected value and variance of linear functions of random variables, and portfolio variance (Wackerly §5.8).
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**

📧 sorujov@ada.edu.az\
🏢 Building D, Room D325\
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- The retailer's revenue needed only the marginals. Which quantity about $R$ will need the covariance as well?

- If you double every return of bank A, what happens to the covariance, and what happens to $\rho$?

- Can you build a joint table where $Y_2 = Y_1^2$ exactly and the covariance is zero? What must the distribution of $Y_1$ look like?
:::
