---
title: "Mathematical Statistics"
subtitle: "Independent Random Variables"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - Independent Random Variables"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
library(ggplot2)
theme_set(theme_minimal(base_size = 18))
```

## 🎯 Learning Objectives

::: {style="font-size: 32px"}
By the end of this lecture, you will be able to:

- **State** Definition 5.8: $Y_1, Y_2$ are independent when $F(y_1, y_2) = F_1(y_1)F_2(y_2)$

- **Test** a joint table cell by cell with Theorem 5.4, and **stop** at the first cell that fails

- **Apply** the factorisation criterion (Theorem 5.5) to a joint density without finding its marginals

- **Recognise** a support that is not a rectangle as a proof of dependence

- **Explain** why co-moving asset returns cannot be priced as independent
:::

---

## 🗺️ Where We Are

::: {style="font-size: 32px"}
**Wackerly §5.4**

Saturday (§5.3): a **marginal** removes a variable, $p_1(y_1) = \sum_{y_2} p(y_1, y_2)$; a **conditional** keeps one slice of the joint and divides by its total, $p(y_1 \mid y_2) = p(y_1, y_2)/p_2(y_2)$.

::: {.fragment}
We ended on the question for today: what does it mean when conditioning on one variable **leaves the other's distribution unchanged**? Then learning $Y_2$ tells you nothing about $Y_1$, and the joint is just the two marginals multiplied.
:::
:::

---

## ❓ The Question

::: {.callout-important}
## Two branches, one cash van
A Baku bank refills the ATMs of two branches, one in Nizami and one in Khatai, from a single cash centre. Does a heavy day of withdrawals in Nizami make a heavy day in Khatai more likely?
:::

::: {style="font-size: 30px"}
If **no**, the treasury can plan each branch from its own history and multiply. If **yes**, it has to plan from the joint distribution, and both branches running dry on the same day is more likely than the product suggests.
:::

---

## 📐 Definition 5.8

::: {style="font-size: 30px"}
::: {.callout-note}
## Definition 5.8 (Wackerly §5.4)
Let $Y_1$ have distribution function $F_1(y_1)$, $Y_2$ have distribution function $F_2(y_2)$, and $Y_1$ and $Y_2$ have joint distribution function $F(y_1, y_2)$. Then $Y_1$ and $Y_2$ are **independent** if and only if
$$F(y_1, y_2) = F_1(y_1)F_2(y_2)$$
for every pair of real numbers $(y_1, y_2)$. Otherwise they are **dependent**.
:::

It is the Chapter 2 rule $P(A \cap B) = P(A)P(B)$, applied to every pair of events $\{Y_1 \le y_1\}$ and $\{Y_2 \le y_2\}$ at once.
:::

---

## 📐 Theorem 5.4: The Working Test

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorem 5.4
**Discrete:** $Y_1$ and $Y_2$ are independent if and only if
$$p(y_1, y_2) = p_1(y_1)\,p_2(y_2) \quad \text{for all pairs } (y_1, y_2).$$
**Continuous:** $Y_1$ and $Y_2$ are independent if and only if
$$f(y_1, y_2) = f_1(y_1)\,f_2(y_2) \quad \text{for all pairs } (y_1, y_2).$$
:::

::: {.fragment}
"For **all** pairs" cuts both ways: to prove independence you check every cell; to prove dependence **one** failing cell is enough.
:::
:::

---

## 🏦 Two Branches: Large Withdrawals

::: {style="font-size: 28px"}
$Y_1$, $Y_2$ = number of withdrawals above 20,000 AZN before noon at Nizami and at Khatai.

| | $y_2 = 0$ | $y_2 = 1$ | $y_2 = 2$ | $p_1(y_1)$ |
|---|---|---|---|---|
| $y_1 = 0$ | 0.30 | 0.15 | 0.05 | 0.50 |
| $y_1 = 1$ | 0.18 | 0.09 | 0.03 | 0.30 |
| $y_1 = 2$ | 0.12 | 0.06 | 0.02 | 0.20 |
| $p_2(y_2)$ | 0.60 | 0.30 | 0.10 | 1.00 |

::: {.fragment}
Cell $(1, 1)$: $p_1(1)\,p_2(1) = 0.30 \times 0.30 = 0.09 = p(1, 1)$. The same holds in all nine cells, so $Y_1$ and $Y_2$ are **independent**.
:::
:::

---

## 📡 Tower Outages After a Storm

::: {style="font-size: 28px"}
Daily cell-tower outages: $Y_1$ in Baku, $Y_2$ in Sumgait. One storm hits both.

| | $y_2 = 0$ | $y_2 = 1$ | $y_2 = 2$ | $p_1(y_1)$ |
|---|---|---|---|---|
| $y_1 = 0$ | 0.40 | 0.08 | 0.02 | 0.50 |
| $y_1 = 1$ | 0.08 | 0.14 | 0.08 | 0.30 |
| $y_1 = 2$ | 0.02 | 0.08 | 0.10 | 0.20 |
| $p_2(y_2)$ | 0.50 | 0.30 | 0.20 | 1.00 |

::: {.fragment}
Cell $(0, 0)$: $p(0, 0) = 0.40$ but $p_1(0)\,p_2(0) = 0.25$. **Dependent.** Also $P(Y_2 = 0 \mid Y_1 = 0) = 0.80$ but $P(Y_2 = 0 \mid Y_1 = 2) = 0.10$.
:::
:::

---

## 💻 Checking Every Cell in R

```{r}
#| label: cell-check
branches <- matrix(c(.30, .15, .05,  .18, .09, .03,  .12, .06, .02),
                   nrow = 3, byrow = TRUE)
towers   <- matrix(c(.40, .08, .02,  .08, .14, .08,  .02, .08, .10),
                   nrow = 3, byrow = TRUE)
# ratio p(y1, y2) / (p1(y1) p2(y2)): every entry is 1 exactly when independent
ratio <- function(p) p / outer(rowSums(p), colSums(p))
round(ratio(branches), 3)
round(ratio(towers), 3)
```

::: {style="font-size: 28px"}
Ratios above 1 on the diagonal: outages **cluster** on the same days.
:::

---

## 📈 Continuous: A Density That Splits

::: {style="font-size: 28px"}
$Y_1$, $Y_2$ = fraction of each branch's ATM cassette dispensed by closing time, with
$$f(y_1, y_2) = 6y_1y_2^2, \qquad 0 \le y_1 \le 1,\ 0 \le y_2 \le 1.$$

::: {.fragment}
Marginals (limits of integration are **constants**):
$$f_1(y_1) = \int_0^1 6y_1y_2^2\,dy_2 = 2y_1, \qquad f_2(y_2) = \int_0^1 6y_1y_2^2\,dy_1 = 3y_2^2$$
:::

::: {.fragment}
$f_1(y_1)f_2(y_2) = 6y_1y_2^2 = f(y_1, y_2)$ everywhere: **independent** (compare Example 5.11). So
$$P(Y_1 > 0.5,\ Y_2 > 0.5) = (1 - 0.5^2)(1 - 0.5^3) = 0.75 \times 0.875 = 0.656$$
:::
:::

---

## 🔺 A Triangle Is Never Independent

::: {style="font-size: 28px"}
$Y_1$ = fraction of the vault's capacity stocked at opening, $Y_2$ = fraction withdrawn during the day. You cannot withdraw more than was stocked:
$$f(y_1, y_2) = 2, \qquad 0 \le y_2 \le y_1 \le 1.$$

::: {.fragment}
$f_1(y_1) = \int_0^{y_1} 2\,dy_2 = 2y_1$ and $f_2(y_2) = \int_{y_2}^{1} 2\,dy_1 = 2(1 - y_2)$ (compare Example 5.12).
:::

::: {.fragment}
At $(0.25, 0.75)$: $f = 0$, but $f_1(0.25)f_2(0.75) = 0.5 \times 0.5 = 0.25$. **Dependent.** The limits of integration depended on the other variable, and that was the warning.
:::
:::

---

## 📐 Theorem 5.5: Factorisation

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorem 5.5
Let $Y_1$ and $Y_2$ have a joint density $f(y_1, y_2)$ that is positive if and only if $a \le y_1 \le b$ and $c \le y_2 \le d$, for constants $a, b, c, d$; and $f(y_1, y_2) = 0$ otherwise. Then $Y_1$ and $Y_2$ are independent if and only if
$$f(y_1, y_2) = g(y_1)\,h(y_2),$$
where $g$ is a nonnegative function of $y_1$ alone and $h$ of $y_2$ alone.
:::

No marginals needed, and $g$, $h$ need not be densities. Two conditions: a **rectangle** support, and a **product** formula.
:::

---

## 🔍 Factor or Not?

::: {style="font-size: 28px"}
| Joint density | Support | Verdict |
|---|---|---|
| $12\,y_1\,y_2(1 - y_2)$ | $[0,1] \times [0,1]$ | $g = y_1$, $h = 12y_2(1-y_2)$: **independent** |
| $\tfrac32(y_1^2 + y_2^2)$ | $[0,1] \times [0,1]$ | a sum, no split: **dependent** |
| $8\,y_1y_2$ | $0 \le y_2 \le y_1 \le 1$ | a product, but a triangle: **dependent** |

::: {.fragment}
Row 2: $f_1(y_1) = \tfrac32 y_1^2 + \tfrac12$, so at $(0,0)$ we have $f = 0$ but $f_1 f_2 = 0.25$.

Row 3 is the trap. The formula factors, the support does not; at $(0.25, 0.75)$, $f = 0$ but $f_1 f_2 = 0.0625 \times 1.3125 = 0.082$.
:::
:::

---

## 💵 Worked Example: Cash Demand

::: {style="font-size: 28px"}
Daily cash withdrawn (units of 100,000 AZN): $Y_1$ at Nizami, $Y_2$ at Khatai, with
$$f(y_1, y_2) = \tfrac12\, y_1 e^{-y_1} e^{-y_2/2}, \qquad y_1 > 0,\ y_2 > 0.$$

::: {.fragment}
**Step 1.** Support $(0, \infty) \times (0, \infty)$ is a rectangle; $g(y_1) = y_1e^{-y_1}$, $h(y_2) = \tfrac12 e^{-y_2/2}$. Independent by Theorem 5.5: $Y_1$ is gamma$(2, 1)$, $Y_2$ exponential with mean 2.
:::

::: {.fragment}
**Step 2.** Both branches above 200,000 AZN: one double integral becomes two single ones,
$$P(Y_1 > 2,\ Y_2 > 2) = 3e^{-2} \times e^{-1} = 0.406 \times 0.368 = 3e^{-3} = 0.149.$$
:::
:::

---

## 💻 Cash Demand in R

```{r}
#| label: cash-sim
n  <- 100000
y1 <- rgamma(n, shape = 2, rate = 1)   # Nizami, mean 2 (200,000 AZN)
y2 <- rexp(n, rate = 1/2)              # Khatai, mean 2
c(P_Y1_gt2  = pgamma(2, 2, 1, lower.tail = FALSE),
  P_Y2_gt2  = pexp(2, 1/2, lower.tail = FALSE),
  product   = 3 * exp(-3),
  simulated = mean(y1 > 2 & y2 > 2)) |> round(4)
```

::: {style="font-size: 28px"}
About one day in seven, both branches need a large top-up: the cash centre plans for that directly from the two marginals.
:::

---

## 🛢️ Co-Moving Returns

```{r}
#| label: comoving
#| echo: false
#| fig-width: 11
#| fig-height: 4.2
set.seed(2026)
n_days <- 750
z1 <- rnorm(n_days); z2 <- rnorm(n_days)
make <- function(rho, lab) data.frame(
  brent = 2.0 * z1,                                  # Brent daily return, %
  bond  = 0.8 * (rho * z1 + sqrt(1 - rho^2) * z2),   # energy-issuer bond, %
  panel = lab)
df <- rbind(make(0, "independent"), make(0.6, "co-moving"))
df$panel <- factor(df$panel, levels = c("independent", "co-moving"))
ggplot(df, aes(brent, bond)) +
  geom_point(alpha = 0.35, size = 1.6, colour = "#2f4b7c") +
  geom_vline(xintercept = -2, linetype = "dashed", colour = "#8b2635") +
  facet_wrap(~ panel) +
  labs(x = "Brent daily return (%)", y = "Bond daily return (%)") +
  theme(strip.text = element_text(size = 20, face = "bold"))
```

::: {style="font-size: 28px"}
Same marginals in both panels. When Brent falls more than 2% (left of the dashes), the bond falls with probability 0.50 on the left but **0.86** on the right.
:::

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 28px"}
$Y_1$ = tanker loadings delayed at an oil terminal today (0 or 1); $Y_2$ = pipeline pumping-station alarms (0, 1, 2).

| | $y_2 = 0$ | $y_2 = 1$ | $y_2 = 2$ |
|---|---|---|---|
| $y_1 = 0$ | 0.42 | 0.21 | 0.07 |
| $y_1 = 1$ | 0.18 | ? | ? |

1. If $Y_1$ and $Y_2$ are independent, fill in the two missing cells.
2. A colleague enters $0.12$ and $0$ instead. Independent?
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 28px"}
1. Row 0 gives $p_1(0) = 0.70$, so $p_1(1) = 0.30$. Independence in row 0 fixes $p_2$:
$$p_2(0) = \tfrac{0.42}{0.70} = 0.6,\quad p_2(1) = \tfrac{0.21}{0.70} = 0.3,\quad p_2(2) = 0.1$$
Check $p(1,0) = 0.3 \times 0.6 = 0.18$ ✓. Then $p(1,1) = 0.09$ and $p(1,2) = 0.03$.

::: {.fragment}
2. The table still sums to 1, but $p(1, 2) = 0$ while $p_1(1)\,p_2(2) = 0.30 \times 0.07 = 0.021$. **Dependent.** A zero cell under two positive marginals always is (compare Example 5.10).
:::
:::

---

## 📝 Quiz #1: Factorisation {.quiz-question}

$f(y_1, y_2) = \tfrac19\, y_1^2 e^{-y_2}$ for $0 \le y_1 \le 3$, $y_2 > 0$, and 0 elsewhere. Are $Y_1$, $Y_2$ independent?

- [Yes: the support is a rectangle and $f = g(y_1)h(y_2)$]{.correct data-explanation="✅ Theorem 5.5: the support [0,3] × (0,∞) is a rectangle, and f splits as g(y1) = y1²/9 times h(y2) = e^(−y2)."}
- No: $y_1$ has a finite range but $y_2$ does not
- Cannot say without computing both marginals
- No: $g(y_1) = y_1^2$ is not a density

---

## 📝 Quiz #2: One Cell {.quiz-question}

In a joint table of two exporters' monthly shipment counts, $p(2, 0) = 0$, while $p_1(2) = 0.2$ and $p_2(0) = 0.4$. What follows?

- [$Y_1$ and $Y_2$ are dependent]{.correct data-explanation="✅ Theorem 5.4 needs p(y1, y2) = p1(y1)p2(y2) for every pair. Here 0 ≠ 0.2 × 0.4 = 0.08, and one failing cell settles it."}
- Nothing, until the other cells are checked
- They are independent, since $0 \le 0.08$
- The table is invalid

---

## 📝 Quiz #3: Using Definition 5.8 {.quiz-question}

Daily losses $Y_1$, $Y_2$ (in million AZN) on two unrelated trading books are independent, with $P(Y_1 \le 1) = 0.7$ and $P(Y_2 \le 1) = 0.4$. What is $P(Y_1 \le 1,\ Y_2 \le 1)$?

- [$0.28$]{.correct data-explanation="✅ Definition 5.8: F(1, 1) = F1(1)F2(1) = 0.7 × 0.4 = 0.28."}
- $0.82$
- $1.10$
- $0.40$

---

## 📋 Key Formulas

::: {style="font-size: 28px"}
| | Independent if and only if |
|---|---|
| Definition 5.8 | $F(y_1, y_2) = F_1(y_1)F_2(y_2)$ for all $(y_1, y_2)$ |
| Theorem 5.4, discrete | $p(y_1, y_2) = p_1(y_1)p_2(y_2)$ for all pairs |
| Theorem 5.4, continuous | $f(y_1, y_2) = f_1(y_1)f_2(y_2)$ for all pairs |
| Theorem 5.5 (rectangle support) | $f(y_1, y_2) = g(y_1)h(y_2)$ |
| Exercise 5.43 | $f(y_1 \mid y_2) = f_1(y_1)$ whenever $f_2(y_2) > 0$ |
| $n$ variables | $F(y_1, \ldots, y_n) = F_1(y_1)\cdots F_n(y_n)$ |
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- Independence means the joint distribution is the product of the marginals, **everywhere**

- Tables: check every cell to prove it, find one cell to refute it; a zero cell is the quickest refutation

- Densities: a rectangle support plus a product formula proves it (Theorem 5.5), no integration needed

- A support that is not a rectangle proves dependence, however neatly the formula factors

- A common factor (oil, a storm) makes variables co-move; at correlation 0.6, $P(\text{both fall}) = 0.352$, not $0.5 \times 0.5 = 0.25$
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition**

- Exercises at the end of §5.4: start with 5.45, 5.48, 5.51 and 5.53, then 5.56, 5.60, 5.61, and 5.63 – 5.64

- Return to the cash-demand example and find $P(Y_1 < 1,\ Y_2 < 1)$ as a product of two marginal probabilities

**Week 13, Problem Set 1** is open now and closes **Sunday 13 December at 23:59** on WeBWorK, covering §5.4.

**Next class (5 December):** Wackerly §§5.5–5.7, expected values of functions of random variables and covariance.
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**

📧 sorujov@ada.edu.az\
🏢 Building D, Room D325\
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- Is a random variable independent of itself? When?

- Can two variables have a triangular support and still have $F(y_1, y_2) = F_1(y_1)F_2(y_2)$ at some points?

- If Brent and the bond were independent, what would the right-hand scatter plot look like?
:::
