---
title: "Mathematical Statistics"
subtitle: "The Multinomial Probability Distribution"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - The Multinomial Distribution"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
library(ggplot2)
theme_set(theme_minimal(base_size = 18))
```

## 🎯 Learning Objectives

::: {style="font-size: 32px"}
By the end of this lecture, you will be able to:

- **Recognise** a multinomial experiment from the five properties of Definition 5.11

- **Evaluate** the multinomial probability function for a given set of cell counts

- **State** Theorem 5.13: the mean, variance and covariance of the cell counts

- **Explain** why every covariance between two cells is negative

- **Combine** Theorem 5.13 with Theorem 5.12 to find the mean and variance of a revenue total
:::

---

## 🗺️ Where We Are

::: {style="font-size: 30px"}
**Wackerly §5.9** · Quiz II is behind us.

Wednesday ended on Theorem 5.12: the variance of $\sum a_i Y_i$ is every weight applied to a variance **and to a covariance**. Portfolio risk lives in those cross terms.

::: {.fragment}
Today, one distribution in which every cross term is known in advance, from the design of the experiment alone. The binomial counted two outcomes per trial; the **multinomial** counts $k$.
:::

::: {.fragment}
A rating review ends in *up*, *stable*, *down* or *default*, not in two classes. How are the counts distributed, and how do they move together?
:::
:::

---

## 📐 A Multinomial Experiment

::: {style="font-size: 30px"}
::: {.callout-important}
## Definition 5.11

1. The experiment consists of $n$ identical trials.
2. The outcome of each trial falls into one of $k$ classes or cells.
3. The probability that a trial falls into cell $i$ is $p_i$, the same from trial to trial, with $p_1 + p_2 + \cdots + p_k = 1$.
4. The trials are independent.
5. $Y_i$ is the number of trials falling into cell $i$, so $Y_1 + Y_2 + \cdots + Y_k = n$.
:::

With $k = 2$ this is exactly the binomial experiment of Chapter 3.
:::

---

## 📐 The Probability Function

::: {style="font-size: 30px"}
::: {.callout-important}
## Definition 5.12

$Y_1, \ldots, Y_k$ have a multinomial distribution with parameters $n$ and $p_1, \ldots, p_k$ if
$$p(y_1, \ldots, y_k) = \frac{n!}{y_1!\,y_2! \cdots y_k!}\; p_1^{y_1} p_2^{y_2} \cdots p_k^{y_k},$$
where each $y_i = 0, 1, \ldots, n$ and $\sum_{i=1}^{k} y_i = n$.
:::

::: {.fragment}
It is Chapter 2 in two factors: **one** ordered sequence with these counts has probability $p_1^{y_1} \cdots p_k^{y_k}$, and the coefficient is the number of ways to partition $n$ trials into groups of sizes $y_1, \ldots, y_k$.
:::
:::

---

## 📊 Worked Example: Rating Transitions

::: {style="font-size: 28px"}
A bank's loan book holds 8 firms rated BB. Over one year each firm, independently, is **upgraded** (0.10), stays **stable** (0.75), is **downgraded** (0.12) or **defaults** (0.03). Find the probability of 1 upgrade, 6 stable, 1 downgrade and no default.

::: {.fragment}
$$p(1, 6, 1, 0) = \frac{8!}{1!\,6!\,1!\,0!}\,(0.10)^1 (0.75)^6 (0.12)^1 (0.03)^0 = 56 \times 0.10 \times 0.17798 \times 0.12 = 0.1196$$
:::

::: {.fragment}
Even the single most likely year, $(0, 7, 1, 0)$, has probability only 0.1281: with four cells, probability is spread over **165** possible outcomes.
:::
:::

---

## 📐 Moments of the Counts

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorem 5.13

If $Y_1, \ldots, Y_k$ have a multinomial distribution with parameters $n$ and $p_1, \ldots, p_k$, then

1. $E(Y_i) = np_i$ and $V(Y_i) = np_iq_i$, where $q_i = 1 - p_i$.
2. $\text{Cov}(Y_s, Y_t) = -np_sp_t$, if $s \neq t$.
:::

::: {.fragment}
**Part 1.** Merge every cell except $i$ into one: each trial lands in cell $i$ or not, so $Y_i$ is **binomial** $(n, p_i)$.
:::

::: {.fragment}
**Part 2.** On a single trial, landing in cell $s$ rules out cell $t$, so the two indicators have covariance $0 - p_sp_t$. Different trials are independent, and Theorem 5.12 adds the $n$ same-trial terms: $-np_sp_t$.
:::
:::

---

## 📱 Worked Example: Tariff Choice

::: {style="font-size: 28px"}
A Baku mobile operator signs 200 new subscribers. Each independently picks **Basic** (10 AZN a month, $p_1 = 0.50$), **Standard** (20 AZN, $p_2 = 0.35$) or **Premium** (35 AZN, $p_3 = 0.15$).

| | Basic | Standard | Premium |
|---|---|---|---|
| $E(Y_i) = np_i$ | 100 | 70 | 30 |
| $V(Y_i) = np_iq_i$ | 50 | 45.5 | 25.5 |

$\text{Cov}(Y_1, Y_2) = -35$, $\ \text{Cov}(Y_1, Y_3) = -15$, $\ \text{Cov}(Y_2, Y_3) = -10.5$.
:::

---

## 💰 Monthly Revenue

::: {style="font-size: 28px"}
Revenue is $R = 10Y_1 + 20Y_2 + 35Y_3$ AZN.

$$E(R) = 10(100) + 20(70) + 35(30) = 3450 \text{ AZN}$$

::: {.fragment}
$$\begin{aligned} V(R) &= 100(50) + 400(45.5) + 1225(25.5) \\ &\quad + 2\big[200(-35) + 350(-15) + 700(-10.5)\big] \end{aligned}$$
:::

::: {.fragment}
$V(R) = 54\,437.5 - 39\,200 = 15\,237.5$, so the standard deviation is **123.4 AZN**. Dropping the covariances would report 233.3 AZN: a subscriber who picks Premium is one who did not pick Basic.
:::
:::

---

## 💻 Checking It by Simulation

```{r}
#| label: tariff-sim
n <- 200; p <- c(0.50, 0.35, 0.15); fee <- c(10, 20, 35)
Y <- t(rmultinom(10000, size = n, prob = p))    # 10,000 cohorts
R <- Y %*% fee
round(c(cov_basic_premium = cov(Y[, 1], Y[, 3]), theory = -n * p[1] * p[3],
        sd_revenue = sd(R), theory = sqrt(15237.5)), 1)
```

```{r}
#| label: tariff-fig
#| echo: false
#| fig-width: 10
#| fig-height: 2.9
cells <- as.data.frame(table(basic = Y[, 1], premium = Y[, 3]))
cells <- subset(cells, Freq > 0)
cells$basic <- as.integer(as.character(cells$basic))
cells$premium <- as.integer(as.character(cells$premium))
ggplot(cells, aes(basic, premium)) +
  geom_tile(aes(fill = Freq)) +
  geom_abline(intercept = 60, slope = -0.3, colour = "#8b2635", linewidth = 1.2) +
  scale_fill_gradient(low = "grey88", high = "#14130f", name = "Cohorts") +
  labs(x = "Basic subscribers", y = "Premium subscribers") +
  theme(legend.title = element_text(size = 16), legend.text = element_text(size = 14))
```

::: {style="font-size: 28px"}
Each tile is a count of cohorts; the red line is $0.3(200 - y_1)$, the average Premium count given $y_1$ Basic. Its slope is the negative covariance made visible.
:::

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 30px"}
A trader classifies each of 20 trading days on the Baku exchange as **up** (0.45), **flat** (0.20) or **down** (0.35), independently.

**Four minutes, in pairs:**

1. What is the probability of exactly 10 up, 4 flat and 6 down days?

2. Let $D = Y_{\text{up}} - Y_{\text{down}}$. Find $E(D)$ and $V(D)$.

3. What is $\text{Cov}(Y_{\text{up}}, Y_{\text{down}})$, and what does its sign say?
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 28px"}
1. Definition 5.12: $\dfrac{20!}{10!\,4!\,6!}(0.45)^{10}(0.20)^4(0.35)^6 = 38\,798\,760 \times \cdots = 0.0389$

::: {.fragment}
2. $E(D) = 20(0.45) - 20(0.35) = 2$. By Theorem 5.12, with the covariance from part 3,
$$V(D) = 20(0.45)(0.55) + 20(0.35)(0.65) - 2(-3.15) = 4.95 + 4.55 + 6.30 = 15.8$$
:::

::: {.fragment}
3. $\text{Cov} = -20(0.45)(0.35) = -3.15$. Every up day is a day that was not down. The negative covariance **widens** the spread of a difference: $-2\,\text{Cov}$ adds to $V(D)$.
:::
:::

---

## 📝 Quiz #1: The Sign of the Covariance {.quiz-question}

Twelve customers each choose Basic (0.50), Standard (0.35) or Premium (0.15), independently. What is $\text{Cov}(Y_1, Y_2)$?

- [$-2.1$]{.correct data-explanation="✅ Theorem 5.13: Cov(Y1, Y2) = −n p1 p2 = −12 × 0.50 × 0.35 = −2.1. It is negative because the counts share a fixed total of 12."}
- $2.1$
- $0$, because the customers choose independently
- $-0.175$

---

## 📝 Quiz #2: Is It Multinomial? {.quiz-question}

Which of these is **not** a multinomial experiment?

- Classifying 30 independent trading days as up, flat or down
- Recording which of three tariffs each of 200 independent new subscribers picks
- [Drawing 5 firms without replacement from a list of 12, four in each of three rating grades]{.correct data-explanation="✅ Without replacement from a small list, each draw changes the probabilities for the next: property 3 (constant p_i) and property 4 (independence) both fail. This is a hypergeometric setting."}
- Rating 50 independent loan applicants as approve, refer or decline

---

## 📋 Key Formulas

::: {style="font-size: 30px"}
| | Statement |
|---|---|
| Definition 5.12 | $p(y_1, \ldots, y_k) = \dfrac{n!}{y_1! \cdots y_k!}\,p_1^{y_1} \cdots p_k^{y_k}$ |
| constraints | $\sum p_i = 1$, $\ \sum y_i = n$ |
| Theorem 5.13, part 1 | $E(Y_i) = np_i$, $\ V(Y_i) = np_iq_i$ |
| Theorem 5.13, part 2 | $\text{Cov}(Y_s, Y_t) = -np_sp_t$, $\ s \neq t$ |
| each marginal | $Y_i \sim$ binomial $(n, p_i)$ |
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- The multinomial is the binomial with $k$ cells instead of two

- Its probability function is one sequence's probability times the number of sequences

- Merge the other cells, and any single count is binomial

- Two counts always covary negatively: they share the fixed total $n$

- For a total such as revenue, Theorem 5.13 supplies the covariances and Theorem 5.12 combines them
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition**

- §5.9: Exercises 5.119 – 5.127; start with 5.119, 5.123, 5.124, 5.125 and 5.126

- In the tariff example, what premium share $p_3$ maximises $V(Y_3)$ for fixed $n$?

**Week 14, Problem Set 2** is open now and closes **Sunday 20 December at 23:59** on WeBWorK, covering §5.9.

**Next class:** 16 December, the bivariate normal distribution and conditional expectations (§§5.10–5.11).
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**

📧 sorujov@ada.edu.az\
🏢 Building D, Room D325\
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- Why can the four counts in the loan book never be independent of one another?

- Given $Y_1 = y_1$, what is the distribution of the remaining counts?

- The correlation of $Y_s$ and $Y_t$ does not depend on $n$. Why?
:::
