---
title: "Mathematical Statistics"
subtitle: "The Normal Probability Distribution"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - The Normal Distribution"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
library(ggplot2)
theme_set(theme_minimal(base_size = 18) +
            theme(legend.text = element_text(size = 18),
                  legend.title = element_text(size = 18)))
```

## 🎯 Learning Objectives

::: {style="font-size: 32px"}
By the end of this lecture, you will be able to:

- **State** Definition 4.8 and read $\mu$ and $\sigma$ off a normal density (Theorem 4.7)

- **Standardise** any normal variable with $Z = (Y - \mu)/\sigma$ and find probabilities from Table 4 and from `pnorm`

- **Invert** the calculation with `qnorm` to find a quantile: a cut-off, a fill setting, a budget

- **Compute** a one-day Value-at-Risk at the normal quantile

- **Explain** why real daily returns have heavier tails than the normal model allows
:::

---

## 🗺️ Where We Are

::: {style="font-size: 30px"}
**Wackerly §4.5**

Saturday: **expected values for continuous random variables**, $E(Y) = \int y f(y)\,dy$, and the **uniform distribution**, where every probability is a length divided by a length.

::: {.fragment}
The uniform made the integral easy. Today's density is the opposite: the one the whole course leans on, and the one whose integral **has no closed form at all**.
:::

::: {.fragment}
So the skill today is not integration. It is **standardising**, and then reading a table or asking R.
:::
:::

---

## ❓ The Question This Lecture Answers

::: {style="font-size: 30px"}
::: {.callout-important}
## A risk desk in Baku
A bank's trading book holds **2,000,000 AZN** in a broad equity index. The daily return has mean $0.04\%$ and standard deviation $1.2\%$.

What loss will the book **not exceed on 99% of trading days**?
:::

::: {.fragment}
That number is the one-day 99% **Value-at-Risk**. To compute it we need a model for the return, a way to find a tail area under that model, and a way to run the calculation backwards from an area to a cut-off.
:::
:::

---

## 📐 Definition 4.8

::: {style="font-size: 30px"}
::: {.callout-note}
## Definition 4.8
A random variable $Y$ is said to have a **normal probability distribution** if and only if, for $\sigma > 0$ and $-\infty < \mu < \infty$, the density function of $Y$ is
$$f(y) = \frac{1}{\sigma\sqrt{2\pi}}\, e^{-(y-\mu)^2/(2\sigma^2)}, \qquad -\infty < y < \infty.$$
:::

::: {.fragment}
Two parameters, $\mu$ and $\sigma$. The exponent is a squared distance from $\mu$, measured in units of $\sigma$: the further from $\mu$, the faster the density falls away.
:::
:::

---

## 📐 Theorem 4.7

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorem 4.7
If $Y$ is normally distributed with parameters $\mu$ and $\sigma$, then
$$E(Y) = \mu \qquad \text{and} \qquad V(Y) = \sigma^2.$$
:::

The proof waits for §4.9, via the moment-generating function.

::: {.fragment}
- $\mu$ **locates** the centre; the density is symmetric about it, so the median is $\mu$ too

- $\sigma$ **measures the spread**: the peak is $1/(\sigma\sqrt{2\pi})$ at $y = \mu$, and the curve bends at $\mu \pm \sigma$ (Exercises 4.78, 4.79)
:::
:::

---

## 📊 $\mu$ Locates, $\sigma$ Spreads

```{r}
#| label: three-normals
#| echo: false
#| fig-width: 10
#| fig-height: 4.6
y  <- seq(-6, 6, by = 0.02)
dd <- rbind(
  data.frame(y = y, f = dnorm(y, 0, 0.8), model = "a"),
  data.frame(y = y, f = dnorm(y, 0, 1.6), model = "b"),
  data.frame(y = y, f = dnorm(y, 2, 0.8), model = "c")
)
ggplot(dd, aes(y, f, colour = model)) +
  geom_line(linewidth = 1.3) +
  scale_colour_manual(values = c("#14130f", "#3d6b9e", "#8b2635"),
    labels = c(expression(paste("calm index:  ", mu == 0, ",  ", sigma == 0.8)),
               expression(paste("volatile index:  ", mu == 0, ",  ", sigma == 1.6)),
               expression(paste("shifted:  ", mu == 2, ",  ", sigma == 0.8)))) +
  labs(x = "Daily return (%)", y = "f(y)", colour = NULL) +
  theme(legend.position = "top", legend.direction = "vertical",
        legend.text = element_text(size = 18, hjust = 0))
```

::: {style="font-size: 28px"}
Doubling $\sigma$ halves the peak and doubles the width. Area stays 1.
:::

---

## 🧮 No Closed Form: Table 4 and R

::: {style="font-size: 30px"}
$$P(a \le Y \le b) = \int_a^b \frac{1}{\sigma\sqrt{2\pi}}\, e^{-(y-\mu)^2/(2\sigma^2)}\,dy$$
has **no closed-form** antiderivative. It is evaluated numerically, and there are two ways to get the answer:

- **Table 4, Appendix 3** gives $A(z) = P(Z > z)$, the area to the **right** of $z$, for the standard normal only

- **R**: `pnorm(y0, mu, sigma)` gives $P(Y \le y_0)$; `qnorm(p, mu, sigma)` gives the $p$th quantile $\phi_p$ with $P(Y \le \phi_p) = p$

::: {.fragment}
The only advantage of software is **more decimal places**. Watch the direction: the table is a right tail, `pnorm` is a left tail.
:::
:::

---

## 🔄 One Table for Every Normal

::: {style="font-size: 30px"}
There are infinitely many normal distributions, but one table is enough, because
$$Z = \frac{Y - \mu}{\sigma}$$
is a **standard normal** random variable: mean 0, standard deviation 1 (proved in Chapter 6).

::: {.fragment}
$z$ is the distance from the mean **in standard deviations**. So
$$P(Y > y_0) = P\!\left(Z > \frac{y_0 - \mu}{\sigma}\right) = A\!\left(\frac{y_0 - \mu}{\sigma}\right).$$
:::

::: {.fragment}
A return of $-2.36\%$ on a day with $\mu = 0.04\%$, $\sigma = 1.2\%$ is $z = -2$: two standard deviations below the mean, whatever the units.
:::
:::

---

## 📊 Example 4.8: Table vs R

::: {style="font-size: 28px"}
$Z$ standard normal, Table 4:

**(a)** $P(Z > 2) = A(2.0) = .0228$\
**(b)** $P(-2 \le Z \le 2) = 1 - 2(.0228) = .9544$\
**(c)** $P(0 \le Z \le 1.73) = .5 - A(1.73) = .5 - .0418 = .4582$
:::

```{r}
#| label: example-4-8
#| code-fold: false
round(c(a = 1 - pnorm(2),
        b = pnorm(2) - pnorm(-2),
        c = pnorm(1.73) - 0.5), 5)
```

::: {style="font-size: 28px"}
Part (b) differs in the fourth decimal: $.9544$ from the rounded table, $.95450$ from R. Symmetry did the work in (b) and (c): the table only covers one side.
:::

---

## 🏭 Worked Example: Filling Flour Sacks

::: {style="font-size: 28px"}
A flour mill in Ganja packs **50 kg** sacks. The fill weight is normal with $\mu = 50.2$ kg and $\sigma = 0.4$ kg. What fraction of sacks is **underweight**?

::: {.fragment}
$$z = \frac{50 - 50.2}{0.4} = -0.5, \qquad P(Y < 50) = P(Z < -0.5) = A(0.5) = .3085$$
Almost a third of the sacks break the label. Symmetry turned a left tail at $-0.5$ into the table's right tail at $+0.5$.
:::

::: {.fragment}
Within $49.5$ to $51$ kg: $z_1 = -1.75$, $z_2 = 2.0$, so
$$P = 1 - A(1.75) - A(2.0) = 1 - .0401 - .0228 = .9371$$
:::
:::

---

## 🎯 Backwards: Setting the Mean

::: {style="font-size: 28px"}
The regulator allows **1%** underweight. Keep $\sigma = 0.4$; where must the machine's mean be set?

::: {.fragment}
We need $P(Y < 50) = .01$, so $50$ must sit $z_{.01}$ standard deviations below $\mu$. Table 4: $A(2.33) = .0099 \approx .01$.
$$\mu = 50 + 2.33(0.4) = 50.932 \text{ kg}$$
:::

```{r}
#| label: fill-mean
#| code-fold: false
50 + qnorm(0.99) * 0.4
```

::: {.fragment}
The price of the rule: about **0.73 kg of extra flour per sack** over today's setting. The alternative is a better machine: a smaller $\sigma$ needs a smaller safety margin.
:::
:::

---

## 💳 Standardising Credit Scores

::: {style="font-size: 28px"}
A Baku bank scores applicants on a scale that is normal with $\mu = 640$, $\sigma = 80$, and approves at **700** or above.

::: {.fragment}
$$z = \frac{700 - 640}{80} = 0.75, \qquad P(Y \ge 700) = A(0.75) = .2266$$
:::

::: {.fragment}
It buys a second bureau's scores, normal with $\mu = 520$, $\sigma = 60$. The **comparable** cut-off is the one with the same $z$:
$$520 + 0.75(60) = 565$$
:::

::: {.fragment}
Between 600 and 720 on the first scale: $1 - A(0.5) - A(1.0) = 1 - .3085 - .1587 = .5328$.
:::
:::

---

## 📉 Value-at-Risk at the Normal Quantile

::: {style="font-size: 28px"}
Back to the trading book: $R \sim$ normal, $\mu = 0.04\%$, $\sigma = 1.2\%$, position $2{,}000{,}000$ AZN.

::: {.fragment}
The 1% quantile of the return is $\mu - z_{.01}\,\sigma$. With the table's $2.33$:
$$\phi_{.01} = 0.04 - 2.33(1.2) = -2.756\%$$
:::

::: {.fragment}
$$\text{VaR}_{99\%} = 0.02756 \times 2{,}000{,}000 \approx 55{,}120 \text{ AZN}$$
On 99 days in 100 the book loses less than this. On the other one day it loses **more**, and the model says nothing about how much more.
:::
:::

---

## 💻 VaR in R

```{r}
#| label: var
V   <- 2e6; mu <- 0.04; sigma <- 1.2
q01 <- qnorm(0.01, mu, sigma)        # 1% quantile of the return, in %
data.frame(quantile_pct = round(q01, 4), VaR_AZN = round(-q01 / 100 * V))
```

```{r}
#| label: var-figure
#| echo: false
#| fig-width: 10
#| fig-height: 3.3
r <- seq(-5, 5, by = 0.01)
tail_df <- data.frame(r = r[r <= q01], f = dnorm(r[r <= q01], mu, sigma))
ggplot(data.frame(r = r, f = dnorm(r, mu, sigma)), aes(r, f)) +
  geom_area(data = tail_df, fill = "#8b2635", alpha = 0.7) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = q01, linetype = "dashed") +
  annotate("text", x = q01 - 0.1, y = 0.25, hjust = 1, size = 6,
           label = "1% of days") +
  labs(x = "Daily return (%)", y = "f(r)")
```

::: {style="font-size: 28px"}
`qnorm` gives $55{,}032$ AZN; the table's rounded $2.33$ overstates by about 90 AZN.
:::

---

## ⚠️ Why Real Tails Are Heavier

::: {style="font-size: 28px"}
Under the normal model a move beyond $4\sigma$ has probability $2A(4) \approx 0.000063$: **once in about 63 years** of 250 trading days. Equity indices produce such days far more often.

A heavier-tailed model with the **same** $\sigma$ (Student's $t$ with 3 df, rescaled; met properly in Chapter 7), simulated for ten years:
:::

```{r}
#| label: heavy-tails
#| code-fold: true
n_days <- 2500
r_t <- 0.04 + 1.2 * rt(n_days, df = 3) / sqrt(3)   # sd = 1.2, fat tails
data.frame(beyond = c("3 sigma", "4 sigma"),
           simulated_days  = c(sum(abs(r_t - 0.04) > 3 * 1.2),
                               sum(abs(r_t - 0.04) > 4 * 1.2)),
           normal_expected = round(n_days * 2 * pnorm(-c(3, 4)), 2))
```

---

## 📈 The Tail, on a Log Scale

```{r}
#| label: tail-figure
#| echo: false
#| fig-width: 10
#| fig-height: 4.4
k  <- seq(1, 5, by = 0.05)
tl <- rbind(data.frame(k = k, p = 2 * pnorm(-k),          model = "normal"),
            data.frame(k = k, p = 2 * pt(-k * sqrt(3), 3), model = "heavy-tailed (t, 3 df)"))
ggplot(tl, aes(k, p, colour = model)) +
  geom_line(linewidth = 1.3) +
  scale_y_log10(breaks = 10^(-6:0), labels = c("0.000001", "0.00001", "0.0001",
                                               "0.001", "0.01", "0.1", "1")) +
  scale_colour_manual(values = c("#8b2635", "#14130f")) +
  labs(x = "k (standard deviations from the mean)",
       y = expression(P(abs(R - mu) > k * sigma)), colour = NULL) +
  theme(legend.position = "top")
```

::: {style="font-size: 28px"}
Same mean, same $\sigma$; the curves cross near $2\sigma$. At $4\sigma$ the heavy tail is about **100 times** as likely, so normal VaR **understates** extreme losses.
:::

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 30px"}
A mobile operator models a subscriber's monthly data use as normal with $\mu = 12$ GB and $\sigma = 4$ GB.

**Four minutes, in pairs:**

1. What fraction of subscribers use more than 20 GB?

2. A "heavy user" tariff targets the top 5%. Where is its threshold?

3. What does the model say about $P(Y < 0)$, and what does that tell you about the model?
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 30px"}
1. $z = (20 - 12)/4 = 2$, so $P(Y > 20) = A(2.0) = .0228$: about 2.3% of subscribers.

2. The top 5% starts at $z_{.05} = 1.645$: $\;12 + 1.645(4) = 18.58$ GB. In R, `qnorm(0.95, 12, 4)`.

::: {.fragment}
3. $P(Y < 0) = P(Z < -3) = A(3.0) = .00135$. Small, but **impossible** usage has positive probability. Data use is bounded below by zero and skewed to the right; the normal is a working approximation near the centre, not a law of nature. Check the support before you trust the tails.
:::
:::

---

## 📝 Quiz #1: Reading the Table {.quiz-question}

$Z$ is standard normal. What is $P(Z > 1.5)$?

- [$0.0668$]{.correct data-explanation="✅ Table 4 tabulates the right tail directly: A(1.5) = .0668. In R, 1 - pnorm(1.5)."}
- $0.9332$
- $0.4332$
- $0.1336$

---

## 📝 Quiz #2: Setting a Budget {.quiz-question}

A power plant's weekly fuel bill is normal with $\mu = 4000$ AZN and $\sigma = 250$ AZN. What budget is exceeded in only 10% of weeks?

- [About $4320$ AZN]{.correct data-explanation="✅ The 90th percentile: 4000 + 1.28(250) = 4320. In R, qnorm(0.90, 4000, 250) = 4320.4."}
- About $4411$ AZN
- About $4490$ AZN
- About $4250$ AZN

---

## 📝 Quiz #3: What the Normal Misses {.quiz-question}

Real daily index returns show many more moves beyond $4\sigma$ than the normal model predicts. What does this do to a 99.9% normal VaR?

- [It understates the loss the book can actually suffer]{.correct data-explanation="✅ The normal quantile sits too close to the mean when the true tail is heavier, so extreme losses are more likely than the VaR suggests."}
- It overstates the loss, so the bank holds too much capital
- Nothing: VaR depends only on $\mu$ and $\sigma$, which are unchanged
- It makes VaR undefined

---

## 📋 Key Formulas

::: {style="font-size: 30px"}
| | Statement |
|---|---|
| Definition 4.8 | $f(y) = \frac{1}{\sigma\sqrt{2\pi}}\, e^{-(y-\mu)^2/(2\sigma^2)}$ |
| Theorem 4.7 | $E(Y) = \mu, \; V(Y) = \sigma^2$ |
| standardising | $Z = (Y - \mu)/\sigma$ |
| Table 4 | $A(z) = P(Z > z)$ |
| R, forwards | `pnorm(y0, mu, sigma)` $= P(Y \le y_0)$ |
| R, backwards | `qnorm(p, mu, sigma)` $= \phi_p$ |
| quantiles | $z_{.10} = 1.28,\ z_{.05} = 1.645,\ z_{.025} = 1.96,\ z_{.01} = 2.33$ |
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- The normal density has two parameters: $\mu$ is the mean and centre, $\sigma$ the standard deviation and spread

- Its integral has no closed form, so every probability goes through $Z = (Y - \mu)/\sigma$ and then Table 4 or `pnorm`

- Quantiles run the same step backwards: $y = \mu + z\sigma$, or `qnorm`

- A fill setting, an equivalent cut-off and a VaR are all one quantile

- Real returns have heavier tails than the normal, so normal VaR is optimistic about the worst days
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition, §4.5**

- Table practice: 4.58, 4.59; then 4.63 – 4.65 and 4.68 – 4.71

- Quantiles and settings: 4.73 – 4.77; for the brave, 4.78 and 4.79

- Do each one twice: once with Table 4, once with `pnorm`/`qnorm`

**Week 10, Problem Set 1** is open now and closes **Sunday 22 November at 23:59** on WeBWorK, covering §4.5.

**Next class:** 14 November, the gamma, exponential and chi-square distributions (Wackerly §4.6).
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**

📧 sorujov@ada.edu.az\
🏢 Building D, Room D325\
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- Why does one table suffice for infinitely many normal distributions?

- If $\sigma$ halves, what happens to the 99% VaR, and to the fill setting in the flour example?

- Where else in finance would a normal model put positive probability on an impossible value?
:::
