---
title: "Mathematical Statistics"
subtitle: "Interval Estimation & Confidence Intervals"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ../ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    margin: 0.04
    # footer: "Mathematical Statistics – Interval Estimation"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
    include-after-body:
      - text: |
          <script src="../qrcode.min.js"></script>
          <script src="../attendance-qr-slide.js"></script>
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::: {style="font-size:39px"}
::: {.learning-objectives}
By the end of this lecture, you will be able to:

- **Construct** large-sample confidence intervals for $\mu$, $p$, $\mu_1 - \mu_2$, and $p_1 - p_2$ using the $z$-distribution

- **Apply** the pivotal method to derive confidence intervals from sampling distributions

- **Select** the minimum sample size to achieve a desired margin of error

- **Build** small-sample confidence intervals for $\mu$ and $\mu_1 - \mu_2$ using the $t$-distribution

- **Derive** confidence intervals for the population variance $\sigma^2$ using the $\chi^2$ distribution
:::
:::

---

## 📱 Attendance Check-in {.center background-color="#f0f4ff"}

<div class="attendance-qr" data-status="present" id="qr-present" style="min-height: 400px; display: flex; align-items: center; justify-content: center;"></div>

## 📋 Overview {.smaller}

::: {style="font-size:50px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Confidence Intervals** – Definition, confidence coefficient, pivotal method

- **Large-Sample CIs** – For $\mu$, $p$, $\mu_1-\mu_2$, $p_1-p_2$ using CLT

- **Sample Size Selection** – How to plan a study for a target margin of error

- **Small-Sample CIs** – $t$-distribution for means when $n$ is small

- **CI for $\sigma^2$** – Using the $\chi^2$ distribution for variance estimation

- **Case Study** – Confidence intervals for stock return parameters
:::
:::
:::

---

## 📖 From Point to Interval {.smaller}

::: {style="font-size:38px"}

A **point estimate** gives a single number — but how much should we trust it?

::: {.columns}
::: {.column width="50%"}
::: {.fragment}
**Point Estimate:**

> "Based on our sample, mean daily return is **0.05%**"

✅ Simple, but gives no sense of uncertainty
:::
:::
::: {.column width="50%"}
::: {.fragment}
**Interval Estimate:**

> "We are 95% confident mean return is between **0.02% and 0.08%**"

✅ Communicates both the estimate and its precision
:::
:::
:::

::: {.fragment}
::: {.callout-tip}
## 💡 The Finance Analogy

A point estimate is like saying "the bond will yield 4%." An interval estimate says "the yield will fall between 3.7% and 4.3% with 95% confidence." Investors act on intervals, not points.
:::
:::
:::

---

## 📖 Definition: Confidence Interval {.smaller}

::: {style="font-size:38px"}
::: {.callout-note}
## 📝 Confidence Interval

If $\hat{\theta}_L$ and $\hat{\theta}_U$ are the (random) lower and upper confidence limits for parameter $\theta$, and:

$$P\left(\hat{\theta}_L \leq \theta \leq \hat{\theta}_U\right) = 1 - \alpha$$

then $(1-\alpha)$ is the **confidence coefficient**, and $(\hat{\theta}_L,\, \hat{\theta}_U)$ is a **two-sided confidence interval**.
:::

::: {.fragment}
**Key terms:**

- **Confidence coefficient $(1-\alpha)$**: Fraction of such intervals that contain $\theta$ in repeated sampling
- **Confidence level**: Usually expressed as a percent — 90%, 95%, 99%
- **Margin of error**: Half-width of the interval = $z_{\alpha/2} \cdot \sigma_{\hat{\theta}}$
:::
:::

---

## 📖 What "95% Confident" Really Means {.smaller}

::: {style="font-size:40px"}
::: {.callout-important}
## ⚠️ Common Misconception

"There is a 95% probability that $\theta$ lies in **this** interval."

❌ Wrong! The true $\theta$ is **fixed** — it either is or isn't in the interval.
:::

::: {.fragment}
::: {.callout-note}
## ✅ Correct Interpretation

If we repeated our sampling procedure many times and computed a CI each time, **approximately 95% of those intervals** would contain the true $\theta$.

For the one interval we have, we are "95% confident" because we used a procedure that works 95% of the time.
:::
:::

::: {.fragment}
> 💼 *Analogy: A 95% CI is like a fishing net that catches the fish 95% of the time you cast it. Whether the fish is in this particular cast — we don't know.*
:::
:::

---

## 📖 The Pivotal Method {.smaller}

::: {style="font-size:28px"}

A **pivotal quantity** $Q$ satisfies two properties:

::: {.fragment}
1. It is a function of the **sample data and the unknown parameter $\theta$** — but $\theta$ is the **only unknown**
2. Its **probability distribution does not depend on $\theta$**
:::

::: {.fragment}
**Key examples we will use:**

| Scenario | Pivotal Quantity | Distribution |
|----------|-----------------|--------------|
| Large $n$, any population | $Z = \dfrac{\hat{\theta}-\theta}{\sigma_{\hat{\theta}}}$ | $N(0,1)$ |
| Small $n$, normal population, $\sigma^2$ unknown | $T = \dfrac{\bar{Y}-\mu}{S/\sqrt{n}}$ | $t_{n-1}$ |
| Normal population, $\sigma^2$ estimation | $\chi^2 = \dfrac{(n-1)S^2}{\sigma^2}$ | $\chi^2_{n-1}$ |

:::
:::

---

## 📖 Large-Sample CI: The General Formula {.smaller}

::: {style="font-size:28px"}

For large samples, by the CLT:

$$Z = \frac{\hat{\theta} - \theta}{\sigma_{\hat{\theta}}} \;\overset{\text{approx.}}{\sim}\; N(0, 1)$$

::: {.fragment}
Starting from $P\!\left(-z_{\alpha/2} \leq Z \leq z_{\alpha/2}\right) = 1-\alpha$ and isolating $\theta$:

$$\boxed{\hat{\theta} \pm z_{\alpha/2}\,\sigma_{\hat{\theta}}}$$

This is the **universal large-sample CI formula**.
:::

::: {.fragment}
**Common $z_{\alpha/2}$ values:**

| Confidence Level | $\alpha$ | $z_{\alpha/2}$ |
|-----------------|----------|----------------|
| 90% | 0.10 | 1.645 |
| 95% | 0.05 | 1.960 |
| 99% | 0.01 | 2.576 |
:::
:::

---

## 📖 Four Large-Sample CIs (Part 1) {.smaller}

::: {style="font-size:35px"}

**CI for a population mean $\mu$:**

$$\bar{Y} \pm z_{\alpha/2} \cdot \frac{\sigma}{\sqrt{n}} \qquad \left(\text{use } s \text{ if } \sigma \text{ unknown and } n \geq 30\right)$$

::: {.fragment}
**CI for a binomial proportion $p$:**

$$\hat{p} \pm z_{\alpha/2} \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

where $\hat{p} = Y/n$ (substitute $\hat{p}$ for unknown $p$ in the SE formula)
:::

::: {.fragment}
::: {.callout-note}
## 📝 Large-Sample Requirement

For the proportion CI, we need the distribution of $\hat{p}$ to be approximately normal. A safe rule: $n\hat{p} \geq 5$ and $n(1-\hat{p}) \geq 5$.
:::
:::
:::

---

## 📖 Four Large-Sample CIs (Part 2) {.smaller}

::: {style="font-size:33px"}

**CI for a difference in means $\mu_1 - \mu_2$ (independent samples):**

$$(\bar{Y}_1 - \bar{Y}_2) \pm z_{\alpha/2} \sqrt{\frac{\sigma_1^2}{n_1} + \frac{\sigma_2^2}{n_2}}$$

(substitute $s_1^2$, $s_2^2$ when $\sigma^2$ unknown and $n_1, n_2 \geq 30$)

::: {.fragment}
**CI for a difference in proportions $p_1 - p_2$:**

$$(\hat{p}_1 - \hat{p}_2) \pm z_{\alpha/2} \sqrt{\frac{\hat{p}_1(1-\hat{p}_1)}{n_1} + \frac{\hat{p}_2(1-\hat{p}_2)}{n_2}}$$
:::

::: {.fragment}
::: {.callout-tip}
## 💡 Interpretation Tip

If the CI for $\mu_1 - \mu_2$ **contains zero**, then zero is a "believable" value — we cannot rule out that the two means are equal at this confidence level.
:::
:::
:::

---

## 📌 Example 1: CI for Mean Bond Duration {.smaller}

::: {style="font-size:36px"}

**Problem:** A portfolio manager samples $n = 64$ sovereign bonds. The mean maturity is $\bar{y} = 7.3$ years with $s^2 = 9.0$. Construct a 90% CI for the true mean maturity $\mu$.

::: {.fragment}
**Solution:**

$\hat{\theta} = \bar{y} = 7.3$, $s = 3.0$, $n = 64$, $z_{0.05} = 1.645$

$$\bar{y} \pm z_{\alpha/2} \cdot \frac{s}{\sqrt{n}} = 7.3 \pm 1.645 \cdot \frac{3.0}{\sqrt{64}} = 7.3 \pm 1.645 \cdot 0.375$$

$$= 7.3 \pm 0.617$$
:::

::: {.fragment}
**90% Confidence Interval: $(6.68,\; 7.92)$ years**

We are 90% confident the true mean bond maturity lies between **6.68 and 7.92 years**. In repeated sampling, 90% of such intervals would contain $\mu$.
:::
:::

---

## 📌 Example 2: CI for a Default Rate Difference {.smaller}

::: {style="font-size:31px"}

**Problem:** Two loan portfolios are compared. Portfolio A: $n_1 = 50$, 12 defaults ($\hat{p}_1 = 0.24$). Portfolio B: $n_2 = 60$, 12 defaults ($\hat{p}_2 = 0.20$). Construct a 98% CI for $p_1 - p_2$.

::: {.fragment}
**Solution:** $z_{0.01} = 2.33$

$$(\hat{p}_1 - \hat{p}_2) \pm z_{\alpha/2}\sqrt{\frac{\hat{p}_1\hat{q}_1}{n_1} + \frac{\hat{p}_2\hat{q}_2}{n_2}}$$

$$= (0.24 - 0.20) \pm 2.33\sqrt{\frac{(0.24)(0.76)}{50} + \frac{(0.20)(0.80)}{60}}$$

$$= 0.04 \pm 2.33(0.0795) \approx 0.04 \pm 0.185$$
:::

::: {.fragment}
**98% CI: $(-0.145,\; 0.225)$**

Since zero is in the interval, we **cannot conclude** that the default rates differ at the 98% confidence level.
:::
:::

---

## 📖 Sample Size Selection {.smaller}

::: {style="font-size:33px"}

**Goal:** Choose $n$ so the error of estimation is less than $B$ with confidence $1-\alpha$.

**General formula:** Set $z_{\alpha/2} \cdot \sigma_{\hat{\theta}} = B$ and solve for $n$.

::: {.fragment}
**For estimating $\mu$:**

$$n = \left(\frac{z_{\alpha/2} \cdot \sigma}{B}\right)^2$$

(Use a prior estimate $s$ or approximate $\sigma \approx \text{range}/4$ if unknown)
:::

::: {.fragment}
**For estimating $p$:**

$$n = \frac{z_{\alpha/2}^2 \cdot p(1-p)}{B^2}$$

(Use $p = 0.5$ for the most conservative/largest sample size if $p$ is unknown)
:::
:::

---

## 📌 Example 3: Sizing a Market Survey {.smaller}

::: {style="font-size:32px"}

**Problem:** A telecommunications regulator wants to estimate the proportion of subscribers $p$ who report broadband speeds below the advertised rate. The estimate must be correct to within $B = 0.03$ with 95% confidence. How large must the sample be?

::: {.fragment}
**Solution:** $z_{0.025} = 1.96$. No prior info on $p$, so use $p = 0.5$:

$$n = \frac{z_{\alpha/2}^2 \cdot p(1-p)}{B^2} = \frac{(1.96)^2 \cdot (0.5)(0.5)}{(0.03)^2} = \frac{3.8416 \times 0.25}{0.0009} \approx 1,068$$
:::

::: {.fragment}
**Interpretation:** At least **1,068 subscribers** must be sampled.

If we know from a previous study that $p \approx 0.25$, then:

$$n = \frac{(1.96)^2 \cdot (0.25)(0.75)}{(0.03)^2} \approx 801$$

Prior information reduces the required sample by **25%** — saving survey cost.
:::
:::

---

## 📌 Example 4: Sizing a Two-Group Comparison {.smaller}

::: {style="font-size:35px"}

**Problem:** A regulator wants to estimate the difference in mean download speeds between two ISPs, correct to within $B = 5$ Mbps with 95% confidence. Previous data suggest $\sigma \approx 12$ Mbps for both ISPs. Equal sample sizes will be used. Find $n_1 = n_2$.

::: {.fragment}
**Solution:** $z_{0.025} = 1.96$, $\sigma_1 = \sigma_2 = 12$

$$z_{\alpha/2}\sqrt{\frac{\sigma_1^2}{n} + \frac{\sigma_2^2}{n}} = B \implies 1.96\sqrt{\frac{144 + 144}{n}} = 5$$

$$1.96 \cdot \frac{12\sqrt{2}}{\sqrt{n}} = 5 \implies \sqrt{n} = \frac{1.96 \times 16.97}{5} = 6.652 \implies n \approx 45$$
:::

::: {.fragment}
**Each group needs at least $n = 45$ measurements** — totaling 90 speed tests.
:::
:::

---

## 🎮 Interactive: Confidence Level Explorer 

:::::: {style="font-size: 0.72em; margin-top: -8px;"}

**See how confidence level and sample size affect interval width.**

::::: columns
::: {.column width="32%"}

```{ojs}
//| echo: false

viewof conf_level = {
  const input = Inputs.range([0.80, 0.99], {value: 0.95, step: 0.01, label: "Conf. level:"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

viewof n_size = {
  const input = Inputs.range([10, 500], {value: 100, step: 10, label: "Sample size n:"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

viewof sigma_known = {
  const input = Inputs.range([0.5, 5], {value: 2, step: 0.1, label: "Std dev σ:"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

alpha_val   = 1 - conf_level
z_val       = jStat.normal.inv(1 - alpha_val/2, 0, 1)
margin      = z_val * sigma_known / Math.sqrt(n_size)
half_w_pct  = (margin * 100).toFixed(3)

md`
**z\u208A/\u208B = ${z_val.toFixed(3)}**

**Margin of error = ±${margin.toFixed(4)}%**

**Interval width = ${(2*margin).toFixed(4)}%**
`
```

:::

::: {.column width="68%"}

```{ojs}
//| echo: false

jStat = require("https://cdnjs.cloudflare.com/ajax/libs/jstat/1.9.6/jstat.min.js")

{
  const mu = 0.05;
  const nSims = 40;
  const rng = d3.randomNormal(mu, sigma_known / Math.sqrt(n_size));
  const bars = Array.from({length: nSims}, () => {
    const ybar = rng();
    return {lo: ybar - margin, hi: ybar + margin, ybar, covers: ybar - margin <= mu && mu <= ybar + margin};
  });

  const covered = bars.filter(b => b.covers).length;

  return Plot.plot({
    width: 800,
    height: 500,
    marginLeft: 30,
    marginBottom: 40,
    x: {domain: [mu - 4*margin, mu + 4*margin], label: "Parameter value"},
    y: {domain: [-1, nSims], label: ""},
    title: `${nSims} simulated ${(conf_level*100).toFixed(0)}% CIs — ${covered}/${nSims} capture μ`,
    marks: [
      Plot.ruleX([mu], {stroke: "red", strokeWidth: 2.5, strokeDasharray: "6,3"}),
      Plot.link(bars.map((b,i) => ({...b, i})), {
        x1: "lo", x2: "hi", y1: "i", y2: "i",
        stroke: d => d.covers ? "steelblue" : "#e74c3c",
        strokeWidth: 2.2
      }),
      Plot.dot(bars.map((b,i) => ({...b, i})), {x: "ybar", y: "i", r: 3,
        fill: d => d.covers ? "steelblue" : "#e74c3c"}),
      Plot.text([{x: mu, y: nSims + 0.5}], {x:"x", y:"y", text: d => "μ (true)", fill:"red", fontSize:12})
    ]
  });
}
```

:::
:::::
::::::

---

## 📖 Small-Sample CIs: The $t$-Distribution {.smaller}

::: {style="font-size:33px"}

When $n$ is **small** and $\sigma^2$ is **unknown**, we assume normality and use:

$$T = \frac{\bar{Y} - \mu}{S/\sqrt{n}} \;\sim\; t_{n-1}$$

::: {.fragment}
The resulting **100$(1-\alpha)$% CI for $\mu$** is:

$$\boxed{\bar{Y} \pm t_{\alpha/2,\, n-1} \cdot \frac{S}{\sqrt{n}}}$$

where $t_{\alpha/2, n-1}$ is the $t$-critical value with $n-1$ degrees of freedom.
:::

::: {.fragment}
::: {.callout-note}
## 📝 Why $t$ instead of $z$?

The $t$-distribution has **heavier tails** than $N(0,1)$ — reflecting the extra uncertainty from estimating $\sigma^2$ with $S^2$. As $n \to \infty$, $t_{n-1} \to N(0,1)$ (they agree for $\nu > 30$).
:::
:::
:::

---

## 📌 Example 5: Hedge Fund Monthly Returns {.smaller}

::: {style="font-size:35px"}

**Problem:** An analyst observes only $n = 8$ monthly returns (%) from a hedge fund: 2.1, 1.8, 3.0, 0.5, 2.6, 1.2, 2.9, 1.5. Construct a 95% CI for the true mean monthly return $\mu$. Assume normality.

::: {.fragment}
**Solution:** $\bar{y} = 1.95\%$, $s = 0.836\%$, $n-1 = 7$ df, $t_{0.025, 7} = 2.365$

$$\bar{y} \pm t_{\alpha/2,\, 7} \cdot \frac{s}{\sqrt{n}} = 1.95 \pm 2.365 \cdot \frac{0.836}{\sqrt{8}} = 1.95 \pm 0.699$$
:::

::: {.fragment}
**95% CI: $(1.25\%,\; 2.65\%)$ per month**

Annualized: approximately **(15.0%, 31.8%)** — a wide range reflecting the small sample.

> 💡 If we had used $z_{0.025} = 1.96$ instead of $t = 2.365$, the interval would be **artificially narrow**, understating uncertainty.
:::
:::

---

## 📖 Two-Sample $t$-CI for $\mu_1 - \mu_2$ {.smaller}

::: {style="font-size:32px"}

**Assumptions:** Independent random samples from two **normal** populations with **equal but unknown variances** ($\sigma_1^2 = \sigma_2^2 = \sigma^2$).

::: {.fragment}
**Pooled variance estimate:**

$$S_p^2 = \frac{(n_1-1)S_1^2 + (n_2-1)S_2^2}{n_1 + n_2 - 2}$$
:::

::: {.fragment}
**100$(1-\alpha)$% CI for $\mu_1 - \mu_2$:**

$$(\bar{Y}_1 - \bar{Y}_2) \pm t_{\alpha/2,\; n_1+n_2-2} \cdot S_p\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}$$
:::

::: {.fragment}
::: {.callout-tip}
## 💡 When does this apply?

Use whenever sample sizes are small (< 30) and the two population variances are believed to be roughly equal. Check equality of variances with a Levene test or $F$-test first.
:::
:::
:::

---

## 📌 Example 6: Comparing Two Training Programs {.smaller}

::: {style="font-size:33px"}

**Problem:** Two analyst training programs are compared on a standardized performance score. Program A ($n_1 = 9$): $\bar{y}_1 = 35.2$, $s_1^2 = 24.4$. Program B ($n_2 = 9$): $\bar{y}_2 = 31.6$, $s_2^2 = 20.0$. Construct a 95% CI for $\mu_1 - \mu_2$.

::: {.fragment}
**Pooled variance:** $S_p^2 = \dfrac{8(24.4) + 8(20.0)}{16} = 22.2$, so $S_p = 4.71$

$t_{0.025, 16} = 2.120$
:::

::: {.fragment}
$$(\bar{y}_1 - \bar{y}_2) \pm t_{\alpha/2,16} \cdot S_p\sqrt{\frac{1}{9}+\frac{1}{9}} = 3.6 \pm 2.120 \times 4.71 \times 0.471 = 3.6 \pm 4.71$$

**95% CI: $(-1.11,\; 8.31)$**

The interval contains zero → we **cannot conclude** one program is better at 95% confidence.
:::
:::

---

## 📖 CI for Population Variance $\sigma^2$ {.smaller}

::: {style="font-size:30px"}

When the population is **normal**, the pivotal quantity is:

$$\chi^2 = \frac{(n-1)S^2}{\sigma^2} \;\sim\; \chi^2_{n-1}$$

::: {.fragment}
Finding $\chi^2_{1-\alpha/2}$ and $\chi^2_{\alpha/2}$ so the middle area = $1-\alpha$:

$$P\!\left(\chi^2_{1-\alpha/2} \leq \frac{(n-1)S^2}{\sigma^2} \leq \chi^2_{\alpha/2}\right) = 1 - \alpha$$
:::

::: {.fragment}
**100$(1-\alpha)$% CI for $\sigma^2$:**

$$\boxed{\left(\frac{(n-1)S^2}{\chi^2_{\alpha/2}},\;\; \frac{(n-1)S^2}{\chi^2_{1-\alpha/2}}\right)}$$

Note: The $\chi^2$ distribution is **asymmetric** — the CI is not symmetric around $S^2$.
:::
:::

---

## 📌 Example 7: CI for Return Volatility {.smaller}

::: {style="font-size:33px"}

**Problem:** A risk officer records $n = 10$ quarterly returns for a bond fund (%), and computes $s^2 = 3.84$ (%²). Construct a 90% CI for the true variance $\sigma^2$ of quarterly returns. Assume normality.

::: {.fragment}
**Solution:** $n - 1 = 9$ df, $\alpha/2 = 0.05$

From $\chi^2$ table: $\chi^2_{0.05, 9} = 16.919$, $\chi^2_{0.95, 9} = 3.325$

$$\left(\frac{9 \times 3.84}{16.919},\;\; \frac{9 \times 3.84}{3.325}\right) = \left(\frac{34.56}{16.919},\;\; \frac{34.56}{3.325}\right)$$
:::

::: {.fragment}
**90% CI for $\sigma^2$: $(2.04,\; 10.40)$ (%²)**

For the standard deviation: $\sigma \in (\sqrt{2.04},\, \sqrt{10.40}) = (1.43\%,\; 3.22\%)$

> ⚠️ The CI for $\sigma^2$ is highly sensitive to the normality assumption — unlike CIs for means.
:::
:::

---

## 🤝 Think-Pair-Share {.smaller}

::: {style="font-size:49px;"}
::: {.callout-note}
## 💬 Discussion (4 minutes)

A central bank reports: *"Based on a sample of $n=25$ quarterly GDP growth rates, our 95% confidence interval for mean growth is $(1.8\%,\; 3.2\%)$."*

**Discuss with your partner:**

1. What distribution was likely used — $z$ or $t$? Why?
2. What assumptions are required for this CI to be valid?
3. A critic says: "There's a 95% chance true GDP growth is in this interval." Is this correct?
4. If the bank doubled its sample to $n=50$, what would happen to the interval width?
:::
:::

```{r}
#| echo: false
#| eval: true
library(countdown)
countdown(minutes = 4, bottom = 0, right = 0,
          color_running_background = "#2c7bb6",
          color_finished_background = "#d73027")
```

---

## 💰 Case Study: 95% t-CIs for Mean Return 

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

library(tidyverse)
library(tidyquant)
library(knitr)

symbols <- c("AAPL", "JPM", "GLD")
prices <- tq_get(symbols, from = "2022-01-01", to = "2023-12-31")

returns <- prices %>%
  group_by(symbol) %>%
  tq_transmute(select = adjusted, mutate_fun = periodReturn,
               period = "monthly", col_rename = "r")

ci_tbl <- returns %>%
  group_by(symbol) %>%
  summarise(n = n(), mu = mean(r) * 100, s = sd(r) * 100,
            t_cv = qt(0.975, df = n() - 1),
            lo = mu - t_cv * s / sqrt(n()),
            hi = mu + t_cv * s / sqrt(n()))

kable(ci_tbl %>% select(symbol, n, mu, s, lo, hi), digits = 3,
      col.names = c("Asset","n","μ̂ (%)","s (%)","CI Lower (%)","CI Upper (%)"),
      caption = "95% t-Confidence Intervals for Mean Monthly Return")
```

---

## 💰 Case Study: CI Visualisation {.smaller}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 9
#| fig-height: 4

ci_tbl %>%
  ggplot(aes(x = symbol, y = mu, color = symbol)) +
  geom_point(size = 5) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.25, linewidth = 1.4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "95% Confidence Intervals for Mean Monthly Return",
       subtitle = "AAPL, JPM, GLD — 2022–2023",
       x = "Asset", y = "Mean Monthly Return (%)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

---

## 💰 Case Study: CI for Volatility {.smaller}

::: {style="font-size:36px"}
::: {.columns}
::: {.column width="60%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# 90% chi-squared CI for variance of monthly returns
returns_wide <- returns %>%
  pivot_wider(names_from = symbol, values_from = r) %>%
  na.omit()

ci_var <- map_dfr(c("AAPL","JPM","GLD"), function(sym) {
  x  <- returns_wide[[sym]] * 100
  n  <- length(x)
  s2 <- var(x)
  chi_hi <- qchisq(0.95, df = n - 1)
  chi_lo <- qchisq(0.05, df = n - 1)
  tibble(
    Symbol = sym,
    n = n,
    `S² (%²)` = round(s2, 3),
    `CI Lower (%²)` = round((n-1)*s2/chi_hi, 3),
    `CI Upper (%²)` = round((n-1)*s2/chi_lo, 3),
    `σ̂ (%)` = round(sqrt(s2), 3)
  )
})

kable(ci_var, caption = "90% Chi-Squared CIs for Return Variance")
```

:::
::: {.column width="40%"}

::: {.fragment}
::: {.callout-important}
## 📊 Key Takeaways

- **GLD** has the widest CI for mean return — highest volatility

- All mean-return CIs **include zero** — consistent with market efficiency for monthly data

- The **variance CIs are asymmetric** — the chi-squared distribution is skewed right

- With only ~24 observations, CIs are wide — more data needed for precise volatility estimates

:::
:::

:::
:::
:::

---

## 📝 Quiz #1: CI Construction {.smaller .quiz-question}

A sample of $n = 100$ daily ISP speed measurements yields $\bar{y} = 48.2$ Mbps and $s = 12.0$ Mbps. Which formula gives the correct 95% CI for $\mu$?

- [$48.2 \pm 1.96 \times \frac{12.0}{\sqrt{100}}$]{.correct data-explanation="✅ Correct! Large n (≥30), σ unknown → use z = 1.96 with s/√n. The CI is (45.85, 50.55) Mbps."}
- $48.2 \pm 2.326 \times \frac{12.0}{\sqrt{100}}$
- $48.2 \pm 1.96 \times 12.0$
- $48.2 \pm 1.645 \times \frac{12.0}{\sqrt{100}}$

---

## 📝 Quiz #2: Interpreting a CI {.smaller .quiz-question}

A 95% CI for mean GDP growth is $(1.8\%,\; 3.2\%)$. Which statement is correct?

- [If this procedure were repeated many times, about 95% of such intervals would contain the true mean growth rate.]{.correct data-explanation="✅ Correct! The confidence coefficient refers to long-run frequency — not to the probability that this specific interval contains the parameter."}
- There is a 95% probability that the true mean growth lies in $(1.8\%, 3.2\%)$.
- 95% of all GDP growth observations fall in $(1.8\%, 3.2\%)$.
- The true mean growth rate is definitely in $(1.8\%, 3.2\%)$.

---

## 📝 Quiz #3: Sample Size {.smaller .quiz-question}

A regulator wants to estimate a failure rate $p$ to within $B = 0.04$ with 95% confidence. With no prior knowledge of $p$, the minimum sample size is approximately:

- [601]{.correct data-explanation="✅ Correct! n = (1.96² × 0.5 × 0.5) / 0.04² = (3.8416 × 0.25) / 0.0016 = 600.25 → 601."}
- 385
- 1068
- 246

---

## 📝 Quiz #4: Small vs. Large Sample {.smaller .quiz-question}

When would you use a $t$-distribution instead of $z$ for a CI on $\mu$?

- [Small sample ($n < 30$), normal population, and $\sigma^2$ unknown]{.correct data-explanation="✅ Correct! The t-distribution accounts for additional uncertainty from estimating σ² with S². For large n, t ≈ z and either works."}
- Always, because $t$ is more conservative
- Large sample from a non-normal population
- When the population variance $\sigma^2$ is known

---

## 📝 Summary {.smaller}

::: {style="font-size:36px"}
::: {.summary-box}
**✅ Key Takeaways**

- **Confidence Interval** = point estimate ± margin of error; the CI coefficient is a long-run frequency, not a probability about a fixed parameter

- **Large-sample CI** uses $\hat{\theta} \pm z_{\alpha/2}\sigma_{\hat{\theta}}$; applies to $\mu$, $p$, $\mu_1-\mu_2$, $p_1-p_2$ via the CLT

- **Sample Size**: Set $z_{\alpha/2}\sigma_{\hat{\theta}} = B$ and solve for $n$; use $p=0.5$ for proportions with no prior info

- **Small-sample CI** uses $\bar{Y} \pm t_{\alpha/2,n-1} \cdot S/\sqrt{n}$; requires normality; pooled $t$ for two-sample case with equal variances

- **CI for $\sigma^2$** uses the chi-squared distribution; is asymmetric and highly sensitive to the normality assumption
:::
:::

---

## 📚 Practice Problems

::: {style="font-size:42px;"}
::: {.callout-tip}
## 📝 Homework Problems

**Problem 1 (Large-Sample CI):** A random sample of $n = 200$ bond funds yields mean annual return $\bar{y} = 6.8\%$ with $s = 4.2\%$. Construct (a) a 90% CI and (b) a 99% CI for the true mean return. Compare and interpret. (Wackerly §8.6)

**Problem 2 (Sample Size):** You want to estimate the proportion of Azerbaijani broadband subscribers experiencing peak-hour speeds below 50% of their plan, to within 0.02 with 95% confidence. (a) If $p \approx 0.35$, find $n$. (b) Find $n$ if $p$ is unknown. (Wackerly §8.7)

**Problem 3 (Two-Sample $t$):** Fund A: $n_1 = 12$, $\bar{y}_1 = 8.4\%$, $s_1 = 3.1\%$. Fund B: $n_2 = 10$, $\bar{y}_2 = 6.9\%$, $s_2 = 2.8\%$. Assuming equal variances and normality, construct a 95% CI for $\mu_A - \mu_B$. Can you conclude Fund A outperforms Fund B? (Wackerly Ex. 8.85)

**Problem 4 (CI for $\sigma^2$):** A quality control analyst measures quarterly return variance for a risk model. From $n = 15$ observations, $s^2 = 5.76$ (%²). Construct a 95% CI for $\sigma^2$ and interpret in terms of annualized volatility. (Wackerly §8.9)
:::
:::

---

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business

ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Properties of Estimators

**Reading:** Chapter 9 — Sections 9.1–9.5

**Preparation:** Review maximum likelihood from any prior exposure; re-read bias and MSE from Chapter 8

**⏰ Reminders:**

✅ Complete Practice Problems 1–4

✅ Make sure you understand the difference between $z$ and $t$ critical values

✅ Review chi-squared table (Table 6, Appendix 3)

✅ Work hard!
:::
:::

---

## ❓ Questions? {.center}

::: {.callout-note}
## 💬 Open Discussion

**Key Topics for Discussion:**

- Why does the chi-squared CI for $\sigma^2$ fail when the population is not normal, but the $t$-CI for $\mu$ is robust?

- If a 95% CI contains zero, does that mean the true parameter is zero?

- How would you explain "confidence level" versus "probability" to a non-statistician at a bank?

- Why does increasing sample size always narrow a CI — regardless of the true parameter value?
:::

<div class="attendance-qr" data-status="late" id="qr-late" style="min-height: 400px; display: flex; align-items: center; justify-content: center;"></div>
