---
title: "Mathematical Statistics"
subtitle: "Point Estimation"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ../ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    margin: 0.08
    # footer: "Mathematical Statistics – Point Estimation"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
    include-after-body:
      - text: |
          <script src="../qrcode.min.js"></script>
          <script src="../attendance-qr-slide.js"></script>
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives

::: {.learning-objectives style="font-size:38px !important;"}
By the end of this lecture, you will be able to:

- **Define** what a point estimator is and distinguish it from an interval estimator

- **Compute** the bias, variance, and mean square error (MSE) of a point estimator

- **Identify** common unbiased estimators for means, proportions, and differences

- **Apply** the 2-standard-error bound to quantify the precision of any point estimate

- **Interpret** standard errors in financial and economic contexts, including portfolio and market-share estimation
:::

---

## 📱 Attendance Check-in {.center background-color="#f0f4ff"}

<div class="attendance-qr" data-status="present" id="qr-present" style="min-height: 400px; display: flex; align-items: center; justify-content: center;"></div>

## 📋 Overview

::: {style="font-size:50px !important;"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Introduction to Estimation** – Point vs. interval estimates; the role of target parameters

- **Bias & Unbiasedness** – What it means for an estimator to be centered on the truth

- **Mean Square Error (MSE)** – Balancing bias and variance in estimator quality

- **Common Unbiased Estimators** – $\bar{Y}$, $\hat{p}$, $\bar{Y}_1 - \bar{Y}_2$, $\hat{p}_1 - \hat{p}_2$

- **Error of Estimation** – Bounding how far off our estimate could be

- **Case Study** – Estimating mean stock returns and proportions from market data
:::
:::
:::

---

## 📖 Why Estimation?

::: {.callout-note}
## 🎯 Motivation

In finance and economics, population parameters are **never directly observed** — we always work from samples.

::: {.columns}
::: {.column width="50%"}
::: {.fragment}
**Finance Applications:**

- Estimating mean daily return $\mu$ of a stock
- Estimating default probability $p$ of a loan portfolio
- Comparing mean returns $\mu_1 - \mu_2$ of two funds
- Estimating market share difference $p_1 - p_2$
:::
:::

::: {.column width="50%"}
::: {.fragment}
**Regulatory & Policy Applications:**

- Estimating mean broadband speed across ISPs
- Estimating fraud rate in financial transactions
- Estimating GDP growth from survey data
- Estimating unemployment from labor force samples
:::
:::
:::
:::

::: {.callout-important .fragment style="font-size:44px !important;margin-top:-10px"}
## Key Question 

Given a sample, what single number best represents an unknown population parameter?

:::

---

## 📖 Definition: Estimator vs. Estimate 

::: {.callout-note style="font-size:40px !important;"}
## 📝 Definition 8.1: Estimator

An **estimator** is a rule (often a formula) that tells how to calculate the value of an estimate based on the measurements in a sample.

$$\hat{\theta} = g(Y_1, Y_2, \ldots, Y_n)$$

**Interpretation:** The estimator $\hat{\theta}$ (read "theta-hat") is a *random variable* — its value varies from sample to sample.
:::

::: {.fragment style="font-size:40px !important;margin-top:-10px"}
::: {.callout-tip}
## 💡 Point vs. Interval Estimate

| Type | Description | Example |
|------|-------------|---------|
| **Point Estimate** | Single number | $\hat{\mu} = 0.13\%$ daily return |
| **Interval Estimate** | Range of values | $\mu \in (0.07\%, 0.19\%)$ |

A **point estimator** is the *formula*; a **point estimate** is the *resulting number* from your data.
:::
:::

---

## 📖 The Bull's-Eye Analogy {.smaller}

::: {style="font-size:32px;margin-top:-10px;"}
Point estimation is like **firing at a target**:

::: {.columns}
::: {.column width="50%"}
::: {.fragment}
**The Components:**

| Analogy | Statistics |
|---------|-----------|
| Revolver | Estimator $\hat{\theta}$ |
| Single shot | One estimate from one sample |
| Bull's-eye | True parameter $\theta$ |
| Shot pattern | Sampling distribution of $\hat{\theta}$ |

:::
:::

::: {.column width="50%"}
::: {.fragment}
**The Key Insight:**

One lucky shot tells us nothing about a marksman's skill.

Similarly, **one estimate** tells us nothing about an estimator's quality.

We evaluate estimators by their **long-run behavior** — the shape and center of their sampling distribution.
:::
:::
:::

::: {.fragment}
> 💼 *A fund manager who beats the market once may be lucky. One who beats it consistently over 100 quarters is likely skilled.*
:::
:::

---

## 📖 Definition: Bias {.smaller}

::: {style="font-size:33px;"}
::: {.columns}
::: {.column width="50%" .callout-note}
## 📝 Definition 8.2: Unbiased Estimator

Let $\hat{\theta}$ be a point estimator for a parameter $\theta$. Then $\hat{\theta}$ is **unbiased** if:

$$E(\hat{\theta}) = \theta$$

If $E(\hat{\theta}) \neq \theta$, then $\hat{\theta}$ is **biased**.
:::

::: {.column width="50%" .callout-note}
## 📝 Definition 8.3: Bias

The **bias** of a point estimator $\hat{\theta}$ is:

$$B(\hat{\theta}) = E(\hat{\theta}) - \theta$$
:::
:::

::: {.fragment}
**Finance Interpretation:**

- An **unbiased** return estimator: on average, we neither systematically over- nor under-estimate the true expected return

- A **biased** estimator: our estimates consistently overshoot or undershoot the truth — like a fund that consistently over-reports its Sharpe ratio
:::
:::

---

## 📖 Bias Illustrated {.smaller}

::: {style="font-size:38px"}
::: {.columns}
::: {.column width="50%"}
::: {.fragment}
**Unbiased estimator — centered on $\theta$:**

$$E(\hat{\theta}) = \theta$$

The sampling distribution is *centered* at the true value.

✅ Ideal starting property
:::
:::

::: {.column width="50%"}
::: {.fragment}
**Biased estimator — shifted from $\theta$:**

$$E(\hat{\theta}) > \theta \quad \text{(positive bias)}$$
$$E(\hat{\theta}) < \theta \quad \text{(negative bias)}$$

The sampling distribution is *offset* from the true value.

⚠️ Systematically misleading
:::
:::
:::

::: {.fragment}
::: {.callout-important}
## Two Unbiased Estimators — Which is Better?

Even among unbiased estimators, we prefer **smaller variance**. A tight cluster around $\theta$ beats a spread-out one, even if both are centered correctly.
:::
:::
:::

---

## 📖 Definition: Mean Square Error {.smaller}

::: {style="font-size:28px;"}
::: {.callout-note}
## 📝 Definition 8.4: Mean Square Error (MSE)

The **mean square error** of a point estimator $\hat{\theta}$ is:

$$\text{MSE}(\hat{\theta}) = E\left[(\hat{\theta} - \theta)^2\right]$$

**Key decomposition:**

$$\boxed{\text{MSE}(\hat{\theta}) = V(\hat{\theta}) + \left[B(\hat{\theta})\right]^2}$$
:::

::: {.fragment}
::: {.columns}
::: {.column width="55%"}
**Interpretation:**

- MSE captures **total error** = spread + systematic error
- If unbiased: $\text{MSE}(\hat{\theta}) = V(\hat{\theta})$
- A *biased* estimator with **very small variance** can have lower MSE than an unbiased one

:::
::: {.column width="45%"}
**Finance Analogy:**

Like **tracking error** in fund management:

- Bias = consistent over/underperformance vs. benchmark
- Variance = volatility of the tracking difference
- MSE = total risk-adjusted tracking error
:::
:::
:::
:::

---

## 📌 Example 1: Biased vs. Unbiased Variance Estimator {.smaller}

::: {style="font-size:50px; margin-top:-5px;"}

**Problem:** Given a random sample $Y_1, \ldots, Y_n$ with $E(Y_i) = \mu$ and $V(Y_i) = \sigma^2$, compare two estimators of $\sigma^2$:

$$S'^2 = \frac{1}{n}\sum_{i=1}^n (Y_i - \bar{Y})^2, \qquad S^2 = \frac{1}{n-1}\sum_{i=1}^n (Y_i - \bar{Y})^2$$

**Which estimator is unbiased? Derive $E(S'^2)$ and $E(S^2)$.**

:::

---

## 📌 Example 1: Solution {.smaller}

::: {style="font-size:30px; margin-top:-5px;"}

**Solution:** Use $E(Y_i^2) = \sigma^2 + \mu^2$ and $E(\bar{Y}^2) = \sigma^2/n + \mu^2$.

::: {.columns}
::: {.column width="50%" .fragment style="margin-top:-10px;"}

$$E\left[\sum_{i=1}^n (Y_i - \bar{Y})^2\right] = (n-1)\sigma^2$$

$$\Rightarrow E(S'^2) = \frac{n-1}{n}\sigma^2 \neq \sigma^2 \quad \text{Biased!}$$

:::

::: {.column width="50%" .fragment style="margin-top:-10px;"}

$$E(S^2) = \frac{1}{n-1} \cdot (n-1)\sigma^2 = \sigma^2 \quad \text{Unbiased ✅}$$

:::
:::

:::{.fragment}

> 💼 *This is why financial software divides by $n-1$, not $n$, when computing volatility from historical returns.*

:::

:::

---

## 📖 Common Unbiased Estimators {.smaller}

::: {style="font-size:30px"}

| Target Parameter $\theta$ | Sample Size | Estimator $\hat{\theta}$ | Standard Error $\sigma_{\hat{\theta}}$ |
|--------------------------|-------------|--------------------------|----------------------------------------|
| Population mean $\mu$ | $n$ | $\bar{Y}$ | $\sigma/\sqrt{n}$ |
| Binomial proportion $p$ | $n$ | $\hat{p} = Y/n$ | $\sqrt{pq/n}$ |
| Difference of means $\mu_1 - \mu_2$ | $n_1, n_2$ | $\bar{Y}_1 - \bar{Y}_2$ | $\sqrt{\sigma_1^2/n_1 + \sigma_2^2/n_2}$ |
| Difference of proportions $p_1 - p_2$ | $n_1, n_2$ | $\hat{p}_1 - \hat{p}_2$ | $\sqrt{p_1q_1/n_1 + p_2q_2/n_2}$ |

::: {.fragment}
::: {.callout-important}
## Key Facts

- All four estimators are **unbiased** under random sampling
- All four have **approximately normal** sampling distributions for large samples (CLT)
- Standard errors **decrease** as sample size increases — more data → more precision
:::
:::
:::

---

## 📖 Definition: Error of Estimation {.smaller}

::: {style="font-size:31px;"}
::: {.callout-note}
## 📝 Definition 8.5: Error of Estimation

The **error of estimation** $\varepsilon$ is the distance between an estimator and its target parameter:

$$\varepsilon = |\hat{\theta} - \theta|$$
:::

::: {.fragment}
**How large could the error be?**

Since $\hat{\theta}$ is approximately normally distributed (for large $n$):

$$P\left(|\hat{\theta} - \theta| < 2\sigma_{\hat{\theta}}\right) \approx 0.95$$

We use $b = 2\sigma_{\hat{\theta}}$ as a **2-standard-error bound** on the error of estimation.
:::

::: {.fragment}
::: {.callout-tip}
## 💡 Practical Rule

With ~95% confidence, the error of estimation is less than $2\sigma_{\hat{\theta}}$.

This bound gets **tighter** as $n$ increases, since $\sigma_{\hat{\theta}} \propto 1/\sqrt{n}$.
:::
:::
:::

---

## 📌 Example 2: Estimating a Default Rate {.smaller}

::: {style="font-size:32px;"}

**Problem:** A risk officer samples $n = 1000$ corporate loans. Of these, $y = 43$ defaulted within 12 months. Estimate the default probability $p$ and place a 2-standard-error bound on the error.

:::{.fragment}

**Solution:**

Point estimate: $\hat{p} = \dfrac{43}{1000} = 0.043$

:::

:::{.fragment}

Standard error bound (substituting $\hat{p}$ for $p$):

$$b = 2\sigma_{\hat{p}} = 2\sqrt{\frac{\hat{p}(1-\hat{p})}{n}} = 2\sqrt{\frac{(0.043)(0.957)}{1000}} \approx 2 \times 0.0203 = 0.0406$$

:::

:::{.fragment}

**Interpretation:**

We estimate the true default rate is **4.3%**, and we are approximately 95% confident the true value lies within **±0.41 percentage points** of this estimate — i.e., roughly in the interval **(3.9%, 4.7%)**.

:::
:::

---

## 📌 Example 3: Comparing Fund Returns {.smaller}

::: {style="font-size:30px;"}

**Problem:** Two equity funds are compared using independent samples. Fund A: $n_1 = 100$ monthly returns, $\bar{y}_1 = 1.2\%$, $s_1^2 = 4.0$. Fund B: $n_2 = 100$, $\bar{y}_2 = 0.9\%$, $s_2^2 = 5.5$.

Estimate $\mu_1 - \mu_2$ and place a 2-SE bound on the error.

:::{.fragment}

**Point Estimate:**

$$(\bar{y}_1 - \bar{y}_2) = 1.2\% - 0.9\% = 0.3\%$$

:::

:::{.fragment}

**Standard Error:**

$$\sigma_{(\bar{Y}_1 - \bar{Y}_2)} \approx \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}} = \sqrt{\frac{4.0}{100} + \frac{5.5}{100}} = \sqrt{0.095} \approx 0.308\%$$

:::

:::{.fragment}

**2-SE Bound:** $b = 2 \times 0.308\% \approx 0.62\%$

:::

:::{.fragment}
**Interpretation:** Fund A appears to outperform Fund B by **0.3% per month**, but the error bound of **±0.62%** means this difference is not yet statistically distinguishable from zero. *Larger samples needed!*
:::
:::

---

## 🎮 Interactive: MSE Decomposition {.smaller}

:::::: {style="font-size: 0.72em; margin-top: -8px;"}

**Explore how Bias and Variance each contribute to MSE, and why a biased estimator can beat an unbiased one.**

::::: columns
::: {.column width="35%"}

```{ojs}
//| echo: false

viewof bias_val = {
  const input = Inputs.range([-3, 3], {value: 0, step: 0.1, label: "Bias B(θ̂):"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

viewof var_val = {
  const input = Inputs.range([0.1, 5], {value: 1, step: 0.1, label: "Variance V(θ̂):"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

mse_val = var_val + bias_val * bias_val

color_mse = mse_val < 2
  ? html`<div style="background:#d4edda;padding:8px;border-radius:5px;">
      <strong>MSE = ${mse_val.toFixed(3)}</strong> — Low total error ✅
    </div>`
  : html`<div style="background:#f8d7da;padding:8px;border-radius:5px;">
      <strong>MSE = ${mse_val.toFixed(3)}</strong> — High total error ⚠️
    </div>`

md`
**Decomposition:**

- Variance = ${var_val.toFixed(3)}
- Bias² = ${(bias_val*bias_val).toFixed(3)}
- **MSE = ${mse_val.toFixed(3)}**

${color_mse}
`
```

:::

::: {.column width="65%"}

```{ojs}
//| echo: false

{
  const theta = 0;
  const mu = theta + bias_val;
  const sd = Math.sqrt(var_val);
  const xs = Array.from({length: 200}, (_, i) => mu - 4*sd + i * 8*sd/199);
  const ys = xs.map(x => Math.exp(-0.5*((x-mu)/sd)**2) / (sd * Math.sqrt(2*Math.PI)));

  return Plot.plot({
    width: 680,
    height: 360,
    marginLeft: 50,
    marginBottom: 45,
    x: {domain: [-6, 6], label: "Estimator value θ̂"},
    y: {label: "Density"},
    title: `Sampling Distribution: Bias=${bias_val.toFixed(1)}, Var=${var_val.toFixed(1)}, MSE=${mse_val.toFixed(2)}`,
    marks: [
      Plot.line(xs.map((x,i) => ({x, y: ys[i]})), {x:"x", y:"y", stroke:"steelblue", strokeWidth:2.5}),
      Plot.ruleX([theta], {stroke:"red", strokeWidth:3, strokeDasharray:"8,4"}),
      Plot.ruleX([mu], {stroke:"orange", strokeWidth:2}),
      Plot.text([{x: theta, y: Math.max(...ys)*0.05}], {x:"x", y:"y", text: d => "θ (truth)", fill:"red", fontSize:13}),
      Plot.text([{x: mu, y: Math.max(...ys)*0.1}], {x:"x", y:"y", text: d => "E(θ̂)", fill:"orange", fontSize:13}),
      Plot.ruleY([0])
    ]
  });
}
```

:::
:::::
::::::

---

## 🤝 Think-Pair-Share {.smaller}

::: {style="font-size:46px;"}
::: {.callout-note}
## 💬 Discussion (4 minutes)

A central bank samples **200 commercial banks** and finds that 36 reported negative equity in the last quarter. A colleague proposes two estimators for the true failure rate $p$:

$$\hat{p}_1 = \frac{Y}{n} = \frac{36}{200} = 0.18 \qquad \hat{p}_2 = \frac{Y + 2}{n + 4} = \frac{38}{204} \approx 0.186$$

**Discuss with your partner:**

1. Is $\hat{p}_1$ unbiased? What is its standard error?
2. $\hat{p}_2$ is known to be **biased** — can you show why? What is its bias?
3. Under what conditions might $\hat{p}_2$ actually have **lower MSE** than $\hat{p}_1$?
4. Which would you report in a regulatory context, and why?
:::
:::

```{r}
#| echo: false
#| eval: true
library(countdown)
countdown(minutes = 4, bottom = 0, right = 0,
          color_running_background = "#2c7bb6",
          color_finished_background = "#d73027")
```

---

## 💰 Case Study: Point Estimates {.smaller}

```{r}
#| echo: true
#| code-fold: true
#| message: false
#| warning: false
#| eval: true

library(tidyverse)
library(tidyquant)
library(knitr)

symbols <- c("AAPL", "JPM", "XOM")
prices <- tq_get(symbols, from = "2021-01-01", to = "2023-12-31")

returns <- prices %>%
  group_by(symbol) %>%
  tq_transmute(select = adjusted, mutate_fun = periodReturn,
               period = "daily", col_rename = "r")

stats <- returns %>%
  group_by(symbol) %>%
  summarise(n = n(), mu_hat = mean(r) * 100,
            se = sd(r) * 100 / sqrt(n()),
            bound_2se = 2 * sd(r) * 100 / sqrt(n()))

kable(stats, digits = 4,
      col.names = c("Stock", "n", "μ̂ (%)", "SE (%)", "2-SE Bound (%)"),
      caption = "Point Estimates of Mean Daily Return")
```

---

## 💰 Case Study: Return Distributions {.smaller}

```{r}
#| echo: true
#| code-fold: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 10
#| fig-height: 4

returns %>%
  mutate(r_pct = r * 100) %>%
  ggplot(aes(x = r_pct, fill = symbol)) +
  geom_histogram(bins = 60, alpha = 0.6, position = "identity") +
  geom_vline(data = stats,
             aes(xintercept = mu_hat, color = symbol),
             linewidth = 1.2, linetype = "dashed") +
  facet_wrap(~symbol, ncol = 3, scales = "free_y") +
  labs(title = "Daily Return Distributions with Point Estimates",
       x = "Daily Return (%)", y = "Count") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

---

## 💰 Case Study: Key Findings {.smaller}

::: {style="font-size:46px;"}
::: {.callout-important}
## 📊 Interpreting the Point Estimates

::: {.columns}

::: {.column width="33%"}
::: {.fragment}
**Point Estimates $\hat{\mu}$:**

- Apple (AAPL): ~0.07% daily
- JP Morgan (JPM): ~0.06% daily
- Exxon (XOM): ~0.15% daily

All close to zero — consistent with efficient market hypothesis for daily returns
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Standard Errors SE:**

- SE ≈ 0.056–0.069% for all stocks
- 2-SE bounds ≈ 0.11–0.14%
- With $n \approx 750$ observations, SE is small relative to daily volatility (~2%)
- Larger $n$ tightens the SE further

:::
:::

::: {.column width="33%"}
::: {.fragment}
**Takeaways for Estimation:**

1. **Unbiasedness**: $\bar{Y}$ centers on $\mu$
2. **Precision grows** with sample size
3. **2-SE bound** gives a practical confidence region
4. Bias-variance tradeoff matters when data is limited
:::
:::

:::
:::
:::

---

## 📝 Quiz #1: Unbiased Estimator {.smaller .quiz-question}

::: {style="font-size:50px; margin-top:-5px;"}

Given $Y_1, Y_2, Y_3$ i.i.d. with mean $\mu$, which of the following is **NOT** an unbiased estimator of $\mu$?

- [$\frac{Y_1 + Y_2 + Y_3}{3}$]{.correct data-explanation="✅ Wait — this IS unbiased! Its expected value is μ. The question asks which is NOT unbiased. Select again."}
- $\frac{Y_1 + Y_2}{2}$
- [$\frac{2Y_1 + Y_3}{2}$]{.correct data-explanation="✅ Correct! E[(2Y₁ + Y₃)/2] = (2μ + μ)/2 = 3μ/2 ≠ μ — this is biased unless μ = 0."}
- $Y_2$

:::

---

## 📝 Quiz #2: MSE Decomposition {.smaller .quiz-question}

::: {style="font-size:50px; margin-top:-5px;"}

If $\hat{\theta}$ is an unbiased estimator of $\theta$ with variance $V(\hat{\theta}) = 4$, what is $\text{MSE}(\hat{\theta})$?

- [4]{.correct data-explanation="✅ Correct! Since B(θ̂) = 0 for an unbiased estimator, MSE(θ̂) = V(θ̂) + 0² = 4."}
- 2
- 0
- 16

:::

---

## 📝 Quiz #3: Standard Error Bound {.smaller .quiz-question}

::: {style="font-size:50px; margin-top:-5px;"}

A bank estimates the proportion of clients with a credit score above 700. From $n = 400$ clients, 280 qualify. The 2-standard-error bound on the error of estimation is approximately:

- [0.046]{.correct data-explanation="✅ Correct! p̂ = 280/400 = 0.70, SE = √(0.70 × 0.30/400) = 0.0229, bound = 2 × 0.0229 ≈ 0.046."}
- 0.023
- 0.092
- 0.014

:::

---

## 📝 Quiz #4: Comparing Two Estimators {.smaller .quiz-question}

::: {style="font-size:50px; margin-top:-5px;"}

Estimator $\hat{\theta}_1$ is unbiased with $V(\hat{\theta}_1) = 9$. Estimator $\hat{\theta}_2$ has $B(\hat{\theta}_2) = 2$ and $V(\hat{\theta}_2) = 3$. Which has lower MSE?

- [$\hat{\theta}_2$, since $\text{MSE}(\hat{\theta}_2) = 3 + 4 = 7 < 9$]{.correct data-explanation="✅ Correct! MSE(θ̂₁) = 9 + 0 = 9, MSE(θ̂₂) = 3 + 2² = 7. Despite the bias, θ̂₂ has lower total error."}
- $\hat{\theta}_1$, because unbiased estimators always have lower MSE
- They are equal
- Cannot be determined without knowing $\theta$

:::

---

## 📝 Summary {.smaller}

::: {style="font-size:36px"}
::: {.summary-box}
**✅ Key Takeaways**

- **Estimators vs. Estimates**: An estimator is a random variable (formula); an estimate is its realized value from data

- **Unbiasedness**: $E(\hat{\theta}) = \theta$ — the estimator is centered on the true parameter; $\bar{Y}$ and $\hat{p}$ are unbiased

- **MSE = Variance + Bias²**: Total error accounts for both spread and systematic error; a biased estimator can win if its variance is much smaller

- **Common Estimators**: $\bar{Y}$, $\hat{p}$, $\bar{Y}_1 - \bar{Y}_2$, $\hat{p}_1 - \hat{p}_2$ are all unbiased and approximately normal for large $n$

- **Error Bound**: With ~95% confidence, error of estimation $< 2\sigma_{\hat{\theta}}$; this shrinks as $n$ grows
:::
:::

---

## 📚 Practice Problems

::: {style="font-size:46px;"}
::: {.callout-tip}
## 📝 Homework Problems

**Problem 1 (Unbiasedness):** Let $Y_1, Y_2, \ldots, Y_n$ be i.i.d. from a distribution with mean $\mu$ and variance $\sigma^2$. Show that $S^2 = \frac{1}{n-1}\sum(Y_i - \bar{Y})^2$ is an unbiased estimator for $\sigma^2$. (Wackerly Ex. 8.1)

**Problem 2 (Bias & MSE):** Suppose $E(\hat{\theta}) = 2\theta + 1$. Find $B(\hat{\theta})$ and construct an unbiased estimator $\hat{\theta}^*$ from $\hat{\theta}$. (Wackerly Ex. 8.3)

**Problem 3 (Comparing Estimators):** $\hat{p}_1 = Y/n$ and $\hat{p}_2 = (Y+1)/(n+2)$. Derive $B(\hat{p}_2)$, $\text{MSE}(\hat{p}_1)$, and $\text{MSE}(\hat{p}_2)$. For which values of $p$ does $\hat{p}_2$ dominate? (Wackerly Ex. 8.17)

**Problem 4 (Financial Application):** An analyst observes $n = 252$ daily returns of a hedge fund with $\bar{y} = 0.06\%$ and $s = 1.2\%$. (a) Give a point estimate of the true mean daily return. (b) Place a 2-SE bound on the error. (c) Annualize your estimate (×252) and interpret.
:::
:::

---

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business

ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Interval Estimation (Confidence Intervals)

**Reading:** Chapter 8, Sections 8.5–8.9

**Preparation:** Review the standard normal table and $t$-distribution

**⏰ Reminders:**

✅ Complete Practice Problems 1–4

✅ Review the CLT and standard normal distribution

✅ Think about: what does "95% confident" actually mean?

✅ Work hard!
:::
:::

---

## ❓ Questions? {.center}

::: {.callout-note}
## 💬 Open Discussion

**Key Topics for Discussion:**

- Why do we divide by $n-1$ instead of $n$ in the sample variance?

- Can a biased estimator ever be *preferable* to an unbiased one in practice?

- How would you explain the concept of "standard error" to a non-statistician client?

- If you double your sample size, how much does your estimation error shrink?
:::

<div class="attendance-qr" data-status="late" id="qr-late" style="min-height: 400px; display: flex; align-items: center; justify-content: center;"></div>
