---
title: "Mathematical Statistics"
subtitle: "Topic 7: Properties of Point Estimators & Methods of Estimation"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ../ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    margin: 0.04
    # footer: "Mathematical Statistics ŌĆō Properties of Point Estimators"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      shuffleOptions: true
      defaultCorrect: "Ō£ģ Correct! Well done."
      defaultIncorrect: "ŌØī Not quite. Try again."
      includeScore: true
    include-after-body:
      - text: |
          <script src="../qrcode.min.js"></script>
          <script src="../attendance-qr-slide.js"></script>
revealjs-plugins:
  - quiz
---

## ­¤Ä» Learning Objectives

:::: {style="font-size:40px"}
By the end of this lecture, you will be able to:

1. **Compare** two unbiased estimators using relative efficiency and the Cram├®rŌĆōRao lower bound

2. **Verify** whether an estimator is consistent by applying Theorem 9.1 and the Law of Large Numbers

3. **Identify** sufficient statistics using the Factorization Criterion (Theorem 9.4)

4. **Apply** the RaoŌĆōBlackwell theorem to improve estimators and find MVUEs

5. **Derive** point estimators using the Method of Moments and the Method of Maximum Likelihood
::::

---

## ­¤ō▒ Attendance Check-in {.center background-color="#f0f4ff"}

<div class="attendance-qr" data-status="present" id="qr-present" style="min-height: 400px; display: flex; align-items: center; justify-content: center;"></div>

## ­¤ōŗ Overview

:::: {style="font-size:40px"}
**Chapter 9: Properties of Point Estimators and Methods of Estimation**

::: {.columns}
::: {.column width="50%"}
**Part I ŌĆö Evaluating Estimators**

- 9.2 Relative Efficiency
- 9.3 Consistency
- 9.4 Sufficiency & Factorization
- 9.5 RaoŌĆōBlackwell & MVUE
:::
::: {.column width="50%"}
**Part II ŌĆö Finding Estimators**

- 9.6 Method of Moments (MoM)
- 9.7 Method of Maximum Likelihood (MLE)
- 9.8 Large-Sample Properties of MLEs *(optional)*
:::
:::

> **Big Idea:** How do we know if an estimator is *good*, and how do we find the *best possible* one?
::::

---

## ­¤ÆĪ Motivation: Why Does This Matter?

:::: {style="font-size:37px"}
::: {.callout-note}
## The Portfolio Manager's Dilemma

A risk analyst at a Baku investment fund estimates the **expected monthly return** of an equity portfolio. She has collected $n$ months of data.

- **Estimator A:** Uses the full sample mean $\bar{Y}$
- **Estimator B:** Uses only the first two observations $\frac{Y_1 + Y_2}{2}$

Both are *unbiased* ŌĆö but which is *better*?
:::

::: {.fragment}
**The answer requires formal criteria:**

- ­¤Ä» **Efficiency** ŌĆö which has smaller variance?
- ­¤ōē **Consistency** ŌĆö does it converge to the truth as $n \to \infty$?
- ­¤öÆ **Sufficiency** ŌĆö does it retain all information about the parameter?
:::
::::

---

## Part I: Evaluating Estimators {.center}

---

## ­¤ō¢ 9.2 Relative Efficiency

:::: {style="font-size:48px"}
::: {.callout-note}
## ­¤ōØ Definition 9.1 ŌĆö Relative Efficiency

Given **two unbiased estimators** $\hat{\theta}_1$ and $\hat{\theta}_2$ of $\theta$, the efficiency of $\hat{\theta}_1$ **relative to** $\hat{\theta}_2$ is:

$$\text{eff}(\hat{\theta}_1, \hat{\theta}_2) = \frac{V(\hat{\theta}_2)}{V(\hat{\theta}_1)}$$

- If $\text{eff} > 1$ $\Rightarrow$ $\hat{\theta}_1$ is **more efficient** (smaller variance)
- If $\text{eff} < 1$ $\Rightarrow$ $\hat{\theta}_2$ is **more efficient** (preferred)
:::

::: {.fragment}
**Interpretation:** $\text{eff}(\hat{\theta}_1, \hat{\theta}_2) = 1.8$ means $V(\hat{\theta}_2) = 1.8 \cdot V(\hat{\theta}_1)$ ŌĆö you need 80% more data with $\hat\theta_2$ to match $\hat\theta_1$'s precision.
:::
::::

---

## ­¤ōī Example: Mean vs. Median ŌĆö Which Estimator for Stock Returns?

:::: {style="font-size:35px"}
::: {.callout-tip}
## Setting

Daily S&P 500 returns are **normally distributed** with mean $\mu$ and variance $\sigma^2$.

We compare two unbiased estimators of $\mu$:

- $\hat{\mu}_1$ = **Sample median** with $V(\hat{\mu}_1) \approx \frac{(1.2533)^2 \sigma^2}{n}$
- $\hat{\mu}_2$ = **Sample mean** with $V(\hat{\mu}_2) = \frac{\sigma^2}{n}$
:::

::: {.fragment}
$$\text{eff}(\hat{\mu}_1, \hat{\mu}_2) = \frac{V(\hat{\mu}_2)}{V(\hat{\mu}_1)} = \frac{\sigma^2/n}{(1.2533)^2\sigma^2/n} = \frac{1}{1.5708} \approx 0.637$$

**Conclusion:** The sample mean has **~64% of the variance** of the sample median. Ō£ģ Prefer $\bar{Y}$ for estimating expected returns on normally distributed data.
:::
::::

---

## ­¤ō¢ Cram├®rŌĆōRao Lower Bound *(Optional ŌĆö Exercise 9.8)*

:::: {style="font-size:36px"}
::: {.callout-important}
## ­¤¦« Cram├®rŌĆōRao Inequality

For any **unbiased** estimator $\hat\theta$ of $\theta$, under regularity conditions:

$$V(\hat\theta) \geq I(\theta) = \left[ nE\left(-\frac{\partial^2 \ln f(Y|\theta)}{\partial \theta^2}\right) \right]^{-1}$$

An estimator achieving this lower bound is called **efficient**.
:::

::: {.fragment}
**Finance application:** For normally distributed returns, $\bar{Y}$ achieves the Cram├®rŌĆōRao bound ŌĆö it is the **most efficient** unbiased estimator of $\mu$.

> No other unbiased estimator can have *smaller* variance than the sample mean for normal data.
:::
::::

---

## ­¤ō¢ 9.3 Consistency

:::: {style="font-size:50px"}
::: {.callout-note}
## ­¤ōØ Definition 9.2 ŌĆö Consistent Estimator

$\hat\theta_n$ is a **consistent estimator** of $\theta$ if, for any $\varepsilon > 0$:

$$\lim_{n \to \infty} P(|\hat\theta_n - \theta| \leq \varepsilon) = 1$$

Equivalently: $\hat\theta_n$ **converges in probability** to $\theta$.
:::

::: {.fragment}
**Intuition for finance students:** As a portfolio grows in history (more months of returns), your estimator for expected return should get arbitrarily close to the true value ŌĆö not drift away.
:::
::::

---

## ­¤¦« Theorem 9.1 ŌĆö Easy Consistency Check

:::: {style="font-size:41px"}
::: {.callout-important}
## Theorem 9.1

An **unbiased** estimator $\hat\theta_n$ for $\theta$ is **consistent** if:

$$\lim_{n \to \infty} V(\hat\theta_n) = 0$$
:::

::: {.fragment}
**Proof sketch (via Tchebysheff):**

$$0 \leq P(|\hat\theta_n - \theta| > \varepsilon) \leq \frac{V(\hat\theta_n)}{\varepsilon^2} \xrightarrow{n\to\infty} 0 \qquad \square$$
:::

::: {.fragment}
**Example:** $\bar{Y}_n$ is unbiased for $\mu$ and $V(\bar{Y}_n) = \sigma^2/n \to 0$. So $\bar{Y}$ is **consistent** ŌĆö this is the **Law of Large Numbers**.
:::
::::

---

## ­¤¦« Theorem 9.2 ŌĆö Algebra of Consistent Estimators

:::: {style="font-size:40px"}
::: {.callout-important}
## Theorem 9.2

If $\hat\theta_n \xrightarrow{p} \theta$ and $\hat\theta'_n \xrightarrow{p} \theta'$, then:

(a) $\hat\theta_n + \hat\theta'_n \xrightarrow{p} \theta + \theta'$

(b) $\hat\theta_n \times \hat\theta'_n \xrightarrow{p} \theta \times \theta'$

(c) $\hat\theta_n / \hat\theta'_n \xrightarrow{p} \theta / \theta'$ \hspace{1em} (if $\theta' \neq 0$)

(d) If $g(\cdot)$ is continuous at $\theta$: $g(\hat\theta_n) \xrightarrow{p} g(\theta)$
:::

::: {.fragment}
**Finance application:** Since $\bar{Y} \xrightarrow{p} \mu$ and $S^2 \xrightarrow{p} \sigma^2$, by part (b): $S \xrightarrow{p} \sigma$ ŌĆö justifying use of $S$ in large-sample CIs for volatility estimation.
:::
::::

---

## ­¤ōī Example: Sample Variance is Consistent

:::: {style="font-size:30px"}
::: {.callout-tip}
## Setting (Example 9.3)

For a random sample with $E(Y_i) = \mu$, $V(Y_i) = \sigma^2$, show that $S^2_n$ is **consistent** for $\sigma^2$.
:::

::: {.fragment}
**Key steps:**

$$S^2_n = \frac{n}{n-1}\left[\underbrace{\frac{1}{n}\sum Y_i^2}_{\xrightarrow{p}\; \mu'_2} - \underbrace{\bar{Y}^2_n}_{\xrightarrow{p}\; \mu^2}\right]$$

- By LLN: $\frac{1}{n}\sum Y_i^2 \xrightarrow{p} E(Y^2) = \mu'_2$
- By Thm 9.2(d): $\bar{Y}^2_n \xrightarrow{p} \mu^2$
- Since $\frac{n}{n-1} \to 1$: $S^2_n \xrightarrow{p} \mu'_2 - \mu^2 = \sigma^2$ Ō£ģ
:::
::::

---

## ­¤Ä« Interactive: Consistency ŌĆö Estimator Convergence Explorer {.smaller}

:::: {style="font-size: 0.72em; margin-top: -8px;"}

**Watch how $\bar{Y}_n$ converges to $\mu$ as the hedge fund's return history grows.**

::::: columns
::: {.column width="32%"}

```{ojs}
//| echo: false

viewof true_mu = {
  const input = Inputs.range([-2, 5], {value: 2, step: 0.5, label: "True ╬╝ (%):"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

viewof sigma_val = {
  const input = Inputs.range([1, 8], {value: 3, step: 0.5, label: "Std Dev Žā (%):"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

viewof n_paths = {
  const input = Inputs.range([1, 30], {value: 10, step: 1, label: "# Sequences:"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

consistencyInfo = {
  const finalVals = [];
  for (let p = 0; p < n_paths; p++) {
    let sum = 0;
    for (let i = 1; i <= 200; i++) {
      const u1 = Math.random(), u2 = Math.random();
      const z = Math.sqrt(-2*Math.log(u1)) * Math.cos(2*Math.PI*u2);
      sum += true_mu + sigma_val * z;
    }
    finalVals.push(sum / 200);
  }
  const avg = finalVals.reduce((a,b)=>a+b,0)/finalVals.length;
  const spread = Math.max(...finalVals) - Math.min(...finalVals);
  
  return html`<div style="background:#e8f4f8; padding:8px; border-radius:6px; font-size:0.9em;">
    <strong>At n = 200:</strong><br>
    Avg estimate: ${avg.toFixed(3)}%<br>
    True ╬╝: ${true_mu}%<br>
    Range of paths: ${spread.toFixed(3)}%
  </div>`;
}
```

:::
::: {.column width="68%"}

```{ojs}
//| echo: false

consistencyData = {
  const N = 200;
  const paths = [];
  for (let p = 0; p < n_paths; p++) {
    let cumSum = 0;
    for (let i = 1; i <= N; i++) {
      const u1 = Math.random(), u2 = Math.random();
      const z = Math.sqrt(-2*Math.log(u1)) * Math.cos(2*Math.PI*u2);
      cumSum += true_mu + sigma_val * z;
      if (i % 2 === 0 || i <= 10) {
        paths.push({n: i, ybar: cumSum / i, path: p});
      }
    }
  }
  return paths;
}

Plot.plot({
  width: 720, height: 400,
  marginLeft: 55, marginBottom: 45, marginRight: 20,
  x: {label: "Sample size n (months of return history)", domain: [1, 200]},
  y: {label: "Sample mean return (%)", 
      domain: [true_mu - sigma_val*2, true_mu + sigma_val*2]},
  title: `${n_paths} sample paths converging to true ╬╝ = ${true_mu}%`,
  marks: [
    Plot.line(consistencyData, {x: "n", y: "ybar", stroke: "path",
      strokeOpacity: 0.5, strokeWidth: 1.2}),
    Plot.ruleY([true_mu], {stroke: "red", strokeWidth: 3, strokeDasharray: "8,4"}),
    Plot.text([{n: 190, y: true_mu + 0.3}],
      {x: "n", y: "y", text: d => `╬╝ = ${true_mu}%`, fill: "red", fontSize: 13})
  ]
})
```

:::
:::::
::::

---

## ­¤ō¢ 9.4 Sufficiency

:::: {style="font-size:43px"}
::: {.callout-note}
## ­¤ōØ Definition 9.3 ŌĆö Sufficient Statistic

The statistic $U = g(Y_1, \ldots, Y_n)$ is **sufficient** for $\theta$ if the conditional distribution of $Y_1, \ldots, Y_n$ given $U$ **does not depend on** $\theta$.

> *Knowing $U$ extracts all information about $\theta$ from the data.*
:::

::: {.fragment}
**Intuition ŌĆö Loan Default Monitoring:**

Suppose we monitor $n = 100$ loans (success = default, probability = $p$).

After observing the total number of defaults $Y = \sum X_i$, no other function of the individual outcomes tells us anything **new** about $p$.

$\Rightarrow$ $Y = \sum X_i$ is **sufficient** for $p$.
:::
::::

---

## ­¤ō¢ Definition 9.4 ŌĆö The Likelihood Function

:::: {style="font-size:40px"}
::: {.callout-note}
## ­¤ōØ Definition 9.4 ŌĆö Likelihood

The **likelihood** of sample $(y_1, \ldots, y_n)$ is the joint probability (discrete) or joint density (continuous):

$$L(\theta) = L(y_1, \ldots, y_n \mid \theta) = \prod_{i=1}^n f(y_i \mid \theta)$$
:::

::: {.fragment}
**Key insight:** $L(\theta)$ measures how *plausible* the observed data is for different values of $\theta$.

- For the bond default model: $L(p) = p^y(1-p)^{n-y}$
- For normally distributed returns: $L(\mu, \sigma^2) = \left(\frac{1}{\sigma\sqrt{2\pi}}\right)^n \exp\!\left(-\frac{\sum(y_i-\mu)^2}{2\sigma^2}\right)$
:::
::::

---

## ­¤¦« Theorem 9.4 ŌĆö Factorization Criterion

:::: {style="font-size:50px"}
::: {.callout-important}
## Theorem 9.4 (Factorization Criterion)

$U = g(Y_1,\ldots,Y_n)$ is **sufficient** for $\theta$ **if and only if** the likelihood factors as:

$$L(y_1, \ldots, y_n \mid \theta) = \underbrace{g(u, \theta)}_{\text{depends on } u \text{ and } \theta} \times \underbrace{h(y_1, \ldots, y_n)}_{\text{free of } \theta}$$
:::

::: {.fragment}
**Power of the criterion:** Factor the likelihood, identify the part depending on $\theta$, read off the sufficient statistic $U$.
:::
::::

---

## ­¤ōī Example: Sufficient Statistic for Inter-Trade Waiting Times (Example 9.5)

:::: {style="font-size:32px"}
::: {.callout-tip}
## Setting

Times between large equity trades follow an **exponential** distribution with mean $\theta$:

$$f(y_i \mid \theta) = \frac{1}{\theta}e^{-y_i/\theta}, \quad y_i > 0$$
:::

::: {.fragment}
**Apply Factorization:**

$$L(\theta) = \prod_{i=1}^n \frac{1}{\theta}e^{-y_i/\theta} = \frac{e^{-n\bar{y}/\theta}}{\theta^n} = \underbrace{\frac{e^{-n\bar{y}/\theta}}{\theta^n}}_{g(\bar{y},\;\theta)} \times \underbrace{1}_{h(\mathbf{y})}$$

$\Rightarrow$ $\bar{Y}$ (equivalently $\sum Y_i$) is **sufficient** for $\theta$. Ō£ģ

The sample mean retains *all* information about mean inter-trade time.
:::
::::

---

## Part II: Finding the Best Estimator {.center}

---

## ­¤¦« 9.5 The RaoŌĆōBlackwell Theorem

:::: {style="font-size:44px"}
::: {.callout-important}
## Theorem 9.5 ŌĆö RaoŌĆōBlackwell

Let $\hat\theta$ be **unbiased** for $\theta$ with $V(\hat\theta) < \infty$, and let $U$ be a **sufficient statistic** for $\theta$.

Define $\hat\theta^* = E(\hat\theta \mid U)$. Then:

$$E(\hat\theta^*) = \theta \quad \text{and} \quad V(\hat\theta^*) \leq V(\hat\theta)$$
:::

::: {.fragment}
**Translation:** Conditioning any unbiased estimator on a sufficient statistic *never increases* variance ŌĆö and usually *decreases* it.

> Ō×Ī’ĖÅ The best unbiased estimators are functions of sufficient statistics.
:::
::::

---

## ­¤ō¢ Minimum-Variance Unbiased Estimator (MVUE)

:::: {style="font-size:38px"}
::: {.callout-note}
## MVUE Definition

An estimator $\hat\theta^*$ is a **Minimum-Variance Unbiased Estimator (MVUE)** if:

1. $E(\hat\theta^*) = \theta$ (unbiased)
2. $V(\hat\theta^*) \leq V(\hat\theta)$ for **all** other unbiased estimators $\hat\theta$
:::

::: {.fragment}
**Practical recipe for finding MVUEs:**

1. Use **Factorization Criterion** to find the minimal sufficient statistic $U$

2. Find a function $h(U)$ such that $E[h(U)] = \theta$

3. $h(U)$ is the **MVUE** for $\theta$
:::
::::

---

## ­¤ōī Example: MVUE for Bernoulli Default Rate (Example 9.6)

:::: {style="font-size:48px"}
::: {.callout-tip}
## Setting

We observe $n$ independent loans: $Y_i = 1$ (default) with probability $p$.
:::

**Step 1 ŌĆö Find sufficient statistic:**

$$L(p) = p^{\sum y_i}(1-p)^{n - \sum y_i} = \underbrace{p^u(1-p)^{n-u}}_{g(u,p)} \times \underbrace{1}_{h}$$

$\Rightarrow$ $U = \sum_{i=1}^n Y_i$ is sufficient for $p$.
::::

---

## ­¤ōī Example 9.6: Bernoulli MVUE ŌĆö Solution

:::: {style="font-size:50px"}
**Step 2 ŌĆö Find unbiased function of $U$:**

$$E(U) = np \quad \Rightarrow \quad E\!\left(\frac{U}{n}\right) = p$$

$\Rightarrow$ $\hat{p} = \bar{Y} = U/n$ is the **MVUE** for $p$. Ō£ģ

> No other unbiased estimator of the default rate can have smaller variance than $\bar{Y}$.
::::

---

## ­¤ōī Example: MVUE for Normal Return Distribution (Example 9.8)

:::: {style="font-size:31px"}
::: {.callout-tip}
## Setting

Monthly fund returns $Y_i \sim N(\mu, \sigma^2)$ (both unknown). Find MVUEs for $\mu$ and $\sigma^2$.
:::

::: {.fragment}
**Factorize the likelihood:**

$$L(\mu, \sigma^2) \propto \exp\!\left(\!-\!\frac{1}{2\sigma^2}\sum y_i^2 + \frac{\mu}{\sigma^2}\sum y_i - \frac{n\mu^2}{2\sigma^2}\right)$$

$\Rightarrow$ $\left(\sum Y_i,\; \sum Y_i^2\right)$ are **jointly sufficient** for $(\mu, \sigma^2)$.
:::

::: {.fragment}
**MVUEs:**

- $\hat\mu = \bar{Y}$ ŌĆö unbiased for $\mu$, function of $\sum Y_i$ Ō£ģ
- $\hat\sigma^2 = S^2 = \frac{1}{n-1}\sum(Y_i - \bar{Y})^2$ ŌĆö unbiased for $\sigma^2$ Ō£ģ
:::
::::

---

## ­¤żØ Think-Pair-Share {.smaller}

:::: {style="font-size:47px"}
::: {.callout-note}
## ­¤Æ¼ Problem

An analyst tracks **daily trading volume** (in millions USD) for a stock. Volumes follow an **exponential distribution** with unknown mean $\theta$ (average daily volume).

A random sample of $n = 20$ days gives $\sum y_i = 480$ million USD.

**Questions:**

1. Using the **Method of Moments**, derive an estimator for $\theta$.

2. Using the **Factorization Criterion**, show that $\bar{Y}$ is sufficient for $\theta$.

3. Is $\bar{Y}$ the MVUE for $\theta$? Justify your answer.

4. Compute the estimate of $\theta$ from the data above.
:::
::::

```{r}
#| echo: false
library(countdown)
countdown(minutes = 5, bottom = 0, right = 0,
          color_running_background = "#2C7BB6",
          color_finished_background = "#D7191C")
```

---

## Ō£ģ Think-Pair-Share ŌĆö Solution Part 1 {.smaller}

:::: {style="font-size:31px"}
::: {.callout-important}
## ­¤ōŖ Full Solution
:::

**Q1 ŌĆö Method of Moments:**

For Exponential($\theta$): $\mu'_1 = E(Y) = \theta$

Set population moment = sample moment:

$$\theta = m'_1 = \bar{Y} \quad \Rightarrow \quad \hat\theta_{\text{MoM}} = \bar{Y}$$

**Q2 ŌĆö Factorization:**

$$L(\theta) = \prod \frac{1}{\theta}e^{-y_i/\theta} = \frac{1}{\theta^n}e^{-n\bar{y}/\theta}$$

$= \underbrace{\theta^{-n}e^{-n\bar{y}/\theta}}_{g(\bar{y},\,\theta)} \times \underbrace{1}_{h(\mathbf{y})}$

$\Rightarrow$ $\bar{Y}$ is **sufficient** for $\theta$. Ō£ģ
::::

---

## Ō£ģ Think-Pair-Share ŌĆö Solution Part 2 {.smaller}

:::: {style="font-size:32px"}
**Q3 ŌĆö Is $\bar{Y}$ the MVUE?**

- $\bar{Y}$ is **unbiased**: $E(\bar{Y}) = \theta$ Ō£ģ
- $V(\bar{Y}) = \theta^2/n \to 0$ as $n \to \infty$ Ō£ģ (consistent)
- $\bar{Y}$ is a function of the minimal sufficient statistic Ō£ģ
- By RaoŌĆōBlackwell, **$\bar{Y}$ is the MVUE** for $\theta$

No unbiased estimator of mean daily volume has smaller variance.

**Q4 ŌĆö Numerical estimate:**

$$\hat\theta = \bar{Y} = \frac{\sum y_i}{n} = \frac{480}{20} = \mathbf{24} \text{ million USD}$$

> The estimated mean daily trading volume is **\$24 million**.
::::

---

## ­¤ō¢ 9.6 Method of Moments (MoM)

:::: {style="font-size:35px"}
::: {.callout-note}
## Procedure

The **$k$-th population moment**: $\mu'_k = E(Y^k)$

The **$k$-th sample moment**: $m'_k = \frac{1}{n}\sum_{i=1}^n Y_i^k$

**Method:** For $t$ unknown parameters, solve the system:

$$\mu'_k = m'_k, \quad k = 1, 2, \ldots, t$$
:::

::: {.fragment}
**Properties of MoM estimators:**

- Ō£ģ **Always consistent** (sample moments converge to population moments)
- ŌÜĀ’ĖÅ Often **biased** and **not always efficient** (may not be functions of sufficient statistics)
- Ō£ģ **Simple** to compute ŌĆö useful as first-pass estimates
:::
::::

---

## ­¤ōī Example: Estimating Gamma Loss Distribution (Example 9.13)

:::: {style="font-size:50px"}
::: {.callout-tip}
## Setting

Insurance claim losses $Y_i \sim \text{Gamma}(\alpha, \beta)$. Estimate $\alpha$ and $\beta$.
:::

**Gamma moments:** $\mu'_1 = \alpha\beta$ and $\mu'_2 = \alpha\beta^2 + \alpha^2\beta^2$

**Set equal to sample moments and solve:**

$$\alpha\beta = \bar{Y} \qquad \text{and} \qquad \alpha\beta^2 + \alpha^2\beta^2 = \frac{1}{n}\sum Y_i^2$$
::::

---

## ­¤ōī Example 9.13: Gamma MoM Estimators ŌĆö Solution

:::: {style="font-size:50px"}
**Solutions:**

$$\hat\alpha = \frac{n\bar{Y}^2}{\sum_{i=1}^n(Y_i - \bar{Y})^2} \qquad \hat\beta = \frac{\sum_{i=1}^n(Y_i - \bar{Y})^2}{n\bar{Y}}$$

**Application:** Actuaries and risk managers use these to fit loss distributions for capital requirement calculations.
::::

---

## ­¤ō¢ 9.7 Method of Maximum Likelihood (MLE)

:::: {style="font-size:50px"}
::: {.callout-note}
## ­¤ōØ MLE Procedure

Choose as estimates the parameter values that **maximize the likelihood** of the observed sample:

$$\hat\theta = \arg\max_\theta\; L(y_1, \ldots, y_n \mid \theta) = \arg\max_\theta \prod_{i=1}^n f(y_i \mid \theta)$$

**In practice:** Maximize the **log-likelihood** (same solution, easier calculus):

$$\hat\theta = \arg\max_\theta\; \ell(\theta) = \arg\max_\theta \sum_{i=1}^n \ln f(y_i \mid \theta)$$
:::
::::

---

## ­¤ōī Example: MLE for Default Rate (Example 9.14)

:::: {style="font-size:29px"}
::: {.callout-tip}
## Setting

$n$ loans, $y = \sum y_i$ defaults observed. $Y_i \sim \text{Bernoulli}(p)$. Find MLE of $p$.
:::

::: {.fragment}
**Log-likelihood:**

$$\ell(p) = y \ln p + (n-y)\ln(1-p)$$
:::

::: {.fragment}
**Differentiate and set to zero:**

$$\frac{d\ell}{dp} = \frac{y}{p} - \frac{n-y}{1-p} = 0 \quad \Rightarrow \quad \hat{p} = \frac{y}{n} = \bar{Y}$$

Ō£ģ The MLE of the default rate is the **sample proportion** ŌĆö matching intuition.

> MLE confirmed: $\hat{p} = \bar{Y}$ is both the **MVUE** and the **MLE**!
:::
::::

---

## ­¤ōī Example: MLE for Normal Stock Returns (Example 9.15)

:::: {style="font-size:31px"}
::: {.callout-tip}
## Setting

Monthly returns $Y_i \sim N(\mu, \sigma^2)$, both unknown. Find MLEs.
:::

::: {.fragment}
**Log-likelihood:**

$$\ell(\mu, \sigma^2) = -\frac{n}{2}\ln\sigma^2 - \frac{n}{2}\ln 2\pi - \frac{1}{2\sigma^2}\sum(y_i - \mu)^2$$
:::

::: {.fragment}
**Solve $\partial\ell/\partial\mu = 0$ and $\partial\ell/\partial\sigma^2 = 0$:**

$$\hat\mu = \bar{Y} \qquad \hat\sigma^2 = \frac{1}{n}\sum_{i=1}^n(Y_i - \bar{Y})^2$$

ŌÜĀ’ĖÅ $\hat\sigma^2$ is **biased** (divides by $n$ not $n-1$). Adjust to $S^2 = \frac{n}{n-1}\hat\sigma^2$ for unbiasedness.
:::
::::

---

## ­¤ō¢ Invariance Property of MLEs

:::: {style="font-size:38px"}
::: {.callout-important}
## ­¤¦« Invariance Property

If $\hat\theta$ is the MLE of $\theta$, then the MLE of any function $t(\theta)$ is:

$$\widehat{t(\theta)} = t(\hat\theta)$$
:::

::: {.fragment}
**Finance examples:**

| Parameter $\theta$ | Function of interest $t(\theta)$ | MLE |
|---|---|---|
| Default rate $p$ | Variance $np(1-p)$ | $n\hat{p}(1-\hat{p})$ |
| Mean return $\mu$ | Sharpe ratio $\mu/\sigma$ | $\hat\mu/\hat\sigma$ |
| Intensity $\lambda$ | No-event prob $e^{-\lambda}$ | $e^{-\hat\lambda}$ |

:::
::::

---

## ­¤Ä« Interactive: MLE for Normal Returns ŌĆö Likelihood Surface {.smaller}

:::: {style="font-size: 0.72em; margin-top: -8px;"}

**Observe how the log-likelihood peaks at the MLE $(\bar{Y},\; S^2)$.**

::::: columns
::: {.column width="35%"}

```{ojs}
//| echo: false

viewof mu_true2 = {
  const input = Inputs.range([-3, 5], {value: 1.5, step: 0.5, label: "True ╬╝ (%):"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

viewof sig2 = {
  const input = Inputs.range([1, 6], {value: 3, step: 0.5, label: "True Žā (%):"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

viewof n_obs = {
  const input = Inputs.range([5, 60], {value: 20, step: 5, label: "Sample size n:"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

mle_info = {
  let sum = 0, sum2 = 0;
  const seed = Math.floor(mu_true2*10 + sig2*3 + n_obs);
  function lcg(s) { return (1664525 * s + 1013904223) >>> 0; }
  let state = seed;
  const ys = [];
  for (let i = 0; i < n_obs; i++) {
    state = lcg(state);
    const u1 = state / 4294967296;
    state = lcg(state);
    const u2 = state / 4294967296;
    const z = Math.sqrt(-2*Math.log(u1+1e-10)) * Math.cos(2*Math.PI*u2);
    const y = mu_true2 + sig2 * z;
    ys.push(y);
    sum += y; sum2 += y*y;
  }
  const ybar = sum / n_obs;
  const s2 = (sum2 - n_obs*ybar*ybar) / (n_obs - 1);
  return html`<div style="background:#d4edda; padding:8px; border-radius:6px; font-size:0.9em;">
    <strong>MLE Estimates:</strong><br>
    ╬╝╠é = ${ybar.toFixed(3)}%<br>
    Žā╠é = ${Math.sqrt(s2).toFixed(3)}%<br>
    True ╬╝ = ${mu_true2}%, Žā = ${sig2}%
  </div>`;
}
```

:::
::: {.column width="65%"}

```{ojs}
//| echo: false

likelihoodData = {
  let sum = 0, sum2 = 0;
  const seed = Math.floor(mu_true2*10 + sig2*3 + n_obs);
  function lcg(s) { return (1664525 * s + 1013904223) >>> 0; }
  let state = seed;
  for (let i = 0; i < n_obs; i++) {
    state = lcg(state);
    const u1 = state / 4294967296;
    state = lcg(state);
    const u2 = state / 4294967296;
    const z = Math.sqrt(-2*Math.log(u1+1e-10)) * Math.cos(2*Math.PI*u2);
    const y = mu_true2 + sig2 * z;
    sum += y; sum2 += y*y;
  }
  const ybar = sum / n_obs;
  const s2 = (sum2 - n_obs*ybar*ybar) / n_obs;

  const muRange = Array.from({length: 60}, (_, i) => ybar - 4 + i * 8/59);
  return muRange.map(mu_c => {
    const ll = -n_obs/2 * Math.log(s2) - (1/(2*s2)) * (sum2 - 2*mu_c*sum + n_obs*mu_c*mu_c);
    return {mu: mu_c, loglik: ll};
  });
}

Plot.plot({
  width: 650, height: 380,
  marginLeft: 60, marginBottom: 50, marginTop: 30,
  x: {label: "Candidate ╬╝ value (%)"},
  y: {label: "Log-Likelihood Ōäō(╬╝)"},
  title: `Log-Likelihood peaks at MLE: ╬╝╠é = ╚│`,
  marks: [
    Plot.line(likelihoodData, {x: "mu", y: "loglik", stroke: "steelblue", strokeWidth: 2.5}),
    Plot.ruleX([likelihoodData.reduce((a,b) => a.loglik > b.loglik ? a : b).mu],
      {stroke: "red", strokeWidth: 2.5, strokeDasharray: "6,3"}),
    Plot.text([{mu: likelihoodData.reduce((a,b)=>a.loglik>b.loglik?a:b).mu, 
                ll: Math.max(...likelihoodData.map(d=>d.loglik))}],
      {x: "mu", y: "ll", dy: -15, text: d => `MLE: ╬╝╠é=${d.mu.toFixed(2)}%`, fill:"red", fontSize:12})
  ]
})
```

:::
:::::
::::

---

## ­¤Æ░ R Case Study: MoM vs. MLE ŌĆö Estimates {.smaller}

:::: {style="font-size:50px"}

```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: true

library(tidyverse)
library(tidyquant)
library(knitr)
library(MASS)

symbols <- c("AAPL", "XOM", "JPM")
prices <- tq_get(symbols, from = "2021-01-01", to = "2023-12-31")

returns <- prices %>%
  group_by(symbol) %>%
  tq_transmute(select = adjusted,
               mutate_fun = periodReturn,
               period = "monthly",
               col_rename = "return") %>%
  mutate(return_pct = return * 100)

mom_est <- returns %>%
  group_by(symbol) %>%
  summarise(n = n(), mu_MoM = mean(return_pct), sigma_MoM = sd(return_pct),
            .groups = "drop")

kable(mom_est, digits = 3,
      caption = "Method of Moments Estimates (Monthly Returns, %)")
```

::::

---

## ­¤Æ░ R Case Study: MoM vs. MLE ŌĆö Return Distributions {.smaller}

:::: {style="font-size:29px"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 9
#| fig-height: 4

# MLE for normal: mu_MLE = Ybar, sigma_MLE^2 = (1/n)*sum(yi-ybar)^2
mle_est <- returns %>%
  group_by(symbol) %>%
  summarise(
    mu_MLE    = mean(return_pct),
    sigma_MLE = sd(return_pct) * sqrt((n()-1)/n()),
    .groups = "drop"
  )

# Visualise: fitted normal densities vs empirical
returns %>%
  left_join(mle_est, by = "symbol") %>%
  ggplot(aes(x = return_pct)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 12, fill = "steelblue",
                 alpha = 0.5, color = "white") +
  stat_function(fun = dnorm,
    args = list(mean = mle_est$mu_MLE[1],
                sd   = mle_est$sigma_MLE[1]),
    color = "red", linewidth = 1.2, linetype = "dashed") +
  facet_wrap(~symbol, ncol = 3, scales = "free_x") +
  labs(title = "Monthly Returns vs. MLE Normal Fit",
       subtitle = "Red dashed: MLE N(╬╝╠é, Žā╠é┬▓)",
       x = "Monthly Return (%)", y = "Density") +
  theme_minimal(base_size = 10)
```

::::

---

## ­¤Æ░ Case Study: Key Findings {.smaller}

:::: {style="font-size:39px"}
::: {.callout-important}
## ­¤ōŖ MoM vs. MLE ŌĆö Practical Comparison

::: {.columns}
::: {.column width="33%"}
::: {.fragment}
**Estimating $\mu$ (mean return):**

- MoM: $\hat\mu = \bar{Y}$

- MLE: $\hat\mu = \bar{Y}$

- **Identical!** For exponential family distributions, MoM and MLE agree on the mean

- Both are **MVUE** for $\mu$ (normal case)
:::
:::
::: {.column width="33%"}
::: {.fragment}
**Estimating $\sigma$ (volatility):**

- MoM: $\hat\sigma = S$ (divides by $n-1$)

- MLE: $\hat\sigma = \sqrt{\frac{1}{n}\sum(y_i-\bar{y})^2}$ (divides by $n$)

- MLE is **biased** ŌĆö slightly underestimates $\sigma$

- For risk management: prefer unbiased $S$
:::
:::
::: {.column width="33%"}
::: {.fragment}
**Takeaways for practitioners:**

1. **MLE** is consistent, invariant, and often asymptotically efficient

2. **MoM** is simple and always consistent ŌĆö good for multi-parameter models (e.g., Gamma losses)

3. For **normal returns**, both agree on $\mu$; only differ on $\sigma$

4. **MVUE** = the gold standard whenever achievable
:::
:::
:::
:::
::::

---

## ­¤ōØ Quiz #1: Relative Efficiency {.smaller .quiz-question}

Two unbiased estimators for a fund's mean return $\mu$ are computed from $n = 50$ monthly observations. $V(\hat\mu_1) = 0.10$ and $V(\hat\mu_2) = 0.25$. What is $\text{eff}(\hat\mu_1, \hat\mu_2)$?

- [$\text{eff}(\hat\mu_1, \hat\mu_2) = 2.5$, so $\hat\mu_1$ is preferred]{.correct data-explanation="Ō£ģ Correct! eff(╬Ė╠éŌéü, ╬Ė╠éŌéé) = V(╬Ė╠éŌéé)/V(╬Ė╠éŌéü) = 0.25/0.10 = 2.5. Since eff > 1, ╬Ė╠éŌéü has smaller variance and is more efficient."}
- $\text{eff}(\hat\mu_1, \hat\mu_2) = 0.4$, so $\hat\mu_2$ is preferred
- $\text{eff}(\hat\mu_1, \hat\mu_2) = 2.5$, so $\hat\mu_2$ is preferred
- Cannot be determined without knowing the distribution

---

## ­¤ōØ Quiz #2: Consistency {.smaller .quiz-question}

A portfolio analyst uses $\hat\theta_n = Y_1$ (the first observation only) to estimate mean annual return $\mu$. Is $\hat\theta_n$ consistent?

- [No ŌĆö $V(Y_1) = \sigma^2$ does not converge to 0 as $n \to \infty$]{.correct data-explanation="Ō£ģ Correct! By Theorem 9.1, an unbiased estimator is consistent if its variance ŌåÆ 0. Since V(YŌéü) = Žā┬▓ for all n, it never shrinks ŌĆö YŌéü is NOT consistent."}
- Yes ŌĆö $Y_1$ is unbiased for $\mu$ so it must be consistent
- Yes ŌĆö by the Law of Large Numbers, all estimators are consistent
- Cannot determine consistency without more information

---

## ­¤ōØ Quiz #3: Factorization Criterion {.smaller .quiz-question}

For a Poisson sample $Y_1, \ldots, Y_n$ with mean $\lambda$, the likelihood is $L(\lambda) = \frac{\lambda^{\sum y_i} e^{-n\lambda}}{\prod y_i!}$. What is the sufficient statistic for $\lambda$?

- [$U = \sum_{i=1}^n Y_i$]{.correct data-explanation="Ō£ģ Correct! Factor: L(╬╗) = [╬╗^(╬ŻyßĄó) e^(ŌłÆn╬╗)] ├Ś [1/ŌłÅyßĄó!]. The first part depends on the data only through ╬ŻyßĄó = u, and the second part is free of ╬╗. By Theorem 9.4, U = ╬ŻYßĄó is sufficient."}
- $U = \bar{Y}$, since it is unbiased for $\lambda$
- $U = Y_{(n)} = \max(Y_1, \ldots, Y_n)$
- The sufficient statistic is $\lambda$ itself

---

## ­¤ōØ Quiz #4: MLE vs. MoM {.smaller .quiz-question}

For a Uniform$(0, \theta)$ sample, the MoM estimator is $\hat\theta_{\text{MoM}} = 2\bar{Y}$ and the MLE is $\hat\theta_{\text{MLE}} = Y_{(n)} = \max(Y_i)$. Which statement is correct?

- [The MLE is more efficient; the sufficient statistic is $Y_{(n)}$, so the MVUE is based on $Y_{(n)}$, not $\bar{Y}$]{.correct data-explanation="Ō£ģ Correct! The factorization criterion gives YŌéŹŌéÖŌéÄ as the sufficient statistic. eff(MoM, MLE) = 3/(n+2) < 1 for n > 1, meaning the MLE has smaller variance. The MoM estimator 2╚▓ is NOT a function of the sufficient statistic, so it cannot be MVUE."}
- The MoM is preferred because it is unbiased
- Both estimators are equally efficient for all $n$
- The MLE is biased and therefore should never be used

---

## ­¤ōØ Summary {.smaller}

:::: {style="font-size:41px"}
::: {.callout-note}
## Ō£ģ Key Takeaways ŌĆö Chapter 9

::: {.columns}
::: {.column width="50%"}
**Evaluating Estimators**

- **Efficiency** ŌĆö compare via $\text{eff} = V(\hat\theta_2)/V(\hat\theta_1)$; prefer estimator with smaller variance

- **Consistency** ŌĆö $V(\hat\theta_n) \to 0$ for unbiased $\hat\theta_n$ $\Rightarrow$ consistent (Theorem 9.1); $\bar{Y}$ is always consistent (LLN)

- **Sufficiency** ŌĆö $U$ is sufficient for $\theta$ if the likelihood factors as $g(u,\theta) \cdot h(\mathbf{y})$ (Factorization Criterion)

- **MVUE** ŌĆö RaoŌĆōBlackwell: best unbiased estimators are functions of sufficient statistics
:::
::: {.column width="50%"}
**Finding Estimators**

- **Method of Moments** ŌĆö equate $E(Y^k) = m'_k$; always consistent, easy to compute, sometimes inefficient

- **MLE** ŌĆö maximize $L(\theta)$ or $\ell(\theta)$; consistent, invariant $[t(\hat\theta)$ is MLE of $t(\theta)]$, often asymptotically efficient

- **Hierarchy:** MVUE $\geq$ MLE (adjusted) $\geq$ MoM in terms of efficiency

- **Finance link:** $\bar{Y}$ = MLE = MVUE for $\mu$ in normal returns; $S^2$ = MVUE for $\sigma^2$
:::
:::
:::
::::

---

## ­¤ōÜ Practice Problems

:::: {style="font-size:50px"}
::: {.callout-tip}
## ­¤ōØ Homework (Wackerly Ch. 9)

**Problem 1 (Efficiency ŌĆö Ex. 9.5):** Let $\hat\sigma^2_1 = S^2$ and $\hat\sigma^2_2 = \frac{1}{2}(Y_1-Y_2)^2$ be two unbiased estimators of $\sigma^2$ from a normal sample. Find $\text{eff}(\hat\sigma^2_1, \hat\sigma^2_2)$.

**Problem 2 (Consistency ŌĆö Ex. 9.20):** If $Y \sim \text{Binomial}(n, p)$, show that $Y/n$ is a consistent estimator of $p$.

**Problem 3 (Sufficiency ŌĆö Ex. 9.38a):** For a normal sample with known $\sigma^2$, show $\bar{Y}$ is sufficient for $\mu$.

**Problem 4 (MLE ŌĆö Ex. 9.80):** For a Poisson sample with mean $\lambda$: (a) find $\hat\lambda_{\text{MLE}}$; (b) find $E(\hat\lambda)$ and $V(\hat\lambda)$; (c) what is the MLE of $P(Y=0) = e^{-\lambda}$?
:::
::::

---

## ­¤æŗ Thank You!

:::: {style="font-size:44px"}
::: {.columns}
::: {.column width="55%"}
**Contact**

­¤ō¦ sorujov@ada.edu.az

­¤Åø ADA University, School of Business

­¤Åó ICTA Statistics Unit

**Next Lecture**

­¤ōī **Topic 8: Hypothesis Testing**

Chapter 10 ŌĆö Testing large-sample means, proportions, and variances with financial applications
:::
::: {.column width="45%"}
::: {.callout-note}
## ­¤ōī Key Formulas

$$\text{eff}(\hat\theta_1, \hat\theta_2) = \frac{V(\hat\theta_2)}{V(\hat\theta_1)}$$

$$\hat\theta_n \text{ consistent if } V(\hat\theta_n) \to 0$$

$$L(\theta) = g(u,\theta) \cdot h(\mathbf{y}) \Rightarrow U \text{ sufficient}$$

$$\hat\theta^* = E(\hat\theta \mid U) \Rightarrow V(\hat\theta^*) \leq V(\hat\theta)$$
:::
:::
:::
::::

---

## ŌØō Discussion Questions

:::: {style="font-size:50px"}
::: {.callout-note}
## ­¤Æ¼ Open Questions

1. In practice, when would you accept a **biased but consistent** estimator over an **unbiased but inefficient** one? Give an example from portfolio management.

2. The MLE for $\sigma^2$ in the normal model divides by $n$ (biased). Yet many practitioners use it anyway. What trade-off are they making?

3. Can you think of a financial quantity where finding the **sufficient statistic** would substantially simplify estimation? (Hint: think about VaR or CVaR.)

4. How does the **invariance property of MLEs** help when estimating the **Sharpe Ratio** $\mu/\sigma$?
:::
::::

<div class="attendance-qr" data-status="late" id="qr-late" style="min-height: 400px; display: flex; align-items: center; justify-content: center;"></div>
