---
title: "Mathematical Statistics"
subtitle: "The Binomial Probability Distribution"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - The Binomial Distribution"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
library(ggplot2)
```

## 🎯 Learning Objectives

::: {style="font-size: 32px"}
By the end of this lecture, you will be able to:

- **Check** an experiment against the five properties of a binomial experiment (Definition 3.6)

- **Derive** $p(y) = \binom{n}{y} p^y q^{n-y}$ from sample points and the counting rule

- **Compute** binomial probabilities by hand, from Table 1 and with `dbinom` / `pbinom`

- **Use** $E(Y) = np$ and $V(Y) = npq$ (Theorem 3.7) to price and to judge risk

- **Read** an unlikely count as evidence against an assumed $p$
:::

---

## 🗺️ Where We Are

::: {style="font-size: 30px"}
Last class ended on: *§3.4 — the binomial distribution.*

On Wednesday every $p(y)$ came as a **table**, and $E(Y)$ and $V(Y)$ were two sums over it. Today one **formula** covers a whole family of tables — indexed by just two numbers, $n$ and $p$.

::: {.callout-important}
## The question this lecture answers
A Baku bank's SME desk holds 20 loans. Each defaults this year with probability 0.05, independently of the others. **How likely are three or more defaults — and how many should the desk budget for?**
:::
:::

---

## 📐 The Binomial Experiment

::: {style="font-size: 29px"}
::: {.callout-note}
## Definition 3.6
1. The experiment consists of a **fixed number, $n$**, of identical trials.
2. Each trial results in one of **two outcomes**: success, $S$, or failure, $F$.
3. The probability of success on a single trial is equal to some value $p$ and **remains the same** from trial to trial. The probability of a failure is $q = 1 - p$.
4. The trials are **independent**.
5. The random variable of interest is $Y$, the **number of successes** observed during the $n$ trials.
:::

"Success" is only a name for one of the two outcomes. For a credit analyst, a *default* is the success.
:::

---

## 💳 Checking the Five: Card Declines

::: {style="font-size: 29px"}
A payment processor watches the next **8** card transactions at a Baku supermarket. Each is declined with probability $0.03$, independently. $Y$ = number declined. Binomial?

::: {.fragment}
1. **Fixed $n$**: 8 transactions, decided in advance. ✔
2. **Two outcomes**: declined ($S$) or approved ($F$). ✔
3. **Same $p$**: $P(S) = 0.03$ on every transaction. ✔
4. **Independent**: stated. ✔
5. **Counts successes**: $Y$ counts declines. ✔ So $n = 8$, $p = 0.03$.
:::

::: {.fragment}
"Transactions **until** the first decline" breaks property 1 — that is §3.5.
:::
:::

---

## 🌐 Approximately Binomial

::: {style="font-size: 29px"}
An ISP has 250,000 households in Baku; 35% of them reach the advertised 100 Mbps at peak hour. We test **10** households at random. $Y$ = number reaching 100 Mbps.

::: {.fragment}
Sampling is **without** replacement, so strictly the trials are dependent. But removing a few households barely changes the 35%: the conditional probabilities stay very close to $0.35$, and $Y$ is **approximately** binomial with $n = 10$, $p = 0.35$.
:::

::: {.fragment}
If the sample were a large fraction of the population — say 10% — this fails, and the right model is the **hypergeometric** of §3.7.
:::
:::

---

## 🧮 Deriving $p(y)$

::: {style="font-size: 29px"}
A sample point is an $n$-tuple of letters, e.g. $SSFSF\cdots FS$. Take the one with $y$ successes first, then $n - y$ failures:
$$S\,S \cdots S\;F\,F \cdots F \quad\text{has probability}\quad (p \cdot p \cdots p)(q \cdot q \cdots q) = p^y q^{n-y}$$
by independence (Theorem 2.5, applied $n$ times): $y$ factors of $p$, $n - y$ factors of $q$.

::: {.fragment}
Every **other** ordering of $y$ $S$'s and $n - y$ $F$'s has the same probability, and by Theorem 2.3 there are $\binom{n}{y} = \frac{n!}{y!(n-y)!}$ of them.
:::

::: {.fragment}
So $P(Y = y) = \binom{n}{y} p^y q^{n-y}$ — a count times a probability.
:::
:::

---

## 📐 Definition 3.7

::: {style="font-size: 30px"}
::: {.callout-note}
## Definition 3.7
A random variable $Y$ has a **binomial distribution** based on $n$ trials with success probability $p$ if and only if
$$p(y) = \binom{n}{y} p^y q^{n-y}, \qquad y = 0, 1, 2, \dots, n \text{ and } 0 \le p \le 1.$$
:::

::: {.fragment}
Theorem 3.1 holds: each term is positive, and they are the terms of the binomial expansion,
$$\sum_{y=0}^{n} \binom{n}{y} p^y q^{n-y} = (q + p)^n = 1^n = 1.$$
:::
:::

---

## 🏦 Example: The SME Loan Book

::: {style="font-size: 29px"}
$n = 20$ loans, default ($S$) probability $p = 0.05$, so $q = 0.95$.

::: {.fragment}
**No defaults:** $\;p(0) = \binom{20}{0}(0.05)^0(0.95)^{20} = 0.3585$
:::

::: {.fragment}
**At least one** (the complement, as in Lecture 6): $\;1 - p(0) = 0.6415$
:::

::: {.fragment}
**Three or more:** $\;P(Y \ge 3) = 1 - [p(0) + p(1) + p(2)]$
$$= 1 - [0.3585 + 0.3774 + 0.1887] = 1 - 0.9245 = 0.0755$$
(Terms rounded; the unrounded sum is 0.92452.) With 50,000 AZN per loan, a year of 150,000 AZN or more in defaulted principal has about a **7.5%** chance.
:::
:::

---

## 💻 The Same Numbers in R

```{r}
#| label: sme-book
#| code-fold: false
n <- 20; p <- 0.05

round(dbinom(0:4, size = n, prob = p), 4)   # p(0), ..., p(4)

c(none         = dbinom(0, n, p),
  at_least_one = 1 - dbinom(0, n, p),
  three_plus   = 1 - pbinom(2, n, p),        # P(Y >= 3) = 1 - P(Y <= 2)
  total        = sum(dbinom(0:n, n, p))) |> round(4)

# Simulate 100,000 years of the loan book
years <- rbinom(100000, size = n, prob = p)
round(mean(years >= 3), 4)
```

---

## 📈 The Shape Depends on $p$ and $n$

```{r}
#| label: shapes-figure
#| echo: false
#| fig-width: 11
#| fig-height: 4.3
shapes <- rbind(
  data.frame(panel = "n = 10, p = 0.1", y = 0:10, p = dbinom(0:10, 10, 0.1)),
  data.frame(panel = "n = 10, p = 0.5", y = 0:10, p = dbinom(0:10, 10, 0.5)),
  data.frame(panel = "n = 20, p = 0.5", y = 0:20, p = dbinom(0:20, 20, 0.5)))
ggplot(shapes, aes(y, p)) +
  geom_col(width = 0.85, fill = "#8ba3c7") +
  facet_wrap(~ panel, scales = "free_x") +
  scale_x_continuous(breaks = function(l) if (l[2] > 12) seq(0, 20, 5) else seq(0, 10, 2),
                     minor_breaks = NULL) +
  scale_y_continuous(breaks = seq(0, 0.4, 0.1)) +
  labs(x = "Number of successes, y", y = "p(y)") +
  theme_minimal(base_size = 20) +
  theme(strip.text = element_text(size = 20, face = "bold"))
```

::: {style="font-size: 28px"}
Small $p$: piled up near 0 and skewed right. $p = 0.5$: symmetric. Larger $n$: wider and lower.
:::

---

## 📋 Table 1 and `pbinom`: Card Declines

::: {style="font-size: 29px"}
The acquiring bank's contract says a terminal's decline rate is **5%**. An auditor pulls $n = 25$ transactions and finds **4** declines. How surprising is that, if the contract is right?

::: {.fragment}
$$P(Y \ge 4) = 1 - P(Y \le 3) = 1 - \sum_{y=0}^{3} p(y)$$
Table 1, Appendix 3: the table for $n = 25$, column $p = .05$, row $a = 3$ gives $0.966$.

In R: `pbinom(3, 25, 0.05)` returns the same $0.966$.
:::

::: {.fragment}
$$P(Y \ge 4) = 1 - 0.966 = 0.034$$
Either a 3-in-100 event happened, or the true decline rate is **higher than 5%**. As in Example 3.9, the auditor has grounds to question the reported rate.
:::
:::

---

## 📐 Theorem 3.7: Mean and Variance

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorem 3.7
Let $Y$ be a binomial random variable based on $n$ trials and success probability $p$. Then
$$\mu = E(Y) = np \qquad\text{and}\qquad \sigma^2 = V(Y) = npq.$$
:::

::: {.fragment}
**SME book:** $\mu = 20(0.05) = 1$ default a year, $\sigma^2 = 20(0.05)(0.95) = 0.95$, $\sigma = 0.975$.
:::

::: {.fragment}
The mean is what intuition says: 5% of 20. The variance is the new information — and $pq$ is largest at $p = 0.5$, where the outcome is least predictable.
:::
:::

---

## 🔍 The Proof: Two Tricks Worth Keeping

::: {style="font-size: 28px"}
**Mean.** The $y = 0$ term vanishes; cancel $y$ against $y!$, factor out $np$, set $z = y - 1$:
$$E(Y) = np \sum_{z=0}^{n-1} \binom{n-1}{z} p^z q^{n-1-z} = np \cdot 1$$
The sum is a binomial distribution on $n - 1$ trials, so it equals 1.

::: {.fragment}
**Variance.** $E(Y^2)$ does not cancel against $y!$, but $E[Y(Y-1)]$ does. The same device gives $E[Y(Y-1)] = n(n-1)p^2$, so
$$\sigma^2 = E[Y(Y-1)] + \mu - \mu^2 = n(n-1)p^2 + np - n^2p^2 = npq.$$
:::

::: {.fragment}
Trick 1: $\sum p(y) = 1$ evaluates a sum for free. Trick 2: find $E[Y(Y-1)]$, not $E(Y^2)$.
:::
:::

---

## 📞 Worked Example: A Sales Shift

::: {style="font-size: 28px"}
A mobile operator's outbound team makes $n = 40$ calls a shift; each converts to a sale with $p = 0.15$, independently. An agent earns **30 AZN** base plus **12 AZN** per sale.

::: {.fragment}
Sales: $E(Y) = 40(0.15) = 6$, $\;V(Y) = 40(0.15)(0.85) = 5.1$, $\;\sigma = 2.26$.
:::

::: {.fragment}
Pay $W = 30 + 12Y$ — Wednesday's linear rule, Theorems 3.3–3.5 and $V(aY + b) = a^2 V(Y)$:
$$E(W) = 30 + 12(6) = 102 \text{ AZN}, \qquad V(W) = 12^2(5.1) = 734.4, \qquad \sigma_W = 27.10 \text{ AZN}$$
:::

::: {.fragment}
A 10-sale shift (150 AZN): $P(Y \ge 10) = 1 -$ `pbinom(9, 40, 0.15)` $= 0.067$.
:::
:::

---

## 🎯 Using the Model Backwards

::: {style="font-size: 28px"}
The regulator tests 20 households of a new ISP and **6** reach 100 Mbps. Which $p$ makes that result most probable? (Example 3.10)

::: {.fragment}
Maximise $\ln P(Y = 6) = \ln\binom{20}{6} + 6\ln p + 14\ln(1-p)$:
$$\frac{6}{p} - \frac{14}{1-p} = 0 \;\Rightarrow\; \hat{p} = \frac{6}{20} = 0.30$$
:::

```{r}
#| label: likelihood-figure
#| echo: false
#| fig-width: 11
#| fig-height: 2.3
lik <- data.frame(p = seq(0, 1, by = 0.005))
lik$L <- dbinom(6, 20, lik$p)
ggplot(lik, aes(p, L)) +
  geom_line(linewidth = 1.2, colour = "#14130f") +
  geom_vline(xintercept = 0.3, colour = "#8b2635", linewidth = 1.1, linetype = "dashed") +
  annotate("text", x = 0.42, y = 0.17, hjust = 0, size = 6.5, colour = "#8b2635",
           label = "p = 0.30,  P(Y = 6) = 0.192") +
  scale_x_continuous(breaks = seq(0, 1, 0.1)) +
  labs(x = "Candidate p", y = "P(Y = 6)") +
  theme_minimal(base_size = 19)
```
:::

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 30px"}
A microfinance lender in Ganja holds **10** independent loans of **8,000 AZN** each. Each defaults with probability **0.10**, and a default loses the full principal.

**Four minutes, in pairs:**

1. What is the probability that no loan defaults?

2. What is the probability of **two or more** defaults?

3. What are the mean and standard deviation of the AZN loss?
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 29px"}
$Y \sim$ binomial with $n = 10$, $p = 0.10$, $q = 0.90$.

1. $p(0) = (0.9)^{10} = 0.3487$

2. $p(1) = \binom{10}{1}(0.1)(0.9)^9 = 0.3874$, so $P(Y \ge 2) = 1 - 0.3487 - 0.3874 = 0.2639$

::: {.fragment}
3. Loss $L = 8000\,Y$. $E(Y) = np = 1$ and $V(Y) = npq = 0.9$, so
$$E(L) = 8{,}000 \text{ AZN}, \qquad \sigma_L = 8000\sqrt{0.9} = 7{,}589 \text{ AZN}$$
The standard deviation is nearly as large as the mean: ten loans are too few to diversify away default risk.
:::
:::

---

## 🔬 Interactive: How $p$ Moves the Shape {.smaller}

```{ojs}
//| echo: false

viewof p_hit = {
  const input = Inputs.range([0.05, 0.95], {value: 0.35, step: 0.01,
    label: "p, each household hits 100 Mbps:"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}
```

```{ojs}
//| echo: false

nTest = 20
choose = (n, k) => { let c = 1; for (let i = 1; i <= k; i++) c = c * (n - k + i) / i; return c; }
pmfB = Array.from({length: nTest + 1}, (_, y) =>
  ({y: y, p: choose(nTest, y) * Math.pow(p_hit, y) * Math.pow(1 - p_hit, nTest - y)}))

md`n = 20 households · mean **np = ${(nTest * p_hit).toFixed(1)}** · variance **npq = ${(nTest * p_hit * (1 - p_hit)).toFixed(2)}**`

Plot.plot({
  width: 1150,
  height: 300,
  marginTop: 40,
  marginLeft: 78,
  marginBottom: 58,
  style: {fontSize: "18px"},
  x: {label: "Households of 20 reaching 100 Mbps, y", domain: d3.range(0, 21), padding: 0.15},
  y: {label: "p(y)", domain: [0, 0.4], ticks: [0, 0.1, 0.2, 0.3, 0.4], tickFormat: ".1f", grid: true},
  marks: [
    Plot.barY(pmfB, {x: "y", y: "p",
      fill: d => d.y === Math.round(nTest * p_hit) ? "#8b2635" : "#8ba3c7"}),
    Plot.ruleY([0])
  ]
})
```

::: {style="font-size: 28px"}
The red bar is the one nearest $np$. Near $p = 0.05$ or $0.95$ the mass piles against an end; at $p = 0.5$ it is symmetric and widest.
:::

---

## 📝 Quiz #1: Is It Binomial? {.quiz-question}

Which of these random variables has a binomial distribution?

- [The number of 50 card transactions declined, each declined with probability 0.03, independently]{.correct data-explanation="✅ Fixed n = 50, two outcomes, constant p = 0.03, independent trials, and Y counts successes: all five properties of Definition 3.6 hold."}
- The number of cold calls made until the first sale
- The total AZN value of 20 independent transactions
- The number of bad loans in a file review of 6 drawn from a pool of 10 loans, 3 of them bad

---

## 📝 Quiz #2: The Variance {.quiz-question}

A telecom sells to each of $n = 50$ contacted customers with probability $p = 0.2$, independently. What is $V(Y)$?

- [$8$]{.correct data-explanation="✅ Theorem 3.7: V(Y) = npq = 50 × 0.2 × 0.8 = 8. The mean is np = 10; the standard deviation is √8 = 2.83."}
- $10$
- $2.83$
- $40$

---

## 📝 Quiz #3: Which Outcome Is the Success? {.quiz-question}

Each of 8 independent SME loans **defaults** with probability 0.1. $Y$ = number of loans that **repay**. What is $P(Y = 8)$?

- [$0.9^8 = 0.4305$]{.correct data-explanation="✅ Y counts repayments, so the success probability is p = 0.9, not 0.1. P(Y = 8) = 0.9⁸ = 0.4305. Define the success first, then make sure p is its probability."}
- $0.1^8 = 10^{-8}$
- $1 - 0.9^8 = 0.5695$
- $8 \times 0.9 = 7.2$

---

## 📋 Key Formulas

::: {style="font-size: 30px"}
| | Statement |
|---|---|
| Definition 3.7 | $p(y) = \binom{n}{y} p^y q^{n-y}, \quad y = 0, 1, \dots, n$ |
| validity | $\sum_{y=0}^{n} p(y) = (q + p)^n = 1$ |
| at least one | $P(Y \ge 1) = 1 - q^n$ |
| Theorem 3.7 | $E(Y) = np, \qquad V(Y) = npq$ |
| $p(y)$ in R | `dbinom(y, n, p)` |
| $P(Y \le a)$ in R | `pbinom(a, n, p)` |
| estimate of $p$ (Ex. 3.10) | $\hat{p} = y/n$ |
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- A binomial experiment: fixed $n$, two outcomes, constant $p$, independent trials, $Y$ counts successes

- Sampling a small part of a large population is approximately binomial; a large part is not (§3.7)

- $p(y)$ is a count, $\binom{n}{y}$, times the probability of one ordering, $p^y q^{n-y}$

- $E(Y) = np$ and $V(Y) = npq$; with Wednesday's linear rules they price a loss or a pay scheme

- An improbable count under an assumed $p$ is evidence against that $p$

- Name the success first — then make sure $p$ is **its** probability
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition**, exercises at the end of §3.4:

- 3.38, 3.39, 3.40, 3.43, 3.48, 3.54, 3.56, 3.57, 3.59

**Week 6, Problem Set 1** is open now and closes **Sunday 25 October at 23:59** on WeBWorK, covering §3.4.

**Quiz I** is this **Saturday, 17 October**, in the first 30 minutes of class, on **Chapters 1–2**.

**Next class:** Saturday 17 October — the geometric and negative binomial distributions, Wackerly §§3.5–3.6: what happens when the number of trials is the random quantity.
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**

📧 sorujov@ada.edu.az\
🏢 Building D, Room D325\
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- Loan defaults tend to rise together in a recession. Which property of Definition 3.6 does that break, and in which direction does it push $P(Y \ge 3)$?

- If $Y$ is binomial$(n, p)$, what is the distribution of $n - Y$?

- For fixed $n$, which $p$ makes $V(Y)$ largest, and why is that the hardest book to reserve for?
:::
