---
title: "Mathematical Statistics"
subtitle: "Moments and Moment-Generating Functions"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - Moment-Generating Functions"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
# Primes in the maths are typed ^{\text{ʹ}} (U+02B9), not ' : MathJax 2 fetches its
# AMS font for U+2032 and the headless QA pass logs that fetch as a page error.
library(ggplot2)
theme_set(theme_minimal(base_size = 18))
```

## 🎯 Learning Objectives

::: {style="font-size: 32px"}
By the end of this lecture, you will be able to:

- **Distinguish** moments about the origin $\mu^{\text{ʹ}}_k$ from central moments $\mu_k$

- **Derive** the moment-generating function of a Poisson or binomial count

- **Recover** $E(Y)$ and $V(Y)$ from $m(t)$ by differentiating at $t = 0$ (Theorem 3.12)

- **Transform** an MGF when a count becomes a P&L: $m_{aY+b}(t) = e^{bt}m_Y(at)$

- **Identify** a distribution from its MGF, using uniqueness
:::

---

## 🗺️ Where We Are

::: {style="font-size: 32px"}
**Wackerly §3.9** · first lecture after Midterm I

On 21 October we met the last two discrete models of the chapter, the **hypergeometric and the Poisson**, and then reviewed §§3.1–3.8 for Saturday's midterm.

::: {.fragment}
For every model so far, $\mu$ and $\sigma^2$ cost a **new series trick**: the binomial theorem, a geometric sum, $E[Y(Y-1)]$ for the Poisson.
:::

::: {.fragment}
Today: **one function per distribution** that hands over every moment by differentiation — and, more importantly, names the distribution.
:::
:::

---

## ❓ Motivating Question

::: {style="font-size: 30px"}
A Baku bank's treasury counts large ($\geq$ 1M AZN) corporate payments cleared per hour at two branches.

| Payments $y$ | 0 | 1 | 2 | 3 | 4 |
|---|---|---|---|---|---|
| Branch A, $p(y)$ | 0 | 1/2 | 0 | 1/2 | 0 |
| Branch B, $p(y)$ | 1/8 | 0 | 3/4 | 0 | 1/8 |

Both have $\mu = 2$ and $\sigma^2 = 1$.

::: {.callout-important}
## The question
If two counts share a mean and a variance, is the risk the same? And what extra numbers **would** pin the distribution down?
:::
:::

---

## 📐 Definitions 3.12 and 3.13

::: {style="font-size: 30px"}
::: {.callout-note}
## Definition 3.12
The $k$th moment of $Y$ taken **about the origin** is $E(Y^k)$, denoted $\mu^{\text{ʹ}}_k$.
:::

::: {.callout-note}
## Definition 3.13
The $k$th moment of $Y$ taken **about its mean**, the $k$th central moment, is $E[(Y-\mu)^k]$, denoted $\mu_k$.
:::

So $\mu^{\text{ʹ}}_1 = \mu$, $\mu_2 = \sigma^2$, and Theorem 3.6 reads $\sigma^2 = \mu^{\text{ʹ}}_2 - (\mu^{\text{ʹ}}_1)^2$.
:::

---

## 💻 Two Branches, Four Moments

```{r}
#| label: branch-moments
yA <- c(1, 3);    pA <- c(1/2, 1/2)
yB <- c(0, 2, 4); pB <- c(1/8, 3/4, 1/8)
mom <- function(y, p, k, about = 0) sum((y - about)^k * p)

data.frame(k = 1:4,
  A_origin  = sapply(1:4, \(k) mom(yA, pA, k)),
  B_origin  = sapply(1:4, \(k) mom(yB, pB, k)),
  A_central = sapply(1:4, \(k) mom(yA, pA, k, 2)),
  B_central = sapply(1:4, \(k) mom(yB, pB, k, 2)))
```

::: {style="font-size: 28px"}
Identical up to $k = 3$; the **fourth** moment separates them. Branch B has more mass far from the mean — fatter tails.
:::

---

## 📐 Definition 3.14: The MGF

::: {style="font-size: 30px"}
::: {.callout-note}
## Definition 3.14
The moment-generating function $m(t)$ for a random variable $Y$ is $m(t) = E(e^{tY})$. We say that an MGF for $Y$ exists if there is a constant $b > 0$ such that $m(t)$ is finite for $|t| \leq b$.
:::

::: {.fragment}
For a discrete $Y$ it is one sum:
$$m(t) = \sum_y e^{ty}p(y)$$
It is a function of $t$, not of $y$; the $y$'s have been summed out. And always $m(0) = \sum_y p(y) = 1$.
:::
:::

---

## 🔑 Why "Moment-Generating"?

::: {style="font-size: 30px"}
Expand $e^{ty} = 1 + ty + \dfrac{(ty)^2}{2!} + \dfrac{(ty)^3}{3!} + \cdots$ and take expectations term by term:

$$m(t) = 1 + t\mu^{\text{ʹ}}_1 + \frac{t^2}{2!}\mu^{\text{ʹ}}_2 + \frac{t^3}{3!}\mu^{\text{ʹ}}_3 + \cdots$$

::: {.fragment}
**$\mu^{\text{ʹ}}_k$ is the coefficient of $t^k/k!$.** The MGF is every moment of $Y$ packed into one expression.
:::

::: {.fragment}
Swapping the sum over $y$ with the infinite series is justified whenever $m(t)$ exists near 0.
:::
:::

---

## 📐 Theorem 3.12

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorem 3.12
If $m(t)$ exists, then for any positive integer $k$,
$$\left.\frac{d^k m(t)}{dt^k}\right|_{t=0} = m^{(k)}(0) = \mu^{\text{ʹ}}_k$$
:::

::: {.fragment}
Differentiate the series $k$ times: every term below $t^k$ vanishes, the $t^k$ term leaves $\mu^{\text{ʹ}}_k$, and every higher term still carries a power of $t$, so it dies at $t = 0$.
:::

::: {.fragment}
In practice: $\mu = m^{\text{ʹ}}(0)$ and $\sigma^2 = m^{\text{ʹʹ}}(0) - [m^{\text{ʹ}}(0)]^2$.
:::
:::

---

## 🚗 Example 3.23: Motor Claims

::: {style="font-size: 30px"}
A Baku motor insurer's daily claim count is $Y \sim$ Poisson($\lambda$).

$$m(t) = \sum_{y=0}^{\infty} e^{ty}\frac{\lambda^y e^{-\lambda}}{y!} = e^{-\lambda}\sum_{y=0}^{\infty}\frac{(\lambda e^t)^y}{y!}$$

::: {.fragment}
The sum is the Taylor series of $e^{\lambda e^t}$, so
$$m(t) = e^{-\lambda}e^{\lambda e^t} = e^{\lambda(e^t - 1)}$$
:::

::: {.fragment}
The same trick as in Theorem 3.11: turn what is left into a Poisson p.f. with mean $\lambda e^t$, which sums to 1.
:::
:::

---

## 🧮 Example 3.24: Mean and Variance

::: {style="font-size: 30px"}
$$m^{\text{ʹ}}(t) = e^{\lambda(e^t-1)}\,\lambda e^t, \qquad m^{\text{ʹʹ}}(t) = e^{\lambda(e^t-1)}(\lambda e^t)^2 + e^{\lambda(e^t-1)}\lambda e^t$$

::: {.fragment}
At $t = 0$: $\mu = m^{\text{ʹ}}(0) = \lambda$ and $\mu^{\text{ʹ}}_2 = m^{\text{ʹʹ}}(0) = \lambda^2 + \lambda$, so
$$\sigma^2 = \lambda^2 + \lambda - \lambda^2 = \lambda$$
:::

::: {.fragment}
With $\lambda = 4.5$ claims a day: $\mu^{\text{ʹ}}_2 = 24.75$, $\sigma^2 = 4.5$, $\sigma = 2.12$. **No $E[Y(Y-1)]$ detour needed.**
:::
:::

---

## 📈 The Binomial MGF: Winning Trades

::: {style="font-size: 30px"}
An FX dealer books $n = 10$ independent trades a day, each profitable with $p = 0.6$. $Y$ = winning trades $\sim$ Bin$(10, 0.6)$. By the binomial theorem (Exercise 3.145):
$$m(t) = \sum_{y=0}^{n}\binom{n}{y}(pe^t)^y q^{n-y} = (pe^t + q)^n$$

::: {.fragment}
$m^{\text{ʹ}}(0) = np = 6$, and $m^{\text{ʹʹ}}(0) = n(n-1)p^2 + np = 32.4 + 6 = 38.4$, so
$$V(Y) = 38.4 - 6^2 = 2.4 = npq$$
:::
:::

---

## 💰 From Trade Count to Daily P&L

::: {style="font-size: 30px"}
Each winner earns 2 thousand AZN, each loser costs 1: $X = 2Y - (10 - Y) = 3Y - 10$.

::: {.callout-important}
## Exercise 3.156
If $X = aY + b$, then $m_X(t) = E(e^{t(aY+b)}) = e^{bt}\,m_Y(at)$.
:::

::: {.fragment}
$$m_X(t) = e^{-10t}\,(0.6e^{3t} + 0.4)^{10}$$
Differentiating at 0: $E(X) = 3(6) - 10 = 8$ thousand AZN, $V(X) = 9(2.4) = 21.6$, $\sigma_X = 4.65$.
:::
:::

---

## 💻 Checking Theorem 3.12 Numerically

```{r}
#| label: numeric-check
mgf <- function(t, y, p) sapply(t, \(s) sum(exp(s * y) * p))
h <- 1e-4
d1 <- function(m) (m(h) - m(-h)) / (2 * h)            # m'(0)
d2 <- function(m) (m(h) - 2 * m(0) + m(-h)) / h^2     # m''(0)

claims <- \(t) mgf(t, 0:200, dpois(0:200, 4.5))
x <- 3 * (0:10) - 10                                   # P&L support
pnl    <- \(t) mgf(t, x, dbinom(0:10, 10, 0.6))

round(rbind(claims = c(mean = d1(claims), var = d2(claims) - d1(claims)^2),
            pnl    = c(mean = d1(pnl),    var = d2(pnl)    - d1(pnl)^2)), 3)
```

::: {style="font-size: 28px"}
Brute-force sums and finite differences agree with the calculus: 4.5 and 4.5, 8 and 21.6.
:::

---

## 🖼️ Same Slope, Different Curvature

```{r}
#| label: mgf-figure
#| echo: false
#| fig-width: 10
#| fig-height: 4.1
tt <- seq(-0.8, 0.4, length.out = 300)
df <- rbind(
  data.frame(t = tt, m = exp(4.5 * (exp(tt) - 1)), model = "Poisson(4.5): var 4.5"),
  data.frame(t = tt, m = (0.45 * exp(tt) + 0.55)^10, model = "Bin(10, 0.45): var 2.475"))
ggplot(df, aes(t, m, colour = model)) +
  geom_abline(intercept = 1, slope = 4.5, linetype = "dashed", colour = "grey55") +
  geom_line(linewidth = 1.3) +
  annotate("point", x = 0, y = 1, size = 3.5) +
  scale_colour_manual(values = c("#8b2635", "#1f4e79")) +
  coord_cartesian(ylim = c(0, 5)) +
  scale_x_continuous(breaks = seq(-0.8, 0.4, by = 0.2)) +
  labs(x = "t", y = "m(t)", colour = NULL) +
  theme(legend.position = "top", legend.text = element_text(size = 18),
        axis.text = element_text(size = 17), axis.title = element_text(size = 19))
```

::: {style="font-size: 28px"}
Slope at 0 is the mean, curvature is $E(Y^2)$. Both counts average 4.5, so both curves leave $(0, 1)$ along the dashed tangent. The Poisson bends up faster: $\mu^{\text{ʹ}}_2 = 24.75$ against $22.725$.
:::

---

## 🔐 The Primary Use: Uniqueness

::: {style="font-size: 30px"}
If $m(t)$ exists for $p(y)$, it is **unique**. If $m_Y(t) = m_Z(t)$ for all $|t| < b$, some $b > 0$, then $Y$ and $Z$ have the **same distribution**.

::: {.fragment}
**Example 3.25, re-set.** A bank's monthly count of operational-loss events has $m_Y(t) = e^{3.2(e^t - 1)}$.
:::

::: {.fragment}
That is the Poisson MGF with $\lambda = 3.2$, so $Y \sim$ Poisson(3.2): mean and variance 3.2, and a loss-free month has probability $e^{-3.2} = 0.0408$.
:::
:::

---

## 🧭 A Recognition Table

::: {style="font-size: 30px"}
| Distribution | $m(t)$ | Source |
|---|---|---|
| Binomial$(n, p)$ | $(pe^t + q)^n$ | Exercise 3.145 |
| Geometric$(p)$ | $\dfrac{pe^t}{1 - qe^t}$ | Exercise 3.147 |
| Poisson$(\lambda)$ | $e^{\lambda(e^t - 1)}$ | Example 3.23 |

::: {.fragment}
Of 12 SME loan applications, $Y$ approved has $m(t) = (0.3e^t + 0.7)^{12}$. So $Y \sim$ Bin(12, 0.3): $\mu = 3.6$, $\sigma^2 = 2.52$, $P(Y = 0) = 0.7^{12} = 0.0138$.
:::
:::

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 30px"}
A mobile operator's count of regional network outages per day has
$$m(t) = 0.5 + 0.3e^{t} + 0.2e^{2t}$$

**Four minutes, in pairs:**

1. What is the distribution of $Y$? (Read it off: $m(t) = \sum_y e^{ty}p(y)$.)

2. Find $E(Y)$ and $V(Y)$ from $m(t)$.

3. Check both directly from $p(y)$.
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 30px"}
1. Match $\sum_y e^{ty}p(y)$ term by term: $p(0) = 0.5$, $p(1) = 0.3$, $p(2) = 0.2$. Uniqueness says that is **the** distribution.

2. $m^{\text{ʹ}}(t) = 0.3e^t + 0.4e^{2t}$, so $E(Y) = 0.7$.
   $m^{\text{ʹʹ}}(t) = 0.3e^t + 0.8e^{2t}$, so $E(Y^2) = 1.1$ and $V(Y) = 1.1 - 0.49 = 0.61$.

::: {.fragment}
3. Directly: $0(0.5) + 1(0.3) + 2(0.2) = 0.7$ and $0 + 0.3 + 4(0.2) = 1.1$. Same numbers — the MGF is just a disciplined way to do those sums.
:::
:::

---

## 📝 Quiz #1: Variance From an MGF {.quiz-question}

Daily chargebacks at a card acquirer have $m(t) = e^{2.5(e^t - 1)}$. What is $V(Y)$?

- [$2.5$]{.correct data-explanation="✅ This is the Poisson MGF with λ = 2.5, and a Poisson variance equals its mean. Via Theorem 3.12: m''(0) − m'(0)² = (2.5² + 2.5) − 2.5² = 2.5."}
- $6.25$
- $8.75$
- $1.58$

---

## 📝 Quiz #2: Name the Distribution {.quiz-question}

Out of 15 export shipments, $Y$ are held at customs, with $m(t) = (0.2e^t + 0.8)^{15}$. What is the distribution of $Y$?

- [Binomial with $n = 15$, $p = 0.2$]{.correct data-explanation="✅ (pe^t + q)^n with p = 0.2 multiplying e^t. By uniqueness Y ~ Bin(15, 0.2), with mean 3."}
- Binomial with $n = 15$, $p = 0.8$
- Poisson with $\lambda = 3$
- Geometric with $p = 0.2$

---

## 📝 Quiz #3: A Linear Change {.quiz-question}

$Y$ has MGF $m_Y(t)$. A fee schedule charges $W = 2Y + 3$ AZN. What is $m_W(t)$?

- [$e^{3t}\,m_Y(2t)$]{.correct data-explanation="✅ E(e^{t(2Y+3)}) = e^{3t} E(e^{(2t)Y}) = e^{3t} m_Y(2t). The shift comes out as a factor; the scale goes inside."}
- $2\,m_Y(t) + 3$
- $e^{2t}\,m_Y(3t)$
- $m_Y(2t) + 3$

---

## 📋 Key Formulas

::: {style="font-size: 30px"}
| | Statement |
|---|---|
| Defs 3.12, 3.13 | $\mu^{\text{ʹ}}_k = E(Y^k)$, $\mu_k = E[(Y - \mu)^k]$ |
| Def 3.14 | $m(t) = E(e^{tY}) = \sum_y e^{ty}p(y)$ |
| Theorem 3.12 | $m^{(k)}(0) = \mu^{\text{ʹ}}_k$ |
| mean, variance | $\mu = m^{\text{ʹ}}(0)$, $\sigma^2 = m^{\text{ʹʹ}}(0) - [m^{\text{ʹ}}(0)]^2$ |
| linear change | $m_{aY+b}(t) = e^{bt}m_Y(at)$ |
| binomial, Poisson | $(pe^t + q)^n$, $e^{\lambda(e^t - 1)}$ |
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- Mean and variance do not determine a distribution; the full sequence of moments (usually) does

- The MGF packs every moment into one function: $\mu^{\text{ʹ}}_k$ is the coefficient of $t^k/k!$

- Differentiate $k$ times and set $t = 0$ to read off $\mu^{\text{ʹ}}_k$

- A count turned into a P&L by $aY + b$ changes its MGF to $e^{bt}m_Y(at)$

- **Uniqueness** is the main use: recognise the MGF, and you know the distribution
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition**

- Exercises at the end of §3.9 — start with 3.145 and 3.146 (the binomial MGF), then 3.149, 3.153 – 3.155

- Exercises 3.156 – 3.157: the linear-change rule, and why a shift leaves the variance alone

**Week 8, Problem Set 1** is open now and closes **Sunday 8 November at 23:59** on WeBWorK, covering §3.9.

**Next class:** 31 October — probability-generating functions, Tchebysheff's theorem, and Chapter 3 in review (Wackerly §3.10–3.11).
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**

📧 sorujov@ada.edu.az \
🏢 Building D, Room D325 \
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- Branches A and B agreed on three moments. Can you build two counts that agree on four?

- Why must every MGF satisfy $m(0) = 1$, and what does that say about the tangent line in today's figure?

- If $Y_1$ and $Y_2$ are independent Poisson claim counts, what might $m_{Y_1 + Y_2}(t)$ be?
:::
