---
title: "Mathematical Statistics"
subtitle: "Moment Generating Functions"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics - Moment Generating Functions"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives

::: {style="font-size: 30px"}
::: {.learning-objectives}
By the end of this lecture, you will be able to:

- **Define and understand** moment-generating functions and their fundamental properties (essential for distribution theory)
- **Derive MGFs** for common probability distributions analytically and numerically (Poisson, exponential, normal)
- **Apply the MGF derivative theorem** to compute moments: mean, variance, and higher moments (critical for risk modeling)
- **Use MGF uniqueness property** to identify probability distributions and prove equivalence (powerful theoretical tool)
- **Interpret MGFs in financial contexts** including applications to option pricing and portfolio risk management
:::
:::

## 📋 Overview

::: {style="font-size:38px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Moments Review** – Raw moments and central moments as foundation
- **Definition of MGF** – Why and how moment-generating functions work
- **Key Theorem** – Computing moments via derivatives of the MGF
- **Examples** – Deriving MGFs for Poisson and other distributions
- **Applications** – Real-world case study using MGF theory
:::
:::
:::

## 📖 Definition: Moments {.larger}

::: {.callout-note}
## 📝 Definition 1: Raw Moments

The **k-th moment** of a random variable $Y$ taken about the origin is defined to be $E(Y^k)$ and is denoted by $\mu_k'$.

::: {.incremental}
- **First moment** ($\mu_1' = E(Y)$) – Expected value (mean return of an asset)
- **Second moment** ($\mu_2' = E(Y^2)$) – Used to compute variance (volatility measure)
- **Third moment** ($\mu_3'$) – Related to skewness (asymmetry of return distribution)
- **Fourth moment** ($\mu_4'$) – Related to kurtosis (tail risk, probability of extreme events)
:::
:::

## 📖 Definition: Central Moments {.larger}

::: {.callout-note}
## 📝 Definition 2: Central Moments

The **k-th central moment** of a random variable $Y$ is defined to be $E[(Y-\mu)^k]$ and is denoted by $\mu_k$.

::: {.incremental}
- **First central moment** ($\mu_1 = 0$) – Always zero by definition
- **Second central moment** ($\mu_2 = \sigma^2$) – Variance (risk measure in finance)
- **Third central moment** ($\mu_3$) – Measures skewness (asymmetric tail behavior)
- **Fourth central moment** ($\mu_4$) – Measures kurtosis (fat tails, black swan events)
:::
:::

. . .

::: {.callout-important}
## 💡 Financial Insight

Investors care deeply about higher moments: positive skewness indicates higher probability of large gains, while high kurtosis signals increased tail risk (crash probability).
:::

## 📖 Uniqueness Property {.large}

::: {.callout-note}
## 🔑 Remark 1: Uniqueness of Moments

If two random variables $Y$ and $Z$ possess finite moments with
$$
\mu'_{iY} = \mu'_{iZ} \quad \text{for all } i \in \mathbb{Z}^+
$$
then $Y$ and $Z$ have **identical probability distributions**.

**Financial Application**: If two portfolio strategies produce identical moments at all orders, they are statistically equivalent in terms of return distribution.
:::

## 📖 Definition: Moment-Generating Function {.larger}

::: {.callout-note}
## 📝 Definition 3: MGF

The **moment-generating function** $m(t)$ for a random variable $Y$ is defined to be 
$$
m(t) = E(e^{tY})
$$

We say that an MGF exists if there exists a positive constant $b$ such that $m(t)$ is finite for $|t| \leq b$.

**Why "moment-generating"?** The derivatives of $m(t)$ evaluated at $t=0$ give us the moments!
:::

## 🧮 Why It Generates Moments (Part 1) {.smaller}

::: {.callout-tip}
## 🔍 Key Question

Why is $E(e^{tY})$ called the moment-generating function?
:::

. . .

From the Taylor series expansion for $e^{ty}$:
$$
e^{ty} = 1 + ty + \frac{(ty)^2}{2!} + \frac{(ty)^3}{3!} + \frac{(ty)^4}{4!} + \cdots
$$

. . .

Taking the expectation (assuming $\mu'_k$ is finite for all $k$):
$$
E(e^{tY}) = E\left[1 + tY + \frac{(tY)^2}{2!} + \frac{(tY)^3}{3!} + \cdots\right]
$$

## 🧮 Why It Generates Moments (Part 2) {.smaller}

$$
E(e^{tY}) = 1 + tE(Y) + \frac{t^2}{2!}E(Y^2) + \frac{t^3}{3!}E(Y^3) + \cdots
$$

. . .

Substituting the moment notation:
$$
\boxed{m(t) = 1 + t\mu'_1 + \frac{t^2}{2!}\mu'_2 + \frac{t^3}{3!}\mu'_3 + \cdots}
$$

. . .

::: {.callout-important}
## 🎯 Key Insight

The coefficient of $\frac{t^k}{k!}$ in the series expansion of $m(t)$ is exactly $\mu'_k$, the k-th moment!
:::

## 🔑 Key Applications of MGF {.large}

::: {.callout-note}
## 💡 Remark 2: Two Important Applications

::: {.incremental}
1. **Finding Moments**: If $E(e^{tY})$ exists, we can find any moment for $Y$ by differentiation (easier than direct integration for complex distributions)

2. **Uniqueness Property**: If the moment-generating functions for two random variables $Y$ and $Z$ are equal, then $Y$ and $Z$ must have the **same probability distribution** (powerful tool for proving distributional equivalence)
:::
:::

. . .

**Financial Application**: MGFs are used in mathematical finance to derive option pricing formulas and to prove that transformed random variables (e.g., log-returns) follow specific distributions.

## 📐 Theorem: Computing Moments via MGF {.large}

::: {.callout-important}
## 🎓 Theorem 1: Derivative Formula

If $m(t)$ exists, then for any positive integer $k$:
$$
\boxed{\frac{d^km(t)}{dt^k} \bigg|_{t=0} = m^{(k)}(0) = \mu_k'}
$$

**Interpretation**: The k-th derivative of $m(t)$ evaluated at $t=0$ gives the k-th moment.
:::

. . .

::: {.callout-tip}
## 💼 Practical Use

Instead of computing $\int y^k p(y) dy$ directly (which can be messy), we:
1. Find $m(t) = E(e^{tY})$
2. Differentiate $k$ times
3. Set $t=0$
:::

## 📌 Example 1: MGF of Poisson Distribution {.large}

::: {.callout-note}
## 🎯 Problem Setup

Find the moment-generating function $m(t)$ for a Poisson distributed random variable with mean $\lambda$.

**Context**: Poisson distributions model counts of events (e.g., number of trades per minute, customer arrivals, insurance claims).
:::

## 🧮 Example 1: Solution (Part 1) {.smaller}

Starting from the definition:
$$
m(t) = E(e^{tY}) = \sum_{y=0}^{\infty} e^{ty} p(y) = \sum_{y=0}^{\infty} e^{ty} \frac{\lambda^y e^{-\lambda}}{y!}
$$

. . .

Combine the exponential terms:
$$
m(t) = \sum_{y=0}^{\infty} \frac{(\lambda e^t)^y e^{-\lambda}}{y!} = e^{-\lambda} \sum_{y=0}^{\infty} \frac{(\lambda e^t)^y}{y!}
$$

. . .

Recognize that $\sum_{y=0}^{\infty} \frac{a^y}{y!} = e^a$:
$$
m(t) = e^{-\lambda} \cdot e^{\lambda e^t}
$$

## 🧮 Example 1: Solution (Part 2) {.smaller}

Final result:
$$
\boxed{m(t) = e^{\lambda(e^t - 1)}}
$$

. . .

::: {.callout-tip}
## 🔍 Verification Technique

We multiplied and divided by $e^{\lambda e^t}$ to recognize that:
$$
\sum_{y=0}^{\infty} \frac{(\lambda e^t)^y e^{-\lambda e^t}}{y!} = 1
$$
This is the sum of a Poisson PMF with parameter $\lambda e^t$, which equals 1.
:::

. . .

**Financial Context**: This MGF is used in modeling high-frequency trading events and queue theory in financial markets.

## 📌 Example 2: Mean and Variance via MGF {.large}

::: {.callout-note}
## 🎯 Problem

Use the MGF from Example 1 to find the mean $\mu$ and variance $\sigma^2$ for the Poisson random variable.

**Method**: Apply Theorem 1 by computing $m'(0)$ and $m''(0)$.
:::

## 🧮 Example 2: Solution (Part 1) {.smaller}

Recall $m(t) = e^{\lambda(e^t-1)}$. 

. . .

**First derivative** (for the mean):
$$
m'(t) = \frac{d}{dt}\left[e^{\lambda(e^t-1)}\right] = e^{\lambda(e^t-1)} \cdot \lambda e^t
$$

. . .

**Second derivative** (for the second moment):
$$
m''(t) = \frac{d}{dt}\left[e^{\lambda(e^t-1)} \cdot \lambda e^t\right]
$$
$$
= e^{\lambda(e^t-1)} \cdot (\lambda e^t)^2 + e^{\lambda(e^t-1)} \cdot \lambda e^t
$$

## 🧮 Example 2: Solution (Part 2) {.smaller}

Evaluate at $t=0$:

. . .

**Mean**:
$$
\mu = m'(0) = e^{\lambda(e^0-1)} \cdot \lambda e^0 = e^0 \cdot \lambda = \boxed{\lambda}
$$

. . .

**Second moment**:
$$
\mu_2' = m''(0) = e^0 \cdot \lambda^2 + e^0 \cdot \lambda = \lambda^2 + \lambda
$$

. . .

**Variance**:
$$
\sigma^2 = E(Y^2) - \mu^2 = \mu_2' - \mu^2 = \lambda^2 + \lambda - \lambda^2 = \boxed{\lambda}
$$

. . .

::: {.callout-important}
## 🎯 Key Result for Poisson Distribution

For Poisson($\lambda$): Mean = Variance = $\lambda$
:::

## 🎮 Interactive: Visualizing the Poisson MGF {.smaller}

::: {style="font-size: 0.8em;"}

**Explore MGF Properties:** Adjust $\lambda$ to see how the MGF $m(t) = e^{\lambda(e^t - 1)}$ changes, and observe how derivatives at $t=0$ give moments.

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof lambda_mgf = Inputs.range([0.5, 20], {
  value: 20, 
  step: 0.5, 
  label: "λ (Poisson parameter):"
})

// Compute moments from MGF derivatives at t=0
mean_from_mgf = lambda_mgf
variance_from_mgf = lambda_mgf

html`<div style="font-size: 14px; line-height: 1.6;">
  <p><strong>MGF Formula:</strong><br>
  m(t) = exp(λ(e<sup>t</sup> - 1))</p>
  
  <p><strong>Moments from MGF:</strong><br>
  m'(0) = Mean = ${mean_from_mgf.toFixed(2)}<br>
  m''(0) - [m'(0)]² = Var = ${variance_from_mgf.toFixed(2)}</p>
  
  <p><strong>Key Insight:</strong><br>
  At t = 0: m(0) = 1<br>
  Derivatives at t = 0 give moments!</p>
</div>`
```

**Pedagogical Value:**  

- Visualize MGF curve m(t)

- See first derivative m'(t)

- Understand t=0 is where moments are extracted

- Blue curve = MGF, Orange = derivative

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// MGF and its derivatives for Poisson
function poissonMGF(t, lambda) {
  return Math.exp(lambda * (Math.exp(t) - 1));
}

function poissonMGF_derivative1(t, lambda) {
  return lambda * Math.exp(t) * Math.exp(lambda * (Math.exp(t) - 1));
}

// Generate data for plotting
t_range = d3.range(-1, 1, 0.02)
mgf_data = t_range.map(t => ({
  t: t,
  mgf: poissonMGF(t, lambda_mgf),
  derivative1: poissonMGF_derivative1(t, lambda_mgf)
}))

// Values at t=0
mgf_at_0 = poissonMGF(0, lambda_mgf)
deriv1_at_0 = poissonMGF_derivative1(0, lambda_mgf)

// Dynamic y-axis with logarithmic scale
y_min_all = Math.min(...mgf_data.map(d => Math.min(d.mgf, d.derivative1)))
y_max_all = Math.max(...mgf_data.map(d => Math.max(d.mgf, d.derivative1)))

// Use log scale if range is large
use_log_scale = (y_max_all / y_min_all) > 100
y_scale_type = use_log_scale ? "log" : "linear"

// Set domain with padding - no truncation
y_min_domain = use_log_scale ? Math.max(0.001, y_min_all * 0.1) : Math.max(0, y_min_all - (y_max_all - y_min_all) * 0.1)
y_max_domain = use_log_scale ? y_max_all * 10 : y_max_all * 1.15

Plot.plot({
  width: 800,
  height: 450,
  marginLeft: 80,
  marginBottom: 50,
  x: {
    label: "t (MGF parameter)",
    grid: true,
    domain: [-1, 1]
  },
  y: {
    label: "Function Value" + (use_log_scale ? " (log scale)" : ""),
    grid: true,
    type: y_scale_type,
    domain: [y_min_domain, y_max_domain],
    ticks: use_log_scale ? 6 : undefined
  },
  marks: [
    // MGF curve
    Plot.line(mgf_data, {
      x: "t", 
      y: "mgf", 
      stroke: "steelblue", 
      strokeWidth: 3,
      tip: true
    }),
    // First derivative
    Plot.line(mgf_data, {
      x: "t", 
      y: "derivative1", 
      stroke: "orange", 
      strokeWidth: 2.5, 
      strokeDasharray: "5,5",
      tip: true
    }),
    // t=0 vertical line
    Plot.ruleX([0], {
      stroke: "red", 
      strokeWidth: 2.5,
      strokeDasharray: "3,3"
    }),
    // Point at m(0)
    Plot.dot([{t: 0, y: mgf_at_0}], {
      x: "t", 
      y: "y", 
      fill: "steelblue", 
      r: 8,
      stroke: "white",
      strokeWidth: 2
    }),
    // Point at m'(0)
    Plot.dot([{t: 0, y: deriv1_at_0}], {
      x: "t", 
      y: "y", 
      fill: "orange", 
      r: 8,
      stroke: "white",
      strokeWidth: 2
    }),
    Plot.ruleY([0])
  ],
  caption: html`
    <div style="font-size: 14px; text-align: center; margin-top: 10px; line-height: 1.8;">
      <div style="margin-bottom: 5px;">
        <span style="color: steelblue; font-weight: bold; font-size: 16px;">━━━</span> 
        <strong>m(t)</strong> = exp(λ(e<sup>t</sup>−1)) [MGF] &nbsp;&nbsp;|&nbsp;&nbsp; 
        <span style="color: orange; font-weight: bold; font-size: 16px;">━ ━ ━</span> 
        <strong>m'(t)</strong> [First Derivative] &nbsp;&nbsp;|&nbsp;&nbsp; 
        <span style="color: red; font-weight: bold;">┊</span> <strong>t = 0</strong>
      </div>
      <div style="font-size: 13px; color: #555;">
        <span style="color: steelblue; font-size: 18px;">●</span> m(0) = ${mgf_at_0.toFixed(3)} &nbsp;&nbsp;|&nbsp;&nbsp; 
        <span style="color: orange; font-size: 18px;">●</span> m'(0) = ${deriv1_at_0.toFixed(3)} = <strong>Mean</strong>
      </div>
    </div>
  `
})
```

:::

:::

:::


## 📌 Example 3: MGF of Exponential Distribution {.large}

::: {.callout-note}
## 🎯 Problem

Find the moment-generating function for an Exponential($\lambda$) random variable with PDF $f(y) = \lambda e^{-\lambda y}$ for $y > 0$.

**Context**: Exponential distributions model waiting times (e.g., time until next trade, customer service duration).
:::

## 🧮 Example 3: Solution {.smaller}

Starting from the definition:
$$
m(t) = E(e^{tY}) = \int_0^{\infty} e^{ty} \lambda e^{-\lambda y} dy
$$

. . .

Combine exponentials:
$$
m(t) = \lambda \int_0^{\infty} e^{(t-\lambda)y} dy
$$

::: {.columns}
::: {.column width="50%"}
For $t < \lambda$, evaluate the integral:
$$
m(t) = \lambda \left[\frac{e^{(t-\lambda)y}}{t-\lambda}\right]_0^{\infty} = \lambda \cdot \frac{0 - 1}{t-\lambda}
$$
:::

::: {.column width="50%" .fragment}
Final result:
$$
\boxed{m(t) = \frac{\lambda}{\lambda - t} \quad \text{for } t < \lambda}
$$

**Practical Application**: This MGF is fundamental in queueing theory and option pricing models.

:::
:::


## 📌 Example 4: Using Exponential MGF to Find Moments {.large}

::: {.callout-note}
## 🎯 Problem

Use the Exponential MGF $m(t) = \frac{\lambda}{\lambda - t}$ to find $E(Y)$, $E(Y^2)$, and $\text{Var}(Y)$.
:::

## 🧮 Example 4: Solution (Part 1) {.smaller}

**First derivative** (for the mean):
$$
m'(t) = \frac{d}{dt}\left[\frac{\lambda}{\lambda - t}\right] = \frac{\lambda}{(\lambda - t)^2}
$$

. . .

**Mean**:
$$
\mu = m'(0) = \frac{\lambda}{\lambda^2} = \boxed{\frac{1}{\lambda}}
$$

::: {.columns}
::: {.column width="50%"}

**Second derivative** (for the second moment):
$$
m''(t) = \frac{d}{dt}\left[\frac{\lambda}{(\lambda - t)^2}\right] = \frac{2\lambda}{(\lambda - t)^3}
$$
:::

::: {.column width="50%" .fragment}

**Second moment**:
$$
E(Y^2) = m''(0) = \frac{2\lambda}{\lambda^3} = \boxed{\frac{2}{\lambda^2}}
$$

:::
:::

## 🧮 Example 4: Solution (Part 2) {.smaller}

**Variance**:
$$
\sigma^2 = E(Y^2) - [E(Y)]^2 = \frac{2}{\lambda^2} - \frac{1}{\lambda^2} = \boxed{\frac{1}{\lambda^2}}
$$

. . .

::: {.callout-important}
## 🎯 Key Result for Exponential Distribution

For Exponential($\lambda$):
- Mean: $E(Y) = \frac{1}{\lambda}$
- Variance: $\text{Var}(Y) = \frac{1}{\lambda^2}$
- **Memoryless property** (unique among continuous distributions)
:::

. . .

**Financial Note**: In option pricing, exponential MGFs appear in Lévy process models for asset price dynamics.

## 📌 Example 5: MGF of Normal Distribution {.large}

::: {.callout-note}
## 🎯 Problem Setup

Derive the MGF for a Normal($\mu, \sigma^2$) random variable.

**Context**: The normal distribution is fundamental in finance for modeling returns, though empirical returns often exhibit fat tails.
:::

## 🧮 Example 5: Solution {.smaller}

For $Y \sim N(\mu, \sigma^2)$, the MGF is:
$$
m(t) = E(e^{tY})
$$

. . .

Using properties of the normal distribution and completing the square in the exponent:
$$
m(t) = \exp\left(\mu t + \frac{\sigma^2 t^2}{2}\right)
$$

. . .

::: {.callout-important}
## 🎯 Key Result for Normal Distribution

For $Y \sim N(\mu, \sigma^2)$:
$$
\boxed{m(t) = e^{\mu t + \frac{\sigma^2 t^2}{2}}}
$$

**Insight**: All moments can be extracted from this elegant closed-form MGF!
:::

. . .

**Taking derivatives** at $t=0$:
- $m'(0) = \mu$ (recovers the mean)
- $m''(0) = \mu^2 + \sigma^2$ (gives $\text{Var}(Y) = \sigma^2$)

## 📝 Quiz #1: MGF Definition {.quiz-question}

Which of the following correctly defines the moment-generating function?

- [$m(t) = E(e^{tY})$]{.correct data-explanation="✅ Correct! This is the standard definition. The MGF evaluated at t=0 gives the MGF value of 1, and derivatives at t=0 produce the moments."}
- $m(t) = E(Y e^t)$
- $m(t) = e^{E(tY)}$
- $m(t) = E(e^Y)$

## 📝 Quiz #2: Computing Moments from MGF {.quiz-question}

If the MGF of a distribution is $m(t) = e^{3t + 2t^2}$, what is $E(Y)$?

- [3]{.correct data-explanation="✅ Correct! Take the derivative and evaluate at t=0 to get 3."}
- $1$
- $4$
- Cannot determine from this information

## 📝 Quiz #3: MGF Uniqueness {.quiz-question}

What does it mean if two random variables have identical MGFs?

- [They have the same probability distribution]{.correct data-explanation="✅ Correct! The uniqueness theorem states that if m_Y(t) = m_Z(t) for all t in some interval, then Y and Z are identically distributed."}
- They have the same mean
- They have the same variance
- They are independent

## 💰 Case Study: Portfolio Return Analysis (Real Data) {.smaller}

::: {style="font-size:28px"}
::: {.columns}
::: {.column width="50%"}
::: {.callout-note}
## 📈 Investment Problem

**Context**: A portfolio manager wants to understand whether returns follow a normal distribution and use MGF theory to assess risk.

**Key Questions**:

- What are the moments of portfolio daily returns?
- Can we estimate the MGF from empirical data?
- How well does the normal MGF fit observed returns?
:::
:::

::: {.column width="50%" .fragment}
::: {.callout-tip}
## 📊 Data Source

We analyze **S&P 500 ETF (SPY) daily returns** from **2020-2025**.

**Source**: Yahoo Finance API (via quantmod)

**Period**: January 2020 - November 2025

**Data Quality**: Adjusted closing prices

**Verification**: Cross-checked with CRSP database
:::
:::
:::
:::

## 📊 Data Loading and Moments Calculation {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Load required libraries
library(quantmod)
library(tidyverse)
library(knitr)
library(moments)

# Download S&P 500 ETF data
getSymbols("SPY", 
           from = "2020-01-01",
           to = Sys.Date(),
           auto.assign = TRUE)

# Calculate daily returns (percentage)
spy_returns <- dailyReturn(SPY, 
                           type = "log") * 100

# Convert to data frame
spy_df <- data.frame(
  Date = index(spy_returns),
  Return = as.numeric(spy_returns)
)

# Display first few observations
head(spy_df, 5) %>% 
  kable(digits = 3, 
        caption = "SPY Daily Log Returns (%)")
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

#Compute first four moments
mu <- mean(spy_df$Return)
sigma <- sd(spy_df$Return)
skew <- skewness(spy_df$Return)
kurt <- kurtosis(spy_df$Return)

#Display summary statistics
cat(sprintf("Mean (μ):        %.4f%%", mu))
cat(sprintf("Std Dev (σ):     %.4f%%", sigma))
cat(sprintf("Skewness:        %.4f", skew))
cat(sprintf("Kurtosis:        %.4f", kurt))
cat(sprintf("Excess Kurtosis: %.4f", 
            kurt - 3))
cat("Sample size:", nrow(spy_df), "days")
```

:::
:::
:::

## 📊 MGF Estimation from Empirical Data {.smaller}

::: {style="font-size:26px"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Function to estimate MGF numerically from data
estimate_mgf <- function(data, t) {
  mean(exp(t * data))
}

# Compute empirical MGF at various t values
t_vals <- seq(-0.3, 0.3, by = 0.02)
empirical_mgf <- sapply(t_vals, function(t) 
  estimate_mgf(spy_df$Return, t))

# Compute normal MGF assuming normal distribution
normal_mgf <- sapply(t_vals, function(t) 
  exp(mu * t + 0.5 * sigma^2 * t^2))

# Create comparison table
mgf_comparison <- data.frame(
  t = t_vals,
  Empirical_MGF = empirical_mgf,
  Normal_MGF = normal_mgf,
  Difference = abs(empirical_mgf - normal_mgf)
)

# Display comparison
head(mgf_comparison, 10) %>%
  kable(digits = 6, 
        caption = "Empirical vs Normal MGF")
```

:::

## 📊 Visualization: Empirical vs Normal MGF {.smaller .center}

::: {style="font-size:26px"}

```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: true
#| fig-width: 11
#| fig-height: 6
#| fig-align: center

library(ggplot2)

# Create visualization comparing MGFs
ggplot(mgf_comparison, aes(x = t)) +
  geom_line(aes(y = Empirical_MGF, color = "Empirical MGF"), 
            linewidth = 1.2) +
  geom_line(aes(y = Normal_MGF, color = "Normal MGF"), 
            linewidth = 1.2, linetype = "dashed") +
  geom_point(aes(y = Empirical_MGF), color = "steelblue", size = 2, alpha = 0.6) +
  scale_color_manual(values = c("Empirical MGF" = "steelblue", 
                                "Normal MGF" = "red")) +
  labs(title = "Empirical vs Normal MGF: S&P 500 Returns",
       subtitle = sprintf("μ = %.2f%% | σ = %.2f%% | Excess Kurt = %.2f", 
                         mu, sigma, kurt - 3),
       x = "t (parameter)",
       y = "m(t) = E(e^tY)",
       color = "MGF Type") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11),
        legend.position = "topright")
```

:::

## 📊 Distribution of Returns {.smaller}

::: {style="font-size:26px"}

```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: true
#| fig-width: 11
#| fig-height: 6
#| fig-align: center

# Create histogram with normal overlay
ggplot(spy_df, aes(x = Return)) +
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 60, fill = "steelblue", alpha = 0.6) +
  geom_density(color = "darkblue", linewidth = 1) +
  stat_function(fun = dnorm, 
                args = list(mean = mu, sd = sigma),
                color = "red", linewidth = 1.2, linetype = "dashed",
                aes(label = "Normal Fit")) +
  geom_vline(xintercept = mu, color = "green", 
             linetype = "solid", linewidth = 1) +
  annotate("text", x = mu + 0.3, y = 0.4, 
           label = sprintf("μ = %.2f%%", mu), 
           color = "green", size = 3.5) +
  labs(title = "S&P 500 Daily Return Distribution (2020-2025)",
       subtitle = "Empirical Distribution vs Normal Approximation",
       x = "Daily Return (%)",
       y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11))
```

:::

## 💡 Case Study Insights {.large}

::: {style="font-size: 30px"}

::: {.callout-important}
## 🎯 Key Findings

1. **Negative skewness** (≈ -0.30 to -0.50) indicates asymmetric downside risk in equity returns

2. **Excess kurtosis** (≈ 3-5 depending on period) shows **fat tails** - more extreme events than normal distribution predicts

3. **MGF divergence**: Empirical MGF deviates from normal MGF at large |t| values, confirming non-normality

4. **Risk modeling**: Normal MGF underestimates tail probabilities; alternative models (Lévy processes) needed for accurate VaR
:::

:::

. . .

::: {.callout-tip}
## 💼 Practical Application

For portfolio construction, using MGF-based risk measures beyond variance captures tail risk more accurately. This is why practitioners use Value-at-Risk (VaR) and Expected Shortfall (ES).
:::

## 📝 Summary {.smaller}

::: {.summary-box}
✅ Key Takeaways

**Moment-generating functions characterize distributions completely**: The MGF uniquely identifies a probability distribution; if two random variables share the same MGF, they are identically distributed

**MGFs simplify moment calculations**: Instead of direct integration, compute $m^{(k)}(0)$ to find the $k^{th}$ moment - often algebraically simpler and more numerically stable

**Common MGF results**: Poisson: $e^{\lambda(e^t-1)}$, Exponential: $\frac{\lambda}{\lambda-t}$, Normal: $e^{\mu t + \frac{\sigma^2 t^2}{2}}$

**Moment relationships**: Mean = $m'(0)$, Variance = $m''(0) - [m'(0)]^2$, and higher moments follow from higher derivatives

**Financial applications**: MGFs are essential in derivatives pricing, risk management via characteristic functions, and proving convergence results (e.g., Central Limit Theorem)
:::

## 📚 Practice Problems

::: {.callout-tip}
## 📝 Homework Problems

**Problem 1**: Find the MGF for a geometric distribution with success probability $p$. Use it to derive $E(Y)$ and $\text{Var}(Y)$.

**Problem 2**: For the Poisson MGF $m(t) = e^{\lambda(e^t-1)}$, compute the third derivative $m'''(t)$ and evaluate at $t=0$ to find the third moment $E(Y^3)$.

**Problem 3**: Show that if $Y \sim N(\mu, \sigma^2)$ with MGF $m(t) = e^{\mu t + \frac{\sigma^2 t^2}{2}}$, then $E(Y^4) = \mu^4 + 6\mu^2\sigma^2 + 3\sigma^4$.

**Problem 4**: Download 3 years of daily returns for an asset. Estimate the empirical MGF and compare to the MGF of a fitted normal distribution. Discuss where they diverge.

**Problem 5**: If $X \sim \text{Poisson}(\lambda)$ and $Y = 2X + 3$, find the MGF of $Y$ using the linear transformation properties and verify your answer.
:::

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
### 📬 Contact Information:

**Samir Orujov, PhD**  
Assistant Professor  
School of Business  
ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)  
🏢 Office: D312  
⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
### 📅 Next Class:

**Topic**: Characteristic Functions & Lévy Processes

**Reading**: Chapter 5, Sections 5.1-5.3

**Preparation**: Review MGF properties

### ⏰ Reminders:

✅ Practice problems due next week  
✅ Study MGF derivations thoroughly  
✅ Work hard!
:::
:::

## ❓ Questions? {.center}

::: {.callout-note}
## 💬 Open Discussion (5 minutes)

- **Questions about MGF derivations?** Let's work through another example together

- **Confused about the Poisson MGF?** We'll break down the exponential manipulation step-by-step

- **Want to see more financial applications?** We can explore characteristic functions in option pricing

- **Need help with practice problems?** Office hours available by appointment
:::
