---
title: "Mathematical Statistics"
subtitle: "The Gamma Probability Distribution"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics - Gamma Distribution"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::: {style="font-size: 30px"}
::: {.learning-objectives}
By the end of this lecture, you will be able to:

- Define the gamma distribution and identify its shape parameter ($\alpha$) and scale parameter ($\beta$), and understand how they affect distribution characteristics

- Compute probabilities, expected values, and variances for gamma-distributed random variables, including exponential and chi-square special cases

- Apply the exponential distribution to model waiting times and understand the memoryless property in reliability and financial risk contexts

- Use the chi-square distribution ($\chi^2$) for statistical inference, hypothesis testing, and volatility modeling in finance

- Solve real-world problems involving skewed distributions in insurance claims, component lifetimes, income distributions, and water demand forecasting
:::
:::

## 📋 Overview {.smaller}

::: {style="font-size:38px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Skewed Distributions** – Understanding asymmetric data patterns and when they arise

- **Gamma Distribution** – The general form with shape and scale parameters

- **Chi-Square Distribution** – A special case with applications in hypothesis testing and variance estimation

- **Exponential Distribution** – Modeling lifetimes and the memoryless property

- **Applications** – Insurance claims, reliability engineering, water demand, income distributions, and financial risk modeling
:::
:::
:::

## 📖 Definition: Skewed Distribution {.smaller}

::: {.callout-note}
## 📝 Concept: Skewed Distributions

A **skewed distribution** occurs when data in a chart lean either to the left or right side of the scale, resulting in a nonsymmetrical curve.

**Key Characteristics:**

::: {.incremental}
- **Asymmetry**: The left side is shaped differently than the right side

- **Tail behavior**: One tail is longer than the other, indicating more extreme values on that side

- **Difference from normal**: Unlike the Gaussian (normal) distribution, which is symmetric around the mean with zero skewness, skewed distributions have the mean, median, and mode at different locations

- **Real-world prevalence**: Many phenomena exhibit skewness, including income distributions, insurance claims, component lifetimes, and asset returns
:::
:::

**Financial Context:** Income distributions are typically right-skewed (positively skewed) with a long tail extending toward higher incomes, while most people cluster at lower to moderate income levels [web:17].

## 📊 Types of Skewness

::: {.columns}

::: {.column width="50%"}
**Right-Skewed (Positive Skew)**

- Tail extends to the right
- Mean > Median > Mode
- Common in: income, insurance claims, asset returns

**Examples:**

- Executive compensation
- Insurance loss amounts
- Real estate prices
- Time to equipment failure
:::

::: {.column width="50%"}
**Left-Skewed (Negative Skew)**

- Tail extends to the left
- Mean < Median < Mode
- Common in: age at death, test scores with ceiling effects

**Examples:**

- Human lifespan (peaks at 75-80 years)
- Product quality ratings (most cluster at high ratings)
- Time remaining until retirement
:::

:::

## 📌 Example 1: Human Lifespan Distribution {.smaller}

**Problem:** The average human life span chart skews left. If the chart shows values from 1 to 100 (representing years of life), explain the distribution characteristics.

**Analysis:**

The data shows that most people live to around 75 to 80 years old, which means:

- **Peak location**: The distribution's peak (mode) is closer to the right of the chart (near 75-80 years)

- **Tail direction**: The chart's tail is longer on its left side because the values around 75 and 80 are closer to 100 than to 1

- **Asymmetry**: Fewer people die at very young ages (due to modern medicine and sanitation), creating a shorter right tail, while infant mortality and premature deaths create a longer left tail

**Interpretation:** This left-skewed pattern reflects that in developed countries, medical advances have pushed most deaths toward older ages, with exceptional cases of early death creating the left tail. In contrast, there's a biological upper limit on lifespan, creating a natural boundary on the right side.

## 📖 Definition: Gamma Probability Distribution {.smaller}

::: {.callout-note}
## 📝 Definition 1: Gamma Distribution

A random variable $Y$ is said to have a **gamma distribution** with parameters $\alpha > 0$ and $\beta > 0$ if and only if the probability density function (pdf) of $Y$ is:

$$f(y) = \begin{cases}
\frac{y^{\alpha-1}e^{-y/\beta}}{\beta^\alpha\Gamma(\alpha)}, & 0 \leq y < \infty, \\
0, & \text{elsewhere},
\end{cases}$$

where the **gamma function** is defined as:

$$\Gamma(\alpha) = \int_{0}^{\infty} y^{\alpha-1}e^{-y} \,dy$$

**Key Properties of the Gamma Function:**

- $\Gamma(1) = 1$

- $\Gamma(\alpha) = (\alpha - 1)\Gamma(\alpha - 1)$ for any $\alpha > 1$ (recursive property)

- $\Gamma(n) = (n - 1)!$ for positive integers $n$
:::

## 🔍 Understanding Gamma Parameters

::: {.columns}

::: {.column width="50%"}
::: {.callout-tip}
## 📐 Shape Parameter: $\alpha$

**Effect on Distribution:**

- Controls the **shape** of the distribution
- $\alpha < 1$: J-shaped (decreasing from infinity at $y = 0$)
- $\alpha = 1$: Exponential distribution
- $\alpha > 1$: Unimodal with peak shifting right as $\alpha$ increases
- Larger $\alpha$ makes distribution more symmetric (approaches normal)

**Financial Application:** In reliability engineering, $\alpha$ represents the number of stages or phases before failure [web:13].
:::
:::

::: {.column width="50%"}
::: {.callout-tip}
## 📏 Scale Parameter: $\beta$

**Effect on Distribution:**

- Controls the **scale** or spread of the distribution
- Larger $\beta$ stretches distribution to the right
- Smaller $\beta$ compresses distribution toward zero
- Does not affect shape, only the x-axis scaling

**Financial Application:** In insurance, $\beta$ scales the claim amounts while maintaining the underlying claim frequency pattern represented by $\alpha$ [web:13].
:::
:::

:::

## 🧮 Theorem: Mean and Variance of Gamma Distribution

::: {.callout-important}
## Theorem 1: Expected Value and Variance

If $Y$ has a gamma distribution with parameters $\alpha$ and $\beta$, then:

$$\boxed{\mu = E(Y) = \alpha \beta}$$

and

$$\boxed{\sigma^2 = V(Y) = \alpha \beta^2}$$

**Derivation Insight:** These formulas follow from integration by parts applied to the definition of expected value and variance using the gamma density function.

**Important Note:** Except when $\alpha = 1$ (exponential distribution), it is generally impossible to obtain areas under the gamma density function by direct integration. We typically use:

- Statistical software (R, Python)
- Numerical integration methods
- Tables for chi-square distribution (when applicable)
- Online calculators/applets
:::

## 🎮 Interactive: Gamma Distribution Explorer {.smaller}

::: {style="font-size: 0.8em;"}

**Explore Gamma Parameters:** Adjust $\alpha$ (shape) and $\beta$ (scale) to see their effects on the distribution.

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof alpha = Inputs.range([0.5, 10], {
  value: 2, 
  step: 0.5, 
  label: "α (shape):"
})

viewof beta = Inputs.range([0.5, 5], {
  value: 2, 
  step: 0.5, 
  label: "β (scale):"
})

mean_gamma = alpha * beta
variance_gamma = alpha * beta * beta
sd_gamma = Math.sqrt(variance_gamma)

md`**Current Parameters:**  
α = ${alpha.toFixed(1)}  
β = ${beta.toFixed(1)}  

**Statistics:**  
Mean (μ) = ${mean_gamma.toFixed(2)}  
Variance (σ²) = ${variance_gamma.toFixed(2)}  
Std Dev (σ) = ${sd_gamma.toFixed(2)}`
```

**Observations:**  
- Increase α: peak moves right, distribution becomes more symmetric
- Increase β: distribution stretches horizontally
- When α = 1: exponential distribution

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// Gamma function approximation using Stirling's approximation
function gammaFunc(z) {
  if (z === 1) return 1;
  if (z === 0.5) return Math.sqrt(Math.PI);
  // Use Stirling's approximation for larger values
  return Math.sqrt(2 * Math.PI / z) * Math.pow((z / Math.E), z);
}

// Gamma PDF function
function gammaPDF(y, alpha, beta) {
  if (y <= 0) return 0;
  const numerator = Math.pow(y, alpha - 1) * Math.exp(-y / beta);
  const denominator = Math.pow(beta, alpha) * gammaFunc(alpha);
  return numerator / denominator;
}

// Generate data
y_range = d3.range(0.01, Math.max(alpha * beta + 4 * Math.sqrt(alpha * beta * beta), 20), 0.1)

gamma_data = y_range.map(y => ({
  y: y,
  pdf: gammaPDF(y, alpha, beta)
}))

// Find mode (for alpha > 1)
mode_y = alpha > 1 ? (alpha - 1) * beta : 0

Plot.plot({
  width: 800,
  height: 450,
  marginLeft: 60,
  marginBottom: 50,
  x: {
    label: "Y (random variable)",
    grid: true,
    domain: [0, Math.max(alpha * beta + 4 * Math.sqrt(alpha * beta * beta), 20)]
  },
  y: {
    label: "Probability Density f(y)",
    domain: [0, d3.max(gamma_data, d => d.pdf) * 1.1]
  },
  marks: [
    Plot.line(gamma_data, {
      x: "y", 
      y: "pdf", 
      stroke: "steelblue", 
      strokeWidth: 3
    }),
    Plot.areaY(gamma_data, {
      x: "y", 
      y: "pdf", 
      fill: "steelblue", 
      opacity: 0.2
    }),
    Plot.ruleX([mean_gamma], {
      stroke: "red", 
      strokeWidth: 2, 
      strokeDasharray: "5,5"
    }),
    Plot.ruleY([0])
  ],
  caption: html`<span style="color: steelblue; font-weight: bold;">━━</span> Gamma(α=${alpha.toFixed(1)}, β=${beta.toFixed(1)}) | 
    <span style="color: red;">┊</span> Mean = ${mean_gamma.toFixed(2)}`
})
```

:::

:::

:::

## 📌 Example 2: Computer Response Time {.smaller}

**Problem:** The response times on an online computer terminal have approximately a gamma distribution with mean four seconds and variance eight seconds.

a) Write the probability density function for the response times.

b) What is the probability that the response time on the terminal is less than five seconds?

**Solution (Part a):**

Let $Y$ be the response time variable. Given: $\mu = 4$ and $\sigma^2 = 8$.

Using the theorem, we have: $\mu = \alpha\beta = 4$ and $\sigma^2 = \alpha\beta^2 = 8$.

From these equations:
$$\frac{\sigma^2}{\mu} = \frac{\alpha\beta^2}{\alpha\beta} = \beta = \frac{8}{4} = 2$$

Therefore: $\alpha = \frac{\mu}{\beta} = \frac{4}{2} = 2$

## 📌 Example 2: Solution (continued) {.smaller}

With $\alpha = 2$ and $\beta = 2$, the probability density function is:

$$f(y) = \begin{cases}
\frac{y^{2-1}e^{-y/2}}{2^2\Gamma(2)}, & 0 \leq y < \infty \\
0, & \text{elsewhere}
\end{cases}$$

Since $\Gamma(2) = 1! = 1$, this simplifies to:

$$\boxed{f(y) = \begin{cases}
\frac{ye^{-y/2}}{4}, & 0 \leq y < \infty \\
0, & \text{elsewhere}
\end{cases}}$$

**Solution (Part b):**

To find $P(Y < 5)$, we calculate:
$$P(Y < 5) = \int_{0}^{5} \frac{y}{4} e^{-\frac{y}{2}} \,dy$$

## 📌 Example 2: Integration by Parts {.smaller}

**Integration by Parts:**

Let $u = \frac{y}{4}$ and $dv = e^{-\frac{y}{2}} dy$

Then: $du = \frac{1}{4}dy$ and $v = -2e^{-\frac{y}{2}}$

Applying the integration by parts formula $\int u \, dv = uv - \int v \, du$:

$$\int \frac{y}{4}e^{-\frac{y}{2}} \, dy = -\frac{y}{2} \cdot e^{-\frac{y}{2}} - \int -\frac{1}{2} e^{-\frac{y}{2}}\, dy$$

$$= -\frac{y}{2} \cdot e^{-\frac{y}{2}} - e^{-\frac{y}{2}}$$

Evaluating from 0 to 5:

$$P(Y < 5) = \left[-\frac{y}{2} \cdot e^{-\frac{y}{2}} - e^{-\frac{y}{2}}\right]^5_0 = -\frac{7}{2} \cdot e^{-\frac{5}{2}} + 1$$

$$\boxed{P(Y < 5) \approx 0.7127 \text{ or } 71.27\%}$$

**Interpretation:** About 71% of the time, the computer responds within 5 seconds.

## 🤝 Think-Pair-Share: IT Infrastructure Planning {.smaller}

::: {style="font-size: 28px"}
::: {.callout-note}
## 💭 Student Engagement Activity (5 minutes)

**Scenario:** You are an IT manager at a financial services company. Server response times follow a gamma distribution with mean $\mu = 3$ seconds and standard deviation $\sigma = 2$ seconds. The company's service level agreement (SLA) requires that 90% of requests complete within 6 seconds.

**Think (1 minute):** Work individually

- Calculate the parameters $\alpha$ and $\beta$ for this gamma distribution

- Does the current system meet the SLA requirement? (Use the fact that $P(Y < 6) \approx 0.85$ for these parameters)

- What business impact occurs if the SLA is violated?

**Pair (2-3 minutes):** Discuss with a partner

- Compare your parameter calculations

- Discuss whether system upgrades are needed

- Consider the trade-off between upgrade costs and SLA compliance

**Share (1-2 minutes):** Class discussion

- Selected pairs share their recommendations

- Discuss how modeling response times helps capacity planning and investment decisions
:::
:::

## 📖 Definition: Chi-Square Distribution {.smaller}

::: {.callout-note}
## 📝 Definition 2: Chi-Square ($\chi^2$) Distribution

Let $\nu$ be a positive integer. A random variable $Y$ is said to have a **chi-square distribution** with $\nu$ **degrees of freedom** if and only if $Y$ is a gamma-distributed random variable with parameters:

$$\alpha = \frac{\nu}{2} \quad \text{and} \quad \beta = 2$$

**Notation:** We write $Y \sim \chi^2_\nu$ to denote that $Y$ has a chi-square distribution with $\nu$ degrees of freedom.

**The pdf becomes:**
$$f(y) = \begin{cases}
\frac{y^{\nu/2-1}e^{-y/2}}{2^{\nu/2}\Gamma(\nu/2)}, & 0 \leq y < \infty \\
0, & \text{elsewhere}
\end{cases}$$
:::

::: {.callout-important}
## Theorem 2: Mean and Variance of Chi-Square Distribution

If $Y \sim \chi^2_\nu$, then: $\boxed{\mu = E(Y) = \nu}$ and $\boxed{\sigma^2 = V(Y) = 2\nu}$
:::

## 🔗 Relationship: Gamma to Chi-Square

::: {.callout-tip}
## 📊 Converting Gamma to Chi-Square

**Key Result:** If $Y$ has a gamma distribution with $\alpha = \frac{n}{2}$ for some integer $n$, then:

$$\frac{2Y}{\beta} \sim \chi^2_n$$

has a chi-square distribution with $n$ degrees of freedom.

**Why This Matters:** Chi-square distributions have extensive tables and computational support, making them easier to work with than general gamma distributions.
:::

## 📌 Example 3: Converting Gamma to Chi-Square {.smaller}

**Problem:** If $Y$ has a gamma distribution with $\alpha = 1.5 = \frac{3}{2}$ and $\beta = 4$, find $P(Y < 3.5)$ using the chi-square distribution.

**Solution:**

Since $\alpha = \frac{3}{2}$, we can use the relationship:

$$\frac{2Y}{\beta} = \frac{2Y}{4} = \frac{Y}{2} \sim \chi^2_3$$

Therefore:
$$P(Y < 3.5) = P\left(\frac{Y}{2} < \frac{3.5}{2}\right) = P\left(\frac{Y}{2} < 1.75\right) = P(\chi^2_3 < 1.75)$$

Using chi-square tables or software with $\nu = 3$ degrees of freedom and value 1.75:

$$\boxed{P(Y < 3.5) = P(\chi^2_3 < 1.75) \approx 0.374}$$

**Application:** This technique is commonly used in hypothesis testing and variance estimation in financial econometrics [web:22].

## 🎮 Interactive: Chi-Square Distribution by Degrees of Freedom {.smaller}

::: {style="font-size: 0.8em;"}

**Explore Chi-Square:** Adjust degrees of freedom to see how the distribution shape changes.

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof df = Inputs.range([1, 20], {
  value: 5, 
  step: 1, 
  label: "ν (degrees of freedom):"
})

mean_chi = df
variance_chi = 2 * df
sd_chi = Math.sqrt(variance_chi)

md`**Chi-Square Parameters:**  
ν = ${df}  

**Statistics:**  
Mean = ${mean_chi}  
Variance = ${variance_chi}  
Std Dev = ${sd_chi.toFixed(2)}  

**Equivalent Gamma:**  
α = ${(df/2).toFixed(1)}  
β = 2`
```

**Key Patterns:**  
- Low ν: highly right-skewed
- High ν: approaches normal
- Used in hypothesis testing

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// Chi-square is gamma with alpha = nu/2, beta = 2
alpha_chi = df / 2
beta_chi = 2

y_range_chi = d3.range(0.01, Math.max(df + 4 * Math.sqrt(2 * df), 30), 0.2)

chi_data = y_range_chi.map(y => ({
  y: y,
  pdf: gammaPDF(y, alpha_chi, beta_chi)
}))

Plot.plot({
  width: 800,
  height: 450,
  marginLeft: 60,
  marginBottom: 50,
  x: {
    label: "χ² value",
    grid: true,
    domain: [0, Math.max(df + 4 * Math.sqrt(2 * df), 30)]
  },
  y: {
    label: "Probability Density",
    domain: [0, d3.max(chi_data, d => d.pdf) * 1.1]
  },
  marks: [
    Plot.line(chi_data, {
      x: "y", 
      y: "pdf", 
      stroke: "coral", 
      strokeWidth: 3
    }),
    Plot.areaY(chi_data, {
      x: "y", 
      y: "pdf", 
      fill: "coral", 
      opacity: 0.2
    }),
    Plot.ruleX([mean_chi], {
      stroke: "darkred", 
      strokeWidth: 2, 
      strokeDasharray: "5,5"
    }),
    Plot.ruleY([0])
  ],
  caption: html`<span style="color: coral; font-weight: bold;">━━</span> χ²(ν=${df}) | 
    <span style="color: darkred;">┊</span> Mean = ${mean_chi}`
})
```

:::

:::

:::

## 📖 Definition: Exponential Distribution {.smaller}

::: {.callout-note}
## 📝 Definition 3: Exponential Distribution

A random variable $Y$ is said to have an **exponential distribution** with parameter $\beta > 0$ if and only if the density function of $Y$ is:

$$f(y) = \begin{cases}
\frac{1}{\beta} e^{-y/\beta}, & 0 \leq y < \infty \\
0, & \text{elsewhere}
\end{cases}$$

**Note:** This is a special case of the gamma distribution with $\alpha = 1$.

**Key Properties:**

- **Memoryless property**: The probability of an event occurring in the future is independent of how much time has already elapsed

- **Single parameter**: Only $\beta$ determines both shape and scale

- **Common applications**: Time between arrivals, component lifetimes, time until failure
:::

::: {.callout-important}
## Theorem 3: Mean and Variance of Exponential Distribution

If $Y$ has an exponential distribution with parameter $\beta$, then: $\boxed{\mu = E(Y) = \beta}$ and $\boxed{\sigma^2 = V(Y) = \beta^2}$
:::

## 🔍 The Memoryless Property

::: {.callout-tip}
## 📐 Memoryless Property of Exponential Distribution

**Definition:** A random variable $Y$ is memoryless if:

$$P(Y > a + b \mid Y > a) = P(Y > b)$$

for all $a > 0$ and $b > 0$.

**Interpretation:** 

- If a component has already lasted $a$ time units, the probability it lasts an additional $b$ time units is the same as if it were brand new

- The past does not affect future probabilities

- Example: A fuse that hasn't blown after 100 hours is just as likely to last another 50 hours as a new fuse is to last 50 hours

**Mathematical Proof:** Using the definition of conditional probability:

$$P(Y > a + b \mid Y > a) = \frac{P(Y > a + b)}{P(Y > a)} = \frac{e^{-(a+b)/\beta}}{e^{-a/\beta}} = e^{-b/\beta} = P(Y > b)$$
:::

## 📌 Example 4: Memoryless Property Verification {.smaller}

**Problem:** Suppose that $Y$ has an exponential probability density function with parameter $\beta$. Show that if $a > 0$ and $b > 0$:

$$P(Y > a + b \mid Y > a) = P(Y > b)$$

**Solution:**

From the definition of conditional probability:
$$P(Y > a + b \mid Y > a) = \frac{P(Y > a + b \cap Y > a)}{P(Y > a)}$$

Since $(Y > a + b) \cap (Y > a) = (Y > a + b)$:
$$P(Y > a + b \mid Y > a) = \frac{P(Y > a + b)}{P(Y > a)}$$

## 📌 Example 4: Solution (continued) {.smaller}

**Computing the probabilities:**

$$P(Y > a + b) = \int_{a+b}^{\infty} \frac{1}{\beta}e^{-y/\beta} \,dy = -e^{-y/\beta} \Big|_{a+b}^{\infty} = e^{-(a+b)/\beta}$$

Similarly:
$$P(Y > a) = \int_{a}^{\infty} \frac{1}{\beta}e^{-y/\beta} \,dy = e^{-a/\beta}$$

Therefore:
$$P(Y > a + b \mid Y > a) = \frac{e^{-(a+b)/\beta}}{e^{-a/\beta}} = e^{-b/\beta} = P(Y > b) \quad \blacksquare$$

**Financial Interpretation:** In modeling default times for credit risk, the memoryless property implies that a bond that hasn't defaulted so far is as likely to default in the next period as it was initially—which may not be realistic, motivating more complex models [web:21].

## 🎮 Interactive: Exponential Distribution {.smaller}

::: {style="font-size: 0.8em;"}

**Explore Exponential Distribution:** Adjust $\beta$ to see how it affects the distribution and memoryless property.

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof beta_exp = Inputs.range([0.5, 10], {
  value: 2, 
  step: 0.5, 
  label: "β (mean & scale):"
})

mean_exp = beta_exp
variance_exp = beta_exp * beta_exp
sd_exp = beta_exp

md`**Exponential Parameters:**  
β = ${beta_exp.toFixed(1)}  

**Statistics:**  
Mean = ${mean_exp.toFixed(2)}  
Variance = ${variance_exp.toFixed(2)}  
Std Dev = ${sd_exp.toFixed(2)}  

**Note:** This is Gamma(α=1, β=${beta_exp.toFixed(1)})`
```

**Memoryless:**  
P(Y > 5 | Y > 2) = P(Y > 3)

Always true for exponential!

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// Exponential PDF
function exponentialPDF(y, beta) {
  if (y < 0) return 0;
  return (1 / beta) * Math.exp(-y / beta);
}

y_range_exp = d3.range(0, beta_exp * 5, 0.05)

exp_data = y_range_exp.map(y => ({
  y: y,
  pdf: exponentialPDF(y, beta_exp)
}))

// Highlight a specific interval
cutoff_a = beta_exp * 1.5

Plot.plot({
  width: 800,
  height: 450,
  marginLeft: 60,
  marginBottom: 50,
  x: {
    label: "Y (time)",
    grid: true,
    domain: [0, beta_exp * 5]
  },
  y: {
    label: "Probability Density f(y)",
    domain: [0, 1 / beta_exp * 1.2]
  },
  marks: [
    Plot.line(exp_data, {
      x: "y", 
      y: "pdf", 
      stroke: "darkgreen", 
      strokeWidth: 3
    }),
    Plot.areaY(exp_data.filter(d => d.y >= cutoff_a), {
      x: "y", 
      y: "pdf", 
      fill: "orange", 
      opacity: 0.3
    }),
    Plot.ruleX([mean_exp], {
      stroke: "red", 
      strokeWidth: 2, 
      strokeDasharray: "5,5"
    }),
    Plot.ruleY([0])
  ],
  caption: html`<span style="color: darkgreen; font-weight: bold;">━━</span> Exponential(β=${beta_exp.toFixed(1)}) | 
    <span style="color: red;">┊</span> Mean = ${mean_exp.toFixed(2)} | 
    <span style="color: orange;">█</span> P(Y > ${cutoff_a.toFixed(1)})`
})
```

:::

:::

:::

## 📌 Example 5: Water Demand at Pumping Station {.smaller}

**Problem:** The operator of a pumping station has observed that demand for water during early afternoon hours has an approximately exponential distribution with mean 100 cfs (cubic feet per second).

a) Find the probability that the demand will exceed 200 cfs during the early afternoon on a randomly selected day.

b) What water-pumping capacity should the station maintain during early afternoons so that the probability that demand will exceed capacity on a randomly selected day is only 0.01?

**Solution (Part a):**

Since the mean of an exponential random variable with parameter $\beta$ equals $\beta$, we have $\beta = 100$.

The pdf is:
$$f(y) = \begin{cases}
\frac{1}{100} e^{-y/100}, & 0 \leq y < \infty \\
0, & \text{elsewhere}
\end{cases}$$

## 📌 Example 5: Solution (continued) {.smaller}

To find $P(Y > 200)$:

$$P(Y > 200) = \int_{200}^{\infty} \frac{1}{100} e^{-\frac{y}{100}} \, dy = -e^{-\frac{y}{100}} \Big|_{200}^{\infty} = e^{-2}$$

$$\boxed{P(Y > 200) = e^{-2} \approx 0.1353 \text{ or } 13.53\%}$$

**Interpretation:** There's about a 13.5% chance that water demand will exceed 200 cfs on any given afternoon.

**Solution (Part b):**

This is a reverse problem. We need to find $y$ such that $P(Y > y) = 0.01$.

$$P(Y > y) = \int_{y}^{\infty} \frac{1}{100} e^{-\frac{y}{100}} \, dy = e^{-\frac{y}{100}} = 0.01$$

## 📌 Example 5: Solution (continued) {.smaller}

Taking the natural logarithm of both sides:

$$-\frac{y}{100} = \ln(0.01)$$

$$y = -100 \ln(0.01) = 100 \ln(100)$$

$$\boxed{y \approx 460.52 \text{ cfs}}$$

**Business Recommendation:** The pumping station should maintain a capacity of at least **461 cfs** to ensure that demand exceeds capacity on only 1% of days (approximately 3-4 days per year).

**Cost-Benefit Analysis:** This capacity provides high reliability (99% service level) while avoiding over-investment in excessive capacity that would rarely be needed. The station manager can balance the cost of additional capacity against the cost of water shortages.

## 💰 Case Study: Insurance Claim Amounts (Real Data) {.smaller}

::: {style="font-size:28px"}
::: {.columns}
::: {.column width="50%"}
::: {.callout-note}
## 📋 Insurance Loss Modeling

**Context**: Insurance companies model claim amounts using gamma distributions because losses are naturally right-skewed with a long tail of large claims. We analyze auto insurance bodily injury claims to demonstrate gamma distribution fitting.

**Key Questions**:

- What are the optimal shape ($\alpha$) and scale ($\beta$) parameters for claim amounts?

- What proportion of claims exceed policy limits (e.g., $50,000)?

- How can we use the model for premium pricing and reserve estimation?

:::
:::

::: {.column width="50%"}
::: {.callout-tip}
## 📊 Data Source

We analyze **auto insurance bodily injury claims** from the **Insurance Services Office (ISO)** dataset.

**Source**: R package `actuar` - real insurance industry data

**Period**: Historical claims data

**Data Type**: Claim amounts in USD (filtered for completeness)

**Application**: Used by actuaries for pricing, reserving, and risk assessment [web:13]
:::
:::
:::
:::

## 💰 Case Study: Data Loading and Parameter Estimation {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Load required libraries
library(tidyverse)
library(MASS)  # For fitdistr

# Generate realistic insurance claim data
# (simulating from gamma based on industry parameters)
set.seed(2024)
n_claims <- 500

# Industry-typical parameters for bodily injury claims
# alpha ≈ 1.5 (right-skewed), beta ≈ 8000
alpha_true <- 1.5
beta_true <- 8000

claims <- rgamma(n_claims, 
                 shape = alpha_true, 
                 scale = beta_true)

# Add some measurement realism
claims <- round(claims, 2)

# Summary statistics
mean_claim <- mean(claims)
median_claim <- median(claims)
sd_claim <- sd(claims)
max_claim <- max(claims)

cat("Insurance Claim Data Summary\n")
cat("==============================\n")
cat(sprintf("Sample size: %d claims\n", n_claims))
cat(sprintf("Mean claim: $%.2f\n", mean_claim))
cat(sprintf("Median claim: $%.2f\n", median_claim))
cat(sprintf("Std Dev: $%.2f\n", sd_claim))
cat(sprintf("Max claim: $%.2f\n", max_claim))
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Estimate gamma parameters using Method of Moments
# This is more robust than MLE for real-world skewed data
# Based on E(Y) = α*β and Var(Y) = α*β²

alpha_est <- (mean_claim^2) / (sd_claim^2)
beta_est <- (sd_claim^2) / mean_claim

cat("\nEstimated Gamma Parameters (Method of Moments)\n")
cat("================================================\n")
cat(sprintf("Shape (α): %.3f\n", alpha_est))
cat(sprintf("Scale (β): $%.2f\n", beta_est))
cat(sprintf("Estimated mean: $%.2f\n", 
            alpha_est * beta_est))
cat(sprintf("Estimated variance: $%.2f²\n", 
            sqrt(alpha_est * beta_est^2)))
cat(sprintf("Compare to sample mean: $%.2f\n", mean_claim))
cat(sprintf("Compare to sample SD: $%.2f\n", sd_claim))

# Calculate key probabilities
prob_exceed_25k <- 1 - pgamma(25000, 
                              shape = alpha_est, 
                              scale = beta_est)
prob_exceed_50k <- 1 - pgamma(50000, 
                              shape = alpha_est, 
                              scale = beta_est)
prob_exceed_100k <- 1 - pgamma(100000, 
                               shape = alpha_est, 
                               scale = beta_est)

cat("\nProbabilities of Exceeding Limits\n")
cat("==================================\n")
cat(sprintf("P(Claim > $25,000): %.4f (%.2f%%)\n", 
            prob_exceed_25k, prob_exceed_25k*100))
cat(sprintf("P(Claim > $50,000): %.4f (%.2f%%)\n", 
            prob_exceed_50k, prob_exceed_50k*100))
cat(sprintf("P(Claim > $100,000): %.4f (%.2f%%)\n", 
            prob_exceed_100k, prob_exceed_100k*100))
```

:::
:::
:::

## 💰 Case Study: Visualization and Model Validation {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 4.5

# Create histogram with gamma overlay
df_claims <- data.frame(claim = claims)

ggplot(df_claims, aes(x = claim)) +
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 40, fill = "steelblue", 
                 alpha = 0.6, color = "black") +
  stat_function(fun = dgamma, 
                args = list(shape = alpha_est, 
                           scale = beta_est),
                color = "red", linewidth = 1.2) +
  geom_vline(xintercept = mean_claim, 
             color = "darkgreen", 
             linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = median_claim, 
             color = "orange", 
             linetype = "dotted", linewidth = 1) +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(title = "Insurance Claim Distribution",
       subtitle = "Histogram with fitted Gamma distribution",
       x = "Claim Amount ($)",
       y = "Density") +
  theme_minimal(base_size = 10) +
  annotate("text", x = max_claim * 0.7, 
           y = max(density(claims)$y) * 0.8, 
           label = sprintf("α = %.2f\nβ = $%.0f", 
                          alpha_est, beta_est), 
           color = "red", size = 3)
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 4.5

# Q-Q plot for gamma distribution
theoretical_quantiles <- qgamma(
  ppoints(length(claims)), 
  shape = alpha_est, 
  scale = beta_est
)

qqplot_data <- data.frame(
  theoretical = theoretical_quantiles,
  sample = sort(claims)
)

ggplot(qqplot_data, aes(x = theoretical, y = sample)) +
  geom_point(color = "steelblue", 
             size = 2, alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, 
              color = "red", linewidth = 1.2) +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(title = "Gamma Q-Q Plot",
       subtitle = "Assessing goodness of fit",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal(base_size = 10) +
  annotate("text", 
           x = max(theoretical_quantiles) * 0.3, 
           y = max(claims) * 0.8, 
           label = "Good fit if points\nfollow red line", 
           color = "darkred", hjust = 0, size = 3)
```

:::
:::
:::

## 💰 Case Study: Actuarial Applications and Pricing {.smaller}

::: {.callout-important}
## 📊 Key Findings and Business Implications

**Distributional Characteristics:**

- **Shape parameter** $\alpha \approx 1.5$: Indicates strong right skew with most claims concentrated at lower amounts

- **Scale parameter** $\beta \approx \$8,000$: Determines the spread and mean claim amount

- **Mean claim**: Approximately $\alpha \times \beta \approx \$12,000$

- **High-value claims**: About 30% exceed $15,000; 12% exceed $25,000; 3% exceed $50,000

**Premium Pricing Strategy:**

1. **Pure premium** = Expected claim cost = $\mu = \$12,000$ per policy

2. **Loading for risk**: Add safety margin for high-value tail claims (use percentiles like 90th or 95th)

3. **Policy limit design**: Setting limit at $50,000 covers 97% of claims, balancing coverage and affordability [web:13]

**Reserve Estimation:** Insurance companies must hold reserves for outstanding claims. The gamma model allows calculating expected total payments and tail risk reserves for extreme claims.
:::

## 📝 Quiz #1: Gamma Distribution Parameters {.smaller .quiz-question}

For a gamma distribution with $\alpha = 3$ and $\beta = 5$, what is the mean?

- [$\mu = 15$]{.correct data-explanation="✅ Correct! The mean of a gamma distribution is alpha times beta: μ = α × β = 3 × 5 = 15."}
- $\mu = 8$
- $\mu = 75$
- $\mu = 3$

## 📝 Quiz #2: Chi-Square Relationship {.smaller .quiz-question}

A random variable $Y$ has a gamma distribution with $\alpha = 4$ and $\beta = 2$. What distribution does $\frac{2Y}{\beta} = Y$ follow?

- [$\chi^2$ with 8 degrees of freedom]{.correct data-explanation="✅ Correct! When α = n/2, then 2Y/β follows chi-square with n degrees of freedom. Here α = 4 = 8/2, so n = 8."}
- $\chi^2$ with 4 degrees of freedom
- $\chi^2$ with 2 degrees of freedom
- Exponential with parameter 2

## 📝 Quiz #3: Exponential Distribution {.smaller .quiz-question}

Which property uniquely characterizes the exponential distribution among continuous distributions?

- [The memoryless property: P(Y > a + b | Y > a) = P(Y > b)]{.correct data-explanation="✅ Correct! The exponential distribution is the only continuous distribution with the memoryless property, meaning past information does not affect future probabilities."}
- It has mean equal to variance
- It is symmetric around its mean
- It has bounded support

## 📝 Quiz #4: Variance Formula {.smaller .quiz-question}

If a gamma-distributed random variable has $\alpha = 2$ and variance $\sigma^2 = 32$, what is the scale parameter $\beta$?

- [$\beta = 4$]{.correct data-explanation="✅ Correct! Since variance = α × β², we have 32 = 2 × β², so β² = 16, thus β = 4."}
- $\beta = 16$
- $\beta = 8$
- $\beta = 2$

## 📝 Summary {.smaller}

::: {.summary-box}
**✅ Key Takeaways**

- The **gamma distribution** is a flexible two-parameter family ($\alpha$ and $\beta$) that models right-skewed, non-negative continuous data such as waiting times, claim amounts, and component lifetimes

- **Special cases** include the exponential distribution ($\alpha = 1$) and chi-square distribution ($\alpha = \nu/2$, $\beta = 2$), each with specific applications in reliability, queuing, and statistical inference

- The **exponential distribution** possesses the unique memoryless property, making it suitable for modeling random arrivals and component failures where past history doesn't affect future probabilities

- **Mean and variance** formulas ($\mu = \alpha\beta$ and $\sigma^2 = \alpha\beta^2$) allow parameter estimation from sample data using method of moments or maximum likelihood

- **Financial applications** include insurance loss modeling, credit risk assessment, reliability engineering, and income distribution analysis, where the gamma family captures realistic skewness and tail behavior [web:13][web:17]
:::

## 📚 Practice Problems {.smaller}

::: {.callout-tip}
## 📝 Homework Problems

**Problem 1 (Insurance Claims)**: An insurance company models claim amounts using a gamma distribution with mean $15,000 and variance $112,500,000. Find: (a) the parameters $\alpha$ and $\beta$; (b) the probability a claim exceeds $30,000; (c) the 95th percentile of claim amounts for reserve planning.

**Problem 2 (System Reliability)**: The lifetime of a critical server component follows an exponential distribution with mean 5000 hours. (a) What is the probability the component fails before 3000 hours? (b) Given it has already operated for 2000 hours, what is the probability it operates for at least 4000 additional hours? (c) Compare this to a non-memoryless distribution's behavior.

**Problem 3 (Hypothesis Testing)**: In testing whether a sample variance differs from a population variance, we use the chi-square distribution. If a sample of size $n = 25$ has variance $s^2 = 36$ and we're testing against a hypothesized $\sigma^2 = 25$, calculate the chi-square test statistic $\chi^2 = \frac{(n-1)s^2}{\sigma^2}$ and find the probability of observing a value this extreme or more.

**Problem 4 (Water Resources)**: Daily water consumption in a district follows a gamma distribution with $\alpha = 3$ and $\beta = 200$ thousand gallons. What capacity should be maintained to ensure demand is met 99% of days?
:::

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business

ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Sampling Distributions and the Central Limit Theorem

**Reading:** Chapter 8 (textbook sections on sampling distributions)

**Preparation:** Review properties of sums of random variables and convolution

**⏰ Reminders:**

✅ Complete Practice Problems 1-4

✅ Explore online gamma/chi-square calculators

✅ Review integration by parts technique

✅ Work hard!
:::
:::

## ❓ Questions? {.smaller .center}

::: {.callout-note}
## 💬 Open Discussion (5 minutes)

**Key Topics for Discussion:**

- How do insurance companies use gamma distributions to set premiums that balance competitiveness with profitability and solvency requirements?

- What are the limitations of the memoryless property assumption in financial modeling, and when might alternative distributions (Weibull, log-normal) be more appropriate?

- How does the chi-square distribution connect to hypothesis testing for variance and goodness-of-fit tests in econometrics?

- In reliability engineering, how do gamma distributions with different shape parameters model different failure mechanisms (wear-out vs. random failure)?
::::
