---
title: "Mathematical Statistics"
subtitle: "Linear Models & Least Squares"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    margin: 0.04
    # footer: "Mathematical Statistics – Linear Models & Least Squares"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
    include-after-body:
      - text: |
          <script src="../qrcode.min.js"></script>
          <script src="../attendance-qr-slide.js"></script>
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::: {style="font-size:39px;"}
::: {.learning-objectives}
By the end of this lecture, you will be able to:

- **Formulate** the simple linear regression model and state its assumptions

- **Derive** the least-squares estimators $\hat\beta_0$ and $\hat\beta_1$ using the $S_{xx}$, $S_{xy}$ shorthand

- **Apply** the unbiasedness results and compute $S^2 = \text{SSE}/(n-2)$ as the estimator of $\sigma^2$

- **Execute** the $t$-test for $H_0: \beta_1 = 0$ and construct a confidence interval for $\beta_1$

- **Interpret** the coefficient of determination $r^2$ and Pearson correlation $r$ in financial contexts
:::
:::

---

## 📱 Attendance Check-in {.center background-color="#f0f4ff"}

<div class="attendance-qr" data-status="present" id="qr-present" style="min-height: 400px; display: flex; align-items: center; justify-content: center;"></div>

## 📋 Overview {.smaller}

::: {style="font-size:50px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **The Linear Model** — definition, assumptions, financial motivation (§11.1–11.2)

- **Least Squares** — minimising SSE, deriving $\hat\beta_1$ and $\hat\beta_0$ (§11.3)

- **Estimator Properties** — unbiasedness, variances, $S^2 = \text{SSE}/(n-2)$ (§11.4)

- **Inference on $\beta_i$** — $t$-test for slope, CI for $\beta_1$ (§11.5)

- **Goodness of Fit** — $r^2$ and Pearson $r$ (§11.6–11.7)

- **Case Study** — Estimating market beta (CAPM) with real stock data
:::
:::
:::

---

## 📖 Motivation: Regression in Finance {.smaller}

::: {style="font-size:45px;"}
::: {.callout-note}
## 🎯 Linear Models are Everywhere in Finance

::: {.columns}
::: {.column width="50%"}
::: {.fragment}
**Asset Pricing:**

- CAPM: $R_i - R_f = \alpha + \beta(R_m - R_f) + \varepsilon$
- Market beta $\beta$ measures systematic risk
- Is $\alpha \neq 0$? → abnormal return (Jensen's alpha)
- Factor models: Fama-French 3 & 5-factor
:::
:::

::: {.column width="50%"}
::: {.fragment}
**Regulation & Forecasting:**

- Broadband penetration as a function of price
- Revenue elasticity to GDP growth
- QoS speed vs infrastructure investment
- Loan default probability vs credit score
:::
:::
:::
:::

**Key idea:** We observe $(x_i, Y_i)$ pairs. We believe $E(Y \mid x) = \beta_0 + \beta_1 x$. How do we estimate $\beta_1$, test whether it is significant, and measure how well $x$ explains $Y$?
:::

---

## 📝 Definition: The Simple Linear Model {.smaller}

::: {style="font-size:35px;"}
::: {.callout-note}
## 📝 Definition 11.1 — Simple Linear Regression Model

$$Y = \beta_0 + \beta_1 x + \varepsilon$$

| Symbol | Meaning |
|--------|---------|
| $Y$ | Random response variable |
| $x$ | Fixed (non-random) predictor variable |
| $\beta_0$ | Unknown intercept parameter |
| $\beta_1$ | Unknown slope parameter |
| $\varepsilon$ | Random error term |

**Assumptions (required for inference):**

- $E(\varepsilon) = 0$ → $E(Y \mid x) = \beta_0 + \beta_1 x$
- $V(\varepsilon) = \sigma^2$ (constant, independent of $x$) — **homoscedasticity**
- $\varepsilon \sim N(0, \sigma^2)$ for $t$-tests and CIs
- Errors are independent across observations

:::
:::

---

## 🧮 The Method of Least Squares (§11.3) {.smaller}

::: {style="font-size:50px"}

Given $n$ data pairs $(x_1, Y_1), \ldots, (x_n, Y_n)$, the predicted value under the model is $\hat Y_i = \hat\beta_0 + \hat\beta_1 x_i$.

**We minimise the Sum of Squares for Error (SSE):**

$$\text{SSE} = \sum_{i=1}^n (Y_i - \hat Y_i)^2 = \sum_{i=1}^n [Y_i - (\hat\beta_0 + \hat\beta_1 x_i)]^2$$

::: {.fragment}
Setting $\partial\,\text{SSE}/\partial\hat\beta_0 = 0$ and $\partial\,\text{SSE}/\partial\hat\beta_1 = 0$ and solving the **normal equations** gives the unique minimisers:
:::
:::

---

## 🧮 Least-Squares Estimators {.smaller}

::: {style="font-size:39px;"}
::: {.callout-important}
## Least-Squares Estimators (Wackerly §11.3)

$$\hat\beta_1 = \frac{S_{xy}}{S_{xx}}, \qquad \hat\beta_0 = \bar Y - \hat\beta_1 \bar x$$

**Computational shorthand:**

$$S_{xx} = \sum(x_i - \bar x)^2 = \sum x_i^2 - \frac{(\sum x_i)^2}{n}$$

$$S_{xy} = \sum(x_i - \bar x)(Y_i - \bar Y) = \sum x_i Y_i - \frac{(\sum x_i)(\sum Y_i)}{n}$$

$$S_{yy} = \sum(Y_i - \bar Y)^2 = \sum Y_i^2 - \frac{(\sum Y_i)^2}{n}$$

:::

**Fitted line:** $\hat Y = \hat\beta_0 + \hat\beta_1 x$ always passes through $(\bar x, \bar Y)$.
:::

---

## 📌 Example 1: CAPM Beta by Hand {.smaller}

::: {style="font-size:36px;"}
**Scenario:** Estimate market beta for a stock using $n = 5$ weekly excess returns.

| $x_i$ (Market) | $Y_i$ (Stock) | $x_i Y_i$ | $x_i^2$ |
|:-:|:-:|:-:|:-:|
| −2 | 0 | 0 | 4 |
| −1 | 0 | 0 | 1 |
|  0 | 1 | 0 | 0 |
|  1 | 1 | 1 | 1 |
|  2 | 3 | 6 | 4 |
| **Σ = 0** | **Σ = 5** | **Σ = 7** | **Σ = 10** |

:::{.fragment}
$$S_{xx} = 10 - \frac{0^2}{5} = 10, \quad S_{xy} = 7 - \frac{(0)(5)}{5} = 7$$
:::
:::

---

## 📌 Example 1: Solution {.smaller}

::: {style="font-size:50px;"}
$$\hat\beta_1 = \frac{7}{10} = 0.7, \quad \hat\beta_0 = \frac{5}{5} - 0.7 \times 0 = 1.0$$

:::{.fragment}
**Fitted line:** $\hat Y = 1.0 + 0.7x$
:::

:::{.fragment}
**Interpretation:** For every 1% rise in the market, the stock is predicted to return an extra $0.7\%$. The intercept $\hat\beta_0 = 1.0$ suggests positive alpha — but is it significant?
:::
:::

---

## 🧮 Properties of LS Estimators (§11.4) {.smaller}

::: {.callout-important}
## Theorem 11.1 — Unbiasedness and Variances

Under $E(\varepsilon) = 0$ and $V(\varepsilon) = \sigma^2$:

$$E(\hat\beta_1) = \beta_1, \qquad V(\hat\beta_1) = \frac{\sigma^2}{S_{xx}}$$

$$E(\hat\beta_0) = \beta_0, \qquad V(\hat\beta_0) = \sigma^2 \cdot \frac{\sum x_i^2}{n \cdot S_{xx}}$$

:::

::: {style="font-size:42px"}
::: {.fragment}
**Estimating $\sigma^2$:** We cannot use $\sigma^2$ directly — we estimate it with:

$$S^2 = \frac{\text{SSE}}{n-2} \qquad \text{where} \qquad \text{SSE} = S_{yy} - \hat\beta_1 S_{xy}$$

$S^2$ is an **unbiased** estimator of $\sigma^2$ with $n-2$ degrees of freedom (two are used estimating $\beta_0$ and $\beta_1$).
:::
:::

---

## 📌 Example 2: Computing SSE and $S^2$ {.smaller}

::: {style="font-size:50px;"}
**Continuing Example 1:** $S_{yy} = \sum Y_i^2 - (\sum Y_i)^2/n$

$$\sum Y_i^2 = 0+0+1+1+9 = 11, \quad S_{yy} = 11 - \frac{25}{5} = 6$$

:::{.fragment}
$$\text{SSE} = S_{yy} - \hat\beta_1 S_{xy} = 6 - 0.7 \times 7 = 6 - 4.9 = 1.1$$
:::

:::{.fragment}
$$S^2 = \frac{\text{SSE}}{n-2} = \frac{1.1}{3} = 0.367, \quad S = \sqrt{0.367} = 0.606$$
:::
:::

---

## 📌 Example 2: Interpretation {.smaller}

::: {style="font-size:50px;"}
::: {.columns}
::: {.column width="50%"}
**Interpretation of $S$:**

$S \approx 0.61\%$ is the estimated standard deviation of the stock return around the fitted regression line — a measure of **residual risk** not explained by market beta.
:::

::: {.column width="50%"}
**Rule of thumb:**

- Small $S$ relative to $\bar Y$ → good fit
- Large $S$ → substantial unexplained variation
- Formally measured by $r^2$ (see later)
:::
:::
:::

---

## 🧮 $t$-Test for the Slope $\beta_1$ (§11.5) {.smaller}

::: {style="font-size:43px;"}
::: {.callout-important}
## Test: $H_0: \beta_1 = \beta_{10}$ — (Most common: $\beta_{10} = 0$)

$$T = \frac{\hat\beta_1 - \beta_{10}}{S\sqrt{1/S_{xx}}} = \frac{\hat\beta_1 - \beta_{10}}{S/\sqrt{S_{xx}}} \sim t_{n-2} \text{ under } H_0$$

**Rejection regions** at level $\alpha$ (df $= n-2$):

| $H_a$ | Rejection Region |
|-------|------------------|
| $\beta_1 > \beta_{10}$ | $t > t_\alpha$ |
| $\beta_1 < \beta_{10}$ | $t < -t_\alpha$ |
| $\beta_1 \neq \beta_{10}$ | $\|t\| > t_{\alpha/2}$ |

:::

**Financial use:** $H_0: \beta_1 = 0$ tests whether $x$ has **any linear relationship** with $Y$.
:::

---

## 📌 Example 3: Testing the Market Beta {.smaller}

::: {style="font-size:45px;"}
**Continuing Example 1:** Test $H_0: \beta_1 = 0$ vs $H_a: \beta_1 \neq 0$ at $\alpha = 0.05$.

$$S = 0.606, \quad S_{xx} = 10, \quad \hat\beta_1 = 0.7$$

:::{.fragment}
$$T = \frac{0.7 - 0}{0.606/\sqrt{10}} = \frac{0.7}{0.1916} = 3.654$$
:::

:::{.fragment}
**Critical value:** $t_{0.025}$ with $\nu = n-2 = 3$ df $= 3.182$.

Since $|T| = 3.654 > 3.182$ → **Reject $H_0$** ✅ — the market beta is significantly different from zero.
:::
:::

---

## 📌 Example 3: Confidence Interval {.smaller}

::: {style="font-size:47px;"}
::: {.columns}
::: {.column width="50%"}
**95% Confidence Interval for $\beta_1$:**

$$\hat\beta_1 \pm t_{0.025} \cdot S/\sqrt{S_{xx}}  $$
$$= 0.7 \pm 3.182 \times 0.1916  $$
$$= 0.7 \pm 0.610$$

$$\Rightarrow (0.090, \; 1.310)$$
:::
::: {.column width="50%"}
**Interpretation:** We are 95% confident the stock's market beta lies between $0.09$ and $1.31$. The wide interval reflects the tiny sample ($n=5$) — real CAPM estimation uses years of weekly data.
:::
:::
:::


## 🧮 Coefficient of Determination $r^2$ {.smaller}

::: {style="font-size:47px;"}
::: {.callout-important}
## Definition — $r^2$ (Wackerly §11.6)

$$r^2 = 1 - \frac{\text{SSE}}{S_{yy}}$$

$$r^2 = \frac{S_{yy} - \text{SSE}}{S_{yy}} = \frac{\hat\beta_1 S_{xy}}{S_{yy}}$$

**Interpretation:** $r^2$ is the **proportion of total variation in $Y$ explained by the linear relationship with $x$**.

- $r^2 = 1.0$ → Perfect fit
- $r^2 = 0.0$ → No linear fit  
- $r^2 = 0.7$ → 70% explained

:::
:::

---

## 🧮 $r^2$ Example {.smaller}

::: {style="font-size:28px;"}
**Recall Example 1:**

$$r^2 = 1 - \frac{\text{SSE}}{S_{yy}} = 1 - \frac{1.1}{6} = 1 - 0.183 = 0.817$$

**Interpretation:** The market index explains **81.7%** of the stock's return variation.

- 81.7% of variation is **systematic risk** (due to market movements)
- 18.3% of variation is **idiosyncratic risk** (firm-specific)
:::

---

## 🧮 Pearson Correlation Coefficient $r$ {.smaller}

::: {.callout-note}
## 📝 Pearson $r$ — Measuring Linear Association

$$r = \frac{S_{xy}}{\sqrt{S_{xx} \cdot S_{yy}}}$$

- $r \in [-1, 1]$
- $r > 0$: positive linear relationship
- $r < 0$: negative linear relationship
- $r^2$ = coefficient of determination in simple regression

:::

::: {style="font-size:35px"}
::: {.fragment}
**Testing $H_0: \rho = 0$ (no linear association):**

$$T = \frac{r\sqrt{n-2}}{\sqrt{1-r^2}} \sim t_{n-2} \text{ under } H_0$$

This is **algebraically equivalent** to the $t$-test for $H_0: \beta_1 = 0$ — same test, same p-value.
:::
:::

---

## 🎮 Interactive: Fit a Regression Line {.smaller}

:::::: {style="font-size: 0.72em; margin-top: -8px;"}
**Adjust the true slope and noise to see how LS estimates, $r^2$, and the $t$-statistic respond.**

::::: columns
::: {.column width="30%"}

```{ojs}
//| echo: false

viewof true_beta1 = {
  const input = Inputs.range([-2, 2], {value: 0.7, step: 0.1, label: "True β₁:"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

viewof noise_sd = {
  const input = Inputs.range([0.1, 3.0], {value: 0.8, step: 0.1, label: "Noise σ:"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

viewof n_reg = {
  const input = Inputs.range([5, 60], {value: 20, step: 1, label: "n (obs):"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}

// Seeded pseudo-random using LCG
rand_data = {
  const seed = 42;
  let s = seed;
  const lcg = () => { s = (1664525 * s + 1013904223) >>> 0; return s / 4294967296; };
  const nrand = () => {
    const u = lcg(), v = lcg();
    return Math.sqrt(-2*Math.log(u+1e-10)) * Math.cos(2*Math.PI*v);
  };
  const pts = [];
  for (let i = 0; i < n_reg; i++) {
    const x = (i / (n_reg - 1)) * 4 - 2;
    const y = 1.0 + true_beta1 * x + noise_sd * nrand();
    pts.push({x, y});
  }
  return pts;
}

// LS estimates
lsfit = {
  const n = rand_data.length;
  const xbar = rand_data.reduce((s, d) => s + d.x, 0) / n;
  const ybar = rand_data.reduce((s, d) => s + d.y, 0) / n;
  const Sxx = rand_data.reduce((s, d) => s + (d.x - xbar)**2, 0);
  const Sxy = rand_data.reduce((s, d) => s + (d.x - xbar)*(d.y - ybar), 0);
  const Syy = rand_data.reduce((s, d) => s + (d.y - ybar)**2, 0);
  const b1 = Sxy / Sxx;
  const b0 = ybar - b1 * xbar;
  const SSE = Syy - b1 * Sxy;
  const S2 = SSE / (n - 2);
  const S = Math.sqrt(S2);
  const r2 = 1 - SSE / Syy;
  const t_stat = b1 / (S / Math.sqrt(Sxx));
  return {b0, b1, S, S2, r2, t_stat, Sxx, SSE, n};
}

stats_html = html`<div style="background:#f0f4ff; padding:8px; border-radius:6px; font-size:0.93em; line-height:1.7;">
  <strong>β̂₁ = ${lsfit.b1.toFixed(3)}</strong> (true: ${true_beta1.toFixed(1)})<br>
  <strong>β̂₀ = ${lsfit.b0.toFixed(3)}</strong><br>
  <strong>S = ${lsfit.S.toFixed(3)}</strong><br>
  <strong>r² = ${lsfit.r2.toFixed(3)}</strong><br>
  <strong>t = ${lsfit.t_stat.toFixed(3)}</strong> (df = ${lsfit.n - 2})
</div>`;

md`${stats_html}`
```

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

line_pts = [
  {x: -2, y: lsfit.b0 + lsfit.b1 * (-2)},
  {x:  2, y: lsfit.b0 + lsfit.b1 * 2}
]

true_line_pts = [
  {x: -2, y: 1.0 + true_beta1 * (-2)},
  {x:  2, y: 1.0 + true_beta1 * 2}
]

Plot.plot({
  width: 680, height: 380,
  marginLeft: 55, marginBottom: 50,
  x: {domain: [-2.3, 2.3], label: "Market excess return x (%)"},
  y: {label: "Stock excess return Y (%)"},
  title: `LS fit | β̂₁=${lsfit.b1.toFixed(2)}, r²=${lsfit.r2.toFixed(3)}, t=${lsfit.t_stat.toFixed(2)}, n=${n_reg}`,
  marks: [
    Plot.dot(rand_data, {x: "x", y: "y", fill: "steelblue", opacity: 0.7, r: 4}),
    Plot.line(line_pts,      {x: "x", y: "y", stroke: "tomato",    strokeWidth: 2.5}),
    Plot.line(true_line_pts, {x: "x", y: "y", stroke: "#aaa",      strokeWidth: 1.5, strokeDasharray: "6,3"}),
    Plot.ruleY([0], {stroke: "#ddd"}),
    Plot.ruleX([0], {stroke: "#ddd"}),
    Plot.text([{x: 1.6, y: lsfit.b0 + lsfit.b1*1.6 + 0.3, text: "LS fit"}],
              {x:"x",y:"y",text:"text", fill:"tomato", fontSize:13}),
    Plot.text([{x: 1.6, y: 1.0 + true_beta1*1.6 + 0.3, text: "True"}],
              {x:"x",y:"y",text:"text", fill:"#999", fontSize:12})
  ]
})
```

*Red = LS fitted line. Grey dashed = true population line. Observe how $r^2$ and $t$ change with noise and $n$.*
:::
:::::
::::::

---

## 🤝 Think-Pair-Share {.smaller}

::: {style="font-size:44px;"}
::: {.callout-tip}
## 💬 Activity (5 minutes)

**Scenario:** An ICTA analyst is studying the relationship between **broadband price** (AZN/month, $x$) and **broadband penetration** (%, $Y$) across 10 regions of Azerbaijan.

| $\sum x_i = 320$ | $\sum Y_i = 480$ | $n = 10$ |
|:-:|:-:|:-:|
| $\sum x_i^2 = 11{,}200$ | $\sum x_i Y_i = 14{,}400$ | $\sum Y_i^2 = 24{,}600$ |

**Questions:**

1. Compute $S_{xx}$, $S_{xy}$, $S_{yy}$ using the shorthand formulas.

2. Find $\hat\beta_1$ and $\hat\beta_0$. Interpret $\hat\beta_1$ in context.

3. Compute SSE and $S^2$. Then compute $r^2$ and interpret it.

4. Test $H_0: \beta_1 = 0$ vs $H_a: \beta_1 \neq 0$ at $\alpha = 0.05$. What is your conclusion?
:::

```{r}
#| echo: false
#| eval: false
library(countdown)
countdown(minutes = 5, bottom = 0, right = 0, color_border = "#003366")
```
:::

---

## ✅ Think-Pair-Share: Solution (1/4) {.smaller}

::: {style="font-size:50px;"}
**1. Sums of squares:**

$$S_{xx} = \sum x_i^2 - \frac{(\sum x_i)^2}{n} = 11200 - \frac{320^2}{10}$$

$$= 11200 - 10240 = 960$$

$$S_{xy} = \sum x_i y_i - \frac{(\sum x_i)(\sum y_i)}{n} = 14400 - \frac{320 \times 480}{10}$$  

$$= 14400 - 15360 = -960$$

$$S_{yy} = \sum y_i^2 - \frac{(\sum y_i)^2}{n} = 24600 - \frac{480^2}{10}$$

$$= 24600 - 23040 = 1560$$
:::

---

## ✅ Think-Pair-Share: Solution (2/5) {.smaller}

::: {style="font-size:34px;"}
**2a. Slope estimate:**

$$\hat\beta_1 = \frac{S_{xy}}{S_{xx}} = \frac{-960}{960} = -1.0$$

Each 1 AZN price increase → **1 pp decrease** in penetration.
:::

---

## ✅ Think-Pair-Share: Solution (3/5) {.smaller}

::: {style="font-size:50px;"}
**2b. Intercept estimate:**

$$\hat\beta_0 = \bar{y} - \hat\beta_1 \bar{x} = 48 - (-1)(32) = 80$$

**Fitted model:** $\hat{y} = 80 - 1.0 x$
:::

---

## ✅ Think-Pair-Share: Solution (4/5) {.smaller}

::: {style="font-size:50px;"}
**3. Error variance and $r^2$:**

$$\text{SSE} = S_{yy} - \hat\beta_1 S_{xy}$$  
$$= 1560 - (-1)(-960) = 600$$

$$S^2 = \frac{600}{8} = 75, \quad S = 8.66$$

$$r^2 = 1 - \frac{600}{1560} = 0.615$$
:::

---

## ✅ Think-Pair-Share: Solution (5/5) {.smaller}

::: {style="font-size:42px;"}
**4. Hypothesis test:** $H_0: \beta_1 = 0$ at $\alpha = 0.05$

$$T = \frac{-1.0}{8.66/\sqrt{960}} = -3.577$$

$t_{0.025, 8} = 2.306$

$|T| > 2.306$ → **Reject $H_0$** ✅
:::

---

## 💰 Case Study: CAPM Regression Table {.smaller}

```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: true

library(tidyverse)
library(tidyquant)
library(knitr)

# Estimate CAPM beta for MSFT vs S&P 500 (SPY)
symbols <- c("MSFT", "SPY")
prices  <- tq_get(symbols, from = "2021-01-01", to = "2023-12-31")

monthly_ret <- prices %>%
  group_by(symbol) %>%
  tq_transmute(select      = adjusted,
               mutate_fun  = periodReturn,
               period      = "monthly",
               col_rename  = "return") %>%
  pivot_wider(names_from = symbol, values_from = return) %>%
  na.omit()

# OLS regression: MSFT ~ SPY
fit <- lm(MSFT ~ SPY, data = monthly_ret)
s   <- summary(fit)

coef_tbl <- data.frame(
  Parameter  = c("β₀ (alpha)", "β₁ (market beta)"),
  Estimate   = round(coef(fit) * 100, 4),
  Std_Error  = round(s$coefficients[,2] * 100, 4),
  t_stat     = round(s$coefficients[,3], 3),
  p_value    = round(s$coefficients[,4], 4)
)

kable(coef_tbl, digits = 4,
      caption = paste0("CAPM Regression: MSFT ~ SPY | R² = ",
                       round(s$r.squared, 3),
                       " | S = ", round(s$sigma * 100, 3), "%"))
```

---

## 💰 Case Study: CAPM Scatter Plot {.smaller}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 8
#| fig-height: 5.5

# Scatter with regression line and CI band
ggplot(monthly_ret, aes(x = SPY * 100, y = MSFT * 100)) +
  geom_point(colour = "steelblue", alpha = 0.7, size = 2.5) +
  geom_smooth(method = "lm", colour = "tomato",
              fill   = "tomato", alpha = 0.15, linewidth = 1.2) +
  geom_hline(yintercept = 0, colour = "#aaa", linewidth = 0.5) +
  geom_vline(xintercept = 0, colour = "#aaa", linewidth = 0.5) +
  annotate("text", x = -8, y = 12,
           label = paste0("β̂₁ = ", round(coef(fit)[2], 3),
                          "\nr² = ", round(s$r.squared, 3)),
           hjust = 0, size = 4.5, colour = "tomato") +
  labs(
    title    = "CAPM Beta: MSFT vs SPY (2021–2023)",
    subtitle = "Monthly excess returns | OLS regression",
    x = "S&P 500 monthly return (%)",
    y = "MSFT monthly return (%)"
  ) +
  theme_minimal(base_size = 12)
```

---

## 💰 Case Study: Key Findings {.smaller}

::: {style="font-size: 41px;"}
::: {.callout-important}
## 📊 MSFT CAPM Regression Results (2021–2023)

::: {.columns}

::: {.column width="33%"}
::: {.fragment}
**Estimated Parameters:**

- $\hat\beta_1 \approx 1.10$ — MSFT is slightly more volatile than the market (high-beta tech stock)

- $\hat\beta_0 \approx 0.3\%$/month — small positive alpha, likely insignificant

- $S \approx 4.5\%$ — residual risk per month not explained by market
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Inference:**

- $t_{\hat\beta_1}$ large → **Reject $H_0: \beta_1 = 0$** — market return significantly explains MSFT

- $t_{\hat\beta_0}$ small → **Fail to reject $H_0: \beta_0 = 0$** — no statistically significant alpha

- $r^2 \approx 0.70$ → market explains ~70% of MSFT's monthly return variation
:::
:::

::: {.column width="33%"}
::: {.fragment}
**Financial Implications:**

1. **Risk decomposition**: ~70% systematic (market), ~30% idiosyncratic (firm-specific)

2. **Portfolio construction**: $\beta > 1$ means MSFT amplifies market swings

3. **Active management**: Alpha of ~0.3%/month is not statistically different from zero — consistent with market efficiency
:::
:::

:::
:::
:::

---

## 📝 Quiz #1: LS Estimator Formula {.smaller .quiz-question}

A portfolio manager regresses fund returns ($Y$) on a factor ($x$) with $n = 12$ monthly observations. She computes $S_{xy} = 84$ and $S_{xx} = 120$. What is $\hat\beta_1$?

- [$\hat\beta_1 = S_{xy}/S_{xx} = 84/120 = 0.70$]{.correct data-explanation="✅ Correct! The LS slope estimator is β̂₁ = Sxy/Sxx = 84/120 = 0.70. This means a 1-unit increase in the factor is associated with a 0.70 unit increase in fund return."}
- $\hat\beta_1 = S_{xx}/S_{xy} = 120/84 = 1.429$
- $\hat\beta_1 = \sqrt{S_{xy}/S_{xx}} = \sqrt{0.70} = 0.837$
- $\hat\beta_1 = S_{xy} \cdot S_{xx} = 84 \times 120 = 10080$

---

## 📝 Quiz #2: Degrees of Freedom for $S^2$ {.smaller .quiz-question}

In simple linear regression with $n = 25$ observations, the SSE is computed to be 48.0. What is the unbiased estimator $S^2$ of the error variance $\sigma^2$?

- [$S^2 = \text{SSE}/(n-2) = 48/(25-2) = 48/23 \approx 2.087$]{.correct data-explanation="✅ Correct! In SLR we lose 2 degrees of freedom estimating β₀ and β₁. So S² = SSE/(n−2) = 48/23 ≈ 2.087."}
- $S^2 = 48/(25-1) = 48/24 = 2.000$
- $S^2 = 48/25 = 1.920$
- $S^2 = 48/(25-3) = 48/22 \approx 2.182$

---

## 📝 Quiz #3: Interpreting $r^2$ {.smaller .quiz-question}

A regression of quarterly GDP growth on broadband penetration gives $r^2 = 0.63$. Which is the best interpretation?

- [63% of the variation in quarterly GDP growth is explained by its linear relationship with broadband penetration]{.correct data-explanation="✅ Correct! r² = 1 − SSE/Syy is the proportion of total variation in Y explained by the linear model. Here, 63% of GDP variation is accounted for by penetration."}
- The correlation between GDP growth and penetration is $0.63$
- GDP growth increases by 63% for each unit increase in penetration
- The model predicts correctly 63% of the time

---

## 📝 Quiz #4: $t$-Test for Slope {.smaller .quiz-question}

For a CAPM regression with $n = 36$ monthly returns, $\hat\beta_1 = 1.25$, $S = 0.042$, $S_{xx} = 0.18$. What is the $t$-statistic for $H_0: \beta_1 = 1$ (testing if beta equals 1)?

- [$t = (1.25 - 1.00)/(0.042/\sqrt{0.18}) = 0.25/0.099 = 2.525$, df $= 34$]{.correct data-explanation="✅ Correct! t = (β̂₁ − β₁₀)/(S/√Sxx) = (1.25−1.00)/(0.042/√0.18) = 0.25/0.0990 = 2.525 with df = n−2 = 34."}
- $t = 1.25/(0.042/\sqrt{0.18}) = 12.63$
- $t = (1.25 - 1.00)/0.042 = 5.952$
- $t = 0.25 \times \sqrt{36}/0.042 = 35.7$

---

## 📝 Summary {.smaller}

::: {style="font-size: 37px"}
::: {.summary-box}
**✅ Key Takeaways**

- **Model**: $Y = \beta_0 + \beta_1 x + \varepsilon$ with $E(\varepsilon) = 0$, $V(\varepsilon) = \sigma^2$; normality needed for $t$-inference

- **LS Estimators**: $\hat\beta_1 = S_{xy}/S_{xx}$, $\hat\beta_0 = \bar Y - \hat\beta_1 \bar x$; both unbiased; $V(\hat\beta_1) = \sigma^2/S_{xx}$

- **Residual Variance**: $S^2 = \text{SSE}/(n-2)$ where $\text{SSE} = S_{yy} - \hat\beta_1 S_{xy}$; uses $n-2$ df

- **$t$-Test for slope**: $T = \hat\beta_1/(S/\sqrt{S_{xx}}) \sim t_{n-2}$; rejecting $H_0: \beta_1 = 0$ means $x$ is a significant linear predictor

- **Goodness of fit**: $r^2 = 1 - \text{SSE}/S_{yy}$ — proportion of variation in $Y$ explained by $x$; $r = S_{xy}/\sqrt{S_{xx} S_{yy}}$ is Pearson correlation
:::
:::

---

## 📚 Practice Problems

::: {style="font-size:47px;"}
::: {.callout-tip}
## 📝 Homework Problems — Chapter 11 (§11.1–11.6)

**Problem 1 (LS by hand):** An analyst has 6 monthly observations: $\sum x = 24$, $\sum Y = 18$, $\sum x^2 = 110$, $\sum xY = 76$, $\sum Y^2 = 62$, $n = 6$. Compute $\hat\beta_0$, $\hat\beta_1$, SSE, $S^2$, and $r^2$.

**Problem 2 ($t$-test for slope):** Using the results of Problem 1, test $H_0: \beta_1 = 0$ at $\alpha = 0.05$. Construct a 95% CI for $\beta_1$ and interpret it in a financial context of your choice.

**Problem 3 (CAPM interpretation):** A regression of stock excess returns on market excess returns gives $\hat\beta_1 = 1.35$, SE$(\hat\beta_1) = 0.21$, $r^2 = 0.82$, $n = 48$. Is the stock significantly different from a "market-neutral" position ($\beta_1 = 1$)? Test at $\alpha = 0.05$.

**Problem 4 (Elasticity):** Regress broadband penetration ($Y$\%) on GDP per capita ($x$, USD thousands) for 20 countries: $S_{xx} = 8{,}400$, $S_{xy} = 1{,}260$, $S_{yy} = 320$. Find $\hat\beta_1$, $r^2$, and test the slope. Interpret the elasticity result for a policy audience.
:::
:::

---

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business, ADA University

📧 [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Multiple Regression & Model Diagnostics (§11.11–11.14)

**Reading:** Chapter 11, Sections 11.11–11.14

**Preparation:** Review matrix notation basics; think about what "controlling for other variables" means

**⏰ Reminders:**

✅ Complete Practice Problems 1–4

✅ Run the CAPM regression on a stock of your choice in R

✅ Verify the equivalence of the $t$-test for $\beta_1$ and the $t$-test for $\rho$

✅ Work hard!
:::
:::

---

## ❓ Questions? {.center}

::: {.callout-note}
## 💬 Open Discussion

**Key Topics for Discussion:**

- Why does the LS line always pass through $(\bar x, \bar Y)$? What does this imply geometrically?

- In CAPM, $\beta_0$ represents Jensen's alpha. What would it mean statistically if $H_0: \beta_0 = 0$ is rejected?

- The same $t$-statistic tests both $H_0: \beta_1 = 0$ and $H_0: \rho = 0$. How can two different null hypotheses lead to the same test?

- You compute $r^2 = 0.95$ for a regression of GDP on broadband penetration over time. Should you be impressed? What hidden problem might this signal?
:::

<div class="attendance-qr" data-status="late" id="qr-late" style="min-height: 400px; display: flex; align-items: center; justify-content: center;"></div>
