---
title: "Mathematical Statistics"
subtitle: "The Law of Total Probability and Bayes' Rule"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - Total Probability and Bayes' Rule"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
library(ggplot2)
```

## 🎯 Learning Objectives

::: {style="font-size: 32px"}
By the end of this lecture, you will be able to:

- **Recognise** a partition of $S$ (Definition 2.11) and choose one that makes the conditionals easy

- **Compute** $P(A)$ by the law of total probability (Theorem 2.8), branch by branch

- **Invert** a conditional with Bayes' rule (Theorem 2.9), reading the answer off a tree diagram

- **Explain** why an accurate screen for a rare event produces mostly false alarms

- **Update** a probability twice, using two conditionally independent signals in sequence
:::

---

## 🗺️ Where We Are

::: {style="font-size: 32px"}
**Wackerly §2.10**

Last class ended on this: *what to do when the event depends on which state the world is in* — the law of total probability and Bayes' rule.

::: {.fragment}
A loan's chance of default depends on its rating grade. A flagged card payment is fraud more often if fraud is common. In both cases $P(A \mid \text{state})$ is easy; $P(A)$ is not.
:::

::: {.fragment}
Today: **add up over the states** to get $P(A)$, then **run the conditional backwards** to learn the state from $A$.
:::
:::

---

## 💳 A Question Before Any Notation

::: {style="font-size: 32px"}
A Baku bank screens every card payment with a fraud model.

- It flags **95%** of fraudulent payments
- It clears **97%** of legitimate ones
- About **1 payment in 500** is fraudulent

::: {.fragment}
A payment has just been flagged. **Is it probably fraud?**

Write down a number now. We will come back to it.
:::
:::

---

## 📐 Definition 2.11: A Partition

::: {style="font-size: 30px"}
::: {.callout-important}
## Definition 2.11

For some positive integer $k$, let the sets $B_1, B_2, \ldots, B_k$ be such that

1. $S = B_1 \cup B_2 \cup \cdots \cup B_k$
2. $B_i \cap B_j = \emptyset$, for $i \neq j$.

Then the collection $\{B_1, B_2, \ldots, B_k\}$ is said to be a **partition** of $S$.
:::

::: {.fragment}
Every loan has **exactly one** rating grade: the grades cover the book and never overlap. "Rated A", "rated B" and "in arrears" are not a partition — a loan can be B and in arrears.
:::
:::

---

## 📐 Theorem 2.8: Total Probability

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorem 2.8

If $\{B_1, \ldots, B_k\}$ is a partition of $S$ with $P(B_i) > 0$ for every $i$, then for any event $A$
$$P(A) = \sum_{i=1}^{k} P(A \mid B_i)\,P(B_i)$$
:::

::: {.fragment}
**Proof in one line.** $A = (A \cap B_1) \cup \cdots \cup (A \cap B_k)$, and these pieces are mutually exclusive, so Axiom 3 adds them. Theorem 2.5 writes each piece as $P(A \mid B_i)P(B_i)$.
:::
:::

---

## 🏦 Default Across Three Grades

::: {style="font-size: 28px"}
A bank's SME loan book is split by internal rating grade. $D$ is the event that a loan defaults within the year.

| Grade | Share of book $P(B_i)$ | Default rate $P(D \mid B_i)$ | Product |
|---|---|---|---|
| A | 0.50 | 0.01 | 0.005 |
| B | 0.35 | 0.04 | 0.014 |
| C | 0.15 | 0.12 | 0.018 |
| | **1.00** | | **0.037** |

::: {.fragment}
$$P(D) = 0.005 + 0.014 + 0.018 = 0.037$$
The book-wide default rate, 3.7%, is a **weighted average** of the grade rates, with the grade shares as weights.
:::
:::

---

## 📐 Theorem 2.9: Bayes' Rule

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorem 2.9 (Bayes' Rule)

If $\{B_1, \ldots, B_k\}$ is a partition of $S$ with $P(B_i) > 0$ for every $i$, then
$$P(B_j \mid A) = \frac{P(A \mid B_j)\,P(B_j)}{\sum_{i=1}^{k} P(A \mid B_i)\,P(B_i)}$$
:::

::: {.fragment}
The numerator is **one row** of the table: $P(A \cap B_j)$. The denominator is **the total**: $P(A)$ from Theorem 2.8. Bayes' rule is Definition 2.9 with both pieces filled in.
:::
:::

---

## 🔄 Running the Loan Book Backwards

::: {style="font-size: 30px"}
A loan has defaulted. Which grade did it come from?

$$P(C \mid D) = \frac{0.018}{0.037} = 0.486, \quad P(B \mid D) = \frac{0.014}{0.037} = 0.378, \quad P(A \mid D) = \frac{0.005}{0.037} = 0.135$$

::: {.fragment}
Grade C is **15%** of the book but **49%** of the defaults. The three posteriors sum to 1 because they divide one partition's products by their own total.
:::

::: {.fragment}
**Prior** $P(B_j)$: what you believed before the default. **Posterior** $P(B_j \mid D)$: what you believe after.
:::
:::

---

## 🌳 Example 2.23: A Card Batch

::: {style="font-size: 29px"}
A card manufacturer personalises chip cards on **five lines** at equal rates. Normally 2% of chips are defective. In March **line 1** malfunctioned and produced 5% defectives.

A bank receives a March batch of 100 cards, tests three, and **exactly one fails**. What is the probability the batch came from line 1?

::: {.fragment}
Let $B$ = "batch from line 1" and $A$ = "exactly one of three fails". Then $P(B) = 0.2$, $P(\bar{B}) = 0.8$, and
$$P(A \mid B) = 3(0.05)(0.95)^2 = 0.135375, \qquad P(A \mid \bar{B}) = 3(0.02)(0.98)^2 = 0.057624$$
:::
:::

---

## 🌳 The Tree

```{r}
#| label: card-tree
#| echo: false
#| fig-width: 11
#| fig-height: 4.6
pB  <- 0.2
a1  <- 3 * 0.05 * 0.95^2
a0  <- 3 * 0.02 * 0.98^2
nodes <- data.frame(
  x = c(0, 1, 1, 2, 2, 2, 2),
  y = c(0, 1.5, -1.5, 2.25, 0.75, -0.75, -2.25),
  lab = c("", "B: line 1", "~B: lines 2-5", "A", "~A", "A", "~A"))
edges <- data.frame(
  x = c(0, 0, 1, 1, 1, 1), y = c(0, 0, 1.5, 1.5, -1.5, -1.5),
  xend = c(1, 1, 2, 2, 2, 2), yend = c(1.5, -1.5, 2.25, 0.75, -0.75, -2.25),
  p = sprintf("%.4f", c(pB, 1 - pB, a1, 1 - a1, a0, 1 - a0)))
joint <- c(pB * a1, pB * (1 - a1), (1 - pB) * a0, (1 - pB) * (1 - a0))
leaves <- data.frame(x = 2.55, y = c(2.25, 0.75, -0.75, -2.25),
  lab = sprintf("%.4f", joint), hit = c(TRUE, FALSE, TRUE, FALSE))
ggplot() +
  geom_segment(data = edges, aes(x, y, xend = xend, yend = yend),
               linewidth = 1, colour = "grey45") +
  geom_label(data = edges, aes((x + xend) / 2, (y + yend) / 2, label = p),
             size = 6.5, label.size = 0, fill = "white") +
  geom_label(data = nodes[-1, ], aes(x, y, label = lab), size = 7,
             fill = "#eef2f8") +
  geom_text(data = leaves, aes(x, y, label = lab,
            colour = hit, fontface = ifelse(hit, "bold", "plain")), size = 7) +
  scale_colour_manual(values = c(`TRUE` = "#8b2635", `FALSE` = "grey35"),
                      guide = "none") +
  annotate("text", x = 2.55, y = 3, label = "P(branch)", size = 6.5) +
  xlim(-0.1, 2.85) + theme_void()
```

::: {style="font-size: 28px"}
Multiply **along** a path (Thm 2.5); add **across** the paths ending in $A$ (Thm 2.8).
:::

---

## 🌳 Reading the Answer Off the Tree

::: {style="font-size: 30px"}
The two red leaves are the paths that end in $A$:
$$P(A) = 0.0271 + 0.0461 = 0.0732$$

::: {.fragment}
Bayes' rule is **one red leaf over both red leaves**:
$$P(B \mid A) = \frac{(0.135375)(0.2)}{0.0731742} = 0.37, \qquad P(\bar{B} \mid A) = 1 - 0.37 = 0.63$$
:::

::: {.fragment}
One failure in three nearly doubles the chance of line 1, from 0.20 to 0.37 — and still leaves the other four lines the likelier source.
:::
:::

---

## ⚠️ Back to the Fraud Screen

::: {style="font-size: 30px"}
$F$ = fraud, $T$ = flagged. $P(F) = 0.002$, $P(T \mid F) = 0.95$, $P(T \mid \bar{F}) = 0.03$.
$$P(F \mid T) = \frac{0.95 \times 0.002}{0.95 \times 0.002 + 0.03 \times 0.998} = \frac{0.0019}{0.03184} = 0.060$$

::: {.fragment}
**Six percent.** In 100,000 payments: 200 are fraud and 190 of them are flagged; 99,800 are legitimate and **2,994** of them are flagged. Of 3,184 flags, 190 are fraud.
:::

::: {.fragment}
This is the **base-rate paradox**: the screen is accurate, but the legitimate pool is so large that its 3% error outnumbers all the fraud there is.
:::
:::

---

## 🔬 Interactive: Posterior vs Base Rate {.smaller}

```{ojs}
//| echo: false

viewof base_pct = {
  const input = Inputs.range([0.1, 50], {value: 0.2, step: 0.1,
    label: "Base rate of fraud (%):"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}
```

```{ojs}
//| echo: false

sens = 0.95
fpr = 0.03
posterior = p => p * sens / (p * sens + (1 - p) * fpr)
grid = Array.from({length: 121}, (_, i) => {
  const pct = 0.1 * Math.pow(500, i / 120);
  return {pct: pct, post: posterior(pct / 100)};
})
current = [{pct: base_pct, post: posterior(base_pct / 100)}]

md`Screen fixed at 95% / 97%. At a base rate of **${base_pct.toFixed(1)}%**, a flag means fraud with probability **${(100 * current[0].post).toFixed(1)}%**.`

Plot.plot({
  width: 1150,
  height: 310,
  marginLeft: 78,
  marginTop: 40,
  marginBottom: 58,
  style: {fontSize: "18px"},
  x: {type: "log", domain: [0.1, 50], label: "Base rate of fraud (%, log scale)",
      ticks: [0.1, 0.2, 0.5, 1, 2, 5, 10, 20, 50], tickFormat: d => String(d)},
  y: {label: "P(fraud | flagged)", domain: [0, 1], tickFormat: ".1f"},
  marks: [
    Plot.ruleY([0.5], {stroke: "#cbb8a9", strokeDasharray: "4 4"}),
    Plot.line(grid, {x: "pct", y: "post", stroke: "#14130f", strokeWidth: 2}),
    Plot.dot(current, {x: "pct", y: "post", r: 9, fill: "#8b2635"}),
    Plot.ruleY([0])
  ]
})
```

---

## 🔧 What Would Actually Help?

```{r}
#| label: screen-designs
posterior <- function(base, sens, fpr) base * sens / (base * sens + (1 - base) * fpr)

data.frame(
  design    = c("current", "sensitivity 99.9%", "false alarms 1%",
                "false alarms 0.1%", "base rate 2%"),
  base      = c(0.002, 0.002, 0.002, 0.002, 0.02),
  sens      = c(0.95, 0.999, 0.95, 0.95, 0.95),
  fpr       = c(0.03, 0.03, 0.01, 0.001, 0.03)
) |> transform(posterior = round(posterior(base, sens, fpr), 3))
```

::: {style="font-size: 28px"}
Catching more fraud barely moves it. **Cutting false alarms** does — the denominator is where the legitimate pool lives. The posterior crosses 50% at a base rate of $0.03/0.98 = 3.1\%$.
:::

---

## 🔁 Sequential Updating

::: {style="font-size: 29px"}
Flagged payments go to a **device check**: it fails for 90% of fraudulent payments and 5% of legitimate ones. Given fraud status, the two signals are **conditionally independent**.

::: {.fragment}
**Today's posterior is tomorrow's prior.** After the flag, $P(F) = 0.0597$. Apply Bayes' rule again with the device check:
$$P(F \mid T \cap V) = \frac{0.90 \times 0.0597}{0.90 \times 0.0597 + 0.05 \times 0.9403} = 0.533$$
:::

::: {.fragment}
Or do it in one step from the start, multiplying the conditionals (independence given $F$):
$$\frac{0.002 \times 0.95 \times 0.90}{0.002 \times 0.95 \times 0.90 + 0.998 \times 0.03 \times 0.05} = \frac{0.00171}{0.003207} = 0.533$$
:::
:::

---

## 💻 Two Routes, One Answer

```{r}
#| label: sequential
bayes <- function(prior, hit, false_alarm) {
  prior * hit / (prior * hit + (1 - prior) * false_alarm)
}
p0 <- 0.002                           # base rate
p1 <- bayes(p0, 0.95, 0.03)           # after the model flags it
p2 <- bayes(p1, 0.90, 0.05)           # after the device check fails too
one_step <- bayes(p0, 0.95 * 0.90, 0.03 * 0.05)
round(c(prior = p0, after_flag = p1, after_both = p2, one_step = one_step), 4)
```

::: {style="font-size: 28px"}
The order of the checks does not matter, and neither does whether you stop in between. Without conditional independence the one-step route would need $P(V \mid T \cap F)$, not $P(V \mid F)$.
:::

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 30px"}
A mobile operator's subscribers are **60% prepaid**, **30% postpaid** and **10% business**. Monthly churn rates are 6%, 2% and 1%.

**Four minutes, in pairs:**

1. What share of all subscribers churn this month?

2. A subscriber has just churned. What is the probability they were prepaid?

3. A subscriber stayed. What is the probability they are business?
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 30px"}
1. Let $L$ = the subscriber leaves. Theorem 2.8 over the three plans:
$$P(L) = 0.6(0.06) + 0.3(0.02) + 0.1(0.01) = 0.036 + 0.006 + 0.001 = 0.043$$

2. Theorem 2.9: $P(\text{prepaid} \mid L) = 0.036 / 0.043 = 0.837$.

::: {.fragment}
3. The same partition, with $\bar{L}$ in place of $L$. $P(\bar{L}) = 1 - 0.043 = 0.957$ and
$$P(\text{business} \mid \bar{L}) = \frac{0.1 \times 0.99}{0.957} = 0.103$$
Staying tells you almost nothing — nearly everyone stays — so the posterior barely moves off the 10% prior.
:::
:::

---

## 📝 Quiz #1: Total Probability {.quiz-question}

$\{B_1, B_2\}$ is a partition with $P(B_1) = 0.7$. $P(A \mid B_1) = 0.1$ and $P(A \mid B_2) = 0.4$. What is $P(A)$?

- [$0.19$]{.correct data-explanation="✅ Theorem 2.8: 0.7 × 0.1 + 0.3 × 0.4 = 0.07 + 0.12 = 0.19. The conditionals are weighted by the partition probabilities, not averaged equally."}
- $0.25$
- $0.50$
- $0.28$

---

## 📝 Quiz #2: Bayes' Rule {.quiz-question}

Same partition as Quiz #1. Given that $A$ occurred, what is $P(B_2 \mid A)$?

- [$0.63$]{.correct data-explanation="✅ Theorem 2.9: P(B₂ | A) = 0.12 / 0.19 = 0.63. The numerator is B₂'s branch of the tree; the denominator is P(A) from Quiz #1."}
- $0.40$
- $0.12$
- $0.30$

---

## 📝 Quiz #3: The Base-Rate Paradox {.quiz-question}

A screen for a rare event (base rate 0.2%) has 95% sensitivity and a 3% false-alarm rate. Which change raises $P(\text{event} \mid \text{flag})$ the most?

- Raising sensitivity to 99.9%
- [Cutting the false-alarm rate to 0.1%]{.correct data-explanation="✅ The posterior rises from 0.06 to 0.66. False alarms come from the huge non-event pool, which is the denominator's dominant term; sensitivity only acts on the tiny event pool."}
- Screening twice as many payments
- Raising sensitivity to 100%

---

## 📋 Key Formulas

::: {style="font-size: 30px"}
| | Statement |
|---|---|
| Definition 2.11 | $S = B_1 \cup \cdots \cup B_k$, $\; B_i \cap B_j = \emptyset$ for $i \neq j$ |
| decomposition | $A = (A \cap B_1) \cup \cdots \cup (A \cap B_k)$ |
| Theorem 2.8 | $P(A) = \sum_{i=1}^k P(A \mid B_i)P(B_i)$ |
| Theorem 2.9 | $P(B_j \mid A) = \dfrac{P(A \mid B_j)P(B_j)}{\sum_{i=1}^k P(A \mid B_i)P(B_i)}$ |
| two signals, independent given $B_j$ | $P(B_j \mid A_1 \cap A_2) \propto P(A_1 \mid B_j)P(A_2 \mid B_j)P(B_j)$ |
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- A partition splits $S$ into states that cover everything and never overlap

- Total probability is a **weighted average** of the conditionals, weighted by the states

- Bayes' rule is one branch of the tree divided by all the branches that end in $A$

- For a rare event, even an accurate screen gives a small posterior: **the base rate dominates**

- False alarms, not missed detections, are what a rare-event screen must cut

- Updating is sequential: yesterday's posterior is today's prior
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition**

- Exercises at the end of §2.10 — start with 2.124, 2.125 and 2.129, then 2.126 – 2.127 on sensitivity and specificity, and 2.132

- Re-draw today's card-batch tree with $P(A \mid B)$ lowered to 0.12, and find $P(B \mid A)$ (compare Exercise 2.123(b))

**Week 4, Problem Set 2** opens today and closes **10 October at 23:59** on WeBWorK, covering §2.10.

**Next class:** turning outcomes into numbers — random variables, random sampling, and a synthesis of Chapter 2 (§§2.11–2.13).
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**

📧 sorujov@ada.edu.az\
🏢 Building D, Room D325\
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- A flag on the fraud screen raised the probability from 0.2% to 6%. Is that a failure of the screen, or a success?

- If the two fraud signals were positively correlated among legitimate payments, would the one-step answer be too high or too low?

- Why must every $P(B_i)$ be strictly positive in Theorems 2.8 and 2.9?
:::
