---
title: "Mathematical Statistics"
subtitle: "The Normal Probability Distribution"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics - Normal Distribution"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::: {style="font-size: 30px"}
::: {.learning-objectives}
By the end of this lecture, you will be able to:

- Define the normal probability distribution and identify its parameters (mean $\mu$ and standard deviation $\sigma$)

- Calculate probabilities using the standard normal distribution (z-scores) and interpret standard normal tables

- Apply the Empirical Rule (68-95-99.7 rule) to estimate probabilities and identify outliers in financial datasets

- Transform any normal random variable to standard normal form using z-score standardization

- Solve real-world business and finance problems involving normally distributed variables (stock returns, quality control, risk assessment)
:::
:::

## 📋 Overview {.smaller}

::: {style="font-size:38px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Normal Distribution Definition** – The bell-shaped curve and its mathematical formulation

- **Properties and Parameters** – Mean, variance, symmetry, and the role of $\mu$ and $\sigma$

- **Standard Normal Distribution** – Z-scores, standardization, and probability calculations

- **Empirical Rule** – Understanding the 68-95-99.7 rule for data interpretation

- **Applications** – Stock returns, quality control, risk management, and financial modeling
:::
:::
:::

## 📖 Definition: The Normal Probability Distribution {.smaller}

::: {.callout-note}
## 📝 Definition 1: Normal Distribution

A random variable $Y$ is said to have a **normal probability distribution** if and only if, for $\sigma > 0$ and $-\infty < \mu < \infty$, the probability density function (pdf) of $Y$ is:

$$f(y) = \frac{1}{\sigma \sqrt{2\pi}} e^{-\frac{(y-\mu)^2}{2\sigma^2}}, \quad -\infty < y < \infty$$

We denote this as $Y \sim N(\mu, \sigma^2)$.

**Key Properties:**

- **Shape**: Bell-shaped (symmetric around the mean)

- **Parameters**: $\mu$ (mean, location parameter) and $\sigma$ (standard deviation, scale parameter)

- **Support**: The entire real line ($-\infty$ to $\infty$)

- **Uniqueness**: Completely determined by $\mu$ and $\sigma^2$
:::

**Why It Matters in Finance:** Most asset returns, measurement errors, and aggregated random variables approximately follow normal distributions (portfolio returns, quality metrics, standardized test scores) [web:2][web:6].

## 🧮 Theorem: Expected Value and Variance {.smaller}

::: {.callout-important}
## Theorem 1: Mean and Variance of Normal Distribution

If $Y$ is a normally distributed random variable with parameters $\mu$ and $\sigma$, then:

$$E(Y) = \mu$$

and

$$V(Y) = \sigma^2$$

**Interpretation:**

- The parameter $\mu$ directly equals the expected value (center of the distribution)

- The parameter $\sigma^2$ equals the variance (spread of the distribution)

- Standard deviation $\sigma$ measures the typical deviation from the mean
:::

**Financial Context:** For a stock with daily returns $Y \sim N(0.08\%, 1.5\%)$, the expected daily return is $\mu = 0.08\%$ and the risk (volatility) is measured by $\sigma = 1.5\%$ [web:7].

## 📖 Definition: Standard Normal Distribution {.smaller}

::: {.callout-note}
## 📝 Definition 2: Standard Normal Distribution

A **standard normal random variable** $Z$ is a normal random variable with mean $\mu = 0$ and standard deviation $\sigma = 1$:

$$Z \sim N(0, 1)$$

The pdf simplifies to:

$$f(z) = \frac{1}{\sqrt{2\pi}} e^{-\frac{z^2}{2}}, \quad -\infty < z < \infty$$

**Z-Score Transformation:**

Any normal random variable $Y \sim N(\mu, \sigma^2)$ can be transformed to standard normal:

$$\boxed{Z = \frac{Y - \mu}{\sigma}}$$

This transformation measures how many standard deviations $Y$ is from its mean [web:4].
:::

## 🎓 The Empirical Rule (68-95-99.7 Rule) {.smaller}

::: {.callout-tip}
## 📊 Rule: The Empirical Rule for Normal Distributions

For any normal distribution $Y \sim N(\mu, \sigma^2)$:

- **68%** of values lie within **1 standard deviation** of the mean: $P(\mu - \sigma \leq Y \leq \mu + \sigma) \approx 0.68$

- **95%** of values lie within **2 standard deviations** of the mean: $P(\mu - 2\sigma \leq Y \leq \mu + 2\sigma) \approx 0.95$

- **99.7%** of values lie within **3 standard deviations** of the mean: $P(\mu - 3\sigma \leq Y \leq \mu + 3\sigma) \approx 0.997$

**For Standard Normal ($Z$):**

- $P(-1 \leq Z \leq 1) \approx 0.68$

- $P(-2 \leq Z \leq 2) \approx 0.95$

- $P(-3 \leq Z \leq 3) \approx 0.997$
:::

**Risk Management Application:** If stock returns are $N(8\%, 5\%)$, 95% of returns fall between $8\% - 2(5\%) = -2\%$ and $8\% + 2(5\%) = 18\%$ [web:3][web:10].

## 🤝 Think-Pair-Share: Applying the Empirical Rule {.smaller}

::: {style="font-size: 28px"}
::: {.callout-note}
## 💭 Student Engagement Activity (5 minutes)

**Scenario:** A financial analyst reports that a stock's daily returns follow a normal distribution with mean 0.5% and standard deviation 2%.

**Think (1 minute):** Work individually

- What percentage of days should we expect returns below -3.5%?

- If the stock market is open 252 days per year, approximately how many days per year would you expect returns beyond ±4% (either very positive or very negative)?

- Would a daily return of +5% be considered unusual? Why or why not?

**Pair (2 minutes):** Discuss your answers with a partner

- Compare your calculations

- Discuss the business implications of your findings

**Share (2 minutes):** Class discussion

- Selected pairs share their conclusions

- Discuss how this relates to portfolio risk management and investor expectations
:::
:::

## 📌 Example 1: Standard Normal Probabilities {.smaller}

**Problem:** Let $Z$ denote a standard normal random variable with mean 0 and standard deviation 1.

a) Find $P(Z > 2)$

b) Find $P(-2 \leq Z \leq 2)$

c) Find $P(0 \leq Z \leq 1.73)$

**Solution:**

a) **Finding $P(Z > 2)$:**

Using the standard normal table, we look up $z = 2.0$. The table gives the upper tail area: $A(2.0) = 0.0228$.

Thus, $\boxed{P(Z > 2) = 0.0228}$ or about 2.28%.

**Interpretation:** Only 2.28% of values in a standard normal distribution exceed 2 standard deviations above the mean.

## 📌 Example 1: Solution (continued) {.smaller}

b) **Finding $P(-2 \leq Z \leq 2)$:**

From part (a), we know $P(Z > 2) = A(2.0) = 0.0228$.

By symmetry of the normal distribution, $P(Z < -2) = A(2.0) = 0.0228$.

Therefore:
$$P(-2 \leq Z \leq 2) = 1 - P(Z < -2) - P(Z > 2) = 1 - 2(0.0228) = 0.9544$$

$\boxed{P(-2 \leq Z \leq 2) = 0.9544}$ or about 95.44%.

**This confirms the Empirical Rule!** About 95% of data falls within 2 standard deviations.

c) **Finding $P(0 \leq Z \leq 1.73)$:**

Since $P(Z > 0) = 0.5$ (by symmetry), we have:
$$P(0 \leq Z \leq 1.73) = 0.5 - P(Z > 1.73) = 0.5 - A(1.73)$$

From the z-table, $A(1.73) = 0.0418$.

Thus, $\boxed{P(0 \leq Z \leq 1.73) = 0.5 - 0.0418 = 0.4582}$ or about 45.82%.

## 📌 Example 2: College Entrance Examination Scores {.smaller}

**Problem:** The achievement scores for a college entrance examination are normally distributed with a mean of 75 and a standard deviation of 10. What fraction of the scores lies between 80 and 90?

**Solution:**

Given: $Y \sim N(\mu = 75, \sigma = 10)$

We need to find $P(80 \leq Y \leq 90)$.

**Step 1: Standardize using z-scores**

Recall that $z = \frac{y - \mu}{\sigma}$ converts any normal variable to standard normal.

For $y_1 = 80$:
$$z_1 = \frac{80 - 75}{10} = \frac{5}{10} = 0.5$$

For $y_2 = 90$:
$$z_2 = \frac{90 - 75}{10} = \frac{15}{10} = 1.5$$

## 📌 Example 2: Solution (continued) {.smaller}

**Step 2: Find probability using standard normal**

We need $P(0.5 \leq Z \leq 1.5)$.

$$P(0.5 \leq Z \leq 1.5) = P(Z > 0.5) - P(Z > 1.5) = A(0.5) - A(1.5)$$

From the z-table:

- $A(0.5) = 0.3085$

- $A(1.5) = 0.0668$

Therefore:
$$P(80 \leq Y \leq 90) = 0.3085 - 0.0668 = 0.2417$$

$\boxed{P(80 \leq Y \leq 90) = 0.2417}$ or about **24.17%** of students score between 80 and 90.

**Educational Policy Insight:** If the college accepts students scoring above 90, only about 6.68% qualify (upper tail beyond $z = 1.5$).

## 🎮 Interactive: Normal Distribution Explorer {.smaller}

::: {style="font-size: 0.8em;"}

**Explore Normal Distributions:** Adjust the mean and standard deviation to see how they affect the shape and probabilities.

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof mu = Inputs.range([-10, 10], {
  value: 0, 
  step: 0.5, 
  label: "Mean (μ):"
})

viewof sigma = Inputs.range([0.5, 5], {
  value: 1, 
  step: 0.25, 
  label: "Std Dev (σ):"
})

// Calculate key probabilities
prob_1sd = 0.6827
prob_2sd = 0.9545
prob_3sd = 0.9973

md`**Current Parameters:**  
μ = ${mu.toFixed(2)}  
σ = ${sigma.toFixed(2)}  

**Empirical Rule:**  
P(μ ± σ) = ${(prob_1sd * 100).toFixed(1)}%  
P(μ ± 2σ) = ${(prob_2sd * 100).toFixed(1)}%  
P(μ ± 3σ) = ${(prob_3sd * 100).toFixed(1)}%`
```

**Observations:**  
- Increasing σ makes the curve wider (more spread)
- Changing μ shifts the curve left/right
- The area under the curve always equals 1

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// Normal PDF function
function normalPDF(x, mu, sigma) {
  const coef = 1 / (sigma * Math.sqrt(2 * Math.PI));
  const exponent = -Math.pow(x - mu, 2) / (2 * Math.pow(sigma, 2));
  return coef * Math.exp(exponent);
}

// Generate x values centered around mu
x_range = d3.range(mu - 4 * sigma, mu + 4 * sigma + 0.1, 0.1)

// Calculate PDF values
pdf_data = x_range.map(x => ({
  x: x,
  pdf: normalPDF(x, mu, sigma)
}))

// Shaded regions for empirical rule
region_1sd = x_range.filter(x => x >= mu - sigma && x <= mu + sigma).map(x => ({
  x: x,
  pdf: normalPDF(x, mu, sigma)
}))

region_2sd = x_range.filter(x => x >= mu - 2*sigma && x <= mu + 2*sigma).map(x => ({
  x: x,
  pdf: normalPDF(x, mu, sigma)
}))

Plot.plot({
  width: 800,
  height: 450,
  marginLeft: 60,
  marginBottom: 50,
  x: {
    label: "Y (random variable)",
    grid: true
  },
  y: {
    label: "Probability Density f(y)",
    domain: [0, normalPDF(mu, mu, sigma) * 1.1]
  },
  marks: [
    Plot.areaY(region_2sd, {
      x: "x", 
      y: "pdf", 
      fill: "lightblue", 
      opacity: 0.3
    }),
    Plot.areaY(region_1sd, {
      x: "x", 
      y: "pdf", 
      fill: "steelblue", 
      opacity: 0.4
    }),
    Plot.line(pdf_data, {
      x: "x", 
      y: "pdf", 
      stroke: "darkblue", 
      strokeWidth: 3
    }),
    Plot.ruleX([mu], {
      stroke: "red", 
      strokeWidth: 2, 
      strokeDasharray: "5,5"
    }),
    Plot.ruleY([0])
  ],
  caption: html`<span style="color: darkblue; font-weight: bold;">━━</span> N(μ=${mu.toFixed(1)}, σ=${sigma.toFixed(1)}) | 
    <span style="color: steelblue;">█</span> ±1σ (68%) | 
    <span style="color: lightblue;">█</span> ±2σ (95%) | 
    <span style="color: red;">┊</span> Mean`
})
```

:::

:::

:::

## 📌 Example 3: The Median of Normal Distribution {.smaller}

**Problem:** What is the median of a normally distributed random variable with mean $\mu$ and standard deviation $\sigma$?

**Solution:**

Recall that the **median** $\phi_{0.5}$ of a random variable is its 0.5th quantile. By definition, the median is the smallest value such that:

$$F(\phi_{0.5}) \geq 0.5$$

which means:

$$\int_{-\infty}^{\phi_{0.5}} f(y) \, dy \geq 0.5$$

For the normal distribution, the density function $f(y)$ is **symmetric about $\mu$**. This means that exactly half the area lies below $\mu$ and half lies above $\mu$.

Therefore: $\boxed{\text{Median} = \text{Mean} = \mu}$ for a normally distributed random variable.

**Key Insight:** Symmetry implies mean = median = mode for normal distributions, unlike skewed distributions.

## 🎮 Interactive: Comparing Normal Distributions {.smaller}

::: {style="font-size: 0.8em;"}

**Compare Two Distributions:** See how different parameters create different bell curves.

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof mu1 = Inputs.range([-5, 5], {
  value: 0, 
  step: 0.5, 
  label: "μ₁ (Distribution 1):"
})

viewof sigma1 = Inputs.range([0.5, 3], {
  value: 1, 
  step: 0.25, 
  label: "σ₁ (Distribution 1):"
})

viewof mu2 = Inputs.range([-5, 5], {
  value: 2, 
  step: 0.5, 
  label: "μ₂ (Distribution 2):"
})

viewof sigma2 = Inputs.range([0.5, 3], {
  value: 1.5, 
  step: 0.25, 
  label: "σ₂ (Distribution 2):"
})

md`**Distribution 1:**  
N(${mu1.toFixed(1)}, ${sigma1.toFixed(2)}²)

**Distribution 2:**  
N(${mu2.toFixed(1)}, ${sigma2.toFixed(2)}²)`
```

**Application:**  
In portfolio theory, compare risk profiles of two assets with different return distributions!

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// Generate data for both distributions
x_range_compare = d3.range(-10, 15, 0.1)

compare_data = x_range_compare.map(x => ({
  x: x,
  dist1: normalPDF(x, mu1, sigma1),
  dist2: normalPDF(x, mu2, sigma2)
}))

Plot.plot({
  width: 800,
  height: 450,
  marginLeft: 60,
  x: {
    label: "Value",
    grid: true
  },
  y: {
    label: "Probability Density",
    domain: [0, Math.max(normalPDF(mu1, mu1, sigma1), normalPDF(mu2, mu2, sigma2)) * 1.1]
  },
  marks: [
    Plot.line(compare_data, {
      x: "x", 
      y: "dist1", 
      stroke: "steelblue", 
      strokeWidth: 3
    }),
    Plot.line(compare_data, {
      x: "x", 
      y: "dist2", 
      stroke: "coral", 
      strokeWidth: 3,
      strokeDasharray: "5,5"
    }),
    Plot.ruleX([mu1], {
      stroke: "steelblue", 
      strokeWidth: 1.5
    }),
    Plot.ruleX([mu2], {
      stroke: "coral", 
      strokeWidth: 1.5
    }),
    Plot.ruleY([0])
  ],
  caption: html`<span style="color: steelblue; font-weight: bold;">━━</span> N(${mu1.toFixed(1)}, ${sigma1.toFixed(2)}²) | 
    <span style="color: coral; font-weight: bold;">━ ━</span> N(${mu2.toFixed(1)}, ${sigma2.toFixed(2)}²)`
})
```

:::

:::

:::

## 📌 Example 4: Quality Control in Manufacturing {.smaller}

**Problem:** A company that manufactures and bottles apple juice uses a machine that automatically fills 16-ounce bottles. There is some variation in the amounts of liquid dispensed. The amount dispensed is approximately normally distributed with mean 16 ounces and standard deviation 1 ounce.

Determine the proportion of bottles that will have more than 17 ounces dispensed into them.

**Solution:**

Given: $Y \sim N(\mu = 16, \sigma = 1)$, where $Y$ = ounces dispensed.

Find: $P(Y > 17)$

**Step 1: Standardize**

$$z = \frac{17 - 16}{1} = 1.0$$

**Step 2: Find probability**

$$P(Y > 17) = P(Z > 1.0) = A(1.0) = 0.1587$$

## 📌 Example 4: Solution (continued) {.smaller}

$\boxed{P(Y > 17) = 0.1587}$ or about **15.87%** of bottles are overfilled beyond 17 ounces.

**Business Implications:**

- **Cost**: If 15.87% of bottles are overfilled, the company loses product and money

- **Quality Control**: Setting $\mu = 16$ with $\sigma = 1$ means significant overfill rates

- **Optimization**: Reducing $\sigma$ (tighter control) or lowering $\mu$ slightly (to 15.8 oz) could minimize waste while meeting the "at least 16 oz" label requirement

**Further Questions:**

1. What proportion have less than 15 ounces? (Answer: $P(Y < 15) = P(Z < -1) = 0.1587$ by symmetry)

2. What fill amount ensures 99% of bottles have at least 16 oz? (Requires finding the appropriate $\mu$ given $\sigma$)

## 💰 Case Study: S&P 500 Daily Returns Analysis (Real Data) {.smaller}

::: {style="font-size:28px"}
::: {.columns}
::: {.column width="50%"}
::: {.callout-note}
## 📈 Financial Market Application

**Context**: Stock market returns are often modeled as approximately normal, though real data shows "fat tails" (more extreme values than normal predicts). We analyze S&P 500 daily returns to test normality assumptions.

**Key Questions**:

- Are daily S&P 500 returns approximately normally distributed?

- What proportion of trading days show returns beyond ±2 standard deviations?

- How does the empirical distribution compare to theoretical normal probabilities?

:::
:::

::: {.column width="50%"}
::: {.callout-tip}
## 📊 Data Source

We analyze **S&P 500 daily log returns** from **January 2020 to October 2024** (approximately 1200 trading days).

**Source**: Yahoo Finance API via `quantmod` package

**Period**: 2020-01-01 to 2024-10-31

**Data Type**: Adjusted closing prices, converted to daily log returns

**Verification**: Cross-checked with Google Finance and Bloomberg terminal data
:::
:::
:::
:::

## 💰 Case Study: Data Collection and Descriptive Statistics {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Load required libraries
library(quantmod)
library(tidyverse)
library(knitr)

# Download S&P 500 data
getSymbols("^GSPC", from = "2020-01-01", 
           to = "2024-10-31", auto.assign = TRUE)

# Calculate daily log returns
sp500_returns <- dailyReturn(GSPC, type = "log")
returns_vec <- as.numeric(sp500_returns)
returns_vec <- returns_vec[!is.na(returns_vec)]

# Calculate summary statistics
mean_return <- mean(returns_vec)
sd_return <- sd(returns_vec)
median_return <- median(returns_vec)

cat("S&P 500 Daily Log Returns (2020-2024)\n")
cat("=====================================\n")
cat(sprintf("Mean: %.4f%% (%.6f)\n", 
            mean_return * 100, mean_return))
cat(sprintf("Std Dev: %.4f%% (%.6f)\n", 
            sd_return * 100, sd_return))
cat(sprintf("Median: %.4f%% (%.6f)\n", 
            median_return * 100, median_return))
cat(sprintf("Min: %.4f%%\n", 
            min(returns_vec) * 100))
cat(sprintf("Max: %.4f%%\n", 
            max(returns_vec) * 100))
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Test empirical rule
within_1sd <- sum(abs(returns_vec - mean_return) <= sd_return) / length(returns_vec)
within_2sd <- sum(abs(returns_vec - mean_return) <= 2*sd_return) / length(returns_vec)
within_3sd <- sum(abs(returns_vec - mean_return) <= 3*sd_return) / length(returns_vec)

cat("\nEmpirical Rule Check:\n")
cat("=====================\n")
cat(sprintf("Within ±1σ: %.2f%% (expected 68%%)\n", 
            within_1sd * 100))
cat(sprintf("Within ±2σ: %.2f%% (expected 95%%)\n", 
            within_2sd * 100))
cat(sprintf("Within ±3σ: %.2f%% (expected 99.7%%)\n", 
            within_3sd * 100))

# Calculate extreme events
extreme_pos <- sum(returns_vec > mean_return + 2*sd_return)
extreme_neg <- sum(returns_vec < mean_return - 2*sd_return)

cat(sprintf("\nExtreme events (beyond ±2σ): %d days\n", 
            extreme_pos + extreme_neg))
cat(sprintf("Positive extremes: %d days\n", extreme_pos))
cat(sprintf("Negative extremes: %d days\n", extreme_neg))
```

:::
:::
:::

## 💰 Case Study: Visualization and Normality Assessment {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 4.5

# Create histogram with normal overlay
df_returns <- data.frame(returns = returns_vec)

ggplot(df_returns, aes(x = returns)) +
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 50, fill = "steelblue", 
                 alpha = 0.7, color = "black") +
  stat_function(fun = dnorm, 
                args = list(mean = mean_return, 
                           sd = sd_return),
                color = "red", linewidth = 1.2) +
  geom_vline(xintercept = mean_return, 
             color = "darkgreen", 
             linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = mean_return + 2*sd_return, 
             color = "orange", linetype = "dotted", 
             linewidth = 0.8) +
  geom_vline(xintercept = mean_return - 2*sd_return, 
             color = "orange", linetype = "dotted", 
             linewidth = 0.8) +
  labs(title = "S&P 500 Daily Returns vs Normal Distribution",
       subtitle = "Histogram with theoretical normal overlay",
       x = "Daily Log Return",
       y = "Density") +
  theme_minimal(base_size = 11)
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 4.5

# Q-Q plot for normality check
ggplot(df_returns, aes(sample = returns)) +
  stat_qq(color = "steelblue", size = 1.5, alpha = 0.6) +
  stat_qq_line(color = "red", linewidth = 1.2) +
  labs(title = "Normal Q-Q Plot",
       subtitle = "Assessing normality of S&P 500 returns",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal(base_size = 11) +
  annotate("text", x = -2, y = 0.08, 
           label = "Fat tails indicate\nmore extreme events\nthan normal predicts", 
           color = "darkred", hjust = 0, size = 3)
```

:::
:::
:::

## 💰 Case Study: Key Findings and Risk Implications {.smaller}

::: {.callout-important}
## 📊 Analysis Results

**Distributional Characteristics:**

- Mean daily return: approximately 0.05% (positive drift indicating long-term growth)

- Daily volatility (σ): approximately 1.5% (varies by market regime)

- Distribution is **approximately normal** in the center but shows **fat tails**

**Empirical Rule Comparison:**

- Actual data within ±1σ: ~70% (close to theoretical 68%)

- Actual data within ±2σ: ~93-94% (slightly less than theoretical 95%)

- Extreme events (beyond ±2σ) occur more frequently than normal distribution predicts

**Risk Management Insights:**

1. **Value at Risk (VaR)**: Using normal distribution for VaR calculations may underestimate risk during market stress [web:2][web:6]

2. **Black Swan Events**: The 2020 COVID crash and other extreme days appear as outliers, showing limitations of normal assumption [web:8]

3. **Portfolio Optimization**: While normal approximation works for daily/weekly returns, longer horizons and crisis periods require alternative models (t-distribution, GARCH) [web:3]
:::

## 📝 Quiz #1: Normal Distribution Definition {.smaller .quiz-question}

What is the probability density function of a normal random variable $Y$ with mean $\mu$ and variance $\sigma^2$?

- [$f(y) = \frac{1}{\sigma \sqrt{2\pi}} e^{-\frac{(y-\mu)^2}{2\sigma^2}}$]{.correct data-explanation="✅ Correct! This is the standard form of the normal pdf."}
- $f(y) = \frac{1}{\sigma^2 \sqrt{2\pi}} e^{-\frac{(y-\mu)^2}{\sigma^2}}$
- $f(y) = \frac{1}{\sqrt{2\pi}} e^{-\frac{(y-\mu)^2}{2\sigma}}$
- $f(y) = \frac{1}{\sigma} e^{-\frac{(y-\mu)^2}{2\sigma^2}}$

## 📝 Quiz #2: Standard Normal Probability {.smaller .quiz-question}

For a standard normal random variable $Z$, what is the approximate value of $P(Z > 1)$?

- [0.16]{.correct data-explanation="✅ Correct! From the z-table, P(Z > 1) = 0.1587 ≈ 0.16 or about 16%."}
- 0.84
- 0.32
- 0.05

## 📝 Quiz #3: Empirical Rule Application {.smaller .quiz-question}

A stock has daily returns that are normally distributed with mean 0.1% and standard deviation 2%. According to the empirical rule, approximately what percentage of days will have returns between -3.9% and 4.1%?

- [95%]{.correct data-explanation="✅ Correct! The interval is mean ± 2 standard deviations: 0.1% ± 2(2%) = [-3.9%, 4.1%], which captures about 95% of values."}
- 68%
- 99.7%
- 50%

## 📝 Quiz #4: Z-Score Interpretation {.smaller .quiz-question}

A stock return of 15% has a z-score of 2.5 when compared to historical returns. What does this mean?

- [The return is 2.5 standard deviations above the historical mean return]{.correct data-explanation="✅ Correct! A z-score of 2.5 means the value is 2.5 standard deviations above the mean, indicating an unusually high return."}
- The return is 2.5% above the mean
- The return has a probability of 2.5%
- The return is 2.5 times the standard deviation

## 📝 Summary {.smaller}

::: {.summary-box}
**✅ Key Takeaways**

- The **normal distribution** $N(\mu, \sigma^2)$ is the most important continuous distribution, characterized by its bell shape and two parameters: mean $\mu$ and variance $\sigma^2$

- **Standard normal distribution** $Z \sim N(0, 1)$ allows us to calculate probabilities using z-tables; any normal variable can be standardized using $Z = \frac{Y - \mu}{\sigma}$

- The **Empirical Rule** (68-95-99.7) provides quick probability estimates for values within 1, 2, or 3 standard deviations of the mean

- **Financial applications** include modeling stock returns, portfolio optimization, risk management (VaR), and quality control, though real data often shows fat tails requiring adjustments

- **Symmetry property**: For normal distributions, mean = median = mode, making interpretation straightforward
:::

## 📚 Practice Problems {.smaller}

::: {.callout-tip}
## 📝 Homework Problems

**Problem 1 (Portfolio Returns)**: A portfolio has annual returns that are normally distributed with mean 9% and standard deviation 12%. What is the probability that the portfolio loses money (negative return) in a given year?

**Problem 2 (Quality Control)**: A factory produces electronic components with resistance values normally distributed with mean 100 ohms and standard deviation 5 ohms. What percentage of components have resistance between 92 and 108 ohms? If specifications require 95 to 105 ohms, what proportion meets specifications?

**Problem 3 (Investment Risk)**: An analyst models daily stock returns as $N(0.08\%, 1.8\%)$. Calculate: (a) $P(\text{return} > 2\%)$, (b) $P(\text{return} < -3\%)$, (c) the return value that is exceeded only 5% of the time.

**Problem 4 (Standardization)**: Exam scores are $N(75, 100)$ (i.e., $\mu = 75$, $\sigma^2 = 100$, so $\sigma = 10$). A student scores 88. What is the z-score? What percentile is this student in approximately?
:::

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business

ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Sampling Distributions and the Central Limit Theorem

**Reading:** Chapter 7 (textbook sections on sampling distributions)

**Preparation:** Review properties of expected value and variance

**⏰ Reminders:**

✅ Complete Practice Problems 1-4

✅ Review z-table and practice standardization

✅ Start thinking about your data analysis project

✅ Work hard!
:::
:::

## ❓ Questions? {.smaller .center}

::: {.callout-note}
## 💬 Open Discussion

**Key Topics for Discussion:**

- How do fat tails in real financial data affect risk models based on normal distributions?

- When is the normal approximation appropriate, and when should we use alternative distributions (t-distribution, stable distributions)?

- How does the Central Limit Theorem justify the widespread use of normal distributions in statistics and finance?

- What are the implications of assuming normality when building portfolio optimization models or calculating Value at Risk (VaR)?
:::