---
title: "Mathematical Statistics"
subtitle: "Discrete Random Variables and Expected Value"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - Discrete Random Variables and Expected Value"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
library(ggplot2)
```

## 🎯 Learning Objectives

::: {style="font-size: 32px"}
By the end of this lecture, you will be able to:

- **Identify** a discrete random variable and write its probability distribution as a formula, a table or a graph

- **Check** that a proposed $p(y)$ is a valid distribution using Theorem 3.1

- **Compute** $E(Y)$, $E[g(Y)]$, $V(Y)$ and $\sigma$ for a discrete distribution

- **Use** linearity (Theorems 3.3–3.5) and $V(Y) = E(Y^2) - \mu^2$ to shorten the arithmetic

- **Price** a premium or compare two projects by their mean and their variance
:::

---

## 🗺️ Where We Are

::: {style="font-size: 30px"}
Last class ended on: *Chapter 3 — discrete random variables, their probability distributions, and expected value (§§3.1–3.3).*

Chapter 2 gave us a random variable as a **real-valued function on a sample space**. Today we stop listing sample points and start working with the numbers directly.

::: {.callout-important}
## The question this lecture answers
A motor insurer sells hail cover. Most policies never claim; a few claim a lot. **What should one policy cost, and how far can a year's claims stray from that price?**
:::
:::

---

## 📐 Discrete Random Variables

::: {style="font-size: 30px"}
::: {.callout-note}
## Definition 3.1
A random variable $Y$ is said to be **discrete** if it can assume only a finite or countably infinite number of distinct values.
:::

Discrete random variables are usually **counts**:

- complaints per hour at a mobile operator's call centre: $0, 1, 2, \dots$
- outages per week on an ISP's business network: $0, 1, 2, 3$
- hail claims filed on a motor portfolio in a season

Notation: $Y$ is the random variable; $y$ is a particular value it takes. After the hour is over, $y$ is not random.
:::

---

## 📐 Definitions 3.2 and 3.3, Theorem 3.1

::: {style="font-size: 29px"}
::: {.callout-note}
## Definitions 3.2 and 3.3
$P(Y = y)$ is the **sum of the probabilities of all sample points in $S$ assigned the value $y$**, written $p(y)$ — the *probability function* for $Y$. The **probability distribution** of $Y$ is a formula, a table or a graph giving $p(y)$ for all $y$.
:::

::: {.callout-important}
## Theorem 3.1
For any discrete probability distribution: **(1)** $0 \le p(y) \le 1$ for all $y$; **(2)** $\sum_y p(y) = 1$, summed over all $y$ with nonzero probability.
:::

Any $y$ not assigned a positive probability has $p(y) = 0$.
:::

---

## 📊 Example 3.1: An Index Draw

::: {style="font-size: 28px"}
A new index will add **2 of 6** shortlisted Baku-listed companies, chosen at random: 3 are banks, 3 are energy firms. Let $Y$ = number of banks added.

::: {.fragment}
$S$ has $\binom{6}{2} = 15$ equally likely points. Counting the points assigned each value:
$$p(y) = \frac{\binom{3}{y}\binom{3}{2-y}}{\binom{6}{2}}, \quad y = 0, 1, 2$$
:::

::: {.fragment}
| $y$ | 0 | 1 | 2 |
|---|---|---|---|
| $p(y)$ | $3/15 = 1/5$ | $9/15 = 3/5$ | $3/15 = 1/5$ |

Formula, table — and each $p(y) \in [0,1]$, summing to 1, as Theorem 3.1 requires.
:::
:::

---

## 📐 Mean and Variance

::: {style="font-size: 29px"}
::: {.callout-note}
## Definition 3.4
The **expected value** of $Y$ is $\;E(Y) = \sum_y y\,p(y).$ If $p(y)$ describes the population, $E(Y) = \mu$.
:::

::: {.callout-note}
## Definition 3.5
The **variance** of $Y$ is $\;V(Y) = E[(Y - \mu)^2]$. The **standard deviation** $\sigma$ is its positive square root.
:::

::: {.fragment}
$E(Y)$ is a **long-run average**: run the experiment $n$ times and about $n\,p(y)$ results equal $y$, so the average of the results is close to $\sum_y y\,p(y)$.
:::
:::

---

## 🌐 Example 3.2: ISP Outages

::: {style="font-size: 28px"}
$Y$ = outages per week on an ISP's business network in Baku.

| $y$ | 0 | 1 | 2 | 3 |
|---|---|---|---|---|
| $p(y)$ | $1/8$ | $1/4$ | $3/8$ | $1/4$ |

::: {.fragment}
$$\mu = 0\left(\tfrac18\right) + 1\left(\tfrac14\right) + 2\left(\tfrac38\right) + 3\left(\tfrac14\right) = 1.75$$
:::

::: {.fragment}
$$\sigma^2 = (0-1.75)^2\tfrac18 + (1-1.75)^2\tfrac14 + (2-1.75)^2\tfrac38 + (3-1.75)^2\tfrac14 = 0.9375$$
$\sigma = \sqrt{0.9375} = 0.97$. The interval $\mu \pm \sigma = (0.78, 2.72)$ holds $y = 1, 2$: probability $5/8$.
:::
:::

---

## 💻 The Same Numbers in R

```{r}
#| label: isp-moments
#| code-fold: false
y <- 0:3
p <- c(1, 2, 3, 2) / 8

mu     <- sum(y * p)                 # Definition 3.4
sigma2 <- sum((y - mu)^2 * p)        # Definition 3.5
round(c(total = sum(p), mu = mu, var = sigma2, sd = sqrt(sigma2)), 4)

# A long run of simulated weeks: the average settles on mu
weeks <- sample(y, 100000, replace = TRUE, prob = p)
round(c(avg = mean(weeks), var = mean((weeks - mean(weeks))^2)), 4)
```

---

## 📈 The Probability Histogram

```{r}
#| label: isp-figure
#| echo: false
#| fig-width: 10
#| fig-height: 4.4
s <- sqrt(sigma2)
ggplot(data.frame(y = y, p = p), aes(y, p)) +
  annotate("rect", xmin = mu - s, xmax = mu + s, ymin = 0, ymax = Inf,
           fill = "#cbb8a9", alpha = 0.35) +
  geom_col(width = 0.9, fill = "#8ba3c7") +
  geom_vline(xintercept = mu, colour = "#8b2635", linewidth = 1.2) +
  annotate("text", x = mu + 0.05, y = 0.41, hjust = 0, size = 6,
           label = "mu == 1.75", parse = TRUE, colour = "#8b2635") +
  scale_x_continuous(breaks = 0:3, minor_breaks = NULL) +
  scale_y_continuous(limits = c(0, 0.42), breaks = seq(0, 0.4, 0.1)) +
  labs(x = "Outages per week, y", y = "p(y)") +
  theme_minimal(base_size = 20)
```

::: {style="font-size: 28px"}
$\mu$ sits between bars, at a value $Y$ can never take. The shaded band is $\mu \pm \sigma$.
:::

---

## 📐 Theorem 3.2: $E[g(Y)]$

::: {style="font-size: 28px"}
::: {.callout-important}
## Theorem 3.2
$E[g(Y)] = \sum_{\text{all } y} g(y)\,p(y)$
:::

The ISP's contract credits a corporate client **0 AZN** for 0 or 1 outages, **500 AZN** for 2 and **1,500 AZN** for 3.
$$E[g(Y)] = 0\left(\tfrac18\right) + 0\left(\tfrac14\right) + 500\left(\tfrac38\right) + 1500\left(\tfrac14\right) = 562.5 \text{ AZN per week}$$

::: {.fragment}
No need to derive the distribution of $g(Y)$ first. The proof does exactly that grouping — $P(g = 0) = 3/8$, $P(g = 500) = 3/8$, $P(g = 1500) = 1/4$ — and gets the same sum.
:::
:::

---

## 📐 Theorems 3.3–3.5: Linearity

::: {style="font-size: 30px"}
::: {.callout-important}
## Theorems 3.3, 3.4 and 3.5
For a constant $c$ and functions $g, g_1, \dots, g_k$ of $Y$:
$$E(c) = c, \qquad E[c\,g(Y)] = c\,E[g(Y)],$$
$$E[g_1(Y) + \cdots + g_k(Y)] = E[g_1(Y)] + \cdots + E[g_k(Y)]$$
:::

::: {.fragment}
Each proof is one line: apply Theorem 3.2, then pull the constant or split the sum. $E(c) = c$ uses $\sum_y p(y) = 1$ from Theorem 3.1.
:::
:::

---

## 📐 Theorem 3.6: A Shortcut

::: {style="font-size: 29px"}
::: {.callout-important}
## Theorem 3.6
$V(Y) = \sigma^2 = E[(Y - \mu)^2] = E(Y^2) - \mu^2$
:::

::: {.fragment}
Expand and use 3.3–3.5: $E(Y^2 - 2\mu Y + \mu^2) = E(Y^2) - 2\mu^2 + \mu^2$.
:::

::: {.fragment}
**Example 3.3, ISP outages:** $E(Y^2) = 0 + 1\left(\tfrac14\right) + 4\left(\tfrac38\right) + 9\left(\tfrac14\right) = 4$, so
$$\sigma^2 = 4 - 1.75^2 = 0.9375$$
Note $E(Y^2) = 4 \ne \mu^2 = 3.0625$: **the mean of a function is not the function of the mean.**
:::
:::

---

## 💸 A Remittance Fee, $aY + b$

::: {style="font-size: 28px"}
A Baku transfer office charges **1 AZN plus 2%** of the amount. With $Y$ = amount in hundreds of AZN, the fee is $F = 2Y + 1$.

| $y$ | 1 | 2 | 3 | 5 |
|---|---|---|---|---|
| $p(y)$ | 0.4 | 0.3 | 0.2 | 0.1 |

$E(Y) = 2.1$, $E(Y^2) = 5.9$, so $V(Y) = 5.9 - 2.1^2 = 1.49$.

::: {.fragment}
From Theorems 3.3–3.6: $E(aY + b) = a\mu + b$ and $V(aY + b) = a^2\sigma^2$.
$$E(F) = 2(2.1) + 1 = 5.2 \text{ AZN}, \qquad V(F) = 4(1.49) = 5.96, \quad \sigma_F = 2.44 \text{ AZN}$$
The fixed 1 AZN moves the mean and leaves the spread alone.
:::
:::

---

## 🌨️ Worked Example: Pricing Hail Cover

::: {style="font-size: 28px"}
One car's hail claim $X$ in a season: **0** (prob. 0.95), **2,000 AZN** (0.04), **8,000 AZN** (0.01). Admin costs 40 AZN per policy; the insurer targets an expected profit of 60 AZN. What premium $C$?

::: {.fragment}
$E(X) = 2000(0.04) + 8000(0.01) = 160$ AZN. Profit is $C - 40 - X$, so by linearity
$$E(\text{profit}) = C - 40 - 160 = 60 \;\Rightarrow\; C = 260 \text{ AZN}$$
:::

::: {.fragment}
$E(X^2) = 0.04(2000^2) + 0.01(8000^2) = 800{,}000$, so $\sigma_X = \sqrt{800{,}000 - 160^2} = 880$ AZN — more than three premiums. One policy is a gamble; the business is in **pooling many**.
:::
:::

---

## 🚚 Example 3.4: Two Van Models

::: {style="font-size: 28px"}
A logistics firm runs a van $t$ hours a day. Daily breakdowns $Y_A$ have mean and variance $0.10t$; $Y_B$ has $0.12t$. Daily cost (AZN): $C_A = 10t + 30Y_A^2$, $C_B = 8t + 30Y_B^2$.

::: {.fragment}
The key step is Theorem 3.6 read backwards, $E(Y^2) = V(Y) + \mu^2$:
$$E(C_A) = 10t + 30[0.10t + (0.10t)^2] = 13t + 0.3t^2, \quad E(C_B) = 11.6t + 0.432t^2$$
:::

::: {.fragment}
| $t$ | $E(C_A)$ | $E(C_B)$ | cheaper |
|---|---|---|---|
| 10 h | 160.0 | 159.2 | B |
| 20 h | 380.0 | 404.8 | A |

The two cross at $t = 1.4/0.132 \approx 10.6$ hours.
:::
:::

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 29px"}
A bank can finance one of two projects. NPV in millions of AZN:

| Project | Outcome 1 | Outcome 2 | Outcome 3 |
|---|---|---|---|
| P | $-1$ (prob. 0.2) | $2$ (prob. 0.5) | $4$ (prob. 0.3) |
| Q | $1$ (prob. 0.25) | $2$ (prob. 0.5) | $3$ (prob. 0.25) |

**Four minutes, in pairs:**

1. Find $E(Y)$ for each project.
2. Find $V(Y)$ for each, using Theorem 3.6.
3. Which would a lender who fears losses choose, and what does the mean alone hide?
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 29px"}
1. $E(P) = -0.2 + 1.0 + 1.2 = 2.0$ and $E(Q) = 0.25 + 1.0 + 0.75 = 2.0$. **Same mean.**

2. $E(P^2) = 0.2 + 2.0 + 4.8 = 7.0$, so $V(P) = 7.0 - 4 = 3.0$, $\sigma_P = 1.73$.
   $E(Q^2) = 0.25 + 2.0 + 2.25 = 4.5$, so $V(Q) = 0.5$, $\sigma_Q = 0.71$.

::: {.fragment}
3. The mean cannot tell them apart. P carries a **20% chance of losing 1 million**; Q never loses. The variance is the first number that sees the difference — which is why §1.3 paired every mean with a spread.
:::
:::

---

## 🔬 Interactive: Moving Mass to the Tail {.smaller}

```{ojs}
//| echo: false

viewof shift = {
  const input = Inputs.range([0, 0.30], {value: 0, step: 0.01,
    label: "Mass moved from 1 to 5 complaints:"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}
```

```{ojs}
//| echo: false

base = [0.15, 0.35, 0.25, 0.15, 0.05, 0.05]
pmf = base.map((b, y) => ({y: y, p: y === 1 ? b - shift : y === 5 ? b + shift : b}))
mu = d3.sum(pmf, d => d.y * d.p)
ey2 = d3.sum(pmf, d => d.y * d.y * d.p)
v = ey2 - mu * mu

md`Total **${d3.sum(pmf, d => d.p).toFixed(2)}** · mean **μ = ${mu.toFixed(2)}** · variance **σ² = ${v.toFixed(3)}** · σ = ${Math.sqrt(v).toFixed(3)}`

Plot.plot({
  width: 1150,
  height: 300,
  marginTop: 40,
  marginLeft: 78,
  marginBottom: 58,
  style: {fontSize: "18px"},
  x: {label: "Complaints per hour, y", domain: [0, 1, 2, 3, 4, 5], padding: 0.15},
  y: {label: "p(y)", domain: [0, 0.4], ticks: [0, 0.1, 0.2, 0.3, 0.4], tickFormat: ".1f", grid: true},
  marks: [
    Plot.barY(pmf, {x: "y", y: "p", fill: d => d.y === 1 || d.y === 5 ? "#8b2635" : "#8ba3c7"}),
    Plot.text(pmf, {x: "y", y: "p", text: d => d.p.toFixed(2), dy: -14, fontSize: 18}),
    Plot.ruleY([0])
  ]
})
```

::: {style="font-size: 28px"}
Moving 0.30 of mass four steps raises μ by $4(0.30) = 1.2$ — but nearly **doubles** σ².
:::

---

## 📝 Quiz #1: Is It a Distribution? {.quiz-question}

Which of these can be the probability function of complaints per hour, $y = 0, 1, 2, 3$?

- $p(y) = 0.3, 0.3, 0.3, 0.3$
- [$p(y) = 0.1, 0.4, 0.3, 0.2$]{.correct data-explanation="✅ Every value lies in [0, 1] and they sum to 1, which is all Theorem 3.1 asks. The first sums to 1.2; the third has a negative entry; the fourth sums to 0.9."}
- $p(y) = 0.5, 0.4, 0.2, -0.1$
- $p(y) = 0.2, 0.2, 0.2, 0.3$

---

## 📝 Quiz #2: The Mean {.quiz-question}

A small fleet files $Y$ = 0, 1 or 2 hail claims in a season, with probabilities 0.5, 0.3, 0.2. What is $E(Y)$?

- $1$, the middle value
- [$0.7$]{.correct data-explanation="✅ Definition 3.4: 0(0.5) + 1(0.3) + 2(0.2) = 0.7. The mean weights each value by its probability; it need not be a value Y can take."}
- $0.33$, the average probability
- $1.1$

---

## 📝 Quiz #3: A Fee Rule {.quiz-question}

$V(Y) = 4$ for a transfer amount $Y$. The fee is $F = 3Y + 5$ AZN. What is $V(F)$?

- $17$
- $12$
- [$36$]{.correct data-explanation="✅ V(aY + b) = a²V(Y) = 9 × 4 = 36. The constant 5 shifts every fee equally and adds no spread; the multiplier enters squared."}
- $41$

---

## 📋 Key Formulas

::: {style="font-size: 29px"}
| | Statement |
|---|---|
| Theorem 3.1 | $0 \le p(y) \le 1$, $\;\sum_y p(y) = 1$ |
| Definition 3.4 | $E(Y) = \sum_y y\,p(y)$ |
| Theorem 3.2 | $E[g(Y)] = \sum_y g(y)\,p(y)$ |
| Definition 3.5 | $V(Y) = E[(Y - \mu)^2]$, $\;\sigma = \sqrt{V(Y)}$ |
| Theorems 3.3–3.5 | $E(c) = c$, $\;E[cg(Y)] = cE[g(Y)]$, $\;E$ of a sum is the sum of $E$ |
| Theorem 3.6 | $V(Y) = E(Y^2) - \mu^2$ |
| consequence | $E(aY + b) = a\mu + b$, $\;V(aY + b) = a^2\sigma^2$ |
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- A discrete random variable takes countably many values; its distribution is $p(y)$ as a formula, a table or a graph

- Theorem 3.1 is the validity check: every $p(y)$ in $[0, 1]$, total 1

- $E(Y)$ is a probability-weighted average — a premium is built on it

- $V(Y)$ measures what the mean hides: two projects with the same mean can carry very different risk

- Linearity and $E(Y^2) - \mu^2$ turn most calculations into two sums
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition**

- Exercises at the end of §3.2: 3.1, 3.2, 3.9
- Exercises at the end of §3.3: 3.12, 3.13, 3.14, 3.19, 3.23, 3.27, 3.30

**Week 5, Problem Set 2** opens today and closes **17 October at 23:59** on WeBWorK, covering §§3.1–3.3. The TA-led tutorial runs this week.

**Quiz I** is on **17 October**, first 30 minutes of class, on Chapters 1–2 — today's material is not on it, but it is on Midterm I (24 October).

**Next class:** §3.4 — the binomial distribution.
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**

📧 sorujov@ada.edu.az\
🏢 Building D, Room D325\
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- Can $E(Y)$ be a value that $Y$ never takes? Can $V(Y)$ ever be negative?

- A premium is set at $E(X)$ plus costs. Why does an insurer with 50 policies charge more than one with 50,000?

- When is $E[g(Y)] = g(E(Y))$ exactly true?
:::
