---
title: "Mathematical Statistics"
subtitle: "The Expected Value and Variance of Linear Functions; Portfolio Variance"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: [default, ../slide-fit.scss]
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics I - Linear Functions and Portfolio Variance"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
# Named explicitly so engine selection cannot drift away from R.
engine: knitr
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: png
    dpi: 150
    fig.align: center
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
library(ggplot2)
theme_set(theme_minimal(base_size = 20))
```

## 🎯 Learning Objectives

::: {style="font-size: 32px"}
By the end of this lecture, you will be able to:

- **State** Theorem 5.12 for the mean, variance and covariance of linear functions

- **Compute** the expected return and variance of a two- and three-asset portfolio

- **Explain** why the variance of an equally weighted portfolio falls towards a floor set by the average covariance

- **Find** the hedge ratio that minimises the variance of a position

- **Recognise** $\bar{Y}$ and $\hat{p}$ as linear functions, with $V(\bar{Y}) = \sigma^2/n$
:::

---

## 🗺️ Where We Are

::: {style="font-size: 32px"}
**Wackerly §5.8**

Last class ended on a question: the retailer's expected revenue needed only the marginals. **Which quantity about the revenue will need the covariance as well?**

::: {.fragment}
The variance. Expectation is linear and ignores dependence; the variance of a sum is where $\text{Cov}(Y_1, Y_2)$ enters, and it enters with a factor of 2.
:::

::: {.fragment}
In finance that one fact has a name: **portfolio variance**. It is why holding two risky assets can be less risky than holding either one.
:::
:::

---

## ❓ Motivating Question

::: {style="font-size: 30px"}
A Baku pension fund puts 100,000 AZN into two equity funds for a year:

| | weight | mean return | s.d. of return |
|---|---|---|---|
| Energy fund $(Y_1)$ | 0.6 | 12% | 25% |
| Telecom fund $(Y_2)$ | 0.4 | 8% | 15% |

::: {.fragment}
The expected return is $0.6(12) + 0.4(8) = 10.4\%$, the weighted average. **Is the risk the weighted average too**, $0.6(25) + 0.4(15) = 21\%$?
:::

::: {.fragment}
Only in one special case. Today's theorem says which, and what the risk is otherwise.
:::
:::

---

## 📐 Theorem 5.12

::: {style="font-size: 28px"}
::: {.callout-important}
## Theorem 5.12

Let $E(Y_i) = \mu_i$ and $E(X_j) = \xi_j$. Define $U_1 = \sum_{i=1}^n a_iY_i$ and $U_2 = \sum_{j=1}^m b_jX_j$ for constants $a_i$, $b_j$. Then

a. $E(U_1) = \sum_{i=1}^n a_i\mu_i$

b. $V(U_1) = \sum_{i=1}^n a_i^2V(Y_i) + 2\sum\sum_{1 \le i < j \le n} a_ia_j\text{Cov}(Y_i, Y_j)$

c. $\text{Cov}(U_1, U_2) = \sum_{i=1}^n\sum_{j=1}^m a_ib_j\text{Cov}(Y_i, X_j)$
:::

::: {.fragment}
Part (a) is Theorems 5.7 and 5.8 again. Parts (b) and (c) are new, and (b) is a special case of (c), since $\text{Cov}(Y_i, Y_i) = V(Y_i)$.
:::
:::

---

## 🔍 Reading Part (b) for Two Variables

::: {style="font-size: 30px"}
$$V(a_1Y_1 + a_2Y_2) = a_1^2V(Y_1) + a_2^2V(Y_2) + 2a_1a_2\text{Cov}(Y_1, Y_2)$$

::: {.fragment}
- Constants come out **squared**, so $V(-Y) = V(Y)$
- The sign of the covariance term depends on the signs of $a_1a_2$ **and** of the covariance
- With $a_1 = 1$, $a_2 = -1$: $\; V(Y_1 - Y_2) = V(Y_1) + V(Y_2) - 2\text{Cov}(Y_1, Y_2)$
:::

::: {.fragment}
The proof expands $\left[\sum a_i(Y_i - \mu_i)\right]^2$: the $n$ squared terms give the variances, and the $n(n-1)$ cross terms pair up into $2\sum\sum_{i<j}$.
:::
:::

---

## 💼 The Two-Asset Portfolio

::: {style="font-size: 30px"}
Weight $w$ in asset 1 and $1 - w$ in asset 2. The portfolio return is a linear function, $R_p = wY_1 + (1-w)Y_2$.

::: {.callout-important}
## Portfolio mean and variance
$$E(R_p) = w\mu_1 + (1-w)\mu_2$$
$$V(R_p) = w^2\sigma_1^2 + (1-w)^2\sigma_2^2 + 2w(1-w)\rho\,\sigma_1\sigma_2$$
:::

::: {.fragment}
The last term uses $\text{Cov}(Y_1, Y_2) = \rho\sigma_1\sigma_2$ from Definition 5.11. With $\rho = 1$ the bracket is a perfect square, $V(R_p) = [w\sigma_1 + (1-w)\sigma_2]^2$, and the weighted-average s.d. is exact.
:::
:::

---

## 📊 Worked Example: The Pension Fund

::: {style="font-size: 30px"}
Energy and telecom returns have correlation $\rho = 0.3$, so $\text{Cov}(Y_1, Y_2) = 0.3(25)(15) = 112.5$ (in $\%^2$).

::: {.fragment}
$$V(R_p) = 0.36(625) + 0.16(225) + 2(0.6)(0.4)(112.5) = 225 + 36 + 54 = 315$$
:::

::: {.fragment}
$\sigma_{R_p} = \sqrt{315} = 17.75\%$, not 21%. On 100,000 AZN: an expected gain of **10,400 AZN** with a standard deviation of **17,748 AZN**.
:::

::: {.fragment}
Same expected return, 3.25 points less risk. **That gap is diversification**, and it exists whenever $\rho < 1$.
:::
:::

---

## 💻 Risk Across Correlations

```{r}
#| label: two-asset
w <- 0.6; s1 <- 25; s2 <- 15
rho <- c(-1, -0.5, 0, 0.3, 0.5, 1)

var_p <- w^2 * s1^2 + (1 - w)^2 * s2^2 + 2 * w * (1 - w) * rho * s1 * s2
data.frame(rho = rho, var_p = var_p, sd_p = round(sqrt(var_p), 2))
```

::: {style="font-size: 28px"}
The expected return is 10.4% in every row. Only the risk moves, from 21% at $\rho = 1$ down to 9% at $\rho = -1$.
:::

---

## 🔬 Interactive: The Shape of the Risk Curve {.smaller}

```{ojs}
//| echo: false

viewof rho_ET = {
  const input = Inputs.range([-1, 1], {value: 0.3, step: 0.05,
    label: "Correlation of the two funds:"});
  ['pointerdown','touchstart','mousedown','click','wheel','pointermove','touchmove']
    .forEach(e => input.addEventListener(e, ev => ev.stopPropagation()));
  return input;
}
```

```{ojs}
//| echo: false

sdP = (w, r) => Math.sqrt(w*w*625 + (1-w)*(1-w)*225 + 2*w*(1-w)*r*375)
riskCurve = Array.from({length: 101}, (_, i) => ({w: i/100, s: sdP(i/100, rho_ET)}))
wStar = Math.min(1, Math.max(0, (225 - rho_ET*375) / (625 + 225 - 2*rho_ET*375)))
best = [{w: wStar, s: sdP(wStar, rho_ET)}]

md`Lowest risk: **${best[0].s.toFixed(1)}%** with **${(100*wStar).toFixed(0)}%** in energy.`

Plot.plot({
  width: 1150,
  height: 300,
  marginLeft: 78,
  marginBottom: 58,
  style: {fontSize: "18px"},
  x: {label: "Weight w in the energy fund", domain: [0, 1]},
  y: {label: "Portfolio s.d. (%)", domain: [0, 26]},
  marks: [
    Plot.line([{w: 0, s: 15}, {w: 1, s: 25}], {x: "w", y: "s", stroke: "#cbb8a9", strokeDasharray: "4 4"}),
    Plot.line(riskCurve, {x: "w", y: "s", stroke: "#14130f", strokeWidth: 2.5}),
    Plot.dot(best, {x: "w", y: "s", r: 8, fill: "#8b2635"}),
    Plot.ruleY([0])
  ]
})
```

---

## 🏛️ Three Assets: Adding a Bond Fund

::: {style="font-size: 28px"}
The fund rebalances to 50% energy, 30% telecom and 20% an AZN government bond fund $(Y_3)$: mean 6%, s.d. 5%.

| | $\text{Cov}$ with telecom | $\text{Cov}$ with bonds |
|---|---|---|
| **Energy** | $0.3(25)(15) = 112.5$ | $-0.2(25)(5) = -25$ |
| **Telecom** | | $0.1(15)(5) = 7.5$ |

::: {.fragment}
Theorem 5.12(b): three variance terms and **three** pairs,
$$V(R_p) = 156.25 + 20.25 + 1 + 2(.5)(.3)(112.5) + 2(.5)(.2)(-25) + 2(.3)(.2)(7.5)$$
$$= 177.5 + 33.75 - 5 + 0.9 = 207.15, \qquad \sigma_{R_p} = 14.39\%$$
:::
:::

---

## 💻 The Same Sum as a Matrix

```{r}
#| label: three-asset
w     <- c(energy = 0.5, telecom = 0.3, bonds = 0.2)
mu    <- c(12, 8, 6)
sd    <- c(25, 15, 5)
R     <- matrix(c( 1.0, 0.3, -0.2,
                   0.3, 1.0,  0.1,
                  -0.2, 0.1,  1.0), 3, 3)
Sigma <- diag(sd) %*% R %*% diag(sd)       # covariance matrix

c(mean = sum(w * mu),
  var  = drop(t(w) %*% Sigma %*% w),
  sd   = sqrt(drop(t(w) %*% Sigma %*% w)))
```

::: {style="font-size: 28px"}
$w^\top\Sigma w$ is Theorem 5.12(b) with all $n^2$ terms written out: each off-diagonal pair appears twice, which is the factor 2.
:::

---

## 📐 $n$ Equally Weighted Assets

::: {style="font-size: 28px"}
Take $n$ assets, each with variance $\sigma^2$ and every pair with covariance $\rho\sigma^2$. Put $a_i = 1/n$ in Theorem 5.12(b):

$$V(R_p) = n\cdot\frac{1}{n^2}\sigma^2 + 2\cdot\frac{n(n-1)}{2}\cdot\frac{1}{n^2}\rho\sigma^2 = \frac{\sigma^2}{n} + \left(1 - \frac{1}{n}\right)\rho\sigma^2$$

::: {.fragment}
- With $\rho = 0$ this is **Example 5.27**: $V(\bar{Y}) = \sigma^2/n$, which falls to zero
- With $\rho > 0$ it falls only to $\rho\sigma^2$, the risk **no amount of diversification removes**
:::

::: {.fragment}
The first term is risk specific to each asset; the second is risk shared by the whole market.
:::
:::

---

## 📉 Diversification Has a Floor

```{r}
#| label: diversification
#| fig-width: 10
#| fig-height: 4
n  <- 1:40; s <- 30
df <- rbind(data.frame(n = n, sd = sqrt(s^2 / n + (1 - 1/n) * 0.25 * s^2), case = "r25"),
            data.frame(n = n, sd = sqrt(s^2 / n),                          case = "r0"))
ggplot(df, aes(n, sd, colour = case)) +
  geom_hline(yintercept = 15, linetype = "dashed", colour = "grey55") +
  geom_line(linewidth = 1.4) +
  scale_colour_manual(values = c(r25 = "#8b2635", r0 = "#3d6e8f"),
                      labels = c(r25 = expression(rho == 0.25), r0 = expression(rho == 0)),
                      name = NULL) +
  scale_x_continuous(breaks = c(1, 5, 10, 20, 30, 40)) +
  scale_y_continuous(breaks = seq(0, 30, 5), limits = c(0, 30)) +
  labs(x = "Number of shares held", y = "Portfolio s.d. (%)") +
  theme(legend.position = c(0.8, 0.8),
        legend.text = element_text(size = 22, hjust = 0),
        legend.key.width = unit(1.6, "cm"))
```

::: {style="font-size: 28px"}
Shares with $\sigma = 30\%$: at $\rho = 0.25$, ten shares give 17.1% and thirty give 15.7%, never below $\sqrt{0.25 \times 900} = 15\%$.
:::

---

## 🛡️ Hedging With a Negative Covariance

::: {style="font-size: 30px"}
A Sumqayit fuel distributor's monthly margin $Y_1$ (thousand AZN) shrinks when crude prices rise. It buys $h$ oil-linked contracts, each with value change $Y_2$. The position is $H = Y_1 + hY_2$.

::: {.fragment}
$$V(H) = V(Y_1) + h^2V(Y_2) + 2h\,\text{Cov}(Y_1, Y_2)$$
A parabola in $h$. Setting the derivative to zero:
$$h^* = -\frac{\text{Cov}(Y_1, Y_2)}{V(Y_2)}, \qquad V(H^*) = V(Y_1)(1 - \rho^2)$$
:::

::: {.fragment}
Negative covariance makes $h^* > 0$: **buy** the asset that moves against you.
:::
:::

---

## ⛽ Worked Example: The Hedge Ratio

::: {style="font-size: 30px"}
$\sigma_1 = 40$, $\sigma_2 = 8$ (thousand AZN), $\rho = -0.8$, so $\text{Cov}(Y_1, Y_2) = -0.8(40)(8) = -256$.

::: {.fragment}
$$h^* = \frac{256}{64} = 4 \text{ contracts}$$
$$V(H^*) = 1600 + 16(64) + 2(4)(-256) = 1600 + 1024 - 2048 = 576$$
:::

::: {.fragment}
The s.d. falls from 40 to $\sqrt{576} = 24$ thousand AZN, as $40\sqrt{1 - 0.64} = 24$ predicts.
:::

::: {.fragment}
A hedge removes the **correlated** part of the risk. The remaining 24 comes from what the contract does not track.
:::
:::

---

## 🧠 Think-Pair-Share

```{r}
#| label: tps-timer
#| echo: false
# The timer is the only thing in this deck that needs a package beyond base R.
# Guarded so a machine without it renders the deck anyway, with a static
# figure in the same corner, rather than halting the whole build.
if (requireNamespace("countdown", quietly = TRUE)) {
  countdown::countdown(minutes = 4, seconds = 0,
                       top = 0, right = 0,
                       font_size = "2em", warn_when = 30)
} else {
  htmltools::HTML(paste0(
    '<div style="position:absolute; top:0; right:0; font-size:2em;',
    ' font-weight:700; opacity:0.75">4:00</div>'))
}
```

::: {style="font-size: 30px"}
Back to the energy fund $Y_1$ and telecom fund $Y_2$: means 12 and 8, variances 625 and 225, $\text{Cov}(Y_1, Y_2) = 112.5$.

Client A holds $U = 0.5Y_1 + 0.5Y_2$. Client B runs a long-short book, $W = Y_1 - Y_2$.

**Four minutes, in pairs:**

1. Find $E(W)$ and $V(W)$.

2. Use Theorem 5.12(c) to find $\text{Cov}(U, W)$.

3. Can $U$ and $W$ be independent?
:::

---

## ✅ Think-Pair-Share: Solution

::: {style="font-size: 28px"}
1. $E(W) = 12 - 8 = 4\%$ and $V(W) = 625 + 225 - 2(112.5) = 625$, so $\sigma_W = 25\%$.

2. With $a = (0.5, 0.5)$ and $b = (1, -1)$, four terms:
$$\text{Cov}(U, W) = 0.5(625) - 0.5(112.5) + 0.5(112.5) - 0.5(225) = 200$$

::: {.fragment}
3. No. Theorem 5.11 says independence forces zero covariance, and $200 \ne 0$. With $V(U) = 268.75$, $\rho_{UW} = 200/(16.39 \times 25) = 0.49$.

The long-short book looks "market neutral", yet its return still moves with client A's portfolio.
:::
:::

---

## 📝 Quiz #1: A Difference {.quiz-question}

$V(Y_1) = 4$, $V(Y_2) = 9$ and $\text{Cov}(Y_1, Y_2) = 2$. What is $V(Y_1 - Y_2)$?

- [$9$]{.correct data-explanation="✅ Theorem 5.12(b) with a = (1, −1): 4 + 9 − 2(2) = 9. The constants are squared, and the cross term takes the sign of the product of the constants, (1)(−1) = −1."}
- $17$
- $13$
- $5$

---

## 📝 Quiz #2: Sixteen Independent Shares {.quiz-question}

Sixteen independent shares each have s.d. 20%. What is the s.d. of the equally weighted portfolio?

- [$5\%$]{.correct data-explanation="✅ With zero covariances V = σ²/n = 400/16 = 25, so the s.d. is σ/√n = 20/4 = 5%."}
- $20\%$
- $1.25\%$
- $10\%$

---

## 📝 Quiz #3: A Default-Rate Estimate {.quiz-question}

A bank reviews $n = 400$ loans, each defaulting independently with $p = 0.05$, and estimates the rate by $\hat{p} = Y/n$. What is the s.d. of $\hat{p}$?

- [$0.0109$]{.correct data-explanation="✅ Example 5.28: V(p̂) = pq/n = (0.05)(0.95)/400 = 0.00011875, and its square root is 0.0109."}
- $0.00012$
- $0.2179$
- $0.0475$

---

## 📋 Key Formulas

::: {style="font-size: 28px"}
| | Statement |
|---|---|
| Theorem 5.12(a) | $E\left(\sum a_iY_i\right) = \sum a_i\mu_i$ |
| Theorem 5.12(b) | $V\left(\sum a_iY_i\right) = \sum a_i^2V(Y_i) + 2\sum\sum_{i<j}a_ia_j\text{Cov}(Y_i, Y_j)$ |
| Theorem 5.12(c) | $\text{Cov}\left(\sum a_iY_i, \sum b_jX_j\right) = \sum\sum a_ib_j\text{Cov}(Y_i, X_j)$ |
| two assets | $V(R_p) = w^2\sigma_1^2 + (1-w)^2\sigma_2^2 + 2w(1-w)\rho\sigma_1\sigma_2$ |
| $n$ equal weights | $V(R_p) = \sigma^2/n + (1 - 1/n)\rho\sigma^2$ |
| hedge | $h^* = -\text{Cov}(Y_1, Y_2)/V(Y_2)$, $\; V(H^*) = V(Y_1)(1 - \rho^2)$ |
| Examples 5.27, 5.28 | $V(\bar{Y}) = \sigma^2/n$, $\; V(\hat{p}) = pq/n$ |
:::

---

## 📋 Summary

::: {style="font-size: 30px"}
- The mean of a linear function needs only the means; the variance needs every covariance

- Constants enter the variance squared, and each pair's covariance enters twice

- Portfolio risk is below the weighted-average risk whenever $\rho < 1$

- Diversification drives specific risk to zero, but shared risk sets a floor of $\rho\sigma^2$

- A negatively correlated asset, held in the right amount, cuts variance by the factor $1 - \rho^2$

- $\bar{Y}$ and $\hat{p}$ are linear functions, which is how Chapters 9 and 11 will use this theorem
:::

---

## 📚 Practice Problems

::: {style="font-size: 28px"}
**Wackerly, 7th edition**

- §5.8: Exercises from 5.102 onwards; start with 5.102, 5.103, 5.110, 5.112 and 5.113

- Redo the pension fund with $\rho = -0.3$. What weight in energy minimises the risk?

**Week 14, Problem Set 1** is open now and closes **Sunday 20 December at 23:59** on WeBWorK, covering §5.8.

**Quiz II is on 12 December**, in the first 30 minutes of class: Chapter 4 and §§5.1–5.8, including today.

**Next class:** 12 December, after the quiz: the multinomial probability distribution (Wackerly §5.9).
:::

---

## 🙏 Thank You

::: {style="font-size: 34px"}
**Dr. Samir Orujov**

📧 sorujov@ada.edu.az\
🏢 Building D, Room D325\
🕓 Office hours: Wednesday, 16:00 – 18:00

Slides and readings: **sorujov.net/teaching**
:::

---

## ❓ Questions

::: {style="font-size: 32px"}
- At $\rho = -1$, which weights make the two-fund portfolio riskless? Is a riskless portfolio of risky assets a paradox?

- Doubling every weight doubles the expected return. What does it do to the variance?

- Why can no amount of diversification across Baku-listed shares remove the risk of an oil-price fall?
:::
