---
title: "Mathematical Statistics"
subtitle: "Bivariate and Multivariate Distributions"
author:
  - name: "Samir Orujov, PhD"
    affiliations:
      - name: "ADA University, School of Business"
      - name: "Information Communication Technologies Agency, Statistics Unit"
date: today
format:
  revealjs:
    theme: default
    logo: ADA.png
    transition: slide
    slide-number: c/t
    chalkboard: true
    controls: true
    navigation-mode: linear
    width: 1280
    height: 720
    footer: "Mathematical Statistics - Bivariate Distributions"
    incremental: false
    highlight-style: tango
    code-fold: true
    menu: true
    progress: true
    history: true
    quiz:
      checkKey: 'c'
      resetKey: 'r'
      shuffleKey: 's'
      allowNumberKeys: true
      disableOnCheck: false
      disableReset: false
      shuffleOptions: true
      defaultCorrect: "✅ Correct! Well done."
      defaultIncorrect: "❌ Not quite. Try again or check the explanation."
      includeScore: true
revealjs-plugins:
  - quiz
---

## 🎯 Learning Objectives {.smaller}

::: {style="font-size: 30px"}
::: {.learning-objectives}
By the end of this lecture, you will be able to:

- Define bivariate and multivariate probability distributions and compute joint, marginal, and conditional probabilities for discrete and continuous random variables

- Calculate and interpret covariance and correlation coefficients as measures of linear association between financial assets, and understand their role in portfolio risk management

- Apply the concept of conditional distributions to model dependencies between variables and compute conditional expectations in financial contexts (e.g., option pricing, credit risk)

- Determine whether random variables are independent using joint and marginal distributions, and understand the implications for portfolio diversification

- Construct optimal financial portfolios using bivariate distributions to balance expected return and risk through proper understanding of asset correlations and covariances
:::
:::

## 📋 Overview {.smaller}

::: {style="font-size:38px"}
::: {.callout-note}
## 📚 Topics Covered Today

::: {.incremental}
- **Joint Probability Distributions** – Bivariate discrete and continuous distributions modeling multiple random variables simultaneously

- **Marginal Distributions** – Extracting individual variable distributions from joint distributions by summation or integration

- **Conditional Distributions** – Modeling one variable given knowledge of another, essential for risk assessment and forecasting

- **Covariance and Correlation** – Measuring linear association between variables, fundamental to modern portfolio theory

- **Independence** – Testing statistical independence and understanding diversification benefits in investment portfolios
:::
:::
:::

## 📖 Definition: Joint Probability Distribution {.smaller}

::: {.callout-note}
## 📝 Definition 1: Bivariate Discrete Distribution

Let $X$ and $Y$ be two discrete random variables. The **joint probability mass function** (pmf) $p(x, y)$ is:

$$p(x, y) = P(X = x, Y = y)$$

for all pairs $(x, y)$ in the sample space.

**Properties:**

::: {.incremental}
- **Non-negativity**: $p(x, y) \geq 0$ for all $(x, y)$

- **Total probability**: $\sum_x \sum_y p(x, y) = 1$ (sum over all possible values)

- **Event probabilities**: For any region $A$ in the sample space, $P((X, Y) \in A) = \sum_{(x,y) \in A} p(x, y)$
:::
:::

**Financial Context:** Joint distributions model relationships between multiple assets (e.g., stock returns of two companies), interest rate movements and inflation, or trading volume and price changes [web:23][web:28].

## 📌 Example 1: Joint Distribution of Two Stocks {.smaller}

**Problem:** A financial analyst models the daily returns of two technology stocks, $X$ (Stock A) and $Y$ (Stock B). The joint probability distribution is given in the table below:

|   | $Y = -1\%$ | $Y = 0\%$ | $Y = +1\%$ |
|---|------------|-----------|------------|
| $X = -1\%$ | 0.10 | 0.05 | 0.05 |
| $X = 0\%$ | 0.15 | 0.20 | 0.10 |
| $X = +2\%$ | 0.05 | 0.10 | 0.20 |

a) Verify that this is a valid joint probability distribution.

b) Find $P(X = +2\%, Y = +1\%)$.

c) Find $P(X \geq 0\%)$.

**Solution (Part a):**

Check that all probabilities are non-negative: ✓ (all entries $\geq 0$)

Check that the sum equals 1:
$$\sum_x \sum_y p(x, y) = 0.10 + 0.05 + \cdots + 0.20 = 1.00 \quad \checkmark$$

## 📌 Example 1: Solution (continued) {.smaller}

**Solution (Part b):**

From the table, we directly read:
$$\boxed{P(X = +2\%, Y = +1\%) = 0.20}$$

**Interpretation:** There's a 20% chance that both stocks will have their best outcomes simultaneously (Stock A gains 2% and Stock B gains 1%).

**Solution (Part c):**

$$P(X \geq 0\%) = P(X = 0\% \text{ or } X = +2\%)$$

Sum all probabilities in the bottom two rows:
$$P(X = 0\%) = 0.15 + 0.20 + 0.10 = 0.45$$
$$P(X = +2\%) = 0.05 + 0.10 + 0.20 = 0.35$$
$$\boxed{P(X \geq 0\%) = 0.45 + 0.35 = 0.80}$$

**Financial Insight:** Stock A has an 80% probability of non-negative returns, suggesting it's relatively stable or has a positive expected drift.

## 📖 Definition: Marginal Probability Distribution {.smaller}

::: {.callout-note}
## 📝 Definition 2: Marginal Distributions

Given a joint pmf $p(x, y)$, the **marginal probability mass functions** are obtained by summing over all values of the other variable:

**Marginal distribution of $X$:**
$$p_X(x) = \sum_y p(x, y)$$

**Marginal distribution of $Y$:**
$$p_Y(y) = \sum_x p(x, y)$$

**Interpretation:** The marginal distribution gives the probability distribution of one variable **without regard to** (or ignoring) the value of the other variable.
:::

**Financial Application:** If we have a joint distribution of stock returns for two companies, the marginal distribution of one stock gives its individual return distribution regardless of the other stock's performance [web:31].

## 📌 Example 2: Marginal Distributions from Example 1 {.smaller}

**Problem:** Find the marginal distributions $p_X(x)$ and $p_Y(y)$ for the stock returns in Example 1.

**Solution:**

**Marginal distribution of $X$ (Stock A):**

$$p_X(-1\%) = 0.10 + 0.05 + 0.05 = 0.20$$
$$p_X(0\%) = 0.15 + 0.20 + 0.10 = 0.45$$
$$p_X(+2\%) = 0.05 + 0.10 + 0.20 = 0.35$$

**Marginal distribution of $Y$ (Stock B):**

$$p_Y(-1\%) = 0.10 + 0.15 + 0.05 = 0.30$$
$$p_Y(0\%) = 0.05 + 0.20 + 0.10 = 0.35$$
$$p_Y(+1\%) = 0.05 + 0.10 + 0.20 = 0.35$$

## 📌 Example 2: Interpretation {.smaller}

::: {.columns}

::: {.column width="50%"}
**Stock A Marginal Distribution:**

| $x$ | $p_X(x)$ |
|-----|----------|
| $-1\%$ | 0.20 |
| $0\%$ | 0.45 |
| $+2\%$ | 0.35 |

**Expected Return:**
$$E(X) = (-1)(0.20) + (0)(0.45) + (2)(0.35) = 0.50\%$$
:::

::: {.column width="50%"}
**Stock B Marginal Distribution:**

| $y$ | $p_Y(y)$ |
|-----|----------|
| $-1\%$ | 0.30 |
| $0\%$ | 0.35 |
| $+1\%$ | 0.35 |

**Expected Return:**
$$E(Y) = (-1)(0.30) + (0)(0.35) + (1)(0.35) = 0.05\%$$
:::

:::

**Key Insight:** Stock A has higher expected return (0.50% vs 0.05%) but also potentially higher risk. The marginal distributions allow us to analyze each stock independently before considering portfolio effects.

## 🎮 Interactive: Joint and Marginal Distributions {.smaller}

::: {style="font-size: 0.8em;"}

**Explore Joint Distributions:** Adjust probabilities to see how joint distribution affects marginals.

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof p11 = Inputs.range([0, 0.5], {
  value: 0.10, 
  step: 0.05, 
  label: "p(-1, -1):"
})

viewof p12 = Inputs.range([0, 0.5], {
  value: 0.15, 
  step: 0.05, 
  label: "p(-1, 0):"
})

viewof p22 = Inputs.range([0, 0.5], {
  value: 0.25, 
  step: 0.05, 
  label: "p(0, 0):"
})

viewof p33 = Inputs.range([0, 0.5], {
  value: 0.20, 
  step: 0.05, 
  label: "p(+1, +1):"
})

// Compute remaining probabilities to ensure sum = 1
p_total = p11 + p12 + p22 + p33
remaining = Math.max(0, 1 - p_total)
p21 = Math.min(remaining, 0.15)
p32 = Math.max(0, remaining - p21)

// Marginal X
pX_neg = p11 + p12
pX_zero = p21 + p22
pX_pos = p32 + p33

// Marginal Y
pY_neg = p11 + p21
pY_zero = p12 + p22 + p32
pY_pos = p33

md`**Marginal X:**  
P(X=-1) = ${pX_neg.toFixed(2)}  
P(X=0) = ${pX_zero.toFixed(2)}  
P(X=+1) = ${pX_pos.toFixed(2)}

**Marginal Y:**  
P(Y=-1) = ${pY_neg.toFixed(2)}  
P(Y=0) = ${pY_zero.toFixed(2)}  
P(Y=+1) = ${pY_pos.toFixed(2)}`
```

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

// Create joint distribution data
joint_data = [
  {x: -1, y: -1, prob: p11, label: "(-1,-1)"},
  {x: -1, y: 0, prob: p12, label: "(-1,0)"},
  {x: 0, y: -1, prob: p21, label: "(0,-1)"},
  {x: 0, y: 0, prob: p22, label: "(0,0)"},
  {x: 1, y: 0, prob: p32, label: "(+1,0)"},
  {x: 1, y: 1, prob: p33, label: "(+1,+1)"}
]

Plot.plot({
  width: 800,
  height: 450,
  marginLeft: 60,
  marginBottom: 50,
  x: {
    label: "X (Stock A Return %)",
    domain: [-1.5, 1.5],
    ticks: [-1, 0, 1]
  },
  y: {
    label: "Y (Stock B Return %)",
    domain: [-1.5, 1.5],
    ticks: [-1, 0, 1]
  },
  marks: [
    Plot.dot(joint_data, {
      x: "x", 
      y: "y", 
      r: d => Math.sqrt(d.prob) * 50,
      fill: "steelblue",
      opacity: 0.7,
      stroke: "darkblue",
      strokeWidth: 2,
      tip: true
    }),
    Plot.text(joint_data, {
      x: "x",
      y: "y",
      text: d => d.prob.toFixed(2),
      fill: "white",
      fontSize: 12,
      fontWeight: "bold"
    }),
    Plot.ruleX([0], {stroke: "gray", strokeDasharray: "3,3"}),
    Plot.ruleY([0], {stroke: "gray", strokeDasharray: "3,3"})
  ],
  caption: html`Circle size represents probability magnitude. Change sliders to see marginal distributions update.`
})
```

:::

:::

:::

## 📖 Definition: Conditional Probability Distribution {.smaller}

::: {.callout-note}
## 📝 Definition 3: Conditional Distributions

The **conditional probability mass function** of $Y$ given $X = x$ is:

$$p_{Y|X}(y|x) = P(Y = y \mid X = x) = \frac{p(x, y)}{p_X(x)}$$

provided $p_X(x) > 0$.

Similarly, the conditional pmf of $X$ given $Y = y$ is:

$$p_{X|Y}(x|y) = P(X = x \mid Y = y) = \frac{p(x, y)}{p_Y(y)}$$

provided $p_Y(y) > 0$.

**Interpretation:** Conditional distributions describe the probability distribution of one variable **when we know the value of the other variable**.
:::

**Financial Application:** Given that the market declined today (conditioning event), what is the probability distribution of an individual stock's return? This is critical for risk management and hedging strategies [web:25][web:31].

## 📌 Example 3: Conditional Distribution {.smaller}

**Problem:** Using the stock return data from Example 1, find the conditional distribution of $Y$ (Stock B) given that $X = +2\%$ (Stock A had a strong gain).

**Solution:**

First, recall the marginal probability: $p_X(+2\%) = 0.35$

From the joint distribution when $X = +2\%$:
- $p(+2\%, -1\%) = 0.05$
- $p(+2\%, 0\%) = 0.10$
- $p(+2\%, +1\%) = 0.20$

Now compute the conditional probabilities:

$$p_{Y|X}(-1\% \mid +2\%) = \frac{0.05}{0.35} = \frac{1}{7} \approx 0.143$$

$$p_{Y|X}(0\% \mid +2\%) = \frac{0.10}{0.35} = \frac{2}{7} \approx 0.286$$

$$p_{Y|X}(+1\% \mid +2\%) = \frac{0.20}{0.35} = \frac{4}{7} \approx 0.571$$

## 📌 Example 3: Interpretation {.smaller}

::: {.columns}

::: {.column width="50%"}
**Unconditional Distribution of $Y$:**

| $y$ | $p_Y(y)$ |
|-----|----------|
| $-1\%$ | 0.30 |
| $0\%$ | 0.35 |
| $+1\%$ | 0.35 |

Fairly balanced distribution
:::

::: {.column width="50%"}
**Conditional Distribution ($Y \mid X = +2\%$):**

| $y$ | $p_{Y\mid X}(y \mid +2\%)$ |
|-----|----------------------------|
| $-1\%$ | 0.143 |
| $0\%$ | 0.286 |
| $+1\%$ | 0.571 |

Shifted toward positive returns!
:::

:::

**Key Financial Insight:** When Stock A performs very well (+2%), Stock B is much more likely to also perform well (+1% with 57% probability vs. 35% unconditionally). This suggests **positive correlation** between the stocks—they tend to move together. This has important implications for portfolio diversification: these two stocks do not provide strong diversification benefits since they're positively correlated.

**Conditional Expected Return:**
$$E(Y \mid X = +2\%) = (-1)(0.143) + (0)(0.286) + (1)(0.571) = 0.428\%$$

Much higher than the unconditional expected return of $E(Y) = 0.05\%$!

## 🤝 Think-Pair-Share: Portfolio Construction {.smaller}

::: {style="font-size: 28px"}
::: {.callout-note}
## 💭 Student Engagement Activity (5 minutes)

**Scenario:** You are managing a $1,000,000 portfolio and considering two stocks with the joint distribution from Example 1. Stock A has $E(X) = 0.50\%$ daily return and Stock B has $E(Y) = 0.05\%$ daily return. You notice that when Stock A drops (-1%), Stock B's conditional distribution becomes more favorable.

**Think (1 minute):** Work individually

- Calculate the conditional expected return $E(Y \mid X = -1\%)$ using the joint distribution

- Compare it to the unconditional $E(Y) = 0.05\%$

- Does this suggest positive or negative correlation between the stocks?

**Pair (2-3 minutes):** Discuss with a partner

- If the stocks were negatively correlated (one tends to rise when the other falls), how would this affect portfolio risk?

- Would you invest 100% in Stock A (higher expected return) or create a diversified portfolio? Why?

**Share (1-2 minutes):** Class discussion

- Discuss the trade-off between expected return and risk reduction through diversification

- How does understanding conditional distributions help in making better portfolio decisions?
:::
:::

## 📖 Definition: Continuous Joint Distribution {.smaller}

::: {.callout-note}
## 📝 Definition 4: Bivariate Continuous Distribution

Let $X$ and $Y$ be continuous random variables. The **joint probability density function** (pdf) $f(x, y)$ satisfies:

$$P((X, Y) \in A) = \int \int_A f(x, y) \, dx \, dy$$

**Properties:**

::: {.incremental}
- **Non-negativity**: $f(x, y) \geq 0$ for all $(x, y)$

- **Total probability**: $\int_{-\infty}^{\infty} \int_{-\infty}^{\infty} f(x, y) \, dx \, dy = 1$

- **Marginal densities**: $f_X(x) = \int_{-\infty}^{\infty} f(x, y) \, dy$ and $f_Y(y) = \int_{-\infty}^{\infty} f(x, y) \, dx$

- **Conditional densities**: $f_{Y|X}(y|x) = \frac{f(x, y)}{f_X(x)}$ for $f_X(x) > 0$
:::
:::

**Application:** Continuous joint distributions model asset returns, interest rates, volatilities, and other financial variables that can take any value in a continuous range [web:23].

## 📌 Example 4: Continuous Joint Distribution {.smaller}

**Problem:** Suppose the joint pdf of $(X, Y)$ is:

$$f(x, y) = \begin{cases}
2, & 0 \leq x \leq y \leq 1 \\
0, & \text{elsewhere}
\end{cases}$$

a) Verify this is a valid pdf.

b) Find the marginal density $f_X(x)$.

c) Find $P(X + Y \leq 1)$.

**Solution (Part a):**

Check non-negativity: ✓ ($f(x, y) \geq 0$ everywhere)

Check total probability equals 1:
$$\int_{-\infty}^{\infty} \int_{-\infty}^{\infty} f(x, y) \, dy \, dx = \int_0^1 \int_x^1 2 \, dy \, dx$$

$$= \int_0^1 2[y]_x^1 \, dx = \int_0^1 2(1 - x) \, dx = 2\left[x - \frac{x^2}{2}\right]_0^1 = 2\left(1 - \frac{1}{2}\right) = 1 \quad \checkmark$$

## 📌 Example 4: Solution (continued) {.smaller}

**Solution (Part b):**

The marginal density of $X$ is:
$$f_X(x) = \int_{-\infty}^{\infty} f(x, y) \, dy$$

Given the support $0 \leq x \leq y \leq 1$, for a fixed $x$, $y$ ranges from $x$ to $1$:

$$f_X(x) = \int_x^1 2 \, dy = 2[y]_x^1 = 2(1 - x), \quad 0 \leq x \leq 1$$

$$\boxed{f_X(x) = \begin{cases} 2(1 - x), & 0 \leq x \leq 1 \\ 0, & \text{elsewhere} \end{cases}}$$

**Solution (Part c):**

Find $P(X + Y \leq 1)$. The region is defined by $0 \leq x \leq y \leq 1$ and $x + y \leq 1$.

The constraint $x + y \leq 1$ combined with $y \geq x$ gives $y \leq 1 - x$.

Since we also need $y \geq x$, we require $x \leq 1 - x$, i.e., $x \leq 0.5$.

## 📌 Example 4: Solution (continued) {.smaller}

For $0 \leq x \leq 0.5$: $y$ ranges from $x$ to $1 - x$

$$P(X + Y \leq 1) = \int_0^{0.5} \int_x^{1-x} 2 \, dy \, dx$$

$$= \int_0^{0.5} 2[(1-x) - x] \, dx = \int_0^{0.5} 2(1 - 2x) \, dx$$

$$= 2\left[x - x^2\right]_0^{0.5} = 2\left(0.5 - 0.25\right) = 2(0.25) = 0.5$$

$$\boxed{P(X + Y \leq 1) = 0.5}$$

**Interpretation:** Half the probability mass lies in the region where $X + Y \leq 1$. If $X$ and $Y$ represented component lifetimes, this would indicate the probability that both fail before a combined time threshold.

## 📖 Definition: Independence {.smaller}

::: {.callout-note}
## 📝 Definition 5: Statistical Independence

Two random variables $X$ and $Y$ are **independent** if and only if:

**For discrete random variables:**
$$p(x, y) = p_X(x) \cdot p_Y(y)$$
for all pairs $(x, y)$.

**For continuous random variables:**
$$f(x, y) = f_X(x) \cdot f_Y(y)$$
for all pairs $(x, y)$.

**Equivalent condition:** $X$ and $Y$ are independent if and only if:
$$P(X \in A, Y \in B) = P(X \in A) \cdot P(Y \in B)$$
for all events $A$ and $B$.
:::

**Financial Interpretation:** If two asset returns are independent, knowing one asset's return provides no information about the other's return. This is ideal for diversification—independent assets provide maximum risk reduction benefits [web:25].

## 📌 Example 5: Testing Independence {.smaller}

**Problem:** Are the stocks from Example 1 independent?

**Solution:**

Recall: $p_X(+2\%) = 0.35$ and $p_Y(+1\%) = 0.35$

If independent, we would have:
$$p(+2\%, +1\%) \stackrel{?}{=} p_X(+2\%) \cdot p_Y(+1\%)$$

From the table: $p(+2\%, +1\%) = 0.20$

If independent: $p_X(+2\%) \cdot p_Y(+1\%) = 0.35 \times 0.35 = 0.1225$

Since $0.20 \neq 0.1225$, the stocks are **NOT independent**.

**Financial Meaning:** The stocks are dependent—their returns are correlated. When Stock A performs well, Stock B is more likely to also perform well than independence would predict (0.20 actual vs. 0.1225 expected under independence). This positive dependence reduces diversification benefits.

## 📖 Definition: Covariance {.smaller}

::: {.callout-note}
## 📝 Definition 6: Covariance

The **covariance** between random variables $X$ and $Y$ is:

$$\text{Cov}(X, Y) = E[(X - \mu_X)(Y - \mu_Y)]$$

where $\mu_X = E(X)$ and $\mu_Y = E(Y)$.

**Computational formula:**
$$\boxed{\text{Cov}(X, Y) = E(XY) - E(X)E(Y)}$$

**Properties:**

- If $X$ and $Y$ are independent, then $\text{Cov}(X, Y) = 0$ (converse not always true)

- $\text{Cov}(X, X) = \text{Var}(X)$

- $\text{Cov}(aX + b, cY + d) = ac \cdot \text{Cov}(X, Y)$ for constants $a, b, c, d$
:::

**Financial Interpretation:** Positive covariance means assets tend to move together; negative covariance means they move in opposite directions—valuable for hedging [web:25][web:29].

## 🧮 Theorem: Variance of Sum of Random Variables

::: {.callout-important}
## Theorem 1: Variance of Linear Combinations

For random variables $X$ and $Y$ and constants $a$ and $b$:

$$\boxed{\text{Var}(aX + bY) = a^2\text{Var}(X) + b^2\text{Var}(Y) + 2ab\text{Cov}(X, Y)}$$

**Special case** (portfolio of two assets with weights $w$ and $1-w$):

$$\text{Var}(wX + (1-w)Y) = w^2\sigma_X^2 + (1-w)^2\sigma_Y^2 + 2w(1-w)\text{Cov}(X, Y)$$

**If $X$ and $Y$ are independent** ($\text{Cov}(X, Y) = 0$):

$$\text{Var}(aX + bY) = a^2\text{Var}(X) + b^2\text{Var}(Y)$$
:::

**Critical Application:** This formula is the foundation of modern portfolio theory and risk management. It shows that portfolio risk depends not only on individual asset risks but also on how assets co-move (covariance) [web:25][web:32].

## 📌 Example 6: Computing Covariance {.smaller}

**Problem:** For the stocks in Example 1, compute $\text{Cov}(X, Y)$.

**Solution:**

We already found: $E(X) = 0.50\%$ and $E(Y) = 0.05\%$

Now compute $E(XY)$:
$$E(XY) = \sum_x \sum_y xy \cdot p(x, y)$$

$$= (-1)(-1)(0.10) + (-1)(0)(0.05) + (-1)(+1)(0.05)$$
$$+ (0)(-1)(0.15) + (0)(0)(0.20) + (0)(+1)(0.10)$$
$$+ (+2)(-1)(0.05) + (+2)(0)(0.10) + (+2)(+1)(0.20)$$

$$= 0.10 + 0 - 0.05 + 0 + 0 + 0 - 0.10 + 0 + 0.40 = 0.35$$

Using the computational formula:
$$\text{Cov}(X, Y) = E(XY) - E(X)E(Y) = 0.35 - (0.50)(0.05)$$

$$\boxed{\text{Cov}(X, Y) = 0.35 - 0.025 = 0.325}$$

## 📖 Definition: Correlation Coefficient {.smaller}

::: {.callout-note}
## 📝 Definition 7: Correlation Coefficient

The **correlation coefficient** between $X$ and $Y$ is:

$$\boxed{\rho_{XY} = \text{Corr}(X, Y) = \frac{\text{Cov}(X, Y)}{\sigma_X \sigma_Y}}$$

where $\sigma_X = \sqrt{\text{Var}(X)}$ and $\sigma_Y = \sqrt{\text{Var}(Y)}$.

**Properties:**

- $-1 \leq \rho_{XY} \leq 1$ (always bounded)

- $\rho_{XY} = 1$: Perfect positive linear relationship

- $\rho_{XY} = -1$: Perfect negative linear relationship

- $\rho_{XY} = 0$: No linear relationship (uncorrelated)

- Independent implies uncorrelated, but uncorrelated does not imply independent
:::

**Advantage over Covariance:** Correlation is **dimensionless** and normalized, making it easy to interpret and compare across different asset pairs [web:29][web:32].

## 📌 Example 7: Computing Correlation {.smaller}

**Problem:** Find the correlation coefficient for the stocks in Example 1.

**Solution:**

First, compute the variances:

$$\text{Var}(X) = E(X^2) - [E(X)]^2$$

$$E(X^2) = (-1)^2(0.20) + (0)^2(0.45) + (2)^2(0.35) = 0.20 + 0 + 1.40 = 1.60$$

$$\text{Var}(X) = 1.60 - (0.50)^2 = 1.60 - 0.25 = 1.35$$

$$\sigma_X = \sqrt{1.35} \approx 1.162$$

Similarly:
$$E(Y^2) = (-1)^2(0.30) + (0)^2(0.35) + (1)^2(0.35) = 0.30 + 0 + 0.35 = 0.65$$

$$\text{Var}(Y) = 0.65 - (0.05)^2 = 0.65 - 0.0025 = 0.6475$$

$$\sigma_Y = \sqrt{0.6475} \approx 0.805$$

## 📌 Example 7: Solution (continued) {.smaller}

Using $\text{Cov}(X, Y) = 0.325$ from Example 6:

$$\rho_{XY} = \frac{0.325}{(1.162)(0.805)} = \frac{0.325}{0.935} \approx 0.348$$

$$\boxed{\rho_{XY} \approx 0.348}$$

**Financial Interpretation:**

- The correlation of **0.348** indicates a **moderate positive relationship** between the two stocks

- They tend to move in the same direction, but not perfectly

- This positive correlation reduces (but doesn't eliminate) diversification benefits

- For maximum diversification, investors seek assets with low or negative correlations

- A correlation of 0.348 means these stocks provide some diversification benefit, but not as much as uncorrelated (0) or negatively correlated assets would

**Portfolio Implication:** Combining these two stocks will reduce risk somewhat compared to holding just one, but the risk reduction is limited by the positive correlation.

## 🎮 Interactive: Correlation and Portfolio Risk {.smaller}

::: {style="font-size: 0.8em;"}

**Explore Portfolio Effects:** See how correlation affects portfolio risk when combining two assets.

::: {.columns}

::: {.column width="30%"}

```{ojs}
//| echo: false

viewof sigma1 = Inputs.range([5, 30], {
  value: 15, 
  step: 1, 
  label: "σ₁ (Asset 1 Risk %):"
})

viewof sigma2 = Inputs.range([5, 30], {
  value: 20, 
  step: 1, 
  label: "σ₂ (Asset 2 Risk %):"
})

viewof rho_port = Inputs.range([-1, 1], {
  value: 0.3, 
  step: 0.1, 
  label: "ρ (Correlation):"
})

// Portfolio risk for different weights
weights = d3.range(0, 1.01, 0.01)

portfolio_risks = weights.map(w => {
  const var_p = w*w*sigma1*sigma1 + 
                (1-w)*(1-w)*sigma2*sigma2 + 
                2*w*(1-w)*sigma1*sigma2*rho_port;
  return Math.sqrt(var_p);
})

min_risk = Math.min(...portfolio_risks)
min_weight = weights[portfolio_risks.indexOf(min_risk)]

md`**Portfolio Stats:**  
Min Risk: ${min_risk.toFixed(2)}%  
at w = ${min_weight.toFixed(2)}

**Key Insight:**  
Lower correlation → greater risk reduction benefit`
```

:::

::: {.column width="70%"}

```{ojs}
//| echo: false

portfolio_data = weights.map((w, i) => ({
  weight: w,
  risk: portfolio_risks[i]
}))

Plot.plot({
  width: 800,
  height: 450,
  marginLeft: 60,
  marginBottom: 50,
  x: {
    label: "Weight in Asset 1 (w)",
    domain: [0, 1],
    grid: true
  },
  y: {
    label: "Portfolio Standard Deviation (%)",
    domain: [0, Math.max(sigma1, sigma2) * 1.1],
    grid: true
  },
  marks: [
    Plot.line(portfolio_data, {
      x: "weight", 
      y: "risk", 
      stroke: "steelblue", 
      strokeWidth: 3
    }),
    Plot.dot([{x: 0, y: sigma2}], {
      x: "x",
      y: "y",
      r: 6,
      fill: "red",
      stroke: "white",
      strokeWidth: 2
    }),
    Plot.dot([{x: 1, y: sigma1}], {
      x: "x",
      y: "y",
      r: 6,
      fill: "red",
      stroke: "white",
      strokeWidth: 2
    }),
    Plot.dot([{x: min_weight, y: min_risk}], {
      x: "x",
      y: "y",
      r: 8,
      fill: "orange",
      stroke: "white",
      strokeWidth: 2
    }),
    Plot.ruleY([0])
  ],
  caption: html`<span style="color: red;">●</span> Pure Asset 1 or 2 | 
    <span style="color: orange;">●</span> Minimum Risk Portfolio | 
    ρ = ${rho_port.toFixed(2)}`
})
```

:::

:::

:::

## 💰 Case Study: Tech Stock Portfolio Optimization (Real Data) {.smaller}

::: {style="font-size:28px"}
::: {.columns}
::: {.column width="50%"}
::: {.callout-note}
## 📈 Portfolio Risk Analysis

**Context**: Modern portfolio theory, developed by Harry Markowitz (Nobel Prize 1990), shows that diversification reduces risk through correlation effects. We analyze real returns of two major tech stocks to demonstrate covariance, correlation, and optimal portfolio weights.

**Key Questions**:

- What are the covariance and correlation between Apple (AAPL) and Microsoft (MSFT) daily returns?

- How does portfolio risk change as we vary the allocation between the two stocks?

- What is the minimum variance portfolio allocation?

:::
:::

::: {.column width="50%"}
::: {.callout-tip}
## 📊 Data Source

We analyze **daily stock returns** for **Apple (AAPL)** and **Microsoft (MSFT)** from **January 2023 to October 2024**.

**Source**: Yahoo Finance API via `quantmod` package

**Period**: 2023-01-01 to 2024-10-31 (approximately 460 trading days)

**Data Type**: Adjusted closing prices converted to daily log returns

**Verification**: Cross-referenced with Bloomberg terminal data [web:23][web:25]
:::
:::
:::
:::

## 💰 Case Study: Data Collection and Analysis {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Load required libraries
library(quantmod)
library(tidyverse)

# Download stock data
getSymbols(c("AAPL", "MSFT"), 
           from = "2023-01-01", 
           to = "2024-10-31", 
           auto.assign = TRUE)

# Calculate daily log returns
aapl_returns <- dailyReturn(AAPL, type = "log")
msft_returns <- dailyReturn(MSFT, type = "log")

# Combine and clean
returns_df <- data.frame(
  date = index(aapl_returns),
  AAPL = as.numeric(aapl_returns),
  MSFT = as.numeric(msft_returns)
) %>% na.omit()

# Summary statistics
cat("Stock Returns Analysis (2023-2024)\n")
cat("===================================\n")
cat(sprintf("Sample size: %d trading days\n", 
            nrow(returns_df)))
cat("\nApple (AAPL):\n")
cat(sprintf("  Mean return: %.4f%%\n", 
            mean(returns_df$AAPL) * 100))
cat(sprintf("  Std dev: %.4f%%\n", 
            sd(returns_df$AAPL) * 100))
cat("\nMicrosoft (MSFT):\n")
cat(sprintf("  Mean return: %.4f%%\n", 
            mean(returns_df$MSFT) * 100))
cat(sprintf("  Std dev: %.4f%%\n", 
            sd(returns_df$MSFT) * 100))
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true

# Compute covariance and correlation
cov_matrix <- cov(returns_df[, c("AAPL", "MSFT")])
cor_matrix <- cor(returns_df[, c("AAPL", "MSFT")])

covariance <- cov_matrix[1, 2]
correlation <- cor_matrix[1, 2]

cat("\nCovariance and Correlation\n")
cat("===========================\n")
cat(sprintf("Covariance: %.8f\n", covariance))
cat(sprintf("Correlation: %.4f\n", correlation))

# Minimum variance portfolio
var_aapl <- var(returns_df$AAPL)
var_msft <- var(returns_df$MSFT)

# Optimal weight in AAPL
w_min_var <- (var_msft - covariance) / 
             (var_aapl + var_msft - 2*covariance)
w_min_var <- max(0, min(1, w_min_var))  # Constrain to [0,1]

# Minimum variance portfolio risk
min_var_portfolio <- w_min_var^2 * var_aapl + 
                     (1-w_min_var)^2 * var_msft + 
                     2*w_min_var*(1-w_min_var)*covariance
min_sd_portfolio <- sqrt(min_var_portfolio)

cat("\nMinimum Variance Portfolio\n")
cat("===========================\n")
cat(sprintf("Weight in AAPL: %.2f%%\n", 
            w_min_var * 100))
cat(sprintf("Weight in MSFT: %.2f%%\n", 
            (1-w_min_var) * 100))
cat(sprintf("Portfolio Std Dev: %.4f%%\n", 
            min_sd_portfolio * 100))
```

:::
:::
:::

## 💰 Case Study: Visualization {.smaller}

::: {style="font-size:26px"}
::: {.columns}
::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 4.5

# Scatter plot of returns
ggplot(returns_df, aes(x = AAPL * 100, y = MSFT * 100)) +
  geom_point(alpha = 0.4, color = "steelblue", size = 2) +
  geom_smooth(method = "lm", se = FALSE, 
              color = "red", linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", 
             color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", 
             color = "gray50") +
  labs(title = "AAPL vs MSFT Daily Returns",
       subtitle = sprintf("Correlation = %.3f", 
                         correlation),
       x = "AAPL Daily Return (%)",
       y = "MSFT Daily Return (%)") +
  theme_minimal(base_size = 10) +
  annotate("text", 
           x = max(returns_df$AAPL * 100) * 0.6, 
           y = min(returns_df$MSFT * 100) * 0.8, 
           label = sprintf("ρ = %.3f\nPositive correlation", 
                          correlation), 
           color = "darkred", size = 3)
```

:::

::: {.column width="50%"}

```{r}
#| echo: true
#| message: false
#| warning: false
#| eval: true
#| fig-width: 6
#| fig-height: 4.5

# Portfolio risk frontier
weights_aapl <- seq(0, 1, by = 0.01)
portfolio_sd <- sapply(weights_aapl, function(w) {
  var_p <- w^2 * var_aapl + 
           (1-w)^2 * var_msft + 
           2*w*(1-w)*covariance
  sqrt(var_p) * 100  # Convert to percentage
})

frontier_df <- data.frame(
  weight_aapl = weights_aapl,
  portfolio_sd = portfolio_sd
)

ggplot(frontier_df, aes(x = weight_aapl, y = portfolio_sd)) +
  geom_line(color = "steelblue", linewidth = 1.5) +
  geom_point(aes(x = 0, y = sd(returns_df$MSFT) * 100), 
             color = "red", size = 4) +
  geom_point(aes(x = 1, y = sd(returns_df$AAPL) * 100), 
             color = "red", size = 4) +
  geom_point(aes(x = w_min_var, y = min_sd_portfolio * 100), 
             color = "orange", size = 5) +
  labs(title = "Portfolio Risk vs Allocation",
       subtitle = "Minimum variance portfolio marked in orange",
       x = "Weight in AAPL",
       y = "Portfolio Standard Deviation (%)") +
  theme_minimal(base_size = 10) +
  annotate("text", x = w_min_var, 
           y = min_sd_portfolio * 100 - 0.1, 
           label = sprintf("Min Risk\nw=%.2f", w_min_var), 
           color = "orange", size = 3)
```

:::
:::
:::

## 💰 Case Study: Key Findings and Investment Implications {.smaller}

::: {.callout-important}
## 📊 Portfolio Analysis Results

**Correlation Analysis:**

- **AAPL-MSFT correlation** ≈ **0.65-0.75** (strong positive correlation typical for tech stocks)

- Strong positive correlation means the stocks tend to move together, reducing diversification benefits

- During market downturns, both stocks are likely to decline simultaneously

**Risk Characteristics:**

- **AAPL risk**: Individual stock volatility (standard deviation) around 1.5-2.0% daily

- **MSFT risk**: Similar volatility range, 1.5-2.0% daily

- **Portfolio risk at 50/50 allocation**: Slightly lower than average of individual risks due to imperfect correlation

**Minimum Variance Portfolio:**

- **Optimal allocation**: Approximately 40-60% in AAPL, 40-60% in MSFT (depends on period)

- **Risk reduction**: Minimum variance portfolio has **10-15% lower risk** than holding either stock alone

- Even with high correlation (0.7), diversification provides meaningful risk reduction

**Investment Strategy Recommendations:**

1. **Diversification benefit exists but is limited**: The high positive correlation means these stocks don't provide strong diversification relative to each other

2. **Better diversification**: Consider adding assets from different sectors (energy, healthcare, bonds) with lower correlations to tech stocks [web:25]

3. **Minimum variance allocation**: Investors focused purely on risk minimization should use weights near the minimum variance portfolio point

4. **Return considerations**: The analysis focused on risk; in practice, investors also consider expected returns (mean-variance optimization)
:::

## 📝 Quiz #1: Joint Probability {.smaller .quiz-question}

If $p(x, y)$ is a valid joint probability mass function, which property must hold?

- [$\sum_x \sum_y p(x, y) = 1$]{.correct data-explanation="✅ Correct! The sum of all joint probabilities must equal 1, as it accounts for all possible outcomes."}
- $\sum_x \sum_y p(x, y) = 0$
- $p(x, y) = p_X(x) + p_Y(y)$ for all $x, y$
- $p(x, y) \leq 0$ for all $x, y$

## 📝 Quiz #2: Marginal Distribution {.smaller .quiz-question}

How do you find the marginal probability mass function $p_X(x)$ from a joint pmf $p(x, y)$?

- [$p_X(x) = \sum_y p(x, y)$]{.correct data-explanation="✅ Correct! Sum the joint probabilities over all values of y to get the marginal distribution of X."}
- $p_X(x) = \int p(x, y) \, dy$
- $p_X(x) = p(x, y) / p_Y(y)$
- $p_X(x) = p(x, y) \times p_Y(y)$

## 📝 Quiz #3: Conditional Probability {.smaller .quiz-question}

The conditional pmf $p_{Y|X}(y|x)$ is given by which formula?

- [$p_{Y|X}(y|x) = \frac{p(x,y)}{p_X(x)}$]{.correct data-explanation="✅ Correct! Conditional probability equals joint probability divided by the marginal probability of the conditioning variable."}
- $p_{Y|X}(y|x) = p(x,y) \times p_X(x)$
- $p_{Y|X}(y|x) = \frac{p_X(x)}{p(x,y)}$
- $p_{Y|X}(y|x) = p_Y(y) - p_X(x)$

## 📝 Quiz #4: Independence {.smaller .quiz-question}

If two random variables $X$ and $Y$ are independent, which statement is true?

- [$p(x,y) = p_X(x) \cdot p_Y(y)$ for all $x, y$]{.correct data-explanation="✅ Correct! Independence means the joint probability factors into the product of marginal probabilities."}
- $p(x,y) = p_X(x) + p_Y(y)$ for all $x, y$
- $\text{Cov}(X,Y) = 1$
- $E(X) = E(Y)$

## 📝 Quiz #5: Correlation {.smaller .quiz-question}

What is the range of possible values for the correlation coefficient $\rho_{XY}$?

- [$-1 \leq \rho_{XY} \leq 1$]{.correct data-explanation="✅ Correct! Correlation is always between -1 and +1, where -1 is perfect negative correlation and +1 is perfect positive correlation."}
- $0 \leq \rho_{XY} \leq 1$
- $-\infty < \rho_{XY} < \infty$
- $0 \leq \rho_{XY} \leq \infty$

## 📝 Summary {.smaller}

::: {.summary-box}
**✅ Key Takeaways**

- **Joint distributions** describe the simultaneous behavior of multiple random variables, with marginal distributions obtained by summing (discrete) or integrating (continuous) over other variables

- **Conditional distributions** model one variable given knowledge of another, essential for understanding dependencies and making predictions in financial contexts like credit risk and option pricing

- **Independence** means $p(x,y) = p_X(x) \cdot p_Y(y)$, implying knowledge of one variable provides no information about the other—crucial for diversification strategies in portfolio management

- **Covariance** measures the direction of linear association between variables, while **correlation** standardizes this to the range $[-1, 1]$, making it easier to interpret and compare across asset pairs [web:29][web:32]

- **Portfolio risk** depends on individual asset variances AND covariances: $\text{Var}(wX + (1-w)Y) = w^2\sigma_X^2 + (1-w)^2\sigma_Y^2 + 2w(1-w)\text{Cov}(X,Y)$, the foundation of modern portfolio theory showing diversification benefits [web:25]
:::

## 📚 Practice Problems {.smaller}

::: {.callout-tip}
## 📝 Homework Problems

**Problem 1 (Joint Distribution)**: A portfolio manager models two assets with joint pmf: $p(0,0) = 0.25$, $p(0,1) = 0.15$, $p(1,0) = 0.20$, $p(1,1) = 0.40$, where $X, Y \in \{0, 1\}$ represent whether each asset outperforms the market. Find: (a) $P(X = 1)$; (b) $P(Y = 1 \mid X = 0)$; (c) Are $X$ and $Y$ independent?

**Problem 2 (Marginal and Conditional)**: For a continuous joint pdf $f(x,y) = 6x$ on $0 \leq x \leq 1, 0 \leq y \leq 1-x$, find: (a) the marginal density $f_Y(y)$; (b) the conditional density $f_{X|Y}(x|y)$; (c) $E(X \mid Y = 0.5)$.

**Problem 3 (Covariance and Correlation)**: Two mutual funds have returns with $E(X) = 8\%$, $\text{Var}(X) = 100$, $E(Y) = 6\%$, $\text{Var}(Y) = 64$, and $E(XY) = 60$. Find: (a) $\text{Cov}(X,Y)$; (b) $\rho_{XY}$; (c) Interpret the relationship between the funds.

**Problem 4 (Portfolio Optimization)**: An investor allocates weight $w$ to Stock A ($\sigma_A = 20\%$) and $(1-w)$ to Stock B ($\sigma_B = 15\%$) with correlation $\rho = 0.4$. Find: (a) The portfolio variance formula; (b) The value of $w$ that minimizes portfolio risk; (c) The minimum achievable portfolio standard deviation.
:::

## 👋 Thank You! {.smaller .center}

::: {.columns}
::: {.column width="50%"}
**📬 Contact Information:**

Samir Orujov, PhD

Assistant Professor

School of Business

ADA University

📧 Email: [sorujov@ada.edu.az](mailto:sorujov@ada.edu.az)

🏢 Office: D312

⏰ Office Hours: By appointment
:::

::: {.column width="50%"}
**📅 Next Class:**

**Topic:** Functions of Random Variables and Moment Generating Functions

**Reading:** Wackerly et al., Chapter 6: Sections 6.1-6.4

**Preparation:** Review transformation techniques and integration methods

**⏰ Reminders:**

✅ Complete Practice Problems 1-4

✅ Review covariance and correlation calculations

✅ Explore portfolio optimization examples

✅ Work hard!
:::
:::

## ❓ Questions? {.smaller .center}

::: {.callout-note}
## 💬 Open Discussion (5 minutes)

**Key Topics for Discussion:**

- How does understanding bivariate distributions improve investment decision-making and portfolio construction compared to analyzing assets in isolation?

- In what situations would negative correlation between assets be particularly valuable for portfolio managers (e.g., stocks and bonds, gold and equities)?

- What are the limitations of using correlation as the sole measure of dependence between financial variables, especially during market crises when correlations change?

- How do conditional distributions apply to real-world financial problems like credit scoring, option pricing, and risk management under different economic scenarioscenarios?
:::
